### 2024-02-24 index.qmd
initialize quarto manuscript file

---
title: "Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer"
authors:
  - name: Roman Ludwig
    orcid: 0000-0001-9434-328X
    email: roman.ludwig@usz.ch
  - name: Yoel Perez Haas
    email: yoel.perezhaas@usz.ch
  - name: Jan Unkelbach
    orcid: 0000-0002-4275-990X
    email: jan.unkelbach@usz.ch
format:
  html:
    code-fold: true
jupyter: python3
---

# Introduction

- head and neck cancer spreads lymphatically
- may sometimes spread to contralateral side
- spreads more frequently and to larger extent contralaterally when tumor extends the mid-sagittal line
- we describe a model based on previously published hidden Markov model
- we extend it to cover the contralateral side, too
- ipsi- and contralateral side are correlated via time of diagnosis, which is correlated with T-category
- tumor extension over mid-sagittal line is modelled as random variable
- spread probabilities from tumor to contralateral LNLs in case of midline extesnion are linear combinations of these probabilities in case of ipsilateral spread and contralateral spread when tumor is clearly lateralized

```{python}
from lymph import models

graph_dict = {
  ("tumor", "T"): ["II", "III"],
  ("lnl", "II"): ["III"],
  ("lnl", "III"): [],
}
model = models.Midline(
  graph_dict=graph_dict,
  is_symmetric={"lnl_spread": True},
  use_mixing=True,
  use_central=False,
  use_midext_evo=True,
)
model.set_modality("max_llh", spec=1.0, sens=1.0)
```

### 2024-02-25 index.qmd
set up simple sampling script

@@
 
+# Methods
+
+Prepare the model and load the data:
+
 ```{python}
+import numpy as np
+import pandas as pd
+
 from lymph import models
+from lyscripts.utils import binom_pmf, parametric_binom_pmf
+
+np.random.seed(42)
 
@@
 }
-model = models.Midline(
+model = models.Bilateral(
   graph_dict=graph_dict,
   is_symmetric={"lnl_spread": True},
-  use_mixing=True,
-  use_central=False,
-  use_midext_evo=True,
+  # use_mixing=True,
+  # use_central=False,
+  # use_midext_evo=True,
 )
 model.set_modality("max_llh", spec=1.0, sens=1.0)
+model.set_distribution(
+  "early",
+  binom_pmf(np.arange(model.max_time+1), model.max_time, p=0.3),
+)
+model.set_distribution("late", parametric_binom_pmf)
+model.load_patient_data(
+  pd.read_csv("data/enhanced.csv", header=[0,1,2]),
+)
 ```
+
+Do the sampling:
+
+```{python}
+from emcee import EnsembleSampler, backends, moves
+
+ndim = model.get_num_dims()
+initial = np.random.uniform(size=(20 * ndim, ndim))
+
+backend = backends.HDFBackend(filename="models/test.hdf5")
+moves = [(moves.DEMove(), 0.8), (moves.DESnookerMove(), 0.2)]
+sampler = EnsembleSampler(
+  nwalkers=20 * ndim,
+  ndim=ndim,
+  log_prob_fn=model.likelihood,
+  backend=backend,
+  moves=moves,
+  parameter_names=list(model.get_params().keys()),
+)
+sampler.run_mcmc(initial, 1000, progress=True)
+```
+
+Look at the results:
+
+```{python}
+from corner import corner
+
+samples = sampler.get_chain(discard=500, thin=10, flat=True)
+corner(
+  samples,
+  titles=list(model.get_params().keys()),
+  show_titles=True,
+);
+```
+
+NOTE: For the `Bilateral` model the results look absolutely plausible, but the `Midline` model seems to have some issues (`late_p` was learned to be 0.17 !?).

### 2024-02-26 index.qmd
switch to 2021 CLB data

--- 2024-02-24 index.qmd
@@
 
+# Methods
+
+Prepare the model and load the data:
+
 ```{python}
+import numpy as np
+import pandas as pd
+from emcee import EnsembleSampler, backends, moves
+from corner import corner
+
 from lymph import models
+from lyscripts.utils import binom_pmf, parametric_binom_pmf
 
+np.random.seed(42)
+```
+
+```{python}
 graph_dict = {
@@
 model.set_modality("max_llh", spec=1.0, sens=1.0)
+model.set_distribution(
+  "early",
+  binom_pmf(np.arange(model.max_time+1), model.max_time, p=0.3),
+)
+model.set_distribution("late", parametric_binom_pmf)
+model.load_patient_data(
+  pd.read_csv("data/enhanced.csv", header=[0,1,2]),
+)
 ```
+
+Do the sampling:
+
+```{python}
+ndim = model.get_num_dims()
+initial = np.random.uniform(size=(20 * ndim, ndim))
+
+backend = backends.HDFBackend(filename="models/midline.hdf5")
+moves_mix = [(moves.DEMove(), 0.8), (moves.DESnookerMove(), 0.2)]
+sampler = EnsembleSampler(
+  nwalkers=20 * ndim,
+  ndim=ndim,
+  log_prob_fn=model.likelihood,
+  backend=backend,
+  moves=moves_mix,
+  parameter_names=list(model.get_params().keys())[:ndim],
+)
+sampler.run_mcmc(initial, 1000, progress=True)
+```
+
+Look at the results:
+
+```{python}
+backend = backends.HDFBackend(
+  filename="models/midline.hdf5",
+  read_only=True,
+)
+
+samples = backend.get_chain(discard=500, thin=10, flat=True)
+corner(
+  samples,
+  titles=list(model.get_params().keys()),
+  show_titles=True,
+);
+```
+
+```{python}
+mean_samples = {
+  k: s.mean() for k, s in zip(model.get_params(), samples.T)
+}
+model.set_params(**mean_samples)
+```
+
+NOTE: For the `Bilateral` model the results look absolutely plausible, but the `Midline` model seems to have some issues (`late_p` was learned to be 0.17 !?).

### 2024-02-26 index.qmd
first successful run of midline evo (fixed model)

--- 2024-02-24 index.qmd
@@
 jupyter: python3
+execute: 
+  cache: true
 ---
@@
 
+# Methods
+
+# Results
+
 ```{python}
-from lymph import models
+from emcee import backends
+from corner import corner
+import matplotlib.pyplot as plt
+from tueplots import figsizes, fontsizes
 
-graph_dict = {
-  ("tumor", "T"): ["II", "III"],
-  ("lnl", "II"): ["III"],
-  ("lnl", "III"): [],
-}
-model = models.Midline(
-  graph_dict=graph_dict,
-  is_symmetric={"lnl_spread": True},
-  use_mixing=True,
-  use_central=False,
-  use_midext_evo=True,
+backend = backends.HDFBackend(
+  filename="models/midline_fix.hdf5",
+  read_only=True,
 )
-model.set_modality("max_llh", spec=1.0, sens=1.0)
+samples = backend.get_chain(thin=20)
+ndim = samples.shape[-1]
+
+plt.rcParams.update(**figsizes.icml2022_full(nrows=ndim))
+
+fig, axes = plt.subplots(nrows=ndim, sharex=True)
+for i, ax in enumerate(axes):
+  ax.plot(samples[:,::10,i], color="gray", alpha=0.2)
+  ax.axhline(samples[200:,:,i].mean(), color="red")
+  ax.set_xlim(0., samples.shape[0])
+  ax.set_ylim(0., 1.)
 ```

### 2024-02-26 scripts/sampling.py

@@ -1,0 +1,77 @@
+"""
+Implement the sampling in this module.
+"""
+import numpy as np
+import pandas as pd
+from emcee import EnsembleSampler, backends, moves
+
+from lymph import models
+
+
+from scipy.special import factorial
+
+def binom_pmf(k: np.ndarray, n: int, p: float):
+    """Binomial PMF"""
+    if p > 1. or p < 0.:
+        # This value error is important to enable seamless sampling!
+        raise ValueError("Binomial prob must be btw. 0 and 1")
+    q = 1. - p
+    binom_coeff = factorial(n) / (factorial(k) * factorial(n - k))
+    return binom_coeff * p**k * q**(n - k)
+
+def late_binomial(support: np.ndarray, p: float = 0.5) -> np.ndarray:
+    """Parametrized binomial distribution."""
+    return binom_pmf(k=support, n=support[-1], p=p)
+
+
+def create_model() -> models.Midline:
+    """Create the model."""
+    graph_dict = {
+        ("tumor", "T"): ["II", "III"],
+        ("lnl", "II"): ["III"],
+        ("lnl", "III"): [],
+    }
+    model = models.Midline(
+        graph_dict=graph_dict,
+        is_symmetric={"lnl_spread": True},
+        use_mixing=True,
+        use_central=False,
+        use_midext_evo=False,
+    )
+    model.set_modality("max_llh", spec=1.0, sens=1.0)
+    frozen_binom_pmf = binom_pmf(np.arange(model.max_time+1), model.max_time, p=0.3)
+    model.set_distribution("early", frozen_binom_pmf)
+    model.set_distribution("late", late_binomial)
+    model.load_patient_data(pd.read_csv("data/enhanced.csv", header=[0,1,2]))
+
+    return model
+
+
+def run_sampling(model: models.Midline, nstep: int = 5000) -> None:
+    """Run the sampling."""
+    ndim = model.get_num_dims()
+    initial = np.random.uniform(size=(20 * ndim, ndim))
+
+    suffix = "evo" if model.use_midext_evo else "fix"
+    backend = backends.HDFBackend(filename=f"models/midline_{suffix}.hdf5")
+    moves_mix = [(moves.DEMove(), 0.8), (moves.DESnookerMove(), 0.2)]
+    sampler = EnsembleSampler(
+        nwalkers=20 * ndim,
+        ndim=ndim,
+        log_prob_fn=model.likelihood,
+        backend=backend,
[…]

### 2024-02-27 index.qmd
write more scaffolding & about bilateral formalism

@@
   cache: true
+crossref:
+  fig-prefix: "figure"
+  tbl-prefix: "table"
+  eq-prefix: "equation"
+  sec-prefix: "section"
 ---
@@
 # Methods
+
+## Modelling Unilateral Lymphatic Spread
+
+- what we previously did
+
+$$
+P \left( \mathbf{X} \mid \mathbf{Z} \right) = \frac{P \left( \mathbf{Z} \mid \mathbf{X} \right) P \left( \mathbf{X} \right)}{P \left( \mathbf{Z} \right)}
+$$ {#eq-uni-bayes-law}
+
+Let us for later use define the matrices $\boldsymbol{\Lambda}$:
+
+$$
+\boldsymbol{\Lambda} = P \left( \mathbf{X} \mid \mathbf{t} \right) = \begin{pmatrix}
+\boldsymbol{\pi}^\intercal \cdot \mathbf{A}^0 \\
+\boldsymbol{\pi}^\intercal \cdot \mathbf{A}^1 \\
+\vdots \\
+\boldsymbol{\pi}^\intercal \cdot \mathbf{A}^{t_\text{max}} \\
+\end{pmatrix}
+$$ {#eq-lambda-matrix}
+
+The $n$-th row in this matrix corresponds to the distribution over hidden states after $t=n-1$ time-steps.
+
+## Extending the Formalism to Both Sides of the Neck
+
+In analogy to @eq-uni-bayes-law, we can write down the posterior joint probability of an involvement $\mathbf{X}^\text{i}$ ipsilaterally _and_ an involvement $\mathbf{X}^\text{c}$ contralaterally, given a diagnosis of the ipsilateral LNLs $\mathbf{Z}^\text{i}$ and of the contralateral ones $\mathbf{Z}^\text{c}$:
+
+$$
+P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \mid \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \right) = \frac{P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \mid \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right) P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right)}{P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \right)}
+$$
+
+The likelihood of the diagnoses given a hidden state simply factorise: $P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \mid \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right) = P \left( \mathbf{Z}^\text{i} \mid \mathbf{X}^\text{i} \right) \cdot P \left( \mathbf{Z}^\text{c} \mid \mathbf{X}^\text{c} \right)$.
+And the two factors are contained in the observation matrices $\mathbf{B}^\text{i}$ and $\mathbf{B}^\text{c}$.
+
+The term representing the model's prior probability of hidden involvement does not factorize.
+However, if we assume that lymphatic spread typically does not cross the mid-sagittal line, we can write it as a factorising sum:
+
+$$
+\begin{aligned}
+P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right) &= \sum_{t=0}^{t_\text{max}} p(t) \cdot P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \mid t \right) \\
+&= \sum_{t=0}^{t_\text{max}} p(t) \cdot P \left( \mathbf{X}^\text{i} \mid t \right) \cdot P \left( \mathbf{X}^\text{c} \mid t \right)
+\end{aligned}
+$$
+
+This assumption makes intuitive sense: The two sides of the lymphatic network in a typical patient are approximately mirror images of each other.
+Thus, no major vessels cross the mid-sagittal line.
+There may, however, be diffusion of lymph fluid accross this line or bulky involvement that redirects lymphatic drainage significantly.
+
+Using this assumption along with @eq-lambda-matrix, we can write the above distribution algebraically as a product:
+
+$$
+P \left( \mathbf{X}^\text{i} = \boldsymbol{\xi}_n, \mathbf{X}^\text{c} = \boldsymbol{\xi}_m \right) = \left[ \boldsymbol{\Lambda}^\intercal_\text{i} \cdot \operatorname{diag} p(\mathbf{t}) \cdot \boldsymbol{\Lambda}^\intercal_\text{c} \right]_{n,m}
+$$
+
+## Parameter Symmetries
+
+- spread from LNL to LNL probably symmetric
+
+## Tumors Extending Over the Mid-Sagittal Line
+
+- higher prevalence of contralateral involvement when tumor crosses the mid-sagittal plane
+- but still lower than on ipsilateral side
+- spread to contralateral side in case of midline extension may be describes as linear combination of ipsi- and contralateral spread in case of no midline extension.
+- only one additional "mixing parameter" $\alpha$
+- tumor's midline extension modelled as a binary RV (lateralized or crossing/extending)
 

### 2024-02-29 scripts/sampling.py
add more data, join them, and filter it

@@ -1,14 +1,17 @@
 """
-Implement the sampling in this module.
+Run the sampling in this module.
 """
+import argparse
+from pathlib import Path
+
 import numpy as np
+from scipy.special import factorial
 import pandas as pd
 from emcee import EnsembleSampler, backends, moves
 
 from lymph import models
 
 
-from scipy.special import factorial
 
 def binom_pmf(k: np.ndarray, n: int, p: float):
     """Binomial PMF"""
@@ -24,7 +27,25 @@ def late_binomial(support: np.ndarray, p: float = 0.5) -> np.ndarray:
     return binom_pmf(k=support, n=support[-1], p=p)
 
 
-def create_model() -> models.Midline:
+def create_parser() -> argparse.ArgumentParser:
+    """Create the parser."""
+    parser = argparse.ArgumentParser(description=__doc__)
+    parser.add_argument(
+        "--data", type=Path, default="data/enhanced.csv",
+        help="Path to the data file."
+    )
+    parser.add_argument(
+        "--samples", type=Path, default="models/midline.hdf5",
+        help="Path to the generated samples HDF5 file."
+    )
+    parser.add_argument(
+        "--nstep", type=int, default=5000,
+        help="Number of steps for the sampling."
+    )
+    return parser
+
+
+def create_model(data_path: Path) -> models.Midline:
     """Create the model."""
     graph_dict = {
         ("tumor", "T"): ["II", "III"],
@@ -42,18 +63,17 @@ def create_model() -> models.Midline:
     frozen_binom_pmf = binom_pmf(np.arange(model.max_time+1), model.max_time, p=0.3)
     model.set_distribution("early", frozen_binom_pmf)
     model.set_distribution("late", late_binomial)
-    model.load_patient_data(pd.read_csv("data/enhanced.csv", header=[0,1,2]))
+    model.load_patient_data(pd.read_csv(data_path, header=[0,1,2]))
 
     return model
 
 
-def run_sampling(model: models.Midline, nstep: int = 5000) -> None:
+def run_sampling(model: models.Midline, samples_path: Path, nstep: int = 5000) -> None:
     """Run the sampling."""
     ndim = model.get_num_dims()
     initial = np.random.uniform(size=(20 * ndim, ndim))
 
-    suffix = "evo" if model.use_midext_evo else "fix"
-    backend = backends.HDFBackend(filename=f"models/midline_{suffix}.hdf5")
[…]

### 2024-03-11 scripts/__init__.py
sample with fixed lymph package

@@ -1,0 +1,18 @@
+"""
+Scripts and other reused code for this paper.
+"""
+from collections import namedtuple
+
+
+Columns = namedtuple("Columns", ["t_stage", "midext", "ipsi_III"])
+COL = Columns(
+    t_stage=("tumor", "1", "t_stage"),
+    midext=("tumor", "1", "extension"),
+    ipsi_III=("max_llh", "ipsi", "III"),
+)
+CONTRA_LNLS = [
+    ("max_llh", "contra", "I"),
+    ("max_llh", "contra", "II"),
+    ("max_llh", "contra", "III"),
+    ("max_llh", "contra", "IV"),
+]

### 2024-03-11 scripts/paths.py

@@ -1,0 +1,10 @@
+"""
+Define paths for data and models.
+"""
+from pathlib import Path
+
+data_dir = Path("data")
+filtered = data_dir / "filtered.csv"
+
+model_dir = Path("models")
+midline = model_dir / "midline.hdf5"

### 2024-03-11 scripts/sampling.py

@@ -3,14 +3,17 @@ Run the sampling in this module.
 """
 import argparse
 from pathlib import Path
+import warnings
 
 import numpy as np
 from scipy.special import factorial
 import pandas as pd
 from emcee import EnsembleSampler, backends, moves
 
-from lymph import models
+from lymph import models, types
 
+
+warnings.filterwarnings("ignore", category=types.DataWarning)
 
 
 def binom_pmf(k: np.ndarray, n: int, p: float):
@@ -39,7 +42,7 @@ def create_parser() -> argparse.ArgumentParser:
         help="Path to the generated samples HDF5 file."
     )
     parser.add_argument(
-        "--nstep", type=int, default=5000,
+        "--nstep", type=int, default=2000,
         help="Number of steps for the sampling."
     )
     return parser
@@ -48,9 +51,10 @@ def create_parser() -> argparse.ArgumentParser:
 def create_model(data_path: Path) -> models.Midline:
     """Create the model."""
     graph_dict = {
-        ("tumor", "T"): ["II", "III"],
+        ("tumor", "T"): ["II", "III", "IV"],
         ("lnl", "II"): ["III"],
-        ("lnl", "III"): [],
+        ("lnl", "III"): ["IV"],
+        ("lnl", "IV"): [],
     }
     model = models.Midline(
         graph_dict=graph_dict,
@@ -77,6 +81,7 @@ def run_sampling(model: models.Midline, samples_path: Path, nstep: int = 5000) -
 
     backend = backends.HDFBackend(filename=samples_path)
     moves_mix = [(moves.DEMove(), 0.8), (moves.DESnookerMove(), 0.2)]
+
     sampler = EnsembleSampler(
         nwalkers=20 * ndim,
         ndim=ndim,

### 2024-03-11 manuscript.qmd
add contra inv table

--- 2024-02-27 index.qmd
@@
     code-fold: true
+  pdf:
+    keep-tex: true
+    include-in-header:
+    - text: |
+        \usepackage{multirow}
 jupyter: python3
@@
 - we extend it to cover the contralateral side, too
+- naive approach: make two independent models for each side, but that is not supported by the data
 - ipsi- and contralateral side are correlated via time of diagnosis, which is correlated with T-category
@@
 # Methods
+
+## Data
+
+```{python}
+#| echo: false
+#| label: tbl-contra
+#| tbl-cap: Contralateral involvement depending on whether the primary tumor extends over the mid-sagittal line, the T-category, and whether the ipsilateral LNL III was involved or healthy.
+
+import pandas as pd
+from scripts import paths, COL, CONTRA_LNLS
+
+data = pd.read_csv(paths.filtered, header=[0,1,2])
+data[COL.t_stage] = data[COL.t_stage].apply(lambda x: "early" if x <= 2 else "late")
+grouped = data.groupby(by=[COL.midext, COL.t_stage, COL.ipsi_III])
+
+total = grouped.count()[CONTRA_LNLS]["max_llh", "contra"]
+idx = total.index.rename(["Mid. ext.", "T-cat.", "ipsi III"])
+
+num_involved = grouped.sum()[CONTRA_LNLS]["max_llh", "contra"]
+percent_involved = num_involved / total
+total.index = idx
+num_involved.index = idx
+percent_involved.index = idx
+
+involved = num_involved.join(
+  100 * percent_involved,
+  rsuffix=" (%)",
+).sort_index(axis="columns")
+involved.columns = pd.MultiIndex.from_product([["I", "II", "III", "IV"], ["n", "%"]])
+involved.style \
+  .format(precision=2)
+```
 

### 2024-03-12 scripts/sampling.py
dont ignore unknown midline extension

@@ -61,14 +61,13 @@ def create_model(data_path: Path) -> models.Midline:
         is_symmetric={"lnl_spread": True},
         use_mixing=True,
         use_central=False,
-        use_midext_evo=False,
+        use_midext_evo=True,
     )
     model.set_modality("max_llh", spec=1.0, sens=1.0)
     frozen_binom_pmf = binom_pmf(np.arange(model.max_time+1), model.max_time, p=0.3)
     model.set_distribution("early", frozen_binom_pmf)
     model.set_distribution("late", late_binomial)
     df = pd.read_csv(data_path, header=[0,1,2])
-    df["tumor", "1", "extension"] = df["tumor", "1", "extension"].astype(bool)
     model.load_patient_data(df)
 
     return model

### 2024-03-12 manuscript.qmd
write more about formalism of extension

@@
 
-# Methods
-
-## Data
-
-```{python}
-#| echo: false
-#| label: tbl-contra
-#| tbl-cap: Contralateral involvement depending on whether the primary tumor extends over the mid-sagittal line, the T-category, and whether the ipsilateral LNL III was involved or healthy.
-
-import pandas as pd
-from scripts import paths, COL, CONTRA_LNLS
-
-data = pd.read_csv(paths.filtered, header=[0,1,2])
-data[COL.t_stage] = data[COL.t_stage].apply(lambda x: "early" if x <= 2 else "late")
-grouped = data.groupby(by=[COL.midext, COL.t_stage, COL.ipsi_III])
-
-total = grouped.count()[CONTRA_LNLS]["max_llh", "contra"]
-idx = total.index.rename(["Mid. ext.", "T-cat.", "ipsi III"])
-
-num_involved = grouped.sum()[CONTRA_LNLS]["max_llh", "contra"]
-percent_involved = num_involved / total
-total.index = idx
-num_involved.index = idx
-percent_involved.index = idx
-
-involved = num_involved.join(
-  100 * percent_involved,
-  rsuffix=" (%)",
-).sort_index(axis="columns")
-involved.columns = pd.MultiIndex.from_product([["I", "II", "III", "IV"], ["n", "%"]])
-involved.style \
-  .format(precision=2)
-```
-
-## Modelling Unilateral Lymphatic Spread
+# Unilateral Model for Lymphatic Progression {#sec-unilateral}
 
@@
 
-## Extending the Formalism to Both Sides of the Neck
+# Extension to a Bilateral Model
 
-In analogy to @eq-uni-bayes-law, we can write down the posterior joint probability of an involvement $\mathbf{X}^\text{i}$ ipsilaterally _and_ an involvement $\mathbf{X}^\text{c}$ contralaterally, given a diagnosis of the ipsilateral LNLs $\mathbf{Z}^\text{i}$ and of the contralateral ones $\mathbf{Z}^\text{c}$:
+A naive approach to model the contralateral lymphatic spread would be to simply employ two independent unilateral models as introduced in @sec-unilateral.
+During training, one could even enforce some shared parameters between these two models, like the parameterization of the distributions over diagnose times or the spread among the LNLs.
+Additionally, we could think of an approach to incorporate the primary tumor's mid-sagittal extension as a risk factor.
+
+However, we did not consider this straightforward approach, as it lacks a way to describe the correlation between ipsi- and contralateral involvement.
+This is displayed in @tbl-contra and shows how often the contralateral LNLs I, II, III, and IV were involved, given all possible combinations of midline extension, T-category, and ipsilateral LNL III involvement.
+Unsurprisingly, the prevalence for contralateral involvement is consistently higher when the tumor extends over the mid-sagittal line or is of later T-category.
+But it is also more frequent when the ipsilateral side shows more severe involvement, which is here shown via the surrogate LNL III.
+
+```{python}
+#| echo: false
+#| label: tbl-contra
+#| tbl-cap: Contralateral involvement depending on whether the primary tumor extends over the mid-sagittal line, the T-category, and whether the ipsilateral LNL III was involved or healthy.
+
+import pandas as pd
+import numpy as np
+from scripts import paths, COL, CONTRA_LNLS
+
+def right_align(column) -> str:
+  """Right align some columns."""
+  return ["text-align: right"] * len(column)
+
+data = pd.read_csv(paths.filtered, header=[0,1,2])
+data[COL.t_stage] = data[COL.t_stage].apply(lambda x: "early" if x <= 2 else "late")
+grouped = data.groupby(by=[COL.midext, COL.t_stage, COL.ipsi_III])
+
+total = grouped.count()[CONTRA_LNLS]["max_llh", "contra"]
+idx = total.index.rename(["Mid. ext.", "T-cat.", "ipsi III"])
+
+num_involved = grouped.sum()[CONTRA_LNLS]["max_llh", "contra"]
+percent_involved = num_involved / total
+total.index = idx
+num_involved.index = idx
+percent_involved.index = idx
+
+involved = num_involved.join(
+  100 * percent_involved,
+  rsuffix=" (%)",
+).sort_index(axis="columns")
+involved.columns = pd.MultiIndex.from_product([["I", "II", "III", "IV"], ["n", "%"]])
+(
+  involved.style
+  .format(precision=2)
+  .apply(right_align, axis="index")
+)
+```
+
+Thus, we attempt to extend the formalism in @sec-unilateral in such a way that the model's ipsi- and contralateral side evolve synchronously.
+To achieve that, we start by writing down the posterior distribution of involvement an analogy to @eq-uni-bayes-law, which is now a joint probability of an involvement $\mathbf{X}^\text{i}$ ipsilaterally *and* an involvement $\mathbf{X}^\text{c}$ contralaterally, given a diagnosis of the ipsilateral LNLs $\mathbf{Z}^\text{i}$ and of the contralateral ones $\mathbf{Z}^\text{c}$:
 
@@
 $$
-P \left( \mathbf{X}^\text{i} = \boldsymbol{\xi}_n, \mathbf{X}^\text{c} = \boldsymbol{\xi}_m \right) = \left[ \boldsymbol{\Lambda}^\intercal_\text{i} \cdot \operatorname{diag} p(\mathbf{t}) \cdot \boldsymbol{\Lambda}^\intercal_\text{c} \right]_{n,m}
+P \left( \mathbf{X}^\text{i} = \boldsymbol{\xi}_n, \mathbf{X}^\text{c} = \boldsymbol{\xi}_m \right) = \left[ \boldsymbol{\Lambda}^\intercal_\text{i} \cdot \operatorname{diag} p(\mathbf{t}) \cdot \boldsymbol{\Lambda}_\text{c} \right]_{n,m}
 $$
+
+## Incorporating Midline Crossing
+
+To account for the increased prevalence of involvement on the contralateral side when the tumor is not clearly lateralized anymore, we also model the tumor's extension over the mid-sagittal line as a binary random variable.
+It starts lateralized (state $\centernot{\text{e}}$) and every time-step there is a finite probability $\epsilon$ that the tumor grows over the symmetry plane of the patient (into state $\text{e}$).
+
+Technically, the introduction of this additional random variable doubles the space of the hidden states.
+However, since we assume no correlation between the tumor's lateralization and the metastases in the LNLs, we can evolve the two parts separately.
+
+As with the state of the lymphatic network, we can also define a small transition matrix $\mathbf{A}_\epsilon$ that evolves the lateralized starting state $\boldsymbol{\pi}_\text{e} = \begin{pmatrix} 1 & 0 \end{pmatrix}$:
+
+$$
+\boldsymbol{\xi}^\text{e}[t] = \boldsymbol{\pi}_\text{e} \cdot \mathbf{A}_\epsilon^t = \begin{pmatrix} 1 & 0 \end{pmatrix} \cdot \begin{pmatrix} 1 - \epsilon & \epsilon \\ 0 & 1 \end{pmatrix}^t
+$$
+
+Just as the lymphatic states, this lateralization state's evolution over the time-steps must then be marginalized over the time of diagnosis using $p(t)$.
+
+- multiply first or second index of evolution with time-prior
+- choose appropriate resulting joint, depending on patient's tumor lateralization
 
@@
 
-- spread from LNL to LNL probably symmetric
+In general, the matrices $\boldsymbol{\Lambda}_\text{i}$ and $\boldsymbol{\Lambda}_\text{c}$ could be parameterized using a disjoint set of parameters.
+I.e., the ipsi- and contralateral spread rates are entirely different.
+However, using two sensible assumptions, we can reduce the parameter space by sharing some parameters between the sides:
+
+1.
+We assume the spread *among* the LNLs to be same on both sides.
+It is reasonable to assume the lymphatic system is symmetric.
+Thus, the spread rates from one LNL to the other should be symmetric, too.
+Formally, this means \
+   $$
+   \begin{aligned}
+   b_v^\text{c} &\neq b_v^\text{i} \\
+   t_{rv}^\text{c} &= t_{rv}^\text{i}
+   \end{aligned}
+   $$ {#eq-symmetries}
+   for all $v \leq V$ and $r \in \operatorname{pa}(v)$.
+2.
+The tumor's spread to the contralateral side in case of an extension over the midline is larger than if it was clearly lateralized, but smaller than its spread to the ipsilateral side.
+This assumption stems from a simple thought experiment: Consider moving the tumor from a clearly lateralized position accross the mid-sagittal plane to its mirror position.
+In the beginning we would have $b_v^\text{c} < b_v^\text{i}$, while in the end, the situation is reversed.
+If a tumor extends over the mid-sagittal line, its contralateral spread rate can be expected to be in between these two extremes.
+We encode this in a *mixing parameter* $\alpha$ that captures a "degree of asymmetry": \
+   $$
+   b_v^\text{c,e} = \alpha \cdot b_v^\text{i} + (1 - \alpha) \cdot b_v^{\text{c},\centernot{\text{e}}}
+   $$ {#eq-mixing}
 

### 2024-03-16 manuscript.qmd
sample using new lyscripts & try quarto plots

@@
         \usepackage{multirow}
+        \usepackage{centernot}
 jupyter: python3
@@
 ```{python}
-from emcee import backends
-from corner import corner
+#| echo: false
+#| label: fig-burnin-history
+#| layout-ncol: 2
+#| fig-cap: Monitoring quantities during the burn-in phase of the parameter sampling.
+#| fig-subcap: 
+#|   - "Autocorrelation time"
+#|   - "log likelihood"
+
+import pandas as pd
 import matplotlib.pyplot as plt
@@
 
-backend = backends.HDFBackend(
-  filename="models/midline_fix.hdf5",
-  read_only=True,
+plt.rcParams.update(figsizes.icml2024_half())
+plt.rcParams.update(fontsizes.icml2024())
+
+burnin_history = pd.read_csv("models/midline_history.csv").set_index("steps")
+
+plt.plot(
+  burnin_history.index,
+  burnin_history.acor_times,
 )
-samples = backend.get_chain(thin=20)
-ndim = samples.shape[-1]
+plt.show()
 
-plt.rcParams.update(**figsizes.icml2022_full(nrows=ndim))
+plt.plot(
+  burnin_history.index,
+  burnin_history.max_log_probs,
+)
+plt.show()
+```
 
-fig, axes = plt.subplots(nrows=ndim, sharex=True)
-for i, ax in enumerate(axes):
-  ax.plot(samples[:,::10,i], color="gray", alpha=0.2)
-  ax.axhline(samples[200:,:,i].mean(), color="red")
-  ax.set_xlim(0., samples.shape[0])
-  ax.set_ylim(0., 1.)
-```
+@Fig-burnin-history is only a temporary placeholder for now.

### 2024-03-16 scripts/sampling.py

@@ -1,106 +1,0 @@
-"""
-Run the sampling in this module.
-"""
-import argparse
-from pathlib import Path
-import warnings
-
-import numpy as np
-from scipy.special import factorial
-import pandas as pd
-from emcee import EnsembleSampler, backends, moves
-
-from lymph import models, types
-
-
-warnings.filterwarnings("ignore", category=types.DataWarning)
-
-
-def binom_pmf(k: np.ndarray, n: int, p: float):
-    """Binomial PMF"""
-    if p > 1. or p < 0.:
-        # This value error is important to enable seamless sampling!
-        raise ValueError("Binomial prob must be btw. 0 and 1")
-    q = 1. - p
-    binom_coeff = factorial(n) / (factorial(k) * factorial(n - k))
-    return binom_coeff * p**k * q**(n - k)
-
-def late_binomial(support: np.ndarray, p: float = 0.5) -> np.ndarray:
-    """Parametrized binomial distribution."""
-    return binom_pmf(k=support, n=support[-1], p=p)
-
-
-def create_parser() -> argparse.ArgumentParser:
-    """Create the parser."""
-    parser = argparse.ArgumentParser(description=__doc__)
-    parser.add_argument(
-        "--data", type=Path, default="data/filtered.csv",
-        help="Path to the data file."
-    )
-    parser.add_argument(
-        "--samples", type=Path, default="models/midline.hdf5",
-        help="Path to the generated samples HDF5 file."
-    )
-    parser.add_argument(
-        "--nstep", type=int, default=2000,
-        help="Number of steps for the sampling."
-    )
-    return parser
-
-
-def create_model(data_path: Path) -> models.Midline:
-    """Create the model."""
-    graph_dict = {
-        ("tumor", "T"): ["II", "III", "IV"],
-        ("lnl", "II"): ["III"],
-        ("lnl", "III"): ["IV"],
-        ("lnl", "IV"): [],
-    }
-    model = models.Midline(
-        graph_dict=graph_dict,
-        is_symmetric={"lnl_spread": True},
-        use_mixing=True,
-        use_central=False,
-        use_midext_evo=True,
-    )
-    model.set_modality("max_llh", spec=1.0, sens=1.0)
-    frozen_binom_pmf = binom_pmf(np.arange(model.max_time+1), model.max_time, p=0.3)
-    model.set_distribution("early", frozen_binom_pmf)
[…]

### 2024-03-18 manuscript.qmd
sample different simple models

@@
 #| label: fig-burnin-history
-#| layout-ncol: 2
+#| layout-ncol: 1
+#| layout-nrow: 1
 #| fig-cap: Monitoring quantities during the burn-in phase of the parameter sampling.
@@
 #|   - "log likelihood"
+from collections import namedtuple
 
 import pandas as pd
+import numpy as np
 import matplotlib.pyplot as plt
@@
 
-plt.rcParams.update(figsizes.icml2024_half())
+nrows, ncols = 1, 3
+label_map = {
+  "acor_times": "Autocorrelation [steps]",
+  "accept_fracs": "Sample Acceptance [%]",
+  "max_log_probs": "Maximum log-Likelihood",
+}
+
+plt.rcParams.update(figsizes.icml2024_full(nrows=nrows, ncols=ncols))
 plt.rcParams.update(fontsizes.icml2024())
 
-burnin_history = pd.read_csv("models/midline_history.csv").set_index("steps")
+fig, axes = plt.subplots(
+  nrows=nrows,
+  ncols=ncols,
+  sharex=True,
+)
 
-plt.plot(
-  burnin_history.index,
-  burnin_history.acor_times,
-)
-plt.show()
+for model in ["ipsi", "contra", "bilateral", "midline"]:
+  history = pd.read_csv(f"models/{model}/history.csv").set_index("steps")
+  for i, column in enumerate(history.columns):
+    history.plot(y=column, ax=axes[i], label=model)
+    axes[i].autoscale(enable=True, tight=True, axis="x")
+    axes[i].set_title(label_map[column])
 
-plt.plot(
-  burnin_history.index,
-  burnin_history.max_log_probs,
-)
 plt.show()

### 2024-03-18 scripts/reduce.py

@@ -1,0 +1,36 @@
+"""
+Remove all unnecessary columns from the joined, enhanced and (row-wise) filtered data.
+"""
+import argparse
+from pathlib import Path
+
+import pandas as pd
+
+
+def create_parser():
+    """Create the argument parser."""
+    parser = argparse.ArgumentParser(description=__doc__)
+    parser.add_argument('input', type=Path, help='Path to the input file.')
+    parser.add_argument('output', type=Path, help='Path to the output file.')
+    return parser
+
+
+def main():
+    """Main function."""
+    parser = create_parser()
+    args = parser.parse_args()
+
+    data = pd.read_csv(
+        args.input, header=[0,1,2],
+    ).drop(columns=[
+        "enbloc_dissected",
+        "enbloc_positive",
+        "positive_dissected",
+        "total_dissected",
+    ])
+
+    data.to_csv(args.output, index=False)
+
+
+if __name__ == '__main__':
+    main()

### 2024-03-19 manuscript.qmd
add table with param estimates

@@
 execute: 
-  cache: true
+  cache: false
 crossref:
@@
 data[COL.t_stage] = data[COL.t_stage].apply(lambda x: "early" if x <= 2 else "late")
+early_midext_data = (data.loc[data[COL.t_stage] == "early", COL.midext] == True).mean()
+late_midext_data = (data.loc[data[COL.t_stage] == "late", COL.midext] == True).mean()
 grouped = data.groupby(by=[COL.midext, COL.t_stage, COL.ipsi_III])
@@
 ).sort_index(axis="columns")
-involved.columns = pd.MultiIndex.from_product([["I", "II", "III", "IV"], ["n", "%"]])
+involved.columns = pd.MultiIndex.from_product(
+  [["I", "II", "III", "IV"],
+   ["n", "%"]],
+  names=["LNL", ""]
+)
 (
@@
 #| fig-cap: Monitoring quantities during the burn-in phase of the parameter sampling.
-#| fig-subcap: 
-#|   - "Autocorrelation time"
-#|   - "log likelihood"
 from collections import namedtuple
@@
 @Fig-burnin-history is only a temporary placeholder for now.
+
+## Parameter Estimates
+
+```{python}
+#| label: tbl-midline-params
+#| tbl-cap: Mean sampled parameter estimates of the midline model and the respective standard deviation.
+#| echo: false
+
+import yaml
+
+import pandas as pd
+from emcee.backends import HDFBackend
+from lyscripts.utils import create_model
+
+with open("models/midline/params.yaml", mode="r") as file:
+  config = yaml.safe_load(file)
+
+model = create_model(config)
+backend = HDFBackend("models/midline/samples.hdf5", read_only=True)
+chain = backend.get_chain(flat=True)
+
+param_names = list(model.get_params().keys())
+param_means = chain.mean(axis=0)
+param_stds = chain.std(axis=0)
+model.set_params(*param_means)
+early_midext_prob = model.get_distribution("early").pmf @ model.midext_evo()[:,1]
+late_midext_prob = model.get_distribution("late").pmf @ model.midext_evo()[:,1]
+
+params_table = pd.DataFrame({
+  "name": param_names,
+  "mean": param_means,
+  "std": param_stds,
+})
+(
+  params_table.style
+  .format("{:.2%}", subset=["mean", "std"])
+  .apply(right_align, subset=["mean", "std"])
+  .hide()
+)
+```
+
+The predicted prevalence of midline extension for early T-category tumors is `{python} f"{early_midext_prob:.2%}"`. For a late T-category tumor, it is `{python} f"{late_midext_prob:.2%}"`.
+
+In the data, the prevalence of midline extension for early T-category tumors is `{python} f"{early_midext_data:.2%}"`. For late T-category tumors it is `{python} f"{late_midext_data:.2%}"`.
+
+- Discrepancy between data and model is due to late T-stage's time prior
+- still tied to other parameters
+- cannot fully represent spread of midline extension and also represent spread params

### 2024-04-02 manuscript.qmd
add prevalences test plot

@@
 
+## Prevalence Comparisons
+
+Below we compare the predicted with the observed prevalence of some involvement patterns.
+
+```{python}
+#| echo: false
+#| label: fig-prevalences
+#| fig-cap: Prevalences of different scenarios.
+
+from pathlib import Path
+
+import matplotlib.pyplot as plt
+from lymph import models
+from lyscripts import utils
+from lyscripts.scenario import Scenario
+from lyscripts.plot.utils import draw, BetaPosterior, Histogram, COLOR_CYCLE
+
+model_dir = Path("models/midline")
+filename = model_dir / "prevalences.hdf5"
+
+params = utils.load_yaml_params(model_dir / "params.yaml")
+model = utils.create_model(params)
+scenarios_config = utils.load_yaml_params("scenarios.yaml")
+scenarios = Scenario.list_from_params(
+  params=scenarios_config,
+  is_uni=isinstance(model, models.Unilateral),
+  side=params["model"].get("side", "ipsi")
+)
+
+fig, ax = plt.subplots()
+
+content = []
+color = next(COLOR_CYCLE)
+content += [
+  Histogram.from_hdf5(
+    filename=filename,
+    dataname=scenarios[0].md5_hash("prevalences"),
+    color=color,
+  ),
+  BetaPosterior.from_hdf5(
+    filename=filename,
+    dataname=scenarios[0].md5_hash("prevalences"),
+  ),
+]
+color = next(COLOR_CYCLE)
+content += [
+  Histogram.from_hdf5(
+    filename=filename,
+    dataname=scenarios[1].md5_hash("prevalences"),
+    color=color,
+  ),
+  BetaPosterior.from_hdf5(
+    filename=filename,
+    dataname=scenarios[1].md5_hash("prevalences"),
+  ),
+]
+draw(
+  axes=ax,
+  contents=content,
+  percent_lims=(2., 2.),
+)
+
+```
+
 ## Parameter Estimates

### 2024-04-02 manuscript.qmd
refactor shared code & add to formalism

--- 2024-03-19 manuscript.qmd
@@
 ---
+
+```{python}
+#| echo: false
+
+import pandas as pd
+import numpy as np
+import matplotlib.pyplot as plt
+from tueplots import figsizes, fontsizes
+
+from lymph import models
+from lyscripts import utils
+
+from scripts import shared, paths, COL, CONTRA_LNLS
+```
 
@@
 
-However, we did not consider this straightforward approach, as it lacks a way to describe the correlation between ipsi- and contralateral involvement.
+However, this approach lacks a way to describe the correlation between ipsi- and contralateral involvement.
 This is displayed in @tbl-contra and shows how often the contralateral LNLs I, II, III, and IV were involved, given all possible combinations of midline extension, T-category, and ipsilateral LNL III involvement.
@@
 #| tbl-cap: Contralateral involvement depending on whether the primary tumor extends over the mid-sagittal line, the T-category, and whether the ipsilateral LNL III was involved or healthy.
-
-import pandas as pd
-import numpy as np
-from scripts import paths, COL, CONTRA_LNLS
-
 def right_align(column) -> str:
@@
 
-data = pd.read_csv(paths.filtered, header=[0,1,2])
+data = pd.read_csv(paths.data, header=[0,1,2])
 data[COL.t_stage] = data[COL.t_stage].apply(lambda x: "early" if x <= 2 else "late")
@@
 To account for the increased prevalence of involvement on the contralateral side when the tumor is not clearly lateralized anymore, we also model the tumor's extension over the mid-sagittal line as a binary random variable.
-It starts lateralized (state $\centernot{\text{e}}$) and every time-step there is a finite probability $\epsilon$ that the tumor grows over the symmetry plane of the patient (into state $\text{e}$).
+It starts lateralized and every time-step there is a finite probability $p_\epsilon$ that the tumor grows over the symmetry plane of the patient.
 
@@
 
-As with the state of the lymphatic network, we can also define a small transition matrix $\mathbf{A}_\epsilon$ that evolves the lateralized starting state $\boldsymbol{\pi}_\text{e} = \begin{pmatrix} 1 & 0 \end{pmatrix}$:
+As with the state of the lymphatic network, we can also define a small transition matrix $\mathbf{A}_\epsilon$ that evolves the lateralized starting state $\boldsymbol{\epsilon}[0] = \begin{pmatrix} 1 & 0 \end{pmatrix}$:
 
 $$
-\boldsymbol{\xi}^\text{e}[t] = \boldsymbol{\pi}_\text{e} \cdot \mathbf{A}_\epsilon^t = \begin{pmatrix} 1 & 0 \end{pmatrix} \cdot \begin{pmatrix} 1 - \epsilon & \epsilon \\ 0 & 1 \end{pmatrix}^t
+\boldsymbol{\epsilon}[t] = \boldsymbol{\epsilon}[0] \cdot \mathbf{A}_\epsilon^t = \begin{pmatrix} 1 & 0 \end{pmatrix} \cdot \begin{pmatrix} 1 - \epsilon & \epsilon \\ 0 & 1 \end{pmatrix}^t = \begin{pmatrix} (1 - \epsilon)^t & 1 - (1 - \epsilon)^t \end{pmatrix}
 $$
+
+We can again collect the entire evolution in a matrix $\boldsymbol{\Lambda}_\epsilon$ that stores the probabilies $P(\boldsymbol{\epsilon} \mid t)$ for each time-step $t$:
+
+$$
+\boldsymbol{\Lambda}_\epsilon = P(\boldsymbol{\epsilon} \mid t) = \begin{pmatrix}
+\boldsymbol{\epsilon}[0] \cdot \mathbf{A}_\epsilon^0 \\
+\boldsymbol{\epsilon}[1] \cdot \mathbf{A}_\epsilon^1 \\
+\vdots \\
+\boldsymbol{\epsilon}[t] \cdot \mathbf{A}_\epsilon^t \\
+\end{pmatrix}
+$$
+
+This quantity has $t$ rows, one for each time-step, and two columns: one for the decreasing probability of a lateralized tumor and one for the increasing probability of a tumor that has crossed the mid-sagittal line.
+
+```{python}
+#| echo: false
+#| label: fig-midext-evo
+#| fig-cap: The matrix $\boldsymbol{\Lambda}_\epsilon$ shown as heatmap of probabilities.
+from matplotlib.colors import LinearSegmentedColormap
+from mpl_toolkits.axes_grid1 import make_axes_locatable
+
+from lyscripts.plot.utils import COLORS
+
+nrows, ncols = 1, 1
+
+plt.rcParams.update(figsizes.icml2024_full(nrows=nrows, ncols=ncols))
+plt.rcParams.update(fontsizes.icml2024())
+
+model = shared.get_model(which="midline", load_samples=True)
+
+fig, ax = plt.subplots(tight_layout=True)
+
+cmap = LinearSegmentedColormap.from_list("redusz", [COLORS["gray"], COLORS["red"]])
+im = ax.imshow(model.midext_evo().T, cmap=cmap)
+
+# create an Axes on the right side of ax. The width of cax will be 5%
+# of ax and the padding between cax and ax will be fixed at 0.05 inch.
+divider = make_axes_locatable(ax);
+cax = divider.append_axes("right", size="5%", pad=0.05);
+plt.colorbar(im, cax=cax, label="Prob. of midext");
+
+ax.set_xticks(np.arange(model.max_time + 1));
+ax.set_xlabel("time-step $t$");
+
+ax.set_yticks([0., 1.], labels=["lat.", "ext."])
+```
 
@@
 
-import pandas as pd
-import numpy as np
-import matplotlib.pyplot as plt
-from tueplots import figsizes, fontsizes
-
 nrows, ncols = 1, 3
@@
 
+## Prevalence Comparisons
+
+Below we compare the predicted with the observed prevalence of some involvement patterns.
+
+```{python}
+#| echo: false
+#| label: fig-prevalences
+#| fig-cap: Prevalences of different scenarios.
+from lyscripts.scenario import Scenario
+from lyscripts.plot.utils import draw, BetaPosterior, Histogram, COLOR_CYCLE
+
+model = shared.get_model(which="midline")
+scenarios_config = utils.load_yaml_params("scenarios.yaml")
+scenarios = Scenario.list_from_params(scenarios_config)
+
+fig, ax = plt.subplots()
+
+content = []
+color = next(COLOR_CYCLE)
+content += [
+  Histogram.from_hdf5(
+    filename=paths.get_filename(model="midline", kind="prevalences"),
+    dataname=scenarios[0].md5_hash("prevalences"),
+    color=color,
+  ),
+  BetaPosterior.from_hdf5(
+    filename=paths.get_filename(model="midline", kind="prevalences"),
+    dataname=scenarios[0].md5_hash("prevalences"),
+  ),
+]
+color = next(COLOR_CYCLE)
+content += [
+  Histogram.from_hdf5(
+    filename=paths.get_filename(model="midline", kind="prevalences"),
+    dataname=scenarios[1].md5_hash("prevalences"),
+    color=color,
+  ),
+  BetaPosterior.from_hdf5(
+    filename=paths.get_filename(model="midline", kind="prevalences"),
+    dataname=scenarios[1].md5_hash("prevalences"),
+  ),
+]
+draw(
+  axes=ax,
+  contents=content,
+  percent_lims=(2., 2.),
+)
+```
+
 ## Parameter Estimates
@@
 ```{python}
+#| echo: false
 #| label: tbl-midline-params
 #| tbl-cap: Mean sampled parameter estimates of the midline model and the respective standard deviation.
-#| echo: false
+model = shared.get_model(which="midline", load_samples=True)
+samples = shared.get_samples(which="midline")
 
-import yaml
+names = list(model.get_params().keys())
+means, stds = samples.mean(axis=0), samples.std(axis=0)
 
-import pandas as pd
-from emcee.backends import HDFBackend
-from lyscripts.utils import create_model
-
-with open("models/midline/params.yaml", mode="r") as file:
-  config = yaml.safe_load(file)
-
-model = create_model(config)
-backend = HDFBackend("models/midline/samples.hdf5", read_only=True)
-chain = backend.get_chain(flat=True)
-
-param_names = list(model.get_params().keys())
-param_means = chain.mean(axis=0)
-param_stds = chain.std(axis=0)
-model.set_params(*param_means)
 early_midext_prob = model.get_distribution("early").pmf @ model.midext_evo()[:,1]
@@
 
-params_table = pd.DataFrame({
-  "name": param_names,
-  "mean": param_means,
-  "std": param_stds,
-})
+params_table = pd.DataFrame({"name": names, "mean": means, "std": stds})
 (

### 2024-04-02 scripts/paths.py

@@ -2,9 +2,16 @@
 Define paths for data and models.
 """
 from pathlib import Path
+from typing import Literal
 
 data_dir = Path("data")
-filtered = data_dir / "filtered.csv"
-
+data = data_dir / "reduced.csv"
 model_dir = Path("models")
-midline = model_dir / "midline.hdf5"
+
+
+def get_filename(
+    model: Literal["ipsi", "contra", "bilateral", "midline"],
+    kind: Literal["params", "samples", "priors", "posteriors", "prevalences", "risks"],
+) -> Path:
+    """Get the filename of the samples of one of the four trained models."""
+    return next((model_dir / model).glob(f"{kind}.*"))

### 2024-04-02 scripts/shared.py

@@ -1,0 +1,36 @@
+"""
+Some shared and reduced functions for the scripts.
+"""
+from typing import Literal
+
+from lymph import types
+from lyscripts import utils
+import numpy as np
+
+from . import paths
+
+
+def get_model(
+    which: Literal["ipsi", "contra", "bilateral", "midline"],
+    load_samples: bool = True,
+) -> types.Model:
+    """Get one of the four trained models."""
+    params_path = paths.model_dir / which / "params.yaml"
+    params = utils.load_yaml_params(params_path)
+    model = utils.create_model(params)
+
+    if not load_samples:
+        return model
+
+    samples_path = paths.model_dir / which / "samples.hdf5"
+    samples = utils.load_model_samples(samples_path)
+    model.set_params(*samples.mean(axis=0))
+    return model
+
+
+def get_samples(
+    which: Literal["ipsi", "contra", "bilateral", "midline"]
+) -> np.ndarray:
+    """Get the samples of one of the four trained models."""
+    samples_path = paths.model_dir / which / "samples.hdf5"
+    return utils.load_model_samples(samples_path)

### 2024-04-03 manuscript.qmd
use custom fig-/fontsizes

@@
 import matplotlib.pyplot as plt
-from tueplots import figsizes, fontsizes
+from matplotlib import ticker
 
@@
 
-plt.rcParams.update(figsizes.icml2024_full(nrows=nrows, ncols=ncols))
-plt.rcParams.update(fontsizes.icml2024())
+plt.rcParams.update(shared.get_fontsizes())
+plt.rcParams.update(shared.get_figsizes(nrows=nrows, ncols=ncols))
 
@@
 #| label: fig-burnin-history 
-#| layout-ncol: 1
-#| layout-nrow: 1
 #| fig-cap: Monitoring quantities during the burn-in phase of the parameter sampling.
@@
 
+def custom(x, pos):
+  return "{:.1f}".format(x/1000)
+
 nrows, ncols = 1, 3
 label_map = {
-  "acor_times": "Autocorrelation [steps]",
-  "accept_fracs": "Sample Acceptance [%]",
-  "max_log_probs": "Maximum log-Likelihood",
+  "acor_times": "autocorrelation [steps]",
+  "accept_fracs": "acceptance fraction [%]",
+  "max_log_probs": r"max. log-likelihood [$\times 10^3$]",
 }
 
-plt.rcParams.update(figsizes.icml2024_full(nrows=nrows, ncols=ncols))
-plt.rcParams.update(fontsizes.icml2024())
+plt.rcParams.update(shared.get_fontsizes())
+plt.rcParams.update(shared.get_figsizes(nrows=nrows, ncols=ncols, aspect_ratio=1.))
 
@@
     axes[i].autoscale(enable=True, tight=True, axis="x")
-    axes[i].set_title(label_map[column])
+    axes[i].set_ylabel(label_map[column])
+    axes[i].xaxis.set_major_formatter(ticker.FuncFormatter(custom))
+    axes[i].set_xlabel(r"steps [$\times 10^3$]")
 
+axes[1].yaxis.set_major_formatter(ticker.StrMethodFormatter("{x:.0%}"))
+axes[2].yaxis.set_major_formatter(ticker.FuncFormatter(custom))
 plt.show()
@@
 from lyscripts.scenario import Scenario
-from lyscripts.plot.utils import draw, BetaPosterior, Histogram, COLOR_CYCLE
+from lyscripts.plot.utils import draw, BetaPosterior, Histogram, COLORS
 
@@
 
+plt.rcParams.update(shared.get_fontsizes())
+plt.rcParams.update(shared.get_figsizes(aspect_ratio=2.))
+
 fig, ax = plt.subplots()
@@
 content = []
-color = next(COLOR_CYCLE)
 content += [
@@
     dataname=scenarios[0].md5_hash("prevalences"),
-    color=color,
+    color=COLORS["blue"],
   ),
@@
     dataname=scenarios[0].md5_hash("prevalences"),
+    color=COLORS["blue"],
   ),
 ]
-color = next(COLOR_CYCLE)
+
 content += [
@@
     dataname=scenarios[1].md5_hash("prevalences"),
-    color=color,
+    color=COLORS["orange"],
   ),
@@
     dataname=scenarios[1].md5_hash("prevalences"),
+    color=COLORS["orange"],
   ),
@@
 
-early_midext_prob = model.get_distribution("early").pmf @ model.midext_evo()[:,1]
-late_midext_prob = model.get_distribution("late").pmf @ model.midext_evo()[:,1]
+early_midext_prob = model.state_dist(t_stage="early")[1].sum()
+late_midext_prob = model.state_dist(t_stage="late")[1].sum()
 

### 2024-04-03 scripts/shared.py

@@ -1,13 +1,17 @@
 """
 Some shared and reduced functions for the scripts.
 """
-from typing import Literal
+from typing import Any, Literal
 
 from lymph import types
 from lyscripts import utils
 import numpy as np
 
 from . import paths
+
+
+CM_TO_INCH = 0.393701
+GOLDEN_RATIO = 1.61803398875
 
 
 def get_model(
@@ -34,3 +38,49 @@ def get_samples(
     """Get the samples of one of the four trained models."""
     samples_path = paths.model_dir / which / "samples.hdf5"
     return utils.load_model_samples(samples_path)
+
+
+def get_fontsizes(base: int = 8, offset: int = 2) -> dict[str, int]:
+    """Get fontsizes that can be used to update the matplotlib rcParams.
+
+    This is essentially taken from the tueplots package v0.0.14.
+    """
+    return {
+        "font.size": base,
+        "axes.labelsize": base,
+        "legend.fontsize": base - offset,
+        "xtick.labelsize": base - offset,
+        "ytick.labelsize": base - offset,
+        "axes.titlesize": base,
+    }
+
+
+def get_figsizes(
+    nrows: int = 1,
+    ncols: int = 1,
+    aspect_ratio: float = GOLDEN_RATIO,
+    width: float = 17.0,
+    pad: float = 0.005,
+    unit: Literal["cm", "in"] = "cm",
+    constrained_layout: bool = True,
+    tight_layout: bool = True,
+) -> dict[str, Any]:
+    """Get figure sizes that can be used to update the matplotlib rcParams.
+
+    This is heavily inspired by the tueplots package v0.0.14.
+    """
+    if unit == "cm":
+        width *= CM_TO_INCH
+        pad *= CM_TO_INCH
+
+    subplot_width = width / ncols
+    subplot_height = subplot_width / aspect_ratio
+    height = subplot_height * nrows
+
+    return {
+        "figure.figsize": (width, height),
+        "figure.constrained_layout.use": constrained_layout,
+        "figure.autolayout": tight_layout,
+        "savefig.bbox": "tight",
+        "savefig.pad_inches": pad,
+    }

### 2024-04-04 scripts/shared.py
deactivate copy on write as it breaks upsetplot

@@ -6,10 +6,12 @@ from typing import Any, Literal
 from lymph import types
 from lyscripts import utils
 import numpy as np
+import pandas as pd
 
 from . import paths
 
 
+pd.options.mode.copy_on_write = False
 CM_TO_INCH = 0.393701
 GOLDEN_RATIO = 1.61803398875
 
@@ -40,6 +42,16 @@ def get_samples(
     return utils.load_model_samples(samples_path)
 
 
+def get_data(map_t_stage: bool = True) -> pd.DataFrame:
+    """Get the data."""
+    data = utils.load_patient_data(paths.data)
+
+    if map_t_stage:
+        data.ly.map_t_stage(lambda x: "early" if x <= 2 else "late")
+
+    return data
+
+
 def get_fontsizes(base: int = 8, offset: int = 2) -> dict[str, int]:
     """Get fontsizes that can be used to update the matplotlib rcParams.
 
@@ -64,6 +76,7 @@ def get_figsizes(
     unit: Literal["cm", "in"] = "cm",
     constrained_layout: bool = True,
     tight_layout: bool = True,
+    individual: bool = False,
 ) -> dict[str, Any]:
     """Get figure sizes that can be used to update the matplotlib rcParams.
 
@@ -74,8 +87,10 @@ def get_figsizes(
         pad *= CM_TO_INCH
 
     subplot_width = width / ncols
-    subplot_height = subplot_width / aspect_ratio
-    height = subplot_height * nrows
+    height = subplot_width / aspect_ratio
+
+    if not individual:
+        height *= nrows
 
     return {
         "figure.figsize": (width, height),

### 2024-04-04 manuscript.qmd
improve midline ext. formalism

@@
 from matplotlib import ticker
+from matplotlib.colors import LinearSegmentedColormap
+import upsetplot
 
@@
 from lyscripts import utils
+from lyscripts.plot.utils import COLORS
 
 from scripts import shared, paths, COL, CONTRA_LNLS
+
+midline_model = shared.get_model("midline", load_samples=True)
+```
+
+```{python}
+#| tags: [parameters]
+#| echo: false
+
+# width of figures. May depend on number of text columns
+full = 17 # cm
+half = full
+# half =  8 # cm
 ```
@@
 \begin{aligned}
-P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right) &= \sum_{t=0}^{t_\text{max}} p(t) \cdot P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \mid t \right) \\
-&= \sum_{t=0}^{t_\text{max}} p(t) \cdot P \left( \mathbf{X}^\text{i} \mid t \right) \cdot P \left( \mathbf{X}^\text{c} \mid t \right)
+P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right) &= \sum_{t=0}^{t_\text{max}} P(t) \cdot P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \mid t \right) \\
+&= \sum_{t=0}^{t_\text{max}} P(t) \cdot P \left( \mathbf{X}^\text{i} \mid t \right) \cdot P \left( \mathbf{X}^\text{c} \mid t \right)
 \end{aligned}
-$$
+$$ {#eq-bilateral-marginal}
 
@@
 $$
-P \left( \mathbf{X}^\text{i} = \boldsymbol{\xi}_n, \mathbf{X}^\text{c} = \boldsymbol{\xi}_m \right) = \left[ \boldsymbol{\Lambda}^\intercal_\text{i} \cdot \operatorname{diag} p(\mathbf{t}) \cdot \boldsymbol{\Lambda}_\text{c} \right]_{n,m}
-$$
+P \left( \mathbf{X}^\text{i} = \boldsymbol{\xi}_n, \mathbf{X}^\text{c} = \boldsymbol{\xi}_m \right) = \left[ \boldsymbol{\Lambda}^\intercal_\text{i} \cdot \operatorname{diag} P(\mathbf{t}) \cdot \boldsymbol{\Lambda}_\text{c} \right]_{n,m}
+$$ {#eq-bilateral-marginal-algebra}
 
@@
 To account for the increased prevalence of involvement on the contralateral side when the tumor is not clearly lateralized anymore, we also model the tumor's extension over the mid-sagittal line as a binary random variable.
-It starts lateralized and every time-step there is a finite probability $p_\epsilon$ that the tumor grows over the symmetry plane of the patient.
+It starts lateralized and at every time-step there is a finite probability $p_\epsilon$ that the tumor grows over the symmetry plane of the patient.
 
-Technically, the introduction of this additional random variable doubles the space of the hidden states.
+Technically, the introduction of this additional random variable doubles the space of the hidden states and therefore quadruples the size of the transition matrix $\mathbf{A}$.
 However, since we assume no correlation between the tumor's lateralization and the metastases in the LNLs, we can evolve the two parts separately.
 
-As with the state of the lymphatic network, we can also define a small transition matrix $\mathbf{A}_\epsilon$ that evolves the lateralized starting state $\boldsymbol{\epsilon}[0] = \begin{pmatrix} 1 & 0 \end{pmatrix}$:
+The probabilities to find a patient with a clearly lateralized tumor or one that extends over the mid-sagittal line after $t$ time-steps are then given by
 
 $$
-\boldsymbol{\epsilon}[t] = \boldsymbol{\epsilon}[0] \cdot \mathbf{A}_\epsilon^t = \begin{pmatrix} 1 & 0 \end{pmatrix} \cdot \begin{pmatrix} 1 - \epsilon & \epsilon \\ 0 & 1 \end{pmatrix}^t = \begin{pmatrix} (1 - \epsilon)^t & 1 - (1 - \epsilon)^t \end{pmatrix}
+\begin{aligned}
+P(\epsilon = \texttt{False} \mid t) &= (1 - p_\epsilon)^t \\
+P(\epsilon = \texttt{True} \mid t) &= 1 - (1 - p_\epsilon)^t
+\end{aligned}
 $$
 
-We can again collect the entire evolution in a matrix $\boldsymbol{\Lambda}_\epsilon$ that stores the probabilies $P(\boldsymbol{\epsilon} \mid t)$ for each time-step $t$:
+To get the joint probability over the ipsi- and contralateral hidden states, as well as the state of the tumor's midline extension $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)$, we simply add the above terms to the marginalization in @eq-bilateral-marginal:
 
 $$
-\boldsymbol{\Lambda}_\epsilon = P(\boldsymbol{\epsilon} \mid t) = \begin{pmatrix}
-\boldsymbol{\epsilon}[0] \cdot \mathbf{A}_\epsilon^0 \\
-\boldsymbol{\epsilon}[1] \cdot \mathbf{A}_\epsilon^1 \\
-\vdots \\
-\boldsymbol{\epsilon}[t] \cdot \mathbf{A}_\epsilon^t \\
-\end{pmatrix}
+P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right) = \sum_{t=0}^{t_\text{max}} P(t) \cdot P(\epsilon \mid t) \cdot P \left( \mathbf{X}^\text{i} \mid t \right) \cdot P \left( \mathbf{X}^\text{c} \mid t \right)
 $$
 
-This quantity has $t$ rows, one for each time-step, and two columns: one for the decreasing probability of a lateralized tumor and one for the increasing probability of a tumor that has crossed the mid-sagittal line.
+Again, this can be written algebraically, by defining the vector $P(\mathbf{t}, \epsilon) = P(\mathbf{t}) \cdot P(\epsilon \mid \mathbf{t})$, to achieve the same form as in @eq-bilateral-marginal-algebra:
+
+$$
+P \left( \mathbf{X}^\text{i} = \boldsymbol{\xi}_n, \mathbf{X}^\text{c} = \boldsymbol{\xi}_m, \epsilon \right) = \left[ \boldsymbol{\Lambda}^\intercal_\text{i} \cdot \operatorname{diag} P(\mathbf{t}, \epsilon) \cdot \boldsymbol{\Lambda}_\text{c} \right]_{n,m}
+$$
+
+In @fig-midext-evo, we visualize how the prior distribution over diagnose times $P(t)$, the conditional probability of midline extension $P(\epsilon \mid t)$, and their joint $P(\epsilon, t)$ evolve over the course of a patient evolution.
 
@@
 #| label: fig-midext-evo
-#| fig-cap: The matrix $\boldsymbol{\Lambda}_\epsilon$ shown as heatmap of probabilities.
-from matplotlib.colors import LinearSegmentedColormap
-from mpl_toolkits.axes_grid1 import make_axes_locatable
+#| fig-cap: The evolution of the probability for the tumor to remain lateralized (green) or to cross the mid-sagittal plane (red).
+t = np.linspace(0, 10, 11)
+p_midline = {
+  "lateralized": (1 - midline_model.midext_prob)**t,
+  "extension": 1 - (1 - midline_model.midext_prob)**t,
+}
+dist = {
+  "early": midline_model.get_distribution("early").pmf,
+  "late": midline_model.get_distribution("late").pmf,
+}
+p_colors = {"lateralized": COLORS["green"], "extension": COLORS["red"]}
+t_colors = {"early": COLORS["blue"], "late": COLORS["orange"]}
+markers = {}
 
-from lyscripts.plot.utils import COLORS
+nrows, ncols = 2, 1
+plt.rcParams.update(shared.get_fontsizes())
+plt.rcParams.update(shared.get_figsizes(
+  nrows=nrows,
+  ncols=ncols,
+  width=half,
+  aspect_ratio=2.5,
+))
 
-nrows, ncols = 1, 1
+fig, axes = plt.subplots(nrows=nrows, ncols=ncols, sharex=True)
+w = 0.3
 
-plt.rcParams.update(shared.get_fontsizes())
-plt.rcParams.update(shared.get_figsizes(nrows=nrows, ncols=ncols))
+for label, p in p_midline.items():
+  axes[0].plot(
+    t, p_midline[label], "o-",
+    label=label,
+    color=p_colors[label],
+  )
+  for i, t_stage in enumerate(["early", "late"]):
+    if label == "lateralized":
+      axes[0].bar(
+        t + i*w - w/2, dist[t_stage],
+        color=t_colors[t_stage],
+        width=w,
+        label=f"{t_stage} T-cat.",
+      )
+    axes[1].plot(
+      t, p * dist[t_stage], "o-",
+      c=p_colors[label],
+      mfc=t_colors[t_stage],
+      mec=t_colors[t_stage],
+      label=f"{label}, {t_stage}",
+    )
 
-model = shared.get_model(which="midline", load_samples=True)
+axes[0].set_ylim(0., 1.)
+axes[0].yaxis.set_major_formatter(ticker.StrMethodFormatter("{x:.0%}"))
+axes[0].set_ylabel("probability")
+axes[0].grid(axis="y", color=COLORS["gray"], alpha=0.5)
 
-fig, ax = plt.subplots(tight_layout=True)
+axes[0].legend()
 
-cmap = LinearSegmentedColormap.from_list("redusz", [COLORS["gray"], COLORS["red"]])
-im = ax.imshow(model.midext_evo().T, cmap=cmap)
+axes[1].set_xlim(min(t), max(t))
+axes[1].set_xticks(t)
+axes[1].set_xlabel("time-step $t$")
 
-# create an Axes on the right side of ax. The width of cax will be 5%
-# of ax and the padding between cax and ax will be fixed at 0.05 inch.
-divider = make_axes_locatable(ax);
-cax = divider.append_axes("right", size="5%", pad=0.05);
-plt.colorbar(im, cax=cax, label="Prob. of midext");
+axes[1].set_ylim(0., 0.25)
+axes[1].yaxis.set_major_formatter(ticker.StrMethodFormatter("{x:.0%}"))
+axes[1].set_ylabel("probability")
+axes[1].grid(axis="y", color=COLORS["gray"], alpha=0.5)
 
-ax.set_xticks(np.arange(model.max_time + 1));
-ax.set_xlabel("time-step $t$");
+axes[1].legend()
 
-ax.set_yticks([0., 1.], labels=["lat.", "ext."])
+plt.show()
 ```
-
-Just as the lymphatic states, this lateralization state's evolution over the time-steps must then be marginalized over the time of diagnosis using $p(t)$.
-
-- multiply first or second index of evolution with time-prior
-- choose appropriate resulting joint, depending on patient's tumor lateralization
 
@@
 plt.rcParams.update(shared.get_fontsizes())
-plt.rcParams.update(shared.get_figsizes(nrows=nrows, ncols=ncols, aspect_ratio=1.))
+plt.rcParams.update(shared.get_figsizes(
+  nrows=nrows,
+  ncols=ncols,
+  aspect_ratio=1.,
+  width=full,
+))
 
@@
 plt.rcParams.update(shared.get_fontsizes())
-plt.rcParams.update(shared.get_figsizes(aspect_ratio=2.))
+plt.rcParams.update(shared.get_figsizes(aspect_ratio=2., width=half))
 
@@
 - cannot fully represent spread of midline extension and also represent spread params
+
+## Distributions over States
+
+```{python}
+plt.rcdefaults()
+
+mi = pd.MultiIndex.from_product(
+  [[False, True]] * len(midline_model.ext.ipsi.graph.lnls),
+  names=midline_model.ext.ipsi.graph.lnls.keys(),
+)
+series = pd.Series(midline_model.ext.ipsi.state_dist(), index=mi)
+upsetplot.plot(series)
+plt.show()
+```

### 2024-04-04 scripts/shared.py
fix typo

@@ -1,5 +1,5 @@
 """
-Some shared and reduced functions for the scripts.
+Some shared and reused functions for the scripts.
 """
 from typing import Any, Literal
 

### 2024-04-08 manuscript.qmd
add upsetplot

@@
     email: jan.unkelbach@usz.ch
+bibliography: references.bib
 format:
   html:
+    toc: true
+    toc-depth: 3
     code-fold: true
@@
 execute: 
-  cache: false
+  cache: true
 crossref:
@@
   sec-prefix: "section"
+code-links:
+  - text: GitHub Repo
+    href: https://github.com/rmnldwg/bilateral-paper
+    icon: github
+resources: 
+  - data/reduced.csv
+meca-bundle: true
 ---
@@
 
-- head and neck cancer spreads lymphatically
+- blabla
+- head and neck cancer spreads through the lymphatic network
 - may sometimes spread to contralateral side
@@
 
+# Detailed Data on Lymphatic Involvement
+
+To train our models, we used detailed tabular data specifying the lymphatic involvement of individual patients on a per LNL basis.
+
+```{python}
+#| echo: false
+#| label: fig-data-upset
+#| fig-cap: Upset plot of the ipsilateral lymphatic involvement patterns [@lex_upset_2014].
+Here, the horizontal bars depict the overall prevalence of LNL involvemen.
+I.e., LNL II is the only level that is involved in more than 500 patients.
+The vertical bar plots depict joint involvements.
+E.g., the fourth bar from the left depicts how often the LNLs II, III, and IV are jointly involved while level I and V are not.
+This is indicated by the light and dark shaded dots in the matrix underneath.
+#| warning: false
+lnls = ["I", "II", "III", "IV", "V"]
+
+indicator_data = upsetplot.from_indicators(lnls, data=data["max_llh", "ipsi"])
+upset = upsetplot.UpSet(
+  indicator_data,
+  sort_by="cardinality",
+  sort_categories_by="-input",
+  min_subset_size=10,
+  facecolor=COLORS["blue"],
+)
+upset.style_subsets(absent=lnls, facecolor=COLORS["green"], label="N0 (all healthy)")
+axes = upset.plot()
+axes["totals"].set_xlabel("Prevalence")
+axes["intersections"].set_ylabel("Scenario size")
+plt.show()
+```
+
 # Results
@@
 
-The predicted prevalence of midline extension for early T-category tumors is `{python} f"{early_midext_prob:.2%}"`. For a late T-category tumor, it is `{python} f"{late_midext_prob:.2%}"`.
+The predicted prevalence of midline extension for early T-category tumors is <!-- insert computational code later -->. For a late T-category tumor, it is <!-- insert computational code later -->.
 
-In the data, the prevalence of midline extension for early T-category tumors is `{python} f"{early_midext_data:.2%}"`. For late T-category tumors it is `{python} f"{late_midext_data:.2%}"`.
+In the data, the prevalence of midline extension for early T-category tumors is <!-- insert computational code later -->. For late T-category tumors it is <!-- insert computational code later -->.
 

### 2024-04-08 manuscript.qmd
define scenarios, write methods, add upsetplot

@@
 
-## Parameter Symmetries
+## Parameter Symmetries {#sec-params-symmetry}
 
@@
 
-## Tumors Extending Over the Mid-Sagittal Line
+## Tumors Extending Over the Mid-Sagittal Line {#sec-midline}
 
@@
 # Detailed Data on Lymphatic Involvement
+
+- describe origin of data
+- why is this data special?
+- mention [LyProX](https://lyprox.org)
+- visualize data, e.g. using upsetplots
 
@@
 ```
+
+# Methods
+
+We compared three different aprroaches to modelling the contralateral involvement in OPSCC:
+
+1. A naive model of the contralateral side only. Neither the patterns of ipsilateral involvement, nor the information on whether the patients' tumors crossed the mid-sagittal line enter this model.
+2. The bilateral model as introduced in the sections up to and including @sec-params-symmetry. The involvement patterns of both sides enter this model, but not whether the tumors extended the midline.
+3.
+An additional bilateral model that tracks the midline extension of the patients' tumors as a random variable.
+This includes the formalism introduced up to and including @sec-midline.
+In contrast to the other two models, it does consider the reported midline extension in the data to attempt a more accurate prediction of the risk for contralateral occult disease.
+
+- all models initialized with the same settings
+  - same graph
+  - same data
+  - same distributions over diagnose time
+- parameter samples were drawn for all three models:
+  - the sampling used {{< var sampling.walkers_per_dim >}} walkers per parameter space dimension.
+  - Every {{< var sampling.check_interval >}} steps, convergence was checked
+  - convergence was reached when two criteria were met:
+      1. the relative difference between subsequent autocorrelation estimates of the chain does not change more than {{< var sampling.rel_thresh |  >}}.
+      2. the autocorrelation estimate of the chain drops below the length of the chain divided by {{< var sampling.trust_fac >}}.
+  - after convergence, {{< var sampling.nsteps >}} samples (from the parallel walkers) were drawn that were spaced {{< var sampling.thin >}} steps apart, resulting in `num params x {{< var sampling.walkers_per_dim >}} x {{< var sampling.nsteps >}}` parameter samples.
+
+For every model, we compare the predicted prevalence of a certain scenario to how often this scenario is seen in the data.
+The considered scenarios are choosen to test if the models can take into account T-category, the correlation of ipsi- and contralateral involvement, and the primary tumor's lateralization (whether or not it extends over the mid-sagittal line).
+
+- [ ] list and describe the scenarios
 

### 2024-04-09 manuscript.qmd
change to manuscript type & move config

@@
 title: "Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer"
-authors:
-  - name: Roman Ludwig
-    orcid: 0000-0001-9434-328X
-    email: roman.ludwig@usz.ch
-  - name: Yoel Perez Haas
-    email: yoel.perezhaas@usz.ch
-  - name: Jan Unkelbach
-    orcid: 0000-0002-4275-990X
-    email: jan.unkelbach@usz.ch
-bibliography: references.bib
-format:
-  html:
-    toc: true
-    toc-depth: 3
-    code-fold: true
-  pdf:
-    keep-tex: true
-    include-in-header:
-    - text: |
-        \usepackage{multirow}
-        \usepackage{centernot}
-jupyter: python3
-execute: 
-  cache: true
-crossref:
-  fig-prefix: "figure"
-  tbl-prefix: "table"
-  eq-prefix: "equation"
-  sec-prefix: "section"
-code-links:
-  - text: GitHub Repo
-    href: https://github.com/rmnldwg/bilateral-paper
-    icon: github
-resources: 
-  - data/reduced.csv
-meca-bundle: true
 ---
@@
 
-- blabla
 - head and neck cancer spreads through the lymphatic network

### 2024-04-09 manuscript.qmd
expand the unilateral HMM recap

@@
 
-- what we previously did
+![Directed acyclic graph (DAG) representing the abstract lymphatic network in the head and neck region.
+Blue nodes are the LNLs' hidden random variables, the red node represents the tumor, and the orange square nodes depict the binary observed variables.
+Red and blue arcs symbolize the probability of lymphatic spread along that edge during one time-step.
+The orange arcs represent the sensitivity and specificity of the observational modality (e.g.
+CT, MRI, pathology, ...).](static/small-graph.png){#fig-small-graph width=60%}
+
+Our first model to predict the lymphatic progression of HNSCC was introduced using Bayesian networks [@pouymayou_bayesian_2019].
+We subsequently extended this work to a hidden Markov model (HMM) [@ludwig_hidden_2020].
+We will briefly summarize this HMM's formalism before building on it to include the contralateral spread in @sec-ext-to-contra.
+
+We model a patient's state of involvement at an abstract time-step $t$ as a vector of hidden binary random variables:
+
+$$
+\mathbf{X}[t] = \begin{pmatrix} X_v[t] \end{pmatrix} \qquad v \in \left\{ 1, 2, \ldots, V \right\}
+$$
+
+Here, $V$ is the number of LNLs the model considers.
+The values a LNL's hidden binary RV may take on are $X_v[t] = 0$ (`False`), meaning the LNL $v$ is healthy or free of metastatic disease, or $X_v[t] = 1$ (`True`), corresponding to some for of tumor presence (i.e., occult or clinical).
+
+Since the state vector $\mathbf{X}[t]$ is $V$-dimensional and binary, there are $2^V$ distinct possible lymphatic involvement patterns, which we enumerate from $\boldsymbol{\xi}_0 = \begin{pmatrix} 0 & 0 & \cdots & 0 \end{pmatrix}$ to $\boldsymbol{\xi}_{2^V} = \begin{pmatrix} 1 & 1 & \cdots & 1 \end{pmatrix}$.
+
+Any hidden Markov model is fully described by three quantities:
+
+1. A starting state $\mathbf{X}[t=0]$ at time $t=0$ just before the patient's tumor formed. In our case, this is always the state where all LNLs are still healthy $\boldsymbol{\xi}_0$.
+2. The *transition matrix*
+   $$
+   \mathbf{A} = \left( A_{ij} \right) = \big( P \left( \mathbf{X}[t+1] = \boldsymbol{\xi}_j \mid \mathbf{X}[t] = \boldsymbol{\xi}_i \right) \big)
+   $$
+   where the value at row $i$ and column $j$ represents the probability to transition from state $\boldsymbol{\xi}_i$ to $\boldsymbol{\xi}_j$ during the time-step from $t$ to $t+1$.
+Note that we prohibit self-healing, meaning that during a transition, no LNL may change their state from $X_v[t]=1$ to $X_v[t+1]=0$.
+This effectively masks large parts of the transition matrix to be zero.
+3. Lastly, the *observation matrix*
+   $$
+   \mathbf{B} = \left( B_{ij} \right) = \big( P \left( \mathbf{Z} = \boldsymbol{\zeta}_j \mid \mathbf{X}[t_D] = \boldsymbol{\xi}_i \right) \big)
+   $$
+   where in row $i$ and at column $j$ we find the probability to *observe* a lymphatic involvement pattern $\mathbf{Z} = \boldsymbol{\zeta}_j$, given that the true (but hidden) state of involvement at the time of diagnosis $t_D$ is $\mathbf{X}[t_D] = \boldsymbol{\xi}_i$.
+
+The transition matrix $\mathbf{A}$ is parametrized using the spread probabilities of a directed acyclic graph (DAG) that we define as the underlying mechanistic representation of the lymphatic network.
+An example of such a DAG is shown in @fig-small-graph.
+
+...
 
@@
 
-# Extension to a Bilateral Model
+# Extension to a Bilateral Model {#sec-ext-to-contra}
 

### 2024-04-12 manuscript.qmd
start writing on data section

@@
 
-## Incorporating Midline Crossing
+## Modelling Midline Extension {#sec-midline}
 
@@
 P(\epsilon = \texttt{False} \mid t) &= (1 - p_\epsilon)^t \\
-P(\epsilon = \texttt{True} \mid t) &= 1 - (1 - p_\epsilon)^t
+P(\epsilon = \texttt{True} \mid t) &= 1 - P(\epsilon = \texttt{False} \mid t)
 \end{aligned}
@@
 #| label: fig-midext-evo
-#| fig-cap: The evolution of the probability for the tumor to remain lateralized (green) or to cross the mid-sagittal plane (red).
+#| fig-cap: The top panel shows the prior probability to get diagnosed at time-step $t$ for early and late T-category tumors as bars.
+Also in the top panel, we plot the conditional probability of the tumor's midline extension ($\epsilon=\texttt{True}$), given the time-step $t$ as a line plot.
+In the bottom panel, we show the joint probability of getting diagnosed in time-step $t$ *and* having a tumor that crosses the midline.
 t = np.linspace(0, 10, 11)
@@
     t, p_midline[label], "o-",
-    label=label,
+    label=f"cond. prob. $P(\epsilon={label=='extension'} \mid t)$",
     color=p_colors[label],
@@
         width=w,
-        label=f"{t_stage} T-cat.",
+        label=f"{t_stage} T-cat. prior $P(t)$",
       )
@@
       mec=t_colors[t_stage],
-      label=f"{label}, {t_stage}",
+      label=f"{label} ($\epsilon={label=='extension'}$) for {t_stage} T-cat.",
     )
@@
 axes[1].yaxis.set_major_formatter(ticker.StrMethodFormatter("{x:.0%}"))
-axes[1].set_ylabel("probability")
+axes[1].set_ylabel("joint probability $P(\epsilon, t)$")
 axes[1].grid(axis="y", color=COLORS["gray"], alpha=0.5)
@@
 The tumor's spread to the contralateral side in case of an extension over the midline is larger than if it was clearly lateralized, but smaller than its spread to the ipsilateral side.
-This assumption stems from a simple thought experiment: Consider moving the tumor from a clearly lateralized position accross the mid-sagittal plane to its mirror position.
+This assumption stems from a simple thought experiment: Consider moving the tumor from a clearly lateralized position accross the mid-sagittal plane to the same position, but on the contralateral side.
 In the beginning we would have $b_v^\text{c} < b_v^\text{i}$, while in the end, the situation is reversed.
@@
    $$
-   b_v^\text{c,e} = \alpha \cdot b_v^\text{i} + (1 - \alpha) \cdot b_v^{\text{c},\centernot{\text{e}}}
+   b_v^{\text{c},\epsilon=\texttt{True}} = \alpha \cdot b_v^\text{i} + (1 - \alpha) \cdot b_v^{\text{c},\epsilon=\texttt{False}}
    $$ {#eq-mixing}
+   This means the model now uses three different sets of parameters to describe the spread from the tumor to the LNLs: $b^\text{i}_v$ for the spread to the ipsilateral LNLs, $b_v^{\text{c},\epsilon=\texttt{False}}$ for the spread to the contralateral LNLs as long as the tumor is clearly lateralized, and finally $b_v^{\text{c},\epsilon=\texttt{True}}$ when it crosses the midline.
+Note, however, that these three sets of spread rates only account for $2 \cdot 2^V + 1$ parameters, since they are coupled via the mixing parameter $\alpha$.
 
-## Tumors Extending Over the Mid-Sagittal Line {#sec-midline}
-
-- higher prevalence of contralateral involvement when tumor crosses the mid-sagittal plane
-- but still lower than on ipsilateral side
-- spread to contralateral side in case of midline extension may be describes as linear combination of ipsi- and contralateral spread in case of no midline extension.
-- only one additional "mixing parameter" $\alpha$
-- tumor's midline extension modelled as a binary RV (lateralized or crossing/extending)
+Together with the explicit modelling of the tumor's midline extension $\epsilon$ from @sec-midline, we now have a model that may be capable of capturing the higher prevalence of contralateral involvment that comes with tumors extending over the mid-sagittal line.
 
@@
 
-- describe origin of data
+The data used to train our models consists of <!-- `{python} len(data)` --> patients with newly diagnosed oropharyngeal squamous cell carcinomas.
+It reports the involvement of every patient individually and per lymph node level in tabular form, in addition to other clinico-pathological information such as age, T-category, and HPV p16 status.
+Their patient records have been collected at four different instituions:
+
+1. <!-- `{python} len(usz_data)` --> patients from the University Hospital Zurich [@ludwig_dataset_2022].
+2.
+
 - why is this data special?

### 2024-04-15 manuscript.qmd
continue with data section/table

@@
 
-data = pd.read_csv(paths.data, header=[0,1,2])
+raw = pd.read_csv(paths.data, header=[0,1,2])
+data = raw.copy()
 data[COL.t_stage] = data[COL.t_stage].apply(lambda x: "early" if x <= 2 else "late")
@@
 
-The data used to train our models consists of <!-- `{python} len(data)` --> patients with newly diagnosed oropharyngeal squamous cell carcinomas.
+The data used to train our models consists of <!-- {python} len(data) --> patients with newly diagnosed oropharyngeal squamous cell carcinomas.
 It reports the involvement of every patient individually and per lymph node level in tabular form, in addition to other clinico-pathological information such as age, T-category, and HPV p16 status.
@@
 
-1. <!-- `{python} len(usz_data)` --> patients from the University Hospital Zurich [@ludwig_dataset_2022].
-2.
+```{python}
+#| echo: false
+#| label: tbl-data
+
+col_map = {
+  ("patient", "#", "institution"): ("Institution", lambda x: None),
+  ("patient", "#", "age"): ("Age", np.mean),
+  ("tumor", "1", "t_stage"): ("T-category", np.mean),
+}
+reduced = raw[col_map.keys()]
+reduced.columns = reduced.columns.get_level_values(2)
+renamed = (
+  reduced
+  .rename(columns={tpl[-1]: val[0] for tpl, val in col_map.items()})
+  .convert_dtypes()
+)
+grouped = renamed.groupby(by="Institution")
+aggregated = grouped.agg({
+  name: func for name, func in col_map.values()
+  if name != "Institution"
+})
+```
+
+```{python}
+#| echo: false
+#| output: asis
+
+for i, institution in enumerate([
+  "University Hospital Zurich",
+  "Centre Léon Bérard",
+  "Inselspital Bern",
+  "Vall d'Hebron Barcelona Hospital",
+]):
+  is_from_institution = data["patient", "#", "institution"] == institution
+  inst_data = data.loc[is_from_institution]
+  locations = inst_data["tumor", "1", "location"].unique()
+  locations = locations[~pd.isna(locations)]
+
+  if len(locations) == 1:
+    locations_str = str(locations[0])
+  else:
+    locations_str = ", ".join(locations[:-1]) + ", and " + str(locations[-1])
+
+  print(f"{i}. {len(inst_data)} patients with newly diagnosed HNSCC in the {locations_str}, extracted at the {institution}.")
+```
 

### 2024-04-16 manuscript.qmd
add data overview table

--- 2024-04-12 manuscript.qmd
@@
 (
-  involved.style
+  involved
+  .reset_index()
+  .style
   .format(precision=2)
   .apply(right_align, axis="index")
+  .hide()
 )
@@
 
-The data used to train our models consists of <!-- `{python} len(data)` --> patients with newly diagnosed oropharyngeal squamous cell carcinomas.
+The data used to train our models consists of <!-- num patients --> patients with newly diagnosed oropharyngeal squamous cell carcinomas.
 It reports the involvement of every patient individually and per lymph node level in tabular form, in addition to other clinico-pathological information such as age, T-category, and HPV p16 status.
-Their patient records have been collected at four different instituions:
 
-1. <!-- `{python} len(usz_data)` --> patients from the University Hospital Zurich [@ludwig_dataset_2022].
-2.
+Their patient records have been collected at four different institutions and a brief overview over some of their patients' characteristics are shown in @tbl-data.
+Note that the data from the Inselspital Bern and the Centre Léon Bérard only consists of patients treated with some form of neck dissection.
+Since this treatment is more commonly chosen for early T-category patients, they also make up a larger portion of the respective dataset.
 
-- why is this data special?
-- mention [LyProX](https://lyprox.org)
-- visualize data, e.g. using upsetplots
+The entire data, including additional patients with tumors in other primary locations than the oropharynx, is publicly available: It may be [downloaded from LyProX](https://lyprox.org/patients/dataset) where it may also be interactively explored, [from GitHub](https://github.com/rmnldwg/lydata), [from zenodo](https://zenodo.org/search?q=lydata), or via the *Data-in-Brief* publications @ludwig_dataset_2022 and @ludwig_multicentric_2023.
 
-To train our models, we used detailed tabular data specifying the lymphatic involvement of individual patients on a per LNL basis.
+```{python}
+#| echo: false
+#| label: tbl-data
+#| tbl-cap: Overview over the five datasets from four different institutions used to train and evaluate our model.
+Here, we briefly characterize the total number of OPSCC patients from the respective institution, their median age, what proportion received some form of neck dissection, the N0 portion of patients, what percentage presented with early T-category, and the prevalence of primary tumor midline extension.
+For a much more detailed look at the data, visit [lyprox.org](https://lyprox.org).
+col_map = {
+  COL.inst: "Institution",
+  COL.age: "Age",
+  COL.nd: "Neck Dissection",
+  COL.t_stage: "T-category",
+  COL.n_stage: "N-category",
+}
+short_col_map = {tpl[-1]: val for tpl, val in col_map.items()}
+
+subdata = raw[col_map.keys()]
+subdata.columns = subdata.columns.get_level_values(2)
+subdata = (
+  subdata
+  .rename(columns=short_col_map)
+  .reset_index(drop=True)
+)
+
+def n0_mean(n_stage: pd.Series) -> float:
+  """Compute portion of N0 patients."""
+  return (n_stage == 0).mean()
+
+def early_mean(t_stage: pd.Series) -> float:
+  """Compute early T-category portion."""
+  return t_stage.isin([0,1,2]).mean()
+
+grouped = (
+  raw.groupby(
+    by=COL.inst,
+  ).aggregate(**{
+    "Total": pd.NamedAgg(column=COL.age, aggfunc="count"),
+    "Age (median)": pd.NamedAgg(column=COL.age, aggfunc="median"),
+    "Neck Dissection": pd.NamedAgg(column=COL.nd, aggfunc="mean"),
+    "N0": pd.NamedAgg(column=COL.n_stage, aggfunc=n0_mean),
+    "Early T-Cat.": pd.NamedAgg(column=COL.t_stage, aggfunc=early_mean),
+    "Mid. Ext.": pd.NamedAgg(column=COL.midext, aggfunc="mean"),
+  }).convert_dtypes()
+)
+grouped.index.name = "Institution"
+(
+  grouped
+  .reset_index()
+  .style
+  .format(
+    formatter="{:>.0%}",
+    subset=["Neck Dissection", "N0", "Early T-Cat.", "Mid. Ext."],
+  )
+  .apply(
+    func=right_align,
+    axis="index",
+    subset=grouped.columns[0:],
+  )
+  .hide()
+)
+```
+
+To the best of our knowledge, this data still remains the only publicly available data on patterns of lymphatic progression containing this level of detail.
+It allows us to visualize these patterns in so-called *upset plots* [@lex_upset_2014] as shown in @fig-data-upset.
 

### 2024-04-16 scripts/__init__.py

@@ -4,9 +4,22 @@ Scripts and other reused code for this paper.
 from collections import namedtuple
 
 
-Columns = namedtuple("Columns", ["t_stage", "midext", "ipsi_III"])
+Columns = namedtuple(
+    "Columns", [
+        "inst",
+        "age",
+        "nd",
+        "t_stage",
+        "n_stage",
+        "midext",
+        "ipsi_III",
+    ])
 COL = Columns(
+    inst=("patient", "#", "institution"),
+    age=("patient", "#", "age"),
+    nd=("patient", "#", "neck_dissection"),
     t_stage=("tumor", "1", "t_stage"),
+    n_stage=("patient", "#", "n_stage"),
     midext=("tumor", "1", "extension"),
     ipsi_III=("max_llh", "ipsi", "III"),
 )

### 2024-04-16 scripts/paths.py

@@ -7,6 +7,7 @@ from typing import Literal
 data_dir = Path("data")
 data = data_dir / "reduced.csv"
 model_dir = Path("models")
+cache_dir = Path(".cache")
 
 
 def get_filename(

### 2024-04-16 manuscript.qmd
colorize contra table

@@
 #| tbl-cap: Contralateral involvement depending on whether the primary tumor extends over the mid-sagittal line, the T-category, and whether the ipsilateral LNL III was involved or healthy.
-def right_align(column) -> str:
+def right_align(column: pd.Series) -> list[str]:
   """Right align some columns."""
   return ["text-align: right"] * len(column)
+
+def highlight_bool(
+  column: pd.Series,
+  c_false: str = COLORS["green"],
+  c_true: str = COLORS["red"],
+) -> list[str]:
+  """Make cell read (`False`) or green (`True`)."""
+  n = len(column)
+  trues = pd.Series([c_true] * n)
+  falses = pd.Series([c_false] * n)
+  colors = trues.where(column, falses)
+  return colors.map(lambda x: f"color: {x}")
+
+def highlight_t_stage(
+  column: pd.Series,
+  c_early: str = COLORS["green"],
+  c_late: str = COLORS["red"],
+) -> list[str]:
+  """Highlight `early` and `late`."""
+  is_early = column == "early"
+  return highlight_bool(is_early, c_false=c_late, c_true=c_early)
 
@@
   .format(precision=2)
-  .apply(right_align, axis="index")
+  .apply(right_align, subset=involved.columns[0:])
+  .apply(highlight_bool, subset=["Mid. ext.", "ipsi III"])
+  .apply(highlight_t_stage, subset=["T-cat."])
   .hide()

### 2024-04-16 manuscript.qmd
setup pipeline with 3 LNLs

@@
 #| echo: false
+import re
 
@@
 
-Below we compare the predicted with the observed prevalence of some involvement patterns.
+First, we look at the prevalence of contralateral involvement for the LNLs II, III, and IV. This is plotted in @fig-prevalences, both for early and advanced T-category.
+
+- [ ] Run sampling for all LNLs
+- [ ] compute prevalences for set of interesting scenarios
 
@@
 #| tbl-cap: Mean sampled parameter estimates of the midline model and the respective standard deviation.
+def map_param_names(name: str) -> str:
+  """Make parameter names more readable."""
+  try:
+    return {
+      "midext_prob": "Mid. ext. probability",
+      "mixing": "Mixing ⍺",
+      "late_p": "late T-cat. binom. prob.",
+    }[name]
+  except KeyError:
+    return re.sub(
+      r"(ipsi|contra)_", r"\g<1>: ", name
+    ).replace(
+      "to", " ➜ "
+    ).replace(
+      "_spread", ""
+    )
+
 model = shared.get_model(which="midline", load_samples=True)
@@
 
-names = list(model.get_params().keys())
+names = [map_param_names(p) for p in model.get_params()]
 means, stds = samples.mean(axis=0), samples.std(axis=0)
@@
 
-params_table = pd.DataFrame({"name": names, "mean": means, "std": stds})
+params_table = pd.DataFrame({"Parameter": names, "Mean": means, "Std. Dev.": stds})
 (
   params_table.style
-  .format("{:.2%}", subset=["mean", "std"])
-  .apply(right_align, subset=["mean", "std"])
+  .format("{:.2%}", subset=["Mean"])
+  .format("± {:.2%}", subset=["Std. Dev."])
+  .apply(right_align, subset=["Mean", "Std. Dev."])
   .hide()

### 2024-04-16 manuscript.qmd
run new pipeline

@@
 
-for model in ["ipsi", "contra", "bilateral", "midline"]:
+for model in ["contra", "bilateral", "midline"]:
   history = pd.read_csv(f"models/{model}/history.csv").set_index("steps")
@@
 model = shared.get_model(which="midline")
-scenarios_config = utils.load_yaml_params("scenarios.yaml")
+scenarios_config = utils.load_yaml_params("scenarios/base.yaml")
 scenarios = Scenario.list_from_params(scenarios_config)
@@
 series = pd.Series(midline_model.ext.ipsi.state_dist(), index=mi)
-upsetplot.plot(series)
+upsetplot.plot(
+  series,
+  sort_by="cardinality",
+  sort_categories_by="-input",
+)
 plt.show()

### 2024-04-17 manuscript.qmd
start writing results

@@
 
+## MCMC Sampling {#sec-sampling}
+
 - all models initialized with the same settings
@@
 
-First, we look at the prevalence of contralateral involvement for the LNLs II, III, and IV. This is plotted in @fig-prevalences, both for early and advanced T-category.
+First, we look at the prevalence of contralateral involvement for the LNLs II, III, and IV.
+The three models' predictions are plotted against beta posteriors of the observed prevalence in the data in @fig-prevalences, both for early and advanced T-category.
 
-- [ ] Run sampling for all LNLs
-- [ ] compute prevalences for set of interesting scenarios
+The beta posteriors (drawn as lines in @fig-prevalences) are a convenient way of indicating not only the percentage of patients that presented with any given involvement, T-category, and/or other characteristics, which sits at the maximum of the beta distribution.
+It also shows the size of the population via its variance: If we observe e.g.
+3 out of 10 patients to have contralateral LNL II involvement when the primary tumor is of advanced T-stage, then we cannot be particularly sure about that rate to be truly 30%.
+However, when we observe 300 out of 1000, we can be much more certain.
+A beta posterior intuitively reflects that in its width and is thus better suited to display such data than e.g.
+a vertical line only.
+
+The shaded histograms in @fig-prevalences show the corresponding prevalence as computed by one of our models.
+They are formed by computing the prevalence for a number of parameter samples drawn during the MCMC learning process, as detailed in @sec-sampling.
+If both the location and width of the histograms matches those of the beta posteriors, then our model is both accurate and precise enough to describe the data.
 
@@
 #| label: fig-prevalences
-#| fig-cap: Prevalences of different scenarios.
+#| fig-cap: Prevalences of contralateral LNL involvement for different scenarios and the three considered models.
+The colored histograms show the computed distribution over the prevalence of involvment in the contralateral levels II (blue), III (orange), and IV (red).
+Each model's prediction is shown in a separate column, while the rows show these prevalences computed for early (top row) and advanced T-category (bottom row) respectively.
+The lines depict beta-posteriors over the observed prevalence of the corresponding scenario using a uniform prior.
+Its maximum coincides with the percentage of patients with the involvement of interest and its width indicates how many patients were considered in total.
+from itertools import cycle
 from lyscripts.scenario import Scenario
@@
 
-model = shared.get_model(which="midline")
 scenarios_config = utils.load_yaml_params("scenarios/base.yaml")
@@
 
+nrows, ncols = 2, 3
+
 plt.rcParams.update(shared.get_fontsizes())
-plt.rcParams.update(shared.get_figsizes(aspect_ratio=2., width=half))
+plt.rcParams.update(shared.get_figsizes(
+  nrows=nrows,
+  ncols=ncols,
+  width=full,
+))
 
-fig, ax = plt.subplots()
+fig, axes = plt.subplots(nrows=nrows, ncols=ncols, sharex="col")
 
+axes[0,0].set_ylabel("early")
+axes[-1,0].set_ylabel("advanced")
+
+for col, model in enumerate(["contra", "bilateral", "midline"]):
+  axes[0,col].set_title(model)
+  axes[-1,col].set_xlabel("prevalence [%]")
+  for row, t_stage in enumerate(["early", "late"]):
     content = []
+    color_cycle = cycle(COLORS.values())
+    lnls = iter(["II", "III", "IV"])
+    for scenario in scenarios:
+      if not t_stage in scenario.t_stages:
+        continue
+
+      if model == "contra":
+        scenario = scenario.from_dict(
+          scenario.as_dict("prevalences"),
+          is_uni=True,
+          side="contra",
+        )
+
+      color = next(color_cycle)
       content += [
         Histogram.from_hdf5(
-    filename=paths.get_filename(model="midline", kind="prevalences"),
-    dataname=scenarios[0].md5_hash("prevalences"),
-    color=COLORS["blue"],
+          filename=paths.get_filename(model=model, kind="prevalences"),
+          dataname=scenario.md5_hash("prevalences"),
+          color=color,
+          label=f"contra LNL {next(lnls)}"
         ),
         BetaPosterior.from_hdf5(
-    filename=paths.get_filename(model="midline", kind="prevalences"),
-    dataname=scenarios[0].md5_hash("prevalences"),
-    color=COLORS["blue"],
-  ),
-]
-
-content += [
-  Histogram.from_hdf5(
-    filename=paths.get_filename(model="midline", kind="prevalences"),
-    dataname=scenarios[1].md5_hash("prevalences"),
-    color=COLORS["orange"],
-  ),
-  BetaPosterior.from_hdf5(
-    filename=paths.get_filename(model="midline", kind="prevalences"),
-    dataname=scenarios[1].md5_hash("prevalences"),
-    color=COLORS["orange"],
+          filename=paths.get_filename(model=model, kind="prevalences"),
+          dataname=scenario.md5_hash("prevalences"),
+          color=color,
         ),
@@
     draw(
-  axes=ax,
+      axes=axes[row,col],
       contents=content,
       percent_lims=(2., 2.),
+      xlims=(0., 27.)
     )
+    axes[row,col].set_yticks([])
+
+axes[0,0].legend(labelspacing=0.2)
+axes[-1,-1].legend(labelspacing=0.2)
 ```

### 2024-04-17 manuscript.qmd
continue with results

@@
 First, we look at the prevalence of contralateral involvement for the LNLs II, III, and IV.
-The three models' predictions are plotted against beta posteriors of the observed prevalence in the data in @fig-prevalences, both for early and advanced T-category.
+The three models' predictions are plotted against beta posteriors of the observed prevalence in the data in @fig-prevalences-overall, both for early and advanced T-category.
 
-The beta posteriors (drawn as lines in @fig-prevalences) are a convenient way of indicating not only the percentage of patients that presented with any given involvement, T-category, and/or other characteristics, which sits at the maximum of the beta distribution.
+The beta posteriors (drawn as lines in @fig-prevalences-overall) are a convenient way of indicating not only the percentage of patients that presented with any given involvement, T-category, and/or other characteristics, which sits at the maximum of the beta distribution.
 It also shows the size of the population via its variance: If we observe e.g.
@@
 
-The shaded histograms in @fig-prevalences show the corresponding prevalence as computed by one of our models.
+The shaded histograms in @fig-prevalences-overall show the corresponding prevalence as computed by one of our models.
 They are formed by computing the prevalence for a number of parameter samples drawn during the MCMC learning process, as detailed in @sec-sampling.
@@
 #| echo: false
-#| label: fig-prevalences
+#| label: fig-prevalences-overall
 #| fig-cap: Prevalences of contralateral LNL involvement for different scenarios and the three considered models.
@@
 
-scenarios_config = utils.load_yaml_params("scenarios/base.yaml")
+scenarios_config = utils.load_yaml_params("scenarios/overall.yaml")
+scenarios = Scenario.list_from_params(scenarios_config)
+
+nrows, ncols = 2, 3
+
+plt.rcParams.update(shared.get_fontsizes())
+plt.rcParams.update(shared.get_figsizes(
+  nrows=nrows,
+  ncols=ncols,
+  width=full,
+))
+
+fig, axes = plt.subplots(nrows=nrows, ncols=ncols, sharex="col")
+
+axes[0,0].set_ylabel("early")
+axes[-1,0].set_ylabel("advanced")
+
+for col, model in enumerate(["contra", "bilateral", "midline"]):
+  axes[0,col].set_title(model)
+  axes[-1,col].set_xlabel("prevalence [%]")
+  for row, t_stage in enumerate(["early", "late"]):
+    content = []
+    color_cycle = cycle(COLORS.values())
+    lnls = iter(["II", "III", "IV"])
+    for scenario in scenarios:
+      if not t_stage in scenario.t_stages:
+        continue
+
+      if model == "contra":
+        scenario = scenario.from_dict(
+          scenario.as_dict("prevalences"),
+          is_uni=True,
+          side="contra",
+        )
+
+      color = next(color_cycle)
+      content += [
+        Histogram.from_hdf5(
+          filename=paths.get_filename(model=model, kind="prevalences"),
+          dataname=scenario.md5_hash("prevalences"),
+          color=color,
+          label=f"contra LNL {next(lnls)}"
+        ),
+        BetaPosterior.from_hdf5(
+          filename=paths.get_filename(model=model, kind="prevalences"),
+          dataname=scenario.md5_hash("prevalences"),
+          color=color,
+        ),
+      ]
+    draw(
+      axes=axes[row,col],
+      contents=content,
+      percent_lims=(2., 2.),
+      xlims=(0., 27.)
+    )
+    axes[row,col].set_yticks([])
+
+axes[0,0].legend(labelspacing=0.2)
+axes[-1,-1].legend(labelspacing=0.2)
+```
+
+At first, from @fig-prevalences-overall it appears as if the simpler, naive, unilateral-only model describes the prevalence of contralateral involvement better than the bilateral model and even the more complex model that tracks midline extension of the primary tumor.
+
+- [ ] make the data in the paragraph below dynamic
+
+This is mainly due to the early/advanced T-category split of the ipsi- and contralateral side: All models use two different distributions for early and advanced T-category to marginalize over the diagnose times (as introduced in the recap of @sec-unilateral).
+But only the two bilateral models need to balance these distributions to describe the ipsi- and contralateral involvement simultaneaously.
+As can be seen in the second and third column of @fig-prevalences-overall, the prevalence of e.g.
+contralateral LNL II involvement in the data (blue lines) increases from 8% (39 out of 473) to 24% (88 out of 360).
+Which is more than the model predicts (blue histograms).
+Ipsilaterally, however, the involvement of LNL II increases from 71% (343 out of 473) to only 76% (276 out of 360).
+This much weaker increase is well capured by the models.
+
+This issue is quickly resolved, though, as we consider other risk factors that correlate with T-category: Both the ipsilateral involvement and the tumor's extension over the mid-sagittal line influence -- the latter probably even causally -- the extent of contralateral involvement, as shown in @tbl-contra.
+Both these risk factors cannot be considered by the contralateral-only model.
+We should thus expect this naive model to perform worse, as soon as we investigate ipsilateral involvement and midline extension.
+
+```{python}
+#| echo: false
+#| label: fig-prevalences-ipsi
+#| fig-cap: Contralateral prevalences when taking ipsilateral invovlement into account.
+scenarios_config = utils.load_yaml_params("scenarios/with_ipsi.yaml")
 scenarios = Scenario.list_from_params(scenarios_config)

### 2024-04-17 manuscript.qmd
add draft of another prevalence plot

@@
         Histogram.from_hdf5(
-          filename=paths.get_filename(model=model, kind="prevalences"),
+          filename=paths.get_filename(
+            model=model,
+            kind="prevalences_overall",
+          ),
           dataname=scenario.md5_hash("prevalences"),
@@
         BetaPosterior.from_hdf5(
-          filename=paths.get_filename(model=model, kind="prevalences"),
+          filename=paths.get_filename(
+            model=model,
+            kind="prevalences_overall",
+          ),
           dataname=scenario.md5_hash("prevalences"),
@@
         Histogram.from_hdf5(
-          filename=paths.get_filename(model=model, kind="prevalences"),
+          filename=paths.get_filename(
+            model=model,
+            kind="prevalences_with_ipsi",
+          ),
           dataname=scenario.md5_hash("prevalences"),
@@
         BetaPosterior.from_hdf5(
-          filename=paths.get_filename(model=model, kind="prevalences"),
+          filename=paths.get_filename(
+            model=model,
+            kind="prevalences_with_ipsi",
+          ),
           dataname=scenario.md5_hash("prevalences"),

### 2024-04-17 scripts/paths.py

@@ -12,7 +12,7 @@ cache_dir = Path(".cache")
 
 def get_filename(
     model: Literal["ipsi", "contra", "bilateral", "midline"],
-    kind: Literal["params", "samples", "priors", "posteriors", "prevalences", "risks"],
+    kind: Literal["params", "samples", "priors"] | str,
 ) -> Path:
     """Get the filename of the samples of one of the four trained models."""
     return next((model_dir / model).glob(f"{kind}.*"))

### 2024-04-18 data.qmd
put data figs in notebook, restructure main body

---
title: "Visualizing Contralateral Involvement"
---

To motivate our bilateral model extension, we show how the contralateral nodal involvement in the data depends on ipsilateral involvement and the primary tumor's midline extension.

```{python}
#| label: fig-data-strat
#| fig-cap: Contralateral involvement stratified by T-category (left panel), the number of metastatic LNLs ipsilaterally (center panel), and whether the primary tumor extended over the mid-sagittal line or was clearly lateralized (right panel).
import matplotlib.pyplot as plt
from matplotlib import gridspec
from matplotlib.axes import Axes
import numpy as np
import pandas as pd
from lyscripts import utils
from lyscripts.plot.utils import COLORS
from scripts import shared, paths, COL, get_lnl_cols

def group_and_plot(
  df: pd.DataFrame,
  column: str,
  axes: Axes,
  colors: list[str],
) -> None:
  """Group `df` by `column` and plot the result."""
  grouped = df.groupby(by=column)
  counts = grouped.count().T
  aggregated = 100 * grouped.mean().T
  edges = np.linspace(1., 0., len(grouped) + 1)
  positions = edges[1:] + ((edges[0] - edges[1]) / 2)

  for i, ((label, series), (_, count)) in enumerate(zip(
    aggregated.items(),
    counts.items()
  )):
    series.plot(
      ax=axes,
      kind="bar",
      label=f"{label} ({count.iloc[0]})",
      color=colors[i],
      position=positions[i],
      rot=0,
      zorder=2 - i * 0.1
    )

  axes.legend()
  axes.grid(axis="y", color=COLORS["gray"])
  axes.set_axisbelow(True)

nrows, ncols = 1, 3
plt.rcParams.update(**shared.get_fontsizes())
plt.rcParams.update(**shared.get_figsizes(
  nrows=nrows,
  ncols=ncols,
  aspect_ratio=1.1,
))
fig = plt.figure()
gs = gridspec.GridSpec(nrows=nrows, ncols=ncols, figure=fig)

left = fig.add_subplot(gs[0])
center = fig.add_subplot(gs[1], sharey=left)
right = fig.add_subplot(gs[2], sharey=center)
row = fig.add_subplot(gs[:], frame_on=False)

right.set_ylabel("contra prevalence [%]")
right.set_title("Midline Extension")
center.set_title("Ipsilateral Involvement")
left.set_title("T-category")
row.set_xlabel("Lymph Node Level", labelpad=17)
row.set_xticks([])
row.set_yticks([])

raw = utils.load_patient_data(paths.data)
contra_by_midext = raw[[*get_lnl_cols("contra"), COL.midext]].copy()
contra_by_midext.columns = contra_by_midext.columns.droplevel([0,1])

# stratify by midline extension
group_and_plot(
  df=contra_by_midext,
  column="extension",
  axes=right,
  colors=[COLORS["green"], COLORS["red"]],
)

# stratify by ipsilateral involvement
num_ipsi_inv = raw[get_lnl_cols("ipsi")].sum(axis="columns")
contra_by_ipsi = raw[get_lnl_cols("contra")].copy()
contra_by_ipsi.columns = contra_by_ipsi.columns.droplevel([0,1])
contra_by_ipsi["ipsi"] = num_ipsi_inv.map({
  0: "ipsi N0",
  1: "one ipsi LNL",
  2: "≥ two ipsi LNLs",
  3: "≥ two ipsi LNLs",
  4: "≥ two ipsi LNLs",
  5: "≥ two ipsi LNLs",
})

group_and_plot(
  df=contra_by_ipsi,
  column="ipsi",
  axes=center,
  colors=[COLORS["green"], COLORS["orange"], COLORS["red"]],
)

# stratify by T-category
contra_by_t = raw[[*get_lnl_cols("contra"), COL.t_stage]].copy()
contra_by_t.columns = contra_by_t.columns.droplevel([0,1])
contra_by_t["t_stage"] = contra_by_t["t_stage"].map({
  0: "T0-2",
  1: "T0-2",
  2: "T0-2",
  3: "T3-4",
  4: "T3-4",
})

group_and_plot(
  df=contra_by_t,
  column="t_stage",
  axes=left,
  colors=[COLORS["blue"], COLORS["orange"]],
)
```

We also show an upsetplot

```{python}
#| label: fig-data-upset
#| fig-cap: Upset plot of the ipsilateral lymphatic involvement patterns [@lex_upset_2014]. Here, the horizontal bars depict the overall prevalence of LNL involvement. I.e., LNL II is the only level that is involved in more than 500 patients. The vertical bar plots depict joint involvements. E.g., the fourth bar from the left depicts how often the LNLs II, III, and IV are jointly involved while level I and V are not. This is indicated by the light and dark shaded dots in the matrix underneath.
import upsetplot

lnls = [tpl[-1] for tpl in get_lnl_cols("ipsi")]
indicator_data = upsetplot.from_indicators(
  indicators=lnls,
  data=raw["max_llh", "ipsi"],
)
upset = upsetplot.UpSet(
  indicator_data,
  sort_by="cardinality",
  sort_categories_by="-input",
  min_subset_size=10,
  facecolor=COLORS["blue"],
)
upset.style_subsets(
  absent=lnls,
  facecolor=COLORS["green"],
  label="N0 (all healthy)",
)
axes = upset.plot()
axes["totals"].set_xlabel("Prevalence")
axes["intersections"].set_ylabel("Scenario size")
plt.show()
```

### 2024-04-18 manuscript.qmd
put data figs in notebook, restructure main body

@@
 
-from scripts import shared, paths, COL, CONTRA_LNLS
+from scripts import shared, paths, COL, get_lnl_cols
 
@@
 - spread probabilities from tumor to contralateral LNLs in case of midline extesnion are linear combinations of these probabilities in case of ipsilateral spread and contralateral spread when tumor is clearly lateralized
+
+# Data on Lymphatic Progression Patterns
+
+We have collected a detailed dataset of <!-- num patients --> patients with newly diagnosed oropharyngeal squamous cell carcinomas.
+It reports the involvement of every patient individually and per lymph node level in tabular form, in addition to other clinico-pathological information such as age, T-category, and HPV p16 status.
+
+Their patient records have been collected at four different institutions and a brief overview over some of their patients' characteristics are shown in @tbl-data-overview.
+Note that the data from the Inselspital Bern and the Centre Léon Bérard only consists of patients treated with some form of neck dissection.
+Since this treatment is more commonly chosen for early T-category patients, they also make up a larger portion of the respective dataset.
+
+```{python}
+#| echo: false
+#| label: tbl-data-overview
+#| tbl-cap: Overview over the five datasets from four different institutions used to train and evaluate our model.
+Here, we briefly characterize the total number of OPSCC patients from the respective institution, their median age, what proportion received some form of neck dissection, the N0 portion of patients, what percentage presented with early T-category, and the prevalence of primary tumor midline extension.
+For a much more detailed look at the data, visit [lyprox.org](https://lyprox.org).
+from typing import Literal
+
+def align(
+  column: pd.Series,
+  where: Literal["right", "left", "center"],
+) -> list[str]:
+  """Align a column."""
+  return [f"text-align: {where}"] * len(column)
+
+def right_align(column: pd.Series) -> list[str]:
+  """Right align column."""
+  return align(column, where="right")
+
+def highlight_bool(
+  column: pd.Series,
+  c_false: str = COLORS["green"],
+  c_true: str = COLORS["red"],
+) -> list[str]:
+  """Make cell read (`False`) or green (`True`)."""
+  n = len(column)
+  trues = pd.Series([c_true] * n)
+  falses = pd.Series([c_false] * n)
+  colors = trues.where(column, falses)
+  return colors.map(lambda x: f"color: {x}")
+
+def highlight_t_stage(
+  column: pd.Series,
+  c_early: str = COLORS["green"],
+  c_late: str = COLORS["red"],
+) -> list[str]:
+  """Highlight `early` and `late`."""
+  is_early = column == "early"
+  return highlight_bool(is_early, c_false=c_late, c_true=c_early)
+
+col_map = {
+  COL.inst: "Institution",
+  COL.age: "Age",
+  COL.nd: "Neck Dissection",
+  COL.t_stage: "T-category",
+  COL.n_stage: "N-category",
+}
+short_col_map = {tpl[-1]: val for tpl, val in col_map.items()}
+
+raw = utils.load_patient_data(paths.data)
+subdata = raw[col_map.keys()]
+subdata.columns = subdata.columns.get_level_values(2)
+subdata = (
+  subdata
+  .rename(columns=short_col_map)
+  .reset_index(drop=True)
+)
+
+def n0_mean(n_stage: pd.Series) -> float:
+  """Compute portion of N0 patients."""
+  return (n_stage == 0).mean()
+
+def early_mean(t_stage: pd.Series) -> float:
+  """Compute early T-category portion."""
+  return t_stage.isin([0,1,2]).mean()
+
+grouped = (
+  raw.groupby(
+    by=COL.inst,
+  ).aggregate(**{
+    "Total": pd.NamedAgg(column=COL.age, aggfunc="count"),
+    "Age (median)": pd.NamedAgg(column=COL.age, aggfunc="median"),
+    "Neck Dissection": pd.NamedAgg(column=COL.nd, aggfunc="mean"),
+    "N0": pd.NamedAgg(column=COL.n_stage, aggfunc=n0_mean),
+    "Early T-Cat.": pd.NamedAgg(column=COL.t_stage, aggfunc=early_mean),
+    "Mid. Ext.": pd.NamedAgg(column=COL.midext, aggfunc="mean"),
+  }).convert_dtypes()
+)
+grouped.index.name = "Institution"
+(
+  grouped
+  .reset_index()
+  .style
+  .format(
+    formatter="{:>.0%}",
+    subset=["Neck Dissection", "N0", "Early T-Cat.", "Mid. Ext."],
+  )
+  .apply(
+    func=right_align,
+    axis="index",
+    subset=grouped.columns[0:],
+  )
+  .hide()
+)
+```
+
+## Contralateral Involvement Prevalence {#sec-data-strat}
+
+These datasets allow us to investigate correlations between the involvement of individual LNLs, or between risk factors and patterns of involvement.
+In @fig-data-strat, we have plotted the prevalence of each contralateral LNL's involvement, stratified by T-category, ipsilateral number of involved LNLs, and whether the tumor extended over the mid-sagittal line.
+A similar but more complete stratification is also tabulated in the appendix in @tbl-data-strat.
+
+{{< embed data.qmd#fig-data-strat >}}
+
+The left panel in @fig-data-strat indicates that T-category is correlated with contralateral involvement (as it is with overal involvement).
+This is simply because T-category may on average be considered a surrogate for the time between onset of disease and diagnosis.
+I.e., a patient with a T4 tumor was -- on average -- diagnosed later than a patient with a T1 tumor.
+Thus, the former did have more time to develop metastases.
+
+Similarly, ipsilateral involvement correlates with contralateral metastasis.
+The tumor of a patients with many metastases in ipsilateral LNLs was probably able to spread for longer (or faster) compared to a tumor in a patient with no nodal disease.
+This, too, may therefore be considered a surrogate for the duration of the disease.
+In rare cases, bulky nodal disease ipsilaterally may also redirect lymph fluids to the contralateral side.
+<!-- citation needed?
+-->
+
+Lastly, the right panel in @fig-data-strat shows that patients with a tumor crossing the mid-sagittal line show contralateral involvement vastly more often compared to patients with clearly lateralized tumors.
+This makes intutitive sense, because the lymphatic system in the head and neck region is typically symmetric and thus no major vessels cross the midline.
+Therefore, interstitial fluids from the primary tumor -- which we assume to carry living malignant cells -- may only reach the blind-ended lymphatic vessels in the contralateral neck via short-ranged diffusion.
+Which in turn is only possible when the primary tumor is close enough to the mid-sagittal line or crosses it.
+<!-- citation(s) needed?
+or is this intuition good enough?
+-->
+
+## Requirements for a Bilateral Model
+
+Based on the observations of the [previous section](#sec-data-strat), any potential model that aims to also predict the risk for contralateral nodal involvement, should be able to take the following into account:
+
+1.
+More advanced T-category should lead to higher risk for nodal disease.
+One approach to achieve this via the expected time of diagnosis has already been developed in the form of a hidden Markov model [@ludwig_hidden_2021].
+2.
+The degree of ipsilateral involvement should give the model information on the time that may have passed between onset and diagnosis of the disease.
+This should come in addition to what can be inferred about this time from T-category alone.
+3. A tumor that extends over the mid-sagittal line should yield contralateral metastases with much higher probability.
+
+Over the course of this work, we will first briefly recap the mentioned HMM in @sec-unilateral, which was so far used to model ipsilateral lymphatic progression only.
+Then, we intuitively extend it to include the contralateral side as well in @sec-ext-to-contra.
+In this section, we also introduce a way of modelling the tumor's midline extension as a random variable (@sec-midline) and lastly talk about how it may affect the contralateral spread in @sec-params-symmetry.
+
+<!-- TODO: Hint at methods and results... -->
+
+## Data Availability
+
+The entire data, including additional patients with tumors in other primary locations than the oropharynx, is publicly available: It may be [downloaded from LyProX](https://lyprox.org/patients/dataset) where it can be interactively explored too, [from GitHub](https://github.com/rmnldwg/lydata), [from zenodo](https://zenodo.org/search?q=lydata), or via the *Data-in-Brief* publications @ludwig_dataset_2022 and @ludwig_multicentric_2023.
+Although these publications do not include the most recent dataset addition from Vall d'Hebron Barcelona Hospital.
 
@@
 
-...
+Using the introduced quantities, we can evolve the distribution of all possible hidden states from $\mathbf{X}[t=0] = \boldsymbol{\xi}_0$ step by step, by successively multiplying this vector with the transition matrix $\mathbf{A}$.
+At the time of the diagnosis, we multiply the result with the observation matrix $\mathbf{B}$.
+We may then look up the likelihood of a patient presenting with the diagnosis $\mathbf{Z}=\boldsymbol{\zeta}_i$ in the $i$-th entry of the final result.
+
+However, the remaining issue is that we do not know how many time-steps to apply.
+Although a time-step should be considered an abstract quantity that is relative to the tumor's time-frame, we have no way of measuring it directly.
+Hence, we rely on a surrogate: T-category depends mostly on the size of the tumor [@brierley_tnm_2017] and may therefore represent for how long it has been growing and also spreading lymphatically.
+
+We still want to allow for cases where early T-category tumors have already spread extesively to the lymph nodes or where advanced T-category patients present with an N0 neck.
+These cases may be relatively rare, but not implausible.
+Hence, we marginalize over the time of diagnosis and choose different distributions for different (groups of) T-categories.
+
+<!-- TODO: Explain here what role our model plays in a Bayesian context. -->
 
@@
 However, this approach lacks a way to describe the correlation between ipsi- and contralateral involvement.
-This is displayed in @tbl-contra and shows how often the contralateral LNLs I, II, III, and IV were involved, given all possible combinations of midline extension, T-category, and ipsilateral LNL III involvement.
+This is displayed in @tbl-data-strat and shows how often the contralateral LNLs I, II, III, and IV were involved, given all possible combinations of midline extension, T-category, and ipsilateral LNL III involvement.
 Unsurprisingly, the prevalence for contralateral involvement is consistently higher when the tumor extends over the mid-sagittal line or is of later T-category.
 But it is also more frequent when the ipsilateral side shows more severe involvement, which is here shown via the surrogate LNL III.
-
-```{python}
-#| echo: false
-#| label: tbl-contra
-#| tbl-cap: Contralateral involvement depending on whether the primary tumor extends over the mid-sagittal line, the T-category, and whether the ipsilateral LNL III was involved or healthy.
-def right_align(column: pd.Series) -> list[str]:
-  """Right align some columns."""
-  return ["text-align: right"] * len(column)
-
-def highlight_bool(
-  column: pd.Series,
-  c_false: str = COLORS["green"],
-  c_true: str = COLORS["red"],
-) -> list[str]:
-  """Make cell read (`False`) or green (`True`)."""
-  n = len(column)
-  trues = pd.Series([c_true] * n)
-  falses = pd.Series([c_false] * n)
-  colors = trues.where(column, falses)
-  return colors.map(lambda x: f"color: {x}")
-
-def highlight_t_stage(
-  column: pd.Series,
-  c_early: str = COLORS["green"],
-  c_late: str = COLORS["red"],
-) -> list[str]:
-  """Highlight `early` and `late`."""
-  is_early = column == "early"
-  return highlight_bool(is_early, c_false=c_late, c_true=c_early)
-
-raw = pd.read_csv(paths.data, header=[0,1,2])
-data = raw.copy()
-data[COL.t_stage] = data[COL.t_stage].apply(lambda x: "early" if x <= 2 else "late")
-early_midext_data = (data.loc[data[COL.t_stage] == "early", COL.midext] == True).mean()
-late_midext_data = (data.loc[data[COL.t_stage] == "late", COL.midext] == True).mean()
-grouped = data.groupby(by=[COL.midext, COL.t_stage, COL.ipsi_III])
-
-total = grouped.count()[CONTRA_LNLS]["max_llh", "contra"]
-idx = total.index.rename(["Mid. ext.", "T-cat.", "ipsi III"])
-
-num_involved = grouped.sum()[CONTRA_LNLS]["max_llh", "contra"]
-percent_involved = num_involved / total
-total.index = idx
-num_involved.index = idx
-percent_involved.index = idx
-
-involved = num_involved.join(
-  100 * percent_involved,
-  rsuffix=" (%)",
-).sort_index(axis="columns")
-involved.columns = pd.MultiIndex.from_product(
-  [["I", "II", "III", "IV"],
-   ["n", "%"]],
-  names=["LNL", ""]
-)
-(
-  involved
-  .reset_index()
-  .style
-  .format(precision=2)
-  .apply(right_align, subset=involved.columns[0:])
-  .apply(highlight_bool, subset=["Mid. ext.", "ipsi III"])
-  .apply(highlight_t_stage, subset=["T-cat."])
-  .hide()
-)
-```
 
@@
 Together with the explicit modelling of the tumor's midline extension $\epsilon$ from @sec-midline, we now have a model that may be capable of capturing the higher prevalence of contralateral involvment that comes with tumors extending over the mid-sagittal line.
-
-# Detailed Data on Lymphatic Involvement
-
-The data used to train our models consists of <!-- num patients --> patients with newly diagnosed oropharyngeal squamous cell carcinomas.
-It reports the involvement of every patient individually and per lymph node level in tabular form, in addition to other clinico-pathological information such as age, T-category, and HPV p16 status.
-
-Their patient records have been collected at four different institutions and a brief overview over some of their patients' characteristics are shown in @tbl-data.
-Note that the data from the Inselspital Bern and the Centre Léon Bérard only consists of patients treated with some form of neck dissection.
-Since this treatment is more commonly chosen for early T-category patients, they also make up a larger portion of the respective dataset.
-
-The entire data, including additional patients with tumors in other primary locations than the oropharynx, is publicly available: It may be [downloaded from LyProX](https://lyprox.org/patients/dataset) where it may also be interactively explored, [from GitHub](https://github.com/rmnldwg/lydata), [from zenodo](https://zenodo.org/search?q=lydata), or via the *Data-in-Brief* publications @ludwig_dataset_2022 and @ludwig_multicentric_2023.
-
-```{python}
-#| echo: false
-#| label: tbl-data
-#| tbl-cap: Overview over the five datasets from four different institutions used to train and evaluate our model.
-Here, we briefly characterize the total number of OPSCC patients from the respective institution, their median age, what proportion received some form of neck dissection, the N0 portion of patients, what percentage presented with early T-category, and the prevalence of primary tumor midline extension.
-For a much more detailed look at the data, visit [lyprox.org](https://lyprox.org).
-col_map = {
-  COL.inst: "Institution",
-  COL.age: "Age",
-  COL.nd: "Neck Dissection",
-  COL.t_stage: "T-category",
-  COL.n_stage: "N-category",
-}
-short_col_map = {tpl[-1]: val for tpl, val in col_map.items()}
-
-subdata = raw[col_map.keys()]
-subdata.columns = subdata.columns.get_level_values(2)
-subdata = (
-  subdata
-  .rename(columns=short_col_map)
-  .reset_index(drop=True)
-)
-
-def n0_mean(n_stage: pd.Series) -> float:
-  """Compute portion of N0 patients."""
-  return (n_stage == 0).mean()
-
-def early_mean(t_stage: pd.Series) -> float:
-  """Compute early T-category portion."""
-  return t_stage.isin([0,1,2]).mean()
-
-grouped = (
-  raw.groupby(
-    by=COL.inst,
-  ).aggregate(**{
-    "Total": pd.NamedAgg(column=COL.age, aggfunc="count"),
-    "Age (median)": pd.NamedAgg(column=COL.age, aggfunc="median"),
-    "Neck Dissection": pd.NamedAgg(column=COL.nd, aggfunc="mean"),
-    "N0": pd.NamedAgg(column=COL.n_stage, aggfunc=n0_mean),
-    "Early T-Cat.": pd.NamedAgg(column=COL.t_stage, aggfunc=early_mean),
-    "Mid. Ext.": pd.NamedAgg(column=COL.midext, aggfunc="mean"),
-  }).convert_dtypes()
-)
-grouped.index.name = "Institution"
-(
-  grouped
-  .reset_index()
-  .style
-  .format(
-    formatter="{:>.0%}",
-    subset=["Neck Dissection", "N0", "Early T-Cat.", "Mid. Ext."],
-  )
-  .apply(
-    func=right_align,
-    axis="index",
-    subset=grouped.columns[0:],
-  )
-  .hide()
-)
-```
-
-To the best of our knowledge, this data still remains the only publicly available data on patterns of lymphatic progression containing this level of detail.
-It allows us to visualize these patterns in so-called *upset plots* [@lex_upset_2014] as shown in @fig-data-upset.
-
-```{python}
-#| echo: false
-#| label: fig-data-upset
-#| fig-cap: Upset plot of the ipsilateral lymphatic involvement patterns [@lex_upset_2014].
-Here, the horizontal bars depict the overall prevalence of LNL involvemen.
-I.e., LNL II is the only level that is involved in more than 500 patients.
-The vertical bar plots depict joint involvements.
-E.g., the fourth bar from the left depicts how often the LNLs II, III, and IV are jointly involved while level I and V are not.
-This is indicated by the light and dark shaded dots in the matrix underneath.
-#| warning: false
-lnls = ["I", "II", "III", "IV", "V"]
-
-indicator_data = upsetplot.from_indicators(lnls, data=data["max_llh", "ipsi"])
-upset = upsetplot.UpSet(
-  indicator_data,
-  sort_by="cardinality",
-  sort_categories_by="-input",
-  min_subset_size=10,
-  facecolor=COLORS["blue"],
-)
-upset.style_subsets(absent=lnls, facecolor=COLORS["green"], label="N0 (all healthy)")
-axes = upset.plot()
-axes["totals"].set_xlabel("Prevalence")
-axes["intersections"].set_ylabel("Scenario size")
-plt.show()
-```
 
@@
 
-This issue is quickly resolved, though, as we consider other risk factors that correlate with T-category: Both the ipsilateral involvement and the tumor's extension over the mid-sagittal line influence -- the latter probably even causally -- the extent of contralateral involvement, as shown in @tbl-contra.
+This issue is quickly resolved, though, as we consider other risk factors that correlate with T-category: Both the ipsilateral involvement and the tumor's extension over the mid-sagittal line influence -- the latter probably even causally -- the extent of contralateral involvement, as shown in @tbl-data-strat.
 Both these risk factors cannot be considered by the contralateral-only model.
@@
 ```
+
+# Acknowledgement {.appendix}
+
+This work was supported by:
+
+- the Clinical Research Priority Program "Artificial Intelligence in Oncological Imaging" of the University of Zurich
+- the Swiss Cancer Research Foundation under grant number KFS 5645-08-2022
+
+# Appendix {.appendix}
+
+```{python}
+#| echo: false
+#| label: tbl-data-strat
+#| tbl-cap: Contralateral involvement depending on whether the primary tumor extends over the mid-sagittal line, the T-category, and whether the ipsilateral LNL III was involved or healthy.
+data = raw.copy()
+data[COL.t_stage] = data[COL.t_stage].apply(lambda x: "early" if x <= 2 else "late")
+early_midext_data = (data.loc[data[COL.t_stage] == "early", COL.midext] == True).mean()
+late_midext_data = (data.loc[data[COL.t_stage] == "late", COL.midext] == True).mean()
+grouped = data.groupby(by=[COL.midext, COL.t_stage, COL.ipsi_III])
+
+lnl_cols = get_lnl_cols("contra", lnls=["I", "II", "III", "IV"])
+total = grouped.count()[lnl_cols]["max_llh", "contra"]
+idx = total.index.rename(["Mid. ext.", "T-cat.", "ipsi III"])
+
+num_involved = grouped.sum()[lnl_cols]["max_llh", "contra"]
+percent_involved = num_involved / total
+total.index = idx
+num_involved.index = idx
+percent_involved.index = idx
+
+involved = num_involved.join(
+  100 * percent_involved,
+  rsuffix=" (%)",
+).sort_index(axis="columns")
+involved.columns = pd.MultiIndex.from_product(
+  [["I", "II", "III", "IV"],
+   ["n", "%"]],
+  names=["LNL", ""]
+)
+(
+  involved
+  .reset_index()
+  .style
+  .format(precision=2)
+  .apply(right_align, subset=involved.columns[0:])
+  .apply(highlight_bool, subset=["Mid. ext.", "ipsi III"])
+  .apply(highlight_t_stage, subset=["T-cat."])
+  .hide()
+)
+```

### 2024-04-19 manuscript.qmd
start pulling out model notebook

@@
 
-```{python}
-#| echo: false
-#| label: fig-midext-evo
-#| fig-cap: The top panel shows the prior probability to get diagnosed at time-step $t$ for early and late T-category tumors as bars.
-Also in the top panel, we plot the conditional probability of the tumor's midline extension ($\epsilon=\texttt{True}$), given the time-step $t$ as a line plot.
-In the bottom panel, we show the joint probability of getting diagnosed in time-step $t$ *and* having a tumor that crosses the midline.
-t = np.linspace(0, 10, 11)
-p_midline = {
-  "lateralized": (1 - midline_model.midext_prob)**t,
-  "extension": 1 - (1 - midline_model.midext_prob)**t,
-}
-dist = {
-  "early": midline_model.get_distribution("early").pmf,
-  "late": midline_model.get_distribution("late").pmf,
-}
-p_colors = {"lateralized": COLORS["green"], "extension": COLORS["red"]}
-t_colors = {"early": COLORS["blue"], "late": COLORS["orange"]}
-markers = {}
-
-nrows, ncols = 2, 1
-plt.rcParams.update(shared.get_fontsizes())
-plt.rcParams.update(shared.get_figsizes(
-  nrows=nrows,
-  ncols=ncols,
-  width=half,
-  aspect_ratio=2.5,
-))
-
-fig, axes = plt.subplots(nrows=nrows, ncols=ncols, sharex=True)
-w = 0.3
-
-for label, p in p_midline.items():
-  axes[0].plot(
-    t, p_midline[label], "o-",
-    label=f"cond. prob. $P(\epsilon={label=='extension'} \mid t)$",
-    color=p_colors[label],
-  )
-  for i, t_stage in enumerate(["early", "late"]):
-    if label == "lateralized":
-      axes[0].bar(
-        t + i*w - w/2, dist[t_stage],
-        color=t_colors[t_stage],
-        width=w,
-        label=f"{t_stage} T-cat. prior $P(t)$",
-      )
-    axes[1].plot(
-      t, p * dist[t_stage], "o-",
-      c=p_colors[label],
-      mfc=t_colors[t_stage],
-      mec=t_colors[t_stage],
-      label=f"{label} ($\epsilon={label=='extension'}$) for {t_stage} T-cat.",
-    )
-
-axes[0].set_ylim(0., 1.)
-axes[0].yaxis.set_major_formatter(ticker.StrMethodFormatter("{x:.0%}"))
-axes[0].set_ylabel("probability")
-axes[0].grid(axis="y", color=COLORS["gray"], alpha=0.5)
-
-axes[0].legend()
-
-axes[1].set_xlim(min(t), max(t))
-axes[1].set_xticks(t)
-axes[1].set_xlabel("time-step $t$")
-
-axes[1].set_ylim(0., 0.25)
-axes[1].yaxis.set_major_formatter(ticker.StrMethodFormatter("{x:.0%}"))
-axes[1].set_ylabel("joint probability $P(\epsilon, t)$")
-axes[1].grid(axis="y", color=COLORS["gray"], alpha=0.5)
-
-axes[1].legend()
-
-plt.show()
-```
+{{< embed model.qmd#fig-model-midext-evo >}}
 
@@
       if model == "contra":
-        scenario = scenario.from_dict(
-          scenario.as_dict("prevalences"),
-          is_uni=True,
-          side="contra",
-        )
+        scenario.side = "contra"
+        scenario.is_uni = True
+
+      print(scenario)
 

### 2024-04-19 model.qmd
start pulling out model notebook

---
title: "Illustrating Model Properties"
---

# Diagnose Time Prior and Evolution of Midline Extension

First, we illustrate how the prior distribution over diagnose times and the conditional distribution of having developed a tumor that crosses the mid-sagittal line *given* a diagnosis time evolve together.

```{python}
#| label: fig-model-midext-evo
#| fig-cap: The top panel shows the prior probability to get diagnosed at time-step $t$ for early and late T-category tumors as bars. Also in the top panel, we plot the conditional probability of the tumor's midline extension ($\epsilon=\texttt{True}$), given the time-step $t$ as a line plot. In the bottom panel, we show the joint probability of getting diagnosed in time-step $t$ *and* having a tumor that crosses the midline.
import numpy as np
import matplotlib.pyplot as plt
from matplotlib import ticker
from lyscripts.plot.utils import COLORS
from scripts import shared

midline_model = shared.get_model("midline", load_samples=True)

t = np.linspace(0, 10, 11)
p_midline = {
  "lateralized": (1 - midline_model.midext_prob)**t,
  "extension": 1 - (1 - midline_model.midext_prob)**t,
}
dist = {
  "early": midline_model.get_distribution("early").pmf,
  "late": midline_model.get_distribution("late").pmf,
}
p_colors = {"lateralized": COLORS["green"], "extension": COLORS["red"]}
t_colors = {"early": COLORS["blue"], "late": COLORS["orange"]}
markers = {}

nrows, ncols = 2, 1
plt.rcParams.update(shared.get_fontsizes())
plt.rcParams.update(shared.get_figsizes(
  nrows=nrows,
  ncols=ncols,
  width=17/2,
  aspect_ratio=2.5,
))

fig, axes = plt.subplots(nrows=nrows, ncols=ncols, sharex=True)
w = 0.3

for label, p in p_midline.items():
  axes[0].plot(
    t, p_midline[label], "o-",
    label=f"cond. prob. $P(\epsilon={label=='extension'} \mid t)$",
    color=p_colors[label],
  )
  for i, t_stage in enumerate(["early", "late"]):
    if label == "lateralized":
      axes[0].bar(
        t + i*w - w/2, dist[t_stage],
        color=t_colors[t_stage],
        width=w,
        label=f"{t_stage} T-cat. prior $P(t)$",
      )
    axes[1].plot(
      t, p * dist[t_stage], "o-",
      c=p_colors[label],
      mfc=t_colors[t_stage],
      mec=t_colors[t_stage],
      label=f"{label} ($\epsilon={label=='extension'}$) for {t_stage} T-cat.",
    )

axes[0].set_ylim(0., 1.)
axes[0].yaxis.set_major_formatter(ticker.StrMethodFormatter("{x:.0%}"))
axes[0].set_ylabel("probability")
axes[0].grid(axis="y", color=COLORS["gray"], alpha=0.5)

axes[0].legend()

axes[1].set_xlim(min(t), max(t))
axes[1].set_xticks(t)
axes[1].set_xlabel("time-step $t$")

axes[1].set_ylim(0., 0.25)
axes[1].yaxis.set_major_formatter(ticker.StrMethodFormatter("{x:.0%}"))
axes[1].set_ylabel("joint probability $P(\epsilon, t)$")
axes[1].grid(axis="y", color=COLORS["gray"], alpha=0.5)

axes[1].legend()

plt.show()
```

### 2024-04-22 manuscript.qmd
mv history to models.qmd, add totals to data table

@@
 #| echo: false
+
 import re
@@
 
+def highlight(column: pd.Series, mapping: dict) -> list[str]:
+  """Color based on `mapping`."""
+  colors = column.map(mapping)
+  return colors.map(lambda x: f"color: {x}")
+
 def highlight_bool(
@@
   """Make cell read (`False`) or green (`True`)."""
-  n = len(column)
-  trues = pd.Series([c_true] * n)
-  falses = pd.Series([c_false] * n)
-  colors = trues.where(column, falses)
-  return colors.map(lambda x: f"color: {x}")
+  return highlight(column, mapping={True: c_true, False: c_false})
 
@@
 
-In @fig-midext-evo, we visualize how the prior distribution over diagnose times $P(t)$, the conditional probability of midline extension $P(\epsilon \mid t)$, and their joint $P(\epsilon, t)$ evolve over the course of a patient evolution.
+In @fig-model-midext-evo, we visualize how the prior distribution over diagnose times $P(t)$, the conditional probability of midline extension $P(\epsilon \mid t)$, and their joint $P(\epsilon, t)$ evolve over the course of a patient evolution.
 
@@
 
-```{python}
-#| echo: false
-#| label: fig-burnin-history 
-#| fig-cap: Monitoring quantities during the burn-in phase of the parameter sampling.
-from collections import namedtuple
+{{< embed model.qmd#fig-model-burnin-history >}}
 
-def custom(x, pos):
-  return "{:.1f}".format(x/1000)
-
-nrows, ncols = 1, 3
-label_map = {
-  "acor_times": "autocorrelation [steps]",
-  "accept_fracs": "acceptance fraction [%]",
-  "max_log_probs": r"max. log-likelihood [$\times 10^3$]",
-}
-
-plt.rcParams.update(shared.get_fontsizes())
-plt.rcParams.update(shared.get_figsizes(
-  nrows=nrows,
-  ncols=ncols,
-  aspect_ratio=1.,
-  width=full,
-))
-
-fig, axes = plt.subplots(
-  nrows=nrows,
-  ncols=ncols,
-  sharex=True,
-)
-
-for model in ["contra", "bilateral", "midline"]:
-  history = pd.read_csv(f"models/{model}/history.csv").set_index("steps")
-  for i, column in enumerate(history.columns):
-    history.plot(y=column, ax=axes[i], label=model)
-    axes[i].autoscale(enable=True, tight=True, axis="x")
-    axes[i].set_ylabel(label_map[column])
-    axes[i].xaxis.set_major_formatter(ticker.FuncFormatter(custom))
-    axes[i].set_xlabel(r"steps [$\times 10^3$]")
-
-axes[1].yaxis.set_major_formatter(ticker.StrMethodFormatter("{x:.0%}"))
-axes[2].yaxis.set_major_formatter(ticker.FuncFormatter(custom))
-plt.show()
-```
-
-@Fig-burnin-history is only a temporary placeholder for now.
+@fig-model-burnin-history is only a temporary placeholder for now.
 
@@
         scenario.is_uni = True
-
-      print(scenario)
+      else:
+        scenario.side = "ipsi"
+        scenario.is_uni = False
 
@@
 #| tbl-cap: Contralateral involvement depending on whether the primary tumor extends over the mid-sagittal line, the T-category, and whether the ipsilateral LNL III was involved or healthy.
-data = raw.copy()
-data[COL.t_stage] = data[COL.t_stage].apply(lambda x: "early" if x <= 2 else "late")
-early_midext_data = (data.loc[data[COL.t_stage] == "early", COL.midext] == True).mean()
-late_midext_data = (data.loc[data[COL.t_stage] == "late", COL.midext] == True).mean()
-grouped = data.groupby(by=[COL.midext, COL.t_stage, COL.ipsi_III])
+lnl_cols = get_lnl_cols("contra", lnls=["I", "II", "III", "IV"])
+num_ipsi_inv = raw[get_lnl_cols("ipsi")].sum(axis="columns")
 
-lnl_cols = get_lnl_cols("contra", lnls=["I", "II", "III", "IV"])
-total = grouped.count()[lnl_cols]["max_llh", "contra"]
-idx = total.index.rename(["Mid. ext.", "T-cat.", "ipsi III"])
+contra_inv = raw[lnl_cols].copy()
+contra_inv.columns = contra_inv.columns.droplevel([0,1])
+contra_inv["t_stage"] = raw[COL.t_stage].apply(lambda x: "early" if x <= 2 else "advanced")
+contra_inv["ipsi"] = num_ipsi_inv.map(lambda x: str(x) if x <= 1 else "≥ 2")
+contra_inv["midext"] = raw[COL.midext]
 
-num_involved = grouped.sum()[lnl_cols]["max_llh", "contra"]
+grouped = contra_inv.groupby(by=["t_stage", "ipsi", "midext"])
+num_involved = grouped.sum()
+total = grouped.count()
 percent_involved = num_involved / total
+idx = total.index.rename(["T-cat.", "ipsi", "Mid. ext."])
+
 total.index = idx
@@
 )
+involved["total", "n"] = total["I"]
 (
@@
   .reset_index()
+  .sort_values("ipsi", ascending=True)
+  .sort_values("T-cat.", ascending=False)
   .style
@@
   .apply(right_align, subset=involved.columns[0:])
-  .apply(highlight_bool, subset=["Mid. ext.", "ipsi III"])
   .apply(highlight_t_stage, subset=["T-cat."])
+  .apply(highlight_bool, subset=["Mid. ext."])
+  .apply(highlight, subset=["ipsi"], mapping={
+    "0": COLORS["green"],
+    "1": COLORS["orange"],
+    "≥ 2": COLORS["red"],
+  })
   .hide()

### 2024-04-22 model.qmd
mv history to models.qmd, add totals to data table

@@
 ```
+
+# Sampling History
+
+For bookeeping, we also show how different quantities evolve during the burnin of the sampling. For example, the ratio of proposed samples that were accepted or the autocorrelation time.
+
+```{python}
+#| echo: false
+#| label: fig-model-burnin-history 
+#| fig-cap: Monitoring quantities during the burn-in phase of the parameter sampling.
+from collections import namedtuple
+import pandas as pd
+
+def custom(x, pos):
+  return "{:.1f}".format(x/1000)
+
+nrows, ncols = 1, 3
+label_map = {
+  "acor_times": "autocorrelation [steps]",
+  "accept_fracs": "acceptance fraction [%]",
+  "max_log_probs": r"max. log-likelihood [$\times 10^3$]",
+}
+
+plt.rcParams.update(shared.get_fontsizes())
+plt.rcParams.update(shared.get_figsizes(
+  nrows=nrows,
+  ncols=ncols,
+  aspect_ratio=1.,
+  width=17,
+))
+
+fig, axes = plt.subplots(
+  nrows=nrows,
+  ncols=ncols,
+  sharex=True,
+)
+
+for model in ["contra", "bilateral", "midline"]:
+  history = pd.read_csv(f"models/{model}/history.csv").set_index("steps")
+  for i, column in enumerate(history.columns):
+    history.plot(y=column, ax=axes[i], label=model)
+    axes[i].autoscale(enable=True, tight=True, axis="x")
+    axes[i].set_ylabel(label_map[column])
+    axes[i].xaxis.set_major_formatter(ticker.FuncFormatter(custom))
+    axes[i].set_xlabel(r"steps [$\times 10^3$]")
+
+axes[1].yaxis.set_major_formatter(ticker.StrMethodFormatter("{x:.0%}"))
+axes[2].yaxis.set_major_formatter(ticker.FuncFormatter(custom))
+plt.show()
+```

### 2024-04-22 manuscript.qmd
add prior, llh to formalism

@@
 Our first model to predict the lymphatic progression of HNSCC was introduced using Bayesian networks [@pouymayou_bayesian_2019].
-We subsequently extended this work to a hidden Markov model (HMM) [@ludwig_hidden_2020].
+We subsequently extended this work to a hidden Markov model (HMM) [@ludwig_hidden_2020] to allow an intuitive inclusion of T-category into the predictions.
 We will briefly summarize this HMM's formalism before building on it to include the contralateral spread in @sec-ext-to-contra.
@@
 
-The transition matrix $\mathbf{A}$ is parametrized using the spread probabilities of a directed acyclic graph (DAG) that we define as the underlying mechanistic representation of the lymphatic network.
+Note that the transition matrix $\mathbf{A}$ is parametrized using the spread probabilities of a directed acyclic graph (DAG) that we define as the underlying mechanistic representation of the lymphatic network.
 An example of such a DAG is shown in @fig-small-graph.
@@
 Using the introduced quantities, we can evolve the distribution of all possible hidden states from $\mathbf{X}[t=0] = \boldsymbol{\xi}_0$ step by step, by successively multiplying this vector with the transition matrix $\mathbf{A}$.
-At the time of the diagnosis, we multiply the result with the observation matrix $\mathbf{B}$.
+At the time of the diagnosis $t_D$, we multiply the result with the observation matrix $\mathbf{B}$.
 We may then look up the likelihood of a patient presenting with the diagnosis $\mathbf{Z}=\boldsymbol{\zeta}_i$ in the $i$-th entry of the final result.
 
-However, the remaining issue is that we do not know how many time-steps to apply.
-Although a time-step should be considered an abstract quantity that is relative to the tumor's time-frame, we have no way of measuring it directly.
-Hence, we rely on a surrogate: T-category depends mostly on the size of the tumor [@brierley_tnm_2017] and may therefore represent for how long it has been growing and also spreading lymphatically.
+However, the remaining issue is that the value of $t_D$ is unknown, i.e.
+over how many time-steps the HMM should be evolved.
+We solve this problem by marginalizing over the time of diagnosis.
+Different distributions over the diagnosis times can then be choosen based on T-category.
+For instance, the mean of the time-prior to marginalize over the diagnosis time for early T-category patients $P\left( t_D \mid \text{early} \right)$ may be shifted towards earlier times than the one for advanced T-category patients $P\left( t_D \mid \text{early} \right)$.
+This gives us for example
 
-We still want to allow for cases where early T-category tumors have already spread extesively to the lymph nodes or where advanced T-category patients present with an N0 neck.
-These cases may be relatively rare, but not implausible.
-Hence, we marginalize over the time of diagnosis and choose different distributions for different (groups of) T-categories.
+$$
+P\left( \mathbf{X} \mid \text{T}x = \text{early} \right) = \sum_{t=0}^{t_\text{max}} P \left( \mathbf{X} \mid t \right) \cdot P(t \mid \text{early})
+$$
+
+In this work, we use binomial distributions $\mathfrak{B} \left( t_D, p_{\text{T}x} \right)$ as time-priors which have one free parameter $p_{\text{T}x}$ for each group of patients we differentiate based on T-category.
+Also, we fix $t_\text{max} = 10$, which means that the expected number of time-steps from the onset of a patient's disease to their diagnosis is $\mathbb{E}\left[ t_D \right] = 10 \cdot p_{\text{T}x}$.
+
+## Likelihood Function
+
+With the formalism introduced above, we can write the likelihood function for a patient to present with a diagnosis consisting of an observed state and a T-category $d = \left( \boldsymbol{\zeta}_i, \text{T}x \right)$ as follows:
+
+$$
+\ell = \sum_{t=0}^{t_\text{max}} \left[ \boldsymbol{\xi}_0 \cdot \mathbf{A} \cdot \mathbf{B} \right]_i \cdot P \left( t \mid \text{T}x \right)
+$$ {#eq-single-patient-llh}
+
+Note that it is also possible to account for missing involvement information: If a diagnosis (like fine needle aspiration (FNA)) is only available for a subset of all LNLs, we can sum over all those possible complete observed states $\boldsymbol{\zeta}_j$ that match the provided diagnosis.
+
+The quantity $\ell$ in @eq-single-patient-llh depends on the spread parameters shown in @fig-small-graph via the transition matrix $\mathbf{A}$ and on the binomial parameters $p_{\text{T}x}$ via time-priors.
+In this work, we will only differentiate between "early" (T1 & T2) and "advanced" (T3 & T4) T-categories.
+Therefore, our parameter space is:
+
+$$
+\boldsymbol{\theta} = \left( \left\{ b_v \right\}, \left\{ t_{vr} \right\}, \left\{p_\text{T}x \right\} \right) \quad \text{with} \quad \genfrac{}{}{0pt}{2}{v\leq V}{r\in\operatorname{pa}(v)}
+$$
+
+And it is our goal to infer the values of these parameters for a given dataset $\mathcal{D} = \left( d_1, d_2, \ldots, d_N \right)$ of OPSCC patients.
+The likelihood of these $N$ diagnoses is simply the product of their individual likelihoods as defined in @eq-single-patient-llh.
+For numerical reasons, we typically compute the data likelihood in log space:
+
+$$
+\log \mathcal{L} \left( \mathcal{D} \mid \boldsymbol{\theta} \right) = \sum_{i=1}^N \log \ell_i
+$$
+
+## Model Prediction in the Bayesian Context
 

### 2024-04-22 model.qmd
add prior, llh to formalism

@@
   width=17/2,
-  aspect_ratio=2.5,
 ))
@@
 
-axes[0].legend()
+axes[0].legend(labelspacing=0.2)
 
@@
 
-axes[1].legend()
+axes[1].legend(labelspacing=0.2)
 

### 2024-04-23 manuscript.qmd
finalize uni model recap

@@
 \mathbf{X}[t] = \begin{pmatrix} X_v[t] \end{pmatrix} \qquad v \in \left\{ 1, 2, \ldots, V \right\}
-$$
+$$ {#eq-state-def}
 
@@
    \mathbf{A} = \left( A_{ij} \right) = \big( P \left( \mathbf{X}[t+1] = \boldsymbol{\xi}_j \mid \mathbf{X}[t] = \boldsymbol{\xi}_i \right) \big)
-   $$
+   $$ {#eq-trans-matrix}
    where the value at row $i$ and column $j$ represents the probability to transition from state $\boldsymbol{\xi}_i$ to $\boldsymbol{\xi}_j$ during the time-step from $t$ to $t+1$.
@@
    \mathbf{B} = \left( B_{ij} \right) = \big( P \left( \mathbf{Z} = \boldsymbol{\zeta}_j \mid \mathbf{X}[t_D] = \boldsymbol{\xi}_i \right) \big)
-   $$
+   $$ {#eq-obs-matrix}
    where in row $i$ and at column $j$ we find the probability to *observe* a lymphatic involvement pattern $\mathbf{Z} = \boldsymbol{\zeta}_j$, given that the true (but hidden) state of involvement at the time of diagnosis $t_D$ is $\mathbf{X}[t_D] = \boldsymbol{\xi}_i$.
@@
 
+For later use, we define at this point the matrices $\boldsymbol{\Lambda}$:
+
+$$
+\boldsymbol{\Lambda} = P \left( \mathbf{X} \mid \mathbf{t} \right) = \begin{pmatrix}
+\boldsymbol{\pi}^\intercal \cdot \mathbf{A}^0 \\
+\boldsymbol{\pi}^\intercal \cdot \mathbf{A}^1 \\
+\vdots \\
+\boldsymbol{\pi}^\intercal \cdot \mathbf{A}^{t_\text{max}} \\
+\end{pmatrix}
+$$ {#eq-lambda-matrix}
+
+Where the $k$-th row in this matrix corresponds to the distribution over hidden states after $t=k-1$ time-steps.
+
 In this work, we use binomial distributions $\mathfrak{B} \left( t_D, p_{\text{T}x} \right)$ as time-priors which have one free parameter $p_{\text{T}x}$ for each group of patients we differentiate based on T-category.
@@
 
+Above, the quantity inside $\left[ \ldots \right]_i$ denotes the $i$-th component of the vector that is the result of the vector and matrix multiplications in the square brackets.
 Note that it is also possible to account for missing involvement information: If a diagnosis (like fine needle aspiration (FNA)) is only available for a subset of all LNLs, we can sum over all those possible complete observed states $\boldsymbol{\zeta}_j$ that match the provided diagnosis.
 
-The quantity $\ell$ in @eq-single-patient-llh depends on the spread parameters shown in @fig-small-graph via the transition matrix $\mathbf{A}$ and on the binomial parameters $p_{\text{T}x}$ via time-priors.
+The single-patient likelihood $\ell$ in @eq-single-patient-llh depends on the spread parameters shown in @fig-small-graph via the transition matrix $\mathbf{A}$ and on the binomial parameters $p_{\text{T}x}$ via time-priors.
 In this work, we will only differentiate between "early" (T1 & T2) and "advanced" (T3 & T4) T-categories.
@@
 $$
-\boldsymbol{\theta} = \left( \left\{ b_v \right\}, \left\{ t_{vr} \right\}, \left\{p_\text{T}x \right\} \right) \quad \text{with} \quad \genfrac{}{}{0pt}{2}{v\leq V}{r\in\operatorname{pa}(v)}
-$$
+\boldsymbol{\theta} = \left( \left\{ b_v \right\}, \left\{ t_{vr} \right\}, p_\text{early}, p_\text{adv.} \right) \quad \text{with} \quad \genfrac{}{}{0pt}{2}{v\leq V}{r\in\operatorname{pa}(v)}
+$$ {#eq-param-space}
 
@@
 \log \mathcal{L} \left( \mathcal{D} \mid \boldsymbol{\theta} \right) = \sum_{i=1}^N \log \ell_i
-$$
+$$ {#eq-log-likelihood}
+
+The methodology we use to infer the model's parameters is detailed in @sec-sampling.
 
@@
 
-<!-- TODO: Explain here what role our model plays in a Bayesian context. -->
+Our stated goal is to compute the risk for a patient's true nodal involvement state $\mathbf{X}$, *given* their individual diagnosis $d = \left( \boldsymbol{\zeta}_k, \text{T}x \right)$.
+Using Bayes' law, this is written as:
 
 $$
-P \left( \mathbf{X} \mid \mathbf{Z} \right) = \frac{P \left( \mathbf{Z} \mid \mathbf{X} \right) P \left( \mathbf{X} \right)}{P \left( \mathbf{Z} \right)}
+P \big( \mathbf{X} \mid \mathbf{Z}=\boldsymbol{\zeta}_k, \boldsymbol{\hat{\theta}}, \text{T}x \big) = \frac{P \left( \boldsymbol{\zeta}_k \mid \mathbf{X} \right) P \big( \mathbf{X} \mid \boldsymbol{\hat{\theta}}, \text{T}x \big)}{\sum_{i=0}^{2^V} P \left( \boldsymbol{\zeta}_k \mid \mathbf{X}=\boldsymbol{\xi}_i \right) P \big( \mathbf{X}=\boldsymbol{\xi}_i \mid \boldsymbol{\hat{\theta}}, \text{T}x \big)}
 $$ {#eq-uni-bayes-law}
 
-Let us for later use define the matrices $\boldsymbol{\Lambda}$:
+The term $P \left( \boldsymbol{\zeta}_k \mid \mathbf{X} \right)$ is defined solely by sensitivity and specificity of the diagnostic modality.
+Terms like this already appeared in the definition of the observation matrx in @eq-obs-matrix.
+The *prior* $P \big( \mathbf{X} \mid \boldsymbol{\hat{\theta}} \big)$ in the above equation is the crucial term that is supplied by a trained model and its parameters $\boldsymbol{\hat{\theta}}$.
+
+It is possible to compute this *posterior* probability of true involvement not only for one fully defined state $\mathbf{X}$, but also for e.g.
+individual LNLs: For example, the risk for involvement in level IV would be a marginalization over all states $\boldsymbol{\xi}_i$, where $\xi_{i4}=1$.
+Formally:
 
 $$
-\boldsymbol{\Lambda} = P \left( \mathbf{X} \mid \mathbf{t} \right) = \begin{pmatrix}
-\boldsymbol{\pi}^\intercal \cdot \mathbf{A}^0 \\
-\boldsymbol{\pi}^\intercal \cdot \mathbf{A}^1 \\
-\vdots \\
-\boldsymbol{\pi}^\intercal \cdot \mathbf{A}^{t_\text{max}} \\
-\end{pmatrix}
-$$ {#eq-lambda-matrix}
-
-The $n$-th row in this matrix corresponds to the distribution over hidden states after $t=n-1$ time-steps.
+P \big( \text{IV} \mid \mathbf{Z}=\boldsymbol{\zeta}_k, \boldsymbol{\hat{\theta}}, \text{T}x  \big) = \sum_{k \, : \, \xi_{k4}=1} P \big( \mathbf{X} = \boldsymbol{\xi}_k \mid \boldsymbol{\zeta}_k, \boldsymbol{\hat{\theta}}, \text{T}x  \big)
+$$ {#eq-marg-over-posterior}
 
@@
 P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \mid \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \right) = \frac{P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \mid \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right) P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right)}{P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \right)}
-$$
+$$ {#eq-bilateral-bayes}
+
+For the sake of brevity, we omit the dependency on the parameters and the T-category here.
 

### 2024-04-24 manuscript.qmd
run only midline model: one small, one full

@@
 
-midline_model = shared.get_model("midline", load_samples=True)
+simple_model = shared.get_model("simple", load_samples=True)
 ```
@@
 
+In @fig-model-midext-evo, we visualize how the prior distribution over diagnose times $P(t)$, the conditional probability of midline extension $P(\epsilon \mid t)$, and their joint $P(\epsilon, t)$ evolve over the course of a patient evolution.
+
+{{< embed model.qmd#fig-model-midext-evo >}}
+
 To get the joint probability over the ipsi- and contralateral hidden states, as well as the state of the tumor's midline extension $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)$, we simply add the above terms to the marginalization in @eq-bilateral-marginal:
@@
 
-In @fig-model-midext-evo, we visualize how the prior distribution over diagnose times $P(t)$, the conditional probability of midline extension $P(\epsilon \mid t)$, and their joint $P(\epsilon, t)$ evolve over the course of a patient evolution.
+In @fig-midline-state-dist we plot the state distribution for a full midline model as two separate heatmaps.
+To keep it readable, this example model only considers the LNLs II, III, and IV ipsi- and contralaterally.
 
-{{< embed model.qmd#fig-model-midext-evo >}}
+{{< embed model.qmd#fig-midline-state-dist >}}
 
@@
 
-## Prevalence Comparisons
+## Prevalence Predictions
 
-First, we look at the prevalence of contralateral involvement for the LNLs II, III, and IV.
-The three models' predictions are plotted against beta posteriors of the observed prevalence in the data in @fig-prevalences-overall, both for early and advanced T-category.
-
-The beta posteriors (drawn as lines in @fig-prevalences-overall) are a convenient way of indicating not only the percentage of patients that presented with any given involvement, T-category, and/or other characteristics, which sits at the maximum of the beta distribution.
-It also shows the size of the population via its variance: If we observe e.g.
-3 out of 10 patients to have contralateral LNL II involvement when the primary tumor is of advanced T-stage, then we cannot be particularly sure about that rate to be truly 30%.
-However, when we observe 300 out of 1000, we can be much more certain.
-A beta posterior intuitively reflects that in its width and is thus better suited to display such data than e.g.
-a vertical line only.
-
-The shaded histograms in @fig-prevalences-overall show the corresponding prevalence as computed by one of our models.
-They are formed by computing the prevalence for a number of parameter samples drawn during the MCMC learning process, as detailed in @sec-sampling.
-If both the location and width of the histograms matches those of the beta posteriors, then our model is both accurate and precise enough to describe the data.
-
-```{python}
-#| echo: false
-#| label: fig-prevalences-overall
-#| fig-cap: Prevalences of contralateral LNL involvement for different scenarios and the three considered models.
-The colored histograms show the computed distribution over the prevalence of involvment in the contralateral levels II (blue), III (orange), and IV (red).
-Each model's prediction is shown in a separate column, while the rows show these prevalences computed for early (top row) and advanced T-category (bottom row) respectively.
-The lines depict beta-posteriors over the observed prevalence of the corresponding scenario using a uniform prior.
-Its maximum coincides with the percentage of patients with the involvement of interest and its width indicates how many patients were considered in total.
-from itertools import cycle
-from lyscripts.scenario import Scenario
-from lyscripts.plot.utils import draw, BetaPosterior, Histogram, COLORS
-
-scenarios_config = utils.load_yaml_params("scenarios/overall.yaml")
-scenarios = Scenario.list_from_params(scenarios_config)
-
-nrows, ncols = 2, 3
-
-plt.rcParams.update(shared.get_fontsizes())
-plt.rcParams.update(shared.get_figsizes(
-  nrows=nrows,
-  ncols=ncols,
-  width=full,
-))
-
-fig, axes = plt.subplots(nrows=nrows, ncols=ncols, sharex="col")
-
-axes[0,0].set_ylabel("early")
-axes[-1,0].set_ylabel("advanced")
-
-for col, model in enumerate(["contra", "bilateral", "midline"]):
-  axes[0,col].set_title(model)
-  axes[-1,col].set_xlabel("prevalence [%]")
-  for row, t_stage in enumerate(["early", "late"]):
-    content = []
-    color_cycle = cycle(COLORS.values())
-    lnls = iter(["II", "III", "IV"])
-    for scenario in scenarios:
-      if not t_stage in scenario.t_stages:
-        continue
-
-      if model == "contra":
-        scenario.side = "contra"
-        scenario.is_uni = True
-      else:
-        scenario.side = "ipsi"
-        scenario.is_uni = False
-
-      color = next(color_cycle)
-      content += [
-        Histogram.from_hdf5(
-          filename=paths.get_filename(
-            model=model,
-            kind="prevalences_overall",
-          ),
-          dataname=scenario.md5_hash("prevalences"),
-          color=color,
-          label=f"contra LNL {next(lnls)}"
-        ),
-        BetaPosterior.from_hdf5(
-          filename=paths.get_filename(
-            model=model,
-            kind="prevalences_overall",
-          ),
-          dataname=scenario.md5_hash("prevalences"),
-          color=color,
-        ),
-      ]
-    draw(
-      axes=axes[row,col],
-      contents=content,
-      percent_lims=(2., 2.),
-      xlims=(0., 27.)
-    )
-    axes[row,col].set_yticks([])
-
-axes[0,0].legend(labelspacing=0.2)
-axes[-1,-1].legend(labelspacing=0.2)
-```
-
-At first, from @fig-prevalences-overall it appears as if the simpler, naive, unilateral-only model describes the prevalence of contralateral involvement better than the bilateral model and even the more complex model that tracks midline extension of the primary tumor.
-
-- [ ] make the data in the paragraph below dynamic
-
-This is mainly due to the early/advanced T-category split of the ipsi- and contralateral side: All models use two different distributions for early and advanced T-category to marginalize over the diagnose times (as introduced in the recap of @sec-unilateral).
-But only the two bilateral models need to balance these distributions to describe the ipsi- and contralateral involvement simultaneaously.
-As can be seen in the second and third column of @fig-prevalences-overall, the prevalence of e.g.
-contralateral LNL II involvement in the data (blue lines) increases from 8% (39 out of 473) to 24% (88 out of 360).
-Which is more than the model predicts (blue histograms).
-Ipsilaterally, however, the involvement of LNL II increases from 71% (343 out of 473) to only 76% (276 out of 360).
-This much weaker increase is well capured by the models.
-
-This issue is quickly resolved, though, as we consider other risk factors that correlate with T-category: Both the ipsilateral involvement and the tumor's extension over the mid-sagittal line influence -- the latter probably even causally -- the extent of contralateral involvement, as shown in @tbl-data-strat.
-Both these risk factors cannot be considered by the contralateral-only model.
-We should thus expect this naive model to perform worse, as soon as we investigate ipsilateral involvement and midline extension.
-
-```{python}
-#| echo: false
-#| label: fig-prevalences-ipsi
-#| fig-cap: Contralateral prevalences when taking ipsilateral invovlement into account.
-scenarios_config = utils.load_yaml_params("scenarios/with_ipsi.yaml")
-scenarios = Scenario.list_from_params(scenarios_config)
-
-nrows, ncols = 2, 3
-
-plt.rcParams.update(shared.get_fontsizes())
-plt.rcParams.update(shared.get_figsizes(
-  nrows=nrows,
-  ncols=ncols,
-  width=full,
-))
-
-fig, axes = plt.subplots(nrows=nrows, ncols=ncols, sharex="col")
-
-axes[0,0].set_ylabel("early")
-axes[-1,0].set_ylabel("advanced")
-
-for col, model in enumerate(["contra", "bilateral", "midline"]):
-  axes[0,col].set_title(model)
-  axes[-1,col].set_xlabel("prevalence [%]")
-  for row, t_stage in enumerate(["early", "late"]):
-    content = []
-    color_cycle = cycle(COLORS.values())
-    lnls = iter(["II", "III", "IV"])
-    for scenario in scenarios:
-      if not t_stage in scenario.t_stages:
-        continue
-
-      if model == "contra":
-        scenario = scenario.from_dict(
-          scenario.as_dict("prevalences"),
-          is_uni=True,
-          side="contra",
-        )
-
-      color = next(color_cycle)
-      content += [
-        Histogram.from_hdf5(
-          filename=paths.get_filename(
-            model=model,
-            kind="prevalences_with_ipsi",
-          ),
-          dataname=scenario.md5_hash("prevalences"),
-          color=color,
-          label=f"contra LNL {next(lnls)}"
-        ),
-        BetaPosterior.from_hdf5(
-          filename=paths.get_filename(
-            model=model,
-            kind="prevalences_with_ipsi",
-          ),
-          dataname=scenario.md5_hash("prevalences"),
-          color=color,
-        ),
-      ]
-    draw(
-      axes=axes[row,col],
-      contents=content,
-      percent_lims=(2., 2.),
-      xlims=(0., 27.)
-    )
-    axes[row,col].set_yticks([])
-
-axes[0,0].legend(labelspacing=0.2)
-axes[-1,-1].legend(labelspacing=0.2)
-```
+Here we plot how well the model fits the data...
 
@@
 
-model = shared.get_model(which="midline", load_samples=True)
-samples = shared.get_samples(which="midline")
+model = shared.get_model(which="simple", load_samples=True)
+samples = shared.get_samples(which="simple")
 
@@
 mi = pd.MultiIndex.from_product(
-  [[False, True]] * len(midline_model.ext.ipsi.graph.lnls),
-  names=midline_model.ext.ipsi.graph.lnls.keys(),
+  [[False, True]] * len(simple_model.ext.ipsi.graph.lnls),
+  names=simple_model.ext.ipsi.graph.lnls.keys(),
 )
-series = pd.Series(midline_model.ext.ipsi.state_dist(), index=mi)
+series = pd.Series(simple_model.ext.ipsi.state_dist(), index=mi)
 upsetplot.plot(

### 2024-04-24 model.qmd
run only midline model: one small, one full

@@
 import matplotlib.pyplot as plt
+import matplotlib.cm
 from matplotlib import ticker
@@
 
-midline_model = shared.get_model("midline", load_samples=True)
+model = shared.get_model("simple", load_samples=True)
 
@@
 p_midline = {
-  "lateralized": (1 - midline_model.midext_prob)**t,
-  "extension": 1 - (1 - midline_model.midext_prob)**t,
+  "lateralized": (1 - model.midext_prob)**t,
+  "extension": 1 - (1 - model.midext_prob)**t,
 }
 dist = {
-  "early": midline_model.get_distribution("early").pmf,
-  "late": midline_model.get_distribution("late").pmf,
+  "early": model.get_distribution("early").pmf,
+  "late": model.get_distribution("late").pmf,
 }
@@
 
-for model in ["contra", "bilateral", "midline"]:
+for model in ["simple"]:
   history = pd.read_csv(f"models/{model}/history.csv").set_index("steps")
@@
 ```
+
+# State Distribution of Midline Model
+
+It may be illustrative to look at the 3D state distribution of the midline model (at least if we only consider 3 LNLs).
+
+```{python}
+#| label: fig-midline-state-dist
+#| fig-cap: 3D state distribution of a midline model with 3 LNLs (II, III, and IV) in both sides of the neck.
+from matplotlib.colors import LinearSegmentedColormap
+from mpl_toolkits.axes_grid1 import AxesGrid
+
+model = shared.get_model(which="simple", load_samples=True)
+state_dist = 100 * model.state_dist(t_stage="late")
+vmin, vmax = np.min(state_dist), np.max(state_dist)
+
+nrows, ncols = 1, 2
+cmap = LinearSegmentedColormap.from_list(
+  name="custom",
+  colors=[COLORS["green"], COLORS["orange"], COLORS["red"]],
+)
+cmap.set_under(COLORS["gray"])
+
+plt.rcParams.update(shared.get_fontsizes())
+plt.rcParams.update(shared.get_figsizes(
+  nrows=nrows,
+  ncols=ncols,
+  aspect_ratio=1.,
+  width=17,
+))
+
+fig = plt.figure()
+grid = AxesGrid(
+  fig, 111,
+  nrows_ncols=(nrows, ncols),
+  axes_pad=0.2,
+  share_all=True,
+  cbar_location="right",
+  cbar_mode="single",
+  cbar_size="8%"
+)
+noext, midext = grid
+
+im = noext.imshow(state_dist[0], cmap=cmap, vmin=100 * np.exp(-5), vmax=vmax)
+im = midext.imshow(state_dist[1], cmap=cmap, vmin=100 * np.exp(-5), vmax=vmax)
+grid.cbar_axes[0].colorbar(im)
+grid.cbar_axes[0].set_ylabel("Probability [%]")
+
+noext.set_title(r"tumor lateralized $\epsilon=\text{False}$")
+midext.set_title(r"midline extension $\epsilon=\text{True}$")
+
+noext.set_ylabel(r"ipsi state $\boldsymbol{\xi}^\text{i}_\ell$")
+noext.set_xlabel(r"contra state $\boldsymbol{\xi}^\text{c}_k$")
+midext.set_xlabel(r"contra state $\boldsymbol{\xi}^\text{c}_k$")
+
+plt.show()
+```

### 2024-04-24 model.qmd
start with prevalence plots

@@
 
-# State Distribution of Midline Model
+# Prior State Distribution of Midline Model
 
@@
 ```
+
+# Observed vs Predicted Prevalence
+
+To verify that the model does what it is supposed to, we compare how often some selected diagnoses appear in the data and how often the model predicts these diagnoses to be.
+
+```{python}
+#| label: fig-model-prevalences
+#| fig-cap: Comparison of predicted vs observed prevalences.
+from lyscripts.scenario import Scenario
+from lyscripts.plot.utils import Histogram, BetaPosterior, draw
+from scripts import paths
+
+nrows, ncols = 2, 2
+
+plt.rcParams.update(shared.get_fontsizes())
+plt.rcParams.update(shared.get_figsizes(
+  nrows=nrows,
+  ncols=ncols,
+  aspect_ratio=2.,
+  width=17,
+))
+
+fig, axes = plt.subplots(
+  nrows=nrows,
+  ncols=ncols,
+  sharex="col",
+  sharey="row",
+)
+
+for i, t_stage in enumerate(["early", "late"]):
+  axes[i,0].set_yticks([])
+  axes[i,0].set_ylabel(t_stage)
+  for j, midext in enumerate([False, True]):
+    content = []
+    for lnl, color in zip(
+      ["II"],
+      [COLORS["blue"]],
+    ):
+      scenario = Scenario(
+        t_stages=[t_stage],
+        midext=midext,
+        diagnosis={"contra": {"max_llh": {lnl: True}}}
+      )
+      content.append(
+        Histogram.from_hdf5(
+          filename=paths.model_dir / "full" / "prevalences_overall.hdf5",
+          dataname=scenario.md5_hash("prevalences"),
+          color=color,
+          label=f"{t_stage = }, {midext = }"
+        )
+      )
+      content.append(
+        BetaPosterior.from_hdf5(
+          filename=paths.model_dir / "full" / "prevalences_overall.hdf5",
+          dataname=scenario.md5_hash("prevalences"),
+          color=color,
+        )
+      )
+    draw(
+      axes=axes[i,j],
+      contents=content,
+      xlims=(0., 20.)
+    )
+    axes[i,j].legend()
+
+plt.show()
+```

### 2024-04-25 manuscript.qmd
add prevalences plot and description

@@
 $$
-\ell = \sum_{t=0}^{t_\text{max}} \left[ \boldsymbol{\xi}_0 \cdot \mathbf{A} \cdot \mathbf{B} \right]_i \cdot P \left( t \mid \text{T}x \right)
+\ell = P \left( \mathbf{Z} = \boldsymbol{\zeta}_i \mid \text{T}x \right) = \sum_{t=0}^{t_\text{max}} \left[ \boldsymbol{\xi}_0 \cdot \mathbf{A} \cdot \mathbf{B} \right]_i \cdot P \left( t \mid \text{T}x \right)
 $$ {#eq-single-patient-llh}
@@
 
-In @fig-midline-state-dist we plot the state distribution for a full midline model as two separate heatmaps.
+In @fig-model-state-dist we plot the state distribution for a full midline model as two separate heatmaps.
 To keep it readable, this example model only considers the LNLs II, III, and IV ipsi- and contralaterally.
 
-{{< embed model.qmd#fig-midline-state-dist >}}
+{{< embed model.qmd#fig-model-state-dist >}}
 
@@
 
+## Computing the Observed and Predicted Prevalence of Involvement Patterns
+
+We want to assess the model's capability to approximate the distribution of lymphatic involvement patterns seen in the data.
+To that end, we compare the prevalence of some invovlement patterns under selected scenarios with the model's prediction for how often these involvements it expects to see, given these scenarios.
+
+In this context, a "scenario" includes the patient's T-category $\text{T}x$ and whether the patient's tumor extended over the mid-sagittal line, i.e.
+$\epsilon=\texttt{True}$ or $\epsilon=\texttt{False}$.
+
+An involvement pattern specifies for each ipsi- and contralateral LNL whether it is "healthy", "involved", or "masked".
+If it is "masked", we essentially state that we are not interested in the involvement of that LNL and the prevalence will be marginalized over this LNL's involvement.
+
+For example, we may be interested in the prevalence of contralateral LNL II involvement (i.e., contra LNL II "involved" and all other LNLs "masked") under the scenario of early T-category (T0-T2) and no midline extension ($\epsilon=\texttt{False}$).
+To compute this prevalence in the data, we select all patients of this scenario (in our data, this amounts to <!-- TODO: dyamically compute this --> patients).
+Of those, <!-- TODO: dynamically compute this --> were found to harbor metastases in their contralateral LNL II.
+Therefore, the prevalence is <!-- TODO: dynamically compute this -->.
+
+When displaying this data prevalence, we often choose to draw a *Beta posterior* over the "true" prevalence, hinting at the fact that our data merely represents a limited sample.
+The Beta posterior follows from a uniform Beta distribution as prior and a Bernoulli likelihood for the number of patients with the involvement of interest, given the parameer for the "true" prevalence.
+The resulting distribution has its maximum at the observed prevalence, but in addition gives a visual intuition for the variance of of the observed quantity.
+I.e., when we observe 3 out of 10 events, the Beta posterior is much wider than if we observe 300 out of 1000 for he same prevalence.
+It also allows us to check not inly if the model is accurate, but also whether it reflects the uncertainty contained in the data.
+
+Predicting the prevalence using our model amounts to computing the following probability:
+
+$$
+P \left( \text{II}^\text{c} \mid \epsilon=\texttt{False}, \text{T}x=\text{early} \right) = \frac{P \left( \text{II}^\text{c}, \epsilon=\texttt{False} \mid \text{T}x=\text{early} \right)}{P \left( \epsilon=\texttt{False} \mid \text{T}x=\text{early} \right)}
+$$
+
+In the enumator, we marginalize over all ipsi- and contralateral LNLs' involvements, except for LNL II contralaterally.
+This is similar to the marginalization in @eq-marg-over-posterior, although we are summing over different quantities.
+In the denominator, we can simply insert the joint distribution over midline extension and diagnose time $P \left( \epsilon, t \right)$ marginalized over $t$ using the early T-category's time-prior.
+
+Since we compare it to the data, which does not report true but only observed involvement -- although pathologically investigated LNLs may be as close as possible to the ground truth -- we do not consider posteriors of the form $P \left( \mathbf{X} \mid \mathbf{Z} \right)$ here.
+Instead, we compute probabilities of observed involvement $P \left( \mathbf{Z} \right)$, as in the likelihood @eq-single-patient-llh.
+
+When plotted, we usually display histograms over the model's predictions.
+Each of their values was computed from a different parameter set drawn during MCMC sampling, effectively giving us a distribution over the prevalences.
+Ideally, the histograms approximate the location and width of the Beta posteriors when attempting to describe the data they were trained on.
+
 # Results
@@
 
-Here we plot how well the model fits the data...
+<!-- NOTE: This should probably belong to the Methods/Formalism section(s) -->
+
+In @fig-model-prevalences, we plot the prevalence of contralateral involvement of the LNLs II, III, and IV for the four scenarios made up of the possible combinations of early and late T-category, as well as lateralized and midline extending tumors.
+
+{{< embed model.qmd#fig-model-prevalences >}}
+
+@Fig-model-prevalences shows nicely how the model is capable of accurately taking the most important risk factors, i.e.
+T-category and midline extension, into account.
+As observed in the data, the model predicts that the prevalence of contralateral LNL II involvement jumps from below 8% for early T-category lateralized tumors to almost 40% when the tumor is of advanced T-category and crosses the mid-sagittal line.
+
+However, for early T-category scenarios with midline extension, the model does seem to overestimate contralateral LNL II invovlement and underestimate the prevalence for metastases in LNL III.
+This likely stems from the the small sample size of this rare scenario as hinted at by the wide Beta posteriors.
 

### 2024-04-25 model.qmd
add prevalences plot and description

--- 2024-04-24 model.qmd
@@
 ```{python}
-#| label: fig-midline-state-dist
+#| label: fig-model-state-dist
 #| fig-cap: 3D state distribution of a midline model with 3 LNLs (II, III, and IV) in both sides of the neck.
@@
 cmap = LinearSegmentedColormap.from_list(
-  name="custom",
+  name="usz",
   colors=[COLORS["green"], COLORS["orange"], COLORS["red"]],
@@
 ```
+
+# Observed vs Predicted Prevalence
+
+To verify that the model does what it is supposed to, we compare how often some selected diagnoses appear in the data and how often the model predicts these diagnoses to be.
+
+```{python}
+#| label: fig-model-prevalences
+#| fig-cap: Comparison of predicted (histograms) vs observed (beta posteriors) prevalences.
+Shown for the contralateral LNLs II (blue), III (orange), and IV (green).
+The top row shows scenarios with early T-category tumors, the bottom row for late T-category ones.
+The left column depicts scenarios where the primary tumor is clearly lateralized, the right column scenarios of tumors extending over the mid-sagittal line.
+from matplotlib import gridspec
+from lyscripts.scenario import Scenario
+from lyscripts.plot.utils import Histogram, BetaPosterior, draw
+from scripts import paths
+
+def turn_axis_off(axes):
+  axes.set_xticks([])
+  axes.set_yticks([])
+
+nrows, ncols = 2, 2
+
+plt.rcParams.update(shared.get_fontsizes())
+plt.rcParams.update(shared.get_figsizes(
+  nrows=nrows,
+  ncols=ncols,
+  aspect_ratio=2.,
+  width=17,
+))
+
+fig = plt.figure()
+gs = gridspec.GridSpec(nrows=nrows, ncols=ncols, figure=fig)
+axes = np.array([[fig.add_subplot(gs[i,j]) for j in range(ncols)] for i in range(nrows)])
+full = fig.add_subplot(gs[:,:], frame_on=False)
+turn_axis_off(full)
+
+full.set_title(
+  "Observed vs Predicted Prevalence of Contralateral Involvement",
+  pad=20.,
+  fontsize="large",
+)
+
+for i, t_stage in enumerate(["early", "late"]):
+  axes[i,0].set_ylabel(f"{t_stage} T-cat.", fontweight="bold")
+  for j, midext in enumerate([False, True]):
+    axes[0,j].set_title("Mid. ext." if midext else "Lateralized", fontweight="bold")
+    axes[0,j].set_xticks([])
+    axes[-1,j].set_xlabel("prevalence [%]")
+    axes[i,j].set_yticks([])
+    content = []
+    for lnl, color in zip(
+      ["II", "III", "IV"],
+      [COLORS["blue"], COLORS["orange"], COLORS["green"]],
+    ):
+      scenario = Scenario(
+        t_stages=[t_stage],
+        midext=midext,
+        diagnosis={"contra": {"max_llh": {lnl: True}}}
+      )
+      content.append(
+        Histogram.from_hdf5(
+          filename=paths.model_dir / "full" / "prevalences_overall.hdf5",
+          dataname=scenario.md5_hash("prevalences"),
+          color=color,
+          label=f"contra LNL {lnl}"
+        )
+      )
+      content.append(
+        BetaPosterior.from_hdf5(
+          filename=paths.model_dir / "full" / "prevalences_overall.hdf5",
+          dataname=scenario.md5_hash("prevalences"),
+          color=color,
+        )
+      )
+    draw(
+      axes=axes[i,j],
+      contents=content,
+      xlims=(0., 15. if not midext else 45.),
+    )
+    axes[i,j].legend(labelspacing=0.1)
+
+plt.show()
+```

### 2024-05-17 manuscript.qmd
add figure on ipsi & contra prevalence correlation

@@
 
+Note that we decided to omit the y-axis ticks and labels in these figures over prevalences and risks.
+The y-axis in these plots measures the probability density and its numerical values are not intuitively interpretable.
+Instead, we occasionally use the freed space to label e.g.
+rows of subplots.
+
 # Results
@@
 
-In @fig-model-prevalences, we plot the prevalence of contralateral involvement of the LNLs II, III, and IV for the four scenarios made up of the possible combinations of early and late T-category, as well as lateralized and midline extending tumors.
+In @fig-model-prevalences-overall, we plot the prevalence of contralateral involvement of the LNLs II, III, and IV for the four scenarios made up of the possible combinations of early and late T-category, as well as lateralized and midline extending tumors.
 
-{{< embed model.qmd#fig-model-prevalences >}}
+{{< embed model.qmd#fig-model-prevalences-overall >}}
 
-@Fig-model-prevalences shows nicely how the model is capable of accurately taking the most important risk factors, i.e.
+@Fig-model-prevalences-overall shows nicely how the model is capable of accurately taking the most important risk factors, i.e.
 T-category and midline extension, into account.
@@
 This likely stems from the the small sample size of this rare scenario as hinted at by the wide Beta posteriors.
+
+{{< embed model.qmd#fig-model-prevalences-with-ipsi >}}
+
+In @fig-model-prevalences-with-ipsi we display the model's ability to capture the correlation between ipsi- and contralateral involvement.
+It shows that the prevalence of metastases in the two sides of the neck is correlated via the time of diagnosis, despite the model not having any direct connections between the two side.
+However, there are some small discrepancies in the model's prediction: When considering the scenario of LNL II involvement in the ipsi- *and* contralateral side, it cannot quite capture the split between early and late T-category.
+The model slightly overestimates the prevalence of this scenario for early T-category patients and underestimates it for advanced T-category.
+
+## Prediction of Risk for Occult Disease
+
+- compute risks for occult disease, given diagnosis
 
@@
 
+# Discussion
+
+- conceptually extended model to cover both sides of neck
+- explicitly models midline extension (also allows for easy marginalization)
+- midline extension intuitively causes contralateral spread to become more similar to ipsilateral spread
+
 # Acknowledgement {.appendix}

### 2024-05-17 model.qmd
add figure on ipsi & contra prevalence correlation

@@
 axes[0].set_ylim(0., 1.)
+axes[0].tick_params(bottom=False, labelbottom=False)
 axes[0].yaxis.set_major_formatter(ticker.StrMethodFormatter("{x:.0%}"))
@@
 #| label: fig-model-burnin-history 
-#| fig-cap: Monitoring quantities during the burn-in phase of the parameter sampling.
+#| fig-cap: "Monitoring quantities during the burn-in phase of the parameter sampling."
 from collections import namedtuple
@@
 #| label: fig-model-state-dist
-#| fig-cap: 3D state distribution of a midline model with 3 LNLs (II, III, and IV) in both sides of the neck.
+#| fig-cap: "3D state distribution of a midline model with 3 LNLs (II, III, and IV) in both sides of the neck."
 from matplotlib.colors import LinearSegmentedColormap
@@
 
-To verify that the model does what it is supposed to, we compare how often some selected diagnoses appear in the data and how often the model predicts these diagnoses to be.
+To verify that the model does what it is supposed to, we compare how often some selected diagnoses appear in the data and how often the model predicts these diagnoses to be observed.
+
+## Contralteral LNLs II, III, and IV Dependent on T-Cat. and Mid. Ext.
+
+We start witht the most commonly involved LNLs in the contralateral neck: II, III, and IV.
+We investigate their prevalence in the four scenarios early T-cat.
+withour mid.
+ext, late T-cat.
+without mid.
+ext., early T-cat.
+with mid.
+ext., and late T-cat.
+with mid.
+ext.
 
 ```{python}
-#| label: fig-model-prevalences
-#| fig-cap: Comparison of predicted (histograms) vs observed (beta posteriors) prevalences.
+#| label: fig-model-prevalences-overall
+#| fig-cap: "Comparison of predicted (histograms) vs observed (beta posteriors) prevalences.
 Shown for the contralateral LNLs II (blue), III (orange), and IV (green).
@@
 The left column depicts scenarios where the primary tumor is clearly lateralized, the right column scenarios of tumors extending over the mid-sagittal line.
+This figure illustrates the model's ability to describe the prevalence of different combinations of scenarios involving the risk factors T-category and midline extension."
 from matplotlib import gridspec
@@
 from lyscripts.plot.utils import Histogram, BetaPosterior, draw
-from scripts import paths
+from scripts import paths, shared
 
@@
 full.set_title(
-  "Observed vs Predicted Prevalence of Contralateral Involvement",
+  "Observed vs Predicted Prevalence of Contralateral Involvement\nin the LNLs II, III, and IV",
   pad=20.,
@@
           color=color,
-          label=f"contra LNL {lnl}"
+          label="prediction",
         )
       )
-      content.append(
-        BetaPosterior.from_hdf5(
+      beta_post = BetaPosterior.from_hdf5(
         filename=paths.model_dir / "full" / "prevalences_overall.hdf5",
@@
       )
-      )
+      beta_post.kwargs["label"] = f"data: {beta_post.num_success} of {beta_post.num_total}"
+      content.append(beta_post)
     draw(
@@
     )
-    axes[i,j].legend(labelspacing=0.1)
+
+# manually place legends
+shared.split_legends(
+  ax=axes[0,0],
+  titles=["LNL II", "LNL III", "LNL IV"],
+  locs=[(0.57, 0.55), (0.48, 0.99), (0.15, 0.99)],
+)
+shared.split_legends(
+  ax=axes[0,1],
+  titles=["LNL II", "LNL III", "LNL IV"],
+  locs=[(0.67, 0.99), (0.27, 0.6), (0.1, 0.99)],
+)
+shared.split_legends(
+  ax=axes[1,0],
+  titles=["LNL II", "LNL III", "LNL IV"],
+  locs=[(0.64, 0.99), (0.27, 0.6), (0.24, 0.99)],
+)
+shared.split_legends(
+  ax=axes[1,1],
+  titles=["LNL II", "LNL III", "LNL IV"],
+  locs=[(0.64, 0.99), (0.4, 0.6), (0.16, 0.99)],
+)
 
@@
 ```
+
+## Correlation between Ipsi- and Contralateral Involvement
+
+```{python}
+#| label: fig-model-prevalences-with-ipsi
+#| fig-cap: "Comparison of predicted (histograms) vs observed (beta posteriors) prevalences.
+Shown are four scenarios, all including contralateral LNL II involvement: Early T-category and an ipsilateral N0 neck (green), early T-category and ipsilateral LNL II involvement (blue), as well as the same two scenarios but for advanced T-category (orange and red).
+The figure shows that the model is capable of describing the correlation between ipsi- and contralateral involvement.
+Although for the scenario of LNL II involvement in both sides, the prediction's split between early and advanced T-category is not large enough."
+from lyscripts.scenario import Scenario
+from lyscripts.plot.utils import Histogram, BetaPosterior, draw
+from scripts import paths, shared
+
+plt.rcParams.update(shared.get_fontsizes())
+plt.rcParams.update(shared.get_figsizes(width=17./2.))
+
+fig, ax = plt.subplots()
+content = []
+colors = iter([COLORS[c] for c in ["orange", "green", "red", "blue"]])
+
+for t_stage in ["early", "late"]:
+  for diagnosis in [{"II": True}, {lnl: False for lnl in ["I", "II", "III", "IV", "V"]}]:
+    color = next(colors)
+    scenario = Scenario(
+      t_stages=[t_stage],
+      diagnosis={
+        "ipsi": {"max_llh": diagnosis},
+        "contra": {"max_llh": {"II": True}},
+      }
+    )
+    content.append(
+      Histogram.from_hdf5(
+        filename=paths.model_dir / "full" / "prevalences_with_ipsi.hdf5",
+        dataname=scenario.md5_hash("prevalences"),
+        color=color,
+        label="prediction"
+      )
+    )
+    beta_post = BetaPosterior.from_hdf5(
+      filename=paths.model_dir / "full" / "prevalences_with_ipsi.hdf5",
+      dataname=scenario.md5_hash("prevalences"),
+      color=color,
+    )
+    beta_post.kwargs["label"] = f"data: {beta_post.num_success} of {beta_post.num_total}"
+    content.append(beta_post)
+
+draw(ax, content, xlims=(0., 25.))
+ax.set_title("Observed vs Predicted Prevalence\nof Contralateral LNL II Involvement")
+ax.set_yticks([])
+ax.set_xlabel("prevalence [%]")
+
+shared.split_legends(
+  ax=ax,
+  titles=["early with ipsi II", "early ipsi N0", "late with ipsi II", "late with ipsi N0"],
+  locs=[(0.23, 0.65), (0.1, 0.99), (0.6, 0.65), (0.45, 0.99)],
+)
+
+plt.show()
+```

### 2024-05-17 scripts/shared.py

@@ -99,3 +99,29 @@ def get_figsizes(
         "savefig.bbox": "tight",
         "savefig.pad_inches": pad,
     }
+
+
+def split_legends(
+    ax,
+    titles: list[str],
+    locs: list[tuple[float, float]],
+    **kwargs,
+) -> None:
+    """Separate the legends of `ax` into separate ones with `titles` and at `locs`."""
+    legend_kwargs = {
+        "title_fontsize": "small",
+        "labelspacing": 0.1,
+    }
+    legend_kwargs.update(kwargs)
+
+    handles, labels = ax.get_legend_handles_labels()
+    for i, (title, loc) in enumerate(zip(titles, locs)):
+        legend = ax.legend(
+            handles[i * 2 : (i+1) * 2],
+            labels[i * 2 : (i+1) * 2],
+            loc="upper left",
+            bbox_to_anchor=loc,
+            title=title,
+            **legend_kwargs,
+        )
+        ax.add_artist(legend)

### 2024-05-17 model.qmd
start with midext prevalence plot

@@
 ```
+
+## Portion of Midline Extension
+
+Since we model the tumor's midline extension as a random variable we can predict its portion among early or late T-category patients and compare it with the observed prevalence in the data.
+
+```{python}
+#| label: fig-model-prevalence-midext
+#| fig-cap: "Comparing the predicted (histograms) and observed (lines depicting beta posteriors) prevalence of midline extension for early (blue) and late (orange) T-category."
+pass
+```

### 2024-05-17 scripts/compute_midext_prevalence.py

@@ -1,0 +1,45 @@
+"""Compute the prevalence of midline extension (predicted and observed)."""
+import h5py
+
+import paths
+import shared
+
+
+def main():
+    model = shared.get_model("full")
+    samples = shared.get_samples("full")
+    predicted = {"early": [], "late": []}
+
+    data = shared.get_data()
+    data = data[~data.ly.midext.isna()]
+    observed = {
+        "early": (
+            data[data.ly.t_stage == "early"].ly.is_midext(True).sum(),
+            (data.ly.t_stage == "early").sum(),
+        ),
+        "late": (
+            data[data.ly.t_stage == "late"].ly.is_midext(True).sum(),
+            (data.ly.t_stage == "late").sum(),
+        ),
+    }
+
+    for sample in samples[::10]:
+        model.set_params(*sample)
+        predicted["early"].append(model.marginalize(t_stage="early", midext=True))
+        predicted["late"].append(model.marginalize(t_stage="late", midext=True))
+
+    with h5py.File(
+        paths.model_dir / "full" / "prevalence_midext.hdf5",
+        mode="w",
+    ) as h5_file:
+        h5_file["early"] = predicted["early"]
+        h5_file["early"].attrs["num_match"] = observed["early"][0]
+        h5_file["early"].attrs["num_total"] = observed["early"][1]
+
+        h5_file["late"] = predicted["late"]
+        h5_file["late"].attrs["num_match"] = observed["late"][0]
+        h5_file["late"].attrs["num_total"] = observed["late"][1]
+
+
+if __name__ == "__main__":
+    main()

### 2024-05-17 scripts/shared.py

@@ -8,7 +8,10 @@ from lyscripts import utils
 import numpy as np
 import pandas as pd
 
-from . import paths
+try:
+    from . import paths
+except ImportError:
+    import paths
 
 
 pd.options.mode.copy_on_write = False
@@ -17,7 +20,7 @@ GOLDEN_RATIO = 1.61803398875
 
 
 def get_model(
-    which: Literal["ipsi", "contra", "bilateral", "midline"],
+    which: str,
     load_samples: bool = True,
 ) -> types.Model:
     """Get one of the four trained models."""
@@ -35,7 +38,7 @@ def get_model(
 
 
 def get_samples(
-    which: Literal["ipsi", "contra", "bilateral", "midline"]
+    which: str
 ) -> np.ndarray:
     """Get the samples of one of the four trained models."""
     samples_path = paths.model_dir / which / "samples.hdf5"

### 2024-05-21 manuscript.qmd
add section on midext prevalence

@@
 
+{{< embed model.qmd#fig-model-prevalences-midext >}}
+
+Lastly, in @fig-model-prevalences-midext, we plot the prevalence of midline extension in the data versus our model's prediction.
+It is obvious the model cannot match the large spread between early and advanced T-category seen in the data.
+This is because to achieve that, it would need to increase the advanced T-category patient's prior distribution over diagnosis times and at the same time reduce the probability of the tumor to cross the midline during a time-step.
+But since the time-priors parameter is also coupled with the spread probabilities among the LNLs, the model does not have that freedom.
+
+<!-- Should the paragraphs below be part of the discussion? -->
+
+However, we do not consider this discrepancy a major limitation of the model: We will not realistically be interested in the probability of midline extension, as it is always possible to assess it with high certainty.
+That is also the reason why we initially modelled the midline extension *not* as a random variable, but as a global risk factor that would have been turned on or off from the onset of a patient's disease evolution.
+This, however, lead to overly high risks for contralateral involvement in advanced T-category patients with midline extension, because then the model assumes an increased spread to the contralateral side from the onset of the disease.
+Which is probably not true in a majority of those cases.
+Thus, treating it as a random variable that only becomes true during a patient's disease evolution resulted in a better description of the data.
+
+Formally, the wrong prediction of midline extension prevalence makes little difference, since it is always given: Instead of $P\left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \mid \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \right)$, we typically compute $P\left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \mid \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c}, \epsilon \right)$, which does not suffer from the wrong probability of midline extension, as the distribution over hidden states is renormalized:
+
+$$
+P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \mid \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c}, \epsilon \right) = \frac{P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \mid \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right) P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)}{P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c}, \epsilon \right)}
+$$
+
+Note that a distribution over $\epsilon$ appears both in the enumerator and the denominator, which largely cancel each other, leaving only the midline extension's effect on the distribution over hidden states in the prediction.
+
 ## Prediction of Risk for Occult Disease

### 2024-05-21 model.qmd
add section on midext prevalence

--- 2024-05-17 model.qmd
@@
 from mpl_toolkits.axes_grid1 import AxesGrid
+from lyscripts.plot.utils import COLORS
 
@@
 from lyscripts.scenario import Scenario
-from lyscripts.plot.utils import Histogram, BetaPosterior, draw
+from lyscripts.plot.utils import Histogram, BetaPosterior, draw, COLORS
 from scripts import paths, shared
@@
 Although for the scenario of LNL II involvement in both sides, the prediction's split between early and advanced T-category is not large enough."
+import matplotlib.pyplot as plt
 from lyscripts.scenario import Scenario
-from lyscripts.plot.utils import Histogram, BetaPosterior, draw
+from lyscripts.plot.utils import Histogram, BetaPosterior, draw, COLORS
 from scripts import paths, shared
@@
 ```
+
+## Portion of Midline Extension
+
+Since we model the tumor's midline extension as a random variable we can predict its portion among early or late T-category patients and compare it with the observed prevalence in the data.
+
+```{python}
+#| label: fig-model-prevalences-midext
+#| fig-cap: "Comparing the predicted (histograms) and observed (lines depicting beta posteriors) prevalence of midline extension for early (blue) and late (orange) T-category.
+While the prevalence is predicted correctly when marginalizing over T-category, the model cannot capture the degree of separation observed in the data.
+Since the tumor's midline extension is virtually always part of the diagnosis and hence *given* when predicting a patient's risk, we do not consider this discrepancy a major issue."
+import matplotlib.pyplot as plt
+from lyscripts.scenario import Scenario
+from lyscripts.plot.utils import Histogram, BetaPosterior, draw, COLORS
+from scripts import paths, shared
+
+plt.rcParams.update(shared.get_fontsizes())
+plt.rcParams.update(shared.get_figsizes(width=17./2.))
+
+fig, ax = plt.subplots()
+content = []
+colors = iter([COLORS[c] for c in ["blue", "orange"]])
+
+for t_stage in ["early", "late"]:
+  color = next(colors)
+  content.append(
+    Histogram.from_hdf5(
+      filename=paths.model_dir / "full" / "prevalences_midext.hdf5",
+      dataname=t_stage,
+      color=color,
+      label="prediction"
+    )
+  )
+  beta_post = BetaPosterior.from_hdf5(
+    filename=paths.model_dir / "full" / "prevalences_midext.hdf5",
+    dataname=t_stage,
+    color=color,
+  )
+  beta_post.kwargs["label"] = f"data: {beta_post.num_success} of {beta_post.num_total}"
+  content.append(beta_post)
+
+draw(ax, content, xlims=(0., 55.))
+ax.set_title("Observed vs Predicted Prevalence\nof Midline Extension")
+ax.set_yticks([])
+ax.set_ylim(top=0.4)
+ax.set_xlabel("prevalence [%]")
+
+shared.split_legends(
+  ax=ax,
+  titles=["early T-cat.", "late T-cat."],
+  locs=[(0.05, 0.99), (0.6, 0.99)],
+)
+
+plt.show()
+```

### 2024-05-21 scripts/compute_midext_prevalences.py

@@ -29,7 +29,7 @@ def main():
         predicted["late"].append(model.marginalize(t_stage="late", midext=True))
 
     with h5py.File(
-        paths.model_dir / "full" / "prevalence_midext.hdf5",
+        paths.model_dir / "full" / "prevalences_midext.hdf5",
         mode="w",
     ) as h5_file:
         h5_file["early"] = predicted["early"]

### 2024-05-21 model.qmd
small tweaks in prevalence section

@@
 draw(ax, content, xlims=(0., 55.))
-ax.set_title("Observed vs Predicted Prevalence\nof Midline Extension")
+ax.set_title("Observed vs Predicted Prevalence\nof Midline Extension $\epsilon$")
 ax.set_yticks([])

### 2024-05-21 data.qmd

@@
 ---
-title: "Visualizing Contralateral Involvement"
+title: "Visualizing the Data"
 ---
 
-To motivate our bilateral model extension, we show how the contralateral nodal involvement in the data depends on ipsilateral involvement and the primary tumor's midline extension.
+# Contralateral Involvement Stratified by Risk Factors
+
+To motivate our bilateral model extension, we show how the contralateral nodal involvement in the data depends on T-category, the ipsilateral involvement, and midline extension of the primary tumor.
 
@@
 
-We also show an upsetplot
+# Involvement Patterns as UpSet Plots
+
+The joint involvement of different combinations of lymphatic involvements is best visualized using so-called *upset plots* [@lex_upset_2014].
+
+## Ipsilateral Involvement
 
 ```{python}
-#| label: fig-data-upset
+#| label: fig-data-upset-ipsi
 #| fig-cap: Upset plot of the ipsilateral lymphatic involvement patterns [@lex_upset_2014].
@@
 This is indicated by the light and dark shaded dots in the matrix underneath.
+from typing import Literal
 import upsetplot
 
-lnls = [tpl[-1] for tpl in get_lnl_cols("ipsi")]
+def plot_involvement_upset(
+  df: pd.DataFrame,
+  lnls: list[str],
+) -> Axes:
+  """Plot involvement of `side` as upsetplot."""
   indicator_data = upsetplot.from_indicators(
     indicators=lnls,
-  data=raw["max_llh", "ipsi"],
+    data=df,
   )
@@
   axes["intersections"].set_ylabel("Scenario size")
+  return axes
+
+axes = plot_involvement_upset(
+  df=raw["max_llh", "ipsi"],
+  lnls=[tpl[-1] for tpl in get_lnl_cols("ipsi")],
+)
 plt.show()
 ```
+
+## Contralateral Involvement
+
+```{python}
+#| label: fig-data-upset-contra
+#| fig-cap: Upset plot of the contralateral lymphatic involvement patterns [@lex_upset_2014].
+Here, the horizontal bars depict the overall prevalence of LNL involvement.
+I.e., LNL II is the only level that is involved in more than 100 patients.
+The vertical bar plots depict joint involvements.
+E.g., the thirs bar from the left depicts how often the LNLs II, and III are jointly involved while level I, IV, and V are not.
+This is indicated by the light and dark shaded dots in the matrix underneath.
+
+axes = plot_involvement_upset(
+  df=raw["max_llh", "contra"],
+  lnls=[tpl[-1] for tpl in get_lnl_cols("contra")],
+)
+plt.show()
+```

### 2024-05-22 manuscript.qmd
small tweaks in prevalence section
store figures instead of embedding notebook output

@@
 
-{{< embed data.qmd#fig-data-strat >}}
+![Contralateral involvement stratified by T-category (left panel), the number of metastatic LNLs ipsilaterally (center panel), and whether the primary tumor extended over the mid-sagittal line or was clearly lateralized (right panel).](figures/fig-data-strat.png){#fig-data-strat}
 
@@
 
-{{< embed model.qmd#fig-model-midext-evo >}}
+![The top panel shows the prior probability to get diagnosed at time-step $t$ for early and late T-category tumors as bars.
+Also in the top panel, we plot the conditional probability of the tumor's midline extension ($\epsilon=\texttt{True}$), given the time-step $t$ as a line plot.
+In the bottom panel, we show the joint probability of getting diagnosed in time-step $t$ *and* having a tumor that crosses the midline.](figures/fig-model-midext-evo.png){#fig-model-midext-evo width="50%"}
 
@@
 
-{{< embed model.qmd#fig-model-state-dist >}}
+![3D state distribution of a midline model with 3 LNLs (II, III, and IV) in both sides of the neck.](figures/fig-model-state-dist.png){#fig-model-state-dist}
 
@@
 
-{{< embed model.qmd#fig-model-burnin-history >}}
+![Monitoring quantities during the burn-in phase of the parameter sampling.](figures/fig-model-burnin-history.png){#fig-model-burnin-history}
 
@@
 
-{{< embed model.qmd#fig-model-prevalences-overall >}}
+![Comparison of predicted (histograms) vs observed (beta posteriors) prevalences.
+Shown for the contralateral LNLs II (blue), III (orange), and IV (green).
+The top row shows scenarios with early T-category tumors, the bottom row for late T-category ones.
+The left column depicts scenarios where the primary tumor is clearly lateralized, the right column scenarios of tumors extending over the mid-sagittal line.
+This figure illustrates the model's ability to describe the prevalence of different combinations of scenarios involving the risk factors T-category and midline extension.](figures/fig-model-prevalences-overall.png){#fig-model-prevalences-overall}
 
@@
 
-{{< embed model.qmd#fig-model-prevalences-with-ipsi >}}
+### Correlation between Ipsi- and Contralateral Involvement
+
+![Comparison of predicted (histograms) vs observed (beta posteriors) prevalences.
+Shown are four scenarios, all including contralateral LNL II involvement: Early T-category and an ipsilateral N0 neck (green), early T-category and ipsilateral LNL II involvement (blue), as well as the same two scenarios but for advanced T-category (orange and red).
+The figure shows that the model is capable of describing the correlation between ipsi- and contralateral involvement.
+Although for the scenario of LNL II involvement in both sides, the prediction's split between early and advanced T-category is not large enough.](figures/fig-model-prevalences-with-ipsi.png){#fig-model-prevalences-with-ipsi width="50%"}
 
@@
 
-{{< embed model.qmd#fig-model-prevalences-midext >}}
+### Midline Extension
+
+![Comparing the predicted (histograms) and observed (lines depicting beta posteriors) prevalence of midline extension for early (blue) and late (orange) T-category.
+While the prevalence is predicted correctly when marginalizing over T-category, the model cannot capture the degree of separation observed in the data.
+Since the tumor's midline extension is virtually always part of the diagnosis and hence *given* when predicting a patient's risk, we do not consider this discrepancy a major issue.](figures/fig-model-prevalences-midext.png){#fig-model-prevalences-midext width="50%"}
 

### 2024-05-22 model.qmd
store figures instead of embedding notebook output

@@
 #| label: fig-model-midext-evo
-#| fig-cap: The top panel shows the prior probability to get diagnosed at time-step $t$ for early and late T-category tumors as bars.
+#| fig-cap: "The top panel shows the prior probability to get diagnosed at time-step $t$ for early and late T-category tumors as bars.
 Also in the top panel, we plot the conditional probability of the tumor's midline extension ($\epsilon=\texttt{True}$), given the time-step $t$ as a line plot.
-In the bottom panel, we show the joint probability of getting diagnosed in time-step $t$ *and* having a tumor that crosses the midline.
+In the bottom panel, we show the joint probability of getting diagnosed in time-step $t$ *and* having a tumor that crosses the midline."
 import numpy as np
@@
 from lyscripts.plot.utils import COLORS
-from scripts import shared
+from scripts import shared, paths
 
@@
 
+plt.savefig(paths.figure_dir / "fig-model-midext-evo.png", dpi=400)
 plt.show()
@@
 import pandas as pd
+from scripts import shared, paths
 
@@
 axes[2].yaxis.set_major_formatter(ticker.FuncFormatter(custom))
+
+plt.savefig(paths.figure_dir / "fig-model-burnin-history.png", dpi=400)
 plt.show()
@@
 from lyscripts.plot.utils import COLORS
+from scripts import shared, paths
 
@@
 
+plt.savefig(paths.figure_dir / "fig-model-state-dist.png", dpi=400)
 plt.show()
@@
 
+plt.savefig(paths.figure_dir / "fig-model-prevalences-overall.png", dpi=400)
 plt.show()
@@
 
+plt.savefig(paths.figure_dir / "fig-model-prevalences-with-ipsi.png", dpi=400)
 plt.show()
@@
 
+plt.savefig(paths.figure_dir / "fig-model-prevalences-midext.png", dpi=400)
 plt.show()

### 2024-05-22 scripts/paths.py

@@ -8,6 +8,7 @@ data_dir = Path("data")
 data = data_dir / "reduced.csv"
 model_dir = Path("models")
 cache_dir = Path(".cache")
+figure_dir = Path("figures")
 
 
 def get_filename(

### 2024-05-28 data.qmd

@@
 )
+
+plt.savefig(paths.figure_dir / "fig-data-strat.png", dpi=400)
+plt.show()
 ```
@@
 import upsetplot
+from scripts import paths
 
@@
 )
+
+plt.savefig(paths.figure_dir / "fig-data-upset-ipsi.png", dpi=400)
 plt.show()
@@
 This is indicated by the light and dark shaded dots in the matrix underneath.
+from scripts import paths
 
@@
 )
+
+plt.savefig(paths.figure_dir / "fig-data-upset-contra.png", dpi=400)
 plt.show()

### 2024-05-28 manuscript.qmd

@@
 The orange arcs represent the sensitivity and specificity of the observational modality (e.g.
-CT, MRI, pathology, ...).](static/small-graph.png){#fig-small-graph width=60%}
+CT, MRI, pathology, ...).](static/small-graph.png){#fig-small-graph width=50%}
 
 Our first model to predict the lymphatic progression of HNSCC was introduced using Bayesian networks [@pouymayou_bayesian_2019].
-We subsequently extended this work to a hidden Markov model (HMM) [@ludwig_hidden_2020] to allow an intuitive inclusion of T-category into the predictions.
+We subsequently extended this work to a hidden Markov model (HMM) [@ludwig_hidden_2021] to allow an intuitive inclusion of T-category into the predictions.
 We will briefly summarize this HMM's formalism before building on it to include the contralateral spread in @sec-ext-to-contra.

### 2024-05-28 manuscript.qmd
improve burnin display

@@
 
-## Requirements for a Bilateral Model
+## Requirements for a Bilateral Model {#sec-requirements}
 
@@
 
-We compared three different aprroaches to modelling the contralateral involvement in OPSCC:
-
-1. A naive model of the contralateral side only. Neither the patterns of ipsilateral involvement, nor the information on whether the patients' tumors crossed the mid-sagittal line enter this model.
-2. The bilateral model as introduced in the sections up to and including @sec-params-symmetry. The involvement patterns of both sides enter this model, but not whether the tumors extended the midline.
-3.
-An additional bilateral model that tracks the midline extension of the patients' tumors as a random variable.
-This includes the formalism introduced up to and including @sec-midline.
-In contrast to the other two models, it does consider the reported midline extension in the data to attempt a more accurate prediction of the risk for contralateral occult disease.
+In this section, we
 
@@
 
-![Monitoring quantities during the burn-in phase of the parameter sampling.](figures/fig-model-burnin-history.png){#fig-model-burnin-history}
+First, in @fig-model-burnin-history, we verify the sampling converged successfully by inspecting two monitoring quantities: The autocorrelation time of the MCMC chain and the acceptance fractions of the parallel walkers.
 
-@fig-model-burnin-history is only a temporary placeholder for now.
+![Monitoring quantities during the burn-in phase of the parameter sampling.
+Left: The autocorrelation time of the sampling chain estimated at different sampling steps.
+We consider the chain converged when the estimate of the autocorrelation time is stable and drops below the trust threshold of $n/50$ where $n$ is the number of steps.
+Right: Fraction of accepted MCMC proposals averaged over all parallel walkers.
+Values around 30% indicate good mixing of the walkers.](figures/fig-model-burnin-history.png){#fig-model-burnin-history}
 
-## Prevalence Predictions
-
-<!-- NOTE: This should probably belong to the Methods/Formalism section(s) -->
-
-In @fig-model-prevalences-overall, we plot the prevalence of contralateral involvement of the LNLs II, III, and IV for the four scenarios made up of the possible combinations of early and late T-category, as well as lateralized and midline extending tumors.
-
-![Comparison of predicted (histograms) vs observed (beta posteriors) prevalences.
-Shown for the contralateral LNLs II (blue), III (orange), and IV (green).
-The top row shows scenarios with early T-category tumors, the bottom row for late T-category ones.
-The left column depicts scenarios where the primary tumor is clearly lateralized, the right column scenarios of tumors extending over the mid-sagittal line.
-This figure illustrates the model's ability to describe the prevalence of different combinations of scenarios involving the risk factors T-category and midline extension.](figures/fig-model-prevalences-overall.png){#fig-model-prevalences-overall}
-
-@Fig-model-prevalences-overall shows nicely how the model is capable of accurately taking the most important risk factors, i.e.
-T-category and midline extension, into account.
-As observed in the data, the model predicts that the prevalence of contralateral LNL II involvement jumps from below 8% for early T-category lateralized tumors to almost 40% when the tumor is of advanced T-category and crosses the mid-sagittal line.
-
-However, for early T-category scenarios with midline extension, the model does seem to overestimate contralateral LNL II invovlement and underestimate the prevalence for metastases in LNL III.
-This likely stems from the the small sample size of this rare scenario as hinted at by the wide Beta posteriors.
-
-### Correlation between Ipsi- and Contralateral Involvement
-
-![Comparison of predicted (histograms) vs observed (beta posteriors) prevalences.
-Shown are four scenarios, all including contralateral LNL II involvement: Early T-category and an ipsilateral N0 neck (green), early T-category and ipsilateral LNL II involvement (blue), as well as the same two scenarios but for advanced T-category (orange and red).
-The figure shows that the model is capable of describing the correlation between ipsi- and contralateral involvement.
-Although for the scenario of LNL II involvement in both sides, the prediction's split between early and advanced T-category is not large enough.](figures/fig-model-prevalences-with-ipsi.png){#fig-model-prevalences-with-ipsi width="50%"}
-
-In @fig-model-prevalences-with-ipsi we display the model's ability to capture the correlation between ipsi- and contralateral involvement.
-It shows that the prevalence of metastases in the two sides of the neck is correlated via the time of diagnosis, despite the model not having any direct connections between the two side.
-However, there are some small discrepancies in the model's prediction: When considering the scenario of LNL II involvement in the ipsi- *and* contralateral side, it cannot quite capture the split between early and late T-category.
-The model slightly overestimates the prevalence of this scenario for early T-category patients and underestimates it for advanced T-category.
-
-### Midline Extension
-
-![Comparing the predicted (histograms) and observed (lines depicting beta posteriors) prevalence of midline extension for early (blue) and late (orange) T-category.
-While the prevalence is predicted correctly when marginalizing over T-category, the model cannot capture the degree of separation observed in the data.
-Since the tumor's midline extension is virtually always part of the diagnosis and hence *given* when predicting a patient's risk, we do not consider this discrepancy a major issue.](figures/fig-model-prevalences-midext.png){#fig-model-prevalences-midext width="50%"}
-
-Lastly, in @fig-model-prevalences-midext, we plot the prevalence of midline extension in the data versus our model's prediction.
-It is obvious the model cannot match the large spread between early and advanced T-category seen in the data.
-This is because to achieve that, it would need to increase the advanced T-category patient's prior distribution over diagnosis times and at the same time reduce the probability of the tumor to cross the midline during a time-step.
-But since the time-priors parameter is also coupled with the spread probabilities among the LNLs, the model does not have that freedom.
-
-<!-- Should the paragraphs below be part of the discussion? -->
-
-However, we do not consider this discrepancy a major limitation of the model: We will not realistically be interested in the probability of midline extension, as it is always possible to assess it with high certainty.
-That is also the reason why we initially modelled the midline extension *not* as a random variable, but as a global risk factor that would have been turned on or off from the onset of a patient's disease evolution.
-This, however, lead to overly high risks for contralateral involvement in advanced T-category patients with midline extension, because then the model assumes an increased spread to the contralateral side from the onset of the disease.
-Which is probably not true in a majority of those cases.
-Thus, treating it as a random variable that only becomes true during a patient's disease evolution resulted in a better description of the data.
-
-Formally, the wrong prediction of midline extension prevalence makes little difference, since it is always given: Instead of $P\left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \mid \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \right)$, we typically compute $P\left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \mid \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c}, \epsilon \right)$, which does not suffer from the wrong probability of midline extension, as the distribution over hidden states is renormalized:
-
-$$
-P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \mid \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c}, \epsilon \right) = \frac{P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \mid \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right) P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)}{P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c}, \epsilon \right)}
-$$
-
-Note that a distribution over $\epsilon$ appears both in the enumerator and the denominator, which largely cancel each other, leaving only the midline extension's effect on the distribution over hidden states in the prediction.
-
-## Prediction of Risk for Occult Disease
-
-- compute risks for occult disease, given diagnosis
-
-## Parameter Estimates
+In @tbl-midline-params, we tabulate the mean and standard deviation of the sampled parameters for the full midline model.
 
@@
 
-The predicted prevalence of midline extension for early T-category tumors is <!-- insert computational code later -->. For a late T-category tumor, it is <!-- insert computational code later -->.
+## Prevalence Predictions
 
-In the data, the prevalence of midline extension for early T-category tumors is <!-- insert computational code later -->. For late T-category tumors it is <!-- insert computational code later -->.
+We want to investigate whether and to what extent the model can fulfill the requirements laid out in @sec-requirements.
+To that end, we compare its predictions for contralateral involvement against observations in the data.
+This is done given scenarios that differ in T-category and/or midline extension and/or ipsilateral involvement.
 
-- Discrepancy between data and model is due to late T-stage's time prior
-- still tied to other parameters
-- cannot fully represent spread of midline extension and also represent spread params
+### Dependence of Contralateral Involvement on T-Category and Midline Extension
+
+In @fig-model-prevalences-overall, we plot the prevalence of contralateral involvement of the LNLs II, III, and IV for the four scenarios made up of the possible combinations of early and late T-category, as well as lateralized and midline extending tumors.
+
+![Comparison of predicted (histograms) vs observed (beta posteriors) prevalences.
+Shown for the contralateral LNLs II (blue), III (orange), and IV (green).
+The top row shows scenarios with early T-category tumors, the bottom row for late T-category ones.
+The left column depicts scenarios where the primary tumor is clearly lateralized, the right column scenarios of tumors extending over the mid-sagittal line.
+This figure illustrates the model's ability to describe the prevalence of different combinations of scenarios involving the risk factors T-category and midline extension.](figures/fig-model-prevalences-overall.png){#fig-model-prevalences-overall}
+
+@Fig-model-prevalences-overall shows nicely how the model is capable of accurately taking the most important risk factors, i.e.
+T-category and midline extension, into account.
+As observed in the data, the model predicts that the prevalence of contralateral LNL II involvement jumps from below 8% for early T-category lateralized tumors to almost 40% when the tumor is of advanced T-category and crosses the mid-sagittal line.
+
+However, for early T-category scenarios with midline extension, the model does seem to overestimate contralateral LNL II invovlement and underestimate the prevalence for metastases in LNL III.
+This likely stems from the the small sample size of this rare scenario as hinted at by the wide Beta posteriors.
+
+### Correlation between Ipsi- and Contralateral Involvement
+
+![Comparison of predicted (histograms) vs observed (beta posteriors) prevalences.
+Shown are four scenarios, all including contralateral LNL II involvement: Early T-category and an ipsilateral N0 neck (green), early T-category and ipsilateral LNL II involvement (blue), as well as the same two scenarios but for advanced T-category (orange and red).
+The figure shows that the model is capable of describing the correlation between ipsi- and contralateral involvement.
+Although for the scenario of LNL II involvement in both sides, the prediction's split between early and advanced T-category is not large enough.](figures/fig-model-prevalences-with-ipsi.png){#fig-model-prevalences-with-ipsi width="50%"}
+
+In @fig-model-prevalences-with-ipsi we display the model's ability to capture the correlation between ipsi- and contralateral involvement.
+It shows that the prevalence of metastases in the two sides of the neck is correlated via the time of diagnosis, despite the model not having any direct connections between the two side.
+However, there are some small discrepancies in the model's prediction: When considering the scenario of LNL II involvement in the ipsi- *and* contralateral side, it cannot quite capture the split between early and late T-category.
+The model slightly overestimates the prevalence of this scenario for early T-category patients and underestimates it for advanced T-category.
+
+### Prevalence of Midline Extension
+
+![Comparing the predicted (histograms) and observed (lines depicting beta posteriors) prevalence of midline extension for early (blue) and late (orange) T-category.
+While the prevalence is predicted correctly when marginalizing over T-category, the model cannot capture the degree of separation observed in the data.
+Since the tumor's midline extension is virtually always part of the diagnosis and hence *given* when predicting a patient's risk, we do not consider this discrepancy a major issue.](figures/fig-model-prevalences-midext.png){#fig-model-prevalences-midext width="50%"}
+
+Lastly, in @fig-model-prevalences-midext, we plot the prevalence of midline extension in the data versus our model's prediction.
+It is obvious the model cannot match the large spread between early and advanced T-category seen in the data.
+This is because to achieve that, it would need to increase the advanced T-category patient's prior distribution over diagnosis times and at the same time reduce the probability of the tumor to cross the midline during a time-step.
+But since the time-priors parameter is also coupled with the spread probabilities among the LNLs, the model does not have that freedom.
+
+<!-- Should the paragraphs below be part of the discussion? -->
+
+However, we do not consider this discrepancy a major limitation of the model: We will not realistically be interested in the probability of midline extension, as it is always possible to assess it with high certainty.
+That is also the reason why we initially modelled the midline extension *not* as a random variable, but as a global risk factor that would have been turned on or off from the onset of a patient's disease evolution.
+This, however, lead to overly high risks for contralateral involvement in advanced T-category patients with midline extension, because then the model assumes an increased spread to the contralateral side from the onset of the disease.
+Which is probably not true in a majority of those cases.
+Thus, treating it as a random variable that only becomes true during a patient's disease evolution resulted in a better description of the data.
+
+Formally, the wrong prediction of midline extension prevalence makes little difference, since it is always given: Instead of $P\left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \mid \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \right)$, we typically compute $P\left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \mid \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c}, \epsilon \right)$, which does not suffer from the wrong probability of midline extension, as the distribution over hidden states is renormalized:
+
+$$
+P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \mid \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c}, \epsilon \right) = \frac{P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \mid \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right) P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)}{P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c}, \epsilon \right)}
+$$
+
+Note that a distribution over $\epsilon$ appears both in the enumerator and the denominator, which largely cancel each other, leaving only the midline extension's effect on the distribution over hidden states in the prediction.
+
+## Prediction of Risk for Occult Disease
+
+- compute risks for occult disease, given diagnosis
 

### 2024-05-28 model.qmd
improve burnin display

@@
 from scripts import shared, paths
+import matplotlib.pyplot as plt
+from matplotlib import ticker
+
+params = shared.get_params()
 
@@
 
-nrows, ncols = 1, 3
-label_map = {
-  "acor_times": "autocorrelation [steps]",
-  "accept_fracs": "acceptance fraction [%]",
-  "max_log_probs": r"max. log-likelihood [$\times 10^3$]",
+PlotConfig = namedtuple("PlotConfig", ["title", "ylabel", "label"])
+
+nrows, ncols = 1, 2
+config_map = {
+  "acor_times": PlotConfig(
+    title="Estimated Autocorrelation Time",
+    ylabel="steps",
+    label="Autocorrelation Time",
+  ),
+  "accept_fracs": PlotConfig(
+    title="Average Acceptance Fraction of Walkers",
+    ylabel=None,
+    label=None,
+  ),
 }
@@
   ncols=ncols,
-  aspect_ratio=1.,
   width=17,
@@
 
-for model in ["simple"]:
+for model in ["full"]:
   history = pd.read_csv(f"models/{model}/history.csv").set_index("steps")
-  for i, column in enumerate(history.columns):
-    history.plot(y=column, ax=axes[i], label=model)
+  for i, (column, config) in enumerate(config_map.items()):
+    axes[i].set_title(config.title, fontweight="bold")
+    axes[i].plot(history.index, history[column], label=config.label)
     axes[i].autoscale(enable=True, tight=True, axis="x")
-    axes[i].set_ylabel(label_map[column])
+    axes[i].set_ylabel(config.ylabel)
     axes[i].xaxis.set_major_formatter(ticker.FuncFormatter(custom))
@@
 
+axes[0].plot(
+  history.index,
+  history.index / params["sampling"]["trust_fac"],
+  "--",
+  color="black",
+  label="Trust Threshold",
+)
+axes[0].legend()
 axes[1].yaxis.set_major_formatter(ticker.StrMethodFormatter("{x:.0%}"))
-axes[2].yaxis.set_major_formatter(ticker.FuncFormatter(custom))
 

### 2024-05-28 scripts/paths.py

@@ -9,6 +9,7 @@ data = data_dir / "reduced.csv"
 model_dir = Path("models")
 cache_dir = Path(".cache")
 figure_dir = Path("figures")
+params = Path("_variables.yml")
 
 
 def get_filename(

### 2024-05-28 scripts/shared.py

@@ -19,6 +19,11 @@ CM_TO_INCH = 0.393701
 GOLDEN_RATIO = 1.61803398875
 
 
+def get_params() -> dict[str, Any]:
+    """Get the parameters."""
+    return utils.load_yaml_params(paths.params)
+
+
 def get_model(
     which: str,
     load_samples: bool = True,

### 2024-05-28 manuscript.qmd
use pip-tools to specify environment
add parapgraphs on data consensus

@@
 
-In this section, we
+<!-- Maybe I should break this apart into separate chapters on data, sampling, etc. -->
+
+In this section, we detail how the experiments were performed.
+Every figure, table, and result is fully reproducible via the GitHub repository [`rmnldwg/bilateral-paper`].
+It also contains the raw manuscript and instructions on how to recreate all figures, tables, and the final document.
+
+[`rmnldwg/bilateral-paper`]: https://github.com/rmnldwg/bilateral-paper
+
+## Involvement Data Consensus
+
+It is possible to provide our model with multiple different diagnostic modalities, each being characterized by different pairs of sensitivity and specificity.
+However, we instead chose to combine them into a single "consensus" diagnosis before parameter inference.
+We opted for this because the literature values of sensitivity and specificity [@debondt_detection_2007] of imaging modalities like MRI and CT do not plausibly match some of our observations: In the USZ cohort, 78% of OPSCC patients where diagnosed with ipsilateral LNL II involvement via diagnostic imaging.
+This is virtually impossible with sensitivities around 80% and specificities lower than 100%.
+
+Our pre-training consensus was formed by considering all reported diagnostic information for a particular patient and LNL.
+When conflicts arose, we computed the *most likely* true state of involvement using the literature sensitivity and specificity values [@debondt_detection_2007].
 
@@
 
-- all models initialized with the same settings
-  - same graph
-  - same data
-  - same distributions over diagnose time
-- parameter samples were drawn for all three models:
-  - the sampling used {{< var sampling.walkers_per_dim >}} walkers per parameter space dimension.
-  - Every {{< var sampling.check_interval >}} steps, convergence was checked
-  - convergence was reached when two criteria were met:
-      1. the relative difference between subsequent autocorrelation estimates of the chain does not change more than {{< var sampling.rel_thresh |  >}}.
-      2. the autocorrelation estimate of the chain drops below the length of the chain divided by {{< var sampling.trust_fac >}}.
-  - after convergence, {{< var sampling.nsteps >}} samples (from the parallel walkers) were drawn that were spaced {{< var sampling.thin >}} steps apart, resulting in `num params x {{< var sampling.walkers_per_dim >}} x {{< var sampling.nsteps >}}` parameter samples.
+For parameter inference, we used the Python package [`emcee`] [@foreman-mackey_emcee_2013].
+It implements efficient MCMC sampling algorithms that employ multiple parallel samplers for affine invariance and better performance on multi-core CPUs.
+The [`emcee`] library was provided with the likelihood implemented by our [`lymph-model`] Python package.
 
-For every model, we compare the predicted prevalence of a certain scenario to how often this scenario is seen in the data.
-The considered scenarios are choosen to test if the models can take into account T-category, the correlation of ipsi- and contralateral involvement, and the primary tumor's lateralization (whether or not it extends over the mid-sagittal line).
+For each dimension in the parameter space of the model, we initialized {{< var sampling.walkers_per_dim >}} of these parallel samplers, called "walkers", with random values in the unit cube.
+Every time all of these walkers advanced {{< var sampling.check_interval >}} steps, the autocorrelation time of the chains was estimated.
+For short chains, this estimate is not trustworthy, but stabilizes for longer chains.
+We therefore considered a sampling to be converged when two criteria where met:
 
-- [ ] list and describe the scenarios
+1. The change in the autocorrelation time was less then {{< var sampling.rel_thresh >}}.
+2. The estimate of the autocorrelation dropped below $n$ / {{< var sampling.trust_fac >}} where $n$ is the length of the chain up to that point.
+
+All samples up to this convergence - called the *burn-in phase* - were discarded.
+We only kept another {{< var sampling.nsteps >}} samples after that, which were spaced {{< var sampling.thin >}} steps apart.
+
+[`emcee`]: https://emcee.readthedocs.io/en/stable/
+[`lymph-model`]: https://lymph-model.readthedocs.io/en/stable/
 

### 2024-05-29 model.qmd
split ipsi scenarios by midext

@@
 This figure illustrates the model's ability to describe the prevalence of different combinations of scenarios involving the risk factors T-category and midline extension."
+import matplotlib.pyplot as plt
 from matplotlib import gridspec
+import numpy as np
 from lyscripts.scenario import Scenario

### 2024-05-29 model.qmd
work on contra inv dep. on ipsi III

@@
 for t_stage in ["early", "late"]:
-  for diagnosis in [{"II": True}, {lnl: False for lnl in ["I", "II", "III", "IV", "V"]}]:
+  for diagnosis in [{"III": True}, {lnl: False for lnl in ["I", "II", "III", "IV", "V"]}]:
     color = next(colors)
@@
       t_stages=[t_stage],
+      midext=True,
       diagnosis={

### 2024-05-29 scripts/shared.py

@@ -1,6 +1,7 @@
 """
 Some shared and reused functions for the scripts.
 """
+
 from typing import Any, Literal
 
 from lymph import types
@@ -42,9 +43,7 @@ def get_model(
     return model
 
 
-def get_samples(
-    which: str
-) -> np.ndarray:
+def get_samples(which: str) -> np.ndarray:
     """Get the samples of one of the four trained models."""
     samples_path = paths.model_dir / which / "samples.hdf5"
     return utils.load_model_samples(samples_path)
@@ -125,8 +124,8 @@ def split_legends(
     handles, labels = ax.get_legend_handles_labels()
     for i, (title, loc) in enumerate(zip(titles, locs)):
         legend = ax.legend(
-            handles[i * 2 : (i+1) * 2],
-            labels[i * 2 : (i+1) * 2],
+            handles[i * 2 : (i + 1) * 2],
+            labels[i * 2 : (i + 1) * 2],
             loc="upper left",
             bbox_to_anchor=loc,
             title=title,

### 2024-05-30 model.qmd
improve beta pdfs with unknowns

--- 2024-05-29 model.qmd
@@
 import matplotlib.pyplot as plt
+from matplotlib import gridspec
+import numpy as np
 from lyscripts.scenario import Scenario
@@
 from scripts import paths, shared
+from scripts.marginalize_midext import (
+  compute_possible_observed_prevalences,
+  SummedBetaPosterior,
+)
+
+def turn_axis_off(axes):
+  axes.set_xticks([])
+  axes.set_yticks([])
+
+nrows, ncols = 2, 2
 
 plt.rcParams.update(shared.get_fontsizes())
-plt.rcParams.update(shared.get_figsizes(width=17./2.))
+plt.rcParams.update(shared.get_figsizes(
+  nrows=nrows,
+  ncols=ncols,
+  aspect_ratio=2.,
+  width=17,
+))
 
-fig, ax = plt.subplots()
+fig = plt.figure()
+gs = gridspec.GridSpec(nrows=nrows, ncols=ncols, figure=fig)
+axes = np.array([[fig.add_subplot(gs[i,j]) for j in range(ncols)] for i in range(nrows)])
+full = fig.add_subplot(gs[:,:], frame_on=False)
+turn_axis_off(full)
+
+full.set_title(
+  "Observed vs Predicted Prevalence of Contralateral LNL II Involvement\nDependent on Ipsilateral Involvement",
+  pad=20.,
+  fontsize="large",
+)
+
+for i, t_stage in enumerate(["early", "late"]):
+  axes[i,0].set_ylabel(f"{t_stage} T-cat.", fontweight="bold")
+  for j, midext in enumerate([False, True]):
+    axes[0,j].set_title("Mid. ext." if midext else "Lateralized", fontweight="bold")
+    axes[0,j].set_xticks([])
+    axes[-1,j].set_xlabel("prevalence [%]")
+    axes[i,j].set_yticks([])
     content = []
-colors = iter([COLORS[c] for c in ["orange", "green", "red", "blue"]])
-
-for t_stage in ["early", "late"]:
-  for diagnosis in [{"II": True}, {lnl: False for lnl in ["I", "II", "III", "IV", "V"]}]:
-    color = next(colors)
+    for ipsi, color in zip(
+      [{"I": False, "II": False, "III": False, "IV": False, "V": False}, {"III": True}],
+      [COLORS["blue"], COLORS["orange"]],
+    ):
       scenario = Scenario(
         t_stages=[t_stage],
+        midext=midext,
         diagnosis={
-        "ipsi": {"max_llh": diagnosis},
+          "ipsi": {"max_llh": ipsi},
           "contra": {"max_llh": {"II": True}},
@@
           color=color,
-        label="prediction"
+          label="prediction",
         )
@@
       content.append(beta_post)
+    draw(
+      axes=axes[i,j],
+      contents=content,
+      xlims=(0., 30. if midext else 10.),
+    )
 
-draw(ax, content, xlims=(0., 25.))
-ax.set_title("Observed vs Predicted Prevalence\nof Contralateral LNL II Involvement")
-ax.set_yticks([])
-ax.set_xlabel("prevalence [%]")
-
+# manually place legends
 shared.split_legends(
-  ax=ax,
-  titles=["early with ipsi II", "early ipsi N0", "late with ipsi II", "late with ipsi N0"],
-  locs=[(0.23, 0.65), (0.1, 0.99), (0.6, 0.65), (0.45, 0.99)],
+  ax=axes[0,0],
+  titles=["ipsi N0", "ipsi LNL III"],
+  locs=[(0.3, 0.99), (0.45, 0.55)],
+)
+shared.split_legends(
+  ax=axes[0,1],
+  titles=["ipsi N0", "ipsi LNL III"],
+  locs=[(0.3, 0.99), (0.5, 0.55)],
+)
+shared.split_legends(
+  ax=axes[1,0],
+  titles=["ipsi N0", "ipsi LNL III"],
+  locs=[(0.2, 0.99), (0.55, 0.99)],
+)
+shared.split_legends(
+  ax=axes[1,1],
+  titles=["ipsi N0", "ipsi LNL III"],
+  locs=[(0.2, 0.99), (0.65, 0.99)],
 )

### 2024-05-30 scripts/marginalize_midext.py

@@ -1,0 +1,141 @@
+"""
+Marginalize over *observd* midline extension status using sums of beta posteriors.
+"""
+
+from pathlib import Path
+import matplotlib.pyplot as plt
+import pandas as pd
+import numpy as np
+from scipy.stats import beta
+from scipy import integrate
+from lyscripts.data import accessor  # noqa: F401
+from lyscripts.scenario import Scenario
+from lyscripts.plot.utils import BetaPosterior
+
+try:
+    from . import paths, shared
+except ImportError:
+    import paths
+    import shared
+
+
+def max_possible_matches(does_match, num_total):
+    """Maximum number of additional possible matches."""
+    return min(num_total, does_match.sum()) + 1
+
+def min_possible_matches(does_match, num_total):
+    """Minimum number of additional possible matches."""
+    return min(num_total, max(0, num_total - (~does_match).sum()))
+
+
+def compute_possible_observed_prevalences(
+    data: pd.DataFrame, scenario: Scenario, modality: str = "max_llh"
+) -> np.ndarray:
+    """Extract possible combinations of prevalences."""
+    # when looking at the data, we always consider both sides
+    is_uni = scenario.is_uni
+    scenario.is_uni = False
+    diagnosis_pattern = scenario.get_pattern(get_from="diagnosis", modality=modality)
+
+    # reset `is_uni` to the original value. Otherwise hash computation will fail.
+    scenario.is_uni = is_uni
+
+    has_t_stage = data.ly.t_stage.isin(scenario.t_stages)
+    has_midext = data.ly.is_midext(scenario.midext)
+    eligible_data = data.loc[has_t_stage & has_midext].reset_index()
+    does_pattern_match = eligible_data.ly.match(diagnosis_pattern, modality)
+    base_num_total = len(eligible_data)
+
+    try:
+        matching_data = eligible_data.loc[does_pattern_match]
+        base_num_match = len(matching_data)
+    except KeyError:
+        # return X, X if no actual pattern was selected
+        base_num_match = base_num_total
+
+    has_unknown_midext = data.ly.midext.isna()
+    unknown_midext_data = data.loc[has_t_stage & has_unknown_midext].reset_index()
+    does_pattern_match = unknown_midext_data.ly.match(diagnosis_pattern, modality)
+    num_eligible = len(unknown_midext_data)
+
+    results = []
[…]

### 2024-06-03 model.qmd
work on marginal over beta posteriors

--- 2024-05-29 model.qmd
@@
 from lyscripts.scenario import Scenario
-from lyscripts.plot.utils import Histogram, BetaPosterior, draw, COLORS
+from lyscripts.plot.utils import Histogram, BetaPosterior, draw, COLORS, split_legends
 from scripts import paths, shared
@@
 # manually place legends
-shared.split_legends(
-  ax=axes[0,0],
+split_legends(
+  axes=axes[0,0],
   titles=["LNL II", "LNL III", "LNL IV"],
@@
 )
-shared.split_legends(
-  ax=axes[0,1],
+split_legends(
+  axes=axes[0,1],
   titles=["LNL II", "LNL III", "LNL IV"],
@@
 )
-shared.split_legends(
-  ax=axes[1,0],
+split_legends(
+  axes=axes[1,0],
   titles=["LNL II", "LNL III", "LNL IV"],
@@
 )
-shared.split_legends(
-  ax=axes[1,1],
+split_legends(
+  axes=axes[1,1],
   titles=["LNL II", "LNL III", "LNL IV"],
@@
 import matplotlib.pyplot as plt
+from matplotlib import gridspec
+import numpy as np
 from lyscripts.scenario import Scenario
-from lyscripts.plot.utils import Histogram, BetaPosterior, draw, COLORS
+from lyscripts.plot.utils import Histogram, BetaPosterior, draw, COLORS, split_legends
 from scripts import paths, shared
+from scripts.marginalize_midext import (
+  compute_possible_observed_prevalences,
+  SummedBetaPosterior,
+)
+
+def turn_axis_off(axes):
+  axes.set_xticks([])
+  axes.set_yticks([])
+
+nrows, ncols = 2, 2
 
 plt.rcParams.update(shared.get_fontsizes())
-plt.rcParams.update(shared.get_figsizes(width=17./2.))
+plt.rcParams.update(shared.get_figsizes(
+  nrows=nrows,
+  ncols=ncols,
+  aspect_ratio=2.,
+  width=17,
+))
 
-fig, ax = plt.subplots()
+fig = plt.figure()
+gs = gridspec.GridSpec(nrows=nrows, ncols=ncols, figure=fig)
+axes = np.array([[fig.add_subplot(gs[i,j]) for j in range(ncols)] for i in range(nrows)])
+full = fig.add_subplot(gs[:,:], frame_on=False)
+turn_axis_off(full)
+
+full.set_title(
+  "Observed vs Predicted Prevalence of Contralateral LNL II Involvement\nDependent on Ipsilateral Involvement",
+  pad=20.,
+  fontsize="large",
+)
+
+data = shared.get_data()
+
+for i, t_stage in enumerate(["early", "late"]):
+  axes[i,0].set_ylabel(f"{t_stage} T-cat.", fontweight="bold")
+  for j, midext in enumerate([False, True]):
+    axes[0,j].set_title("Mid. ext." if midext else "Lateralized", fontweight="bold")
+    axes[0,j].set_xticks([])
+    axes[-1,j].set_xlabel("prevalence [%]")
+    axes[i,j].set_yticks([])
     content = []
-colors = iter([COLORS[c] for c in ["orange", "green", "red", "blue"]])
-
-for t_stage in ["early", "late"]:
-  for diagnosis in [{"II": True}, {lnl: False for lnl in ["I", "II", "III", "IV", "V"]}]:
-    color = next(colors)
+    for ipsi, color in zip(
+      [{"I": False, "II": False, "III": False, "IV": False, "V": False}, {"III": True}],
+      [COLORS["blue"], COLORS["orange"]],
+    ):
       scenario = Scenario(
         t_stages=[t_stage],
+        midext=midext,
         diagnosis={
-        "ipsi": {"max_llh": diagnosis},
+          "ipsi": {"max_llh": ipsi},
           "contra": {"max_llh": {"II": True}},
@@
           color=color,
-        label="prediction"
+          label="prediction",
         )
       )
-    beta_post = BetaPosterior.from_hdf5(
+      content.append(
+        BetaPosterior.from_hdf5(
           filename=paths.model_dir / "full" / "prevalences_with_ipsi.hdf5",
@@
           color=color,
+          label="beta",
         )
-    beta_post.kwargs["label"] = f"data: {beta_post.num_success} of {beta_post.num_total}"
-    content.append(beta_post)
+      )
+      num_success, num_total = compute_possible_observed_prevalences(data, scenario).T
+      content.append(
+          SummedBetaPosterior(
+          num_success=num_success,
+          num_total=num_total,
+          kwargs={"color": "red"},
+        )
+      )
+    draw(
+      axes=axes[i,j],
+      contents=content,
+      xlims=(0., 30. if midext else 10.),
+    )
 
-draw(ax, content, xlims=(0., 25.))
-ax.set_title("Observed vs Predicted Prevalence\nof Contralateral LNL II Involvement")
-ax.set_yticks([])
-ax.set_xlabel("prevalence [%]")
-
-shared.split_legends(
-  ax=ax,
-  titles=["early with ipsi II", "early ipsi N0", "late with ipsi II", "late with ipsi N0"],
-  locs=[(0.23, 0.65), (0.1, 0.99), (0.6, 0.65), (0.45, 0.99)],
+# manually place legends
+split_legends(
+  axes=axes[0,0],
+  titles=["ipsi N0", "ipsi LNL III"],
+  locs=[(0.3, 0.99), (0.45, 0.55)],
+)
+split_legends(
+  axes=axes[0,1],
+  titles=["ipsi N0", "ipsi LNL III"],
+  locs=[(0.3, 0.99), (0.5, 0.55)],
+)
+split_legends(
+  axes=axes[1,0],
+  titles=["ipsi N0", "ipsi LNL III"],
+  locs=[(0.2, 0.99), (0.55, 0.99)],
+)
+split_legends(
+  axes=axes[1,1],
+  titles=["ipsi N0", "ipsi LNL III"],
+  locs=[(0.2, 0.99), (0.65, 0.99)],
 )
@@
 from lyscripts.scenario import Scenario
-from lyscripts.plot.utils import Histogram, BetaPosterior, draw, COLORS
+from lyscripts.plot.utils import Histogram, BetaPosterior, draw, COLORS, split_legends
 from scripts import paths, shared
+from scripts.marginalize_midext import compute_match_total_combinations, SummedBetaPosterior
 
@@
 
+data = shared.get_data()
+has_unknown_midext = data.ly.midext.isna()
+
 for t_stage in ["early", "late"]:
+  has_t_stage = data.ly.t_stage.isin([t_stage])
+  scenario = Scenario(
+    t_stages=[t_stage],
+    midext=True,
+    diagnosis={"ipsi": {"max_llh": {"II": None}}},
+  )
   color = next(colors)
@@
   )
-  beta_post.kwargs["label"] = f"data: {beta_post.num_success} of {beta_post.num_total}"
-  content.append(beta_post)
+  base_match = beta_post.num_success
+  base_total = beta_post.num_total
+  num_success, num_total = compute_match_total_combinations(
+    base_match=base_match,
+    base_total=base_total,
+    unknown_match=(has_unknown_midext & has_t_stage).sum(),
+    unknown_total=(has_unknown_midext & has_t_stage).sum(),
+  ).T
+  content.append(
+    SummedBetaPosterior(
+      num_success=num_success,
+      num_total=num_total,
+    )
+  )
 
@@
 
-shared.split_legends(
-  ax=ax,
+split_legends(
+  axes=ax,
   titles=["early T-cat.", "late T-cat."],

### 2024-06-03 scripts/marginalize_midext.py

@@ -2,15 +2,17 @@
 Marginalize over *observd* midline extension status using sums of beta posteriors.
 """
 
-from pathlib import Path
+from dataclasses import dataclass
+from typing import Any, TypeVar
 import matplotlib.pyplot as plt
+from matplotlib.axes import Axes as MPLAxes
 import pandas as pd
 import numpy as np
 from scipy.stats import beta
-from scipy import integrate
 from lyscripts.data import accessor  # noqa: F401
 from lyscripts.scenario import Scenario
 from lyscripts.plot.utils import BetaPosterior
+from lyscripts.compute.prevalences import compute_observed_prevalence, get_modality_subset
 
 try:
     from . import paths, shared
@@ -19,13 +21,13 @@ except ImportError:
     import shared
 
 
-def max_possible_matches(does_match, num_total):
+def max_possible_matches(num_match: int, num_total: int) -> int:
     """Maximum number of additional possible matches."""
-    return min(num_total, does_match.sum()) + 1
+    return min(num_total, num_match) + 1
 
-def min_possible_matches(does_match, num_total):
+def min_possible_matches(num_not_match: int, num_total: int) -> int:
     """Minimum number of additional possible matches."""
-    return min(num_total, max(0, num_total - (~does_match).sum()))
+    return min(num_total, max(0, num_total - num_not_match))
 
 
 def compute_possible_observed_prevalences(
@@ -33,46 +35,63 @@ def compute_possible_observed_prevalences(
 ) -> np.ndarray:
     """Extract possible combinations of prevalences."""
     # when looking at the data, we always consider both sides
-    is_uni = scenario.is_uni
-    scenario.is_uni = False
-    diagnosis_pattern = scenario.get_pattern(get_from="diagnosis", modality=modality)
-
-    # reset `is_uni` to the original value. Otherwise hash computation will fail.
-    scenario.is_uni = is_uni
+    base_num_match, base_num_total = compute_observed_prevalence(data, scenario)
 
     has_t_stage = data.ly.t_stage.isin(scenario.t_stages)
-    has_midext = data.ly.is_midext(scenario.midext)
-    eligible_data = data.loc[has_t_stage & has_midext].reset_index()
-    does_pattern_match = eligible_data.ly.match(diagnosis_pattern, modality)
-    base_num_total = len(eligible_data)
-
-    try:
[…]

### 2024-06-03 scripts/shared.py

@@ -106,29 +106,3 @@ def get_figsizes(
         "savefig.bbox": "tight",
         "savefig.pad_inches": pad,
     }
-
-
-def split_legends(
-    ax,
-    titles: list[str],
-    locs: list[tuple[float, float]],
-    **kwargs,
-) -> None:
-    """Separate the legends of `ax` into separate ones with `titles` and at `locs`."""
-    legend_kwargs = {
-        "title_fontsize": "small",
-        "labelspacing": 0.1,
-    }
-    legend_kwargs.update(kwargs)
-
-    handles, labels = ax.get_legend_handles_labels()
-    for i, (title, loc) in enumerate(zip(titles, locs)):
-        legend = ax.legend(
-            handles[i * 2 : (i + 1) * 2],
-            labels[i * 2 : (i + 1) * 2],
-            loc="upper left",
-            bbox_to_anchor=loc,
-            title=title,
-            **legend_kwargs,
-        )
-        ax.add_artist(legend)

### 2024-06-04 model.qmd
show comp beta vs marginalized beta dists

@@
 from scripts import paths, shared
+from scripts.marginalize_midext import compute_possible_observed_prevalences, SummedBetaPosterior
 
@@
 
+data = shared.get_data()
+
 for i, t_stage in enumerate(["early", "late"]):
@@
       )
-      beta_post = BetaPosterior.from_hdf5(
+      content.append(
+        BetaPosterior.from_hdf5(
           filename=paths.model_dir / "full" / "prevalences_overall.hdf5",
@@
         )
-      beta_post.kwargs["label"] = f"data: {beta_post.num_success} of {beta_post.num_total}"
-      content.append(beta_post)
+      )
+      num_success, num_total = compute_possible_observed_prevalences(data, scenario).T
+      content.append(
+        SummedBetaPosterior(
+          num_success=num_success,
+          num_total=num_total,
+          kwargs={"label": None, "linestyle": "--"},
+        )
+      )
     draw(
@@
           num_total=num_total,
-          kwargs={"color": "red"},
+          kwargs={"linestyle": "--"},
         )
@@
   )
-  base_match = beta_post.num_success
-  base_total = beta_post.num_total
+  content.append(beta_post)
   num_success, num_total = compute_match_total_combinations(
-    base_match=base_match,
-    base_total=base_total,
+    base_match=beta_post.num_success,
+    base_total=beta_post.num_total,
     unknown_match=(has_unknown_midext & has_t_stage).sum(),
@@
       num_total=num_total,
+      kwargs={"linestyle": "--", "label": None},
     )

### 2024-06-04 scripts/marginalize_midext.py

@@ -187,7 +187,9 @@ class SummedBetaPosterior(BetaPosterior):
 
         plot_kwargs = defaults["plot"].copy()
         plot_kwargs.update(self.kwargs)
-        plot_kwargs["label"] = self.label
+
+        if self.label is not None:
+            plot_kwargs["label"] = self.label
 
         return axes.plot(x, y, **plot_kwargs)
 

### 2024-06-04 model.qmd
change labels, linestyle

@@
           color=color,
-          label="beta",
         )
@@
           num_total=num_total,
-          kwargs={"linestyle": "--"},
+          kwargs={"linestyle": "--", "label": None},
         )

### 2024-06-04 model.qmd
try to fix marginalization (WIP)

@@
 from scripts import paths, shared
-from scripts.marginalize_midext import compute_match_total_combinations, SummedBetaPosterior
+from scripts.marginalize_midext import possible_combinations, SummedBetaPosterior
 
@@
   )
+  unknowns = (has_unknown_midext & has_t_stage).sum()
   color = next(colors)
@@
   content.append(beta_post)
-  num_success, num_total = compute_match_total_combinations(
-    base_match=beta_post.num_success,
-    base_total=beta_post.num_total,
-    unknown_match=(has_unknown_midext & has_t_stage).sum(),
-    unknown_total=(has_unknown_midext & has_t_stage).sum(),
-  ).T
+  combs = possible_combinations(
+    matches=unknowns,
+    max_total=unknowns,
+  )
   content.append(
     SummedBetaPosterior(
-      num_success=num_success,
-      num_total=num_total,
-      kwargs={"linestyle": "--", "label": None},
+      num_success=combs.matches.values + beta_post.num_success,
+      num_total=combs.total.values + beta_post.num_total,
+      permutations=combs.permutations.values,
+      kwargs={"linestyle": "--", "label": None, "color": "red"},
     )
   )
+  print(f"{unknowns = }, {beta_post.num_success = }, {beta_post.num_total = }")
 

### 2024-06-04 scripts/marginalize_midext.py

@@ -2,6 +2,7 @@
 Marginalize over *observd* midline extension status using sums of beta posteriors.
 """
 
+import math
 from dataclasses import dataclass
 from typing import Any, TypeVar
 import matplotlib.pyplot as plt
@@ -26,15 +27,22 @@ def max_possible_matches(num_match: int, num_total: int) -> int:
     return min(num_total, num_match) + 1
 
 def min_possible_matches(num_not_match: int, num_total: int) -> int:
-    """Minimum number of additional possible matches."""
+    """Minimum number of additional possible matches.
+
+    TODO: I think there's an error here.
+    """
     return min(num_total, max(0, num_total - num_not_match))
 
 
-def compute_possible_observed_prevalences(
+def possible_prevalence_combinations(
     data: pd.DataFrame, scenario: Scenario, modality: str = "max_llh"
-) -> np.ndarray:
-    """Extract possible combinations of prevalences."""
-    # when looking at the data, we always consider both sides
+) -> pd.DataFrame:
+    """Extract possible combinations of prevalences.
+
+    Essentially, this computes the prevalence of a scenario in all the data where
+    midline extension is known. To this is adds the `possible_combinations` of the
+    prevalence in the data where midline extension is unknown.
+    """
     base_num_match, base_num_total = compute_observed_prevalence(data, scenario)
 
     has_t_stage = data.ly.t_stage.isin(scenario.t_stages)
@@ -47,88 +55,135 @@ def compute_possible_observed_prevalences(
     unknown_match = does_pattern_match.sum()
     unknown_total = len(unknown_midext_data)
 
-    return compute_match_total_combinations(
-        base_match=base_num_match,
-        base_total=base_num_total,
-        unknown_match=unknown_match,
-        unknown_total=unknown_total,
-    )
+    combs = possible_combinations(unknown_match, unknown_total)
+    combs["matches"] += base_num_match
+    combs["total"] += base_num_total
+
+    return combs
 
 
-def compute_match_total_combinations(
-    base_total: int,
-    base_match: int,
-    unknown_match: int,
-    unknown_total: int,
-) -> np.ndarray:
-    """Compute all possible combinations of match/total pairs."""
-    results = []
-    for pot_num_total in range(unknown_total + 1):
-        for pot_num_match in range(
[…]

### 2024-06-05 model.qmd
switch to marginals over unknown combos

@@
 from scripts import paths, shared
-from scripts.marginalize_midext import compute_possible_observed_prevalences, SummedBetaPosterior
+from scripts.marginalize_midext import possible_prevalence_combinations, SummedBetaPosterior
 
@@
       )
-      content.append(
-        BetaPosterior.from_hdf5(
-          filename=paths.model_dir / "full" / "prevalences_overall.hdf5",
-          dataname=scenario.md5_hash("prevalences"),
-          color=color,
-        )
-      )
-      num_success, num_total = compute_possible_observed_prevalences(data, scenario).T
+      # content.append(
+      #   BetaPosterior.from_hdf5(
+      #     filename=paths.model_dir / "full" / "prevalences_overall.hdf5",
+      #     dataname=scenario.md5_hash("prevalences"),
+      #     color=color,
+      #   )
+      # )
+      combs = possible_prevalence_combinations(data, scenario)
       content.append(
         SummedBetaPosterior(
-          num_success=num_success,
-          num_total=num_total,
-          kwargs={"label": None, "linestyle": "--"},
+          num_success=combs.matches.values,
+          num_total=combs.total.values,
+          permutations=combs.permutations.values,
+          kwargs={"color": color},
         )
@@
   titles=["LNL II", "LNL III", "LNL IV"],
-  locs=[(0.57, 0.55), (0.48, 0.99), (0.15, 0.99)],
+  locs=[(0.53, 0.6), (0.55, 0.99), (0.12, 0.99)],
 )
@@
   titles=["LNL II", "LNL III", "LNL IV"],
-  locs=[(0.67, 0.99), (0.27, 0.6), (0.1, 0.99)],
+  locs=[(0.55, 0.99), (0.27, 0.6), (0.1, 0.99)],
 )
+axes[1,0].set_ylim(top=1.4)
 split_legends(
@@
   titles=["LNL II", "LNL III", "LNL IV"],
-  locs=[(0.64, 0.99), (0.27, 0.6), (0.24, 0.99)],
+  locs=[(0.54, 0.99), (0.24, 0.6), (0.11, 0.99)],
 )
@@
   titles=["LNL II", "LNL III", "LNL IV"],
-  locs=[(0.64, 0.99), (0.4, 0.6), (0.16, 0.99)],
+  locs=[(0.54, 0.99), (0.4, 0.6), (0.12, 0.99)],
 )
@@
 from scripts.marginalize_midext import (
-  compute_possible_observed_prevalences,
+  possible_prevalence_combinations,
   SummedBetaPosterior,
@@
       )
-      content.append(
-        BetaPosterior.from_hdf5(
-          filename=paths.model_dir / "full" / "prevalences_with_ipsi.hdf5",
-          dataname=scenario.md5_hash("prevalences"),
-          color=color,
-        )
-      )
-      num_success, num_total = compute_possible_observed_prevalences(data, scenario).T
+      # content.append(
+      #   BetaPosterior.from_hdf5(
+      #     filename=paths.model_dir / "full" / "prevalences_with_ipsi.hdf5",
+      #     dataname=scenario.md5_hash("prevalences"),
+      #     color=color,
+      #   )
+      # )
+      combs = possible_prevalence_combinations(data, scenario)
       content.append(
         SummedBetaPosterior(
-          num_success=num_success,
-          num_total=num_total,
-          kwargs={"linestyle": "--", "label": None},
+          num_success=combs.matches.values,
+          num_total=combs.total.values,
+          permutations=combs.permutations.values,
+          kwargs={"color": color},
         )
@@
   titles=["ipsi N0", "ipsi LNL III"],
-  locs=[(0.3, 0.99), (0.45, 0.55)],
+  locs=[(0.3, 0.99), (0.5, 0.55)],
 )
@@
   titles=["ipsi N0", "ipsi LNL III"],
-  locs=[(0.2, 0.99), (0.55, 0.99)],
+  locs=[(0.16, 0.99), (0.53, 0.55)],
 )
@@
   titles=["ipsi N0", "ipsi LNL III"],
-  locs=[(0.2, 0.99), (0.65, 0.99)],
+  locs=[(0.13, 0.55), (0.5, 0.99)],
 )
@@
   )
-  beta_post = BetaPosterior.from_hdf5(
+  content.append(
+    BetaPosterior.from_hdf5(
       filename=paths.model_dir / "full" / "prevalences_midext.hdf5",
@@
     )
-  content.append(beta_post)
-  combs = possible_combinations(
-    matches=unknowns,
-    max_total=unknowns,
   )
-  content.append(
-    SummedBetaPosterior(
-      num_success=combs.matches.values + beta_post.num_success,
-      num_total=combs.total.values + beta_post.num_total,
-      permutations=combs.permutations.values,
-      kwargs={"linestyle": "--", "label": None, "color": "red"},
-    )
-  )
-  print(f"{unknowns = }, {beta_post.num_success = }, {beta_post.num_total = }")
 

### 2024-06-05 scripts/marginalize_midext.py

@@ -4,7 +4,7 @@ Marginalize over *observd* midline extension status using sums of beta posterior
 
 import math
 from dataclasses import dataclass
-from typing import Any, TypeVar
+from typing import Any, Generator, TypeVar
 import matplotlib.pyplot as plt
 from matplotlib.axes import Axes as MPLAxes
 import pandas as pd
@@ -27,10 +27,7 @@ def max_possible_matches(num_match: int, num_total: int) -> int:
     return min(num_total, num_match) + 1
 
 def min_possible_matches(num_not_match: int, num_total: int) -> int:
-    """Minimum number of additional possible matches.
-
-    TODO: I think there's an error here.
-    """
+    """Minimum number of additional possible matches."""
     return min(num_total, max(0, num_total - num_not_match))
 
 
@@ -40,7 +37,7 @@ def possible_prevalence_combinations(
     """Extract possible combinations of prevalences.
 
     Essentially, this computes the prevalence of a scenario in all the data where
-    midline extension is known. To this is adds the `possible_combinations` of the
+    midline extension is known. To this is adds the `possible_combinations()` of the
     prevalence in the data where midline extension is unknown.
     """
     base_num_match, base_num_total = compute_observed_prevalence(data, scenario)
@@ -62,6 +59,19 @@ def possible_prevalence_combinations(
     return combs
 
 
+def _possible_combinations(matches: int, max_total: int) -> Generator[dict, None, None]:
+    for possible_total in range(max_total + 1):
+        for possible_matches in range(
+            min_possible_matches(max_total - matches, possible_total),
+            max_possible_matches(matches, possible_total),
+        ):
+            yield {
+                "matches": possible_matches,
+                "total": possible_total,
+                "permutations": math.comb(possible_total, possible_matches),
+            }
+
+
 def possible_combinations(matches: int, max_total: int) -> pd.DataFrame:
     """Tabulate possible combinations of matches, totals, and num of permutations.
 
@@ -69,6 +79,13 @@ def possible_combinations(matches: int, max_total: int) -> pd.DataFrame:
     the respective combo has) of the number of matches in a possible total. This total
     may range from 0 to ``max_total``, while the number of matches may range from
[…]

### 2024-06-06 model.qmd
start with risk prediction figure(s)

@@
 ```
+
+# Risk Predictions
+
+Now, we want to make actual use of the model and predict the risk for occult diease given some diagnoses.
+
+```{python}
+import numpy as np
+import matplotlib.pyplot as plt
+from matplotlib import gridspec
+from lyscripts.scenario import Scenario
+from lyscripts.plot.utils import Histogram, draw, COLORS, split_legends
+from scripts import paths, shared
+
+def turn_axis_off(axes):
+  axes.set_xticks([])
+  axes.set_yticks([])
+
+all_false = {lnl: False for lnl in ["I", "II", "III", "IV", "V"]}
+
+nrows, ncols = 2, 2
+
+plt.rcParams.update(shared.get_fontsizes())
+plt.rcParams.update(shared.get_figsizes(
+  nrows=nrows,
+  ncols=ncols,
+  aspect_ratio=2.,
+  width=17,
+))
+
+fig = plt.figure()
+gs = gridspec.GridSpec(nrows=nrows, ncols=ncols, figure=fig)
+axes = np.array([[fig.add_subplot(gs[i,j]) for j in range(ncols)] for i in range(nrows)])
+full = fig.add_subplot(gs[:,:], frame_on=False)
+turn_axis_off(full)
+
+for i, t_stage in enumerate(["early", "late"]):
+  for j, midext in enumerate([False, True]):
+    contents = []
+    for diagnosis in [
+      {"III": True},
+      all_false,
+    ]:
+      scenario = Scenario(
+        t_stages=[t_stage],
+        midext=midext,
+        diagnosis={
+          "ipsi": {"CT": diagnosis},  # Somehow, this does not seem to have any effect
+          "contra": {"CT": all_false},
+        },
+        involvement={"contra": {"II": True}},
+      )
+      print(scenario.md5_hash("risks"))
+      contents.append(
+        Histogram.from_hdf5(
+          filename=paths.model_dir / "full" / "risks.hdf5",
+          dataname=scenario.md5_hash("risks"),
+        )
+      )
+    draw(axes[i,j], contents, xlims=(0., 50.))
+
+plt.savefig(paths.figure_dir / "fig-model-risks.png", dpi=400)
+plt.show()
+```

### 2024-06-07 manuscript.qmd
split ipsi scenarios by midext
risk with fancy legend

@@
 The figure shows that the model is capable of describing the correlation between ipsi- and contralateral involvement.
-Although for the scenario of LNL II involvement in both sides, the prediction's split between early and advanced T-category is not large enough.](figures/fig-model-prevalences-with-ipsi.png){#fig-model-prevalences-with-ipsi width="50%"}
+Although for the scenario of LNL II involvement in both sides, the prediction's split between early and advanced T-category is not large enough.](figures/fig-model-prevalences-with-ipsi.png){#fig-model-prevalences-with-ipsi}
 
@@
 
-- compute risks for occult disease, given diagnosis
+In this section, we briefly investigate how the model may be applied clinically: We want to estimate the risk for occult disease in some or all LNLs, given the patient's individual diagnosis.
+In terms of our model, this diagnosis consists of the T-category, the lateralization of the tumor (does it extend over the mid-sagittal line?), and which LNLs are clinically involved, e.g.
+because some lymph nodes appear enlarged in an MRI scan or show increased glycolytic activity on an FDG PET/CT scan.
+
+![Histograms over the predicted risk of contralateral LNL II invovlement, shown for all eight possible combinations of the risk factors T-category (early: top row, advanced: bottom row), midline extension (absent: left column, present: right column), and ipsilateral diagnosis (ipsilateral N0: blue, ipsilateral at least LNL III diagnosed: orange).
+The provided diagnosis always included a clinically node negative contralateral neck.
+Each of the three risk factors significantly increase the predicted risk for occult disease](figures/fig-model-risks.png){#fig-model-risks}
 

### 2024-06-07 model.qmd
risk with fancy legend

--- 2024-06-05 model.qmd
@@
 ```
+
+# Risk Predictions
+
+Now, we want to make actual use of the model and predict the risk for occult diease given some diagnoses.
+
+```{python}
+import numpy as np
+import matplotlib.pyplot as plt
+from matplotlib.legend_handler import HandlerTuple
+from lyscripts.scenario import Scenario
+from lyscripts.plot.utils import Histogram, draw, COLORS, split_legends
+from scripts import paths, shared
+
+all_false = {lnl: False for lnl in ["I", "II", "III", "IV", "V"]}
+
+plt.rcParams.update(shared.get_fontsizes())
+plt.rcParams.update(shared.get_figsizes(width=17./2., aspect_ratio=2.))
+
+fig, ax = plt.subplots()
+ax.set_xlim(0., 15.)
+
+contents = []
+defaults = {
+  "density": True,
+  "histtype": "stepfilled",
+  "alpha": 0.7,
+  "bins": 50,
+}
+patches = {
+  "early T-cat.": [],
+  "advanced": [],
+  "lateralized": [],
+  "midline ext.": [],
+  "ipsi N0": [],
+  "≥ ipsi LNL III": [],
+}
+
+scenario = Scenario(
+  t_stages=["early"],
+  midext=False,
+  diagnosis={
+    "ipsi": {"CT": all_false},
+    "contra": {"CT": all_false},
+  },
+  involvement={"contra": {"II": True}},
+)
+_v, bins, patch = Histogram.from_hdf5(
+  filename=paths.model_dir / "full" / "risks.hdf5",
+  dataname=scenario.md5_hash("risks"),
+).draw(ax, hist=defaults)
+
+patches["early T-cat."].append(*patch)
+patches["lateralized"].append(*patch)
+patches["ipsi N0"].append(*patch)
+
+scenario = Scenario(
+  t_stages=["late"],
+  midext=False,
+  diagnosis={
+    "ipsi": {"CT": {"III": True}},
+    "contra": {"CT": all_false},
+  },
+  involvement={"contra": {"II": True}},
+)
+_v, bins, patch = Histogram.from_hdf5(
+  filename=paths.model_dir / "full" / "risks.hdf5",
+  dataname=scenario.md5_hash("risks"),
+).draw(ax, hist=defaults)
+
+patches["advanced"].append(*patch)
+patches["lateralized"].append(*patch)
+patches["≥ ipsi LNL III"].append(*patch)
+
+scenario = Scenario(
+  t_stages=["early"],
+  midext=True,
+  diagnosis={
+    "ipsi": {"CT": all_false},
+    "contra": {"CT": all_false},
+  },
+  involvement={"contra": {"II": True}},
+)
+_v, bins, patch = Histogram.from_hdf5(
+  filename=paths.model_dir / "full" / "risks.hdf5",
+  dataname=scenario.md5_hash("risks"),
+).draw(ax, hist=defaults)
+
+patches["early T-cat."].append(*patch)
+patches["midline ext."].append(*patch)
+patches["ipsi N0"].append(*patch)
+
+scenario = Scenario(
+  t_stages=["late"],
+  midext=True,
+  diagnosis={
+    "ipsi": {"CT": {"III": True}},
+    "contra": {"CT": all_false},
+  },
+  involvement={"contra": {"II": True}},
+)
+_v, bins, patch = Histogram.from_hdf5(
+  filename=paths.model_dir / "full" / "risks.hdf5",
+  dataname=scenario.md5_hash("risks"),
+).draw(ax, hist=defaults)
+
+patches["advanced"].append(*patch)
+patches["midline ext."].append(*patch)
+patches["≥ ipsi LNL III"].append(*patch)
+
+for label in patches:
+  handle = tuple(patches[label])
+  patches[label] = handle
+
+ax.legend(
+  handles=patches.values(),
+  labels=patches.keys(),
+  handler_map={tuple: HandlerTuple(ndivide=None, pad=0.2)},
+  handlelength=2.2,
+  handletextpad=0.4,
+  ncols=3,
+  labelspacing=0.3,
+  columnspacing=1.,
+)
+
+plt.savefig(paths.figure_dir / "fig-model-risks.png", dpi=400)
+plt.show()
+```

### 2024-06-07 manuscript.qmd
simpler legend in risk prediction

@@
 
-![Histograms over the predicted risk of contralateral LNL II invovlement, shown for all eight possible combinations of the risk factors T-category (early: top row, advanced: bottom row), midline extension (absent: left column, present: right column), and ipsilateral diagnosis (ipsilateral N0: blue, ipsilateral at least LNL III diagnosed: orange).
-The provided diagnosis always included a clinically node negative contralateral neck.
-Each of the three risk factors significantly increase the predicted risk for occult disease](figures/fig-model-risks.png){#fig-model-risks}
+![Histograms over the predicted risk of occult contralateral LNL II invovlement, shown for some combinations of T-category, tumor lateralization, and ipsilateral clinical involvement.
+The contralateral side was always assumed to be clinically negative.](figures/fig-model-risks.png){#fig-model-risks width=50%}
 

### 2024-06-07 model.qmd
simpler legend in risk prediction

--- 2024-06-05 model.qmd
@@
 ```
+
+# Risk Predictions
+
+Now, we want to make actual use of the model and predict the risk for occult diease given some diagnoses.
+
+```{python}
+import numpy as np
+import matplotlib.pyplot as plt
+from matplotlib.legend_handler import HandlerTuple
+from lyscripts.scenario import Scenario
+from lyscripts.plot.utils import Histogram, draw, COLORS, split_legends
+from scripts import paths, shared
+
+all_false = {lnl: False for lnl in ["I", "II", "III", "IV", "V"]}
+
+plt.rcParams.update(shared.get_fontsizes())
+plt.rcParams.update(shared.get_figsizes(width=17./2., aspect_ratio=2.))
+
+fig, ax = plt.subplots()
+ax.set_xlim(0., 15.)
+
+contents = []
+defaults = {
+  "density": True,
+  "histtype": "stepfilled",
+  "alpha": 0.7,
+  "bins": 50,
+}
+
+scenario = Scenario(
+  t_stages=["early"],
+  midext=False,
+  diagnosis={
+    "ipsi": {"CT": all_false},
+    "contra": {"CT": all_false},
+  },
+  involvement={"contra": {"II": True}},
+)
+contents.append(Histogram.from_hdf5(
+  filename=paths.model_dir / "full" / "risks.hdf5",
+  dataname=scenario.md5_hash("risks"),
+  color=COLORS["green"],
+  label="early, lateralized; N0",
+))
+
+scenario = Scenario(
+  t_stages=["late"],
+  midext=False,
+  diagnosis={
+    "ipsi": {"CT": {"III": True}},
+    "contra": {"CT": all_false},
+  },
+  involvement={"contra": {"II": True}},
+)
+contents.append(Histogram.from_hdf5(
+  filename=paths.model_dir / "full" / "risks.hdf5",
+  dataname=scenario.md5_hash("risks"),
+  color=COLORS["blue"],
+  label="advanced, lateralized; ipsi LNL III",
+))
+
+scenario = Scenario(
+  t_stages=["early"],
+  midext=True,
+  diagnosis={
+    "ipsi": {"CT": all_false},
+    "contra": {"CT": all_false},
+  },
+  involvement={"contra": {"II": True}},
+)
+contents.append(Histogram.from_hdf5(
+  filename=paths.model_dir / "full" / "risks.hdf5",
+  dataname=scenario.md5_hash("risks"),
+  color=COLORS["orange"],
+  label="early, midline extension; N0",
+))
+
+scenario = Scenario(
+  t_stages=["late"],
+  midext=True,
+  diagnosis={
+    "ipsi": {"CT": {"III": True}},
+    "contra": {"CT": all_false},
+  },
+  involvement={"contra": {"II": True}},
+)
+contents.append(Histogram.from_hdf5(
+  filename=paths.model_dir / "full" / "risks.hdf5",
+  dataname=scenario.md5_hash("risks"),
+  color=COLORS["red"],
+  label="advanced, midline extension; ipsi LNL III",
+))
+
+draw(ax, contents, xlims=(0., 15.))
+ax.legend(labelspacing=0.3)
+ax.set_xlabel("risk [%]")
+ax.set_yticks([])
+
+plt.savefig(paths.figure_dir / "fig-model-risks.png", dpi=400)
+plt.show()
+```

### 2024-06-10 manuscript.qmd
expand results around risk predictions

@@
 The orange arcs represent the sensitivity and specificity of the observational modality (e.g.
-CT, MRI, pathology, ...).](static/small-graph.png){#fig-small-graph width=50%}
+CT, MRI, pathology, ...).](static/small-graph.png){#fig-small-graph width=40%}
 
@@
 T-category and midline extension, into account.
-As observed in the data, the model predicts that the prevalence of contralateral LNL II involvement jumps from below 8% for early T-category lateralized tumors to almost 40% when the tumor is of advanced T-category and crosses the mid-sagittal line.
+As observed in the data, the model predicts that the prevalence of contralateral LNL II involvement jumps from below 8% <!-- TODO: make this dynamic --> for early T-category lateralized tumors to almost 40% <!-- TODO: make this dynamic --> when the tumor is of advanced T-category and crosses the mid-sagittal line.
 
-However, for early T-category scenarios with midline extension, the model does seem to overestimate contralateral LNL II invovlement and underestimate the prevalence for metastases in LNL III.
-This likely stems from the the small sample size of this rare scenario as hinted at by the wide Beta posteriors.
+However, for early T-category scenarios with midline extension, the model does seem to overestimate contralateral LNL II and III invovlement.
+This likely stems from the the small sample size of this relatively rare scenario as hinted at by the wide beta posteriors.
 
@@
 
-In this section, we briefly investigate how the model may be applied clinically: We want to estimate the risk for occult disease in some or all LNLs, given the patient's individual diagnosis.
+In this section, we investigate how the model may be applied clinically: We want to estimate the risk for occult disease in some or all LNLs, given the patient's individual diagnosis.
 In terms of our model, this diagnosis consists of the T-category, the lateralization of the tumor (does it extend over the mid-sagittal line?), and which LNLs are clinically involved, e.g.
@@
 
-## Distributions over States
+@Fig-model-risks shows this predicted risk of occult disease in contralateral LNL II for four interesting combinations of these three risk factors.
+This risk is computed *given* a CT diagnosis assessing the per-LNL clinical involvement for which we assumed a sensitivity of 81% and a specificity of 76% [@debondt_detection_2007].
 
-```{python}
-plt.rcdefaults()
+The variable impacting the prediction for contralateral involvement in our model is the tumor's lateralization.
+For example, a patient with a clearly lateralized early T-category tumor and a clinically N0 neck is assigned a 1-2% <!-- TODO: make this dynamic --> risk for occult disease in contralateral LNL II.
+Under the same scenario, but *with* mid-sagittal extension, the risk jumps to almost 7%.
+<!-- TODO: make this dynamic -->
 
-mi = pd.MultiIndex.from_product(
-  [[False, True]] * len(simple_model.ext.ipsi.graph.lnls),
-  names=simple_model.ext.ipsi.graph.lnls.keys(),
-)
-series = pd.Series(simple_model.ext.ipsi.state_dist(), index=mi)
-upsetplot.plot(
-  series,
-  sort_by="cardinality",
-  sort_categories_by="-input",
-)
-plt.show()
-```
+T-category plays a lesser role: Considering the scenario of a tumor that crosses the mid-sagittal line and an ipsilateral neck where at least LNL III is clinically involved, the risk for occult contralateral LNL II disease is around 8.5%.
+<!-- TODO: make this dynamic --> For the same scenario and an advanced T-category tumor, the risk increases to 11%.
+<!-- TODO: make this dynamic -->
+
+Lastly, the predicted risk is also correlated via the time-steps to the degree of ipsilateral involvement.
+Changing the aforementioned scenario (advanced T-category, midline extension, ipsilateral LNL III clinically involved) to one where the patient presents with a clinically N0 neck, the risk for occult disease in the contralateral LNL II falls from 11% <!-- TODO: make this dynamic --> to 9.5%.
+<!-- TODO: make this dynamic -->
+
+Taken together, T-category and ipsilateral involvement may still considerably impact the risk prediction for contralateral involvement: In @fig-model-risks the scenarios underlying the orange (6.5% <!-- TODO: make this dynamic -->) and the red (11% <!-- TODO: make this dynamic -->) histograms differ in T-category (early vs advanced) and clinical diagnosis (N0 vs ipsilateral at least LNL III involved).
 
@@
 
-# Appendix {.appendix}
+# Contralateral Prevalence of Involvement {.appendix}
 
@@
 ```
+
+# Marginalizing Beta Posteriors over Unknowns {.appendix}
+
+- [ ] write short theory section about why and how to compute the posteriors if some of the data (here midline extension) is unknown

### 2024-06-10 model.qmd
expand results around risk predictions

@@
 plt.rcParams.update(shared.get_fontsizes())
-plt.rcParams.update(shared.get_figsizes(width=17./2., aspect_ratio=2.))
+plt.rcParams.update(shared.get_figsizes(width=17./2.))
 
@@
 ax.set_yticks([])
+ax.set_title("Predicted Risk of Contralateral LNL II Involvement\nfor Selected Diagnosis Scenarios")
 

### 2024-06-11 manuscript.qmd
add appendix about marginal beta posteriors

@@
 
-## Computing the Observed and Predicted Prevalence of Involvement Patterns
+## Computing the Observed and Predicted Prevalence of Involvement Patterns {#sec-prevalence}
 
@@
 For example, we may be interested in the prevalence of contralateral LNL II involvement (i.e., contra LNL II "involved" and all other LNLs "masked") under the scenario of early T-category (T0-T2) and no midline extension ($\epsilon=\texttt{False}$).
-To compute this prevalence in the data, we select all patients of this scenario (in our data, this amounts to <!-- TODO: dyamically compute this --> patients).
-Of those, <!-- TODO: dynamically compute this --> were found to harbor metastases in their contralateral LNL II.
-Therefore, the prevalence is <!-- TODO: dynamically compute this -->.
+To compute this prevalence in the data, we select all patients of this scenario (in our data, this amounts to 379<!-- TODO: dyamically compute this --> patients).
+Of those, 28<!-- TODO: dynamically compute this --> were found to harbor metastases in their contralateral LNL II.
+Therefore, the prevalence is 7%<!-- TODO: dynamically compute this -->.
 
-When displaying this data prevalence, we often choose to draw a *Beta posterior* over the "true" prevalence, hinting at the fact that our data merely represents a limited sample.
-The Beta posterior follows from a uniform Beta distribution as prior and a Bernoulli likelihood for the number of patients with the involvement of interest, given the parameer for the "true" prevalence.
+When displaying this data prevalence, we often choose to draw a *beta posterior* over the "true" prevalence, hinting at the fact that our data merely represents a limited sample.
+The beta posterior follows from a uniform beta distribution as prior and a binomial likelihood for the number of patients with the involvement of interest, given the parameer for the "true" prevalence.
 The resulting distribution has its maximum at the observed prevalence, but in addition gives a visual intuition for the variance of of the observed quantity.
-I.e., when we observe 3 out of 10 events, the Beta posterior is much wider than if we observe 300 out of 1000 for he same prevalence.
-It also allows us to check not inly if the model is accurate, but also whether it reflects the uncertainty contained in the data.
+I.e., when we observe 3 out of 10 events, the beta posterior is much wider than if we observe 300 out of 1000 for he same prevalence.
+It also allows us to check not only if the model is accurate, but also whether it reflects the uncertainty contained in the data.
 
@@
 
-- [ ] write short theory section about why and how to compute the posteriors if some of the data (here midline extension) is unknown
+```{python}
+#| echo: false
+#| label: tbl-data-unknowns
+#| tbl-cap: Combinations of midline extension status and state of contralateral LNL II for early T-category patients.
+import pandas as pd
+import numpy as np
+from scripts import shared, COL
+
+data = shared.get_data()
+is_early = data[COL.t_stage].isin([0,1,2])
+
+reduced = pd.DataFrame({
+  "Midline Extension": data.loc[is_early, COL.midext],
+  "contra LNL II": data.loc[is_early, ("max_llh", "contra", "II")]
+})
+
+result = (
+  reduced
+  .fillna("unknown")
+  .value_counts(dropna=False)
+  .unstack("Midline Extension")
+  .rename(
+    index={
+      False: "healthy",
+      True: "involved",
+    },
+    columns={
+      False: "no",
+      True: "yes",
+    },
+  )
+)
+result.loc["total"] = result.sum()
+result
+```
+
+In @sec-prevalence we explained that we compare the predicted prevalence of an involvement pattern under a scenario to a beta posterior over the "true" prevalence given the observations in the data.
+However, our data contains incomplete observations, especially regarding the tumor's midline extension.
+This information is not present for a large group of 85 patients, most of which treated at the Centre Léon Bérard in France.
+
+To account for this missing information and plot the correct posteriors over the data prevalence, we need to marginalize over the possible combinations of matching involvement patterns and selected patients under a scenario.
+For example, we have only 29<!-- TODO: compute dynamically --> early T-category patients with midline extension.
+But for an additional 65<!-- TODO: compute dynamically --> patients we do not know whether the tumor crossed the mid-sagittal plane.
+If we now want to compute the prevalence of contralateral LNL II invovlement for these early T-category patients with midline extension, then we need to account for the possibility that we are missing information.
+
+In @tbl-data-unknowns we show the combinations of the midline extension status and state of the contralateral LNL II.
+Using only available information, we would get to 6 out of 29<!-- TODO: compute dynamically --> patients for the prevalence of contralateral LNL II invovlement among the early T-category patients with midline extension.
+But the column with the "unknown" numbers suggests that the true prevalence could range from 6 out of 29 + 59 (7%)<!-- TODO: compute dynamically --> up to 12 out of 29 + 6 (34%)<!-- TODO: compute dynamically -->.
+Assuming that every one of the possible combinations has the same a priori probability (uniform prior) and accouting for permutations via the binomial factor, the posterior marginalized over the unknowns could be written as
+
+$$
+\begin{aligned}
+p(x; \alpha=1, \beta=1) &= \sum_{m=0}^{65} \sum_{\ell=0}^{\min(m,6)} \binom{n+m}{k+\ell} x^{k+\ell} (1-x)^{n+m - (k+\ell)} \cdot \frac{1}{\operatorname{B}(\alpha, \beta)} x^\alpha (1-x)^\beta \\
+&= \sum_{m=0}^{65} \sum_{\ell=0}^{\min(m,6)} \binom{n+m}{k+\ell} x^{k+\ell+1} (1-x)^{n+m - (k+\ell) + 1}
+\end{aligned}
+$$
+
+We have plotted this marginal in @fig-model-marginal-pdf (blue curve) and the terms that make up the above sum (light red curves).
+Note that we weighted the plotting opacity with the number of permutations each of the possible combinations has.
+The marginal is compared against the "naive" beta posterior that only takes the data into account for which midline extension is available (green curve).
+
+![Posterior distributions over the data prevalence for contralateral LNL II involvement under the scenario of early T-category and midline extension.
+The green beta distribution shows this prevalence for those patients for which midline extension is known.
+While the blue PDF is a marginal distribution summed over all possible combinations, which are shown in shades of red.
+The more permutations a particular combination has, the more opaque it is plotted.](figures/fig-model-marginal-pdf.png){#fig-model-marginal-pdf width="50%"}

### 2024-06-11 model.qmd
add appendix about marginal beta posteriors

@@
 ```
+
+# Illustrating the Marginalization over Beta PDFs
+
+```{python}
+import matplotlib.pyplot as plt
+from lyscripts.scenario import Scenario
+from lyscripts.plot.utils import draw, COLORS, BetaPosterior
+from lyscripts.compute.prevalences import compute_observed_prevalence
+from scripts import paths, shared, COL
+from scripts.marginalize_midext import (
+  possible_prevalence_combinations,
+  SummedBetaPosterior,
+)
+
+scenario = Scenario(
+  t_stages=["early"],
+  midext=True,
+  diagnosis={"contra": {"max_llh": {"II": True}}},
+)
+combs = possible_prevalence_combinations(data, scenario)
+summed_beta = SummedBetaPosterior(
+  num_success=combs.matches.values,
+  num_total=combs.total.values,
+  permutations=combs.permutations.values,
+  kwargs={"label": "marginalized PDF", "color": COLORS["blue"]},
+)
+num_match, num_total = compute_observed_prevalence(data, scenario, mapping=None)
+naive_beta = BetaPosterior(
+  num_success=num_match,
+  num_total=num_total,
+  kwargs={"color": COLORS["green"], "label": "naive beta PDF"},
+)
+
+plt.rcParams.update(shared.get_fontsizes())
+plt.rcParams.update(shared.get_figsizes(width=17./2.))
+fig, ax = plt.subplots()
+draw(ax, contents=[naive_beta, summed_beta], xlims=(0., 50.))
+
+x = np.linspace(*ax.get_xlim(), 300)
+perm_sum = combs.permutations.sum()
+for _, row in combs.sort_values("permutations").iterrows():
+  beta = BetaPosterior(
+    num_success=row.matches,
+    num_total=row.total,
+  )
+  h, = ax.plot(
+    x, beta.pdf(x),
+    alpha=0.6 * (row.permutations / perm_sum)**0.5,
+    color=COLORS["red"],
+    zorder=1,
+  )
+
+ax.set_ylim(bottom=0.)
+handles, labels = ax.get_legend_handles_labels()
+ax.legend(handles=[*handles, h], labels=[*labels, "component beta PDFs"])
+ax.set_xlabel("prevalence [%]")
+ax.set_yticks([])
+
+plt.savefig(paths.figure_dir / "fig-model-marginal-pdf.png", dpi=400)
+plt.show()
+```

### 2024-06-11 manuscript.qmd
expand appendix by two paragraphs

@@
 The more permutations a particular combination has, the more opaque it is plotted.](figures/fig-model-marginal-pdf.png){#fig-model-marginal-pdf width="50%"}
+
+The distribution that marginalizes over the unknown midline extension cases is shifted towards a lower prevalence.
+This makes intuitive sense: With only six patients among the 65 with unknown midline extension status, the upper bound for the prevalence has moved down compared to six out of 29.
+
+We chose to compare our model against such corrected prevalence posteriors, because the model, too, is capable of marginalizing over unknown midline extension in the data.
+It would therefore be inaccurate to compare the model trained with the full dataset to only that part of the patient cohort where the tumor lateralization was recorded.

### 2024-06-12 manuscript.qmd
flesh out introduction

@@
 
-- head and neck cancer spreads through the lymphatic network
-- may sometimes spread to contralateral side
-- spreads more frequently and to larger extent contralaterally when tumor extends the mid-sagittal line
-- we describe a model based on previously published hidden Markov model
-- we extend it to cover the contralateral side, too
-- naive approach: make two independent models for each side, but that is not supported by the data
-- ipsi- and contralateral side are correlated via time of diagnosis, which is correlated with T-category
-- tumor extension over mid-sagittal line is modelled as random variable
-- spread probabilities from tumor to contralateral LNLs in case of midline extesnion are linear combinations of these probabilities in case of ipsilateral spread and contralateral spread when tumor is clearly lateralized
+When treating head and neck squamous cell carcinomas (HNSCC) with radiotherapy or surgery, not only the primary tumor and the clinically detected lymph node metastases are targeted.
+Per current guidelines, large volumes of the neck - lymph node levels (LNLs) - are part of the the elective clinical target volume (CTV-N) [@gregoire_ctbased_2003;@gregoire_delineation_2014;@gregoire_delineation_2018;@eisbruch_intensitymodulated_2002;@biau_selection_2019;@chao_determination_2002;@vorwerk_guidelines_2011;@ferlito_elective_2009].
+This is to minimize the risk of regional recurrences due to missed microscopic disease which cannot be detected by current in-vivo imaging modalities such as computed tomography (CT), magnetic resonance imaging (MRI), or positron emission tomography (PET).
 
-# Data on Lymphatic Progression Patterns
+Minimizing this risk of missing occult disease in the lymph drainage region must be balanced against the toxicity related to unnecessary treatment of healthy tissue.
 
-We have collected a detailed dataset of <!-- num patients --> patients with newly diagnosed oropharyngeal squamous cell carcinomas.
+The aforementioned guidelines for the definition of the CTV-N quantify this risk via the overall prevalence of lymph node metastases in the LNLs.
+Based on this, they recommend extensive irradiation of both sides of the neck for a majority of patients.
+However, this prevalence the guidelines are based on does not take into account the individual state of tumor progression, which may have a great impact on a patient's personal risk for occult disease in a given LNL.
+For example, a patient with substantial nodal involvement in the ipsilateral side and an advanced primary tumor that crosses the mid-sagittal plane would receive extensive elective irradiation of the contralateral LNLs II, III, and IVa.
+However, if the patient presented with no clinically detectable nodal disease (cN0) and only a small, clearly lateralized T1 tumor, the same contralateral CTV-N would be recommended [@biau_selection_2019].
+
+To better quantify this personalized risk we have previously developed an intuitive, probabilistic hidden Markov model [@ludwig_hidden_2021;@ludwig_modelling_2023] which was originally based on a conceptually similar Bayesian network model [@pouymayou_bayesian_2019].
+So far, these models were limited to the ipsilateral side of the neck.
+The main focus of this work will be to extend the formalism of these previous works to include risk predictions for the contralateral side as well.
+We believe this holds the potential to greatly reduce the electively irradiated volumes in the contralateral neck for patients whose individual risk for occult disease in that side is low.
+In turn, this may substantially reduce the toxicity from irradiation and the associated decrease in quality of life.
+
+In the following, we begin by inspecting the dataset the model is built upon (@sec-data) and from it formulate some properties we require the bilateral model extension to have (@sec-requirements).
+
+After that, in @sec-unilateral, we briefly summarize the original model formalism before we extend it to the contralateral side starting in @sec-ext-to-contra.
+
+Next, we describe how the model training and the computational experiments were set up in @sec-methods and present the results in @sec-results, followed by a closing discussion (@sec-discussion).
+
+# Data on Lymphatic Progression Patterns {#sec-data}
+
+We have collected a detailed dataset of 833<!-- num patients --> patients with newly diagnosed oropharyngeal squamous cell carcinomas.
 It reports the involvement of every patient individually and per lymph node level in tabular form, in addition to other clinico-pathological information such as age, T-category, and HPV p16 status.
@@
 
-# Methods
+# Methods {#sec-methods}
 
@@
 
-# Results
+# Results {#sec-results}
 
@@
 
-# Discussion
+# Discussion {#sec-discussion}
 

### 2024-06-13 model.qmd
add missing imports

@@
 import matplotlib.pyplot as plt
+import numpy as np
 from lyscripts.scenario import Scenario
@@
 )
+
+data = shared.get_data()
 

### 2024-06-14 manuscript.qmd
mention llh as prior in appendix

@@
 But the column with the "unknown" numbers suggests that the true prevalence could range from 6 out of 29 + 59 (7%)<!-- TODO: compute dynamically --> up to 12 out of 29 + 6 (34%)<!-- TODO: compute dynamically -->.
-Assuming that every one of the possible combinations has the same a priori probability (uniform prior) and accouting for permutations via the binomial factor, the posterior marginalized over the unknowns could be written as
+We assume that every one of the possible combinations has the same a priori probability (uniform prior) but we do need to account for permutations via the binomial factor.
+In general, let us say we have $n$ patients with midline extension, of which $k$ harbor metastases in the contralateral LNL II.
+And we have $m$ patients with unknown midline extension, of which $\ell$ presented with diasease in contralateral LNL II, then the posterior distribution - marginalized over the unknown data - over the true prevalence $x \in \left[ 0, 1 \right]$ is given by:
 
 $$
-\begin{aligned}
-p(x; \alpha=1, \beta=1) &= \sum_{m=0}^{65} \sum_{\ell=0}^{\min(m,6)} \binom{n+m}{k+\ell} x^{k+\ell} (1-x)^{n+m - (k+\ell)} \cdot \frac{1}{\operatorname{B}(\alpha, \beta)} x^\alpha (1-x)^\beta \\
-&= \sum_{m=0}^{65} \sum_{\ell=0}^{\min(m,6)} \binom{n+m}{k+\ell} x^{k+\ell+1} (1-x)^{n+m - (k+\ell) + 1}
-\end{aligned}
+p(x \mid n, k, m, \ell, \alpha, \beta) = \left[ \frac{1}{m} \sum_{i=0}^m \frac{1}{\min(i,\ell)} \sum_{j=0}^{\min(i,\ell)} \frac{P(i, j \mid x )}{\binom{n}{k}} \right] \cdot P(n, k \mid x) \cdot \frac{1}{\operatorname{B}(\alpha, \beta)} x^\alpha (1-x)^\beta
+$$
+
+Note the two binomial likelihoods: one being marginalized over and one acting implicitly as a prior to the summed terms.
+This role of the likelihood over the data with known midline extension as a prior to the part where that data is missing corrects for the fact that there are highly unlikely combinations among those we marginalize over.
+E.g., expecting that exactly those 6 patients with contralateral LNL II involvemnt are also coincidentially the only ones of the 65 with a tumor crossing the mid-sagittal line is very unlikely.
+This is accounted for by the fact that this possibility has only one permutation and it is weakened by the likelihood over the data with known midline extension.
+With a uniform prior beta distribution ($\alpha=1$ and $\beta=1$), the above simplifies to:
+
+$$
+p(x \mid n, k, m, \ell) = \left[ \frac{1}{m} \sum_{i=0}^m \frac{1}{\min(i,\ell)} \sum_{j=0}^{\min(i,\ell)} \binom{n+i}{k+j} x^j (1 - x)^{i-j} \right] \cdot x^{k+1} (1 - x)^{n-k+1}
 $$
@@
 The distribution that marginalizes over the unknown midline extension cases is shifted towards a lower prevalence.
-This makes intuitive sense: With only six patients among the 65 with unknown midline extension status, the upper bound for the prevalence has moved down compared to six out of 29.
+This makes intuitive sense: With only 6 patients among the 65 with unknown midline extension status, the upper bound for the prevalence has moved down compared to 6 out of 29.
 

### 2024-06-17 manuscript.qmd
add some refs & write first abstract draft

@@
 title: "Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer"
+abstract: |
+  In current practice, large parts of the contralateral lymph system are included in the elective clinical target volume (CTV-N) of oropharyngeal squamous cell carcinoma (OPSCC) patients.
+Guidelines on how to define the CTV-N are mostly based on prevalences of nodal involvement in the different lymph node levels (LNLs) and do not consider the detailed, individual diagnosis of a patient.
+
+  In this work, we present a probabilistic model for bilateral lymphatic tumor progression in OPSCC.
+Its aim is to predict the personal risk for occult disease in any LNL, given the patient's clincal diagnosis, T-category, and lateralization of the primary tumor.
+The model represents each of the LNLs I, II, III, IV, V, and VII of both sides of the neck as a hidden binary random variable that are connected to the tumor and among each other via arcs that correspond to spread probabilities per time-step.
+These spread probability rates are learned via Markov chain Monte Carlo (MCMC) sampling from a dataset of 833 OPSCC patients.
+
+  The model is able to describe the data well and may in the future be used to only include LNLs in a CTV-N that show a risk above a specific threshold for that particular patient's diagnosis.
+E.g., it predicts the risk for occult disease in contralateral LNL II to be below 5% unless the tumor crosses the mid-sagittal line.
 ---
@@
 
-We have collected a detailed dataset of 833<!-- num patients --> patients with newly diagnosed oropharyngeal squamous cell carcinomas.
+Originally, the only available data of detailed - i.e.
+per patient and per LNL - lymphatic involvement was a reconstructed table from the findings in @sanguineti_defining_2009, which included only ipsilateral nodal involvement for the LNLs I to IV.
+
+To be able to create models for lymphatic tumor progression that include more LNLs and the contralateral side, we have collected a detailed dataset of 833<!-- num patients --> patients with newly diagnosed oropharyngeal squamous cell carcinomas.
 It reports the involvement of every patient individually and per lymph node level in tabular form, in addition to other clinico-pathological information such as age, T-category, and HPV p16 status.
@@
 However, we instead chose to combine them into a single "consensus" diagnosis before parameter inference.
-We opted for this because the literature values of sensitivity and specificity [@debondt_detection_2007] of imaging modalities like MRI and CT do not plausibly match some of our observations: In the USZ cohort, 78% of OPSCC patients where diagnosed with ipsilateral LNL II involvement via diagnostic imaging.
+We opted for this because the literature values of sensitivity and specificity [@debondt_detection_2007;@kyzas_18ffluorodeoxyglucose_2008] of imaging modalities like MRI and CT do not plausibly match some of our observations: In the USZ cohort, 78% of OPSCC patients where diagnosed with ipsilateral LNL II involvement via diagnostic imaging.
 This is virtually impossible with sensitivities around 80% and specificities lower than 100%.
@@
 Our pre-training consensus was formed by considering all reported diagnostic information for a particular patient and LNL.
-When conflicts arose, we computed the *most likely* true state of involvement using the literature sensitivity and specificity values [@debondt_detection_2007].
+When conflicts arose, we computed the *most likely* true state of involvement using the literature sensitivity and specificity values [@debondt_detection_2007;@kyzas_18ffluorodeoxyglucose_2008].
 
@@
 It implements efficient MCMC sampling algorithms that employ multiple parallel samplers for affine invariance and better performance on multi-core CPUs.
+The sample proposal algirothms used by us are based on differential evolution moves [@terbraak_differential_2008;@nelson_run_2013].
 The [`emcee`] library was provided with the likelihood implemented by our [`lymph-model`] Python package.

### 2024-06-18 manuscript.qmd
start fleshing out discussion

@@
 
+Also, the discrepancy in midline extension prevalence between early and advanced T-category is particularly pronounced in oropharyngeal SCC patients.
+For example, in oral cavity SCC, the midline extension only increases from 15.4% (20 out of 130) to 33.3% (13 out of 39).
+
 ## Prediction of Risk for Occult Disease
@@
 
-- conceptually extended model to cover both sides of neck
-- explicitly models midline extension (also allows for easy marginalization)
-- midline extension intuitively causes contralateral spread to become more similar to ipsilateral spread
+## Summary
+
+In this work we present a formalism to model the ipsi- and contralateral lymphatic involvement of oropharyngeal SCC patients.
+An ipsilateral model has been previously developed and published [@ludwig_hidden_2021;@ludwig_modelling_2023].
+Based on this, we introduce an extension that leaves the ipsilateral model untouched, but extends it to the contralateral side in an intuitive and comprehensible manner.
+
+The model's performance w.r.t.
+its ability to describe the data on contralateral nodal involvement was evaluated based on a dataset of 833<!-- TODO: make this dynamic --> patients from four institutions.
+To the best of our knowledge, this is the first time lymphatic tumor progression has been modelled in such detail and to such an extent.
+While there was work on regional tumor spread, these models were conceptually different, more limited in the LNLs for which they make predictions, and not trained with real patient data [@benson_markov_2006;@jung_development_2017].
+
+The model takes the clincal diagnosis of the lymph node levels, the primary tumor's T-category, as well as lateralization into account when predicting the personalized risk for occult disease in any LNL of interest.
+Owing to the relatively few parameters, the model is highly interpretable and every parameter can be intuitively explained, while still offering good accuracy in its predictions.
+
+## Implications for Contralateral Elective Nodal Treatment
+
+The predictions of this model are being used to inform a clinical trial on volume deescalation at the University Hospital Zurich [citation needed<!-- TODO: add -->].
+When accepting a 5% risk of occult disease in any given LNL, the model suggests the following contralateral elective irradiation (assuming the respective LNL appears clinically healthy):
+
+- LNL II when the tumor extends over the mid-sagittal line
+- LNL III when LNL II is clinically, the tumor is of advanced T-category, crosses the midline, and the ipsilateral neck shows involvement of at least LNL II
+- LNL IV only for severe ipsilateral involvement, advanced T-category with obvious midline extension and contralateral upstream involvement
 

### 2024-06-19 manuscript.qmd
fix references

@@
 When treating head and neck squamous cell carcinomas (HNSCC) with radiotherapy or surgery, not only the primary tumor and the clinically detected lymph node metastases are targeted.
-Per current guidelines, large volumes of the neck - lymph node levels (LNLs) - are part of the the elective clinical target volume (CTV-N) [@gregoire_ctbased_2003;@gregoire_delineation_2014;@gregoire_delineation_2018;@eisbruch_intensitymodulated_2002;@biau_selection_2019;@chao_determination_2002;@vorwerk_guidelines_2011;@ferlito_elective_2009].
+Per current guidelines, large volumes of the neck - lymph node levels (LNLs) - are part of the the elective clinical target volume (CTV-N) [@gregoire_ct-based_2003;@gregoire_delineation_2014;@gregoire_delineation_2018;@eisbruch_intensity-modulated_2002;@biau_selection_2019;@chao_determination_2002;@vorwerk_guidelines_2011;@ferlito_elective_2009].
 This is to minimize the risk of regional recurrences due to missed microscopic disease which cannot be detected by current in-vivo imaging modalities such as computed tomography (CT), magnetic resonance imaging (MRI), or positron emission tomography (PET).
@@
 
-The entire data, including additional patients with tumors in other primary locations than the oropharynx, is publicly available: It may be [downloaded from LyProX](https://lyprox.org/patients/dataset) where it can be interactively explored too, [from GitHub](https://github.com/rmnldwg/lydata), [from zenodo](https://zenodo.org/search?q=lydata), or via the *Data-in-Brief* publications @ludwig_dataset_2022 and @ludwig_multicentric_2023.
+The entire data, including additional patients with tumors in other primary locations than the oropharynx, is publicly available: It may be [downloaded from LyProX](https://lyprox.org/patients/dataset) where it can be interactively explored too, [from GitHub](https://github.com/rmnldwg/lydata), [from zenodo](https://zenodo.org/search?q=lydata), or via the *Data-in-Brief* publications @ludwig_dataset_2022 and @ludwig_multi-centric_2023.
 Although these publications do not include the most recent dataset addition from Vall d'Hebron Barcelona Hospital.
@@
 However, we instead chose to combine them into a single "consensus" diagnosis before parameter inference.
-We opted for this because the literature values of sensitivity and specificity [@debondt_detection_2007;@kyzas_18ffluorodeoxyglucose_2008] of imaging modalities like MRI and CT do not plausibly match some of our observations: In the USZ cohort, 78% of OPSCC patients where diagnosed with ipsilateral LNL II involvement via diagnostic imaging.
+We opted for this because the literature values of sensitivity and specificity [@de_bondt_detection_2007;@kyzas_18f-fluorodeoxyglucose_2008] of imaging modalities like MRI and CT do not plausibly match some of our observations: In the USZ cohort, 78% of OPSCC patients where diagnosed with ipsilateral LNL II involvement via diagnostic imaging.
 This is virtually impossible with sensitivities around 80% and specificities lower than 100%.
@@
 Our pre-training consensus was formed by considering all reported diagnostic information for a particular patient and LNL.
-When conflicts arose, we computed the *most likely* true state of involvement using the literature sensitivity and specificity values [@debondt_detection_2007;@kyzas_18ffluorodeoxyglucose_2008].
+When conflicts arose, we computed the *most likely* true state of involvement using the literature sensitivity and specificity values [@de_bondt_detection_2007;@kyzas_18f-fluorodeoxyglucose_2008].
 
@@
 It implements efficient MCMC sampling algorithms that employ multiple parallel samplers for affine invariance and better performance on multi-core CPUs.
-The sample proposal algirothms used by us are based on differential evolution moves [@terbraak_differential_2008;@nelson_run_2013].
+The sample proposal algirothms used by us are based on differential evolution moves [@ter_braak_differential_2008;@nelson_run_2013].
 The [`emcee`] library was provided with the likelihood implemented by our [`lymph-model`] Python package.
@@
 @Fig-model-risks shows this predicted risk of occult disease in contralateral LNL II for four interesting combinations of these three risk factors.
-This risk is computed *given* a CT diagnosis assessing the per-LNL clinical involvement for which we assumed a sensitivity of 81% and a specificity of 76% [@debondt_detection_2007].
+This risk is computed *given* a CT diagnosis assessing the per-LNL clinical involvement for which we assumed a sensitivity of 81% and a specificity of 76% [@de_bondt_detection_2007].
 

### 2024-06-20 model.qmd
add scenarios for more risk computations

@@
 
+nrows, ncols = 1, 3
 plt.rcParams.update(shared.get_fontsizes())
-plt.rcParams.update(shared.get_figsizes(width=17./2.))
+plt.rcParams.update(shared.get_figsizes(
+  nrows=nrows,
+  ncols=ncols,
+  width=17,
+  aspect_ratio=1.3,
+))
 
-fig, ax = plt.subplots()
-ax.set_xlim(0., 15.)
+fig, ax = plt.subplots(nrows=nrows, ncols=ncols)
 
@@
 
+# contra LNL II
 scenario = Scenario(
@@
   color=COLORS["blue"],
-  label="advanced, lateralized; ipsi LNL III",
+  label="adv., lateralized; ipsi LNL III",
 ))
@@
   color=COLORS["orange"],
-  label="early, midline extension; N0",
+  label="early, mid. ext.; N0",
 ))
@@
   color=COLORS["red"],
-  label="advanced, midline extension; ipsi LNL III",
+  label="adv., mid. ext.; ipsi LNL III",
 ))
 
-draw(ax, contents, xlims=(0., 15.))
-ax.legend(labelspacing=0.3)
-ax.set_xlabel("risk [%]")
-ax.set_yticks([])
-ax.set_title("Predicted Risk of Contralateral LNL II Involvement\nfor Selected Diagnosis Scenarios")
+draw(ax[0], contents, xlims=(0., 15.))
+ax[0].legend(labelspacing=0.1)
+ax[0].set_xlabel("risk [%]")
+ax[0].set_yticks([])
+ax[0].set_xticks(np.linspace(0, 15, 6))
+ax[0].set_title("Contralateral LNL II")
+
+# contra LNL III
+contents = []
+
+ax[1].set_yticks([])
+ax[1].set_xticks(np.linspace(0, 15, 6))
+
+ax[2].set_yticks([])
+ax[2].set_xticks(np.linspace(0, 15, 6))
 

### 2024-06-21 model.qmd
rerun with all LNLs & use fixed lymph model
RelatedL #5

@@
     for ipsi, color in zip(
-      [{"I": False, "II": False, "III": False, "IV": False, "V": False}, {"III": True}],
+      [{"I": False, "II": False, "III": False, "IV": False, "V": False, "VII": False}, {"III": True}],
       [COLORS["blue"], COLORS["orange"]],
@@
 
-all_false = {lnl: False for lnl in ["I", "II", "III", "IV", "V"]}
+all_false = {lnl: False for lnl in ["I", "II", "III", "IV", "V", "VII"]}
 

### 2024-06-21 manuscript.qmd
change order of subsections
explain switching of trans mats

@@
 
-## Modelling Midline Extension {#sec-midline}
-
-To account for the increased prevalence of involvement on the contralateral side when the tumor is not clearly lateralized anymore, we also model the tumor's extension over the mid-sagittal line as a binary random variable.
-It starts lateralized and at every time-step there is a finite probability $p_\epsilon$ that the tumor grows over the symmetry plane of the patient.
-
-Technically, the introduction of this additional random variable doubles the space of the hidden states and therefore quadruples the size of the transition matrix $\mathbf{A}$.
-However, since we assume no correlation between the tumor's lateralization and the metastases in the LNLs, we can evolve the two parts separately.
-
-The probabilities to find a patient with a clearly lateralized tumor or one that extends over the mid-sagittal line after $t$ time-steps are then given by
-
-$$
-\begin{aligned}
-P(\epsilon = \texttt{False} \mid t) &= (1 - p_\epsilon)^t \\
-P(\epsilon = \texttt{True} \mid t) &= 1 - P(\epsilon = \texttt{False} \mid t)
-\end{aligned}
-$$
-
-In @fig-model-midext-evo, we visualize how the prior distribution over diagnose times $P(t)$, the conditional probability of midline extension $P(\epsilon \mid t)$, and their joint $P(\epsilon, t)$ evolve over the course of a patient evolution.
-
-![The top panel shows the prior probability to get diagnosed at time-step $t$ for early and late T-category tumors as bars.
-Also in the top panel, we plot the conditional probability of the tumor's midline extension ($\epsilon=\texttt{True}$), given the time-step $t$ as a line plot.
-In the bottom panel, we show the joint probability of getting diagnosed in time-step $t$ *and* having a tumor that crosses the midline.](figures/fig-model-midext-evo.png){#fig-model-midext-evo width="50%"}
-
-To get the joint probability over the ipsi- and contralateral hidden states, as well as the state of the tumor's midline extension $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)$, we simply add the above terms to the marginalization in @eq-bilateral-marginal:
-
-$$
-P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right) = \sum_{t=0}^{t_\text{max}} P(t) \cdot P(\epsilon \mid t) \cdot P \left( \mathbf{X}^\text{i} \mid t \right) \cdot P \left( \mathbf{X}^\text{c} \mid t \right)
-$$
-
-Again, this can be written algebraically, by defining the vector $P(\mathbf{t}, \epsilon) = P(\mathbf{t}) \cdot P(\epsilon \mid \mathbf{t})$, to achieve the same form as in @eq-bilateral-marginal-algebra:
-
-$$
-P \left( \mathbf{X}^\text{i} = \boldsymbol{\xi}_n, \mathbf{X}^\text{c} = \boldsymbol{\xi}_m, \epsilon \right) = \left[ \boldsymbol{\Lambda}^\intercal_\text{i} \cdot \operatorname{diag} P(\mathbf{t}, \epsilon) \cdot \boldsymbol{\Lambda}_\text{c} \right]_{n,m}
-$$
-
-In @fig-model-state-dist we plot the state distribution for a full midline model as two separate heatmaps.
-To keep it readable, this example model only considers the LNLs II, III, and IV ipsi- and contralaterally.
-
-![3D state distribution of a midline model with 3 LNLs (II, III, and IV) in both sides of the neck.](figures/fig-model-state-dist.png){#fig-model-state-dist}
-
 ## Parameter Symmetries {#sec-params-symmetry}
@@
 Together with the explicit modelling of the tumor's midline extension $\epsilon$ from @sec-midline, we now have a model that may be capable of capturing the higher prevalence of contralateral involvment that comes with tumors extending over the mid-sagittal line.
+
+## Modelling Midline Extension {#sec-midline}
+
+To account for the increased prevalence of involvement on the contralateral side when the tumor is not clearly lateralized anymore, we also model the tumor's extension over the mid-sagittal line as a binary random variable.
+It starts lateralized and at every time-step there is a finite probability $p_\epsilon$ that the tumor grows over the symmetry plane of the patient.
+
+Technically, the introduction of this additional random variable doubles the space of the hidden states and therefore quadruples the size of the transition matrix $\mathbf{A}$.
+But since only the transition matrix in case of midline extension depends on the lateralization of the tumor, we implement it differently and thereby make the computation slightly more efficient.
+
+First, the overall probabilities to find a patient with a clearly lateralized tumor or one that extends over the mid-sagittal line after $t$ time-steps are respectively given by
+
+$$
+\begin{aligned}
+P(\epsilon = \texttt{False} \mid t) &= (1 - p_\epsilon)^t \\
+P(\epsilon = \texttt{True} \mid t) &= 1 - P(\epsilon = \texttt{False} \mid t)
+\end{aligned}
+$$
+
+In @fig-model-midext-evo, we visualize how the prior distribution over diagnose times $P(t)$, the conditional probability of midline extension $P(\epsilon \mid t)$, and their joint $P(\epsilon, t)$ evolve over the course of a patient evolution.
+
+![The top panel shows the prior probability to get diagnosed at time-step $t$ for early and late T-category tumors as bars.
+Also in the top panel, we plot the conditional probability of the tumor's midline extension ($\epsilon=\texttt{True}$), given the time-step $t$ as a line plot.
+In the bottom panel, we show the joint probability of getting diagnosed in time-step $t$ *and* having a tumor that crosses the midline.](figures/fig-model-midext-evo.png){#fig-model-midext-evo width="50%"}
+
+Using this, it is straightforward to write down the joint probability of ipsi- and contralalteral states in case of a tumor that does *not* extend over the mid-sagittal line similar to the marginalization in @eq-bilateral-marginal:
+
+$$
+P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\texttt{False} \right) = \sum_{t=0}^{t_\text{max}} P(t) \cdot (1 - p_\epsilon)^t \cdot P \left( \mathbf{X}^\text{i} \mid t \right) \cdot P \left( \mathbf{X}^\text{c} \mid t, \epsilon=\texttt{False} \right)
+$$ {#eq-joint-noext}
+
+where for the contralateral evolution we used the transition matrix that depends on the base spread parameters $b_v^{\text{c},\epsilon=\texttt{False}}$.
+
+The case when midline extension is eventually present is more complicated: We already marginalize over the exact time-step when the tumor grows over the mid-sagittal line.
+But whenever that happens, we also need to change the contralateral transition matrix to use the increased spread rates from the tumor to the contralateral LNLs, given by the linear mix in @eq-mixing.
+We can achieve the correct marginalization by iteratively building the joint $P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid t \right)$ from $t=0$, where we know that all LNLs are healthy and the tumor is lateralized:
+
+$$
+P \left( \mathbf{X}^\text{c} = \boldsymbol{\xi}_0, \epsilon=\texttt{False} \mid t=0 \right) = 1
+$$
+
+Then we consider an arbitrary later time-step $t=\tau+1$. There are two possible scenarios we need to marginalize over:
+
+1.
+The tumor was still lateralized at $t=\tau$ and just grew over the midline.
+Under this scenario, the probability for having a midline extension at $t=\tau+1$ is $p_\epsilon$ and we use this to weight the contralateral state distribution that was so far evolved without increased contralateral spread.
+2.
+The mid-sagittal line was already crossed by the tumor.
+In that case, the probability is 1 to remain in that lateralization state.
+Thus, for this case we simply add the same distribution we are computing, but from one time-step earlier, arriving at a recursion relation:
+
+$$
+P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid \tau + 1 \right) = \big[ p_\epsilon P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{False} \mid \tau \right) + P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid \tau \right) \big]^\top \cdot \mathbf{A}^{\text{c},\epsilon=\texttt{True}}
+$$
+
+Using this, we can write the joint of ipsi- and contralateral involvement also for the case of mid-sagittal extension:
+
+$$
+P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\texttt{True} \right) = \sum_{t=0}^{t_\text{max}} P(t) \cdot P \left( \mathbf{X}^\text{i} \mid t \right) \cdot P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid t \right)
+$$ {#eq-joint-ext}
+
+Subsequently, @eq-joint-noext and @eq-joint-ext can be used to compute the likelihoods of patients with the respective tumor lateralization.
+If, for some reason, it is not known whether the tumor crossed the midline or not, we can simply add the two distributions to marginalize over midline extension as well.
+
+In @fig-model-state-dist we plot the state distribution for a full midline model as two separate heatmaps.
+To keep it readable, this example model only considers the LNLs II, III, and IV ipsi- and contralaterally.
+
+![3D state distribution of a midline model with 3 LNLs (II, III, and IV) in both sides of the neck.](figures/fig-model-state-dist.png){#fig-model-state-dist}
 

### 2024-06-24 manuscript.qmd
finish risk figure, update some captions

@@
 
-model = shared.get_model(which="simple", load_samples=True)
-samples = shared.get_samples(which="simple")
+model = shared.get_model(which="full", load_samples=True)
+samples = shared.get_samples(which="full")
 
@@
 
-![Comparison of predicted (histograms) vs observed (beta posteriors) prevalences.
-Shown are four scenarios, all including contralateral LNL II involvement: Early T-category and an ipsilateral N0 neck (green), early T-category and ipsilateral LNL II involvement (blue), as well as the same two scenarios but for advanced T-category (orange and red).
-The figure shows that the model is capable of describing the correlation between ipsi- and contralateral involvement.
-Although for the scenario of LNL II involvement in both sides, the prediction's split between early and advanced T-category is not large enough.](figures/fig-model-prevalences-with-ipsi.png){#fig-model-prevalences-with-ipsi}
+![Comparison of the computed and observed prevalences for scenarios that illustrate the model's capability of accounting for the correlation between ipsi- and contralateral involvement.
+We show two scenarios: One where the ipsilateral neck shows no involvement (at least LNLs I to V are healthy, LNL VII was ignored because data on it is missing for some patients) in blue and one where at least LNL III was involved in orange.
+These two scenarios are plotted for all combinations of T-category (early in top row, advanced in bottom row) and tumor lateralization (lateralized in left column, extending over mid-sagittal line in right column).](figures/fig-model-prevalences-with-ipsi.png){#fig-model-prevalences-with-ipsi}
 
@@
 It shows that the prevalence of metastases in the two sides of the neck is correlated via the time of diagnosis, despite the model not having any direct connections between the two side.
-However, there are some small discrepancies in the model's prediction: When considering the scenario of LNL II involvement in the ipsi- *and* contralateral side, it cannot quite capture the split between early and late T-category.
-The model slightly overestimates the prevalence of this scenario for early T-category patients and underestimates it for advanced T-category.
+However, there are some discrepancies in the model's prediction: It cannot quite capture the correlation of ipsi- and contralateral involvement that is seen in the data.
+This is notable for example in the bottom right subplot where the prevalence is overestimated for the scenario of an (almost) healthy ipsilateral neck and underestimated when at least LNL III shows metastases.
+One reason for this may be the width of the time-prior: If it was wider, the posterior over the diagnosis time given the ipsilateral involvement could more flexibly shift to reflect a more severe contralateral involvement.
+
+<!-- NOTE: This last sentence is purely speculative! We need to discuss this. -->
 
@@
 
-![Histograms over the predicted risk of occult contralateral LNL II invovlement, shown for some combinations of T-category, tumor lateralization, and ipsilateral clinical involvement.
-The contralateral side was always assumed to be clinically negative.](figures/fig-model-risks.png){#fig-model-risks width=50%}
+![Histograms over the predicted risk of occult contralateral LNL II (left), III (middle), and IV (right) invovlement, shown for some combinations of T-category, tumor lateralization, and ipsilateral clinical involvement.
+Unless otherwise mentioned in the legend, the entire contralateral side was assumed to be clinically node negative.](figures/fig-model-risks.png){#fig-model-risks}
 
-@Fig-model-risks shows this predicted risk of occult disease in contralateral LNL II for four interesting combinations of these three risk factors.
+@Fig-model-risks shows this predicted risk of occult disease in the contralateral LNLs II, III, and IV for interesting combinations of these three risk factors.
 This risk is computed *given* a CT diagnosis assessing the per-LNL clinical involvement for which we assumed a sensitivity of 81% and a specificity of 76% [@de_bondt_detection_2007].
 
-The variable impacting the prediction for contralateral involvement in our model is the tumor's lateralization.
+### Contralateral LNL II
+
+The variable impacting the prediction for contralateral level II involvement in our model is the tumor's lateralization.
 For example, a patient with a clearly lateralized early T-category tumor and a clinically N0 neck is assigned a 1-2% <!-- TODO: make this dynamic --> risk for occult disease in contralateral LNL II.
@@
 Taken together, T-category and ipsilateral involvement may still considerably impact the risk prediction for contralateral involvement: In @fig-model-risks the scenarios underlying the orange (6.5% <!-- TODO: make this dynamic -->) and the red (11% <!-- TODO: make this dynamic -->) histograms differ in T-category (early vs advanced) and clinical diagnosis (N0 vs ipsilateral at least LNL III involved).
+
+### Contralalteral LNL III
+
+As shown in the center subplot of @fig-model-risks, the risk for occult disease in the contralateral LNL III only ever crosses a 5% threshold if the upstream level II is involved too, resulting in a risk of 5% to 6%<!-- TODO: make dynamic -->.
+As long as the contralateral neck appears clinically node negative, however, even an advanced T-category and midline extension do not yield a risk higher than 2%.
+
+### Contralateral LNL IV
+
+Our model's prediction for the contralateral LNL IV show that even in the most extreme case, with an advanced T-category tumor extending over the mid-sagittal line as well as metastases in every ipsilateral LNL and every contralateral upstream level, the risk does not exceed 5%
 

### 2024-06-25 model.qmd

--- 2024-06-20 model.qmd
@@
   axes=axes[0,0],
-  titles=["ipsi N0", "ipsi LNL III"],
+  titles=["ipsi LNLs I-V healthy", "ipsi LNL III"],
   locs=[(0.3, 0.99), (0.5, 0.55)],
@@
   axes=axes[0,1],
-  titles=["ipsi N0", "ipsi LNL III"],
+  titles=["ipsi LNLs I-V healthy", "ipsi LNL III"],
   locs=[(0.3, 0.99), (0.5, 0.55)],
@@
   axes=axes[1,0],
-  titles=["ipsi N0", "ipsi LNL III"],
+  titles=["ipsi LNLs I-V healthy", "ipsi LNL III"],
   locs=[(0.16, 0.99), (0.53, 0.55)],
@@
   axes=axes[1,1],
-  titles=["ipsi N0", "ipsi LNL III"],
+  titles=["ipsi LNLs I-V healthy", "ipsi LNL III"],
   locs=[(0.13, 0.55), (0.5, 0.99)],
@@
 
-all_false = {lnl: False for lnl in ["I", "II", "III", "IV", "V"]}
+all_false = {lnl: False for lnl in ["I", "II", "III", "IV", "V", "VII"]}
+all_true = {lnl: True for lnl in ["I", "II", "III", "IV", "V", "VII"]}
 
@@
 
+scenario = Scenario(
+  t_stages=["early"],
+  midext=True,
+  diagnosis={
+    "ipsi": {"CT": all_false},
+    "contra": {"CT": all_false},
+  },
+  involvement={"contra": {"III": True}},
+)
+contents.append(Histogram.from_hdf5(
+  filename=paths.model_dir / "full" / "risks.hdf5",
+  dataname=scenario.md5_hash("risks"),
+  color=COLORS["green"],
+  label="early, mid. ext.; N0",
+))
+
+scenario = Scenario(
+  t_stages=["late"],
+  midext=True,
+  diagnosis={
+    "ipsi": {"CT": all_false},
+    "contra": {"CT": all_false},
+  },
+  involvement={"contra": {"III": True}},
+)
+contents.append(Histogram.from_hdf5(
+  filename=paths.model_dir / "full" / "risks.hdf5",
+  dataname=scenario.md5_hash("risks"),
+  color=COLORS["blue"],
+  label="adv., mid. ext.; N0",
+))
+
+scenario = Scenario(
+  t_stages=["late"],
+  midext=True,
+  diagnosis={
+    "ipsi": {"CT": {"III": True}},
+    "contra": {"CT": {"I": False, "II": True, "III": False, "IV": False, "V": False, "VII": False}},
+  },
+  involvement={"contra": {"III": True}},
+)
+contents.append(Histogram.from_hdf5(
+  filename=paths.model_dir / "full" / "risks.hdf5",
+  dataname=scenario.md5_hash("risks"),
+  color=COLORS["red"],
+  label="adv., mid. ext.;\nipsi LNL III & contra LNL II",
+))
+
+draw(ax[1], contents, xlims=(0., 8.))
+ax[1].legend(labelspacing=0.1)
+ax[1].set_xlabel("risk [%]")
 ax[1].set_yticks([])
-ax[1].set_xticks(np.linspace(0, 15, 6))
+ax[1].set_xticks(np.linspace(0, 8, 9))
+ax[1].set_title("Contralateral LNL III")
 
+# contra LNL IV
+contents = []
+
+scenario = Scenario(
+  t_stages=["late"],
+  midext=True,
+  diagnosis={
+    "ipsi": {"CT": {"III": True}},
+    "contra": {"CT": all_false},
+  },
+  involvement={"contra": {"IV": True}},
+)
+contents.append(Histogram.from_hdf5(
+  filename=paths.model_dir / "full" / "risks.hdf5",
+  dataname=scenario.md5_hash("risks"),
+  color=COLORS["green"],
+  label="adv., mid. ext.; ipsi LNL III",
+))
+
+scenario = Scenario(
+  t_stages=["late"],
+  midext=True,
+  diagnosis={
+    "ipsi": {"CT": all_true},
+    "contra": {"CT": {"I": True, "II": True, "III": True, "IV": False, "V": False, "VII": False}},
+  },
+  involvement={"contra": {"IV": True}},
+)
+contents.append(Histogram.from_hdf5(
+  filename=paths.model_dir / "full" / "risks.hdf5",
+  dataname=scenario.md5_hash("risks"),
+  color=COLORS["red"],
+  label="adv., mid. ext.;\nsevere bilat. inv.",
+))
+
+draw(ax[2], contents, xlims=(0., 5.))
+ax[2].legend(labelspacing=0.1)
+ax[2].set_xlabel("risk [%]")
 ax[2].set_yticks([])
-ax[2].set_xticks(np.linspace(0, 15, 6))
+ax[2].set_xticks(np.linspace(0, 5, 6))
+ax[2].set_title("Contralateral LNL IV")
 

### 2024-06-26 manuscript.qmd
adjust mid ext theory section

@@
 
-For later use, we define at this point the matrices $\boldsymbol{\Lambda}$:
+For later use, we define at this point the matrices of the dsitributions over all hidden states, given all time-steps $\boldsymbol{\Lambda}$:
 
@@
 In this work, we will only differentiate between "early" (T1 & T2) and "advanced" (T3 & T4) T-categories.
-Therefore, our parameter space is:
+Therefore, the parameter space of the unilateral model is:
 
@@
 
-Together with the explicit modelling of the tumor's midline extension $\epsilon$ from @sec-midline, we now have a model that may be capable of capturing the higher prevalence of contralateral involvment that comes with tumors extending over the mid-sagittal line.
+Our parameter space has now expanded to
+
+$$
+\boldsymbol{\theta} = \left( \left\{ b_v^\text{i} \right\}, \left\{ b_v^\text{c} \right\}, \alpha, \left\{ t_{vr} \right\}, p_\text{early}, p_\text{adv.} \right) \quad \text{with} \quad \genfrac{}{}{0pt}{2}{v\leq V}{r\in\operatorname{pa}(v)}
+$$ {#eq-bi-param-space}
+
+which is less than double in size, compared to the unilateral model.
+From these parameters, we now build three different transition matrices: The unchanged $\mathbf{A}_\text{i}$ for the ipsilateral side, one $\mathbf{A}_\text{c}^{\epsilon=\texttt{False}}$ for the contralateral side that covers the progression as long as the tumor is lateralized, and $\mathbf{A}_\text{c}^{\epsilon=\texttt{True}}$ in case of a tumor that has crossed the mid-sagittal line.
 
@@
 
-Technically, the introduction of this additional random variable doubles the space of the hidden states and therefore quadruples the size of the transition matrix $\mathbf{A}$.
-But since only the transition matrix in case of midline extension depends on the lateralization of the tumor, we implement it differently and thereby make the computation slightly more efficient.
-
-First, the overall probabilities to find a patient with a clearly lateralized tumor or one that extends over the mid-sagittal line after $t$ time-steps are respectively given by
+The overall probabilities to find a patient with a clearly lateralized tumor or one that extends over the mid-sagittal line after $t$ time-steps are then respectively given by
 
@@
 
-Using this, it is straightforward to write down the joint probability of ipsi- and contralalteral states in case of a tumor that does *not* extend over the mid-sagittal line similar to the marginalization in @eq-bilateral-marginal:
+Using this, it is straightforward to write down the matrix of state distributions for all time-steps, as in @eq-lambda-matrix covering the contralateral hidden state evolution:
 
 $$
-P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\texttt{False} \right) = \sum_{t=0}^{t_\text{max}} P(t) \cdot (1 - p_\epsilon)^t \cdot P \left( \mathbf{X}^\text{i} \mid t \right) \cdot P \left( \mathbf{X}^\text{c} \mid t, \epsilon=\texttt{False} \right)
-$$ {#eq-joint-noext}
+\boldsymbol{\Lambda}_\text{c}^{\epsilon=\texttt{False}} =
+\left(
+\begin{array}{r}
+\boldsymbol{\pi}^\intercal \cdot \left( \mathbf{A}_\text{c}^{\epsilon=\texttt{False}} \right)^{0\phantom{t_\text{max}}} \\
+(1-p_\epsilon) \cdot \boldsymbol{\pi}^\intercal \cdot \left( \mathbf{A}_\text{c}^{\epsilon=\texttt{False}} \right)^{1\phantom{t_\text{max}}} \\
+\hfill \vdots \hfill \\
+(1-p_\epsilon)^{t_\text{max}} \cdot \boldsymbol{\pi}^\intercal \cdot \left( \mathbf{A}_\text{c}^{\epsilon=\texttt{False}} \right)^{t_\text{max}\phantom{0}} \\
+\end{array}
+\right)
+$$
 
-where for the contralateral evolution we used the transition matrix that depends on the base spread parameters $b_v^{\text{c},\epsilon=\texttt{False}}$.
+where we used the transition matrix $\mathbf{A}_\text{c}^{\epsilon=\texttt{False}}$ that depends on the base spread parameters $b_v^{\text{c},\epsilon=\texttt{False}}$.
 
 The case when midline extension is eventually present is more complicated: We already marginalize over the exact time-step when the tumor grows over the mid-sagittal line.
-But whenever that happens, we also need to change the contralateral transition matrix to use the increased spread rates from the tumor to the contralateral LNLs, given by the linear mix in @eq-mixing.
-We can achieve the correct marginalization by iteratively building the joint $P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid t \right)$ from $t=0$, where we know that all LNLs are healthy and the tumor is lateralized:
+But whenever that happens, we also need to change the contralateral transition matrix to use the increased spread rates $b_v^{\text{c}, \epsilon=\texttt{True}}$ from the tumor to the contralateral LNLs, given by the linear mix in @eq-mixing.
+We can achieve the correct marginalization by iteratively building the joint $P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid t \right)$.
+We start from $t=0$, where we know that all LNLs are healthy - i.e., the contralateral neck is in the state $\mathbf{X}_\text{c}=\boldsymbol{\xi}_0$ - and the tumor is lateralized ($\epsilon=\texttt{False}$):
 
 $$
-P \left( \mathbf{X}^\text{c} = \boldsymbol{\xi}_0, \epsilon=\texttt{False} \mid t=0 \right) = 1
+P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{False} \mid t=0 \right) = \boldsymbol{\xi}_0
 $$
@@
 The tumor was still lateralized at $t=\tau$ and just grew over the midline.
-Under this scenario, the probability for having a midline extension at $t=\tau+1$ is $p_\epsilon$ and we use this to weight the contralateral state distribution that was so far evolved without increased contralateral spread.
+Under this scenario, the probability for having a midline extension at $t=\tau+1$ is given by $p_\epsilon$.
+We use this to weight the contralateral state distribution that was so far evolved without increased contralateral spread.
 2.
@@
 $$
-P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid \tau + 1 \right) = \big[ p_\epsilon P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{False} \mid \tau \right) + P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid \tau \right) \big]^\top \cdot \mathbf{A}^{\text{c},\epsilon=\texttt{True}}
+P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid \tau + 1 \right) = \big[ p_\epsilon P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{False} \mid \tau \right) + P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid \tau \right) \big]^\top \cdot \mathbf{A}_\text{c}^{\epsilon=\texttt{True}}
 $$
 
-Using this, we can write the joint of ipsi- and contralateral involvement also for the case of mid-sagittal extension:
+We can collect the iteratively computed distributions for the midline extension case to define the matrix over the states given all time-steps, in analogy to @eq-lambda-matrix:
 
 $$
-P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\texttt{True} \right) = \sum_{t=0}^{t_\text{max}} P(t) \cdot P \left( \mathbf{X}^\text{i} \mid t \right) \cdot P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid t \right)
-$$ {#eq-joint-ext}
+\boldsymbol{\Lambda}_\text{c}^{\epsilon=\texttt{True}} = \begin{pmatrix}
+P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid 0 \right) \\
+P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid 1 \right) \\
+\vdots \\
+P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid t_\text{max} \right) \\
+\end{pmatrix}
+$$
 
-Subsequently, @eq-joint-noext and @eq-joint-ext can be used to compute the likelihoods of patients with the respective tumor lateralization.
-If, for some reason, it is not known whether the tumor crossed the midline or not, we can simply add the two distributions to marginalize over midline extension as well.
+Using this, we can again write the joint of ipsi- and contralateral involvement - now also for the case of mid-sagittal extension - algebraically as before in @eq-bilateral-marginal-algebra:
+
+$$
+P \left( \mathbf{X}^\text{i} = \boldsymbol{\xi}_n, \mathbf{X}^\text{c} = \boldsymbol{\xi}_m, \epsilon \right) = \left[ \boldsymbol{\Lambda}^\intercal_\text{i} \cdot \operatorname{diag} P(\mathbf{t}) \cdot \boldsymbol{\Lambda}_\text{c}^\epsilon \right]_{n,m}
+$$
+
+We can use this term to compute the likelihood of all patients with and without midline extension separately.
+And if for some patients the information of tumor lateralization is not available, we can simply sum the above term once for $\epsilon=\texttt{False}$ and once for $\epsilon=\texttt{True}$ to marginalize over the unknown variable.
+
+The final parameter space of our extended model has now reached this size:
+
+$$
+\boldsymbol{\theta} = \left( \left\{ b_v^\text{i} \right\}, \left\{ b_v^\text{c} \right\}, \alpha, \left\{ t_{vr} \right\}, p_\text{early}, p_\text{adv.}, p_\epsilon \right) \quad \text{with} \quad \genfrac{}{}{0pt}{2}{v\leq V}{r\in\operatorname{pa}(v)}
+$$ {#eq-ext-param-space}
 
 In @fig-model-state-dist we plot the state distribution for a full midline model as two separate heatmaps.
-To keep it readable, this example model only considers the LNLs II, III, and IV ipsi- and contralaterally.
+To keep it manageable to interpret, this example model only considers the LNLs II, III, and IV ipsi- and contralaterally, resulting in $2 \times 2^3 \times 2^3 = 128$ distinct states.
 
-![3D state distribution of a midline model with 3 LNLs (II, III, and IV) in both sides of the neck.](figures/fig-model-state-dist.png){#fig-model-state-dist}
+![3D state distribution of a midline model with 3 LNLs (II, III, and IV) in both sides of the neck.
+Note the higher probability for contralateral involvement in the right subplot, where the tumor extends over the mid-sagittal line.](figures/fig-model-state-dist.png){#fig-model-state-dist}
 

### 2024-06-26 data.qmd
update risk scenarios & switch to SVG

@@
 
-plt.savefig(paths.figure_dir / "fig-data-strat.png", dpi=400)
+plt.savefig(paths.figure_dir / "fig-data-strat.svg")
 plt.show()
@@
 
-plt.savefig(paths.figure_dir / "fig-data-upset-ipsi.png", dpi=400)
+plt.savefig(paths.figure_dir / "fig-data-upset-ipsi.svg")
 plt.show()
@@
 
-plt.savefig(paths.figure_dir / "fig-data-upset-contra.png", dpi=400)
+plt.savefig(paths.figure_dir / "fig-data-upset-contra.svg")
 plt.show()

### 2024-06-26 manuscript.qmd
update risk scenarios & switch to SVG

@@
 
-![Contralateral involvement stratified by T-category (left panel), the number of metastatic LNLs ipsilaterally (center panel), and whether the primary tumor extended over the mid-sagittal line or was clearly lateralized (right panel).](figures/fig-data-strat.png){#fig-data-strat}
+![Contralateral involvement stratified by T-category (left panel), the number of metastatic LNLs ipsilaterally (center panel), and whether the primary tumor extended over the mid-sagittal line or was clearly lateralized (right panel).](figures/fig-data-strat.svg){#fig-data-strat}
 
@@
 The orange arcs represent the sensitivity and specificity of the observational modality (e.g.
-CT, MRI, pathology, ...).](static/small-graph.png){#fig-small-graph width=40%}
+CT, MRI, pathology, ...).](static/small-graph.svg){#fig-small-graph width=40%}
 
@@
 Also in the top panel, we plot the conditional probability of the tumor's midline extension ($\epsilon=\texttt{True}$), given the time-step $t$ as a line plot.
-In the bottom panel, we show the joint probability of getting diagnosed in time-step $t$ *and* having a tumor that crosses the midline.](figures/fig-model-midext-evo.png){#fig-model-midext-evo width="50%"}
+In the bottom panel, we show the joint probability of getting diagnosed in time-step $t$ *and* having a tumor that crosses the midline.](figures/fig-model-midext-evo.svg){#fig-model-midext-evo width="50%"}
 
@@
 ![3D state distribution of a midline model with 3 LNLs (II, III, and IV) in both sides of the neck.
-Note the higher probability for contralateral involvement in the right subplot, where the tumor extends over the mid-sagittal line.](figures/fig-model-state-dist.png){#fig-model-state-dist}
+Note the higher probability for contralateral involvement in the right subplot, where the tumor extends over the mid-sagittal line.](figures/fig-model-state-dist.svg){#fig-model-state-dist}
 
@@
 Right: Fraction of accepted MCMC proposals averaged over all parallel walkers.
-Values around 30% indicate good mixing of the walkers.](figures/fig-model-burnin-history.png){#fig-model-burnin-history}
+Values around 30% indicate good mixing of the walkers.](figures/fig-model-burnin-history.svg){#fig-model-burnin-history}
 
@@
 The left column depicts scenarios where the primary tumor is clearly lateralized, the right column scenarios of tumors extending over the mid-sagittal line.
-This figure illustrates the model's ability to describe the prevalence of different combinations of scenarios involving the risk factors T-category and midline extension.](figures/fig-model-prevalences-overall.png){#fig-model-prevalences-overall}
+This figure illustrates the model's ability to describe the prevalence of different combinations of scenarios involving the risk factors T-category and midline extension.](figures/fig-model-prevalences-overall.svg){#fig-model-prevalences-overall}
 
@@
 We show two scenarios: One where the ipsilateral neck shows no involvement (at least LNLs I to V are healthy, LNL VII was ignored because data on it is missing for some patients) in blue and one where at least LNL III was involved in orange.
-These two scenarios are plotted for all combinations of T-category (early in top row, advanced in bottom row) and tumor lateralization (lateralized in left column, extending over mid-sagittal line in right column).](figures/fig-model-prevalences-with-ipsi.png){#fig-model-prevalences-with-ipsi}
+These two scenarios are plotted for all combinations of T-category (early in top row, advanced in bottom row) and tumor lateralization (lateralized in left column, extending over mid-sagittal line in right column).](figures/fig-model-prevalences-with-ipsi.svg){#fig-model-prevalences-with-ipsi}
 
@@
 While the prevalence is predicted correctly when marginalizing over T-category, the model cannot capture the degree of separation observed in the data.
-Since the tumor's midline extension is virtually always part of the diagnosis and hence *given* when predicting a patient's risk, we do not consider this discrepancy a major issue.](figures/fig-model-prevalences-midext.png){#fig-model-prevalences-midext width="50%"}
+Since the tumor's midline extension is virtually always part of the diagnosis and hence *given* when predicting a patient's risk, we do not consider this discrepancy a major issue.](figures/fig-model-prevalences-midext.svg){#fig-model-prevalences-midext width="50%"}
 
@@
 
-![Histograms over the predicted risk of occult contralateral LNL II (left), III (middle), and IV (right) invovlement, shown for some combinations of T-category, tumor lateralization, and ipsilateral clinical involvement.
-Unless otherwise mentioned in the legend, the entire contralateral side was assumed to be clinically node negative.](figures/fig-model-risks.png){#fig-model-risks}
+![Histograms over the predicted risk of occult contralateral LNL II (left), III (middle), and IV (right) invovlement, shown for some combinations of T-category, tumor lateralization, and clinical LNL diagnoses.
+All LNLs not explicitly mentioned in the legend, including the LNL for which the risk of occult disease was computed, where assumed to be clinically negative.](figures/fig-model-risks.svg){#fig-model-risks}
 
@@
 While the blue PDF is a marginal distribution summed over all possible combinations, which are shown in shades of red.
-The more permutations a particular combination has, the more opaque it is plotted.](figures/fig-model-marginal-pdf.png){#fig-model-marginal-pdf width="50%"}
+The more permutations a particular combination has, the more opaque it is plotted.](figures/fig-model-marginal-pdf.svg){#fig-model-marginal-pdf width="50%"}
 

### 2024-06-26 model.qmd
update risk scenarios & switch to SVG

--- 2024-06-20 model.qmd
@@
 
-plt.savefig(paths.figure_dir / "fig-model-midext-evo.png", dpi=400)
+plt.savefig(paths.figure_dir / "fig-model-midext-evo.svg")
 plt.show()
@@
 
-plt.savefig(paths.figure_dir / "fig-model-burnin-history.png", dpi=400)
+plt.savefig(paths.figure_dir / "fig-model-burnin-history.svg")
 plt.show()
@@
 
-plt.savefig(paths.figure_dir / "fig-model-state-dist.png", dpi=400)
+plt.savefig(paths.figure_dir / "fig-model-state-dist.svg")
 plt.show()
@@
 
-plt.savefig(paths.figure_dir / "fig-model-prevalences-overall.png", dpi=400)
+plt.savefig(paths.figure_dir / "fig-model-prevalences-overall.svg")
 plt.show()
@@
 
-plt.savefig(paths.figure_dir / "fig-model-prevalences-with-ipsi.png", dpi=400)
+plt.savefig(paths.figure_dir / "fig-model-prevalences-with-ipsi.svg")
 plt.show()
@@
 
-plt.savefig(paths.figure_dir / "fig-model-prevalences-midext.png", dpi=400)
+plt.savefig(paths.figure_dir / "fig-model-prevalences-midext.svg")
 plt.show()
@@
 import numpy as np
+import h5py
 import matplotlib.pyplot as plt
@@
 from scripts import paths, shared
-
-all_false = {lnl: False for lnl in ["I", "II", "III", "IV", "V"]}
 
@@
 
-fig, ax = plt.subplots(nrows=nrows, ncols=ncols)
+fig, axes = plt.subplots(nrows=nrows, ncols=ncols)
 
-contents = []
+contents = {"II": [], "III": [], "IV": []}
+mean_lists = {"II": [], "III": [], "IV": []}
 defaults = {
@@
 }
+t_stage_map = {"early": "early", "late": "adv."}
+midext_map = {False: "lateralized", True: "mid. ext."}
+colors = [COLORS["green"], COLORS["blue"], COLORS["orange"], COLORS["red"]]
 
-# contra LNL II
-scenario = Scenario(
-  t_stages=["early"],
-  midext=False,
-  diagnosis={
-    "ipsi": {"CT": all_false},
-    "contra": {"CT": all_false},
+def get_diag_label(diagnosis) -> str:
+  label = ""
+  is_n0 = True
+
+  ipsi_positive = [lnl for lnl, status in diagnosis.get("ipsi", {}).get("CT", {}).items() if status]
+  if len(ipsi_positive) > 0:
+    is_n0 = False
+    label += "i:" + ",".join(ipsi_positive)
+
+  contra_positive = [lnl for lnl, status in diagnosis.get("contra", {}).get("CT", {}).items() if status]
+  if len(contra_positive) > 0:
+    is_n0 = False
+    label += "; c:" + ",".join(contra_positive)
+
+  return label if not is_n0 else "N0"
+
+def get_label(scenario: Scenario) -> str:
+  t = scenario.t_stages[0]
+  m = scenario.midext
+  d = eval(scenario.diagnosis)
+  return f"{t_stage_map[t]}; {midext_map[m]}; {get_diag_label(d)}"
+
+with h5py.File(name=paths.model_dir / "full" / "risks.hdf5", mode="r") as h5file:
+  for dset in h5file.values():
+    scenario = Scenario.from_dict(dict(dset.attrs))
+    for_subplot = list(eval(scenario.involvement)["contra"].keys()).pop()
+    mean_lists[for_subplot].append(dset[:].mean())
+
+indices = {}
+for lnl, means in mean_lists.items():
+  indices[lnl] = np.argsort(means)
+
+counter = {"II": 0, "III": 0, "IV": 0}
+with h5py.File(name=paths.model_dir / "full" / "risks.hdf5", mode="r") as h5file:
+  for dset in h5file.values():
+    scenario = Scenario.from_dict(dict(dset.attrs))
+    for_subplot = list(eval(scenario.involvement)["contra"].keys()).pop()
+    c = counter[for_subplot]
+    counter[for_subplot] += 1
+    contents[for_subplot].append(
+      Histogram(
+        values=dset[:],
+        kwargs={
+          "label": get_label(scenario),
+          "color": colors[indices[for_subplot][c]],
         },
-  involvement={"contra": {"II": True}},
       )
-contents.append(Histogram.from_hdf5(
-  filename=paths.model_dir / "full" / "risks.hdf5",
-  dataname=scenario.md5_hash("risks"),
-  color=COLORS["green"],
-  label="early, lateralized; N0",
-))
+    )
 
-scenario = Scenario(
-  t_stages=["late"],
-  midext=False,
-  diagnosis={
-    "ipsi": {"CT": {"III": True}},
-    "contra": {"CT": all_false},
-  },
-  involvement={"contra": {"II": True}},
-)
-contents.append(Histogram.from_hdf5(
-  filename=paths.model_dir / "full" / "risks.hdf5",
-  dataname=scenario.md5_hash("risks"),
-  color=COLORS["blue"],
-  label="adv., lateralized; ipsi LNL III",
-))
+xlims = {
+  "II": (0, 15),
+  "III": (0, 7),
+  "IV": (0, 5),
+}
+for ax, (lnl, content) in zip(axes, contents.items()):
+  draw(ax, content, xlims=xlims[lnl])
+  ax.set_title(f"Contralateral LNL {lnl}")
+  ax.set_yticks([])
+  ax.set_xlabel("risk [%]")
+  ax.legend(labelspacing=0.1)
 
-scenario = Scenario(
-  t_stages=["early"],
-  midext=True,
-  diagnosis={
-    "ipsi": {"CT": all_false},
-    "contra": {"CT": all_false},
-  },
-  involvement={"contra": {"II": True}},
-)
-contents.append(Histogram.from_hdf5(
-  filename=paths.model_dir / "full" / "risks.hdf5",
-  dataname=scenario.md5_hash("risks"),
-  color=COLORS["orange"],
-  label="early, mid. ext.; N0",
-))
+axes[0].set_xticks(np.linspace(xlims["II"][0], xlims["II"][1], 6))
 
-scenario = Scenario(
-  t_stages=["late"],
-  midext=True,
-  diagnosis={
-    "ipsi": {"CT": {"III": True}},
-    "contra": {"CT": all_false},
-  },
-  involvement={"contra": {"II": True}},
-)
-contents.append(Histogram.from_hdf5(
-  filename=paths.model_dir / "full" / "risks.hdf5",
-  dataname=scenario.md5_hash("risks"),
-  color=COLORS["red"],
-  label="adv., mid. ext.; ipsi LNL III",
-))
-
-draw(ax[0], contents, xlims=(0., 15.))
-ax[0].legend(labelspacing=0.1)
-ax[0].set_xlabel("risk [%]")
-ax[0].set_yticks([])
-ax[0].set_xticks(np.linspace(0, 15, 6))
-ax[0].set_title("Contralateral LNL II")
-
-# contra LNL III
-contents = []
-
-ax[1].set_yticks([])
-ax[1].set_xticks(np.linspace(0, 15, 6))
-
-ax[2].set_yticks([])
-ax[2].set_xticks(np.linspace(0, 15, 6))
-
-plt.savefig(paths.figure_dir / "fig-model-risks.png", dpi=400)
+plt.savefig(paths.figure_dir / "fig-model-risks.svg")
 plt.show()
@@
 
-plt.savefig(paths.figure_dir / "fig-model-marginal-pdf.png", dpi=400)
+plt.savefig(paths.figure_dir / "fig-model-marginal-pdf.svg")
 plt.show()

### 2024-07-18 manuscript.qmd
incorporate Jan's first round of comments

@@
 abstract: |
-  In current practice, large parts of the contralateral lymph system are included in the elective clinical target volume (CTV-N) of oropharyngeal squamous cell carcinoma (OPSCC) patients.
-Guidelines on how to define the CTV-N are mostly based on prevalences of nodal involvement in the different lymph node levels (LNLs) and do not consider the detailed, individual diagnosis of a patient.
+  Purpose: According to current guidelines for elective nodal irradiation of oropharyngeal squamous cell carcinoma (OPSCC) patients, large parts of the contralateral lymphatic system are included in the elective clinical target volume (CTV-N), even for lateralized tumors without clinical lymph node involvement in the contralateral neck.
+In this work, we present a probabilistic model for bilateral lymphatic tumor progression in OPSCC to predict the personal risk for occult disease in any lymph node level (LNL), given the patient's clinical lymph node involvement, T-stage, and lateralization of the primary tumor.
 
-  In this work, we present a probabilistic model for bilateral lymphatic tumor progression in OPSCC.
-Its aim is to predict the personal risk for occult disease in any LNL, given the patient's clincal diagnosis, T-category, and lateralization of the primary tumor.
-The model represents each of the LNLs I, II, III, IV, V, and VII of both sides of the neck as a hidden binary random variable that are connected to the tumor and among each other via arcs that correspond to spread probabilities per time-step.
+  Methods: We extend a hidden markov model for lymphatic tumor progression, which was previously developed for ipsilateral lymph node involvement, to the contralateral neck.
+The model represents each of the LNLs I, II, III, IV, V, and VII of both sides of the neck as a hidden binary random variable with states healthy and involved.
+LNLs are connected to the tumor and among each other via arcs that correspond to spread probabilities.
 These spread probability rates are learned via Markov chain Monte Carlo (MCMC) sampling from a dataset of 833 OPSCC patients.
 
-  The model is able to describe the data well and may in the future be used to only include LNLs in a CTV-N that show a risk above a specific threshold for that particular patient's diagnosis.
-E.g., it predicts the risk for occult disease in contralateral LNL II to be below 5% unless the tumor crosses the mid-sagittal line.
+  Results: The model is able to describe the data on lymph node involvement well with a small number of interpretable parameters.
+Midline extension of the primary tumor is the main risk factor for contralateral involvement.
+In addition, the risk of occult metastases in contralateral lymph node level increases with more advanced T-stage and more severe ipsilateral involvement.
+The probability of involvement in contralateral level III is very low if the upstream level II is clinically negative.
+Similarly, the probability of involvement in contralateral level IV is very low if the upstream level III is clinically negative.
+
+  Conclusions: The model may guide personalized volume reduction of the elective CTV-N.
+It suggests that the contralateral neck may be spared from elective irradiation for lateralized tumors not crossing the midline.
+In patients with primary tumors crossing the midline but clinically negative contralateral neck, the contralateral elective irradiation may be limited to level II.
 ---
@@
 #| echo: false
-
 import re
@@
 When treating head and neck squamous cell carcinomas (HNSCC) with radiotherapy or surgery, not only the primary tumor and the clinically detected lymph node metastases are targeted.
-Per current guidelines, large volumes of the neck - lymph node levels (LNLs) - are part of the the elective clinical target volume (CTV-N) [@gregoire_ct-based_2003;@gregoire_delineation_2014;@gregoire_delineation_2018;@eisbruch_intensity-modulated_2002;@biau_selection_2019;@chao_determination_2002;@vorwerk_guidelines_2011;@ferlito_elective_2009].
-This is to minimize the risk of regional recurrences due to missed microscopic disease which cannot be detected by current in-vivo imaging modalities such as computed tomography (CT), magnetic resonance imaging (MRI), or positron emission tomography (PET).
+Per current guidelines, large volumes of the neck are part of the the elective clinical target volume (CTV-N) [@gregoire_ct-based_2003;@gregoire_delineation_2014;@gregoire_delineation_2018;@eisbruch_intensity-modulated_2002;@biau_selection_2019;@chao_determination_2002;@vorwerk_guidelines_2011;@ferlito_elective_2009].
+This is to minimize the risk of regional recurrences due to untreated microscopic disease which cannot be detected by current in-vivo imaging modalities such as computed tomography (CT), magnetic resonance imaging (MRI), or positron emission tomography (PET).
+However, minimizing the risk of missing occult disease in the lymph drainage region must be balanced against the toxicity related to unnecessary treatment of healthy tissue.
 
-Minimizing this risk of missing occult disease in the lymph drainage region must be balanced against the toxicity related to unnecessary treatment of healthy tissue.
-
-The aforementioned guidelines for the definition of the CTV-N quantify this risk via the overall prevalence of lymph node metastases in the LNLs.
+The aforementioned guidelines for the definition of the CTV-N are based on anatomically defined lymph node levels (LNL) [@gregoire_delineation_2014] and the overall prevalence of lymph node metastases in the LNLs.
 Based on this, they recommend extensive irradiation of both sides of the neck for a majority of patients.
-However, this prevalence the guidelines are based on does not take into account the individual state of tumor progression, which may have a great impact on a patient's personal risk for occult disease in a given LNL.
+However, this prevalence the guidelines are based on does not take into account an individual patient's state of tumor progression, which may have a great impact on a patient's personal risk for occult disease in a given LNL.
 For example, a patient with substantial nodal involvement in the ipsilateral side and an advanced primary tumor that crosses the mid-sagittal plane would receive extensive elective irradiation of the contralateral LNLs II, III, and IVa.
@@
 
-To better quantify this personalized risk we have previously developed an intuitive, probabilistic hidden Markov model [@ludwig_hidden_2021;@ludwig_modelling_2023] which was originally based on a conceptually similar Bayesian network model [@pouymayou_bayesian_2019].
+To better quantify this personalized risk of occult disease we have previously developed an intuitive, probabilistic hidden Markov model [@ludwig_hidden_2021;@ludwig_modelling_2023] which was originally based on a conceptually similar Bayesian network model [@pouymayou_bayesian_2019].
 So far, these models were limited to the ipsilateral side of the neck.
-The main focus of this work will be to extend the formalism of these previous works to include risk predictions for the contralateral side as well.
-We believe this holds the potential to greatly reduce the electively irradiated volumes in the contralateral neck for patients whose individual risk for occult disease in that side is low.
+The main contribution of this work will be to extend the formalism to include risk predictions for the contralateral side as well.
+We believe this holds the potential to reduce the electively irradiated volumes in the contralateral neck for patients whose individual risk for occult disease is low.
 In turn, this may substantially reduce the toxicity from irradiation and the associated decrease in quality of life.
 
-In the following, we begin by inspecting the dataset the model is built upon (@sec-data) and from it formulate some properties we require the bilateral model extension to have (@sec-requirements).
+The main contributions of this paper are as follows:
 
-After that, in @sec-unilateral, we briefly summarize the original model formalism before we extend it to the contralateral side starting in @sec-ext-to-contra.
+1.
+In section @sec-data we present a multi-centric dataset on lymph node involvement in 833 OPSCC patients.
+Based on the dataset the main risk factors for contralateral lymph node involvement are identified and the requirements for the bilateral model extension are described (@sec-requirements).
 
-Next, we describe how the model training and the computational experiments were set up in @sec-methods and present the results in @sec-results, followed by a closing discussion (@sec-discussion).
+2.
+In @sec-ext-to-contra we present a bilateral HMM of lymphatic progression that accounts for the T-stage lateralization of the primary tumor as well as clinical involvement as risk factors for occult contralateral involvement.
+In @sec-methods we describe how the model training and the computational experiments were set up.
+
+3.
+In the results in @sec-results, we demonstrate that the model is able to describe the patterns of contralateral lymph node involvement observed in the dataset, and we apply the model to estimating risk of occult disease for typical patients.
+In @sec-discussion, implications for volume-deescalated radiotherapy are discussed.
 
@@
 
-Originally, the only available data of detailed - i.e.
-per patient and per LNL - lymphatic involvement was a reconstructed table from the findings in @sanguineti_defining_2009, which included only ipsilateral nodal involvement for the LNLs I to IV.
-
-To be able to create models for lymphatic tumor progression that include more LNLs and the contralateral side, we have collected a detailed dataset of 833<!-- num patients --> patients with newly diagnosed oropharyngeal squamous cell carcinomas.
-It reports the involvement of every patient individually and per lymph node level in tabular form, in addition to other clinico-pathological information such as age, T-category, and HPV p16 status.
-
+To be able to create models for lymphatic tumor progression that includes all relevant LNLs including the contralateral side, we have collected a detailed dataset of 833<!-- num patients --> patients with newly diagnosed oropharyngeal squamous cell carcinomas.
+It reports the lymph node involvement per LNL of every patient individually in tabular form, in addition to other primary tumor and patient characteristics such as T-category, subside, and lateralization of the primary tumor, and HPV p16 status.
 Their patient records have been collected at four different institutions and a brief overview over some of their patients' characteristics are shown in @tbl-data-overview.
-Note that the data from the Inselspital Bern and the Centre Léon Bérard only consists of patients treated with some form of neck dissection.
+The data from the Inselspital Bern and the Centre Léon Bérard only consists of patients who received a neck dissection.
 Since this treatment is more commonly chosen for early T-category patients, they also make up a larger portion of the respective dataset.
+The majority of patients from the University Hospital Zürich and the Vall d'Hebron Barcelona Hospital was treated with definitive radiotherapy.
 
@@
 
-## Contralateral Involvement Prevalence {#sec-data-strat}
+## Patterns of Contralateral Involvement {#sec-data-strat}
 

### 2024-07-19 model.qmd
use beta over known data in figs

@@
 from scripts import paths, shared
-from scripts.marginalize_midext import possible_prevalence_combinations, SummedBetaPosterior
 
@@
       )
-      # content.append(
-      #   BetaPosterior.from_hdf5(
-      #     filename=paths.model_dir / "full" / "prevalences_overall.hdf5",
-      #     dataname=scenario.md5_hash("prevalences"),
-      #     color=color,
-      #   )
-      # )
-      combs = possible_prevalence_combinations(data, scenario)
       content.append(
-        SummedBetaPosterior(
-          num_success=combs.matches.values,
-          num_total=combs.total.values,
-          permutations=combs.permutations.values,
-          kwargs={"color": color},
+        BetaPosterior.from_hdf5(
+          filename=paths.model_dir / "full" / "prevalences_overall.hdf5",
+          dataname=scenario.md5_hash("prevalences"),
+          color=color,
         )
@@
 from scripts import paths, shared
-from scripts.marginalize_midext import (
-  possible_prevalence_combinations,
-  SummedBetaPosterior,
-)
 
@@
       )
-      # content.append(
-      #   BetaPosterior.from_hdf5(
-      #     filename=paths.model_dir / "full" / "prevalences_with_ipsi.hdf5",
-      #     dataname=scenario.md5_hash("prevalences"),
-      #     color=color,
-      #   )
-      # )
-      combs = possible_prevalence_combinations(data, scenario)
       content.append(
-        SummedBetaPosterior(
-          num_success=combs.matches.values,
-          num_total=combs.total.values,
-          permutations=combs.permutations.values,
-          kwargs={"color": color},
+        BetaPosterior.from_hdf5(
+          filename=paths.model_dir / "full" / "prevalences_with_ipsi.hdf5",
+          dataname=scenario.md5_hash("prevalences"),
+          color=color,
         )
@@
 ```
-
-# Illustrating the Marginalization over Beta PDFs
-
-```{python}
-import matplotlib.pyplot as plt
-import numpy as np
-from lyscripts.scenario import Scenario
-from lyscripts.plot.utils import draw, COLORS, BetaPosterior
-from lyscripts.compute.prevalences import compute_observed_prevalence
-from scripts import paths, shared, COL
-from scripts.marginalize_midext import (
-  possible_prevalence_combinations,
-  SummedBetaPosterior,
-)
-
-data = shared.get_data()
-
-scenario = Scenario(
-  t_stages=["early"],
-  midext=True,
-  diagnosis={"contra": {"max_llh": {"II": True}}},
-)
-combs = possible_prevalence_combinations(data, scenario)
-summed_beta = SummedBetaPosterior(
-  num_success=combs.matches.values,
-  num_total=combs.total.values,
-  permutations=combs.permutations.values,
-  kwargs={"label": "marginalized PDF", "color": COLORS["blue"]},
-)
-num_match, num_total = compute_observed_prevalence(data, scenario, mapping=None)
-naive_beta = BetaPosterior(
-  num_success=num_match,
-  num_total=num_total,
-  kwargs={"color": COLORS["green"], "label": "naive beta PDF"},
-)
-
-plt.rcParams.update(shared.get_fontsizes())
-plt.rcParams.update(shared.get_figsizes(width=17./2.))
-fig, ax = plt.subplots()
-draw(ax, contents=[naive_beta, summed_beta], xlims=(0., 50.))
-
-x = np.linspace(*ax.get_xlim(), 300)
-perm_sum = combs.permutations.sum()
-for _, row in combs.sort_values("permutations").iterrows():
-  beta = BetaPosterior(
-    num_success=row.matches,
-    num_total=row.total,
-  )
-  h, = ax.plot(
-    x, beta.pdf(x),
-    alpha=0.6 * (row.permutations / perm_sum)**0.5,
-    color=COLORS["red"],
-    zorder=1,
-  )
-
-ax.set_ylim(bottom=0.)
-handles, labels = ax.get_legend_handles_labels()
-ax.legend(handles=[*handles, h], labels=[*labels, "component beta PDFs"])
-ax.set_xlabel("prevalence [%]")
-ax.set_yticks([])
-
-plt.savefig(paths.figure_dir / "fig-model-marginal-pdf.svg")
-plt.show()
-```

### 2024-07-19 scripts/marginalize_midext.py
add note on unused code

@@ -1,5 +1,8 @@
-"""
-Marginalize over *observd* midline extension status using sums of beta posteriors.
+"""Plot observed and predicted prevalence of midline extension.
+
+Note: This module contains functions and classes for marginalizing over matching
+patterns where midline extension is unknown. We decided not to use this approach
+in the final analysis, but the code is kept here for reference.
 """
 
 import math

### 2024-07-19 manuscript.qmd

@@
 ```
-
-# Marginalizing Beta Posteriors over Unknowns {.appendix}
-
-```{python}
-#| echo: false
-#| label: tbl-data-unknowns
-#| tbl-cap: Combinations of midline extension status and state of contralateral LNL II for early T-category patients.
-import pandas as pd
-import numpy as np
-from scripts import shared, COL
-
-data = shared.get_data()
-is_early = data[COL.t_stage].isin([0,1,2])
-
-reduced = pd.DataFrame({
-  "Midline Extension": data.loc[is_early, COL.midext],
-  "contra LNL II": data.loc[is_early, ("max_llh", "contra", "II")]
-})
-
-result = (
-  reduced
-  .fillna("unknown")
-  .value_counts(dropna=False)
-  .unstack("Midline Extension")
-  .rename(
-    index={
-      False: "healthy",
-      True: "involved",
-    },
-    columns={
-      False: "no",
-      True: "yes",
-    },
-  )
-)
-result.loc["total"] = result.sum()
-result
-```
-
-In @sec-prevalence we explained that we compare the predicted prevalence of an involvement pattern under a scenario to a beta posterior over the "true" prevalence given the observations in the data.
-However, our data contains incomplete observations, especially regarding the tumor's midline extension.
-This information is not present for a large group of 85 patients, most of which treated at the Centre Léon Bérard in France.
-
-To account for this missing information and plot the correct posteriors over the data prevalence, we need to marginalize over the possible combinations of matching involvement patterns and selected patients under a scenario.
-For example, we have only 29<!-- TODO: compute dynamically --> early T-category patients with midline extension.
-But for an additional 65<!-- TODO: compute dynamically --> patients we do not know whether the tumor crossed the mid-sagittal plane.
-If we now want to compute the prevalence of contralateral LNL II invovlement for these early T-category patients with midline extension, then we need to account for the possibility that we are missing information.
-
-In @tbl-data-unknowns we show the combinations of the midline extension status and state of the contralateral LNL II.
-Using only available information, we would get to 6 out of 29<!-- TODO: compute dynamically --> patients for the prevalence of contralateral LNL II invovlement among the early T-category patients with midline extension.
-But the column with the "unknown" numbers suggests that the true prevalence could range from 6 out of 29 + 59 (7%)<!-- TODO: compute dynamically --> up to 12 out of 29 + 6 (34%)<!-- TODO: compute dynamically -->.
-We assume that every one of the possible combinations has the same a priori probability (uniform prior) but we do need to account for permutations via the binomial factor.
-In general, let us say we have $n$ patients with midline extension, of which $k$ harbor metastases in the contralateral LNL II.
-And we have $m$ patients with unknown midline extension, of which $\ell$ presented with diasease in contralateral LNL II, then the posterior distribution - marginalized over the unknown data - over the true prevalence $x \in \left[ 0, 1 \right]$ is given by:
-
-$$
-p(x \mid n, k, m, \ell, \alpha, \beta) = \left[ \frac{1}{m} \sum_{i=0}^m \frac{1}{\min(i,\ell)} \sum_{j=0}^{\min(i,\ell)} \frac{P(i, j \mid x )}{\binom{n}{k}} \right] \cdot P(n, k \mid x) \cdot \frac{1}{\operatorname{B}(\alpha, \beta)} x^\alpha (1-x)^\beta
-$$
-
-Note the two binomial likelihoods: one being marginalized over and one acting implicitly as a prior to the summed terms.
-This role of the likelihood over the data with known midline extension as a prior to the part where that data is missing corrects for the fact that there are highly unlikely combinations among those we marginalize over.
-E.g., expecting that exactly those 6 patients with contralateral LNL II involvemnt are also coincidentially the only ones of the 65 with a tumor crossing the mid-sagittal line is very unlikely.
-This is accounted for by the fact that this possibility has only one permutation and it is weakened by the likelihood over the data with known midline extension.
-With a uniform prior beta distribution ($\alpha=1$ and $\beta=1$), the above simplifies to:
-
-$$
-p(x \mid n, k, m, \ell) = \left[ \frac{1}{m} \sum_{i=0}^m \frac{1}{\min(i,\ell)} \sum_{j=0}^{\min(i,\ell)} \binom{n+i}{k+j} x^j (1 - x)^{i-j} \right] \cdot x^{k+1} (1 - x)^{n-k+1}
-$$
-
-We have plotted this marginal in @fig-model-marginal-pdf (blue curve) and the terms that make up the above sum (light red curves).
-Note that we weighted the plotting opacity with the number of permutations each of the possible combinations has.
-The marginal is compared against the "naive" beta posterior that only takes the data into account for which midline extension is available (green curve).
-
-![Posterior distributions over the data prevalence for contralateral LNL II involvement under the scenario of early T-category and midline extension.
-The green beta distribution shows this prevalence for those patients for which midline extension is known.
-While the blue PDF is a marginal distribution summed over all possible combinations, which are shown in shades of red.
-The more permutations a particular combination has, the more opaque it is plotted.](figures/fig-model-marginal-pdf.svg){#fig-model-marginal-pdf width="50%"}
-
-The distribution that marginalizes over the unknown midline extension cases is shifted towards a lower prevalence.
-This makes intuitive sense: With only 6 patients among the 65 with unknown midline extension status, the upper bound for the prevalence has moved down compared to 6 out of 29.
-
-We chose to compare our model against such corrected prevalence posteriors, because the model, too, is capable of marginalizing over unknown midline extension in the data.
-It would therefore be inaccurate to compare the model trained with the full dataset to only that part of the patient cohort where the tumor lateralization was recorded.

### 2024-07-22 model.qmd
adapt risk plot to new risks

@@
 t_stage_map = {"early": "early", "late": "adv."}
-midext_map = {False: "lateralized", True: "mid. ext."}
+midext_map = {False: "lateral", True: "mid-ext"}
 colors = [COLORS["green"], COLORS["blue"], COLORS["orange"], COLORS["red"]]
@@
 def get_label(scenario: Scenario) -> str:
+  lnl = list(eval(scenario.involvement)["contra"].keys()).pop()
+  d = eval(scenario.diagnosis)
+  has_upstream_fna = "FNA" in d["contra"]
+
+  if lnl == "IV":
+    return get_diag_label(d) + (" & c:III FNA+" if has_upstream_fna else "")
+
+  m = scenario.midext
+
+  if lnl == "III":
+    return f"{midext_map[m]}; {get_diag_label(d)}" + (" & c:II FNA+" if has_upstream_fna else "")
+
   t = scenario.t_stages[0]
-  m = scenario.midext
-  d = eval(scenario.diagnosis)
   return f"{t_stage_map[t]}; {midext_map[m]}; {get_diag_label(d)}"
@@
 for lnl, means in mean_lists.items():
-  indices[lnl] = np.argsort(means)
+  indices[lnl] = np.argsort(np.argsort(means))
 
@@
 with h5py.File(name=paths.model_dir / "full" / "risks.hdf5", mode="r") as h5file:
-  for dset in h5file.values():
+  for md5, dset in h5file.items():
     scenario = Scenario.from_dict(dict(dset.attrs))
@@
   "II": (0, 15),
-  "III": (0, 7),
-  "IV": (0, 5),
+  "III": (0, 9),
+  "IV": (0, 7),
 }
@@
   ax.set_xlabel("risk [%]")
-  ax.legend(labelspacing=0.1)
 
 axes[0].set_xticks(np.linspace(xlims["II"][0], xlims["II"][1], 6))
+axes[0].legend(labelspacing=0.1)
+axes[1].legend(
+  title="advanced T-category",
+  title_fontsize="small",
+  labelspacing=0.1,
+)
+axes[2].legend(
+  title="advanced T-category\n& midline extension",
+  title_fontsize="small",
+  labelspacing=0.1,
+)
 

### 2024-07-22 manuscript.qmd

@@
 ![Histograms over the predicted risk of occult contralateral LNL II (left), III (middle), and IV (right) invovlement, shown for some combinations of T-category, tumor lateralization, and clinical LNL diagnoses.
-All LNLs not explicitly mentioned in the legend, including the LNL for which the risk of occult disease was computed, where assumed to be clinically negative.](figures/fig-model-risks.svg){#fig-model-risks}
+All LNLs not explicitly mentioned in the legend, including the LNL for which the risk of occult disease was computed, where assumed to be clinically negative, based on CT imaging (specificity 76%, sensitivity 81%).](figures/fig-model-risks.svg){#fig-model-risks}
 
 @Fig-model-risks shows this predicted risk of occult disease in the contralateral LNLs II, III, and IV for interesting combinations of these three risk factors.
-This risk is computed *given* a CT diagnosis assessing the per-LNL clinical involvement for which we assumed a sensitivity of 81% and a specificity of 76% [@de_bondt_detection_2007].
+This risk is computed *given* a CT diagnosis assessing the per-LNL clinical involvement for which we assumed a sensitivity of 81% and a specificity of 76%.
+For the involvement in the LNls III and IV contralaterally, we also cosidered how the risk would be affected if the upstream LNL was confirmed to be metastatic using fine needle aspiration (FNA) with a specificity of 98% and a sensitivity of 80% [@de_bondt_detection_2007].
+In the figure legends, this is noted e.g.
+for the orange historgam over the risk of contralateral LNL IV involvement as `c:III FNA+`, meaning that the contralateral LNL III was pathologically confirmed to harbor metastases via FNA.
 
@@
 
-As shown in the center subplot of @fig-model-risks, the risk for occult disease in the contralateral LNL III only ever crosses a 5% threshold if the upstream level II is involved too, resulting in a risk of 5% to 6%<!-- TODO: make dynamic -->.
-As long as the contralateral neck appears clinically node negative, however, even an advanced T-category and midline extension do not yield a risk higher than 2%.
+As shown in the center subplot of @fig-model-risks, the risk for occult disease in the contralateral LNL III may only crosses a 5% threshold if the upstream level II is clinically involved, too.
+If the clinical diagnosis of the contralateral upstream level II is pathologically confirmed by FNA the risk increases by another percentage point<!-- TODO: make dynamic -->.
+When the tumor additionally crosses the mid-sagittal line, the risk may reach around 7%<!-- TODO: make dynamic -->.
 
@@
 
-Our model's prediction for the contralateral LNL IV show that even in the most extreme case, with an advanced T-category tumor extending over the mid-sagittal line as well as metastases in every ipsilateral LNL and every contralateral upstream level, the risk does not exceed 5%
+Our model's prediction for the contralateral LNL IV show that even in the most extreme case, with an advanced T-category tumor extending over the mid-sagittal line as well as metastases in every ipsilateral LNL and every contralateral upstream level, the risk stays just below 3%<!-- TODO: make dynamic -->.
+We only observe a risk of 5-6% when the involvement of the upstream level III is pathologically confirmed after taking a biopsy using FNA.
+This is because the model's prior probability for contralateral LNL IV involvement is so low that even given a clinical diagnosis in that level, a false positive in LNL III is a more likely explanation.
+This possibility, however, drops virtually to zero when tumor cells are pathologically confirmed in level III.
 
@@
 - LNL II when the tumor extends over the mid-sagittal line
-- LNL III when LNL II is clinically, the tumor is of advanced T-category, crosses the midline, and the ipsilateral neck shows involvement of at least LNL II
-- LNL IV only for severe ipsilateral involvement, advanced T-category with obvious midline extension and contralateral upstream involvement
+- LNL III when LNL II is clinically involved, may need confirmation via FNA
+- LNL IV only when upstream level III pathologically confirmed
 

### 2024-08-02 manuscript.qmd

@@
 
-To be able to create models for lymphatic tumor progression that includes all relevant LNLs including the contralateral side, we have collected a detailed dataset of 833<!-- num patients --> patients with newly diagnosed oropharyngeal squamous cell carcinomas.
+To be able to create models for lymphatic tumor progression that include all relevant LNLs including the contralateral side, we have collected a detailed dataset of 833<!-- num patients --> patients with newly diagnosed oropharyngeal squamous cell carcinomas.
 It reports the lymph node involvement per LNL of every patient individually in tabular form, in addition to other primary tumor and patient characteristics such as T-category, subside, and lateralization of the primary tumor, and HPV p16 status.
 Their patient records have been collected at four different institutions and a brief overview over some of their patients' characteristics are shown in @tbl-data-overview.
-The data from the Inselspital Bern and the Centre Léon Bérard only consists of patients who received a neck dissection.
+The data from the Inselspital Bern and the Centre Léon Bérard only consist of patients who received a neck dissection.
 Since this treatment is more commonly chosen for early T-category patients, they also make up a larger portion of the respective dataset.
@@
 The degree of ipsilateral involvement should give the model information on the time that may have passed between onset and diagnosis of the disease.
-This should come in addition to what can be inferred about this time from T-category alone.
+This should come in addition to what can be inferred about the time from T-category alone.
 3. A tumor that extends over the mid-sagittal line should yield contralateral metastases with much higher probability.
@@
 $$
-\ell = P \left( \mathbf{Z} = \boldsymbol{\zeta}_i \mid \text{T}x \right) = \sum_{t=0}^{t_\text{max}} \left[ \boldsymbol{\xi}_0 \cdot \mathbf{A} \cdot \mathbf{B} \right]_i \cdot P \left( t \mid \text{T}x \right)
+\ell = P \left( \mathbf{Z} = \boldsymbol{\zeta}_i \mid \text{T}x \right) = \sum_{t=0}^{t_\text{max}} \left[ \boldsymbol{\xi}_0 \cdot \mathbf{A}^t \cdot \mathbf{B} \right]_i \cdot P \left( t \mid \text{T}x \right)
 $$ {#eq-single-patient-llh}
@@
 For short chains, this estimate is not trustworthy, but stabilizes for longer chains.
-We therefore considered a sampling to be converged when two criteria where met:
+We therefore considered a sampling to be converged when two criteria were met:
 
@@
 
-As shown in the center subplot of @fig-model-risks, the risk for occult disease in the contralateral LNL III may only crosses a 5% threshold if the upstream level II is clinically involved, too.
+As shown in the center subplot of @fig-model-risks, the risk for occult disease in the contralateral LNL III may only cross a 5% threshold if the upstream level II is clinically involved, too.
 If the clinical diagnosis of the contralateral upstream level II is pathologically confirmed by FNA the risk increases by another percentage point<!-- TODO: make dynamic -->.

### 2024-08-06 manuscript.qmd
add prevs of cIII depending on cII
fixes #9

@@
 
+### Influence of Upstream Involvement
+
+![Predicted (histograms) vs observed (beta posteriors) prevalence of scnearios that show how much more often the contralateral LNL III is involved when the upstream level II is involved (red) against when it is healthy (green).
+As in @fig-model-prevalences-with-ipsi, this is plotted for all four combinations of T-category and midline extension.
+Note that the contralateral LNL III is extremely rarely involved without the upstream level II.](figures/fig-model-prevalences-with-upstream.svg){#fig-model-prevalences-with-upstream}
+
+Similarily to how we have shown the dependence of contralateral LNL II involvement on some extent of ipsilateral metastasis, we can look at the dependence of contralateral LNL III involvement depending on whether or not the upstream level harbors disease.
+In @fig-model-prevalences-with-upstream this is shown in analogy to @fig-model-prevalences-with-ipsi.
+Note that the prevalence of contralateral LNL III involvement is generally much lower than that of LNL II.
+Also, level III involvement is exceedingly rare when the upstream level II is not metastatic.
+The model captures this well.
+
 ### Prevalence of Midline Extension

### 2024-08-06 model.qmd
add prevs of cIII depending on cII
fixes #9

@@
 
+## Contra LNL III Depending on Upstream LNL II Involvement
+
+```{python}
+import matplotlib.pyplot as plt
+from matplotlib import gridspec
+import numpy as np
+from lyscripts.scenario import Scenario
+from lyscripts.plot.utils import Histogram, BetaPosterior, draw, COLORS, split_legends
+from scripts import paths, shared
+
+def turn_axis_off(axes):
+  axes.set_xticks([])
+  axes.set_yticks([])
+
+nrows, ncols = 2, 2
+
+plt.rcParams.update(shared.get_fontsizes())
+plt.rcParams.update(shared.get_figsizes(
+  nrows=nrows,
+  ncols=ncols,
+  aspect_ratio=2.,
+  width=17,
+))
+
+fig = plt.figure()
+gs = gridspec.GridSpec(nrows=nrows, ncols=ncols, figure=fig)
+axes = np.array([[fig.add_subplot(gs[i,j]) for j in range(ncols)] for i in range(nrows)])
+full = fig.add_subplot(gs[:,:], frame_on=False)
+turn_axis_off(full)
+
+full.set_title(
+  "Observed vs Predicted Prevalence of Contralateral LNL III Involvement\nDependent on Upstream LNL III Involvement",
+  pad=20.,
+  fontsize="large",
+)
+
+data = shared.get_data()
+
+for i, t_stage in enumerate(["early", "late"]):
+  axes[i,0].set_ylabel(f"{t_stage} T-cat.", fontweight="bold")
+  for j, midext in enumerate([False, True]):
+    axes[0,j].set_title("Mid. ext." if midext else "Lateralized", fontweight="bold")
+    axes[0,j].set_xticks([])
+    axes[-1,j].set_xlabel("prevalence [%]")
+    axes[i,j].set_yticks([])
+    content = []
+    for upstream, color in zip([False, True], [COLORS["green"], COLORS["red"]]):
+      scenario = Scenario(
+        t_stages=[t_stage],
+        midext=midext,
+        diagnosis={
+          "contra": {"max_llh": {"II": upstream, "III": True}},
+        }
+      )
+      content.append(
+        Histogram.from_hdf5(
+          filename=paths.model_dir / "full" / "prevalences_with_upstream.hdf5",
+          dataname=scenario.md5_hash("prevalences"),
+          color=color,
+          label="prediction",
+        )
+      )
+      content.append(
+        BetaPosterior.from_hdf5(
+          filename=paths.model_dir / "full" / "prevalences_with_upstream.hdf5",
+          dataname=scenario.md5_hash("prevalences"),
+          color=color,
+        )
+      )
+    draw(
+      axes=axes[i,j],
+      contents=content,
+      xlims=(0., 15. if midext else 4.),
+    )
+
+split_legends(
+  axes=axes[0,0],
+  titles=["upstream healthy", "upstream involved"],
+  locs=[(0.4, 0.99), (0.5, 0.55)],
+)
+split_legends(
+  axes=axes[0,1],
+  titles=["upstream healthy", "upstream involved"],
+  locs=[(0.3, 0.99), (0.6, 0.55)],
+)
+split_legends(
+  axes=axes[1,0],
+  titles=["upstream healthy", "upstream involved"],
+  locs=[(0.2, 0.99), (0.65, 0.99)],
+)
+split_legends(
+  axes=axes[1,1],
+  titles=["upstream healthy", "upstream involved"],
+  locs=[(0.3, 0.99), (0.65, 0.99)],
+)
+
+plt.savefig(paths.figure_dir / "fig-model-prevalences-with-upstream.svg")
+plt.show()
+```
+
 ## Portion of Midline Extension

### 2024-08-09 manuscript.qmd
add Jan's comments

--- 2024-08-02 manuscript.qmd
@@
 The majority of patients from the University Hospital Zürich and the Vall d'Hebron Barcelona Hospital was treated with definitive radiotherapy.
+
+Pathological involvement is thus available only for a subset of the patients and only for the dissected levels.
+For the remaining patients, only clinical involvement is available.
+For the analysis performed in this paper, all available diagnostic information has been combined into a consensus decision representing the most likely state of involvement for each patient and LNL, using literature values for sensitivity and specificity of different modalities [@de_bondt_detection_2007;@kyzas_18f-fluorodeoxyglucose_2008].
+In brief, pathology was considered the gold standard if available, overwriting possibly conflicting reports on clinical involvement.
+If pathology was not available, clinical involvement determined the consensus decision.
+For a subset of 83 patients in the CLB dataset, the lateralization of the primary tumor was not reported.
 
@@
 
+## Data Availability
+
+The entire data, including additional patients with tumors in other primary locations than the oropharynx, is publicly available: It may be [downloaded from LyProX](https://lyprox.org/patients/dataset) where it can be interactively explored too, [from GitHub](https://github.com/rmnldwg/lydata), [from zenodo](https://zenodo.org/search?q=lydata), or via the *Data-in-Brief* publications @ludwig_dataset_2022 and @ludwig_multi-centric_2023.
+The Data-in-Brief publications contain a detailed description of the datasets and the data format.
+However, these publications do not include the most recent dataset addition from Vall d'Hebron Barcelona Hospital.
+
 ## Patterns of Contralateral Involvement {#sec-data-strat}
@@
 This, too, may therefore be considered a surrogate for the duration of the disease.
-In rare cases, bulky nodal disease ipsilaterally may also redirect lymph fluids to the contralateral side.
+In addition, it has been hypothesized that bulky nodal disease ipsilaterally may also redirect lymph fluids to the contralateral side.
 <!-- citation needed?
@@
 Lastly, the right panel in @fig-data-strat shows that patients with a tumor crossing the mid-sagittal line show contralateral involvement vastly more often compared to patients with clearly lateralized tumors.
-This makes intutitive sense, because the lymphatic system in the head and neck region is typically symmetric and thus no major vessels cross the midline.
+This makes intuitive sense, because the lymphatic system in the head and neck region is typically symmetric and thus no major vessels cross the midline.
 Therefore, interstitial fluids from the primary tumor -- which we assume to carry living malignant cells -- may only reach the blind-ended lymphatic vessels in the contralateral neck via short-ranged diffusion.
@@
 -->
+
+The three risk factors for contralateral involvement, midline extension, T-category, and ipsilateral involvement are correlated.
+For example, midline extension was present in 48% of advanced T-category tumors but only in 7% of early T-category tumors.
+The higher prevalence of contralateral metastases in advanced T-category tumors may thus be partially explained by a higher rate of midline extension in these tumors.
+However, figure xxx indicates that advanced T-category and severe ipsilateral involvement do represent additional risk factors .
+The figure considers only patients with lateralized tumors not extending over the midline.
+For early T-category and healthy ipsilateral levels I-V, only 1% (1 out of 84 patients) shows involvement of contralateral level II.
+This increases to 5% (9 out of 165 patients) if ipsilateral level II is involved, to 15% (9 out of 60 patients) if ipsilateral levels II and III are involved, and to 23% (6 out of 26 patients) for advanced T-category with ipsilateral levels II and III involved.
 
@@
 
-## Data Availability
-
-The entire data, including additional patients with tumors in other primary locations than the oropharynx, is publicly available: It may be [downloaded from LyProX](https://lyprox.org/patients/dataset) where it can be interactively explored too, [from GitHub](https://github.com/rmnldwg/lydata), [from zenodo](https://zenodo.org/search?q=lydata), or via the *Data-in-Brief* publications @ludwig_dataset_2022 and @ludwig_multi-centric_2023.
-Although these publications do not include the most recent dataset addition from Vall d'Hebron Barcelona Hospital.
-
 # Unilateral Model for Lymphatic Progression {#sec-unilateral}
 
-![Directed acyclic graph (DAG) representing the abstract lymphatic network in the head and neck region.
-Blue nodes are the LNLs' hidden random variables, the red node represents the tumor, and the orange square nodes depict the binary observed variables.
-Red and blue arcs symbolize the probability of lymphatic spread along that edge during one time-step.
-The orange arcs represent the sensitivity and specificity of the observational modality (e.g.
-CT, MRI, pathology, ...).](static/small-graph.svg){#fig-small-graph width=40%}
+This paper builds on the previously developed unilateral model for ipsilateral lymph node involvement presented in [cite 2024 graph extension paper].
+In this section we provide a brief recap of the unilateral model to introduce the notation needed to extend the framework to a bilateral model describing both ipsilateral and contralateral lymph node involvement in @sec-ext-to-contra.
+For further details in the ipsilateral model, the reader is referred to the earlier publications [@ludwig_hidden_2021, cite 2024 graph extension paper]
 
+<!-- 
 Our first model to predict the lymphatic progression of HNSCC was introduced using Bayesian networks [@pouymayou_bayesian_2019].
@@
 We will briefly summarize this HMM's formalism before building on it to include the contralateral spread in @sec-ext-to-contra.
+-->
 
@@
 Here, $V$ is the number of LNLs the model considers.
-The values a LNL's hidden binary RV may take on are $X_v[t] = 0$ (`False`), meaning the LNL $v$ is healthy or free of metastatic disease, or $X_v[t] = 1$ (`True`), corresponding to some for of tumor presence (i.e., occult or clinical).
-
+The values a LNL's hidden binary random variables may take on are $X_v[t] = 0$ (`False`), meaning the LNL $v$ is healthy or free of metastatic disease, or $X_v[t] = 1$ (`True`), corresponding to some form of tumor presence (i.e., occult or clinically detected).
 Since the state vector $\mathbf{X}[t]$ is $V$-dimensional and binary, there are $2^V$ distinct possible lymphatic involvement patterns, which we enumerate from $\boldsymbol{\xi}_0 = \begin{pmatrix} 0 & 0 & \cdots & 0 \end{pmatrix}$ to $\boldsymbol{\xi}_{2^V} = \begin{pmatrix} 1 & 1 & \cdots & 1 \end{pmatrix}$.
+In addition, each LNL is associated with an observed binary random variable $Z_v$ that describes the clinical involvement of a LNL based on imaging: $Z_v = 0$ (`False`) indicates that the LNL $v$ is healthy based on clinical diagnosis, and $Z_v = 1$ (`True`) indicates that suspicious lymph nodes were detected that are deemed metastatic.
 
@@
 Note that we prohibit self-healing, meaning that during a transition, no LNL may change their state from $X_v[t]=1$ to $X_v[t+1]=0$.
-This effectively masks large parts of the transition matrix to be zero.
+Consequently, many elements of the transition matrix are zero.
 3. Lastly, the *observation matrix*
@@
 
-Note that the transition matrix $\mathbf{A}$ is parametrized using the spread probabilities of a directed acyclic graph (DAG) that we define as the underlying mechanistic representation of the lymphatic network.
-An example of such a DAG is shown in @fig-small-graph.
+The transition matrix $\mathbf{A}$ is parametrized using a directed acyclic graph (DAG) as an abstract representation of the underlying lymphatic network.
+Directed arcs from the primary tumor to a LNL are associated with a probability $b_v$ for direct spread to LNL $v$ during one time step.
+Directed arcs from a LNL $v$ to a LNL $r$ are associated with a probability $t_{vr}$ for the tumor to progress from one LNL to the next.
+In this paper, we build on the DAG shown in @fig-small-graph which was obtained by maximizing the model evidence as described in [cite 2024 graph extension paper].
+
+![Directed acyclic graph (DAG) representing the abstract lymphatic network in the head and neck region.
+Blue nodes are the LNLs' hidden random variables, the red node represents the tumor, and the orange square nodes depict the binary observed variables.
+Red and blue arcs symbolize the probability of lymphatic spread along that edge during one time-step.
+The orange arcs represent the sensitivity and specificity of the observational modality (e.g.
+CT, MRI, pathology, ...).](static/small-graph.svg){#fig-small-graph width=40%}
 
 Using the introduced quantities, we can evolve the distribution of all possible hidden states from $\mathbf{X}[t=0] = \boldsymbol{\xi}_0$ step by step, by successively multiplying this vector with the transition matrix $\mathbf{A}$.
-At the time of the diagnosis $t_D$, we multiply the result with the observation matrix $\mathbf{B}$.
-We may then look up the likelihood of a patient presenting with the diagnosis $\mathbf{Z}=\boldsymbol{\zeta}_i$ in the $i$-th entry of the final result.
-
-However, the remaining issue is that the value of $t_D$ is unknown, i.e.
-over how many time-steps the HMM should be evolved.
-We solve this problem by marginalizing over the time of diagnosis.
-Different distributions over the diagnosis times can then be choosen based on T-category.
-For instance, the mean of the time-prior to marginalize over the diagnosis time for early T-category patients $P\left( t_D \mid \text{early} \right)$ may be shifted towards earlier times than the one for advanced T-category patients $P\left( t_D \mid \text{early} \right)$.
-This gives us for example
-
-$$
-P\left( \mathbf{X} \mid \text{T}x = \text{early} \right) = \sum_{t=0}^{t_\text{max}} P \left( \mathbf{X} \mid t \right) \cdot P(t \mid \text{early})
-$$
-
 For later use, we define at this point the matrices of the dsitributions over all hidden states, given all time-steps $\boldsymbol{\Lambda}$:
@@
 
-Where the $k$-th row in this matrix corresponds to the distribution over hidden states after $t=k-1$ time-steps.
+Where the $k$-th row in this matrix corresponds to the probability distribution over hidden states after $t=k-1$ time-steps.
+
+At the time of the diagnosis $t_D$, we multiply the result with the observation matrix $\mathbf{B}$.
+We may then look up the likelihood of a patient presenting with the diagnosis $\mathbf{Z}=\boldsymbol{\zeta}_i$ in the $i$-th entry of the final result.
+However, the remaining issue is that the value of $t_D$ is unknown, i.e.
+over how many time-steps the HMM should be evolved.
+We solve this problem by marginalizing over the time of diagnosis.
+Different distributions over the diagnosis times can then be choosen based on T-category.
+For instance, the mean of the time-prior to marginalize over the diagnosis time for early T-category patients $P\left( t_D \mid \text{early} \right)$ may be shifted towards earlier times than the one for advanced T-category patients $P\left( t_D \mid \text{early} \right)$.
+This gives us for example
+
+$$
+P\left( \mathbf{X} \mid \text{T}x = \text{early} \right) = \sum_{t=0}^{t_\text{max}} P \left( \mathbf{X} \mid t \right) \cdot P(t \mid \text{early})
+$$
 
@@
 A naive approach to model the contralateral lymphatic spread would be to simply employ two independent unilateral models as introduced in @sec-unilateral.
-During training, one could even enforce some shared parameters between these two models, like the parameterization of the distributions over diagnose times or the spread among the LNLs.
-Additionally, we could think of an approach to incorporate the primary tumor's mid-sagittal extension as a risk factor.
+During training, one could enforce that some parameters are shared between these two models, e.g.
+the parameterization of the distributions over diagnose times or the spread among the LNLs ($t_{vr}).
+However, this approach lacks a way to describe the correlation between ipsi- and contralateral involvement discussed in section {#sec-data-strat}.
+Two independent unilateral models would not describe the observation that contralateral involvement becomes more likely with more severe ipsilateral involvement.
 
-However, this approach lacks a way to describe the correlation between ipsi- and contralateral involvement.
-This is displayed in @tbl-data-strat and shows how often the contralateral LNLs I, II, III, and IV were involved, given all possible combinations of midline extension, T-category, and ipsilateral LNL III involvement.
+<!-- This is displayed in @tbl-data-strat and shows how often the contralateral LNLs I, II, III, and IV were involved, given all possible combinations of midline extension, T-category, and ipsilateral LNL III involvement.
 Unsurprisingly, the prevalence for contralateral involvement is consistently higher when the tumor extends over the mid-sagittal line or is of later T-category.
 But it is also more frequent when the ipsilateral side shows more severe involvement, which is here shown via the surrogate LNL III.
+-->
 
-Thus, we attempt to extend the formalism in @sec-unilateral in such a way that the model's ipsi- and contralateral side evolve synchronously.
+<!-- Additionally, we could think of an approach to incorporate the primary tumor's mid-sagittal extension as a risk factor. -->
+
+Thus, we extend the formalism in @sec-unilateral in such a way that the model's ipsi- and contralateral side evolve synchronously over time.
 To achieve that, we start by writing down the posterior distribution of involvement an analogy to @eq-uni-bayes-law, which is now a joint probability of an involvement $\mathbf{X}^\text{i}$ ipsilaterally *and* an involvement $\mathbf{X}^\text{c}$ contralaterally, given a diagnosis of the ipsilateral LNLs $\mathbf{Z}^\text{i}$ and of the contralateral ones $\mathbf{Z}^\text{c}$:
@@
 
-The likelihood of the diagnoses given a hidden state simply factorise: $P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \mid \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right) = P \left( \mathbf{Z}^\text{i} \mid \mathbf{X}^\text{i} \right) \cdot P \left( \mathbf{Z}^\text{c} \mid \mathbf{X}^\text{c} \right)$.
-And the two factors are contained in the observation matrices $\mathbf{B}^\text{i}$ and $\mathbf{B}^\text{c}$.
+The probability of the diagnoses given a hidden state factorises: $P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \mid \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right) = P \left( \mathbf{Z}^\text{i} \mid \mathbf{X}^\text{i} \right) \cdot P \left( \mathbf{Z}^\text{c} \mid \mathbf{X}^\text{c} \right)$, and the two factors are described through observation matrices $\mathbf{B}^\text{i}$ and $\mathbf{B}^\text{c}$.
 
 The term representing the model's prior probability of hidden involvement does not factorize.
-However, if we assume that lymphatic spread typically does not cross the mid-sagittal line, we can write it as a factorising sum:
+However, we assume that there is no direct lymph drainage from an ipsilateral LNL to a contralateral LNL.
+No major lymph vessels cross the mid-sagittal plane.
+In the graphical model, this means that we assume no directed arcs between ipsilateral and contralateral LNLs, i.e.
+tumor spread to the contralateral side is due to the primary tumor alone.
+We can thus write the joint probability $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right)$ as a factorising sum:
 
@@
 
-This assumption makes intuitive sense: The two sides of the lymphatic network in a typical patient are approximately mirror images of each other.
-Thus, no major vessels cross the mid-sagittal line.
-There may, however, be diffusion of lymph fluid accross this line or bulky involvement that redirects lymphatic drainage significantly.
+This assumption makes intuitive sense: The state of the ipsilateral and contralateral sides of the lymphatic network evolves independently over time as there are no major lymph vessels crossing the midline.
+However, both sides are coupled via time.
+Qualitatively speaking, according to the model in {#eq-bilateral-marginal}, a joint state with severe contralateral involvement and limited ipsilateral involvement would be unlikely, because severe contralateral involvement becomes likely only at later time steps when limited ipsilateral involvement becomes unlikely.
 
-Using this assumption along with @eq-lambda-matrix, we can write the above distribution algebraically as a product:
+<!--There may, however, be diffusion of lymph fluid accross this line or bulky involvement that redirects lymphatic drainage significantly.-->
+
+Using {#eq-bilateral-marginal} along with @eq-lambda-matrix, we can write the above distribution algebraically as a product:
 
@@
 I.e., the ipsi- and contralateral spread rates are entirely different.
-However, using two sensible assumptions, we can reduce the parameter space by sharing some parameters between the sides:
+However, using three sensible assumptions, we can reduce the parameter space by sharing some parameters between the sides:
 
-1.
+1. We assume that both ipsilateral and contralateral spread is described through the same graph shown in {#fig-small-graph}.
+2.
 We assume the spread *among* the LNLs to be same on both sides.
@@
    for all $v \leq V$ and $r \in \operatorname{pa}(v)$.
-2.
+3.
 The tumor's spread to the contralateral side in case of an extension over the midline is larger than if it was clearly lateralized, but smaller than its spread to the ipsilateral side.
@@
 If a tumor extends over the mid-sagittal line, its contralateral spread rate can be expected to be in between these two extremes.
-We encode this in a *mixing parameter* $\alpha$ that captures a "degree of asymmetry": \
+We encode this in a *mixing parameter* $\alpha \in \[0,1\]$ that captures a "degree of asymmetry": \
    $$
@@
 which is less than double in size, compared to the unilateral model.
-From these parameters, we now build three different transition matrices: The unchanged $\mathbf{A}_\text{i}$ for the ipsilateral side, one $\mathbf{A}_\text{c}^{\epsilon=\texttt{False}}$ for the contralateral side that covers the progression as long as the tumor is lateralized, and $\mathbf{A}_\text{c}^{\epsilon=\texttt{True}}$ in case of a tumor that has crossed the mid-sagittal line.
+From these parameters, we now build three different transition matrices: The unchanged matrix $\mathbf{A}_\text{i}$ for the ipsilateral side, one $\mathbf{A}_\text{c}^{\epsilon=\texttt{False}}$ for the contralateral side that covers the progression as long as the tumor is lateralized, and $\mathbf{A}_\text{c}^{\epsilon=\texttt{True}}$ in case of a tumor that has crossed the mid-sagittal line.
 
@@
 
-To account for the increased prevalence of involvement on the contralateral side when the tumor is not clearly lateralized anymore, we also model the tumor's extension over the mid-sagittal line as a binary random variable.
-It starts lateralized and at every time-step there is a finite probability $p_\epsilon$ that the tumor grows over the symmetry plane of the patient.
-
+It can be assumed that most tumors that cross the midline at the time of diagnose have started as lateralized tumors and grew over the midline at a later time step.
+Hence, the transition matrix $\mathbf{A}_\text{c}^{\epsilon=\texttt{True}}$ only applies for a subset of the time steps.
+Therefore, we also model the tumor's extension over the mid-sagittal line as a binary random variable.
+A tumor starts lateralized and at every time-step there is a finite probability $p_\epsilon$ that the tumor grows over the midsaggital plane.
 The overall probabilities to find a patient with a clearly lateralized tumor or one that extends over the mid-sagittal line after $t$ time-steps are then respectively given by
@@
 $$
-
-In @fig-model-midext-evo, we visualize how the prior distribution over diagnose times $P(t)$, the conditional probability of midline extension $P(\epsilon \mid t)$, and their joint $P(\epsilon, t)$ evolve over the course of a patient evolution.
-
-![The top panel shows the prior probability to get diagnosed at time-step $t$ for early and late T-category tumors as bars.
-Also in the top panel, we plot the conditional probability of the tumor's midline extension ($\epsilon=\texttt{True}$), given the time-step $t$ as a line plot.
-In the bottom panel, we show the joint probability of getting diagnosed in time-step $t$ *and* having a tumor that crosses the midline.](figures/fig-model-midext-evo.svg){#fig-model-midext-evo width="50%"}
 
@@
 The case when midline extension is eventually present is more complicated: We already marginalize over the exact time-step when the tumor grows over the mid-sagittal line.
-But whenever that happens, we also need to change the contralateral transition matrix to use the increased spread rates $b_v^{\text{c}, \epsilon=\texttt{True}}$ from the tumor to the contralateral LNLs, given by the linear mix in @eq-mixing.
-We can achieve the correct marginalization by iteratively building the joint $P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid t \right)$.
-We start from $t=0$, where we know that all LNLs are healthy - i.e., the contralateral neck is in the state $\mathbf{X}_\text{c}=\boldsymbol{\xi}_0$ - and the tumor is lateralized ($\epsilon=\texttt{False}$):
+But whenever that happens, we also need to change the contralateral transition matrix to use the increased spread rates $b_v^{\text{c}, \epsilon=\texttt{True}}$ from the tumor to the contralateral LNLs, given by the linear mixing in @eq-mixing.
+We can achieve the correct marginalization by iteratively building the joint distribution $P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid t \right)$.
+We start from $t=0$, where we know that all LNLs are healthy (the contralateral neck is in the state $\mathbf{X}_\text{c}=\boldsymbol{\xi}_0$) and the tumor is lateralized ($\epsilon=\texttt{False}$):
 
 $$
-P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{False} \mid t=0 \right) = \boldsymbol{\xi}_0
+P \left( \mathbf{X}^\text{c} = \boldsymbol{\xi}_0, \epsilon=\texttt{False} \mid t=0 \right) = 1
 $$
 
-Then we consider an arbitrary later time-step $t=\tau+1$. There are two possible scenarios we need to marginalize over:
+whereas all other states have zero probability. Then we consider an arbitrary later time-step $t=\tau+1$. There are two possible scenarios we need to marginalize over:
 
@@
 
-In @fig-model-state-dist we plot the state distribution for a full midline model as two separate heatmaps.
-To keep it manageable to interpret, this example model only considers the LNLs II, III, and IV ipsi- and contralaterally, resulting in $2 \times 2^3 \times 2^3 = 128$ distinct states.
-
-![3D state distribution of a midline model with 3 LNLs (II, III, and IV) in both sides of the neck.
-Note the higher probability for contralateral involvement in the right subplot, where the tumor extends over the mid-sagittal line.](figures/fig-model-state-dist.svg){#fig-model-state-dist}
-
-# Methods {#sec-methods}
+# Computational methods {#sec-methods}
 
@@
 [`lymph-model`]: https://lymph-model.readthedocs.io/en/stable/
+
+First, in @fig-model-burnin-history, we verify the sampling converged successfully by inspecting two monitoring quantities: The autocorrelation time of the MCMC chain and the acceptance fractions of the parallel walkers.
+
+![Monitoring quantities during the burn-in phase of the parameter sampling.
+Left: The autocorrelation time of the sampling chain estimated at different sampling steps.
+We consider the chain converged when the estimate of the autocorrelation time is stable and drops below the trust threshold of $n/50$ where $n$ is the number of steps.
+Right: Fraction of accepted MCMC proposals averaged over all parallel walkers.
+Values around 30% indicate good mixing of the walkers.](figures/fig-model-burnin-history.svg){#fig-model-burnin-history}
 
@@
 
-# Results {#sec-results}
-
-First, in @fig-model-burnin-history, we verify the sampling converged successfully by inspecting two monitoring quantities: The autocorrelation time of the MCMC chain and the acceptance fractions of the parallel walkers.
-
-![Monitoring quantities during the burn-in phase of the parameter sampling.
-Left: The autocorrelation time of the sampling chain estimated at different sampling steps.
-We consider the chain converged when the estimate of the autocorrelation time is stable and drops below the trust threshold of $n/50$ where $n$ is the number of steps.
-Right: Fraction of accepted MCMC proposals averaged over all parallel walkers.
-Values around 30% indicate good mixing of the walkers.](figures/fig-model-burnin-history.svg){#fig-model-burnin-history}
+# Results: Model evaluation {#sec-results}
 
 In @tbl-midline-params, we tabulate the mean and standard deviation of the sampled parameters for the full midline model.
+Considering the parameters of the bilateral model that describe the ipsilateral spread ($b^i_v$, $t_{vr}$ and $p_{adv}$, we note that the bilateral model mostly reproduces the parameter values reported in the earlier publication [] for the unilateral model.
+The small differences may be mostly due to the differences in training dataset.
+Therefore, the analysis of the models capability to adequately describe ipsilateral lymph node involvement is not repeated here and we focus instead on the analysis of contralateral involvement.
+We can further note that the contralateral spread parameters $b^c_v$ are small compared to the ipsilateral parameters $b^i_v$, reflecting the low prevalence of contralateral lymph node involvement for lateralized tumors.
 
@@
 ```
+## Illustration of the model
+In this subsection we illustrate parts of the mathematical framework defined in the previous section.
+In @fig-model-midext-evo (top panel), we visualize the prior distribution over diagnose times $P(t)$.
+By parameter choice, early T-category tumors are diagnosed on average after 3 time steps.
+Advanced T-category tumors are diagnosed on average after 4.5 time steps ($p_{adv}=0.45)$, a parameter learned from the data to describe the higher involvement for advanced T-category tumors.
+The probability per time step for the tumor to grow over the midline, $p_\epsilon = 0.0816$ yields the probability of midline extension $P(\epsilon \mid t)$ at a given time step $t$ (red line).
+The parameter is adjusted such that the model yields the correct proportion of patients with tumors extending over the midline seen in the dataset.
+The bottom panel in @fig-model-midext-evo shows the joint probability $P(\epsilon, t)$ to be diagnosed at time step $t$ with a given state of midline extension and T-category.
 
-## Prevalence Predictions
+![The top panel shows the prior probability to get diagnosed at time-step $t$ for early and late T-category tumors as bars.
+Also in the top panel, we plot the conditional probability of the tumor's midline extension ($\epsilon=\texttt{True}$), given the time-step $t$ as a line plot.
+In the bottom panel, we show the joint probability of getting diagnosed in time-step $t$ *and* having a tumor that crosses the midline.](figures/fig-model-midext-evo.svg){#fig-model-midext-evo width="50%"}
 
-We want to investigate whether and to what extent the model can fulfill the requirements laid out in @sec-requirements.
-To that end, we compare its predictions for contralateral involvement against observations in the data.
+The model introduced in the previous section defines a model of the joint probability distribution over midline extension and ipsi- and contralateral lymph node involvement, $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right, \epsilon)$, which is illustrated in @fig-model-state-dist.
+To keep it manageable to interpret, this example only considers the LNLs II, III, and IV ipsi- and contralaterally, resulting in $2 \times 2^3 \times 2^3 = 128$ distinct states.
+LNLs I, V, and VII and the associated spread parameters have been removed from the model while the remaining parameters are set to the mean values in table xxx.
+@fig-model-state-dist shows the joint distribution as two separate heatmaps for the two midline extension states for early/advanced???
+T-category tumors.
+The most likely state is a lateralized tumor with involvement of ipsilateral level II and no contralateral involvement, which has a probability of approximately 25%.
+The second most probable state is a lateralized tumor with ipsilateral involvement of levels II and III and no contralateral involvement.
+The most likely state with contralateral involvement are tumors with midline extension with involvement of contralateral level II and ipsilateral level II and III.
+
+![3D state distribution of a midline model with 3 LNLs (II, III, and IV) in both sides of the neck.
+Note the higher probability for contralateral involvement in the right subplot, where the tumor extends over the mid-sagittal line.](figures/fig-model-state-dist.svg){#fig-model-state-dist}
+
+## Prevalence predictions for contralateral involvement
+
+The bilateral model was designed to fulfil the requirements laid out in @sec-requirements.
+Now, we investigate to what extent the model can quantitatively describe the patterns of lymph node involvement observed in the dataset.
+To that end, we compare the model's predictions for contralateral involvement against observations in the data.
 This is done given scenarios that differ in T-category and/or midline extension and/or ipsilateral involvement.
@@
 The left column depicts scenarios where the primary tumor is clearly lateralized, the right column scenarios of tumors extending over the mid-sagittal line.
-This figure illustrates the model's ability to describe the prevalence of different combinations of scenarios involving the risk factors T-category and midline extension.](figures/fig-model-prevalences-overall.svg){#fig-model-prevalences-overall}
+This figure illustrates the model's ability to describe the prevalence of involvement for different combinations of the risk factors T-category and midline extension.](figures/fig-model-prevalences-overall.svg){#fig-model-prevalences-overall}
 
-@Fig-model-prevalences-overall shows nicely how the model is capable of accurately taking the most important risk factors, i.e.
-T-category and midline extension, into account.
-As observed in the data, the model predicts that the prevalence of contralateral LNL II involvement jumps from below 8% <!-- TODO: make this dynamic --> for early T-category lateralized tumors to almost 40% <!-- TODO: make this dynamic --> when the tumor is of advanced T-category and crosses the mid-sagittal line.
+@Fig-model-prevalences-overall shows nicely that the model is capable of accurately accounting for the most important risk factors, i.e.
+T-category and midline extension.
+As observed in the data, the model predicts that the prevalence of contralateral LNL II involvement increases from below 8% <!--  TODO: make this dynamic --> for early T-category lateralized tumors to almost 40% <!-- TODO: make this dynamic --> when the tumor is of advanced T-category and crosses the mid-sagittal line.
+Similarly, the prevalence of contralateral LNL III involvement increases from around 2% <!--  TODO: make this dynamic --> for early T-category lateralized tumors to almost 15% <!-- TODO: make this dynamic --> when the tumor is of advanced T-category and crosses the mid-sagittal line.
 
-However, for early T-category scenarios with midline extension, the model does seem to overestimate contralateral LNL II and III invovlement.
+<!--  However, for early T-category scenarios with midline extension, the model does seem to overestimate contralateral LNL II and III invovlement.
 This likely stems from the the small sample size of this relatively rare scenario as hinted at by the wide beta posteriors.
+-->
 
@@
 
-## Prediction of Risk for Occult Disease
+# Results: Prediction of Risk for Occult Disease
 
-In this section, we investigate how the model may be applied clinically: We want to estimate the risk for occult disease in some or all LNLs, given the patient's individual diagnosis.
-In terms of our model, this diagnosis consists of the T-category, the lateralization of the tumor (does it extend over the mid-sagittal line?), and which LNLs are clinically involved, e.g.
-because some lymph nodes appear enlarged in an MRI scan or show increased glycolytic activity on an FDG PET/CT scan.
+For a clinical application of the model we are interested in the risk of occult metastases rather than the observed patterns of lymph node involvement.
+We want to estimate the risk for occult disease in clinically negative LNLs, given the patient's individual diagnosis.
+In terms of our model, the diagnosis consists of the T-category, the lateralization of the tumor (if it extend over the mid-sagittal plane), and which LNLs are clinically involved based on imaging and possibly fine needle aspiration (FNA).
+For the results shown in this section, we assume that involvement of a LNL is clinically diagnosed through imaging with a sensitivity of 81% and a specificity of 76% [@de_bondt_detection_2007].
+For FNA we assume a specificity of 98% and a sensitivity of 80%.
+This amounts to the assumption that lymph node involvement confirmed by FNA is almost certainly true involvement, i.e.
+there are almost no false positive diagnoses.
 
@@
 
-@Fig-model-risks shows this predicted risk of occult disease in the contralateral LNLs II, III, and IV for interesting combinations of these three risk factors.
+<!-- @Fig-model-risks shows this predicted risk of occult disease in the contralateral LNLs II, III, and IV for interesting combinations of these three risk factors.
 This risk is computed *given* a CT diagnosis assessing the per-LNL clinical involvement for which we assumed a sensitivity of 81% and a specificity of 76%.
@@
 In the figure legends, this is noted e.g.
-for the orange historgam over the risk of contralateral LNL IV involvement as `c:III FNA+`, meaning that the contralateral LNL III was pathologically confirmed to harbor metastases via FNA.
+for the orange historgam over the risk of contralateral LNL IV involvement as `c:III FNA+`, meaning that the contralateral LNL III was pathologically confirmed to harbor metastases via FNA.-->
 
@@
 
-The variable impacting the prediction for contralateral level II involvement in our model is the tumor's lateralization.
-For example, a patient with a clearly lateralized early T-category tumor and a clinically N0 neck is assigned a 1-2% <!-- TODO: make this dynamic --> risk for occult disease in contralateral LNL II.
-Under the same scenario, but *with* mid-sagittal extension, the risk jumps to almost 7%.
+@Fig-model-risks (left panel) shows the predicted risk of occult disease in contralateral LNL II.
+The most important variable impacting the prediction for contralateral level II involvement in our model is the tumor's lateralization.
+A patient with a clearly lateralized early T-category tumor and a clinically N0 neck is predicted to have a 1-2% <!-- TODO: make this dynamic, and include this scenario in the plot --> risk for occult disease in contralateral LNL II.
+For a patient with a clearly lateralized early T-category tumor but *with* mid-sagittal extension, the risk increases to almost 7%.
 <!-- TODO: make this dynamic -->
 
-T-category plays a lesser role: Considering the scenario of a tumor that crosses the mid-sagittal line and an ipsilateral neck where at least LNL III is clinically involved, the risk for occult contralateral LNL II disease is around 8.5%.
+Also advanced T-category increases the risk of occult disease but plays a lesser role: Considering the scenario of a tumor that crosses the midline and an ipsilateral neck where at least LNL III is clinically involved, the risk for occult contralateral LNL II disease is around 8.5%.
 <!-- TODO: make this dynamic --> For the same scenario and an advanced T-category tumor, the risk increases to 11%.

### 2024-08-09 manuscript.qmd

@@
 If a tumor extends over the mid-sagittal line, its contralateral spread rate can be expected to be in between these two extremes.
-We encode this in a *mixing parameter* $\alpha \in \[0,1\]$ that captures a "degree of asymmetry": \
+We encode this in a *mixing parameter* $\alpha \in [0,1]$ that captures a "degree of asymmetry": \
    $$
@@
 In @tbl-midline-params, we tabulate the mean and standard deviation of the sampled parameters for the full midline model.
-Considering the parameters of the bilateral model that describe the ipsilateral spread ($b^i_v$, $t_{vr}$ and $p_{adv}$, we note that the bilateral model mostly reproduces the parameter values reported in the earlier publication [] for the unilateral model.
+Considering the parameters of the bilateral model that describe the ipsilateral spread ($b^i_v$, $t_{vr}$ and $p_\text{adv.}$), we note that the bilateral model mostly reproduces the parameter values reported in the earlier publication [@ludwig_modelling_2023] for the unilateral model.
 The small differences may be mostly due to the differences in training dataset.
@@
 ```
+
 ## Illustration of the model
+
 In this subsection we illustrate parts of the mathematical framework defined in the previous section.
@@
 By parameter choice, early T-category tumors are diagnosed on average after 3 time steps.
-Advanced T-category tumors are diagnosed on average after 4.5 time steps ($p_{adv}=0.45)$, a parameter learned from the data to describe the higher involvement for advanced T-category tumors.
-The probability per time step for the tumor to grow over the midline, $p_\epsilon = 0.0816$ yields the probability of midline extension $P(\epsilon \mid t)$ at a given time step $t$ (red line).
+Advanced T-category tumors are diagnosed on average after 4.5 time steps ($p_\text{adv.}=0.45$) <!-- TODO: make dynamic -->, a parameter learned from the data to describe the higher involvement for advanced T-category tumors.
+The probability per time step for the tumor to grow over the midline, $p_\epsilon = 0.0816$ <!-- TODO: make dynamic --> yields the probability of midline extension $P(\epsilon \mid t)$ at a given time step $t$ (red line).
 The parameter is adjusted such that the model yields the correct proportion of patients with tumors extending over the midline seen in the dataset.
@@
 
-The model introduced in the previous section defines a model of the joint probability distribution over midline extension and ipsi- and contralateral lymph node involvement, $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right, \epsilon)$, which is illustrated in @fig-model-state-dist.
+The model introduced in the previous section defines a model of the joint probability distribution over midline extension and ipsi- and contralateral lymph node involvement, $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)$, which is illustrated in @fig-model-state-dist.
 To keep it manageable to interpret, this example only considers the LNLs II, III, and IV ipsi- and contralaterally, resulting in $2 \times 2^3 \times 2^3 = 128$ distinct states.
@@
 
-### Contralateral LNL II
+## Contralateral LNL II
 
@@
 
-### Contralalteral LNL III
+## Contralalteral LNL III
 
@@
 
-### Contralateral LNL IV
+## Contralateral LNL IV
 

### 2024-08-14 scripts/model_prevalences_upstream.py
pull out scripts from supporting `.qmd`

@@ -1,0 +1,112 @@
+"""Predicted/observed prevalences of contra III involvement, dep. on upstream level."""
+
+import matplotlib.pyplot as plt
+import numpy as np
+import paths
+import shared
+from matplotlib import gridspec
+
+from lyscripts.plot.utils import COLORS, BetaPosterior, Histogram, draw, split_legends
+from lyscripts.scenario import Scenario
+
+
+def main():
+    """Plot figure."""
+    nrows, ncols = 2, 2
+
+    plt.rcParams.update(shared.get_fontsizes())
+    plt.rcParams.update(
+        shared.get_figsizes(
+            nrows=nrows,
+            ncols=ncols,
+            aspect_ratio=2.0,
+            width=17,
+        )
+    )
+
+    fig = plt.figure()
+    gs = gridspec.GridSpec(nrows=nrows, ncols=ncols, figure=fig)
+    axes = np.array(
+        [[fig.add_subplot(gs[i, j]) for j in range(ncols)] for i in range(nrows)]
+    )
+    full = fig.add_subplot(gs[:, :], frame_on=False)
+    shared.turn_axis_off(full)
+
+    full.set_title(
+        (
+            "Observed vs Predicted Prevalence of Contralateral LNL III Involvement"
+            "\nDependent on Upstream LNL III Involvement"
+        ),
+        pad=20.0,
+        fontsize="large",
+    )
+
+    for i, t_stage in enumerate(["early", "late"]):
+        axes[i, 0].set_ylabel(f"{t_stage} T-cat.", fontweight="bold")
+        for j, midext in enumerate([False, True]):
+            axes[0, j].set_title(
+                "Mid. ext." if midext else "Lateralized", fontweight="bold"
+            )
+            axes[0, j].set_xticks([])
+            axes[-1, j].set_xlabel("prevalence [%]")
+            axes[i, j].set_yticks([])
+            content = []
+            for upstream, color in zip([False, True], [COLORS["green"], COLORS["red"]]):
+                scenario = Scenario(
+                    t_stages=[t_stage],
+                    midext=midext,
+                    diagnosis={
+                        "contra": {"max_llh": {"II": upstream, "III": True}},
+                    },
+                )
+                content.append(
+                    Histogram.from_hdf5(
+                        filename=paths.model_dir
+                        / "full"
+                        / "prevalences_with_upstream.hdf5",
+                        dataname=scenario.md5_hash("prevalences"),
+                        color=color,
[…]

### 2024-08-16 manuscript.qmd

@@
 
-![Contralateral involvement stratified by T-category (left panel), the number of metastatic LNLs ipsilaterally (center panel), and whether the primary tumor extended over the mid-sagittal line or was clearly lateralized (right panel).](figures/fig-data-strat.svg){#fig-data-strat}
+![Contralateral involvement stratified by T-category (left panel), the number of metastatic LNLs ipsilaterally (center panel), and whether the primary tumor extended over the mid-sagittal line or was clearly lateralized (right panel).](figures/fig_data_strat.svg){#fig-data-strat}
 
@@
 Right: Fraction of accepted MCMC proposals averaged over all parallel walkers.
-Values around 30% indicate good mixing of the walkers.](figures/fig-model-burnin-history.svg){#fig-model-burnin-history}
+Values around 30% indicate good mixing of the walkers.](figures/fig_model_burnin_history.svg){#fig-model-burnin-history}
 
@@
 Also in the top panel, we plot the conditional probability of the tumor's midline extension ($\epsilon=\texttt{True}$), given the time-step $t$ as a line plot.
-In the bottom panel, we show the joint probability of getting diagnosed in time-step $t$ *and* having a tumor that crosses the midline.](figures/fig-model-midext-evo.svg){#fig-model-midext-evo width="50%"}
+In the bottom panel, we show the joint probability of getting diagnosed in time-step $t$ *and* having a tumor that crosses the midline.](figures/fig_model_midext_evo.svg){#fig-model-midext-evo width="50%"}
 
@@
 ![3D state distribution of a midline model with 3 LNLs (II, III, and IV) in both sides of the neck.
-Note the higher probability for contralateral involvement in the right subplot, where the tumor extends over the mid-sagittal line.](figures/fig-model-state-dist.svg){#fig-model-state-dist}
+Note the higher probability for contralateral involvement in the right subplot, where the tumor extends over the mid-sagittal line.](figures/fig_model_state_dist.svg){#fig-model-state-dist}
 
@@
 The left column depicts scenarios where the primary tumor is clearly lateralized, the right column scenarios of tumors extending over the mid-sagittal line.
-This figure illustrates the model's ability to describe the prevalence of involvement for different combinations of the risk factors T-category and midline extension.](figures/fig-model-prevalences-overall.svg){#fig-model-prevalences-overall}
+This figure illustrates the model's ability to describe the prevalence of involvement for different combinations of the risk factors T-category and midline extension.](figures/fig_model_prevalences_overall.svg){#fig-model-prevalences-overall}
 
@@
 We show two scenarios: One where the ipsilateral neck shows no involvement (at least LNLs I to V are healthy, LNL VII was ignored because data on it is missing for some patients) in blue and one where at least LNL III was involved in orange.
-These two scenarios are plotted for all combinations of T-category (early in top row, advanced in bottom row) and tumor lateralization (lateralized in left column, extending over mid-sagittal line in right column).](figures/fig-model-prevalences-with-ipsi.svg){#fig-model-prevalences-with-ipsi}
+These two scenarios are plotted for all combinations of T-category (early in top row, advanced in bottom row) and tumor lateralization (lateralized in left column, extending over mid-sagittal line in right column).](figures/fig_model_prevalences_with_ipsi.svg){#fig-model-prevalences-with-ipsi}
 
@@
 While the prevalence is predicted correctly when marginalizing over T-category, the model cannot capture the degree of separation observed in the data.
-Since the tumor's midline extension is virtually always part of the diagnosis and hence *given* when predicting a patient's risk, we do not consider this discrepancy a major issue.](figures/fig-model-prevalences-midext.svg){#fig-model-prevalences-midext width="50%"}
+Since the tumor's midline extension is virtually always part of the diagnosis and hence *given* when predicting a patient's risk, we do not consider this discrepancy a major issue.](figures/fig_model_prevalences_midext.svg){#fig-model-prevalences-midext width="50%"}
 
@@
 ![Histograms over the predicted risk of occult contralateral LNL II (left), III (middle), and IV (right) invovlement, shown for some combinations of T-category, tumor lateralization, and clinical LNL diagnoses.
-All LNLs not explicitly mentioned in the legend, including the LNL for which the risk of occult disease was computed, where assumed to be clinically negative, based on CT imaging (specificity 76%, sensitivity 81%).](figures/fig-model-risks.svg){#fig-model-risks}
+All LNLs not explicitly mentioned in the legend, including the LNL for which the risk of occult disease was computed, where assumed to be clinically negative, based on CT imaging (specificity 76%, sensitivity 81%).](figures/fig_model_risks.svg){#fig-model-risks}
 

### 2024-08-16 manuscript.qmd
plot joint state dist as algebra equation
Related: #11

@@
 
-The model introduced in the previous section defines a model of the joint probability distribution over midline extension and ipsi- and contralateral lymph node involvement, $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)$, which is illustrated in @fig-model-state-dist.
-To keep it manageable to interpret, this example only considers the LNLs II, III, and IV ipsi- and contralaterally, resulting in $2 \times 2^3 \times 2^3 = 128$ distinct states.
+The model introduced in the previous section defines a model of the joint probability distribution over midline extension and ipsi- and contralateral lymph node involvement, $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)$.
+How this quantity is computed is illustrated in @fig-model-state-dist, visually representing @eq-bilateral-marginal-algebra.
+To keep it manageable to interpret, this example only considers the LNLs II, III, and IV ipsi- and contralaterally, resulting in $2^3 = 16$ distinct states per side and thus $2 \times 8 \times 8 = 128$ total states.
 LNLs I, V, and VII and the associated spread parameters have been removed from the model while the remaining parameters are set to the mean values in table xxx.
@@
 
-![3D state distribution of a midline model with 3 LNLs (II, III, and IV) in both sides of the neck.
-Note the higher probability for contralateral involvement in the right subplot, where the tumor extends over the mid-sagittal line.](figures/fig_model_state_dist.svg){#fig-model-state-dist}
+![Visual representation of @eq-bilateral-marginal-algebra for the case of midline extention.
+The left and right matrix represent the evolution of the possible hidden states for the ipsi- and contralateral neck respectively.
+In particular, the right matrix describes the evolution of the contralateral hidden states *and* midline extension.
+In the center, the time-prior is plotted as a diagonal matrix.
+The result of this matrix computation is the joint distribution $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\texttt{True} \right)$](figures/fig_model_state_dist.svg){#fig-model-state-dist}
 

### 2024-08-16 scripts/model_state_dist.py

@@ -1,62 +1,93 @@
-"""Plot the bilateral state distribution of the model."""
+"""Plot the term P(X_i | t)^T @ P(t) @ P(X_c | t) for the simple model."""
 
+import operator
+from matplotlib.gridspec import GridSpec
 import matplotlib.pyplot as plt
 import numpy as np
-import paths
 import shared
 from matplotlib.colors import LinearSegmentedColormap
-from mpl_toolkits.axes_grid1 import AxesGrid
 
 from lyscripts.plot.utils import COLORS
 
 
 def main():
     """Plot figure."""
-    model = shared.get_model(which="simple", load_samples=True)
-    state_dist = 100 * model.state_dist(t_stage="late")
-    vmax = np.max(state_dist)
+    # use the full model's parameters in the smaller, simple model
+    smpl_model = shared.get_model(which="simple", load_samples=False)
+    full_model = shared.get_model(which="full", load_samples=True)
+    smpl_model.set_params(**full_model.get_params())
 
-    nrows, ncols = 1, 2
-    cmap = LinearSegmentedColormap.from_list(
-        name="usz",
-        colors=[COLORS["green"], COLORS["orange"], COLORS["red"]],
+    ipsi_state_dist_evo = smpl_model.ext.ipsi.state_dist_evo() * 100.0
+    _, contra_state_dist_evo = smpl_model.contra_state_dist_evo()
+    contra_state_dist_evo *= 100.0
+    time_prior = np.diag(smpl_model.get_distribution("late").pmf) * 100.0
+
+    vmin = np.min(
+        [
+            ipsi_state_dist_evo.min(),
+            time_prior.min(),
+            contra_state_dist_evo.min(),
+        ]
     )
-    cmap.set_under(COLORS["gray"])
+    vmax = np.max(
+        [
+            ipsi_state_dist_evo.max(),
+            time_prior.max(),
+            contra_state_dist_evo.max(),
+        ]
+    )
 
+    nrows, ncols = 1, 3
     plt.rcParams.update(shared.get_fontsizes())
     plt.rcParams.update(
         shared.get_figsizes(
             nrows=nrows,
             ncols=ncols,
-            aspect_ratio=1.0,
+            aspect_ratio=0.7675,
[…]

### 2024-08-16 manuscript.qmd
use full graph for fig. 2
Related: #11

@@
 Directed arcs from a LNL $v$ to a LNL $r$ are associated with a probability $t_{vr}$ for the tumor to progress from one LNL to the next.
-In this paper, we build on the DAG shown in @fig-small-graph which was obtained by maximizing the model evidence as described in [cite 2024 graph extension paper].
+In this paper, we build on the DAG shown in @fig-full-graph which was obtained by maximizing the model evidence as described in [cite 2024 graph extension paper].
 
@@
 The orange arcs represent the sensitivity and specificity of the observational modality (e.g.
-CT, MRI, pathology, ...).](static/small-graph.svg){#fig-small-graph width=40%}
+CT, MRI, pathology, ...).](static/full-graph.svg){#fig-full-graph width=40%}
 
@@
 
-The single-patient likelihood $\ell$ in @eq-single-patient-llh depends on the spread parameters shown in @fig-small-graph via the transition matrix $\mathbf{A}$ and on the binomial parameters $p_{\text{T}x}$ via time-priors.
+The single-patient likelihood $\ell$ in @eq-single-patient-llh depends on the spread parameters shown in @fig-full-graph via the transition matrix $\mathbf{A}$ and on the binomial parameters $p_{\text{T}x}$ via time-priors.
 In this work, we will only differentiate between "early" (T1 & T2) and "advanced" (T3 & T4) T-categories.
@@
 
-1. We assume that both ipsilateral and contralateral spread is described through the same graph shown in {#fig-small-graph}.
+1. We assume that both ipsilateral and contralateral spread is described through the same graph shown in {#fig-full-graph}.
 2.

### 2024-08-19 manuscript.qmd
add references to DiB papers
Related #11

@@
 
-To be able to create models for lymphatic tumor progression that include all relevant LNLs including the contralateral side, we have collected a detailed dataset of 833<!-- num patients --> patients with newly diagnosed oropharyngeal squamous cell carcinomas.
+To be able to create models for lymphatic tumor progression that include all relevant LNLs including the contralateral side, we have collected a detailed dataset of 833<!-- num patients --> patients with newly diagnosed oropharyngeal squamous cell carcinomas [@ludwig_dataset_2022;@ludwig_multi-centric_2023].
 It reports the lymph node involvement per LNL of every patient individually in tabular form, in addition to other primary tumor and patient characteristics such as T-category, subside, and lateralization of the primary tumor, and HPV p16 status.

### 2024-08-19 manuscript.qmd
move and expand consensus section
Related: #11

@@
 The majority of patients from the University Hospital Zürich and the Vall d'Hebron Barcelona Hospital was treated with definitive radiotherapy.
-
-Pathological involvement is thus available only for a subset of the patients and only for the dissected levels.
-For the remaining patients, only clinical involvement is available.
-For the analysis performed in this paper, all available diagnostic information has been combined into a consensus decision representing the most likely state of involvement for each patient and LNL, using literature values for sensitivity and specificity of different modalities [@de_bondt_detection_2007;@kyzas_18f-fluorodeoxyglucose_2008].
-In brief, pathology was considered the gold standard if available, overwriting possibly conflicting reports on clinical involvement.
-If pathology was not available, clinical involvement determined the consensus decision.
 For a subset of 83 patients in the CLB dataset, the lateralization of the primary tumor was not reported.
@@
 ```
+
+## Consensus on Involvement Status
+
+Pathological involvement is thus available only for a subset of the patients and only for the dissected levels.
+For the remaining patients, only clinical involvement is available.
+For the analysis performed in this paper, all available diagnostic information has been combined into a consensus decision representing the most likely state of involvement for each patient and LNL, using literature values for sensitivity and specificity of different modalities [@de_bondt_detection_2007;@kyzas_18f-fluorodeoxyglucose_2008].
+These values are tabulated in @tbl-spec-sens.
+
+This consensus was formed as follows: Suppose the involvement status $X_v$ of LNL $v$ was assessed using different diagnostic modalities $\mathcal{O} = \{ \text{MRI}, \text{CT}, \text{pathology}, \ldots \}$, each characterized by their own pair of sensitivity and specificity values $s_N^\mathcal{o}$ and $s_P^\mathcal{o}$, with $\mathcal{o} \in \mathcal{O}$.
+Then we have $|\mathcal{O}|$ observations $z_v^\mathcal{o} \in \left[ 0, 1 \right]$, where 0 stands for "healthy" and 1 for "involved".
+We can then compute the most likely true involvement $X_v$ using the likelihood function
+
+$$
+\begin{aligned}
+\ell \left( X_v \mid \{ z_v^\mathcal{o} \}_{\mathcal{o} \in \mathcal{O}} \right) = \prod_{\mathcal{o} \in \mathcal{O}}
+\left( 1 - X_v \right) \cdot &\left[ z_v^\mathcal{o} \cdot \left( 1 - s_P^\mathcal{o} \right) + \left( 1 - z_v^\mathcal{o} \right) \cdot s_P^\mathcal{o} \right] \\
++ X_v \cdot &\left[ z_v^\mathcal{o} \cdot s_N^\mathcal{o} + \left( 1 - z_v^\mathcal{o} \right) \cdot (1 - s_N^\mathcal{o}) \right]
+\end{aligned}
+$$
+
+We now assume the true state $X_v$ to take on the value 1 if $\ell \left( X_v = 1 \mid \ldots \right) > \ell \left( X_v = 0 \mid \ldots \right)$ and 0 otherwise.
+For example, if we have $z_\text{II}^\text{CT} = 0$ and $z_\text{II}^\text{MRI} = 1$ we would compute the following likelihoods:
+
+$$
+\begin{aligned}
+\ell \left( X_\text{II} = 1 \mid z_\text{II}^\text{CT} = 0, z_\text{II}^\text{MRI} = 1 \right) &= \left( 1 - s_N^\text{CT} \right) \cdot s_N^\text{MRI} = 15.39\% \\
+\ell \left( X_\text{II} = 0 \mid z_\text{II}^\text{CT} = 0, z_\text{II}^\text{MRI} = 1 \right) &= s_P^\text{CT} \cdot \left(1 - s_N^\text{MRI}\right) = 14.44\%
+\end{aligned}
+$$
+
+In this example, we would thus assume the true state to be involved ($X_\text{II} = 1$).
+
+This method of computing a consensus also ensures that the pathology reports always override any conflicting clinical diagnosis, due to pathology's high sensitivity and specificity.
+
+| Modality  | Specificity | Sensitivity |
+| :-------- | ----------: | ----------: |
+| CT        |         76% |         81% |
+| PET       |         86% |         79% |
+| MRI       |         63% |         81% |
+| FNA       |         98% |         80% |
+| pathology |        100% |        100% |
+
+: Specificity and sensitivity values from the literature [@de_bondt_detection_2007;@kyzas_18f-fluorodeoxyglucose_2008]. {#tbl-spec-sens}
 

### 2024-08-19 manuscript.qmd
add fig for lateral tumors
Related: #11

@@
 The higher prevalence of contralateral metastases in advanced T-category tumors may thus be partially explained by a higher rate of midline extension in these tumors.
-However, figure xxx indicates that advanced T-category and severe ipsilateral involvement do represent additional risk factors .
+However, @fig-data-strat-uncorr indicates that advanced T-category and severe ipsilateral involvement do represent additional risk factors .
 The figure considers only patients with lateralized tumors not extending over the midline.
 For early T-category and healthy ipsilateral levels I-V, only 1% (1 out of 84 patients) shows involvement of contralateral level II.
-This increases to 5% (9 out of 165 patients) if ipsilateral level II is involved, to 15% (9 out of 60 patients) if ipsilateral levels II and III are involved, and to 23% (6 out of 26 patients) for advanced T-category with ipsilateral levels II and III involved.
+This increases to 5% (9 out of 165 patients) if ipsilateral level II is involved, to 15% (9 out of 60 patients) if ipsilateral levels II and III are involved, and to 23% (6 out of 26 patients) for advanced T-category with ipsilateral levels II and III involved.<!-- TODO: make values here dynamic.
+-->
+
+![Contralateral involvement prevalence by LNL for lateralized tumors.
+Shown are four different scenarios: 1) early T-category with an ipsilateral N0 neck (green), 2) early T-category with ipsilaterally LNL II involved (blue), 3) late T-category with ipsilateral LNL II metastatic (orange), and 4) early T-category with ipsilateral both levels II and III involved (red).
+This illustrates that even when controlling for the tumor's lateralization, T-category and ipsilaeral involvement are predictive of contralateral metastases.](figures/fig_data_strat_uncorr.svg){#fig-data-strat-uncorr}
 

### 2024-08-19 scripts/data_strat_uncorr.py

@@ -1,0 +1,83 @@
+"""Plot how T-stage and ipsi inv. predict contra involv. when controlling for midext."""
+
+import numpy as np
+import paths
+import shared
+from matplotlib import pyplot as plt
+from shared import COL, CONTRA_LNLS
+
+from lyscripts import utils
+from lyscripts.plot.utils import COLORS
+
+
+def main():
+    """Plot figure."""
+    nrows, ncols = 1, 1
+    plt.rcParams.update(**shared.get_fontsizes())
+    plt.rcParams.update(
+        **shared.get_figsizes(
+            nrows=nrows,
+            ncols=ncols,
+            width=17/2,
+        )
+    )
+    _fig, ax = plt.subplots()
+
+    raw = utils.load_patient_data(paths.data)
+    lateral = raw[raw[COL.midext] == False]
+
+    early = lateral[lateral[COL.t_stage] <= 2]
+    early_n0 = early[early["max_llh", "ipsi"].sum(axis=1) == 0]
+    early_II = early[early["max_llh", "ipsi", "II"] == True]
+    early_II_and_III = early_II[early_II["max_llh", "ipsi", "III"] == True]
+
+    late = lateral[lateral[COL.t_stage] > 2]
+    late_II = late[late["max_llh", "ipsi", "II"] == True]
+
+    early_n0 = early_n0[CONTRA_LNLS]["max_llh", "contra"].mean(axis=0)
+    early_II = early_II[CONTRA_LNLS]["max_llh", "contra"].mean(axis=0)
+    early_II_and_III = early_II_and_III[CONTRA_LNLS]["max_llh", "contra"].mean(axis=0)
+    late_II = late_II[CONTRA_LNLS]["max_llh", "contra"].mean(axis=0)
+
+    pos = np.arange(len(early_n0))
+    kwargs = {"width": 0.6, "zorder": 2}
+
+    plt.bar(
+        x=pos+0.2,
+        height=100 * early_II_and_III,
+        color=COLORS["red"],
+        label="early; ipsi II and III",
+        **kwargs,
+    )
+    plt.bar(
+        x=pos+0.1,
+        height=100 * late_II,
+        color=COLORS["orange"],
+        label="late; ipsi II",
+        **kwargs,
+    )
+    plt.bar(
+        x=pos,
+        height=100 * early_II,
+        color=COLORS["blue"],
+        label="early; ipsi II",
+        **kwargs,
+    )
+    plt.bar(
[…]

### 2024-08-19 scripts/shared.py

@@ -1,5 +1,4 @@
-"""Some shared and reused functions for the scripts.
-"""
+"""Some shared and reused functions for the scripts."""
 
 from collections import namedtuple
 from pathlib import Path
@@ -10,7 +9,6 @@ import pandas as pd
 from lymph import types
 
 from lyscripts import utils
-
 
 try:
     from . import paths

### 2024-08-19 manuscript.qmd
trim trailing whitespace
rerun and publish

@@
 
-This consensus was formed as follows: Suppose the involvement status $X_v$ of LNL $v$ was assessed using different diagnostic modalities $\mathcal{O} = \{ \text{MRI}, \text{CT}, \text{pathology}, \ldots \}$, each characterized by their own pair of sensitivity and specificity values $s_N^\mathcal{o}$ and $s_P^\mathcal{o}$, with $\mathcal{o} \in \mathcal{O}$.
-Then we have $|\mathcal{O}|$ observations $z_v^\mathcal{o} \in \left[ 0, 1 \right]$, where 0 stands for "healthy" and 1 for "involved".
+This consensus was formed as follows: Suppose the involvement status $X_v$ of LNL $v$ was assessed using different diagnostic modalities $\mathcal{O} = \{ \text{MRI}, \text{CT}, \text{pathology}, \ldots \}$, each characterized by their own pair of sensitivity and specificity values $s_N^{\mathcal{o}}$ and $s_P^{\mathcal{o}}$, with $\mathcal{o} \in \mathcal{O}$.
+Then we have $|\mathcal{O}|$ observations $z_v^{\mathcal{o}} \in \left[ 0, 1 \right]$, where 0 stands for "healthy" and 1 for "involved".
 We can then compute the most likely true involvement $X_v$ using the likelihood function
@@
 \begin{aligned}
-\ell \left( X_v \mid \{ z_v^\mathcal{o} \}_{\mathcal{o} \in \mathcal{O}} \right) = \prod_{\mathcal{o} \in \mathcal{O}}
-\left( 1 - X_v \right) \cdot &\left[ z_v^\mathcal{o} \cdot \left( 1 - s_P^\mathcal{o} \right) + \left( 1 - z_v^\mathcal{o} \right) \cdot s_P^\mathcal{o} \right] \\
-+ X_v \cdot &\left[ z_v^\mathcal{o} \cdot s_N^\mathcal{o} + \left( 1 - z_v^\mathcal{o} \right) \cdot (1 - s_N^\mathcal{o}) \right]
+\ell \left( X_v \mid \{ z_v^{\mathcal{o}} \}_{\mathcal{o} \in \mathcal{O}} \right) = \prod_{\mathcal{o} \in \mathcal{O}}
+\left( 1 - X_v \right) \cdot &\left[ z_v^{\mathcal{o}} \cdot \left( 1 - s_P^{\mathcal{o}} \right) + \left( 1 - z_v^{\mathcal{o}} \right) \cdot s_P^{\mathcal{o}} \right] \\
++ X_v \cdot &\left[ z_v^{\mathcal{o}} \cdot s_N^{\mathcal{o}} + \left( 1 - z_v^{\mathcal{o}} \right) \cdot (1 - s_N^{\mathcal{o}}) \right]
 \end{aligned}

### 2024-08-20 manuscript.qmd
describe llh in Bayes' context for bilateral
Related: #11

@@
 
-## Model Prediction in the Bayesian Context
-
-Our stated goal is to compute the risk for a patient's true nodal involvement state $\mathbf{X}$, *given* their individual diagnosis $d = \left( \boldsymbol{\zeta}_k, \text{T}x \right)$.
-Using Bayes' law, this is written as:
-
-$$
-P \big( \mathbf{X} \mid \mathbf{Z}=\boldsymbol{\zeta}_k, \boldsymbol{\hat{\theta}}, \text{T}x \big) = \frac{P \left( \boldsymbol{\zeta}_k \mid \mathbf{X} \right) P \big( \mathbf{X} \mid \boldsymbol{\hat{\theta}}, \text{T}x \big)}{\sum_{i=0}^{2^V} P \left( \boldsymbol{\zeta}_k \mid \mathbf{X}=\boldsymbol{\xi}_i \right) P \big( \mathbf{X}=\boldsymbol{\xi}_i \mid \boldsymbol{\hat{\theta}}, \text{T}x \big)}
-$$ {#eq-uni-bayes-law}
-
-The term $P \left( \boldsymbol{\zeta}_k \mid \mathbf{X} \right)$ is defined solely by sensitivity and specificity of the diagnostic modality.
-Terms like this already appeared in the definition of the observation matrx in @eq-obs-matrix.
-The *prior* $P \big( \mathbf{X} \mid \boldsymbol{\hat{\theta}} \big)$ in the above equation is the crucial term that is supplied by a trained model and its parameters $\boldsymbol{\hat{\theta}}$.
-
-It is possible to compute this *posterior* probability of true involvement not only for one fully defined state $\mathbf{X}$, but also for e.g.
-individual LNLs: For example, the risk for involvement in level IV would be a marginalization over all states $\boldsymbol{\xi}_i$, where $\xi_{i4}=1$.
-Formally:
-
-$$
-P \big( \text{IV} \mid \mathbf{Z}=\boldsymbol{\zeta}_k, \boldsymbol{\hat{\theta}}, \text{T}x  \big) = \sum_{k \, : \, \xi_{k4}=1} P \big( \mathbf{X} = \boldsymbol{\xi}_k \mid \boldsymbol{\zeta}_k, \boldsymbol{\hat{\theta}}, \text{T}x  \big)
-$$ {#eq-marg-over-posterior}
-
 # Extension to a Bilateral Model {#sec-ext-to-contra}
@@
 $$ {#eq-ext-param-space}
+
+## Model Prediction in the Bayesian Context
+
+Our stated goal is to compute the risk for a patient's true ipsi- and contralateral nodal involvement states $\mathbf{X}^\text{i}$ and $\mathbf{X}^\text{c}$, *given* their individual diagnosis $d = \left( \boldsymbol{\zeta}^\text{i}_k, \boldsymbol{\zeta}^\text{c}_\ell, \epsilon, \text{T}x \right)$.
+Here, this diagnosis consists of the observed ipsi- and contralateral nodal involvements, the patient's midline extension $\epsilon$, and their tumor's T-category $\text{T}x$.
+Using Bayes' law, we can write this risk as:
+
+$$
+P \big( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \mid d, \boldsymbol{\hat{\theta}} \big)
+= \frac{P \left( \boldsymbol{\zeta}^\text{i}_k \mid \mathbf{X}^\text{i} \right) P \left( \boldsymbol{\zeta}^\text{c}_\ell \mid \mathbf{X}^\text{c} \right) P \big( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \mid \boldsymbol{\hat{\theta}}, \text{T}x \big)}
+{\sum_{i=0}^{2^V} \sum_{j=0}^{2^V} P \left( \boldsymbol{\zeta}^\text{i}_k \mid \mathbf{X}^\text{i}=\boldsymbol{\xi}^\text{i}_i \right) P \big( \boldsymbol{\zeta}^\text{c}_\ell \mid \mathbf{X}^\text{c}=\boldsymbol{\xi}^\text{c}_j \big) P \big( \mathbf{X}^\text{i}=\boldsymbol{\xi}^\text{i}_i, \mathbf{X}^\text{c}=\boldsymbol{\xi}^\text{c}_j, \epsilon \mid \boldsymbol{\hat{\theta}}, \text{T}x \big)}
+$$ {#eq-uni-bayes-law}
+
+The terms $P \left( \boldsymbol{\zeta}^\text{i}_k \mid \mathbf{X}^\text{i} \right)$ and $P \left( \boldsymbol{\zeta}^\text{c}_\ell \mid \mathbf{X}^\text{c} \right)$ are defined solely by sensitivity and specificity of the diagnostic modality.
+Terms like these already appeared in the definition of the observation matrx in @eq-obs-matrix.
+The *prior* $P \big( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \mid \boldsymbol{\hat{\theta}}, \text{T}x \big)$ in the above equation is the crucial term that is supplied by a trained model and its parameters $\boldsymbol{\hat{\theta}}$.
+
+It is possible to compute this *posterior* probability of true involvement not only for one fully defined state $(\mathbf{X}^\text{i}, \mathbf{X}^\text{c})$, but also for e.g.
+individual LNLs: For example, the risk for involvement in the contralateral level IV would be a marginalization over all ipsilateral states $\boldsymbol{\xi}^\text{i}_i$ and all contralateral states $\boldsymbol{\xi}^\text{c}_j$ where $\xi^\text{c}_{j4}=1$.
+Formally:
+
+$$
+P \big( \text{cIV} \mid \mathbf{Z}^\text{i}=\boldsymbol{\zeta}^\text{i}_k, \mathbf{Z}^\text{c}=\boldsymbol{\zeta}^\text{c}_\ell, \boldsymbol{\hat{\theta}}, \text{T}x \big) = \sum_k \sum_{\ell \, : \, \xi_{\ell 4}=1} P \big( \mathbf{X}^\text{i} = \boldsymbol{\xi}^\text{i}_k, \mathbf{X}^\text{c} = \boldsymbol{\xi}^\text{c}_\ell \mid \boldsymbol{\zeta}^\text{i}_k, \boldsymbol{\zeta}^\text{c}_\ell, \epsilon, \boldsymbol{\hat{\theta}}, \text{T}x \big)
+$$ {#eq-marg-over-posterior}
 

### 2024-08-20 manuscript.qmd
add upstream prevalence plot

@@
 
+### Influence of Upstream Involvement on Contralateral Metastasis
+
+![The influence of the upstream LNL II's involvement on the prevalence of contralateral level III for the four combinations of tumor lateralization (lateralized or extending over midline) and T-ctageory (early or advanced).
+Our model predictions (histograms) are plotted against the observations in the data (beta posteriors).](figures/fig_model_prevalences_upstream.svg){#fig-model-prevalences-upstream}
+
+@Fig-model-prevalences-upstream illustrates how rarely the contralateral LNL III harbors metastases when its upstream level II is healthy.
+Also, the model can capture this correlation rather well, with the exception of early T-category tumors that extend over the mid-sagittal line.
+As in @fig-model-prevalences-overall and @fig-model-prevalences-with-ipsi, these cases are relatively rare resulting in a broad distribution over the true prevalence.
+
 ### Prevalence of Midline Extension

### 2024-08-20 scripts/model_prevalences_upstream.py
add scenario to risks
Related: #11

@@ -35,7 +35,7 @@ def main():
     full.set_title(
         (
             "Observed vs Predicted Prevalence of Contralateral LNL III Involvement"
-            "\nDependent on Upstream LNL III Involvement"
+            "\nDependent on Upstream LNL II Involvement"
         ),
         pad=20.0,
         fontsize="large",

### 2024-08-20 scripts/model_risks.py

@@ -73,7 +73,7 @@ def main():
 
     contents = {"II": [], "III": [], "IV": []}
     mean_lists = {"II": [], "III": [], "IV": []}
-    colors = [COLORS["green"], COLORS["blue"], COLORS["orange"], COLORS["red"]]
+    colors = [COLORS["green"], COLORS["blue"], COLORS["orange"], COLORS["red"], "#6821ab"]
 
     with h5py.File(name=paths.model_dir / "full" / "risks.hdf5", mode="r") as h5file:
         for dset in h5file.values():

### 2024-08-21 scripts/model_prevalences_with_ipsi.py
add scenario to prevalences with ipsi
Related: #11

@@ -54,9 +54,10 @@ def main():
             for ipsi, color in zip(
                 [
                     {"I": False, "II": False, "III": False, "IV": False, "V": False},
-                    {"III": True},
+                    {"II": True},
+                    {"II": True, "III": True},
                 ],
-                [COLORS["blue"], COLORS["orange"]],
+                [COLORS["green"], COLORS["orange"], COLORS["red"]],
             ):
                 scenario = Scenario(
                     t_stages=[t_stage],
@@ -88,29 +89,29 @@ def main():
             draw(
                 axes=axes[i, j],
                 contents=content,
-                xlims=(0.0, 30.0 if midext else 10.0),
+                xlims=(0.0, 40.0 if midext else 14.0),
             )
 
     # manually place legends
     split_legends(
         axes=axes[0, 0],
-        titles=["ipsi LNLs I-V healthy", "ipsi LNL III"],
-        locs=[(0.3, 0.99), (0.5, 0.55)],
+        titles=["ipsi LNLs I-V healthy", "ipsi LNL II", "ipsi LNLs II & III"],
+        locs=[(0.2, 0.99), (0.5, 0.6), (0.55, 0.99)],
     )
     split_legends(
         axes=axes[0, 1],
-        titles=["ipsi LNLs I-V healthy", "ipsi LNL III"],
-        locs=[(0.3, 0.99), (0.5, 0.55)],
+        titles=["ipsi LNLs I-V healthy", "ipsi LNL II", "ipsi LNLs II & III"],
+        locs=[(0.15, 0.99), (0.7, 0.6), (0.5, 0.99)],
     )
     split_legends(
         axes=axes[1, 0],
-        titles=["ipsi LNLs I-V healthy", "ipsi LNL III"],
-        locs=[(0.16, 0.99), (0.53, 0.55)],
+        titles=["ipsi LNLs I-V healthy", "ipsi LNL II", "ipsi LNLs II & III"],
+        locs=[(0.1, 0.99), (0.65, 0.99), (0.35, 0.6)],
     )
     split_legends(
         axes=axes[1, 1],
-        titles=["ipsi LNLs I-V healthy", "ipsi LNL III"],
-        locs=[(0.13, 0.55), (0.5, 0.99)],
+        titles=["ipsi LNLs I-V healthy", "ipsi LNL II", "ipsi LNLs II & III"],
+        locs=[(0.1, 0.99), (0.65, 0.6), (0.45, 0.99)],
     )
 
     plt.savefig(shared.get_figure_path(__file__))

### 2024-08-23 manuscript.qmd

@@
 
-## Consensus on Involvement Status
+## Consensus on Involvement Status {#sec-data-consensus}
 
@@
 
-These datasets allow us to investigate correlations between the involvement of individual LNLs, or between risk factors and patterns of involvement.
-In @fig-data-strat, we have plotted the prevalence of each contralateral LNL's involvement, stratified by T-category, ipsilateral number of involved LNLs, and whether the tumor extended over the mid-sagittal line.
+These datasets allow us to investigate correlations between the involvement of individual LNLs, or between risk factors and involvement.
+In @fig-data-strat, we have plotted the prevalence of each contralateral LNL's involvement, stratified by T-category, number of ipsilaterally involved LNLs, and whether the tumor extended over the mid-sagittal line.
 A similar but more complete stratification is also tabulated in the appendix in @tbl-data-strat.
@@
 The higher prevalence of contralateral metastases in advanced T-category tumors may thus be partially explained by a higher rate of midline extension in these tumors.
-However, @fig-data-strat-uncorr indicates that advanced T-category and severe ipsilateral involvement do represent additional risk factors .
+However, @fig-data-strat-uncorr indicates that advanced T-category and severe ipsilateral involvement do represent additional risk factors.
 The figure considers only patients with lateralized tumors not extending over the midline.
@@
 
-Based on the observations of the [previous section](#sec-data-strat), any potential model that aims to also predict the risk for contralateral nodal involvement, should be able to take the following into account:
+Based on the observations of the [previous section](#sec-data-strat), any model to predict the risk for contralateral nodal involvement, should be able to describe the following:
 
@@
 
-Over the course of this work, we will first briefly recap the mentioned HMM in @sec-unilateral, which was so far used to model ipsilateral lymphatic progression only.
+Over the course of this work, we will first briefly recap the HMM in @sec-unilateral, which was so far used to model ipsilateral lymphatic progression only.
 Then, we intuitively extend it to include the contralateral side as well in @sec-ext-to-contra.
@@
 
-## Likelihood Function
+## Likelihood Function of the Unilateral Model
 
@@
 
-<!--There may, however, be diffusion of lymph fluid accross this line or bulky involvement that redirects lymphatic drainage significantly.-->
+<!-- There may, however, be diffusion of lymph fluid accross this line or bulky involvement that redirects lymphatic drainage significantly.-->
 
@@
 
-In general, the matrices $\boldsymbol{\Lambda}_\text{i}$ and $\boldsymbol{\Lambda}_\text{c}$ could be parameterized using a disjoint set of parameters.
-I.e., the ipsi- and contralateral spread rates are entirely different.
-However, using three sensible assumptions, we can reduce the parameter space by sharing some parameters between the sides:
+In general, the matrices $\boldsymbol{\Lambda}_\text{i}$ and $\boldsymbol{\Lambda}_\text{c}$ could be parameterized using a disjoint set of parameters, i.e., the ipsi- and contralateral spread rates are entirely different.
+However, using three sensible assumptions, we can reduce the parameter space by sharing some parameters between the two sides:
 
@@
 In that case, the probability is 1 to remain in that lateralization state.
-Thus, for this case we simply add the same distribution we are computing, but from one time-step earlier, arriving at a recursion relation:
+Thus, to account for this case we simply add the distribution $P\left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid \tau \right)$ from one time-step earlier, arriving at a recursion relation:
 
@@
 It is possible to provide our model with multiple different diagnostic modalities, each being characterized by different pairs of sensitivity and specificity.
-However, we instead chose to combine them into a single "consensus" diagnosis before parameter inference.
+However, we instead chose to combine them into a single "consensus" diagnosis before parameter inference, as introduced in @sec-sec-data-consensus.
 We opted for this because the literature values of sensitivity and specificity [@de_bondt_detection_2007;@kyzas_18f-fluorodeoxyglucose_2008] of imaging modalities like MRI and CT do not plausibly match some of our observations: In the USZ cohort, 78% of OPSCC patients where diagnosed with ipsilateral LNL II involvement via diagnostic imaging.
 This is virtually impossible with sensitivities around 80% and specificities lower than 100%.
-
-Our pre-training consensus was formed by considering all reported diagnostic information for a particular patient and LNL.
-When conflicts arose, we computed the *most likely* true state of involvement using the literature sensitivity and specificity values [@de_bondt_detection_2007;@kyzas_18f-fluorodeoxyglucose_2008].
 

### 2024-08-26 scripts/data_strat.py
start combining data-strat figures

@@ -45,27 +45,28 @@ def group_and_plot(
 
 def main():
     """Plot figure."""
-    nrows, ncols = 1, 3
+    nrows, ncols = 2, 2
     plt.rcParams.update(**shared.get_fontsizes())
     plt.rcParams.update(
         **shared.get_figsizes(
             nrows=nrows,
             ncols=ncols,
-            aspect_ratio=1.1,
+            aspect_ratio=1.5,
         )
     )
     fig = plt.figure()
     gs = gridspec.GridSpec(nrows=nrows, ncols=ncols, figure=fig)
 
-    left = fig.add_subplot(gs[0])
-    center = fig.add_subplot(gs[1], sharey=left)
-    right = fig.add_subplot(gs[2], sharey=center)
+    t_cat_ax = fig.add_subplot(gs[0,0])
+    ipsi_inv_ax = fig.add_subplot(gs[0,1], sharey=t_cat_ax)
+    midext_ax = fig.add_subplot(gs[1,0], sharey=ipsi_inv_ax)
     row = fig.add_subplot(gs[:], frame_on=False)
 
-    right.set_ylabel("contra prevalence [%]")
-    right.set_title("Midline Extension")
-    center.set_title("Ipsilateral Involvement")
-    left.set_title("T-category")
+    t_cat_ax.set_ylabel("contra prevalence [%]")
+    midext_ax.set_ylabel("contra prevalence [%]")
+    midext_ax.set_title("Midline Extension", fontweight="bold")
+    ipsi_inv_ax.set_title("Ipsilateral Involvement", fontweight="bold")
+    t_cat_ax.set_title("T-category", fontweight="bold")
     row.set_xlabel("Lymph Node Level", labelpad=17)
     row.set_xticks([])
     row.set_yticks([])
@@ -78,7 +79,7 @@ def main():
     group_and_plot(
         df=contra_by_midext,
         column="extension",
-        axes=right,
+        axes=midext_ax,
         colors=[COLORS["green"], COLORS["red"]],
     )
 
@@ -100,7 +101,7 @@ def main():
     group_and_plot(
         df=contra_by_ipsi,
         column="ipsi",
-        axes=center,
+        axes=ipsi_inv_ax,
         colors=[COLORS["green"], COLORS["orange"], COLORS["red"]],
     )
 
@@ -120,7 +121,7 @@ def main():
     group_and_plot(
         df=contra_by_t,
         column="t_stage",
[…]

### 2024-08-27 manuscript.qmd
put data-strat panels into one figure

@@
 These datasets allow us to investigate correlations between the involvement of individual LNLs, or between risk factors and involvement.
-In @fig-data-strat, we have plotted the prevalence of each contralateral LNL's involvement, stratified by T-category, number of ipsilaterally involved LNLs, and whether the tumor extended over the mid-sagittal line.
+In @fig-data-strat, we have plotted the prevalence of each contralateral LNL's involvement, stratified by risk factors such as T-category, number of ipsilaterally involved LNLs, and whether the tumor extended over the mid-sagittal line.
+We also show in the fourth subfigre that T-category and ipsilateral involvement are not just rpedictive of contralateral involvement because they correlate with midline extension.
 A similar but more complete stratification is also tabulated in the appendix in @tbl-data-strat.
 
-![Contralateral involvement stratified by T-category (left panel), the number of metastatic LNLs ipsilaterally (center panel), and whether the primary tumor extended over the mid-sagittal line or was clearly lateralized (right panel).](figures/fig_data_strat.svg){#fig-data-strat}
+![Contralateral involvement stratified by T-category (top left panel), the number of metastatic LNLs ipsilaterally (top right panel), and whether the primary tumor extended over the mid-sagittal line or was clearly lateralized (bottom left panel).
+Additionally, in the bottom right, we show that even for lateralized tumors, T-category and ipsilateral involvement are correlated with contralateral metastasis by comparing the contralateral involvement prevalence for selected scenarios that vary in their T-category and ipsilateral involvement extent.](figures/fig_data_strat.svg){#fig-data-strat}
 
-The left panel in @fig-data-strat indicates that T-category is correlated with contralateral involvement (as it is with overal involvement).
+The top left panel in @fig-data-strat indicates that T-category is correlated with contralateral involvement (as it is with overal involvement).
 This is simply because T-category may on average be considered a surrogate for the time between onset of disease and diagnosis.
@@
 
-Lastly, the right panel in @fig-data-strat shows that patients with a tumor crossing the mid-sagittal line show contralateral involvement vastly more often compared to patients with clearly lateralized tumors.
+The bottom left panel in @fig-data-strat shows that patients with a tumor crossing the mid-sagittal line show contralateral involvement vastly more often compared to patients with clearly lateralized tumors.
 This makes intuitive sense, because the lymphatic system in the head and neck region is typically symmetric and thus no major vessels cross the midline.
@@
 The higher prevalence of contralateral metastases in advanced T-category tumors may thus be partially explained by a higher rate of midline extension in these tumors.
-However, @fig-data-strat-uncorr indicates that advanced T-category and severe ipsilateral involvement do represent additional risk factors.
+However, the bottom right panel of @fig-data-strat indicates that advanced T-category and severe ipsilateral involvement do represent additional risk factors.
 The figure considers only patients with lateralized tumors not extending over the midline.
@@
 -->
-
-![Contralateral involvement prevalence by LNL for lateralized tumors.
-Shown are four different scenarios: 1) early T-category with an ipsilateral N0 neck (green), 2) early T-category with ipsilaterally LNL II involved (blue), 3) late T-category with ipsilateral LNL II metastatic (orange), and 4) early T-category with ipsilateral both levels II and III involved (red).
-This illustrates that even when controlling for the tumor's lateralization, T-category and ipsilaeral involvement are predictive of contralateral metastases.](figures/fig_data_strat_uncorr.svg){#fig-data-strat-uncorr}
 

### 2024-08-27 scripts/data_strat.py

@@ -5,11 +5,10 @@ import numpy as np
 import pandas as pd
 import paths
 import shared
-from matplotlib import gridspec
-from matplotlib.axes import Axes
-
 from lyscripts import utils
 from lyscripts.plot.utils import COLORS
+from matplotlib import gridspec
+from matplotlib.axes import Axes
 
 
 def group_and_plot(
@@ -51,23 +50,31 @@ def main():
         **shared.get_figsizes(
             nrows=nrows,
             ncols=ncols,
-            aspect_ratio=1.5,
+            aspect_ratio=1.6,
         )
     )
+    plt.rcParams.update({
+        "xtick.bottom": False,
+        "xtick.major.pad": 1,
+        "ytick.left": False,
+        "ytick.major.pad": 1,
+    })
     fig = plt.figure()
     gs = gridspec.GridSpec(nrows=nrows, ncols=ncols, figure=fig)
 
     t_cat_ax = fig.add_subplot(gs[0,0])
     ipsi_inv_ax = fig.add_subplot(gs[0,1], sharey=t_cat_ax)
     midext_ax = fig.add_subplot(gs[1,0], sharey=ipsi_inv_ax)
+    uncorr_ax = fig.add_subplot(gs[1,1], sharey=midext_ax)
     row = fig.add_subplot(gs[:], frame_on=False)
 
-    t_cat_ax.set_ylabel("contra prevalence [%]")
-    midext_ax.set_ylabel("contra prevalence [%]")
-    midext_ax.set_title("Midline Extension", fontweight="bold")
     ipsi_inv_ax.set_title("Ipsilateral Involvement", fontweight="bold")
     t_cat_ax.set_title("T-category", fontweight="bold")
-    row.set_xlabel("Lymph Node Level", labelpad=17)
+    midext_ax.set_title("Midline Extension", fontweight="bold")
+    uncorr_ax.set_title("Lateralized Tumors", fontweight="bold")
+
+    row.set_xlabel("Lymph Node Level", labelpad=15)
+    row.set_ylabel("Contralateral Involvement Prevalence [%]", labelpad=17)
     row.set_xticks([])
     row.set_yticks([])
 
@@ -124,6 +131,59 @@ def main():
         axes=t_cat_ax,
         colors=[COLORS["blue"], COLORS["orange"]],
     )
+
+    # plot showing that T-stage and ipsi involvement are still predictive, when
[…]

### 2024-08-27 scripts/data_strat_uncorr.py

@@ -1,83 +1,0 @@
-"""Plot how T-stage and ipsi inv. predict contra involv. when controlling for midext."""
-
-import numpy as np
-import paths
-import shared
-from matplotlib import pyplot as plt
-from shared import COL, CONTRA_LNLS
-
-from lyscripts import utils
-from lyscripts.plot.utils import COLORS
-
-
-def main():
-    """Plot figure."""
-    nrows, ncols = 1, 1
-    plt.rcParams.update(**shared.get_fontsizes())
-    plt.rcParams.update(
-        **shared.get_figsizes(
-            nrows=nrows,
-            ncols=ncols,
-            width=17/2,
-        )
-    )
-    _fig, ax = plt.subplots()
-
-    raw = utils.load_patient_data(paths.data)
-    lateral = raw[raw[COL.midext] == False]
-
-    early = lateral[lateral[COL.t_stage] <= 2]
-    early_n0 = early[early["max_llh", "ipsi"].sum(axis=1) == 0]
-    early_II = early[early["max_llh", "ipsi", "II"] == True]
-    early_II_and_III = early_II[early_II["max_llh", "ipsi", "III"] == True]
-
-    late = lateral[lateral[COL.t_stage] > 2]
-    late_II = late[late["max_llh", "ipsi", "II"] == True]
-
-    early_n0 = early_n0[CONTRA_LNLS]["max_llh", "contra"].mean(axis=0)
-    early_II = early_II[CONTRA_LNLS]["max_llh", "contra"].mean(axis=0)
-    early_II_and_III = early_II_and_III[CONTRA_LNLS]["max_llh", "contra"].mean(axis=0)
-    late_II = late_II[CONTRA_LNLS]["max_llh", "contra"].mean(axis=0)
-
-    pos = np.arange(len(early_n0))
-    kwargs = {"width": 0.6, "zorder": 2}
-
-    plt.bar(
-        x=pos+0.2,
-        height=100 * early_II_and_III,
-        color=COLORS["red"],
-        label="early; ipsi II and III",
-        **kwargs,
-    )
-    plt.bar(
-        x=pos+0.1,
-        height=100 * late_II,
-        color=COLORS["orange"],
-        label="late; ipsi II",
-        **kwargs,
-    )
-    plt.bar(
-        x=pos,
-        height=100 * early_II,
-        color=COLORS["blue"],
-        label="early; ipsi II",
-        **kwargs,
-    )
-    plt.bar(
[…]

### 2024-08-27 scripts/data_strat.py
update rcParams in some figures

@@ -38,8 +38,7 @@ def group_and_plot(
         )
 
     axes.legend()
-    axes.grid(axis="y", color=COLORS["gray"])
-    axes.set_axisbelow(True)
+    axes.grid(visible=False, axis="x")
 
 
 def main():
@@ -53,12 +52,7 @@ def main():
             aspect_ratio=1.6,
         )
     )
-    plt.rcParams.update({
-        "xtick.bottom": False,
-        "xtick.major.pad": 1,
-        "ytick.left": False,
-        "ytick.major.pad": 1,
-    })
+    plt.rcParams.update(shared.get_axes_params())
     fig = plt.figure()
     gs = gridspec.GridSpec(nrows=nrows, ncols=ncols, figure=fig)
 
@@ -183,7 +177,6 @@ def main():
 
     uncorr_ax.legend()
     uncorr_ax.set_xticks(pos, labels=["I", "II", "III", "IV", "V"])
-    uncorr_ax.grid(axis="y", color=COLORS["gray"])
 
     plt.savefig(shared.get_figure_path(__file__))
 

### 2024-08-27 scripts/model_burnin_history.py

@@ -4,7 +4,6 @@ from collections import namedtuple
 
 import matplotlib.pyplot as plt
 import pandas as pd
-import paths
 import shared
 from matplotlib import ticker
 
@@ -23,12 +22,12 @@ def main():
     config_map = {
         "acor_times": PlotConfig(
             title="Estimated Autocorrelation Time",
-            ylabel="steps",
+            ylabel="estimate [steps]",
             label="Autocorrelation Time",
         ),
         "accept_fracs": PlotConfig(
             title="Average Acceptance Fraction of Walkers",
-            ylabel=None,
+            ylabel="fraction [%]",
             label=None,
         ),
     }
@@ -41,6 +40,7 @@ def main():
             width=17,
         )
     )
+    plt.rcParams.update(shared.get_axes_params())
 
     fig, axes = plt.subplots(
         nrows=nrows,
@@ -68,7 +68,9 @@ def main():
         label="Trust Threshold",
     )
     axes[0].legend()
-    axes[1].yaxis.set_major_formatter(ticker.StrMethodFormatter("{x:.0%}"))
+    axes[1].yaxis.set_major_formatter(
+        ticker.FuncFormatter(lambda x, *_: f"{100 * x:.0f}")
+    )
 
     plt.savefig(shared.get_figure_path(__file__))
 

### 2024-08-27 scripts/model_midext_evo.py

@@ -2,11 +2,9 @@
 
 import matplotlib.pyplot as plt
 import numpy as np
-import paths
 import shared
-from matplotlib import ticker
-
 from lyscripts.plot.utils import COLORS
+from matplotlib import ticker
 
 
 def main():
@@ -34,6 +32,7 @@ def main():
             width=17 / 2,
         )
     )
+    plt.rcParams.update(shared.get_axes_params())
 
     fig, axes = plt.subplots(nrows=nrows, ncols=ncols, sharex=True)
     w = 0.3

### 2024-08-27 scripts/shared.py

@@ -4,6 +4,7 @@ from collections import namedtuple
 from pathlib import Path
 from typing import Any, Literal
 
+from lyscripts.plot.utils import COLORS
 import numpy as np
 import pandas as pd
 from lymph import types
@@ -168,3 +169,16 @@ CONTRA_LNLS = [
     ("max_llh", "contra", "IV"),
     ("max_llh", "contra", "V"),
 ]
+
+
+def get_axes_params() -> dict[str, Any]:
+    """Return rc params regarding axes, ticks and grid."""
+    return {
+        "xtick.bottom": False,
+        "xtick.major.pad": 1,
+        "ytick.left": False,
+        "ytick.major.pad": 1,
+        "axes.grid": True,
+        "axes.grid.axis": "y",
+        "grid.color": COLORS["gray"],
+    }

### 2024-08-27 manuscript.qmd
move prev. of midext to appendix for now

@@
 It is possible to provide our model with multiple different diagnostic modalities, each being characterized by different pairs of sensitivity and specificity.
-However, we instead chose to combine them into a single "consensus" diagnosis before parameter inference, as introduced in @sec-sec-data-consensus.
+However, we instead chose to combine them into a single "consensus" diagnosis before parameter inference, as introduced in @sec-data-consensus.
 We opted for this because the literature values of sensitivity and specificity [@de_bondt_detection_2007;@kyzas_18f-fluorodeoxyglucose_2008] of imaging modalities like MRI and CT do not plausibly match some of our observations: In the USZ cohort, 78% of OPSCC patients where diagnosed with ipsilateral LNL II involvement via diagnostic imaging.
@@
 
-### Prevalence of Midline Extension
-
-![Comparing the predicted (histograms) and observed (lines depicting beta posteriors) prevalence of midline extension for early (blue) and late (orange) T-category.
-While the prevalence is predicted correctly when marginalizing over T-category, the model cannot capture the degree of separation observed in the data.
-Since the tumor's midline extension is virtually always part of the diagnosis and hence *given* when predicting a patient's risk, we do not consider this discrepancy a major issue.](figures/fig_model_prevalences_midext.svg){#fig-model-prevalences-midext width="50%"}
-
-Lastly, in @fig-model-prevalences-midext, we plot the prevalence of midline extension in the data versus our model's prediction.
-It is obvious the model cannot match the large spread between early and advanced T-category seen in the data.
-This is because to achieve that, it would need to increase the advanced T-category patient's prior distribution over diagnosis times and at the same time reduce the probability of the tumor to cross the midline during a time-step.
-But since the time-priors parameter is also coupled with the spread probabilities among the LNLs, the model does not have that freedom.
-
-<!-- Should the paragraphs below be part of the discussion? -->
-
-However, we do not consider this discrepancy a major limitation of the model: We will not realistically be interested in the probability of midline extension, as it is always possible to assess it with high certainty.
-That is also the reason why we initially modelled the midline extension *not* as a random variable, but as a global risk factor that would have been turned on or off from the onset of a patient's disease evolution.
-This, however, lead to overly high risks for contralateral involvement in advanced T-category patients with midline extension, because then the model assumes an increased spread to the contralateral side from the onset of the disease.
-Which is probably not true in a majority of those cases.
-Thus, treating it as a random variable that only becomes true during a patient's disease evolution resulted in a better description of the data.
-
-Formally, the wrong prediction of midline extension prevalence makes little difference, since it is always given: Instead of $P\left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \mid \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \right)$, we typically compute $P\left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \mid \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c}, \epsilon \right)$, which does not suffer from the wrong probability of midline extension, as the distribution over hidden states is renormalized:
-
-$$
-P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \mid \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c}, \epsilon \right) = \frac{P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \mid \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right) P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)}{P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c}, \epsilon \right)}
-$$
-
-Note that a distribution over $\epsilon$ appears both in the enumerator and the denominator, which largely cancel each other, leaving only the midline extension's effect on the distribution over hidden states in the prediction.
-
-Also, the discrepancy in midline extension prevalence between early and advanced T-category is particularly pronounced in oropharyngeal SCC patients.
-For example, in oral cavity SCC, the midline extension only increases from 15.4% (20 out of 130) to 33.3% (13 out of 39).
-
 # Results: Prediction of Risk for Occult Disease
@@
 ```
+
+# Prevalence of Midline Extension {.appendix}
+
+<!-- I moved this here, because we think this is probably irrelevant or distracting w.r.t. the main findings of the paper. But I do think it is somewhat interesting. -->
+
+![Comparing the predicted (histograms) and observed (lines depicting beta posteriors) prevalence of midline extension for early (blue) and late (orange) T-category.
+While the prevalence is predicted correctly when marginalizing over T-category, the model cannot capture the degree of separation observed in the data.
+Since the tumor's midline extension is virtually always part of the diagnosis and hence *given* when predicting a patient's risk, we do not consider this discrepancy a major issue.](figures/fig_model_prevalences_midext.svg){#fig-model-prevalences-midext width="50%"}
+
+In @fig-model-prevalences-midext, we plot the prevalence of midline extension in the data versus our model's prediction.
+It is obvious the model cannot match the large spread between early and advanced T-category seen in the data.
+This is because to achieve that, it would need to increase the advanced T-category patient's prior distribution over diagnosis times and at the same time reduce the probability of the tumor to cross the midline during a time-step.
+But since the time-priors parameter is also coupled with the spread probabilities among the LNLs, the model does not have that freedom.
+
+<!-- Should the paragraphs below be part of the discussion? -->
+
+However, we do not consider this discrepancy a major limitation of the model: We will not realistically be interested in the probability of midline extension, as it is always possible to assess it with high certainty.
+That is also the reason why we initially modelled the midline extension *not* as a random variable, but as a global risk factor that would have been turned on or off from the onset of a patient's disease evolution.
+This, however, lead to overly high risks for contralateral involvement in advanced T-category patients with midline extension, because then the model assumes an increased spread to the contralateral side from the onset of the disease.
+Which is probably not true in a majority of those cases.
+Thus, treating it as a random variable that only becomes true during a patient's disease evolution resulted in a better description of the data.
+
+Formally, the wrong prediction of midline extension prevalence makes little difference, since it is always given: Instead of $P\left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \mid \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \right)$, we typically compute $P\left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \mid \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c}, \epsilon \right)$, which does not suffer from the wrong probability of midline extension, as the distribution over hidden states is renormalized:
+
+$$
+P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \mid \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c}, \epsilon \right) = \frac{P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \mid \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right) P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)}{P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c}, \epsilon \right)}
+$$
+
+Note that a distribution over $\epsilon$ appears both in the enumerator and the denominator, which largely cancel each other, leaving only the midline extension's effect on the distribution over hidden states in the prediction.
+
+Also, the discrepancy in midline extension prevalence between early and advanced T-category is particularly pronounced in oropharyngeal SCC patients.
+For example, in oral cavity SCC, the midline extension only increases from 15.4% (20 out of 130) to 33.3% (13 out of 39).

### 2024-08-28 scripts/compile_variables.py
start with variables script

@@ -1,0 +1,29 @@
+"""Script to compile variables from data and models and stores them in a YAML file."""
+
+import argparse
+from pathlib import Path
+
+import pandas as pd
+import yaml
+
+
+def create_parser() -> argparse.ArgumentParser:
+    """Create an argument parser for the script."""
+    parser = argparse.ArgumentParser(description=__doc__)
+    parser.add_argument(
+        "--output",
+        type=Path,
+        default=Path("_variables.yaml"),
+        help="Path to the output YAML.",
+    )
+    return parser
+
+
+def main() -> None:
+    """Merge the variables from data and models and store them in `_variables.yaml`."""
+    args = create_parser().parse_args()
+
+    variables = {}
+
+    with open(args.output, mode="w", encoding="utf-8") as file:
+        yaml.dump(variables, file)

### 2024-08-29 scripts/compile_variables.py
use pydantic for cmd args

@@ -5,25 +5,20 @@ from pathlib import Path
 
 import pandas as pd
 import yaml
+from pydantic_settings import BaseSettings
 
 
-def create_parser() -> argparse.ArgumentParser:
-    """Create an argument parser for the script."""
-    parser = argparse.ArgumentParser(description=__doc__)
-    parser.add_argument(
-        "--output",
-        type=Path,
-        default=Path("_variables.yaml"),
-        help="Path to the output YAML.",
-    )
-    return parser
+class CmdSettings(BaseSettings, cli_parse_args=True):
+    """Settings for the command-line arguments."""
+
+    output: Path = Path("_variables.yaml")
 
 
 def main() -> None:
     """Merge the variables from data and models and store them in `_variables.yaml`."""
-    args = create_parser().parse_args()
+    cmd = CmdSettings()
 
     variables = {}
 
-    with open(args.output, mode="w", encoding="utf-8") as file:
+    with open(cmd.output, mode="w", encoding="utf-8") as file:
         yaml.dump(variables, file)

### 2024-09-03 manuscript.qmd
make vars in data section dynamic

@@
 
-from scripts import shared, paths, COL, get_lnl_cols
+from scripts import shared, paths
+from scripts.shared import COL, get_lnl_cols
 
@@
 1.
-In section @sec-data we present a multi-centric dataset on lymph node involvement in 833 OPSCC patients.
+In section @sec-data we present a multi-centric dataset on lymph node involvement in {{< var data.num_patients >}} OPSCC patients.
 Based on the dataset the main risk factors for contralateral lymph node involvement are identified and the requirements for the bilateral model extension are described (@sec-requirements).
@@
 
-To be able to create models for lymphatic tumor progression that include all relevant LNLs including the contralateral side, we have collected a detailed dataset of 833<!-- num patients --> patients with newly diagnosed oropharyngeal squamous cell carcinomas [@ludwig_dataset_2022;@ludwig_multi-centric_2023].
+To be able to create models for lymphatic tumor progression that include all relevant LNLs including the contralateral side, we have collected a detailed dataset of {{< var data.num_patients >}} patients with newly diagnosed oropharyngeal squamous cell carcinomas [@ludwig_dataset_2022;@ludwig_multi-centric_2023].
 It reports the lymph node involvement per LNL of every patient individually in tabular form, in addition to other primary tumor and patient characteristics such as T-category, subside, and lateralization of the primary tumor, and HPV p16 status.
@@
 The three risk factors for contralateral involvement, midline extension, T-category, and ipsilateral involvement are correlated.
-For example, midline extension was present in 48% of advanced T-category tumors but only in 7% of early T-category tumors.
+For example, midline extension was present in {{< var data.advanced_with_midext_percent >}}% of advanced T-category tumors but only in {{< var data.early_with_midext_percent >}}% of early T-category tumors.
 The higher prevalence of contralateral metastases in advanced T-category tumors may thus be partially explained by a higher rate of midline extension in these tumors.
@@
 The figure considers only patients with lateralized tumors not extending over the midline.
-For early T-category and healthy ipsilateral levels I-V, only 1% (1 out of 84 patients) shows involvement of contralateral level II.
-This increases to 5% (9 out of 165 patients) if ipsilateral level II is involved, to 15% (9 out of 60 patients) if ipsilateral levels II and III are involved, and to 23% (6 out of 26 patients) for advanced T-category with ipsilateral levels II and III involved.<!-- TODO: make values here dynamic.
--->
+For early T-category and healthy ipsilateral levels I-V, only {{< var data.early_ipsin0_cII_percent >}}% ({{< var data.early_ipsin0_cII_match >}} out of {{< var data.early_ipsin0_cII_total >}} patients) shows involvement of contralateral level II.
+This increases to {{< var data.early_ipsiII_cII_percent >}}% ({{< var data.early_ipsiII_cII_match >}} out of {{< var data.early_ipsiII_cII_total >}} patients) if ipsilateral level II is involved, to {{< var data.early_ipsiIIandIII_cII_percent >}}% ({{< var data.early_ipsiIIandIII_cII_match >}} out of {{< var data.early_ipsiIIandIII_cII_total >}} patients) if ipsilateral levels II and III are involved, and to {{< var data.late_ipsiIIandIII_cII_percent >}}% ({{< var data.late_ipsiIIandIII_cII_match >}} out of {{< var data.late_ipsiIIandIII_cII_total >}} patients) for advanced T-category with ipsilateral levels II and III involved.
 
@@
 
-grouped = contra_inv.groupby(by=["t_stage", "ipsi", "midext"])
+grouped = contra_inv.groupby(by=["t_stage", "ipsi", "midext"], dropna=False)
 num_involved = grouped.sum()
@@
   rsuffix=" (%)",
-).sort_index(axis="columns")
+)
 involved.columns = pd.MultiIndex.from_product(
@@
   .reset_index()
-  .sort_values("ipsi", ascending=True)
-  .sort_values("T-cat.", ascending=False)
+  .sort_values(
+    by=["T-cat.", "ipsi", "Mid. ext."],
+    ascending=[False, True, True],
+  )
   .style

### 2024-09-03 scripts/compile_variables.py

@@ -1,17 +1,141 @@
 """Script to compile variables from data and models and stores them in a YAML file."""
 
-import argparse
 from pathlib import Path
+from typing import Any
 
+import lydata  # noqa: F401
 import pandas as pd
+import shared
 import yaml
+from lydata.accessor import Q
+from pandas.api.typing import DataFrameGroupBy
 from pydantic_settings import BaseSettings
 
 
 class CmdSettings(BaseSettings, cli_parse_args=True):
     """Settings for the command-line arguments."""
 
-    output: Path = Path("_variables.yaml")
+    output: Path = Path("_variables.dynamic.yaml")
+    precision: int = 2
+
+
+def group_contra_inv(data: pd.DataFrame) -> DataFrameGroupBy:
+    """Group the data by T-stage, num of ipsi involved, and midline extension."""
+    lnl_cols = shared.get_lnl_cols("contra", lnls=["I", "II", "III", "IV"])
+    num_ipsi_inv = data[shared.get_lnl_cols("ipsi")].sum(axis="columns")
+    contra_inv = data[lnl_cols].copy()
+    contra_inv.columns = contra_inv.columns.droplevel([0,1])
+
+    contra_inv["t_stage"] = data[shared.COL.t_stage]
+    contra_inv["ipsi"] = num_ipsi_inv.map(lambda x: str(x) if x <= 1 else "≥ 2")
+    contra_inv["midext"] = data[shared.COL.midext]
+
+    return contra_inv.groupby(by=["t_stage", "ipsi", "midext"])
+
+
+def cast_numpy_to_native(variables: dict[str, Any]) -> dict[str, Any]:
+    """Convert numpy dtypes to native Python types."""
+    result = {}
+
+    for key, value in variables.items():
+        try:
+            result[key] = value.item()
+        except AttributeError:
+            result[key] = value
+
+    return result
+
+
+def get_data_variables(precision: int) -> dict:
+    """Get the variables from the data."""
+    variables = {}
+    data = shared.get_data()
+    grouped_contra_inv = group_contra_inv(data)
+    absolute = grouped_contra_inv.sum()
+    percent = absolute / grouped_contra_inv.count()
+
+    variables["num_patients"] = len(data)
+
+    is_early = Q("t_stage", "==", "early")
+    is_late = Q("t_stage", "==", "late")
[…]

### 2024-09-03 scripts/data_strat.py

@@ -5,10 +5,11 @@ import numpy as np
 import pandas as pd
 import paths
 import shared
-from lyscripts import utils
-from lyscripts.plot.utils import COLORS
 from matplotlib import gridspec
 from matplotlib.axes import Axes
+
+from lyscripts import utils
+from lyscripts.plot.utils import COLORS
 
 
 def group_and_plot(

### 2024-09-03 scripts/shared.py

@@ -4,12 +4,12 @@ from collections import namedtuple
 from pathlib import Path
 from typing import Any, Literal
 
-from lyscripts.plot.utils import COLORS
 import numpy as np
 import pandas as pd
 from lymph import types
 
 from lyscripts import utils
+from lyscripts.plot.utils import COLORS
 
 try:
     from . import paths
@@ -20,6 +20,34 @@ except ImportError:
 pd.options.mode.copy_on_write = False
 CM_TO_INCH = 0.393701
 GOLDEN_RATIO = 1.61803398875
+
+Columns = namedtuple(
+    "Columns", [
+        "inst",
+        "age",
+        "nd",
+        "t_stage",
+        "n_stage",
+        "midext",
+        "ipsi_III",
+    ])
+
+COL = Columns(
+    inst=("patient", "#", "institution"),
+    age=("patient", "#", "age"),
+    nd=("patient", "#", "neck_dissection"),
+    t_stage=("tumor", "1", "t_stage"),
+    n_stage=("patient", "#", "n_stage"),
+    midext=("tumor", "1", "extension"),
+    ipsi_III=("max_llh", "ipsi", "III"),
+)
+CONTRA_LNLS = [
+    ("max_llh", "contra", "I"),
+    ("max_llh", "contra", "II"),
+    ("max_llh", "contra", "III"),
+    ("max_llh", "contra", "IV"),
+    ("max_llh", "contra", "V"),
+]
 
 
 def get_params() -> dict[str, Any]:
@@ -56,7 +84,9 @@ def get_data(map_t_stage: bool = True) -> pd.DataFrame:
     data = utils.load_patient_data(paths.data)
 
     if map_t_stage:
-        data.ly.map_t_stage(lambda x: "early" if x <= 2 else "late")
+        data[COL.t_stage] = data[COL.t_stage].map(
+            lambda x: "early" if x <= 2 else "late"
+        )
 
     return data
 
@@ -139,36 +169,6 @@ def get_lnl_cols(
     """Get the columns of the LNL involvements."""
     lnls = lnls or ["I", "II", "III", "IV", "V"]
     return [("max_llh", side, lnl) for lnl in lnls]
-
-
-Columns = namedtuple(
-    "Columns", [
-        "inst",
-        "age",
-        "nd",
-        "t_stage",
-        "n_stage",
-        "midext",
-        "ipsi_III",
-    ])
-
-
[…]

### 2024-09-03 manuscript.qmd
make model vars dynamic

@@
 The three risk factors for contralateral involvement, midline extension, T-category, and ipsilateral involvement are correlated.
-For example, midline extension was present in {{< var data.advanced_with_midext_percent >}}% of advanced T-category tumors but only in {{< var data.early_with_midext_percent >}}% of early T-category tumors.
+For example, midline extension was present in {{< var data.late_with_midext_percent >}}% of advanced T-category tumors but only in {{< var data.early_with_midext_percent >}}% of early T-category tumors.
 The higher prevalence of contralateral metastases in advanced T-category tumors may thus be partially explained by a higher rate of midline extension in these tumors.
@@
 For example, we may be interested in the prevalence of contralateral LNL II involvement (i.e., contra LNL II "involved" and all other LNLs "masked") under the scenario of early T-category (T0-T2) and no midline extension ($\epsilon=\texttt{False}$).
-To compute this prevalence in the data, we select all patients of this scenario (in our data, this amounts to 379<!-- TODO: dyamically compute this --> patients).
-Of those, 28<!-- TODO: dynamically compute this --> were found to harbor metastases in their contralateral LNL II.
-Therefore, the prevalence is 7%<!-- TODO: dynamically compute this -->.
+To compute this prevalence in the data, we select all patients of this scenario (in our data, this amounts to {{< var data.early_nomidext_cII_total >}} patients).
+Of those, {{< var data.early_nomidext_cII_match >}} were found to harbor metastases in their contralateral LNL II.
+Therefore, the prevalence is {{< var data.early_nomidext_cII_percent >}}%.
 
@@
 In @fig-model-midext-evo (top panel), we visualize the prior distribution over diagnose times $P(t)$.
-By parameter choice, early T-category tumors are diagnosed on average after 3 time steps.
-Advanced T-category tumors are diagnosed on average after 4.5 time steps ($p_\text{adv.}=0.45$) <!-- TODO: make dynamic -->, a parameter learned from the data to describe the higher involvement for advanced T-category tumors.
-The probability per time step for the tumor to grow over the midline, $p_\epsilon = 0.0816$ <!-- TODO: make dynamic --> yields the probability of midline extension $P(\epsilon \mid t)$ at a given time step $t$ (red line).
-The parameter is adjusted such that the model yields the correct proportion of patients with tumors extending over the midline seen in the dataset.
-The bottom panel in @fig-model-midext-evo shows the joint probability $P(\epsilon, t)$ to be diagnosed at time step $t$ with a given state of midline extension and T-category.
+By parameter choice, early T-category tumors are diagnosed on average after {{< var model.early_expected_time >}} time steps.
+Advanced T-category tumors are diagnosed on average after {{< var model.late_expected_time >}} time steps, because the parameter $p_\text{adv.}$ was learned to be {{< var model.params_percent.late_p >}}% for advanced T-category tumors.
+The probability per time step for the tumor to grow over the midline ($p_\epsilon$) was on average {{< var model.params_percent.midext_prob >}}%.
+Using this, one can compute the probability of midline extension $P(\epsilon \mid t)$ at a given time step $t$ (red line in @fig-model-midext-evo).
+The bottom panel in this figure shows the joint probability $P(\epsilon, t)$ to be diagnosed at time step $t$ with a given state of midline extension and T-category.
 
@@
 The most important variable impacting the prediction for contralateral level II involvement in our model is the tumor's lateralization.
-A patient with a clearly lateralized early T-category tumor and a clinically N0 neck is predicted to have a 1-2% <!-- TODO: make this dynamic, and include this scenario in the plot --> risk for occult disease in contralateral LNL II.
+A patient with a clearly lateralized early T-category tumor and a clinically N0 neck is predicted to have a 1-2% <!-- TODO: make this dynamic --> risk for occult disease in contralateral LNL II.
 For a patient with a clearly lateralized early T-category tumor but *with* mid-sagittal extension, the risk increases to almost 7%.

### 2024-09-03 scripts/compile_variables.py

@@ -7,7 +7,7 @@ import lydata  # noqa: F401
 import pandas as pd
 import shared
 import yaml
-from lydata.accessor import Q
+from lydata.accessor import Q, QueryPortion
 from pandas.api.typing import DataFrameGroupBy
 from pydantic_settings import BaseSettings
 
@@ -46,19 +46,30 @@ def cast_numpy_to_native(variables: dict[str, Any]) -> dict[str, Any]:
     return result
 
 
-def get_data_variables(precision: int) -> dict:
+def variables_from_portion(
+    key: str,
+    portion: QueryPortion,
+    precision: int = 2,
+) -> dict[str, Any]:
+    """Get a dictionary from a QueryPortion object."""
+    return {
+        f"{key}_match": portion.match,
+        f"{key}_total": portion.total,
+        f"{key}_percent": round(portion.percent, precision),
+    }
+
+
+def get_data_variables(precision: int = 2) -> dict[str, Any]:
     """Get the variables from the data."""
     variables = {}
     data = shared.get_data()
-    grouped_contra_inv = group_contra_inv(data)
-    absolute = grouped_contra_inv.sum()
-    percent = absolute / grouped_contra_inv.count()
 
     variables["num_patients"] = len(data)
 
     is_early = Q("t_stage", "==", "early")
     is_late = Q("t_stage", "==", "late")
     has_midext = Q("midext", "==", True)
+    not_has_midext = Q("midext", "==", False)
     is_cII_involved = Q(("max_llh", "contra", "II"), "==", True)
     is_ipsi_healthy = (
         Q(("max_llh", "ipsi", "I"), "==", False)
@@ -82,60 +93,93 @@ def get_data_variables(precision: int) -> dict:
         & Q(("max_llh", "ipsi", "V"), "==", False)
     )
 
-    variables["early_with_midext_percent"] = round(
-        number=data.ly.portion(query=has_midext, given=is_early).percent,
-        ndigits=precision,
-    )
-    variables["advanced_with_midext_percent"] = round(
-        number=data.ly.portion(query=has_midext, given=is_late).percent,
-        ndigits=precision,
-    )
+    variables.update(variables_from_portion(
+        key="early_with_midext",
+        portion=data.ly.portion(query=has_midext, given=is_early),
+        precision=precision,
+    ))
+    variables.update(variables_from_portion(
+        key="late_with_midext",
+        portion=data.ly.portion(query=has_midext, given=is_late),
[…]

### 2024-09-03 manuscript.qmd
make risks dynamic, too
Also adds paragraph about LNLs I, V, and VII.

Related: #12

@@
 A patient with a clearly lateralized early T-category tumor and a clinically N0 neck is predicted to have a 1-2% <!-- TODO: make this dynamic --> risk for occult disease in contralateral LNL II.
-For a patient with a clearly lateralized early T-category tumor but *with* mid-sagittal extension, the risk increases to almost 7%.
-<!-- TODO: make this dynamic -->
+For a patient with a clearly lateralized early T-category tumor but *with* mid-sagittal extension, the risk increases to {{< var risk.early_ext_iN0_cN0_II >}}%.
 
-Also advanced T-category increases the risk of occult disease but plays a lesser role: Considering the scenario of a tumor that crosses the midline and an ipsilateral neck where at least LNL III is clinically involved, the risk for occult contralateral LNL II disease is around 8.5%.
-<!-- TODO: make this dynamic --> For the same scenario and an advanced T-category tumor, the risk increases to 11%.
-<!-- TODO: make this dynamic -->
+Also advanced T-category increases the risk of occult disease but plays a lesser role: Considering the scenario of an early T-category tumor that crosses the midline and an ipsilateral neck where the LNLs II and III are clinically involved, the risk for occult contralateral LNL II disease is around {{< var risk.early_ext_iII+III_cN0_II >}}%.
+For the same scenario and an advanced T-category tumor, the risk increases to {{< var risk.late_ext_iII+III_cN0_II >}}%.
 
 Lastly, the predicted risk is also correlated via the time-steps to the degree of ipsilateral involvement.
-Changing the aforementioned scenario (advanced T-category, midline extension, ipsilateral LNL III clinically involved) to one where the patient presents with a clinically N0 neck, the risk for occult disease in the contralateral LNL II falls from 11% <!-- TODO: make this dynamic --> to 9.5%.
-<!-- TODO: make this dynamic -->
+When an early T-category patient with a tumor that extends over the mid-sagittal line presents with a clinically N0 ipsilateral neck instead of involvement in the levels II and III, the risk is {{< var risk.early_ext_iN0_cN0_II >}}% instead of {{< var risk.early_ext_iII+III_cN0_II >}}%.
 
-Taken together, T-category and ipsilateral involvement may still considerably impact the risk prediction for contralateral involvement: In @fig-model-risks the scenarios underlying the orange (6.5% <!-- TODO: make this dynamic -->) and the red (11% <!-- TODO: make this dynamic -->) histograms differ in T-category (early vs advanced) and clinical diagnosis (N0 vs ipsilateral at least LNL III involved).
+Taken together, T-category and ipsilateral involvement may still considerably impact the risk prediction for contralateral involvement: In @fig-model-risks the scenarios underlying the orange ({{< var risk.early_ext_iN0_cN0_II >}}%) and the purple ({{< var risk.late_ext_iII+III_cN0_II >}}%) histograms differ in T-category (early vs advanced) and clinical diagnosis (N0 vs ipsilateral LNL II and III involved).
 
@@
 As shown in the center subplot of @fig-model-risks, the risk for occult disease in the contralateral LNL III may only cross a 5% threshold if the upstream level II is clinically involved, too.
-If the clinical diagnosis of the contralateral upstream level II is pathologically confirmed by FNA the risk increases by another percentage point<!-- TODO: make dynamic -->.
-When the tumor additionally crosses the mid-sagittal line, the risk may reach around 7%<!-- TODO: make dynamic -->.
+If the clinical diagnosis of the contralateral upstream level II is pathologically confirmed by FNA the risk increases by another percentage point to {{< var risk.late_noext_iII_cII_FNA_III >}}%.
+When the tumor additionally crosses the mid-sagittal line, the risk may reach around {{< var risk.late_ext_iII_cII_FNA_III >}}%.
 
@@
 
-Our model's prediction for the contralateral LNL IV show that even in the most extreme case, with an advanced T-category tumor extending over the mid-sagittal line as well as metastases in every ipsilateral LNL and every contralateral upstream level, the risk stays just below 3%<!-- TODO: make dynamic -->.
-We only observe a risk of 5-6% when the involvement of the upstream level III is pathologically confirmed after taking a biopsy using FNA.
+Our model's prediction for the contralateral LNL IV show that even in the most extreme case, with an advanced T-category tumor extending over the mid-sagittal line as well as metastases in the ipsilateral LNLs II-IV and the contralateral upstream levels II and III, the risk stays at just {{< var risk.late_ext_iII+III+IV_cII+III_IV >}}%.
+We only observe a risk of {{< var risk.late_ext_iII+III+IV_cII+III_FNA_IV >}}% when the involvement of the upstream level III is pathologically confirmed after taking a biopsy using FNA.
 This is because the model's prior probability for contralateral LNL IV involvement is so low that even given a clinical diagnosis in that level, a false positive in LNL III is a more likely explanation.
 This possibility, however, drops virtually to zero when tumor cells are pathologically confirmed in level III.
+
+## Contralateral LNLs I, V, and VII
+
+Even for advanced T-category patients whose tumor crosses the midline and who present with extensive ipsi- and contralateral clinical involvement, the risk for involvement in the LNLs I and VII never exceeds 3%.
+LNL V contralaterally also shows only a very low risk for occult disease, but it does rise to {{< var risk.late_ext_iI+II+III+IV+V+VII_cII+III+IV_FNA_V >}}% in the same extreme case when additionally the upstream level IV involvement has been confirmed with FNA.
 
@@
 - LNL IV only when upstream level III pathologically confirmed
+- LNLs I and VII only when clinically involved
+- LNL V only in extreme cases (adv. T-cat., midext, all ipsi LNLs metastatic, contra II, III, and IV clinically involved plus cIV via FNA confirmed)
 

### 2024-09-03 scripts/compile_variables.py

@@ -3,18 +3,25 @@
 from pathlib import Path
 from typing import Any
 
+import h5py
 import lydata  # noqa: F401
+import numpy as np
 import pandas as pd
+import paths
 import shared
 import yaml
 from lydata.accessor import Q, QueryPortion
 from pandas.api.typing import DataFrameGroupBy
 from pydantic_settings import BaseSettings
 
+from lyscripts.scenario import Scenario
+
 
 class CmdSettings(BaseSettings, cli_parse_args=True):
     """Settings for the command-line arguments."""
 
+    risks: Path = paths.model_dir / "full" / "risks.hdf5"
+    scenarios: Path = paths.scenario_dir / "risks.yaml"
     output: Path = Path("_variables.dynamic.yaml")
     precision: int = 2
 
@@ -175,9 +182,60 @@ def get_model_variables(precision: int = 2) -> dict[str, Any]:
     return variables
 
 
-def get_risk_variables(precision: int = 2) -> dict[str, Any]:
+def get_short_label(
+    side_diagnosis,
+    prefix: str = "i",
+    joiner: str = "+",
+) -> str:
+    """Create a short label from a diagnosis."""
+    label = ""
+    is_n0 = True
+
+    positive = [
+        lnl for lnl, status in side_diagnosis.get("CT", {}).items() if status
+    ]
+    if len(positive) > 0:
+        is_n0 = False
+        label += prefix + joiner.join(positive)
+
+    if "FNA" in side_diagnosis:
+        is_n0 = False
+        label += "_FNA"
+
+    return label if not is_n0 else f"{prefix}N0"
+
+
+def get_variable_name(scenario: Scenario) -> str:
+    """Get the variable name for the scenario."""
+    return "_".join([
+        str(scenario.t_stages[0]),
+        "ext" if scenario.midext else "noext",
+        get_short_label(scenario.diagnosis.get("ipsi", {}), prefix="i"),
+        get_short_label(scenario.diagnosis.get("contra", {}), prefix="c"),
+        list(scenario.involvement["contra"].keys()).pop()
+    ])
+
+
+def get_risk_variables(
+    risks_file: Path,
+    scenarios_file: Path,
+    precision: int = 2,
+) -> dict[str, Any]:
     """Get the variables for the risk calculations."""
+    with h5py.File(risks_file, mode="r") as file:
+        risks = {key: dset[:] for key, dset in file.items()}
+
+    with open(scenarios_file, mode="r", encoding="utf-8") as file:
[…]

### 2024-09-03 scripts/paths.py

@@ -1,5 +1,5 @@
-"""Define paths for data and models.
-"""
+"""Define paths for data and models."""
+
 from pathlib import Path
 from typing import Literal
 
@@ -8,6 +8,7 @@ data = data_dir / "reduced.csv"
 model_dir = Path("models")
 cache_dir = Path(".cache")
 figure_dir = Path("figures")
+scenario_dir = Path("scenarios")
 params = Path("_variables.yml")
 
 

### 2024-09-04 manuscript.qmd
make prevalences dynamic

@@
 T-category and midline extension.
-As observed in the data, the model predicts that the prevalence of contralateral LNL II involvement increases from below 8% <!--  TODO: make this dynamic --> for early T-category lateralized tumors to almost 40% <!-- TODO: make this dynamic --> when the tumor is of advanced T-category and crosses the mid-sagittal line.
-Similarly, the prevalence of contralateral LNL III involvement increases from around 2% <!--  TODO: make this dynamic --> for early T-category lateralized tumors to almost 15% <!-- TODO: make this dynamic --> when the tumor is of advanced T-category and crosses the mid-sagittal line.
+As observed in the data, the model predicts that the prevalence of contralateral LNL II involvement increases from {{< var prevalence.early_noext_iN0_cII.predicted >}}% for early T-category lateralized tumors to {{< var prevalence.late_ext_iN0_cII.predicted >}}% when the tumor is of advanced T-category and crosses the mid-sagittal line.
+Similarly, the prevalence of contralateral LNL III involvement increases from around {{< var prevalence.early_noext_iN0_cIII.predicted >}}% for early T-category lateralized tumors to almost {{< var prevalence.late_ext_iN0_cIII.predicted >}}% when the tumor is of advanced T-category and crosses the mid-sagittal line.
 
@@
 The model's performance w.r.t.
-its ability to describe the data on contralateral nodal involvement was evaluated based on a dataset of 833<!-- TODO: make this dynamic --> patients from four institutions.
+its ability to describe the data on contralateral nodal involvement was evaluated based on a dataset of {{< var data.num_patients >}} patients from four institutions.
 To the best of our knowledge, this is the first time lymphatic tumor progression has been modelled in such detail and to such an extent.

### 2024-09-04 scripts/compile_variables.py

@@ -20,8 +20,10 @@ from lyscripts.scenario import Scenario
 class CmdSettings(BaseSettings, cli_parse_args=True):
     """Settings for the command-line arguments."""
 
-    risks: Path = paths.model_dir / "full" / "risks.hdf5"
-    scenarios: Path = paths.scenario_dir / "risks.yaml"
+    risk_hdf5: Path = paths.model_dir / "full" / "risks.hdf5"
+    risk_scenarios: Path = paths.scenario_dir / "risks.yaml"
+    prevalence_hdf5: Path = paths.model_dir / "full" / "prevalences_overall.hdf5"
+    prevalence_scenarios: Path = paths.scenario_dir / "overall.yaml"
     output: Path = Path("_variables.dynamic.yaml")
     precision: int = 2
 
@@ -64,6 +66,59 @@ def variables_from_portion(
         f"{key}_total": portion.total,
         f"{key}_percent": round(portion.percent, precision),
     }
+
+
+def get_diagnosis_label(
+    side_diagnosis,
+    modality: str = "CT",
+    prefix: str = "i",
+    joiner: str = "+",
+) -> str:
+    """Create a short label from a diagnosis."""
+    label = ""
+    is_n0 = True
+
+    positive = [
+        lnl for lnl, status in side_diagnosis.get(modality, {}).items() if status
+    ]
+    if len(positive) > 0:
+        is_n0 = False
+        label += prefix + joiner.join(positive)
+
+    if "FNA" in side_diagnosis:
+        is_n0 = False
+        label += "_FNA"
+
+    return label if not is_n0 else f"{prefix}N0"
+
+
+def get_variable_name_for_prevalence(scenario: Scenario) -> str:
+    """Get the variable name for the scenario."""
+    return "_".join([
+        str(scenario.t_stages[0]),
+        "ext" if scenario.midext else "noext",
+        get_diagnosis_label(
+            scenario.diagnosis.get("ipsi", {}),
+            modality="max_llh",
+            prefix="i",
+        ),
+        get_diagnosis_label(
+            scenario.diagnosis.get("contra", {}),
+            modality="max_llh",
+            prefix="c",
+        ),
+    ])
+
+
+def get_variable_name_for_risk(scenario: Scenario) -> str:
+    """Get the variable name for the scenario."""
+    return "_".join([
+        str(scenario.t_stages[0]),
+        "ext" if scenario.midext else "noext",
+        get_diagnosis_label(scenario.diagnosis.get("ipsi", {}), prefix="i"),
[…]

### 2024-09-18 manuscript.qmd
testing overleaf setup
first changes

--- 2024-09-03 manuscript.qmd
@@
 ---
-title: "Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer"
+title: "A mixture of hidden Markov models to predict the lymphatic spread in head and neck cancer depending on primary tumor location"
 abstract: |
-  Purpose: According to current guidelines for elective nodal irradiation of oropharyngeal squamous cell carcinoma (OPSCC) patients, large parts of the contralateral lymphatic system are included in the elective clinical target volume (CTV-N), even for lateralized tumors without clinical lymph node involvement in the contralateral neck.
-In this work, we present a probabilistic model for bilateral lymphatic tumor progression in OPSCC to predict the personal risk for occult disease in any lymph node level (LNL), given the patient's clinical lymph node involvement, T-stage, and lateralization of the primary tumor.
+  Purpose: to be done
 
-  Methods: We extend a hidden markov model for lymphatic tumor progression, which was previously developed for ipsilateral lymph node involvement, to the contralateral neck.
-The model represents each of the LNLs I, II, III, IV, V, and VII of both sides of the neck as a hidden binary random variable with states healthy and involved.
-LNLs are connected to the tumor and among each other via arcs that correspond to spread probabilities.
-These spread probability rates are learned via Markov chain Monte Carlo (MCMC) sampling from a dataset of 833 OPSCC patients.
+  Methods: to be done
 
-  Results: The model is able to describe the data on lymph node involvement well with a small number of interpretable parameters.
-Midline extension of the primary tumor is the main risk factor for contralateral involvement.
-In addition, the risk of occult metastases in contralateral lymph node level increases with more advanced T-stage and more severe ipsilateral involvement.
-The probability of involvement in contralateral level III is very low if the upstream level II is clinically negative.
-Similarly, the probability of involvement in contralateral level IV is very low if the upstream level III is clinically negative.
+  Results: to be done
 
-  Conclusions: The model may guide personalized volume reduction of the elective CTV-N.
-It suggests that the contralateral neck may be spared from elective irradiation for lateralized tumors not crossing the midline.
-In patients with primary tumors crossing the midline but clinically negative contralateral neck, the contralateral elective irradiation may be limited to level II.
+  Conclusions: to be done
 ---
 
-```{python}
-#| echo: false
-import re
+# Introduction
+Head and Neck Squamous Cell Carcinoma (HNSCC) are known to spread through the lymphatic system and form metastases in lymph nodes [@mukherji_cervical_2001, @shah_patterns_1990].
+To avoid nodal recurrences, lymph node levels (LNL) at risk of harbouring occult metastases are electively irradiated.
+Guidelines for different tumor locations [@biau_selection_2019] are based on the overall prevalence of nodal disease as reported in the literature [@mukherji_cervical_2001, @shah_patterns_1990].
+To personalize the prediction of the risk of occult metastases, given a patient's individual diagnosis, we first published a large, multi-centric dataset where the lymphatic involvement per LNL is available for each patient[@ludwig_dataset_2022, @ludwig_multi-centric_2023].
+Subsequently we published an interpretable hidden Markov model (HMM), trained with the data to predict the risk for occult nodal disease, given an individual patient's diagnosis [@ludwig_hidden_2021].
+A personalized risk prediction may allow clinicians to safely reduce the elective clinical target volume (CTV-N) and reduce side-effects that degrade the patient's quality of life without compromising treatment efficacy [@batth_practical_2014].
 
-import pandas as pd
-import numpy as np
-import matplotlib.pyplot as plt
-from matplotlib import ticker
-from matplotlib.colors import LinearSegmentedColormap
-import upsetplot
-
-from lymph import models
-from lyscripts import utils
-from lyscripts.plot.utils import COLORS
-
-from scripts import shared, paths
-from scripts.shared import COL, get_lnl_cols
-
-simple_model = shared.get_model("simple", load_samples=True)
-```
-
-```{python}
-#| tags: [parameters]
-#| echo: false
-
-# width of figures. May depend on number of text columns
-full = 17 # cm
-half = full
-# half =  8 # cm
-```
-
-# Introduction
-
-When treating head and neck squamous cell carcinomas (HNSCC) with radiotherapy or surgery, not only the primary tumor and the clinically detected lymph node metastases are targeted.
-Per current guidelines, large volumes of the neck are part of the the elective clinical target volume (CTV-N) [@gregoire_ct-based_2003;@gregoire_delineation_2014;@gregoire_delineation_2018;@eisbruch_intensity-modulated_2002;@biau_selection_2019;@chao_determination_2002;@vorwerk_guidelines_2011;@ferlito_elective_2009].
-This is to minimize the risk of regional recurrences due to untreated microscopic disease which cannot be detected by current in-vivo imaging modalities such as computed tomography (CT), magnetic resonance imaging (MRI), or positron emission tomography (PET).
-However, minimizing the risk of missing occult disease in the lymph drainage region must be balanced against the toxicity related to unnecessary treatment of healthy tissue.
-
-The aforementioned guidelines for the definition of the CTV-N are based on anatomically defined lymph node levels (LNL) [@gregoire_delineation_2014] and the overall prevalence of lymph node metastases in the LNLs.
-Based on this, they recommend extensive irradiation of both sides of the neck for a majority of patients.
-However, this prevalence the guidelines are based on does not take into account an individual patient's state of tumor progression, which may have a great impact on a patient's personal risk for occult disease in a given LNL.
-For example, a patient with substantial nodal involvement in the ipsilateral side and an advanced primary tumor that crosses the mid-sagittal plane would receive extensive elective irradiation of the contralateral LNLs II, III, and IVa.
-However, if the patient presented with no clinically detectable nodal disease (cN0) and only a small, clearly lateralized T1 tumor, the same contralateral CTV-N would be recommended [@biau_selection_2019].
-
-To better quantify this personalized risk of occult disease we have previously developed an intuitive, probabilistic hidden Markov model [@ludwig_hidden_2021;@ludwig_modelling_2023] which was originally based on a conceptually similar Bayesian network model [@pouymayou_bayesian_2019].
-So far, these models were limited to the ipsilateral side of the neck.
-The main contribution of this work will be to extend the formalism to include risk predictions for the contralateral side as well.
-We believe this holds the potential to reduce the electively irradiated volumes in the contralateral neck for patients whose individual risk for occult disease is low.
-In turn, this may substantially reduce the toxicity from irradiation and the associated decrease in quality of life.
-
-The main contributions of this paper are as follows:
-
-1.
-In section @sec-data we present a multi-centric dataset on lymph node involvement in {{< var data.num_patients >}} OPSCC patients.
-Based on the dataset the main risk factors for contralateral lymph node involvement are identified and the requirements for the bilateral model extension are described (@sec-requirements).
-
-2.
-In @sec-ext-to-contra we present a bilateral HMM of lymphatic progression that accounts for the T-stage lateralization of the primary tumor as well as clinical involvement as risk factors for occult contralateral involvement.
-In @sec-methods we describe how the model training and the computational experiments were set up.
-
-3.
-In the results in @sec-results, we demonstrate that the model is able to describe the patterns of contralateral lymph node involvement observed in the dataset, and we apply the model to estimating risk of occult disease for typical patients.
-In @sec-discussion, implications for volume-deescalated radiotherapy are discussed.
+So far we trained different models for different tumor locations such as oropharynx and oral cavity tumors.
+However, this approach does not describe differences in lymphatic spread between different subsites within the oropharynx and oral cavity.
+This resulted in distorted predictions for each subsite, as the spread patterns were pooled into one single model .To address this issue, we present an approach using mixtures of HMMs.
+The intuition is that the lymphatic spread of a tumor that lies anatomically at the border of oropharynx and oral cavity (e.g.
+tumors in the palate) may be described by a mixture of different models.
 
@@
 
-To be able to create models for lymphatic tumor progression that include all relevant LNLs including the contralateral side, we have collected a detailed dataset of {{< var data.num_patients >}} patients with newly diagnosed oropharyngeal squamous cell carcinomas [@ludwig_dataset_2022;@ludwig_multi-centric_2023].
-It reports the lymph node involvement per LNL of every patient individually in tabular form, in addition to other primary tumor and patient characteristics such as T-category, subside, and lateralization of the primary tumor, and HPV p16 status.
-Their patient records have been collected at four different institutions and a brief overview over some of their patients' characteristics are shown in @tbl-data-overview.
-The data from the Inselspital Bern and the Centre Léon Bérard only consist of patients who received a neck dissection.
-Since this treatment is more commonly chosen for early T-category patients, they also make up a larger portion of the respective dataset.
-The majority of patients from the University Hospital Zürich and the Vall d'Hebron Barcelona Hospital was treated with definitive radiotherapy.
-For a subset of 83 patients in the CLB dataset, the lateralization of the primary tumor was not reported.
-
-```{python}
-#| echo: false
-#| label: tbl-data-overview
-#| tbl-cap: Overview over the five datasets from four different institutions used to train and evaluate our model.
-Here, we briefly characterize the total number of OPSCC patients from the respective institution, their median age, what proportion received some form of neck dissection, the N0 portion of patients, what percentage presented with early T-category, and the prevalence of primary tumor midline extension.
-For a much more detailed look at the data, visit [lyprox.org](https://lyprox.org).
-from typing import Literal
-
-def align(
-  column: pd.Series,
-  where: Literal["right", "left", "center"],
-) -> list[str]:
-  """Align a column."""
-  return [f"text-align: {where}"] * len(column)
-
-def right_align(column: pd.Series) -> list[str]:
-  """Right align column."""
-  return align(column, where="right")
-
-def highlight(column: pd.Series, mapping: dict) -> list[str]:
-  """Color based on `mapping`."""
-  colors = column.map(mapping)
-  return colors.map(lambda x: f"color: {x}")
-
-def highlight_bool(
-  column: pd.Series,
-  c_false: str = COLORS["green"],
-  c_true: str = COLORS["red"],
-) -> list[str]:
-  """Make cell read (`False`) or green (`True`)."""
-  return highlight(column, mapping={True: c_true, False: c_false})
-
-def highlight_t_stage(
-  column: pd.Series,
-  c_early: str = COLORS["green"],
-  c_late: str = COLORS["red"],
-) -> list[str]:
-  """Highlight `early` and `late`."""
-  is_early = column == "early"
-  return highlight_bool(is_early, c_false=c_late, c_true=c_early)
-
-col_map = {
-  COL.inst: "Institution",
-  COL.age: "Age",
-  COL.nd: "Neck Dissection",
-  COL.t_stage: "T-category",
-  COL.n_stage: "N-category",
-}
-short_col_map = {tpl[-1]: val for tpl, val in col_map.items()}
-
-raw = utils.load_patient_data(paths.data)
-subdata = raw[col_map.keys()]
-subdata.columns = subdata.columns.get_level_values(2)
-subdata = (
-  subdata
-  .rename(columns=short_col_map)
-  .reset_index(drop=True)
-)
-
-def n0_mean(n_stage: pd.Series) -> float:
-  """Compute portion of N0 patients."""
-  return (n_stage == 0).mean()
-
-def early_mean(t_stage: pd.Series) -> float:
-  """Compute early T-category portion."""
-  return t_stage.isin([0,1,2]).mean()
-
-grouped = (
-  raw.groupby(
-    by=COL.inst,
-  ).aggregate(**{
-    "Total": pd.NamedAgg(column=COL.age, aggfunc="count"),
-    "Age (median)": pd.NamedAgg(column=COL.age, aggfunc="median"),
-    "Neck Dissection": pd.NamedAgg(column=COL.nd, aggfunc="mean"),
-    "N0": pd.NamedAgg(column=COL.n_stage, aggfunc=n0_mean),
-    "Early T-Cat.": pd.NamedAgg(column=COL.t_stage, aggfunc=early_mean),
-    "Mid. Ext.": pd.NamedAgg(column=COL.midext, aggfunc="mean"),
-  }).convert_dtypes()
-)
-grouped.index.name = "Institution"
-(
-  grouped
-  .reset_index()
-  .style
-  .format(
-    formatter="{:>.0%}",
-    subset=["Neck Dissection", "N0", "Early T-Cat.", "Mid. Ext."],
-  )
-  .apply(
-    func=right_align,
-    axis="index",
-    subset=grouped.columns[0:],
-  )
-  .hide()
-)
-```
-
-## Consensus on Involvement Status {#sec-data-consensus}
-
-Pathological involvement is thus available only for a subset of the patients and only for the dissected levels.
-For the remaining patients, only clinical involvement is available.
-For the analysis performed in this paper, all available diagnostic information has been combined into a consensus decision representing the most likely state of involvement for each patient and LNL, using literature values for sensitivity and specificity of different modalities [@de_bondt_detection_2007;@kyzas_18f-fluorodeoxyglucose_2008].
-These values are tabulated in @tbl-spec-sens.
-
-This consensus was formed as follows: Suppose the involvement status $X_v$ of LNL $v$ was assessed using different diagnostic modalities $\mathcal{O} = \{ \text{MRI}, \text{CT}, \text{pathology}, \ldots \}$, each characterized by their own pair of sensitivity and specificity values $s_N^{\mathcal{o}}$ and $s_P^{\mathcal{o}}$, with $\mathcal{o} \in \mathcal{O}$.
-Then we have $|\mathcal{O}|$ observations $z_v^{\mathcal{o}} \in \left[ 0, 1 \right]$, where 0 stands for "healthy" and 1 for "involved".
-We can then compute the most likely true involvement $X_v$ using the likelihood function
-
-$$
-\begin{aligned}
-\ell \left( X_v \mid \{ z_v^{\mathcal{o}} \}_{\mathcal{o} \in \mathcal{O}} \right) = \prod_{\mathcal{o} \in \mathcal{O}}
-\left( 1 - X_v \right) \cdot &\left[ z_v^{\mathcal{o}} \cdot \left( 1 - s_P^{\mathcal{o}} \right) + \left( 1 - z_v^{\mathcal{o}} \right) \cdot s_P^{\mathcal{o}} \right] \\
-+ X_v \cdot &\left[ z_v^{\mathcal{o}} \cdot s_N^{\mathcal{o}} + \left( 1 - z_v^{\mathcal{o}} \right) \cdot (1 - s_N^{\mathcal{o}}) \right]
-\end{aligned}
-$$
-
-We now assume the true state $X_v$ to take on the value 1 if $\ell \left( X_v = 1 \mid \ldots \right) > \ell \left( X_v = 0 \mid \ldots \right)$ and 0 otherwise.
-For example, if we have $z_\text{II}^\text{CT} = 0$ and $z_\text{II}^\text{MRI} = 1$ we would compute the following likelihoods:
-
-$$
-\begin{aligned}
-\ell \left( X_\text{II} = 1 \mid z_\text{II}^\text{CT} = 0, z_\text{II}^\text{MRI} = 1 \right) &= \left( 1 - s_N^\text{CT} \right) \cdot s_N^\text{MRI} = 15.39\% \\
-\ell \left( X_\text{II} = 0 \mid z_\text{II}^\text{CT} = 0, z_\text{II}^\text{MRI} = 1 \right) &= s_P^\text{CT} \cdot \left(1 - s_N^\text{MRI}\right) = 14.44\%
-\end{aligned}
-$$
-
-In this example, we would thus assume the true state to be involved ($X_\text{II} = 1$).
-
-This method of computing a consensus also ensures that the pathology reports always override any conflicting clinical diagnosis, due to pathology's high sensitivity and specificity.
-
-| Modality  | Specificity | Sensitivity |
-| :-------- | ----------: | ----------: |
-| CT        |         76% |         81% |
-| PET       |         86% |         79% |
-| MRI       |         63% |         81% |
-| FNA       |         98% |         80% |
-| pathology |        100% |        100% |
-
-: Specificity and sensitivity values from the literature [@de_bondt_detection_2007;@kyzas_18f-fluorodeoxyglucose_2008]. {#tbl-spec-sens}
-
-## Data Availability
-
-The entire data, including additional patients with tumors in other primary locations than the oropharynx, is publicly available: It may be [downloaded from LyProX](https://lyprox.org/patients/dataset) where it can be interactively explored too, [from GitHub](https://github.com/rmnldwg/lydata), [from zenodo](https://zenodo.org/search?q=lydata), or via the *Data-in-Brief* publications @ludwig_dataset_2022 and @ludwig_multi-centric_2023.
-The Data-in-Brief publications contain a detailed description of the datasets and the data format.
-However, these publications do not include the most recent dataset addition from Vall d'Hebron Barcelona Hospital.
-
-## Patterns of Contralateral Involvement {#sec-data-strat}
-
-These datasets allow us to investigate correlations between the involvement of individual LNLs, or between risk factors and involvement.
-In @fig-data-strat, we have plotted the prevalence of each contralateral LNL's involvement, stratified by risk factors such as T-category, number of ipsilaterally involved LNLs, and whether the tumor extended over the mid-sagittal line.
-We also show in the fourth subfigre that T-category and ipsilateral involvement are not just rpedictive of contralateral involvement because they correlate with midline extension.
-A similar but more complete stratification is also tabulated in the appendix in @tbl-data-strat.
-
-![Contralateral involvement stratified by T-category (top left panel), the number of metastatic LNLs ipsilaterally (top right panel), and whether the primary tumor extended over the mid-sagittal line or was clearly lateralized (bottom left panel).
-Additionally, in the bottom right, we show that even for lateralized tumors, T-category and ipsilateral involvement are correlated with contralateral metastasis by comparing the contralateral involvement prevalence for selected scenarios that vary in their T-category and ipsilateral involvement extent.](figures/fig_data_strat.svg){#fig-data-strat}
-
-The top left panel in @fig-data-strat indicates that T-category is correlated with contralateral involvement (as it is with overal involvement).
-This is simply because T-category may on average be considered a surrogate for the time between onset of disease and diagnosis.
-I.e., a patient with a T4 tumor was -- on average -- diagnosed later than a patient with a T1 tumor.
-Thus, the former did have more time to develop metastases.
-
-Similarly, ipsilateral involvement correlates with contralateral metastasis.
-The tumor of a patients with many metastases in ipsilateral LNLs was probably able to spread for longer (or faster) compared to a tumor in a patient with no nodal disease.
-This, too, may therefore be considered a surrogate for the duration of the disease.
-In addition, it has been hypothesized that bulky nodal disease ipsilaterally may also redirect lymph fluids to the contralateral side.
-<!-- citation needed?
--->
-
-The bottom left panel in @fig-data-strat shows that patients with a tumor crossing the mid-sagittal line show contralateral involvement vastly more often compared to patients with clearly lateralized tumors.
-This makes intuitive sense, because the lymphatic system in the head and neck region is typically symmetric and thus no major vessels cross the midline.
-Therefore, interstitial fluids from the primary tumor -- which we assume to carry living malignant cells -- may only reach the blind-ended lymphatic vessels in the contralateral neck via short-ranged diffusion.
-Which in turn is only possible when the primary tumor is close enough to the mid-sagittal line or crosses it.
-<!-- citation(s) needed?
-or is this intuition good enough?
--->
-
-The three risk factors for contralateral involvement, midline extension, T-category, and ipsilateral involvement are correlated.
-For example, midline extension was present in {{< var data.late_with_midext_percent >}}% of advanced T-category tumors but only in {{< var data.early_with_midext_percent >}}% of early T-category tumors.
-The higher prevalence of contralateral metastases in advanced T-category tumors may thus be partially explained by a higher rate of midline extension in these tumors.
-However, the bottom right panel of @fig-data-strat indicates that advanced T-category and severe ipsilateral involvement do represent additional risk factors.
-The figure considers only patients with lateralized tumors not extending over the midline.
-For early T-category and healthy ipsilateral levels I-V, only {{< var data.early_ipsin0_cII_percent >}}% ({{< var data.early_ipsin0_cII_match >}} out of {{< var data.early_ipsin0_cII_total >}} patients) shows involvement of contralateral level II.
-This increases to {{< var data.early_ipsiII_cII_percent >}}% ({{< var data.early_ipsiII_cII_match >}} out of {{< var data.early_ipsiII_cII_total >}} patients) if ipsilateral level II is involved, to {{< var data.early_ipsiIIandIII_cII_percent >}}% ({{< var data.early_ipsiIIandIII_cII_match >}} out of {{< var data.early_ipsiIIandIII_cII_total >}} patients) if ipsilateral levels II and III are involved, and to {{< var data.late_ipsiIIandIII_cII_percent >}}% ({{< var data.late_ipsiIIandIII_cII_match >}} out of {{< var data.late_ipsiIIandIII_cII_total >}} patients) for advanced T-category with ipsilateral levels II and III involved.
-
-## Requirements for a Bilateral Model {#sec-requirements}
-
-Based on the observations of the [previous section](#sec-data-strat), any model to predict the risk for contralateral nodal involvement, should be able to describe the following:
-
-1.
-More advanced T-category should lead to higher risk for nodal disease.
-One approach to achieve this via the expected time of diagnosis has already been developed in the form of a hidden Markov model [@ludwig_hidden_2021].
-2.
-The degree of ipsilateral involvement should give the model information on the time that may have passed between onset and diagnosis of the disease.
-This should come in addition to what can be inferred about the time from T-category alone.
-3. A tumor that extends over the mid-sagittal line should yield contralateral metastases with much higher probability.
-
-Over the course of this work, we will first briefly recap the HMM in @sec-unilateral, which was so far used to model ipsilateral lymphatic progression only.
-Then, we intuitively extend it to include the contralateral side as well in @sec-ext-to-contra.
-In this section, we also introduce a way of modelling the tumor's midline extension as a random variable (@sec-midline) and lastly talk about how it may affect the contralateral spread in @sec-params-symmetry.
-
-<!-- TODO: Hint at methods and results... -->
-
 # Unilateral Model for Lymphatic Progression {#sec-unilateral}
 
-This paper builds on the previously developed unilateral model for ipsilateral lymph node involvement presented in [cite 2024 graph extension paper].
-In this section we provide a brief recap of the unilateral model to introduce the notation needed to extend the framework to a bilateral model describing both ipsilateral and contralateral lymph node involvement in @sec-ext-to-contra.
-For further details in the ipsilateral model, the reader is referred to the earlier publications [@ludwig_hidden_2021, cite 2024 graph extension paper]
+# Mixture Model for Lymphatic Spread {#sec-mixture}
 
-<!--
-Our first model to predict the lymphatic progression of HNSCC was introduced using Bayesian networks [@pouymayou_bayesian_2019].
-We subsequently extended this work to a hidden Markov model (HMM) [@ludwig_hidden_2021] to allow an intuitive inclusion of T-category into the predictions.
-We will briefly summarize this HMM's formalism before building on it to include the contralateral spread in @sec-ext-to-contra.
--->
+Since primary tumors at different subsites have different patterns of lymphatic spread patterns one would need to either generalize over several subsites as introduced in [@ludwig_dynamic_2021], where the HMM has been introduced for oropharyngeal cancer, or train a distinct model for each subsite.
+The former approach would lead to less precise predictions for each subsite, especially those subsites with few patients.
+Training one HMM per subsite would be rather time intensive and lead to large uncertainties for subsites with only few patients.
+Further we would not exploit the anatomical similarities which are expected for anatomically nearby locations.
 
-We model a patient's state of involvement at an abstract time-step $t$ as a vector of hidden binary random variables:
-
-$$
-\mathbf{X}[t] = \begin{pmatrix} X_v[t] \end{pmatrix} \qquad v \in \left\{ 1, 2, \ldots, V \right\}
-$$ {#eq-state-def}
-
-Here, $V$ is the number of LNLs the model considers.
-The values a LNL's hidden binary random variables may take on are $X_v[t] = 0$ (`False`), meaning the LNL $v$ is healthy or free of metastatic disease, or $X_v[t] = 1$ (`True`), corresponding to some form of tumor presence (i.e., occult or clinically detected).
-Since the state vector $\mathbf{X}[t]$ is $V$-dimensional and binary, there are $2^V$ distinct possible lymphatic involvement patterns, which we enumerate from $\boldsymbol{\xi}_0 = \begin{pmatrix} 0 & 0 & \cdots & 0 \end{pmatrix}$ to $\boldsymbol{\xi}_{2^V} = \begin{pmatrix} 1 & 1 & \cdots & 1 \end{pmatrix}$.
-In addition, each LNL is associated with an observed binary random variable $Z_v$ that describes the clinical involvement of a LNL based on imaging: $Z_v = 0$ (`False`) indicates that the LNL $v$ is healthy based on clinical diagnosis, and $Z_v = 1$ (`True`) indicates that suspicious lymph nodes were detected that are deemed metastatic.
-
-Any hidden Markov model is fully described by three quantities:
-
-1. A starting state $\mathbf{X}[t=0]$ at time $t=0$ just before the patient's tumor formed. In our case, this is always the state where all LNLs are still healthy $\boldsymbol{\xi}_0$.
-2. The *transition matrix*
-   $$
-   \mathbf{A} = \left( A_{ij} \right) = \big( P \left( \mathbf{X}[t+1] = \boldsymbol{\xi}_j \mid \mathbf{X}[t] = \boldsymbol{\xi}_i \right) \big)
-   $$ {#eq-trans-matrix}
-   where the value at row $i$ and column $j$ represents the probability to transition from state $\boldsymbol{\xi}_i$ to $\boldsymbol{\xi}_j$ during the time-step from $t$ to $t+1$.
-Note that we prohibit self-healing, meaning that during a transition, no LNL may change their state from $X_v[t]=1$ to $X_v[t+1]=0$.
-Consequently, many elements of the transition matrix are zero.
-3. Lastly, the *observation matrix*
-   $$
-   \mathbf{B} = \left( B_{ij} \right) = \big( P \left( \mathbf{Z} = \boldsymbol{\zeta}_j \mid \mathbf{X}[t_D] = \boldsymbol{\xi}_i \right) \big)
-   $$ {#eq-obs-matrix}
-   where in row $i$ and at column $j$ we find the probability to *observe* a lymphatic involvement pattern $\mathbf{Z} = \boldsymbol{\zeta}_j$, given that the true (but hidden) state of involvement at the time of diagnosis $t_D$ is $\mathbf{X}[t_D] = \boldsymbol{\xi}_i$.
-
-The transition matrix $\mathbf{A}$ is parametrized using a directed acyclic graph (DAG) as an abstract representation of the underlying lymphatic network.
-Directed arcs from the primary tumor to a LNL are associated with a probability $b_v$ for direct spread to LNL $v$ during one time step.
-Directed arcs from a LNL $v$ to a LNL $r$ are associated with a probability $t_{vr}$ for the tumor to progress from one LNL to the next.
-In this paper, we build on the DAG shown in @fig-full-graph which was obtained by maximizing the model evidence as described in [cite 2024 graph extension paper].
-
-![Directed acyclic graph (DAG) representing the abstract lymphatic network in the head and neck region.
-Blue nodes are the LNLs' hidden random variables, the red node represents the tumor, and the orange square nodes depict the binary observed variables.
-Red and blue arcs symbolize the probability of lymphatic spread along that edge during one time-step.
-The orange arcs represent the sensitivity and specificity of the observational modality (e.g.
-CT, MRI, pathology, ...).](static/full-graph.svg){#fig-full-graph width=40%}
-
-Using the introduced quantities, we can evolve the distribution of all possible hidden states from $\mathbf{X}[t=0] = \boldsymbol{\xi}_0$ step by step, by successively multiplying this vector with the transition matrix $\mathbf{A}$.
-For later use, we define at this point the matrices of the dsitributions over all hidden states, given all time-steps $\boldsymbol{\Lambda}$:
-
-$$
-\boldsymbol{\Lambda} = P \left( \mathbf{X} \mid \mathbf{t} \right) = \begin{pmatrix}
-\boldsymbol{\pi}^\intercal \cdot \mathbf{A}^0 \\
-\boldsymbol{\pi}^\intercal \cdot \mathbf{A}^1 \\
-\vdots \\
-\boldsymbol{\pi}^\intercal \cdot \mathbf{A}^{t_\text{max}} \\
-\end{pmatrix}
-$$ {#eq-lambda-matrix}
-
-Where the $k$-th row in this matrix corresponds to the probability distribution over hidden states after $t=k-1$ time-steps.
-
-At the time of the diagnosis $t_D$, we multiply the result with the observation matrix $\mathbf{B}$.
-We may then look up the likelihood of a patient presenting with the diagnosis $\mathbf{Z}=\boldsymbol{\zeta}_i$ in the $i$-th entry of the final result.
-However, the remaining issue is that the value of $t_D$ is unknown, i.e.
-over how many time-steps the HMM should be evolved.
-We solve this problem by marginalizing over the time of diagnosis.
-Different distributions over the diagnosis times can then be choosen based on T-category.
-For instance, the mean of the time-prior to marginalize over the diagnosis time for early T-category patients $P\left( t_D \mid \text{early} \right)$ may be shifted towards earlier times than the one for advanced T-category patients $P\left( t_D \mid \text{early} \right)$.
-This gives us for example
-
-$$
-P\left( \mathbf{X} \mid \text{T}x = \text{early} \right) = \sum_{t=0}^{t_\text{max}} P \left( \mathbf{X} \mid t \right) \cdot P(t \mid \text{early})
-$$
-
-In this work, we use binomial distributions $\mathfrak{B} \left( t_D, p_{\text{T}x} \right)$ as time-priors which have one free parameter $p_{\text{T}x}$ for each group of patients we differentiate based on T-category.
-Also, we fix $t_\text{max} = 10$, which means that the expected number of time-steps from the onset of a patient's disease to their diagnosis is $\mathbb{E}\left[ t_D \right] = 10 \cdot p_{\text{T}x}$.
-
-## Likelihood Function of the Unilateral Model
-
-With the formalism introduced above, we can write the likelihood function for a patient to present with a diagnosis consisting of an observed state and a T-category $d = \left( \boldsymbol{\zeta}_i, \text{T}x \right)$ as follows:
-
-$$
-\ell = P \left( \mathbf{Z} = \boldsymbol{\zeta}_i \mid \text{T}x \right) = \sum_{t=0}^{t_\text{max}} \left[ \boldsymbol{\xi}_0 \cdot \mathbf{A}^t \cdot \mathbf{B} \right]_i \cdot P \left( t \mid \text{T}x \right)
-$$ {#eq-single-patient-llh}
-
-Above, the quantity inside $\left[ \ldots \right]_i$ denotes the $i$-th component of the vector that is the result of the vector and matrix multiplications in the square brackets.
-Note that it is also possible to account for missing involvement information: If a diagnosis (like fine needle aspiration (FNA)) is only available for a subset of all LNLs, we can sum over all those possible complete observed states $\boldsymbol{\zeta}_j$ that match the provided diagnosis.
-
-The single-patient likelihood $\ell$ in @eq-single-patient-llh depends on the spread parameters shown in @fig-full-graph via the transition matrix $\mathbf{A}$ and on the binomial parameters $p_{\text{T}x}$ via time-priors.
-In this work, we will only differentiate between "early" (T1 & T2) and "advanced" (T3 & T4) T-categories.
-Therefore, the parameter space of the unilateral model is:
-
-$$
-\boldsymbol{\theta} = \left( \left\{ b_v \right\}, \left\{ t_{vr} \right\}, p_\text{early}, p_\text{adv.} \right) \quad \text{with} \quad \genfrac{}{}{0pt}{2}{v\leq V}{r\in\operatorname{pa}(v)}
-$$ {#eq-param-space}
-
-And it is our goal to infer the values of these parameters for a given dataset $\mathcal{D} = \left( d_1, d_2, \ldots, d_N \right)$ of OPSCC patients.
-The likelihood of these $N$ diagnoses is simply the product of their individual likelihoods as defined in @eq-single-patient-llh.
-For numerical reasons, we typically compute the data likelihood in log space:
-
-$$
-\log \mathcal{L} \left( \mathcal{D} \mid \boldsymbol{\theta} \right) = \sum_{i=1}^N \log \ell_i
-$$ {#eq-log-likelihood}
-
-The methodology we use to infer the model's parameters is detailed in @sec-sampling.
-
-# Extension to a Bilateral Model {#sec-ext-to-contra}
-
-A naive approach to model the contralateral lymphatic spread would be to simply employ two independent unilateral models as introduced in @sec-unilateral.
-During training, one could enforce that some parameters are shared between these two models, e.g.
-the parameterization of the distributions over diagnose times or the spread among the LNLs ($t_{vr}).
-However, this approach lacks a way to describe the correlation between ipsi- and contralateral involvement discussed in section {#sec-data-strat}.
-Two independent unilateral models would not describe the observation that contralateral involvement becomes more likely with more severe ipsilateral involvement.
-
-<!-- This is displayed in @tbl-data-strat and shows how often the contralateral LNLs I, II, III, and IV were involved, given all possible combinations of midline extension, T-category, and ipsilateral LNL III involvement.
-Unsurprisingly, the prevalence for contralateral involvement is consistently higher when the tumor extends over the mid-sagittal line or is of later T-category.
-But it is also more frequent when the ipsilateral side shows more severe involvement, which is here shown via the surrogate LNL III.
--->
-
-<!-- Additionally, we could think of an approach to incorporate the primary tumor's mid-sagittal extension as a risk factor. -->
-
-Thus, we extend the formalism in @sec-unilateral in such a way that the model's ipsi- and contralateral side evolve synchronously over time.
-To achieve that, we start by writing down the posterior distribution of involvement an analogy to @eq-uni-bayes-law, which is now a joint probability of an involvement $\mathbf{X}^\text{i}$ ipsilaterally *and* an involvement $\mathbf{X}^\text{c}$ contralaterally, given a diagnosis of the ipsilateral LNLs $\mathbf{Z}^\text{i}$ and of the contralateral ones $\mathbf{Z}^\text{c}$:
-
-$$
-P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \mid \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \right) = \frac{P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \mid \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right) P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right)}{P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \right)}
-$$ {#eq-bilateral-bayes}
-
-For the sake of brevity, we omit the dependency on the parameters and the T-category here.
-
-The probability of the diagnoses given a hidden state factorises: $P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \mid \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right) = P \left( \mathbf{Z}^\text{i} \mid \mathbf{X}^\text{i} \right) \cdot P \left( \mathbf{Z}^\text{c} \mid \mathbf{X}^\text{c} \right)$, and the two factors are described through observation matrices $\mathbf{B}^\text{i}$ and $\mathbf{B}^\text{c}$.
-
-The term representing the model's prior probability of hidden involvement does not factorize.
-However, we assume that there is no direct lymph drainage from an ipsilateral LNL to a contralateral LNL.
-No major lymph vessels cross the mid-sagittal plane.
-In the graphical model, this means that we assume no directed arcs between ipsilateral and contralateral LNLs, i.e.
-tumor spread to the contralateral side is due to the primary tumor alone.
-We can thus write the joint probability $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right)$ as a factorising sum:
-
-$$
-\begin{aligned}
-P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right) &= \sum_{t=0}^{t_\text{max}} P(t) \cdot P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \mid t \right) \\
-&= \sum_{t=0}^{t_\text{max}} P(t) \cdot P \left( \mathbf{X}^\text{i} \mid t \right) \cdot P \left( \mathbf{X}^\text{c} \mid t \right)
-\end{aligned}
-$$ {#eq-bilateral-marginal}
-
-This assumption makes intuitive sense: The state of the ipsilateral and contralateral sides of the lymphatic network evolves independently over time as there are no major lymph vessels crossing the midline.
-However, both sides are coupled via time.
-Qualitatively speaking, according to the model in {#eq-bilateral-marginal}, a joint state with severe contralateral involvement and limited ipsilateral involvement would be unlikely, because severe contralateral involvement becomes likely only at later time steps when limited ipsilateral involvement becomes unlikely.
-
-<!-- There may, however, be diffusion of lymph fluid accross this line or bulky involvement that redirects lymphatic drainage significantly.-->
-
-Using {#eq-bilateral-marginal} along with @eq-lambda-matrix, we can write the above distribution algebraically as a product:
-
-$$
-P \left( \mathbf{X}^\text{i} = \boldsymbol{\xi}_n, \mathbf{X}^\text{c} = \boldsymbol{\xi}_m \right) = \left[ \boldsymbol{\Lambda}^\intercal_\text{i} \cdot \operatorname{diag} P(\mathbf{t}) \cdot \boldsymbol{\Lambda}_\text{c} \right]_{n,m}
-$$ {#eq-bilateral-marginal-algebra}
-
-## Parameter Symmetries {#sec-params-symmetry}
-
-In general, the matrices $\boldsymbol{\Lambda}_\text{i}$ and $\boldsymbol{\Lambda}_\text{c}$ could be parameterized using a disjoint set of parameters, i.e., the ipsi- and contralateral spread rates are entirely different.
-However, using three sensible assumptions, we can reduce the parameter space by sharing some parameters between the two sides:
-
-1. We assume that both ipsilateral and contralateral spread is described through the same graph shown in {#fig-full-graph}.
-2.
-We assume the spread *among* the LNLs to be same on both sides.
-It is reasonable to assume the lymphatic system is symmetric.
-Thus, the spread rates from one LNL to the other should be symmetric, too.
-Formally, this means \
-   $$
-   \begin{aligned}
-   b_v^\text{c} &\neq b_v^\text{i} \\
-   t_{rv}^\text{c} &= t_{rv}^\text{i}
-   \end{aligned}
-   $$ {#eq-symmetries}
-   for all $v \leq V$ and $r \in \operatorname{pa}(v)$.
-3.
-The tumor's spread to the contralateral side in case of an extension over the midline is larger than if it was clearly lateralized, but smaller than its spread to the ipsilateral side.
-This assumption stems from a simple thought experiment: Consider moving the tumor from a clearly lateralized position accross the mid-sagittal plane to the same position, but on the contralateral side.
-In the beginning we would have $b_v^\text{c} < b_v^\text{i}$, while in the end, the situation is reversed.
-If a tumor extends over the mid-sagittal line, its contralateral spread rate can be expected to be in between these two extremes.
-We encode this in a *mixing parameter* $\alpha \in [0,1]$ that captures a "degree of asymmetry": \
-   $$
-   b_v^{\text{c},\epsilon=\texttt{True}} = \alpha \cdot b_v^\text{i} + (1 - \alpha) \cdot b_v^{\text{c},\epsilon=\texttt{False}}
-   $$ {#eq-mixing}
-   This means the model now uses three different sets of parameters to describe the spread from the tumor to the LNLs: $b^\text{i}_v$ for the spread to the ipsilateral LNLs, $b_v^{\text{c},\epsilon=\texttt{False}}$ for the spread to the contralateral LNLs as long as the tumor is clearly lateralized, and finally $b_v^{\text{c},\epsilon=\texttt{True}}$ when it crosses the midline.
-Note, however, that these three sets of spread rates only account for $2 \cdot 2^V + 1$ parameters, since they are coupled via the mixing parameter $\alpha$.
-
-Our parameter space has now expanded to
-
-$$
-\boldsymbol{\theta} = \left( \left\{ b_v^\text{i} \right\}, \left\{ b_v^\text{c} \right\}, \alpha, \left\{ t_{vr} \right\}, p_\text{early}, p_\text{adv.} \right) \quad \text{with} \quad \genfrac{}{}{0pt}{2}{v\leq V}{r\in\operatorname{pa}(v)}
-$$ {#eq-bi-param-space}
-
-which is less than double in size, compared to the unilateral model.
-From these parameters, we now build three different transition matrices: The unchanged matrix $\mathbf{A}_\text{i}$ for the ipsilateral side, one $\mathbf{A}_\text{c}^{\epsilon=\texttt{False}}$ for the contralateral side that covers the progression as long as the tumor is lateralized, and $\mathbf{A}_\text{c}^{\epsilon=\texttt{True}}$ in case of a tumor that has crossed the mid-sagittal line.
-
-## Modelling Midline Extension {#sec-midline}
-
-It can be assumed that most tumors that cross the midline at the time of diagnose have started as lateralized tumors and grew over the midline at a later time step.
-Hence, the transition matrix $\mathbf{A}_\text{c}^{\epsilon=\texttt{True}}$ only applies for a subset of the time steps.
-Therefore, we also model the tumor's extension over the mid-sagittal line as a binary random variable.
-A tumor starts lateralized and at every time-step there is a finite probability $p_\epsilon$ that the tumor grows over the midsaggital plane.
-The overall probabilities to find a patient with a clearly lateralized tumor or one that extends over the mid-sagittal line after $t$ time-steps are then respectively given by
-
-$$
-\begin{aligned}
-P(\epsilon = \texttt{False} \mid t) &= (1 - p_\epsilon)^t \\
-P(\epsilon = \texttt{True} \mid t) &= 1 - P(\epsilon = \texttt{False} \mid t)
-\end{aligned}
-$$
-
-Using this, it is straightforward to write down the matrix of state distributions for all time-steps, as in @eq-lambda-matrix covering the contralateral hidden state evolution:
-
-$$
-\boldsymbol{\Lambda}_\text{c}^{\epsilon=\texttt{False}} =
-\left(
-\begin{array}{r}
-\boldsymbol{\pi}^\intercal \cdot \left( \mathbf{A}_\text{c}^{\epsilon=\texttt{False}} \right)^{0\phantom{t_\text{max}}} \\
-(1-p_\epsilon) \cdot \boldsymbol{\pi}^\intercal \cdot \left( \mathbf{A}_\text{c}^{\epsilon=\texttt{False}} \right)^{1\phantom{t_\text{max}}} \\
-\hfill \vdots \hfill \\
-(1-p_\epsilon)^{t_\text{max}} \cdot \boldsymbol{\pi}^\intercal \cdot \left( \mathbf{A}_\text{c}^{\epsilon=\texttt{False}} \right)^{t_\text{max}\phantom{0}} \\
-\end{array}
-\right)
-$$
-
-where we used the transition matrix $\mathbf{A}_\text{c}^{\epsilon=\texttt{False}}$ that depends on the base spread parameters $b_v^{\text{c},\epsilon=\texttt{False}}$.
-
-The case when midline extension is eventually present is more complicated: We already marginalize over the exact time-step when the tumor grows over the mid-sagittal line.
-But whenever that happens, we also need to change the contralateral transition matrix to use the increased spread rates $b_v^{\text{c}, \epsilon=\texttt{True}}$ from the tumor to the contralateral LNLs, given by the linear mixing in @eq-mixing.
-We can achieve the correct marginalization by iteratively building the joint distribution $P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid t \right)$.
-We start from $t=0$, where we know that all LNLs are healthy (the contralateral neck is in the state $\mathbf{X}_\text{c}=\boldsymbol{\xi}_0$) and the tumor is lateralized ($\epsilon=\texttt{False}$):
-
-$$
-P \left( \mathbf{X}^\text{c} = \boldsymbol{\xi}_0, \epsilon=\texttt{False} \mid t=0 \right) = 1
-$$
-
-whereas all other states have zero probability. Then we consider an arbitrary later time-step $t=\tau+1$. There are two possible scenarios we need to marginalize over:
-
-1.
-The tumor was still lateralized at $t=\tau$ and just grew over the midline.
-Under this scenario, the probability for having a midline extension at $t=\tau+1$ is given by $p_\epsilon$.
-We use this to weight the contralateral state distribution that was so far evolved without increased contralateral spread.
-2.
-The mid-sagittal line was already crossed by the tumor.
-In that case, the probability is 1 to remain in that lateralization state.
-Thus, to account for this case we simply add the distribution $P\left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid \tau \right)$ from one time-step earlier, arriving at a recursion relation:
-
-$$
-P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid \tau + 1 \right) = \big[ p_\epsilon P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{False} \mid \tau \right) + P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid \tau \right) \big]^\top \cdot \mathbf{A}_\text{c}^{\epsilon=\texttt{True}}
-$$
-
-We can collect the iteratively computed distributions for the midline extension case to define the matrix over the states given all time-steps, in analogy to @eq-lambda-matrix:
-
-$$
-\boldsymbol{\Lambda}_\text{c}^{\epsilon=\texttt{True}} = \begin{pmatrix}
-P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid 0 \right) \\
-P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid 1 \right) \\
-\vdots \\
-P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid t_\text{max} \right) \\
-\end{pmatrix}
-$$
-
-Using this, we can again write the joint of ipsi- and contralateral involvement - now also for the case of mid-sagittal extension - algebraically as before in @eq-bilateral-marginal-algebra:
-
-$$
-P \left( \mathbf{X}^\text{i} = \boldsymbol{\xi}_n, \mathbf{X}^\text{c} = \boldsymbol{\xi}_m, \epsilon \right) = \left[ \boldsymbol{\Lambda}^\intercal_\text{i} \cdot \operatorname{diag} P(\mathbf{t}) \cdot \boldsymbol{\Lambda}_\text{c}^\epsilon \right]_{n,m}
-$$
-
-We can use this term to compute the likelihood of all patients with and without midline extension separately.
-And if for some patients the information of tumor lateralization is not available, we can simply sum the above term once for $\epsilon=\texttt{False}$ and once for $\epsilon=\texttt{True}$ to marginalize over the unknown variable.
-
-The final parameter space of our extended model has now reached this size:
-
-$$
-\boldsymbol{\theta} = \left( \left\{ b_v^\text{i} \right\}, \left\{ b_v^\text{c} \right\}, \alpha, \left\{ t_{vr} \right\}, p_\text{early}, p_\text{adv.}, p_\epsilon \right) \quad \text{with} \quad \genfrac{}{}{0pt}{2}{v\leq V}{r\in\operatorname{pa}(v)}
-$$ {#eq-ext-param-space}
-
-## Model Prediction in the Bayesian Context
-
-Our stated goal is to compute the risk for a patient's true ipsi- and contralateral nodal involvement states $\mathbf{X}^\text{i}$ and $\mathbf{X}^\text{c}$, *given* their individual diagnosis $d = \left( \boldsymbol{\zeta}^\text{i}_k, \boldsymbol{\zeta}^\text{c}_\ell, \epsilon, \text{T}x \right)$.
-Here, this diagnosis consists of the observed ipsi- and contralateral nodal involvements, the patient's midline extension $\epsilon$, and their tumor's T-category $\text{T}x$.
-Using Bayes' law, we can write this risk as:
-
-$$
-P \big( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \mid d, \boldsymbol{\hat{\theta}} \big)
-= \frac{P \left( \boldsymbol{\zeta}^\text{i}_k \mid \mathbf{X}^\text{i} \right) P \left( \boldsymbol{\zeta}^\text{c}_\ell \mid \mathbf{X}^\text{c} \right) P \big( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \mid \boldsymbol{\hat{\theta}}, \text{T}x \big)}
-{\sum_{i=0}^{2^V} \sum_{j=0}^{2^V} P \left( \boldsymbol{\zeta}^\text{i}_k \mid \mathbf{X}^\text{i}=\boldsymbol{\xi}^\text{i}_i \right) P \big( \boldsymbol{\zeta}^\text{c}_\ell \mid \mathbf{X}^\text{c}=\boldsymbol{\xi}^\text{c}_j \big) P \big( \mathbf{X}^\text{i}=\boldsymbol{\xi}^\text{i}_i, \mathbf{X}^\text{c}=\boldsymbol{\xi}^\text{c}_j, \epsilon \mid \boldsymbol{\hat{\theta}}, \text{T}x \big)}
-$$ {#eq-uni-bayes-law}
-
-The terms $P \left( \boldsymbol{\zeta}^\text{i}_k \mid \mathbf{X}^\text{i} \right)$ and $P \left( \boldsymbol{\zeta}^\text{c}_\ell \mid \mathbf{X}^\text{c} \right)$ are defined solely by sensitivity and specificity of the diagnostic modality.
-Terms like these already appeared in the definition of the observation matrx in @eq-obs-matrix.
-The *prior* $P \big( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \mid \boldsymbol{\hat{\theta}}, \text{T}x \big)$ in the above equation is the crucial term that is supplied by a trained model and its parameters $\boldsymbol{\hat{\theta}}$.
-
-It is possible to compute this *posterior* probability of true involvement not only for one fully defined state $(\mathbf{X}^\text{i}, \mathbf{X}^\text{c})$, but also for e.g.
-individual LNLs: For example, the risk for involvement in the contralateral level IV would be a marginalization over all ipsilateral states $\boldsymbol{\xi}^\text{i}_i$ and all contralateral states $\boldsymbol{\xi}^\text{c}_j$ where $\xi^\text{c}_{j4}=1$.
-Formally:
-
-$$
-P \big( \text{cIV} \mid \mathbf{Z}^\text{i}=\boldsymbol{\zeta}^\text{i}_k, \mathbf{Z}^\text{c}=\boldsymbol{\zeta}^\text{c}_\ell, \boldsymbol{\hat{\theta}}, \text{T}x \big) = \sum_k \sum_{\ell \, : \, \xi_{\ell 4}=1} P \big( \mathbf{X}^\text{i} = \boldsymbol{\xi}^\text{i}_k, \mathbf{X}^\text{c} = \boldsymbol{\xi}^\text{c}_\ell \mid \boldsymbol{\zeta}^\text{i}_k, \boldsymbol{\zeta}^\text{c}_\ell, \epsilon, \boldsymbol{\hat{\theta}}, \text{T}x \big)
-$$ {#eq-marg-over-posterior}
-
-# Computational methods {#sec-methods}
-
-<!-- Maybe I should break this apart into separate chapters on data, sampling, etc. -->
-
-In this section, we detail how the experiments were performed.
-Every figure, table, and result is fully reproducible via the GitHub repository [`rmnldwg/bilateral-paper`].
-It also contains the raw manuscript and instructions on how to recreate all figures, tables, and the final document.
-
-[`rmnldwg/bilateral-paper`]: https://github.com/rmnldwg/bilateral-paper
-
-## Involvement Data Consensus
-
-It is possible to provide our model with multiple different diagnostic modalities, each being characterized by different pairs of sensitivity and specificity.
-However, we instead chose to combine them into a single "consensus" diagnosis before parameter inference, as introduced in @sec-data-consensus.
-We opted for this because the literature values of sensitivity and specificity [@de_bondt_detection_2007;@kyzas_18f-fluorodeoxyglucose_2008] of imaging modalities like MRI and CT do not plausibly match some of our observations: In the USZ cohort, 78% of OPSCC patients where diagnosed with ipsilateral LNL II involvement via diagnostic imaging.
-This is virtually impossible with sensitivities around 80% and specificities lower than 100%.
-
-## MCMC Sampling {#sec-sampling}
-
-For parameter inference, we used the Python package [`emcee`] [@foreman-mackey_emcee_2013].
-It implements efficient MCMC sampling algorithms that employ multiple parallel samplers for affine invariance and better performance on multi-core CPUs.
-The sample proposal algirothms used by us are based on differential evolution moves [@ter_braak_differential_2008;@nelson_run_2013].
-The [`emcee`] library was provided with the likelihood implemented by our [`lymph-model`] Python package.
-
-For each dimension in the parameter space of the model, we initialized {{< var sampling.walkers_per_dim >}} of these parallel samplers, called "walkers", with random values in the unit cube.
-Every time all of these walkers advanced {{< var sampling.check_interval >}} steps, the autocorrelation time of the chains was estimated.
-For short chains, this estimate is not trustworthy, but stabilizes for longer chains.
-We therefore considered a sampling to be converged when two criteria were met:
-
-1. The change in the autocorrelation time was less then {{< var sampling.rel_thresh >}}.
-2. The estimate of the autocorrelation dropped below $n$ / {{< var sampling.trust_fac >}} where $n$ is the length of the chain up to that point.
-
-All samples up to this convergence - called the *burn-in phase* - were discarded.
-We only kept another {{< var sampling.nsteps >}} samples after that, which were spaced {{< var sampling.thin >}} steps apart.
-
-[`emcee`]: https://emcee.readthedocs.io/en/stable/
-[`lymph-model`]: https://lymph-model.readthedocs.io/en/stable/
-
-First, in @fig-model-burnin-history, we verify the sampling converged successfully by inspecting two monitoring quantities: The autocorrelation time of the MCMC chain and the acceptance fractions of the parallel walkers.
-
-![Monitoring quantities during the burn-in phase of the parameter sampling.
-Left: The autocorrelation time of the sampling chain estimated at different sampling steps.
-We consider the chain converged when the estimate of the autocorrelation time is stable and drops below the trust threshold of $n/50$ where $n$ is the number of steps.
-Right: Fraction of accepted MCMC proposals averaged over all parallel walkers.
-Values around 30% indicate good mixing of the walkers.](figures/fig_model_burnin_history.svg){#fig-model-burnin-history}
-
-## Computing the Observed and Predicted Prevalence of Involvement Patterns {#sec-prevalence}
-
-We want to assess the model's capability to approximate the distribution of lymphatic involvement patterns seen in the data.
-To that end, we compare the prevalence of some invovlement patterns under selected scenarios with the model's prediction for how often these involvements it expects to see, given these scenarios.
-
-In this context, a "scenario" includes the patient's T-category $\text{T}x$ and whether the patient's tumor extended over the mid-sagittal line, i.e.
-$\epsilon=\texttt{True}$ or $\epsilon=\texttt{False}$.
-
-An involvement pattern specifies for each ipsi- and contralateral LNL whether it is "healthy", "involved", or "masked".
-If it is "masked", we essentially state that we are not interested in the involvement of that LNL and the prevalence will be marginalized over this LNL's involvement.
-
-For example, we may be interested in the prevalence of contralateral LNL II involvement (i.e., contra LNL II "involved" and all other LNLs "masked") under the scenario of early T-category (T0-T2) and no midline extension ($\epsilon=\texttt{False}$).
-To compute this prevalence in the data, we select all patients of this scenario (in our data, this amounts to {{< var data.early_nomidext_cII_total >}} patients).
-Of those, {{< var data.early_nomidext_cII_match >}} were found to harbor metastases in their contralateral LNL II.
-Therefore, the prevalence is {{< var data.early_nomidext_cII_percent >}}%.
-
-When displaying this data prevalence, we often choose to draw a *beta posterior* over the "true" prevalence, hinting at the fact that our data merely represents a limited sample.
-The beta posterior follows from a uniform beta distribution as prior and a binomial likelihood for the number of patients with the involvement of interest, given the parameer for the "true" prevalence.
-The resulting distribution has its maximum at the observed prevalence, but in addition gives a visual intuition for the variance of of the observed quantity.
-I.e., when we observe 3 out of 10 events, the beta posterior is much wider than if we observe 300 out of 1000 for he same prevalence.
-It also allows us to check not only if the model is accurate, but also whether it reflects the uncertainty contained in the data.
-
-Predicting the prevalence using our model amounts to computing the following probability:
-
-$$
-P \left( \text{II}^\text{c} \mid \epsilon=\texttt{False}, \text{T}x=\text{early} \right) = \frac{P \left( \text{II}^\text{c}, \epsilon=\texttt{False} \mid \text{T}x=\text{early} \right)}{P \left( \epsilon=\texttt{False} \mid \text{T}x=\text{early} \right)}
-$$
-
-In the enumator, we marginalize over all ipsi- and contralateral LNLs' involvements, except for LNL II contralaterally.
-This is similar to the marginalization in @eq-marg-over-posterior, although we are summing over different quantities.
-In the denominator, we can simply insert the joint distribution over midline extension and diagnose time $P \left( \epsilon, t \right)$ marginalized over $t$ using the early T-category's time-prior.
-
-Since we compare it to the data, which does not report true but only observed involvement -- although pathologically investigated LNLs may be as close as possible to the ground truth -- we do not consider posteriors of the form $P \left( \mathbf{X} \mid \mathbf{Z} \right)$ here.
-Instead, we compute probabilities of observed involvement $P \left( \mathbf{Z} \right)$, as in the likelihood @eq-single-patient-llh.
-
-When plotted, we usually display histograms over the model's predictions.
-Each of their values was computed from a different parameter set drawn during MCMC sampling, effectively giving us a distribution over the prevalences.
-Ideally, the histograms approximate the location and width of the Beta posteriors when attempting to describe the data they were trained on.
-
-Note that we decided to omit the y-axis ticks and labels in these figures over prevalences and risks.
-The y-axis in these plots measures the probability density and its numerical values are not intuitively interpretable.
-Instead, we occasionally use the freed space to label e.g.
-rows of subplots.
-
-# Results: Model evaluation {#sec-results}
-
-In @tbl-midline-params, we tabulate the mean and standard deviation of the sampled parameters for the full midline model.
-Considering the parameters of the bilateral model that describe the ipsilateral spread ($b^i_v$, $t_{vr}$ and $p_\text{adv.}$), we note that the bilateral model mostly reproduces the parameter values reported in the earlier publication [@ludwig_modelling_2023] for the unilateral model.
-The small differences may be mostly due to the differences in training dataset.
-Therefore, the analysis of the models capability to adequately describe ipsilateral lymph node involvement is not repeated here and we focus instead on the analysis of contralateral involvement.
-We can further note that the contralateral spread parameters $b^c_v$ are small compared to the ipsilateral parameters $b^i_v$, reflecting the low prevalence of contralateral lymph node involvement for lateralized tumors.
-
-```{python}
-#| echo: false
-#| label: tbl-midline-params
-#| tbl-cap: Mean sampled parameter estimates of the midline model and the respective standard deviation.
-def map_param_names(name: str) -> str:
-  """Make parameter names more readable."""
-  try:
-    return {
-      "midext_prob": "Mid. ext. probability",
-      "mixing": "Mixing ⍺",
-      "late_p": "late T-cat. binom. prob.",
-    }[name]
-  except KeyError:
-    return re.sub(
-      r"(ipsi|contra)_", r"\g<1>: ", name
-    ).replace(
-      "to", " ➜ "
-    ).replace(
-      "_spread", ""
-    )
-
-model = shared.get_model(which="full", load_samples=True)
-samples = shared.get_samples(which="full")
-
-names = [map_param_names(p) for p in model.get_params()]
-means, stds = samples.mean(axis=0), samples.std(axis=0)
-
-early_midext_prob = model.state_dist(t_stage="early")[1].sum()
-late_midext_prob = model.state_dist(t_stage="late")[1].sum()
-
-params_table = pd.DataFrame({"Parameter": names, "Mean": means, "Std. Dev.": stds})
-(
-  params_table.style
-  .format("{:.2%}", subset=["Mean"])
-  .format("± {:.2%}", subset=["Std. Dev."])
-  .apply(right_align, subset=["Mean", "Std. Dev."])
-  .hide()
-)
-```
-
-## Illustration of the model
-
-In this subsection we illustrate parts of the mathematical framework defined in the previous section.
-In @fig-model-midext-evo (top panel), we visualize the prior distribution over diagnose times $P(t)$.
-By parameter choice, early T-category tumors are diagnosed on average after {{< var model.early_expected_time >}} time steps.
-Advanced T-category tumors are diagnosed on average after {{< var model.late_expected_time >}} time steps, because the parameter $p_\text{adv.}$ was learned to be {{< var model.params_percent.late_p >}}% for advanced T-category tumors.
-The probability per time step for the tumor to grow over the midline ($p_\epsilon$) was on average {{< var model.params_percent.midext_prob >}}%.
-Using this, one can compute the probability of midline extension $P(\epsilon \mid t)$ at a given time step $t$ (red line in @fig-model-midext-evo).
-The bottom panel in this figure shows the joint probability $P(\epsilon, t)$ to be diagnosed at time step $t$ with a given state of midline extension and T-category.
-
-![The top panel shows the prior probability to get diagnosed at time-step $t$ for early and late T-category tumors as bars.
-Also in the top panel, we plot the conditional probability of the tumor's midline extension ($\epsilon=\texttt{True}$), given the time-step $t$ as a line plot.
-In the bottom panel, we show the joint probability of getting diagnosed in time-step $t$ *and* having a tumor that crosses the midline.](figures/fig_model_midext_evo.svg){#fig-model-midext-evo width="50%"}
-
-The model introduced in the previous section defines a model of the joint probability distribution over midline extension and ipsi- and contralateral lymph node involvement, $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)$.
-How this quantity is computed is illustrated in @fig-model-state-dist, visually representing @eq-bilateral-marginal-algebra.
-To keep it manageable to interpret, this example only considers the LNLs II, III, and IV ipsi- and contralaterally, resulting in $2^3 = 16$ distinct states per side and thus $2 \times 8 \times 8 = 128$ total states.
-LNLs I, V, and VII and the associated spread parameters have been removed from the model while the remaining parameters are set to the mean values in table xxx.
-@fig-model-state-dist shows the joint distribution as two separate heatmaps for the two midline extension states for early/advanced???
-T-category tumors.
-The most likely state is a lateralized tumor with involvement of ipsilateral level II and no contralateral involvement, which has a probability of approximately 25%.
-The second most probable state is a lateralized tumor with ipsilateral involvement of levels II and III and no contralateral involvement.
-The most likely state with contralateral involvement are tumors with midline extension with involvement of contralateral level II and ipsilateral level II and III.
-
-![Visual representation of @eq-bilateral-marginal-algebra for the case of midline extention.
-The left and right matrix represent the evolution of the possible hidden states for the ipsi- and contralateral neck respectively.
-In particular, the right matrix describes the evolution of the contralateral hidden states *and* midline extension.
-In the center, the time-prior is plotted as a diagonal matrix.
-The result of this matrix computation is the joint distribution $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\texttt{True} \right)$](figures/fig_model_state_dist.svg){#fig-model-state-dist}
-
-## Prevalence predictions for contralateral involvement
-
-The bilateral model was designed to fulfil the requirements laid out in @sec-requirements.
-Now, we investigate to what extent the model can quantitatively describe the patterns of lymph node involvement observed in the dataset.
-To that end, we compare the model's predictions for contralateral involvement against observations in the data.
-This is done given scenarios that differ in T-category and/or midline extension and/or ipsilateral involvement.
-
-### Dependence of Contralateral Involvement on T-Category and Midline Extension
-
-In @fig-model-prevalences-overall, we plot the prevalence of contralateral involvement of the LNLs II, III, and IV for the four scenarios made up of the possible combinations of early and late T-category, as well as lateralized and midline extending tumors.
-
-![Comparison of predicted (histograms) vs observed (beta posteriors) prevalences.
-Shown for the contralateral LNLs II (blue), III (orange), and IV (green).
-The top row shows scenarios with early T-category tumors, the bottom row for late T-category ones.
-The left column depicts scenarios where the primary tumor is clearly lateralized, the right column scenarios of tumors extending over the mid-sagittal line.
-This figure illustrates the model's ability to describe the prevalence of involvement for different combinations of the risk factors T-category and midline extension.](figures/fig_model_prevalences_overall.svg){#fig-model-prevalences-overall}
-
-@Fig-model-prevalences-overall shows nicely that the model is capable of accurately accounting for the most important risk factors, i.e.
-T-category and midline extension.
-As observed in the data, the model predicts that the prevalence of contralateral LNL II involvement increases from below 8% <!--  TODO: make this dynamic --> for early T-category lateralized tumors to almost 40% <!-- TODO: make this dynamic --> when the tumor is of advanced T-category and crosses the mid-sagittal line.
-Similarly, the prevalence of contralateral LNL III involvement increases from around 2% <!--  TODO: make this dynamic --> for early T-category lateralized tumors to almost 15% <!-- TODO: make this dynamic --> when the tumor is of advanced T-category and crosses the mid-sagittal line.
-
-<!--  However, for early T-category scenarios with midline extension, the model does seem to overestimate contralateral LNL II and III invovlement.
-This likely stems from the the small sample size of this relatively rare scenario as hinted at by the wide beta posteriors.
--->
-
-### Correlation between Ipsi- and Contralateral Involvement
-
-![Comparison of the computed and observed prevalences for scenarios that illustrate the model's capability of accounting for the correlation between ipsi- and contralateral involvement.
-We show two scenarios: One where the ipsilateral neck shows no involvement (at least LNLs I to V are healthy, LNL VII was ignored because data on it is missing for some patients) in blue and one where at least LNL III was involved in orange.
-These two scenarios are plotted for all combinations of T-category (early in top row, advanced in bottom row) and tumor lateralization (lateralized in left column, extending over mid-sagittal line in right column).](figures/fig_model_prevalences_with_ipsi.svg){#fig-model-prevalences-with-ipsi}
-
-In @fig-model-prevalences-with-ipsi we display the model's ability to capture the correlation between ipsi- and contralateral involvement.
-It shows that the prevalence of metastases in the two sides of the neck is correlated via the time of diagnosis, despite the model not having any direct connections between the two side.
-However, there are some discrepancies in the model's prediction: It cannot quite capture the correlation of ipsi- and contralateral involvement that is seen in the data.
-This is notable for example in the bottom right subplot where the prevalence is overestimated for the scenario of an (almost) healthy ipsilateral neck and underestimated when at least LNL III shows metastases.
-One reason for this may be the width of the time-prior: If it was wider, the posterior over the diagnosis time given the ipsilateral involvement could more flexibly shift to reflect a more severe contralateral involvement.
-
-<!-- NOTE: This last sentence is purely speculative! We need to discuss this. -->
-
-### Influence of Upstream Involvement on Contralateral Metastasis
-
-![The influence of the upstream LNL II's involvement on the prevalence of contralateral level III for the four combinations of tumor lateralization (lateralized or extending over midline) and T-ctageory (early or advanced).
-Our model predictions (histograms) are plotted against the observations in the data (beta posteriors).](figures/fig_model_prevalences_upstream.svg){#fig-model-prevalences-upstream}
-
-@Fig-model-prevalences-upstream illustrates how rarely the contralateral LNL III harbors metastases when its upstream level II is healthy.
-Also, the model can capture this correlation rather well, with the exception of early T-category tumors that extend over the mid-sagittal line.
-As in @fig-model-prevalences-overall and @fig-model-prevalences-with-ipsi, these cases are relatively rare resulting in a broad distribution over the true prevalence.
-
-# Results: Prediction of Risk for Occult Disease
-
-For a clinical application of the model we are interested in the risk of occult metastases rather than the observed patterns of lymph node involvement.
-We want to estimate the risk for occult disease in clinically negative LNLs, given the patient's individual diagnosis.
-In terms of our model, the diagnosis consists of the T-category, the lateralization of the tumor (if it extend over the mid-sagittal plane), and which LNLs are clinically involved based on imaging and possibly fine needle aspiration (FNA).
-For the results shown in this section, we assume that involvement of a LNL is clinically diagnosed through imaging with a sensitivity of 81% and a specificity of 76% [@de_bondt_detection_2007].
-For FNA we assume a specificity of 98% and a sensitivity of 80%.
-This amounts to the assumption that lymph node involvement confirmed by FNA is almost certainly true involvement, i.e.
-there are almost no false positive diagnoses.
-
-![Histograms over the predicted risk of occult contralateral LNL II (left), III (middle), and IV (right) invovlement, shown for some combinations of T-category, tumor lateralization, and clinical LNL diagnoses.
-All LNLs not explicitly mentioned in the legend, including the LNL for which the risk of occult disease was computed, where assumed to be clinically negative, based on CT imaging (specificity 76%, sensitivity 81%).](figures/fig_model_risks.svg){#fig-model-risks}
-
-<!-- @Fig-model-risks shows this predicted risk of occult disease in the contralateral LNLs II, III, and IV for interesting combinations of these three risk factors.
-This risk is computed *given* a CT diagnosis assessing the per-LNL clinical involvement for which we assumed a sensitivity of 81% and a specificity of 76%.
-For the involvement in the LNls III and IV contralaterally, we also cosidered how the risk would be affected if the upstream LNL was confirmed to be metastatic using fine needle aspiration (FNA) with a specificity of 98% and a sensitivity of 80% [@de_bondt_detection_2007].
-In the figure legends, this is noted e.g.
-for the orange historgam over the risk of contralateral LNL IV involvement as `c:III FNA+`, meaning that the contralateral LNL III was pathologically confirmed to harbor metastases via FNA.-->
-
-## Contralateral LNL II
-
-@Fig-model-risks (left panel) shows the predicted risk of occult disease in contralateral LNL II.
-The most important variable impacting the prediction for contralateral level II involvement in our model is the tumor's lateralization.
-A patient with a clearly lateralized early T-category tumor and a clinically N0 neck is predicted to have a 1-2% <!-- TODO: make this dynamic --> risk for occult disease in contralateral LNL II.
-For a patient with a clearly lateralized early T-category tumor but *with* mid-sagittal extension, the risk increases to almost 7%.
-<!-- TODO: make this dynamic -->
-
-Also advanced T-category increases the risk of occult disease but plays a lesser role: Considering the scenario of a tumor that crosses the midline and an ipsilateral neck where at least LNL III is clinically involved, the risk for occult contralateral LNL II disease is around 8.5%.
-<!-- TODO: make this dynamic --> For the same scenario and an advanced T-category tumor, the risk increases to 11%.
-<!-- TODO: make this dynamic -->
-
-Lastly, the predicted risk is also correlated via the time-steps to the degree of ipsilateral involvement.
-Changing the aforementioned scenario (advanced T-category, midline extension, ipsilateral LNL III clinically involved) to one where the patient presents with a clinically N0 neck, the risk for occult disease in the contralateral LNL II falls from 11% <!-- TODO: make this dynamic --> to 9.5%.
-<!-- TODO: make this dynamic -->
-
-Taken together, T-category and ipsilateral involvement may still considerably impact the risk prediction for contralateral involvement: In @fig-model-risks the scenarios underlying the orange (6.5% <!-- TODO: make this dynamic -->) and the red (11% <!-- TODO: make this dynamic -->) histograms differ in T-category (early vs advanced) and clinical diagnosis (N0 vs ipsilateral at least LNL III involved).
-
-## Contralalteral LNL III
-
-As shown in the center subplot of @fig-model-risks, the risk for occult disease in the contralateral LNL III may only cross a 5% threshold if the upstream level II is clinically involved, too.
-If the clinical diagnosis of the contralateral upstream level II is pathologically confirmed by FNA the risk increases by another percentage point<!-- TODO: make dynamic -->.
-When the tumor additionally crosses the mid-sagittal line, the risk may reach around 7%<!-- TODO: make dynamic -->.
-
-## Contralateral LNL IV
-
-Our model's prediction for the contralateral LNL IV show that even in the most extreme case, with an advanced T-category tumor extending over the mid-sagittal line as well as metastases in every ipsilateral LNL and every contralateral upstream level, the risk stays just below 3%<!-- TODO: make dynamic -->.
-We only observe a risk of 5-6% when the involvement of the upstream level III is pathologically confirmed after taking a biopsy using FNA.
-This is because the model's prior probability for contralateral LNL IV involvement is so low that even given a clinical diagnosis in that level, a false positive in LNL III is a more likely explanation.
-This possibility, however, drops virtually to zero when tumor cells are pathologically confirmed in level III.
-
-# Discussion {#sec-discussion}
-
-## Summary
-
-In this work we present a formalism to model the ipsi- and contralateral lymphatic involvement of oropharyngeal SCC patients.
-An ipsilateral model has been previously developed and published [@ludwig_hidden_2021;@ludwig_modelling_2023].
-Based on this, we introduce an extension that leaves the ipsilateral model untouched, but extends it to the contralateral side in an intuitive and comprehensible manner.
-
-The model's performance w.r.t.
-its ability to describe the data on contralateral nodal involvement was evaluated based on a dataset of 833<!-- TODO: make this dynamic --> patients from four institutions.
-To the best of our knowledge, this is the first time lymphatic tumor progression has been modelled in such detail and to such an extent.
-While there was work on regional tumor spread, these models were conceptually different, more limited in the LNLs for which they make predictions, and not trained with real patient data [@benson_markov_2006;@jung_development_2017].
-
-The model takes the clincal diagnosis of the lymph node levels, the primary tumor's T-category, as well as lateralization into account when predicting the personalized risk for occult disease in any LNL of interest.
-Owing to the relatively few parameters, the model is highly interpretable and every parameter can be intuitively explained, while still offering good accuracy in its predictions.
-
-## Implications for Contralateral Elective Nodal Treatment
-
-The predictions of this model are being used to inform a clinical trial on volume deescalation at the University Hospital Zurich [citation needed<!-- TODO: add -->].
-When accepting a 5% risk of occult disease in any given LNL, the model suggests the following contralateral elective irradiation (assuming the respective LNL appears clinically healthy):
-
-- LNL II when the tumor extends over the mid-sagittal line
-- LNL III when LNL II is clinically involved, may need confirmation via FNA
-- LNL IV only when upstream level III pathologically confirmed
-
-# Acknowledgement {.appendix}
-
-This work was supported by:
-
-- the Clinical Research Priority Program "Artificial Intelligence in Oncological Imaging" of the University of Zurich
-- the Swiss Cancer Research Foundation under grant number KFS 5645-08-2022
-
-# Contralateral Prevalence of Involvement {.appendix}
-
-```{python}
-#| echo: false
-#| label: tbl-data-strat
-#| tbl-cap: Contralateral involvement depending on whether the primary tumor extends over the mid-sagittal line, the T-category, and whether the ipsilateral LNL III was involved or healthy.
-lnl_cols = get_lnl_cols("contra", lnls=["I", "II", "III", "IV"])
-num_ipsi_inv = raw[get_lnl_cols("ipsi")].sum(axis="columns")
-
-contra_inv = raw[lnl_cols].copy()
-contra_inv.columns = contra_inv.columns.droplevel([0,1])
-contra_inv["t_stage"] = raw[COL.t_stage].apply(lambda x: "early" if x <= 2 else "advanced")
-contra_inv["ipsi"] = num_ipsi_inv.map(lambda x: str(x) if x <= 1 else "≥ 2")
-contra_inv["midext"] = raw[COL.midext]
-
-grouped = contra_inv.groupby(by=["t_stage", "ipsi", "midext"], dropna=False)
-num_involved = grouped.sum()
-total = grouped.count()
-percent_involved = num_involved / total
-idx = total.index.rename(["T-cat.", "ipsi", "Mid. ext."])
-
-total.index = idx
-num_involved.index = idx
-percent_involved.index = idx
-
-involved = num_involved.join(
-  100 * percent_involved,
-  rsuffix=" (%)",
-)
-involved.columns = pd.MultiIndex.from_product(
-  [["I", "II", "III", "IV"],
-   ["n", "%"]],
-  names=["LNL", ""]
-)
-involved["total", "n"] = total["I"]
-(
-  involved
-  .reset_index()
-  .sort_values(
-    by=["T-cat.", "ipsi", "Mid. ext."],
-    ascending=[False, True, True],
-  )
-  .style
-  .format(precision=2)
-  .apply(right_align, subset=involved.columns[0:])
-  .apply(highlight_t_stage, subset=["T-cat."])
-  .apply(highlight_bool, subset=["Mid. ext."])
-  .apply(highlight, subset=["ipsi"], mapping={
-    "0": COLORS["green"],
-    "1": COLORS["orange"],
-    "≥ 2": COLORS["red"],
-  })
-  .hide()
-)
-```
-
-# Prevalence of Midline Extension {.appendix}
-
-<!-- I moved this here, because we think this is probably irrelevant or distracting w.r.t. the main findings of the paper. But I do think it is somewhat interesting. -->
-
-![Comparing the predicted (histograms) and observed (lines depicting beta posteriors) prevalence of midline extension for early (blue) and late (orange) T-category.
-While the prevalence is predicted correctly when marginalizing over T-category, the model cannot capture the degree of separation observed in the data.
-Since the tumor's midline extension is virtually always part of the diagnosis and hence *given* when predicting a patient's risk, we do not consider this discrepancy a major issue.](figures/fig_model_prevalences_midext.svg){#fig-model-prevalences-midext width="50%"}
-
-In @fig-model-prevalences-midext, we plot the prevalence of midline extension in the data versus our model's prediction.
-It is obvious the model cannot match the large spread between early and advanced T-category seen in the data.
-This is because to achieve that, it would need to increase the advanced T-category patient's prior distribution over diagnosis times and at the same time reduce the probability of the tumor to cross the midline during a time-step.
-But since the time-priors parameter is also coupled with the spread probabilities among the LNLs, the model does not have that freedom.
-
-<!-- Should the paragraphs below be part of the discussion? -->
-
-However, we do not consider this discrepancy a major limitation of the model: We will not realistically be interested in the probability of midline extension, as it is always possible to assess it with high certainty.
-That is also the reason why we initially modelled the midline extension *not* as a random variable, but as a global risk factor that would have been turned on or off from the onset of a patient's disease evolution.
-This, however, lead to overly high risks for contralateral involvement in advanced T-category patients with midline extension, because then the model assumes an increased spread to the contralateral side from the onset of the disease.
-Which is probably not true in a majority of those cases.
-Thus, treating it as a random variable that only becomes true during a patient's disease evolution resulted in a better description of the data.
-
-Formally, the wrong prediction of midline extension prevalence makes little difference, since it is always given: Instead of $P\left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \mid \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \right)$, we typically compute $P\left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \mid \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c}, \epsilon \right)$, which does not suffer from the wrong probability of midline extension, as the distribution over hidden states is renormalized:
-
-$$
-P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \mid \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c}, \epsilon \right) = \frac{P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \mid \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right) P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)}{P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c}, \epsilon \right)}
-$$
-
-Note that a distribution over $\epsilon$ appears both in the enumerator and the denominator, which largely cancel each other, leaving only the midline extension's effect on the distribution over hidden states in the prediction.
-
-Also, the discrepancy in midline extension prevalence between early and advanced T-category is particularly pronounced in oropharyngeal SCC patients.
-For example, in oral cavity SCC, the midline extension only increases from 15.4% (20 out of 130) to 33.3% (13 out of 39).
+Therefore we introduce a mixture model which combines all subsites in a singular model exploiting anatomical proximities.

### 2024-09-18 scripts/__init__.py

@@ -1,1 +1,0 @@
-"""Scripts and other reused code for this paper."""

### 2024-09-18 scripts/compile_variables.py

@@ -1,321 +1,0 @@
-"""Script to compile variables from data and models and stores them in a YAML file."""
-
-from pathlib import Path
-from typing import Any
-
-import h5py
-import lydata  # noqa: F401
-import numpy as np
-import pandas as pd
-import paths
-import shared
-import yaml
-from lydata.accessor import Q, QueryPortion
-from pandas.api.typing import DataFrameGroupBy
-from pydantic_settings import BaseSettings
-
-from lyscripts.scenario import Scenario
-
-
-class CmdSettings(BaseSettings, cli_parse_args=True):
-    """Settings for the command-line arguments."""
-
-    risk_hdf5: Path = paths.model_dir / "full" / "risks.hdf5"
-    risk_scenarios: Path = paths.scenario_dir / "risks.yaml"
-    prevalence_hdf5: Path = paths.model_dir / "full" / "prevalences_overall.hdf5"
-    prevalence_scenarios: Path = paths.scenario_dir / "overall.yaml"
-    output: Path = Path("_variables.dynamic.yaml")
-    precision: int = 2
-
-
-def group_contra_inv(data: pd.DataFrame) -> DataFrameGroupBy:
-    """Group the data by T-stage, num of ipsi involved, and midline extension."""
-    lnl_cols = shared.get_lnl_cols("contra", lnls=["I", "II", "III", "IV"])
-    num_ipsi_inv = data[shared.get_lnl_cols("ipsi")].sum(axis="columns")
-    contra_inv = data[lnl_cols].copy()
-    contra_inv.columns = contra_inv.columns.droplevel([0,1])
-
-    contra_inv["t_stage"] = data[shared.COL.t_stage]
-    contra_inv["ipsi"] = num_ipsi_inv.map(lambda x: str(x) if x <= 1 else "≥ 2")
-    contra_inv["midext"] = data[shared.COL.midext]
-
-    return contra_inv.groupby(by=["t_stage", "ipsi", "midext"])
-
-
-def cast_numpy_to_native(variables: dict[str, Any]) -> dict[str, Any]:
-    """Convert numpy dtypes to native Python types."""
-    result = {}
-
-    for key, value in variables.items():
-        try:
-            result[key] = value.item()
-        except AttributeError:
-            result[key] = value
-
-    return result
-
-
-def variables_from_portion(
-    key: str,
-    portion: QueryPortion,
-    precision: int = 2,
-) -> dict[str, Any]:
[…]

### 2024-09-18 scripts/data_strat.py

@@ -1,186 +1,0 @@
-"""Contra involvement by T-category, number of metastatic ipsi LNLs, and midext."""
-
-import matplotlib.pyplot as plt
-import numpy as np
-import pandas as pd
-import paths
-import shared
-from matplotlib import gridspec
-from matplotlib.axes import Axes
-
-from lyscripts import utils
-from lyscripts.plot.utils import COLORS
-
-
-def group_and_plot(
-    df: pd.DataFrame,
-    column: str,
-    axes: Axes,
-    colors: list[str],
-) -> None:
-    """Group `df` by `column` and plot the result."""
-    grouped = df.groupby(by=column)
-    counts = grouped.count().T
-    aggregated = 100 * grouped.mean().T
-    edges = np.linspace(1.0, 0.0, len(grouped) + 1)
-    positions = edges[1:] + ((edges[0] - edges[1]) / 2)
-
-    for i, ((label, series), (_, count)) in enumerate(
-        zip(aggregated.items(), counts.items())
-    ):
-        series.plot(
-            ax=axes,
-            kind="bar",
-            label=f"{label} ({count.iloc[0]})",
-            color=colors[i],
-            position=positions[i],
-            rot=0,
-            zorder=2 - i * 0.1,
-        )
-
-    axes.legend()
-    axes.grid(visible=False, axis="x")
-
-
-def main():
-    """Plot figure."""
-    nrows, ncols = 2, 2
-    plt.rcParams.update(**shared.get_fontsizes())
-    plt.rcParams.update(
-        **shared.get_figsizes(
-            nrows=nrows,
-            ncols=ncols,
-            aspect_ratio=1.6,
-        )
-    )
-    plt.rcParams.update(shared.get_axes_params())
-    fig = plt.figure()
-    gs = gridspec.GridSpec(nrows=nrows, ncols=ncols, figure=fig)
-
-    t_cat_ax = fig.add_subplot(gs[0,0])
-    ipsi_inv_ax = fig.add_subplot(gs[0,1], sharey=t_cat_ax)
-    midext_ax = fig.add_subplot(gs[1,0], sharey=ipsi_inv_ax)
-    uncorr_ax = fig.add_subplot(gs[1,1], sharey=midext_ax)
-    row = fig.add_subplot(gs[:], frame_on=False)
-
-    ipsi_inv_ax.set_title("Ipsilateral Involvement", fontweight="bold")
[…]

### 2024-09-18 scripts/data_upset_ipsi.py

@@ -1,52 +1,0 @@
-"""Upset plot of ipsilateral lymphatic involvement patterns."""
-
-import matplotlib.pyplot as plt
-import pandas as pd
-import paths
-import shared
-import upsetplot
-from matplotlib.axes import Axes
-
-from lyscripts import utils
-from lyscripts.plot.utils import COLORS
-
-
-def plot_involvement_upset(
-    df: pd.DataFrame,
-    lnls: list[str],
-) -> Axes:
-    """Plot involvement of `side` as upsetplot."""
-    indicator_data = upsetplot.from_indicators(
-        indicators=lnls,
-        data=df,
-    )
-    upset = upsetplot.UpSet(
-        indicator_data,
-        sort_by="cardinality",
-        sort_categories_by="-input",
-        min_subset_size=10,
-        facecolor=COLORS["blue"],
-    )
-    upset.style_subsets(
-        absent=lnls,
-        facecolor=COLORS["green"],
-        label="N0 (all healthy)",
-    )
-    axes = upset.plot()
-    axes["totals"].set_xlabel("Prevalence")
-    axes["intersections"].set_ylabel("Scenario size")
-    return axes
-
-
-def main():
-    """Plot figure."""
-    raw = utils.load_patient_data(paths.data)
-    plot_involvement_upset(
-        df=raw["max_llh", "ipsi"],
-        lnls=[tpl[-1] for tpl in shared.get_lnl_cols("ipsi")],
-    )
-    plt.savefig(shared.get_figure_path(__file__))
-
-
-if __name__ == "__main__":
-    main()

### 2024-09-18 scripts/model_burnin_history.py

@@ -1,79 +1,0 @@
-"""Plot the burn-in history of the sampling."""
-
-from collections import namedtuple
-
-import matplotlib.pyplot as plt
-import pandas as pd
-import shared
-from matplotlib import ticker
-
-
-def custom(x, pos):
-    """Return custom format."""
-    return f"{x / 1000:.1f}"
-
-
-PlotConfig = namedtuple("PlotConfig", ["title", "ylabel", "label"])
-
-
-def main():
-    """Plot figure."""
-    nrows, ncols = 1, 2
-    config_map = {
-        "acor_times": PlotConfig(
-            title="Estimated Autocorrelation Time",
-            ylabel="estimate [steps]",
-            label="Autocorrelation Time",
-        ),
-        "accept_fracs": PlotConfig(
-            title="Average Acceptance Fraction of Walkers",
-            ylabel="fraction [%]",
-            label=None,
-        ),
-    }
-
-    plt.rcParams.update(shared.get_fontsizes())
-    plt.rcParams.update(
-        shared.get_figsizes(
-            nrows=nrows,
-            ncols=ncols,
-            width=17,
-        )
-    )
-    plt.rcParams.update(shared.get_axes_params())
-
-    fig, axes = plt.subplots(
-        nrows=nrows,
-        ncols=ncols,
-        sharex=True,
-    )
-
-    for model in ["full"]:
-        history = pd.read_csv(f"models/{model}/history.csv").set_index("steps")
-        for i, (column, config) in enumerate(config_map.items()):
-            axes[i].set_title(config.title, fontweight="bold")
-            axes[i].plot(history.index, history[column], label=config.label)
-            axes[i].autoscale(enable=True, tight=True, axis="x")
-            axes[i].set_ylabel(config.ylabel)
-            axes[i].xaxis.set_major_formatter(ticker.FuncFormatter(custom))
-            axes[i].set_xlabel(r"steps [$\times 10^3$]")
-
-    params = shared.get_params()
-
-    axes[0].plot(
-        history.index,
-        history.index / params["sampling"]["trust_fac"],
-        "--",
-        color="black",
-        label="Trust Threshold",
-    )
-    axes[0].legend()
-    axes[1].yaxis.set_major_formatter(
-        ticker.FuncFormatter(lambda x, *_: f"{100 * x:.0f}")
-    )
-
-    plt.savefig(shared.get_figure_path(__file__))
-
-
-if __name__ == "__main__":
-    main()

### 2024-09-18 scripts/model_midext_evo.py

@@ -1,90 +1,0 @@
-"""Time-prior and evolution over midline extension probability."""
-
-import matplotlib.pyplot as plt
-import numpy as np
-import shared
-from lyscripts.plot.utils import COLORS
-from matplotlib import ticker
-
-
-def main():
-    """Plot figure."""
-    model = shared.get_model("simple", load_samples=True)
-
-    t = np.linspace(0, 10, 11)
-    p_midline = {
-        "lateralized": (1 - model.midext_prob) ** t,
-        "extension": 1 - (1 - model.midext_prob) ** t,
-    }
-    dist = {
-        "early": model.get_distribution("early").pmf,
-        "late": model.get_distribution("late").pmf,
-    }
-    p_colors = {"lateralized": COLORS["green"], "extension": COLORS["red"]}
-    t_colors = {"early": COLORS["blue"], "late": COLORS["orange"]}
-
-    nrows, ncols = 2, 1
-    plt.rcParams.update(shared.get_fontsizes())
-    plt.rcParams.update(
-        shared.get_figsizes(
-            nrows=nrows,
-            ncols=ncols,
-            width=17 / 2,
-        )
-    )
-    plt.rcParams.update(shared.get_axes_params())
-
-    fig, axes = plt.subplots(nrows=nrows, ncols=ncols, sharex=True)
-    w = 0.3
-
-    for label, p in p_midline.items():
-        axes[0].plot(
-            t,
-            p_midline[label],
-            "o-",
-            label=rf"cond. prob. $P(\epsilon={label=='extension'} \mid t)$",
-            color=p_colors[label],
-        )
-        for i, t_stage in enumerate(["early", "late"]):
-            if label == "lateralized":
-                axes[0].bar(
-                    t + i * w - w / 2,
-                    dist[t_stage],
-                    color=t_colors[t_stage],
-                    width=w,
-                    label=f"{t_stage} T-cat. prior $P(t)$",
-                )
-            axes[1].plot(
-                t,
-                p * dist[t_stage],
-                "o-",
-                c=p_colors[label],
-                mfc=t_colors[t_stage],
-                mec=t_colors[t_stage],
-                label=rf"{label} ($\epsilon={label=='extension'}$) for {t_stage} T-cat."
-            )
-
[…]

### 2024-09-18 scripts/model_prevalences_midext.py

@@ -1,53 +1,0 @@
-"""Compute the prevalence of midline extension (predicted and observed)."""
-
-import matplotlib.pyplot as plt
-import paths
-import shared
-
-from lyscripts.plot.utils import COLORS, BetaPosterior, Histogram, draw, split_legends
-
-
-def main():
-    """Plot figure."""
-    plt.rcParams.update(shared.get_fontsizes())
-    plt.rcParams.update(shared.get_figsizes(width=17.0 / 2.0))
-
-    fig, ax = plt.subplots()
-    content = []
-    colors = iter([COLORS[c] for c in ["blue", "orange"]])
-
-    for t_stage in ["early", "late"]:
-        color = next(colors)
-        content.append(
-            Histogram.from_hdf5(
-                filename=paths.model_dir / "full" / "prevalences_midext.hdf5",
-                dataname=t_stage,
-                color=color,
-                label="prediction",
-            )
-        )
-        content.append(
-            BetaPosterior.from_hdf5(
-                filename=paths.model_dir / "full" / "prevalences_midext.hdf5",
-                dataname=t_stage,
-                color=color,
-            )
-        )
-
-    draw(ax, content, xlims=(0.0, 55.0))
-    ax.set_title("Observed vs Predicted Prevalence\nof Midline Extension $\\epsilon$")
-    ax.set_yticks([])
-    ax.set_ylim(top=0.4)
-    ax.set_xlabel("prevalence [%]")
-
-    split_legends(
-        axes=ax,
-        titles=["early T-cat.", "late T-cat."],
-        locs=[(0.05, 0.99), (0.6, 0.99)],
-    )
-
-    plt.savefig(shared.get_figure_path(__file__))
-
-
-if __name__ == "__main__":
-    main()

### 2024-09-18 scripts/model_prevalences_overall.py

@@ -1,111 +1,0 @@
-"""Compare overall prevalences in the data with the model predictions."""
-
-import matplotlib.pyplot as plt
-import numpy as np
-import paths
-import shared
-from matplotlib import gridspec
-
-from lyscripts.plot.utils import COLORS, BetaPosterior, Histogram, draw, split_legends
-from lyscripts.scenario import Scenario
-
-
-def main():
-    """Plot the figure."""
-    nrows, ncols = 2, 2
-
-    plt.rcParams.update(shared.get_fontsizes())
-    plt.rcParams.update(
-        shared.get_figsizes(
-            nrows=nrows,
-            ncols=ncols,
-            aspect_ratio=2.0,
-            width=17,
-        )
-    )
-
-    fig = plt.figure()
-    gs = gridspec.GridSpec(nrows=nrows, ncols=ncols, figure=fig)
-    axes = np.array(
-        [[fig.add_subplot(gs[i, j]) for j in range(ncols)] for i in range(nrows)]
-    )
-    full = fig.add_subplot(gs[:, :], frame_on=False)
-    shared.turn_axis_off(full)
-
-    full.set_title(
-        (
-            "Observed vs Predicted Prevalence of Contralateral Involvement"
-            "\nin the LNLs II, III, and IV"
-        ),
-        pad=20.0,
-        fontsize="large",
-    )
-
-    for i, t_stage in enumerate(["early", "late"]):
-        axes[i, 0].set_ylabel(f"{t_stage} T-cat.", fontweight="bold")
-        for j, midext in enumerate([False, True]):
-            axes[0, j].set_title(
-                "Mid. ext." if midext else "Lateralized", fontweight="bold"
-            )
-            axes[0, j].set_xticks([])
-            axes[-1, j].set_xlabel("prevalence [%]")
-            axes[i, j].set_yticks([])
-            content = []
-            for lnl, color in zip(
-                ["II", "III", "IV"],
-                [COLORS["blue"], COLORS["orange"], COLORS["green"]],
-            ):
-                scenario = Scenario(
-                    t_stages=[t_stage],
-                    midext=midext,
-                    diagnosis={"contra": {"max_llh": {lnl: True}}},
-                )
-                content.append(
-                    Histogram.from_hdf5(
-                        filename=paths.model_dir / "full" / "prevalences_overall.hdf5",
-                        dataname=scenario.md5_hash("prevalences"),
[…]

### 2024-09-18 scripts/model_prevalences_upstream.py

@@ -1,112 +1,0 @@
-"""Predicted/observed prevalences of contra III involvement, dep. on upstream level."""
-
-import matplotlib.pyplot as plt
-import numpy as np
-import paths
-import shared
-from matplotlib import gridspec
-
-from lyscripts.plot.utils import COLORS, BetaPosterior, Histogram, draw, split_legends
-from lyscripts.scenario import Scenario
-
-
-def main():
-    """Plot figure."""
-    nrows, ncols = 2, 2
-
-    plt.rcParams.update(shared.get_fontsizes())
-    plt.rcParams.update(
-        shared.get_figsizes(
-            nrows=nrows,
-            ncols=ncols,
-            aspect_ratio=2.0,
-            width=17,
-        )
-    )
-
-    fig = plt.figure()
-    gs = gridspec.GridSpec(nrows=nrows, ncols=ncols, figure=fig)
-    axes = np.array(
-        [[fig.add_subplot(gs[i, j]) for j in range(ncols)] for i in range(nrows)]
-    )
-    full = fig.add_subplot(gs[:, :], frame_on=False)
-    shared.turn_axis_off(full)
-
-    full.set_title(
-        (
-            "Observed vs Predicted Prevalence of Contralateral LNL III Involvement"
-            "\nDependent on Upstream LNL II Involvement"
-        ),
-        pad=20.0,
-        fontsize="large",
-    )
-
-    for i, t_stage in enumerate(["early", "late"]):
-        axes[i, 0].set_ylabel(f"{t_stage} T-cat.", fontweight="bold")
-        for j, midext in enumerate([False, True]):
-            axes[0, j].set_title(
-                "Mid. ext." if midext else "Lateralized", fontweight="bold"
-            )
-            axes[0, j].set_xticks([])
-            axes[-1, j].set_xlabel("prevalence [%]")
-            axes[i, j].set_yticks([])
-            content = []
-            for upstream, color in zip([False, True], [COLORS["green"], COLORS["red"]]):
-                scenario = Scenario(
-                    t_stages=[t_stage],
-                    midext=midext,
-                    diagnosis={
-                        "contra": {"max_llh": {"II": upstream, "III": True}},
-                    },
-                )
-                content.append(
-                    Histogram.from_hdf5(
-                        filename=paths.model_dir
-                        / "full"
-                        / "prevalences_with_upstream.hdf5",
-                        dataname=scenario.md5_hash("prevalences"),
-                        color=color,
[…]

### 2024-09-18 scripts/model_prevalences_with_ipsi.py

@@ -1,121 +1,0 @@
-"""Predicted/observed prevalences of contra II involvement, dep. on ipsi involvement."""
-
-import matplotlib.pyplot as plt
-import numpy as np
-import paths
-import shared
-from matplotlib import gridspec
-
-from lyscripts.plot.utils import COLORS, BetaPosterior, Histogram, draw, split_legends
-from lyscripts.scenario import Scenario
-
-
-def main():
-    """Plot figure."""
-    nrows, ncols = 2, 2
-
-    plt.rcParams.update(shared.get_fontsizes())
-    plt.rcParams.update(
-        shared.get_figsizes(
-            nrows=nrows,
-            ncols=ncols,
-            aspect_ratio=2.0,
-            width=17,
-        )
-    )
-
-    fig = plt.figure()
-    gs = gridspec.GridSpec(nrows=nrows, ncols=ncols, figure=fig)
-    axes = np.array(
-        [[fig.add_subplot(gs[i, j]) for j in range(ncols)] for i in range(nrows)]
-    )
-    full = fig.add_subplot(gs[:, :], frame_on=False)
-    shared.turn_axis_off(full)
-
-    full.set_title(
-        (
-            "Observed vs Predicted Prevalence of Contralateral LNL II Involvement"
-            "\nDependent on Ipsilateral Involvement"
-        ),
-        pad=20.0,
-        fontsize="large",
-    )
-
-    for i, t_stage in enumerate(["early", "late"]):
-        axes[i, 0].set_ylabel(f"{t_stage} T-cat.", fontweight="bold")
-        for j, midext in enumerate([False, True]):
-            axes[0, j].set_title(
-                "Mid. ext." if midext else "Lateralized", fontweight="bold"
-            )
-            axes[0, j].set_xticks([])
-            axes[-1, j].set_xlabel("prevalence [%]")
-            axes[i, j].set_yticks([])
-            content = []
-            for ipsi, color in zip(
-                [
-                    {"I": False, "II": False, "III": False, "IV": False, "V": False},
-                    {"II": True},
-                    {"II": True, "III": True},
-                ],
-                [COLORS["green"], COLORS["orange"], COLORS["red"]],
-            ):
-                scenario = Scenario(
-                    t_stages=[t_stage],
-                    midext=midext,
-                    diagnosis={
-                        "ipsi": {"max_llh": ipsi},
-                        "contra": {"max_llh": {"II": True}},
-                    },
[…]

### 2024-09-18 scripts/model_risks.py

@@ -1,139 +1,0 @@
-"""Predict the risk for selected scenarios and plot the results."""
-
-import h5py
-import matplotlib.pyplot as plt
-import numpy as np
-import paths
-import shared
-
-from lyscripts.plot.utils import COLORS, Histogram, draw
-from lyscripts.scenario import Scenario
-
-
-def get_diag_label(diagnosis) -> str:
-    """Get the diagnosis label."""
-    label = ""
-    is_n0 = True
-
-    ipsi_positive = [
-        lnl for lnl, status in diagnosis.get("ipsi", {}).get("CT", {}).items() if status
-    ]
-    if len(ipsi_positive) > 0:
-        is_n0 = False
-        label += "i:" + ",".join(ipsi_positive)
-
-    contra_positive = [
-        lnl
-        for lnl, status in diagnosis.get("contra", {}).get("CT", {}).items()
-        if status
-    ]
-    if len(contra_positive) > 0:
-        is_n0 = False
-        label += "; c:" + ",".join(contra_positive)
-
-    return label if not is_n0 else "N0"
-
-
-def get_label(scenario: Scenario) -> str:
-    """Get the label for the scenario."""
-    lnl = list(eval(scenario.involvement)["contra"].keys()).pop()
-    d = eval(scenario.diagnosis)
-    has_upstream_fna = "FNA" in d["contra"]
-    t_stage_map = {"early": "early", "late": "adv."}
-    midext_map = {False: "lateral", True: "mid-ext"}
-
-    if lnl == "IV":
-        return get_diag_label(d) + (" & c:III FNA+" if has_upstream_fna else "")
-
-    m = scenario.midext
-
-    if lnl == "III":
-        return f"{midext_map[m]}; {get_diag_label(d)}" + (
-            " & c:II FNA+" if has_upstream_fna else ""
-        )
-
-    t = scenario.t_stages[0]
-    return f"{t_stage_map[t]}; {midext_map[m]}; {get_diag_label(d)}"
-
-
-def main():
-    """Plot figure."""
-    nrows, ncols = 1, 3
-    plt.rcParams.update(shared.get_fontsizes())
-    plt.rcParams.update(
-        shared.get_figsizes(
-            nrows=nrows,
-            ncols=ncols,
-            width=17,
-            aspect_ratio=1.3,
-        )
-    )
-
-    fig, axes = plt.subplots(nrows=nrows, ncols=ncols)
-
[…]

### 2024-09-18 scripts/model_state_dist.py

@@ -1,96 +1,0 @@
-"""Plot the term P(X_i | t)^T @ P(t) @ P(X_c | t) for the simple model."""
-
-import operator
-from matplotlib.gridspec import GridSpec
-import matplotlib.pyplot as plt
-import numpy as np
-import shared
-from matplotlib.colors import LinearSegmentedColormap
-
-from lyscripts.plot.utils import COLORS
-
-
-def main():
-    """Plot figure."""
-    # use the full model's parameters in the smaller, simple model
-    smpl_model = shared.get_model(which="simple", load_samples=False)
-    full_model = shared.get_model(which="full", load_samples=True)
-    smpl_model.set_params(**full_model.get_params())
-
-    ipsi_state_dist_evo = smpl_model.ext.ipsi.state_dist_evo() * 100.0
-    _, contra_state_dist_evo = smpl_model.contra_state_dist_evo()
-    contra_state_dist_evo *= 100.0
-    time_prior = np.diag(smpl_model.get_distribution("late").pmf) * 100.0
-
-    vmin = np.min(
-        [
-            ipsi_state_dist_evo.min(),
-            time_prior.min(),
-            contra_state_dist_evo.min(),
-        ]
-    )
-    vmax = np.max(
-        [
-            ipsi_state_dist_evo.max(),
-            time_prior.max(),
-            contra_state_dist_evo.max(),
-        ]
-    )
-
-    nrows, ncols = 1, 3
-    plt.rcParams.update(shared.get_fontsizes())
-    plt.rcParams.update(
-        shared.get_figsizes(
-            nrows=nrows,
-            ncols=ncols,
-            aspect_ratio=0.7675,
-            width=17,
-        )
-    )
-
-    fig = plt.figure()
-    gs = GridSpec(nrows + 1, 2 * ncols, figure=fig, height_ratios=[1, 0.075])
-
-    ipsi = fig.add_subplot(gs[0, 0:2])
-    ipsi.set_aspect(operator.truediv(*ipsi_state_dist_evo.shape))
-
-    time = fig.add_subplot(gs[0, 2:4])
-    time.set_aspect(operator.truediv(*time_prior.shape))
-
-    contra = fig.add_subplot(gs[0, 4:6], sharey=time)
-    contra.set_aspect(operator.truediv(*contra_state_dist_evo.shape))
-
[…]

### 2024-09-18 scripts/paths.py

@@ -1,20 +1,0 @@
-"""Define paths for data and models."""
-
-from pathlib import Path
-from typing import Literal
-
-data_dir = Path("data")
-data = data_dir / "reduced.csv"
-model_dir = Path("models")
-cache_dir = Path(".cache")
-figure_dir = Path("figures")
-scenario_dir = Path("scenarios")
-params = Path("_variables.yml")
-
-
-def get_filename(
-    model: Literal["ipsi", "contra", "bilateral", "midline"],
-    kind: Literal["params", "samples", "priors"] | str,
-) -> Path:
-    """Get the filename of the samples of one of the four trained models."""
-    return next((model_dir / model).glob(f"{kind}.*"))

### 2024-09-18 scripts/reduce.py

@@ -1,35 +1,0 @@
-"""Remove all unnecessary columns from the joined, enhanced and (row-wise) filtered data.
-"""
-import argparse
-from pathlib import Path
-
-import pandas as pd
-
-
-def create_parser():
-    """Create the argument parser."""
-    parser = argparse.ArgumentParser(description=__doc__)
-    parser.add_argument('input', type=Path, help='Path to the input file.')
-    parser.add_argument('output', type=Path, help='Path to the output file.')
-    return parser
-
-
-def main():
-    """Main function."""
-    parser = create_parser()
-    args = parser.parse_args()
-
-    data = pd.read_csv(
-        args.input, header=[0,1,2],
-    ).drop(columns=[
-        "enbloc_dissected",
-        "enbloc_positive",
-        "positive_dissected",
-        "total_dissected",
-    ])
-
-    data.to_csv(args.output, index=False)
-
-
-if __name__ == '__main__':
-    main()

### 2024-09-18 scripts/shared.py

@@ -1,184 +1,0 @@
-"""Some shared and reused functions for the scripts."""
-
-from collections import namedtuple
-from pathlib import Path
-from typing import Any, Literal
-
-import numpy as np
-import pandas as pd
-from lymph import types
-
-from lyscripts import utils
-from lyscripts.plot.utils import COLORS
-
-try:
-    from . import paths
-except ImportError:
-    import paths
-
-
-pd.options.mode.copy_on_write = False
-CM_TO_INCH = 0.393701
-GOLDEN_RATIO = 1.61803398875
-
-Columns = namedtuple(
-    "Columns", [
-        "inst",
-        "age",
-        "nd",
-        "t_stage",
-        "n_stage",
-        "midext",
-        "ipsi_III",
-    ])
-
-COL = Columns(
-    inst=("patient", "#", "institution"),
-    age=("patient", "#", "age"),
-    nd=("patient", "#", "neck_dissection"),
-    t_stage=("tumor", "1", "t_stage"),
-    n_stage=("patient", "#", "n_stage"),
-    midext=("tumor", "1", "extension"),
-    ipsi_III=("max_llh", "ipsi", "III"),
-)
-CONTRA_LNLS = [
-    ("max_llh", "contra", "I"),
-    ("max_llh", "contra", "II"),
-    ("max_llh", "contra", "III"),
-    ("max_llh", "contra", "IV"),
-    ("max_llh", "contra", "V"),
-]
-
-
-def get_params() -> dict[str, Any]:
-    """Get the parameters."""
-    return utils.load_yaml_params(paths.params)
-
-
-def get_model(
-    which: str,
-    load_samples: bool = True,
-) -> types.Model:
-    """Get one of the four trained models."""
-    params_path = paths.model_dir / which / "params.yaml"
-    params = utils.load_yaml_params(params_path)
-    model = utils.create_model(params)
-
-    if not load_samples:
-        return model
-
-    samples_path = paths.model_dir / which / "samples.hdf5"
-    samples = utils.load_model_samples(samples_path)
-    model.set_params(*samples.mean(axis=0))
-    return model
-
-
-def get_samples(which: str) -> np.ndarray:
-    """Get the samples of one of the four trained models."""
-    samples_path = paths.model_dir / which / "samples.hdf5"
-    return utils.load_model_samples(samples_path)
-
-
-def get_data(map_t_stage: bool = True) -> pd.DataFrame:
-    """Get the data."""
[…]

### 2024-09-26 manuscript.qmd
first version

@@
 However, this approach does not describe differences in lymphatic spread between different subsites within the oropharynx and oral cavity.
-This resulted in distorted predictions for each subsite, as the spread patterns were pooled into one single model .To address this issue, we present an approach using mixtures of HMMs.
+This resulted in distorted predictions for each subsite, as the spread patterns were pooled into one single model.
+To address this issue, we present an approach using mixtures of HMMs.
 The intuition is that the lymphatic spread of a tumor that lies anatomically at the border of oropharynx and oral cavity (e.g.
 tumors in the palate) may be described by a mixture of different models.
+We further extend the analysis to a mixture model that considers oral cavity, oropharynx, hypopharynx and larynx.
 
 # Data on Lymphatic Progression Patterns {#sec-data}
+
+For the analyses in this work, we used seven datasets from 5 institutions resulting in 2741 patients in total.
+
+1. xxx oropharyngeal patients from the University of Zurich in Switzerland
+2. xxx oropharyngeal patients from the Centre Léon Bérard in France
+3. xxx oropharyngeal, larynx and oral cavity patients from the Inselspital Bern in Switzerland
+4. xxx oropharyngeal, larynx and oral cavity patients from the Centre Léon Bérard in France
+5. xxx oropharyngeal, larynx and oral cavity patients from the University of Zurich in Switzerland
+6. 164 oropharyngeal patients from the Hospital Vall d'Hebron in Spain (not yet public)
+7. 979 hypopharynx, larynx and oral cavity patients from University Medical Center Groningen (not yet public)
+
+The datasets 1-4 are publicly available as CSV tables (@ludwig_multi-centric_2023 @ludwig_detailed_2022) and can be interactively explored on [LyProX](https://lyprox.org).
+For each patient the primary tumor subsite is reportd and each indicidual LNL is reported as either metastatic or healthy given the available diagnostic modalities, which include pathology after neck dissection in some patients.
+In this work we will stratify the tumor locations into different ICD codes which are depicted in @fig-subsites.
+
+![Anatomical sketch of the tumor subsites and their corresponding ICD-10 codes.
+Subsite C06 "other parts of mouth" has not been included.
+Further the The tumor locations are color coded in the following pattern: blue-oral cavity, green-oropharynx, red-hypopharynx, orange-larynx.](figures/Subsites.png){#fig-subsites}
+
+The prevelance of involvement in LNLs I, II, III, IV and V is shown in @fig-involvement.
+The involvement is stratified per tumor subsite and t-stage.
+The figure illustrates the variations in LNL involvement between subsites within oral cavity (blue), oropharynx (green), hypopharynx (red) and larynx (orange).
+The involvement pattern presents a continuous change over the tumor subsites.
+Where tumors in the oral cavity show the most prominent LNL I involvement.
+As the tumor location moves towards the oropharynx LNL II involvement increases.
+Moving the tumor location further in caudal direction towards the hypopharynx increases LNL III involvement while LNL I and II involvement decrease.
+Larnygeal tumors show the least LNL I involvement.
+
+![Prevalence of ipsilateral LNL involvement stratified by subsite.
+The subsites are sorted in natural order to represent the continuously changing LNL involvement.
+The different tumo locations are color coded, where oral cavity subsistes are depicted in blue, larynx in green, hypopharynx in red and larynx in orange.
+The patient data is further stratified in early t-stage (0-2) and late t-stage (3-4).
+The legend furter specifies the number of patients in each subsite.](figures/involvement_I_to_V_all_sites_t_staging.png){#fig-involvement}
 
@@
 Since primary tumors at different subsites have different patterns of lymphatic spread patterns one would need to either generalize over several subsites as introduced in [@ludwig_dynamic_2021], where the HMM has been introduced for oropharyngeal cancer, or train a distinct model for each subsite.
-The former approach would lead to less precise predictions for each subsite, especially those subsites with few patients.
-Training one HMM per subsite would be rather time intensive and lead to large uncertainties for subsites with only few patients.
+The former approach leads to less precise predictions for each subsite, especially those subsites with few patients.
+Training one HMM per subsite would be rather time intensive and lead to large uncertainties for subsites with only few patients such as C04 Floor of mouth or C05 Palate.
 Further we would not exploit the anatomical similarities which are expected for anatomically nearby locations.
 
-Therefore we introduce a mixture model which combines all subsites in a singular model exploiting anatomical proximities.
+Therefore we introduce a mixture model which combines all subsites in a single model exploiting anatomical proximities.
+Let us assume that we have a dataset $\mathbf{D}$ that is specified via the number of patients $N_{is}$ that were diagnosed in LNL involvement state $i$ and had a primary tutor in subsite $s$.
+Let us further assume that we want to describe this dataset using a mixture of $M$ HMMs, each with a different set of model parameters $\boldsymbol{\theta_m}$.
+As the generative model of the data, we assume that a patient with subsite $s$ is generated with probability $\pi_m^s$ from model $m$, where the mixing components $\pi_m^s$ need to satisfy $\sum_{m = 1}^{M}\pi_m^s = 1$, with $\pi_m^s \geq 0$ for all $m$ and all subsites $s$.
+The likelihood of the dataset can then be written as
+
+$$
+P \left( \mathbf{D} \mid \boldsymbol{\theta}, \boldsymbol{\pi}\right) = \prod_s \prod_i \left[ \sum_{m=1}^M \pi_m^s P_m \left( \mathbf{X}=\boldsymbol{\xi}_i \mid \boldsymbol{\theta}_m \right) \right]^{N_{is}}
+$$ {#eq-mixture-distribution}
+
+The model now contains two types of parameters.
+The rpobabilities of tumor spread for the different models, $\mathbf{\theta_m}$, and the mixing coefficients $\pi_m^s$.
+The posterior distribution over the parameters $P(\boldsymbol{\theta},\boldsymbol{\pi}|\mathbf{D})$ can be calculated assuming a uniform prior in the interval [0,1] for all parameters.
+However, $P(\boldsymbol{\theta},\boldsymbol{\pi}|\mathbf{D})$ is a multimodal-distribution as one can permute the different models.
+To address this problem, we apply an expectation-maximization (EM) algorithm.
+We introduce latent variables $\boldsymbol{\epsilon}$ which assigns to each observation $\mathbf{B}$ in our dataset the probability to originate from each model $m$.
+The log-likelihood can be expressed as:
+
+$$
+\ln p(\mathbf{X}|\boldsymbol{\theta},\boldsymbol{\pi}) = \left\{\sum_{\boldsymbol{\epsilon}} p(\mathbf{X},\boldsymbol{\epsilon}|\boldsymbol{\theta}, \boldsymbol{\pi})\right\}
+$$ {#eq-mixture-llh}
+
+ The EM algorithm evaluates the model parameters $\boldsymbol{\theta}$ and $\boldsymbol{\pi}$ in two steps.
+In the expectation step the latent variables are evaluated.
+To do so, we fix the current model parameters $\boldsymbol{\theta}$ and $\boldsymbol{\pi}$ and compute the expectation value of the latent variables.
+We can compute the expectation values $\boldsymbol{\gamma}$ of the latent variables for each diagnosis $Z_i$ separatly.
+
+ $$
+\mathbb{E}[\epsilon_m] = \gamma(\epsilon_m^i) = \frac{\pi_m^s p(Z_i|\theta_m)}{\sum_{j=1}^M \pi_j^sp(Z_i|\theta_j)} 
+ $$ {#eq-expectation-step}
+
+The @eq-expectation-step includes a superscript $s$ for the mixture parameters $\pi$ which depends on the tumor subsite of diagnosis $Z_i$. 
+After computing the expectation value for all components $M$ and patients $N$, the model parameters are updated by in the maximization step by maximizing the following likelihood:
+$$
+ \ln p(\mathbf{Z},\boldsymbol{\epsilon}|\boldsymbol{\theta},\boldsymbol{\pi}) = \sum_{i=1}^{N}\sum_{m=1}^M \gamma(\epsilon_m^i)\{\ln \pi_m^s + \ln p(Z_i|\theta_m)\}
+ $$ {#eq-maximization-step}
+
+In @eq-maximization-step one can see, that we can evaluate the mixture parameters $\boldsymbol{\pi}$ and model parameter $\boldsymbol{\theta}$ idependently of each other.
+The mixture parameters can be calculated analytically:
+
+$$
+\pi_m^s = \frac{1}{N^s}\sum_i^{N^s}\gamma(\epsilon_m^i)
+$$ {#eq-mixture-maximization}
+
+The model parameters $\boldsymbol{\theta}$ can be optimized component-wise with an optimization algorithm, since an analytic solution is not feasible:
+
+$$
+\ln p(\mathbf{Z},\boldsymbol{\epsilon}|\boldsymbol{\theta_m}) = \sum_{i=1}^{N} \gamma(\epsilon_m^i)\{\ln p(Z_i|\theta_m)\}
+$$
+
+By repeating these steps the algorithm is guaranteed to converge to a (local) maximum.
+
+# Three component Mixture Model {#sec-3comp}
+
+We illustrate the methodology for a mixture model with M = 3 components, considering the ipsilateral involvement of LNLs I, II, III, IV, and V.
+We include the ICD codes as subsites for oral cavity, hypopharynx and oropharynx.
+In @fig-convergence the convergence of the negative log-likelihood and change in model parameters is depicted.
+After a random inizialization, the algorithm rapidly converges.
+
+![The y-axis on the left shows the negative likelihood convergence depicted in the blue line.
+The y-axis on the right shows the sum of absulte difference between all model parameters showing that the parameter values stabilize rapidly as well.](figures/Convergence_3_comp.png){#fig-convergence}
+
+In @fig-3_simplex, we visualize the resulting mixture coefficients $\boldsymbol{\pi}$ using a spatial representation, where the vertices of the triangle correspond to the three components.
+In @fig-3-matrix, these mixture coefficients are presented in matrix form, with the y-axis representing the ICD codes, showing how the mixture components in each row add up to 1.
+
+The spatial plot in @fig-3_simplex illustrates how the model assigns the three components to different tumor subsites.
+Component 0, located at the bottom right of the triangle, primarily characterizes oropharyngeal subsites.
+For instance, the base of tongue subsite (C01), which exhibits the highest involvement of LNL II, is fully assigned to this component.
+Similarly, subsite C10, which includes several oropharyngeal regions, is assigned roughly 50% to the oropharynx-like component, with the remaining mixture distributed across the other two components.
+
+Hypopharyngeal subsites, on the other hand, are fully assigned to Component 2, located at the top vertex of the triangle.
+Meanwhile, the gum subsite (C03), with predominant LNL I involvement, is entirely assigned to Component 1, situated at the bottom left.
+As subsites anatomically approach the oropharynx, their mixture coefficients for the oropharynx-like component increase.
+This is evident in the subsites C02 (tongue) and C05 (palate), which display a higher proportion of oropharyngeal influence in their mixture.
+These results conform well with the involvement patterns observed in the data.
+
+::: {layout="[75,30]" layout-valign="bottom"}
+
+![Assignment of each subsite to each of the three components.
+The closer a subsite is to a vertex, the more it is assigned to the corresponding component, with component 0 on the bottom right, 1 on the bottom left and 2 on the top.
+The size of the marker (area) corresponds to the number of patients in each subsite.](figures/mixture_components_3_simplex_t_stages_new.png){#fig-3_simplex}
+
+![Matrix representation of component assignment.
+Each row of the matrix corresponds to each ICD code.
+The collums represent the three different components](figures/mixture_components_3_t_stages_new.png){#fig-3-matrix}
+:::

### 2024-09-26 manuscript.qmd
extended by a bit

@@
 
+A patient's state of lymph node involvement $\mathbf{X}[t]$ evolves over discrete time steps $t$.
+Let us enumerate all $2^V$ possible states, representing all combinations of LNLs.
+In this paper, we consider ipsilateral LNLs I, II, III, IV and V, which amounts to 32 possible states.
+The HMM is specified by a transition matrix $\mathbf{A}$:
+
+$$
+\mathbf{A} = \begin{pmatrix} A_{ij} \end{pmatrix} = P \left( \mathbf{X}[t+1] = \boldsymbol{\xi}_j \mid \mathbf{X}[t] = \boldsymbol{\xi}_i \right)
+$$ {#eq-transition-matrix}
+
+whose elements $A_{ij}$ contain the conditional probabilities that a state $\mathbf{X}[t]=\boldsymbol{\xi}_i$ transitions to $\mathbf{X}[t+1]=\boldsymbol{\xi}_j$ over one time step.
+The transition matrix is specified and parameterised via the graphical model shown in @fig-schematic-and-graph.
+The red arcs in the graph of @fig-schematic-and-graph (right panel) are associated with the probability that the primary tumor spreads directly to a LNL (parameters $b_v$).
+The blue arcs describe the spread from an upstream LNL -- given it is already metastatic -- to a downstream level (parameters $t_{v \rightarrow v+1}$).
+
+Now, let $\boldsymbol{\pi}$ be the _starting distribution_
+
+$$
+\boldsymbol{\pi} = \begin{pmatrix} \pi_i \end{pmatrix} = P \left( \mathbf{X}[0] = \boldsymbol{\xi}_i \right)
+$$ {#eq-starting-distribution}
+
+denoting the probability to start in state $\boldsymbol{\xi}_i$ at time step 0.
+Assuming that every patient started with all LNLs being healthy, we set $\pi_i$ to zero for all states  except the completly healthy state $\boldsymbol{\xi} = \begin{pmatrix} 0, 0, 0, 0 \end{pmatrix}$, which has probability one.
+
+Using the quantities introduced so far, the probability $P \left( \mathbf{X}[t]=\boldsymbol{\xi}_i \right)$ to be in state $\boldsymbol{\theta}_i$ in time step $t$ can now be conveniently expressed as a matrix product:
+
+$$
+P \left( \mathbf{X}[t]=\boldsymbol{\xi}_i \right) = \left( \boldsymbol{\pi} \cdot \mathbf{A}^t \right)_i
+$$ {#eq-evolution}
+
+This evolution implicitly marginalizes over all possible paths to arrive at state $\boldsymbol{\xi}_i$ after $t$ time-steps.
+Additionally, we must marginalize over the unknown time of diagnosis using a time-prior $P_T(t)$.
+This finally defines the probability distribution over all states of lymph node involvement used in @eq-bayes-law.
+
+$$
+P \left( \mathbf{X}=\boldsymbol{\xi}_i \mid \boldsymbol{\theta} \right) = \sum_{t=0}^{t_\text{max}} P_T(t) \left( \boldsymbol{\pi} \cdot \mathbf{A}^t \right)_i
+$$ {#eq-marginalized-evolution}
+
+where $\boldsymbol{\theta}=\{ b_v, t_{v \rightarrow v+1} \}$ denotes the set of all model parameters (7 in our case).
+Fortunately, the exact length and shape of this distribution has little impact as previously shown [@ludwig_hidden_2021].
+We set $t_\text{max}=$ {{< var model.max_t >}} and $P_\text{early}(t)$ to a binomial distribution with parameter {{< var model.first_binom_prob >}}.
+Further details on the HMM can be found in @ludwig_hidden_2021 and @zora231470.
+
+![On the left: Anatomical sketch of the tumor subsites and corresponding ICD-10 codes considered in this work.
+The subsite "other parts of mouth" (C06) was not drawn.
+On the right: Parametrized graphical model of the lymphatic network considering four LNLs.
+Blue nodes represent the hidden states of LNLs $X_v$, while the red one is the tumor.
+Arcs represent possible routes of metastatic spread, associated with a probability.](../static/schematic_and_graph.png){#fig-schematic-and-graph fig-pos="th"}
+
 # Mixture Model for Lymphatic Spread {#sec-mixture}
@@
 To do so, we fix the current model parameters $\boldsymbol{\theta}$ and $\boldsymbol{\pi}$ and compute the expectation value of the latent variables.
-We can compute the expectation values $\boldsymbol{\gamma}$ of the latent variables for each diagnosis $Z_i$ separatly.
+We can compute the expectation values $\boldsymbol{\gamma}$ of the latent variables for each observed involvement pattern $\mathbf{X}_i$ in subsite $s$ separatly.
 
  $$
-\mathbb{E}[\epsilon_m] = \gamma(\epsilon_m^i) = \frac{\pi_m^s p(Z_i|\theta_m)}{\sum_{j=1}^M \pi_j^sp(Z_i|\theta_j)} 
+\mathbb{E}[\epsilon_m] = \gamma(\epsilon_m^i) = \frac{\pi_m^s p(\mathbf{X}_i|\theta_m)}{\sum_{j=1}^M \pi_j^sp(\mathbf{X}_i|\theta_j)} 
  $$ {#eq-expectation-step}

### 2024-09-30 manuscript.qmd
extended paper

@@
 
-For the analyses in this work, we used seven datasets from 5 institutions resulting in 2741 patients in total.
+For the analyses in this work, we used seven datasets from 5 institutions resulting in {{< var patients.total >}} patients in total.
 
-1. xxx oropharyngeal patients from the University of Zurich in Switzerland
-2. xxx oropharyngeal patients from the Centre Léon Bérard in France
-3. xxx oropharyngeal, larynx and oral cavity patients from the Inselspital Bern in Switzerland
-4. xxx oropharyngeal, larynx and oral cavity patients from the Centre Léon Bérard in France
-5. xxx oropharyngeal, larynx and oral cavity patients from the University of Zurich in Switzerland
-6. 164 oropharyngeal patients from the Hospital Vall d'Hebron in Spain (not yet public)
-7. 979 hypopharynx, larynx and oral cavity patients from University Medical Center Groningen (not yet public)
+1. {{< var patients.USZ_oropharynx >}} oropharyngeal patients from the University of Zurich in Switzerland
+2. {{< var patients.CLB_oropharynx >}} oropharyngeal patients from the Centre Léon Bérard in France
+3. {{< var patients.ISB_multiside >}} oropharyngeal, larynx and oral cavity patients from the Inselspital Bern in Switzerland
+4. {{< var patients.CLB_multiside >}} oropharyngeal, larynx and oral cavity patients from the Centre Léon Bérard in France
+5. {{< var patients.USZ_multiside >}} oropharyngeal, larynx and oral cavity patients from the University of Zurich in Switzerland
+6. {{< var patients.HVH >}} oropharyngeal patients from the Hospital Vall d'Hebron in Spain (not yet public)
+7. {{< var patients.HMC >}} hypopharynx, larynx and oral cavity patients from University Medical Center Groningen (not yet public)
 
@@
 
-A patient's state of lymph node involvement $\mathbf{X}[t]$ evolves over discrete time steps $t$.
-Let us enumerate all $2^V$ possible states, representing all combinations of LNLs.
+In this chapter we will briefly summarize unilateral model for ipsilateral lymph node involvement introduced in @ludwig_hidden_2021, presenting the notation which is then needed to extend the HMM to a mixture model encompassing multiple tumor subsites.
+
+The HMM describes each LNL $v \in 1, 2, ..., V$ by a binary random variable corresponding to the status of the LNL; healthy (0) or involved (1).
+The entire state of a patient with $V$ LNLs is defined by the $v$-dimensional vector $\mathbf{X} = [X_1, X_2, ...,X_V]$.
+In the HMM, a patient's involvement is modeled over time $t$.
+Thus, a patient's state of lymph node involvement $\mathbf{X}[t]$ evolves over discrete time steps $t$.
+Let us enumerate all $2^V$ possible states, representing all combinations of LNL involvement.
 In this paper, we consider ipsilateral LNLs I, II, III, IV and V, which amounts to 32 possible states.
@@
 whose elements $A_{ij}$ contain the conditional probabilities that a state $\mathbf{X}[t]=\boldsymbol{\xi}_i$ transitions to $\mathbf{X}[t+1]=\boldsymbol{\xi}_j$ over one time step.
-The transition matrix is specified and parameterised via the graphical model shown in @fig-schematic-and-graph.
-The red arcs in the graph of @fig-schematic-and-graph (right panel) are associated with the probability that the primary tumor spreads directly to a LNL (parameters $b_v$).
+The transition matrix is specified and parameterised via the graphical model shown in @fig-graph.
+The red arcs in the graph of @fig-graph (right panel) are associated with the probability that the primary tumor spreads directly to a LNL (parameters $b_v$).
 The blue arcs describe the spread from an upstream LNL -- given it is already metastatic -- to a downstream level (parameters $t_{v \rightarrow v+1}$).
@@
 denoting the probability to start in state $\boldsymbol{\xi}_i$ at time step 0.
-Assuming that every patient started with all LNLs being healthy, we set $\pi_i$ to zero for all states  except the completly healthy state $\boldsymbol{\xi} = \begin{pmatrix} 0, 0, 0, 0 \end{pmatrix}$, which has probability one.
+Assuming that every patient started with all LNLs being healthy, we set $\pi_i$ to zero for all states  except the completly healthy state $\boldsymbol{\xi} = \begin{pmatrix} 0, 0, 0, 0, 0 \end{pmatrix}$, which has probability one.
 
@@
 This evolution implicitly marginalizes over all possible paths to arrive at state $\boldsymbol{\xi}_i$ after $t$ time-steps.
-Additionally, we must marginalize over the unknown time of diagnosis using a time-prior $P_T(t)$.
+Additionally, we must marginalize over the unknown time of diagnosis using a time-prior $P_T(t)$ which is defined by a binomial distribution.
+The t-stage of the tumor can be included in the model by choosing different parameterizations, considering that a tumor in late t-stages was diagnosed later than a tumor in early t-stages, therefore shifting the probability of diagnosis to later time steps.
 This finally defines the probability distribution over all states of lymph node involvement used in @eq-bayes-law.
@@
 
-![On the left: Anatomical sketch of the tumor subsites and corresponding ICD-10 codes considered in this work.
-The subsite "other parts of mouth" (C06) was not drawn.
-On the right: Parametrized graphical model of the lymphatic network considering four LNLs.
+![Parametrized graphical model of the lymphatic network considering four LNLs.
 Blue nodes represent the hidden states of LNLs $X_v$, while the red one is the tumor.
-Arcs represent possible routes of metastatic spread, associated with a probability.](../static/schematic_and_graph.png){#fig-schematic-and-graph fig-pos="th"}
+Arcs represent possible routes of metastatic spread, associated with a probability.](figures/Model_5_LNL.png){#fig-graph width=20%}
+
+For model training we assume that
 

### 2024-09-30 manuscript.qmd
extended again

@@
 In this paper, we consider ipsilateral LNLs I, II, III, IV and V, which amounts to 32 possible states.
-The HMM is specified by a transition matrix $\mathbf{A}$:
+The HMM is then specified by a transition matrix $\mathbf{A}$:
 
@@
 The transition matrix is specified and parameterised via the graphical model shown in @fig-graph.
-The red arcs in the graph of @fig-graph (right panel) are associated with the probability that the primary tumor spreads directly to a LNL (parameters $b_v$).
+The red arcs in the graph of @fig-graph are associated with the probability that the primary tumor spreads directly to a LNL (parameters $b_v$).
 The blue arcs describe the spread from an upstream LNL -- given it is already metastatic -- to a downstream level (parameters $t_{v \rightarrow v+1}$).
+
+![Parametrized graphical model of the lymphatic network considering four LNLs.
+Blue nodes represent the hidden states of LNLs $X_v$, while the red one is the tumor.
+Arcs represent possible routes of metastatic spread, associated with a probability.](figures/Model_5_LNL.png){#fig-graph width=20%}
 
@@
 
-Using the quantities introduced so far, the probability $P \left( \mathbf{X}[t]=\boldsymbol{\xi}_i \right)$ to be in state $\boldsymbol{\theta}_i$ in time step $t$ can now be conveniently expressed as a matrix product:
+Using the quantities introduced so far, the probability $P \left( \mathbf{X}[t]=\boldsymbol{\xi}_i \right)$ to be in state $\boldsymbol{\xi}_i$ in time step $t$ can now be conveniently expressed as a matrix product:
 
@@
 Additionally, we must marginalize over the unknown time of diagnosis using a time-prior $P_T(t)$ which is defined by a binomial distribution.
-The t-stage of the tumor can be included in the model by choosing different parameterizations, considering that a tumor in late t-stages was diagnosed later than a tumor in early t-stages, therefore shifting the probability of diagnosis to later time steps.
+The t-stage of the tumor can be included in the model by choosing different parameterizations of the binomial distribution, considering that a tumor in late t-stages was diagnosed later than a tumor in early t-stages, therefore shifting the probability of diagnosis to later time steps.
 This finally defines the probability distribution over all states of lymph node involvement used in @eq-bayes-law.
@@
 
-![Parametrized graphical model of the lymphatic network considering four LNLs.
-Blue nodes represent the hidden states of LNLs $X_v$, while the red one is the tumor.
-Arcs represent possible routes of metastatic spread, associated with a probability.](figures/Model_5_LNL.png){#fig-graph width=20%}
+With @eq-marginalized-evolution we can compute the probability of a patient being in any state $\xi_i$. Therefore the likelihood of observing a the 
+For model training we assume that the diagnoses in our data $\mathbf{D}$ we observe correspond to the hidden state $\mathbf{X}$ of the patient.
+Thus, learning the model parameters corresponds to maximizing the probability of observing the dataset $\mathbf{D}$:
 
-For model training we assume that
+$$
+P \left( \mathbf{D} \mid \boldsymbol{\theta}, \boldsymbol{\pi}\right) = \prod_i \left[ \sum_{t=0}^{t_\text{max}} P_{T_i}(t) \left( \boldsymbol{\pi} \cdot \mathbf{A}^t \right)_i \right]^{N_{i}}
+$$ {#eq-likelihood-HMM}
+
+In @eq-likelihood-HMM we take the likelihood of observing a patient in t-stage $T_i$ and with involvement $\xi_i$ to the power of $N_i$, corresponding to the number of observations of the specific diagnosis $i$ in the dataset $\mathbf{D}$.
+By multiplying over all diagnoses $i$ we get the likelihood of observing dataset $\mathbf{D}$ given the previously intoduced model with parameters $\boldsymbol{\theta}$
 
@@
 After a random inizialization, the algorithm rapidly converges.
+The algorithm was stopped when the difference of log-likelihood between two iterations was below 0.01.
 
@@
 :::
+
+# Four component Mixture Model {#sec-4comp}
+
+We can extend the mixture model to include the larynx.
+The larynx patients are more finely divided into ICD codes C32.0, C32.1 and C32.2 as there is a notable difference between these ICD codes in @fig-involvement.
+
+Simlarly to the three component model, we can analyze the convergence over the iterations of the EM-algorithm.
+In @fig-fig-convergence4 we can see that in this more complex model, the likelihood space becomes more complex as at around 200 iterations, the negative log-likelihood starts to increase faster again.
+
+![The y-axis on the left shows the negative likelihood convergence depicted in the blue line.
+The y-axis on the right shows the sum of absulte difference between all model parameters.](figures/convergence_4_components_4_loc.png){#fig-convergence4}

### 2024-10-01 manuscript.qmd
finished a first proposal

@@
 # Introduction
-Head and Neck Squamous Cell Carcinoma (HNSCC) are known to spread through the lymphatic system and form metastases in lymph nodes [@mukherji_cervical_2001, @shah_patterns_1990].
-To avoid nodal recurrences, lymph node levels (LNL) at risk of harbouring occult metastases are electively irradiated.
-Guidelines for different tumor locations [@biau_selection_2019] are based on the overall prevalence of nodal disease as reported in the literature [@mukherji_cervical_2001, @shah_patterns_1990].
-To personalize the prediction of the risk of occult metastases, given a patient's individual diagnosis, we first published a large, multi-centric dataset where the lymphatic involvement per LNL is available for each patient[@ludwig_dataset_2022, @ludwig_multi-centric_2023].
-Subsequently we published an interpretable hidden Markov model (HMM), trained with the data to predict the risk for occult nodal disease, given an individual patient's diagnosis [@ludwig_hidden_2021].
-A personalized risk prediction may allow clinicians to safely reduce the elective clinical target volume (CTV-N) and reduce side-effects that degrade the patient's quality of life without compromising treatment efficacy [@batth_practical_2014].
+Head and Neck Squamous Cell Carcinoma (HNSCC) are known to spread through the lymphatic system  often leading to metastases in the lymph nodes [@mukherji_cervical_2001, @shah_patterns_1990].
+To minimize nodal recurrences, lymph node levels (LNL) at risk of harboring occult metastases are typically irradiated electively.
+Current guidelines for different tumor locations are based on the overall prevalence of nodal disease as reported in literature [@biau_selection_2019, @mukherji_cervical_2001, @shah_patterns_1990].
 
-So far we trained different models for different tumor locations such as oropharynx and oral cavity tumors.
-However, this approach does not describe differences in lymphatic spread between different subsites within the oropharynx and oral cavity.
-This resulted in distorted predictions for each subsite, as the spread patterns were pooled into one single model.
-To address this issue, we present an approach using mixtures of HMMs.
-The intuition is that the lymphatic spread of a tumor that lies anatomically at the border of oropharynx and oral cavity (e.g.
-tumors in the palate) may be described by a mixture of different models.
-We further extend the analysis to a mixture model that considers oral cavity, oropharynx, hypopharynx and larynx.
+To personalize the prediction of the risk of occult metastases, given a patient's individual diagnosis, we previously published a large, multi-centric dataset where the lymphatic involvement per LNL is available for each patient[@ludwig_dataset_2022, @ludwig_multi-centric_2023].Building on this dataset, we introduced an interpretable hidden Markov model (HMM), trained to predict the risk for occult nodal disease, given an individual patient's diagnosis [@ludwig_hidden_2021].
+
+Personalized risk predictions could enable clinicians to safely reduce the elective clinical target volume (CTV-N), potentially decreasing treatment-related side effects that impair a patient's quality of life, without compromising the efficacy of the treatment [@batth_practical_2014].
+
+Initially, separate models were trained for distinct tumor locations, such as the oropharynx and oral cavity.
+These tumor locations are also used in guidelines to define the elective target volumes [@biau_selection_2019].
+However, this approach did not account for variations in lymphatic spread between subsites within these tumor regions.
+With data for more than 2700 patients available, we can now further analyze subsite specific spread patterns.
+Closer analysis showed that pooling subsites into a single model led to inaccurate predictions, as it failed to capture distinct lymphatic spread patterns.
+To resolve this, we propose using a mixture of HMMs, which allows us to model the lymphatic spread more accurately for tumors located near anatomical borders, such as those between the oropharynx and oral cavity (e.g., tumors in the palate).
+
+Additionally, we extend the analysis to a broader mixture model that encompasses tumors of the oral cavity, oropharynx, hypopharynx, and larynx, offering a more comprehensive understanding of lymphatic spread across these regions.
 
@@
 The t-stage of the tumor can be included in the model by choosing different parameterizations of the binomial distribution, considering that a tumor in late t-stages was diagnosed later than a tumor in early t-stages, therefore shifting the probability of diagnosis to later time steps.
-This finally defines the probability distribution over all states of lymph node involvement used in @eq-bayes-law.
+This finally defines the probability distribution over all states of lymph node involvement.
 
 $$
-P \left( \mathbf{X}=\boldsymbol{\xi}_i \mid \boldsymbol{\theta} \right) = \sum_{t=0}^{t_\text{max}} P_T(t) \left( \boldsymbol{\pi} \cdot \mathbf{A}^t \right)_i
+P \left( \mathbf{X}=\boldsymbol{\xi}_i \mid \boldsymbol{\theta}, \mathbf{T} \right) = \sum_{t=0}^{t_\text{max}} P_T(t) \left( \boldsymbol{\pi} \cdot \mathbf{A}^t \right)_i
 $$ {#eq-marginalized-evolution}
@@
 $$
-P \left( \mathbf{D} \mid \boldsymbol{\theta}, \boldsymbol{\pi}\right) = \prod_i \left[ \sum_{t=0}^{t_\text{max}} P_{T_i}(t) \left( \boldsymbol{\pi} \cdot \mathbf{A}^t \right)_i \right]^{N_{i}}
+P \left( \mathbf{D} \mid \boldsymbol{\theta}, \boldsymbol{\pi}\right) = \prod_i \left[P \left( \mathbf{X}=\boldsymbol{\xi}_i \mid \boldsymbol{\theta}, T_i \right) \right]^{N_{i}}
 $$ {#eq-likelihood-HMM}
@@
 
-Since primary tumors at different subsites have different patterns of lymphatic spread patterns one would need to either generalize over several subsites as introduced in [@ludwig_dynamic_2021], where the HMM has been introduced for oropharyngeal cancer, or train a distinct model for each subsite.
-The former approach leads to less precise predictions for each subsite, especially those subsites with few patients.
-Training one HMM per subsite would be rather time intensive and lead to large uncertainties for subsites with only few patients such as C04 Floor of mouth or C05 Palate.
-Further we would not exploit the anatomical similarities which are expected for anatomically nearby locations.
+Primary tumors at different subsites exhibit distinct lymphatic spread patterns.
+This presents a challenge when attempting to generalize predictive models across subsites.
+One approach, as introduced in [@ludwig_dynamic_2021], uses a Hidden Markov Model (HMM) trained specifically for oropharyngeal cancer.
+However, extending this model to other subsites would either require generalizing over several subsites or training a separate model for each.
+The former approach sacrifices precision, particularly for subsites with fewer patients, while the latter approach becomes computationally intensive and introduces large uncertainties for subsites with limited patient data, such as C04 (Floor of mouth) or C05 (Palate).
 
-Therefore we introduce a mixture model which combines all subsites in a single model exploiting anatomical proximities.
-Let us assume that we have a dataset $\mathbf{D}$ that is specified via the number of patients $N_{is}$ that were diagnosed in LNL involvement state $i$ and had a primary tutor in subsite $s$.
+To address these challenges and exploit the anatomical similarities between nearby subsites, we introduce a mixture model that combines data from all subsites into a single model.
+This model accounts for anatomical proximities, thereby improving predictive power while maintaining computational efficiency.
+
+## Mixture Model Formulation
+
+Assume we have a dataset $\mathbf{D}$ comprising the number of patients $N_{is}$ diagnosed with LNL involvement state $i$ and having a primary tutor in subsite $s$.
 Let us further assume that we want to describe this dataset using a mixture of $M$ HMMs, each with a different set of model parameters $\boldsymbol{\theta_m}$.
-As the generative model of the data, we assume that a patient with subsite $s$ is generated with probability $\pi_m^s$ from model $m$, where the mixing components $\pi_m^s$ need to satisfy $\sum_{m = 1}^{M}\pi_m^s = 1$, with $\pi_m^s \geq 0$ for all $m$ and all subsites $s$.
-The likelihood of the dataset can then be written as
+
+As the generative model of the data, we assume that a patient with subsite $s$ is generated with probability $\pi_m^s$ from model $m$,which satisfies the constraints $\sum_{m = 1}^{M}\pi_m^s = 1$ and $\pi_m^s \geq 0$ for all $m$ and all subsites $s$.
+
+The likelihood of the dataset is expressed as:
 
 $$
-P \left( \mathbf{D} \mid \boldsymbol{\theta}, \boldsymbol{\pi}\right) = \prod_s \prod_i \left[ \sum_{m=1}^M \pi_m^s P_m \left( \mathbf{X}=\boldsymbol{\xi}_i \mid \boldsymbol{\theta}_m \right) \right]^{N_{is}}
+P \left( \mathbf{D} \mid \boldsymbol{\theta}, \boldsymbol{\pi}\right) = \prod_s \prod_i \left[ \sum_{m=1}^M \pi_m^s P_m \left( \mathbf{X}=\boldsymbol{\xi}_i \mid \boldsymbol{\theta}_m, T_i \right) \right]^{N_{is}}
 $$ {#eq-mixture-distribution}
 
-The model now contains two types of parameters.
-The rpobabilities of tumor spread for the different models, $\mathbf{\theta_m}$, and the mixing coefficients $\pi_m^s$.
-The posterior distribution over the parameters $P(\boldsymbol{\theta},\boldsymbol{\pi}|\mathbf{D})$ can be calculated assuming a uniform prior in the interval [0,1] for all parameters.
-However, $P(\boldsymbol{\theta},\boldsymbol{\pi}|\mathbf{D})$ is a multimodal-distribution as one can permute the different models.
+In @eq-mixture-distribution we compute the probability of each state $\boldsymbol{\xi}_i$ under model $m$ similarly to @eq-likelihood-HMM.
+However, we weight each model with the mixing parameter $\pi_m^s$.
+Since the mixing parameters are subsite dependent, we differ between each subsite.
+
+The model now contains two sets of parameters: the tumor spread probabilities $\mathbf{\theta_m}$ for each model, and the mixing coefficients $\pi_m^s$.
+The posterior distribution over these parameters $P(\boldsymbol{\theta},\boldsymbol{\pi}|\mathbf{D})$ can be estimated assuming a uniform prior on the interval [0,1] for all parameters.
+However, $P(\boldsymbol{\theta},\boldsymbol{\pi}|\mathbf{D})$ is a multimodal-distribution as one can permute the different models and achieve the same result.
 To address this problem, we apply an expectation-maximization (EM) algorithm.
-We introduce latent variables $\boldsymbol{\epsilon}$ which assigns to each observation $\mathbf{B}$ in our dataset the probability to originate from each model $m$.
-The log-likelihood can be expressed as:
+
+## Expectation-Maximization (EM) Algorithm
+To solve for the model parameters, we apply the EM algorithm, which iterates between two steps: the **Expectation (E) step** and **the Maximization (M) step**.
+
+#### Expectation Step
+
+In the E-step, we compute the expected value of the latent variables $\boldsymbol{\epsilon}$, which represent the probability that each observation $\mathbf{X} = \boldsymbol{\xi}_i$ originates from model $m$.
+Given the current estimates of $\boldsymbol{\theta}$ and $\boldsymbol{\pi}$, the expectation value $\gamma(\epsilon_m^i)$ is calculated as:
 
 $$
-\ln p(\mathbf{X}|\boldsymbol{\theta},\boldsymbol{\pi}) = \left\{\sum_{\boldsymbol{\epsilon}} p(\mathbf{X},\boldsymbol{\epsilon}|\boldsymbol{\theta}, \boldsymbol{\pi})\right\}
-$$ {#eq-mixture-llh}
+\mathbb{E}[\epsilon_m] = \gamma(\epsilon_m^i) = \frac{\pi_m^s P(\mathbf{X} = \boldsymbol{\xi}_i \mid \boldsymbol{\theta_m}, T_i)}{\sum_{j=1}^M \pi_j^s P(\mathbf{X} = \boldsymbol{\xi}_i \mid \boldsymbol{\theta_j}, T_i)}
+$${#eq-expectation-step}
 
- The EM algorithm evaluates the model parameters $\boldsymbol{\theta}$ and $\boldsymbol{\pi}$ in two steps.
-In the expectation step the latent variables are evaluated.
-To do so, we fix the current model parameters $\boldsymbol{\theta}$ and $\boldsymbol{\pi}$ and compute the expectation value of the latent variables.
-We can compute the expectation values $\boldsymbol{\gamma}$ of the latent variables for each observed involvement pattern $\mathbf{X}_i$ in subsite $s$ separatly.
+Here, the mixing parameter $\pi_m^s$ depends on the subsite of diagnosis, $Z_i$, as denoted by the superscript $s$.
+
+#### Maximization Step
+
+In the M-step, the model parameters are updated by maximizing the following likelihood, based on the expectations computed in the E-step:
 
 $$
-\mathbb{E}[\epsilon_m] = \gamma(\epsilon_m^i) = \frac{\pi_m^s p(\mathbf{X}_i|\theta_m)}{\sum_{j=1}^M \pi_j^sp(\mathbf{X}_i|\theta_j)} 
- $$ {#eq-expectation-step}
+\ln P(\mathbf{D}, \boldsymbol{\epsilon} \mid \boldsymbol{\theta}, \boldsymbol{\pi}) = \sum_{i=1}^{N} \sum_{m=1}^M \gamma(\epsilon_m^i) \left( \ln \pi_m^s + \ln P(\mathbf{X} = \boldsymbol{\xi}_i \mid \boldsymbol{\theta_m}, T_i) \right)
+$${#eq-maximization-step}
 
-The @eq-expectation-step includes a superscript $s$ for the mixture parameters $\pi$ which depends on the tumor subsite of diagnosis $Z_i$. 
-After computing the expectation value for all components $M$ and patients $N$, the model parameters are updated by in the maximization step by maximizing the following likelihood:
-$$
- \ln p(\mathbf{Z},\boldsymbol{\epsilon}|\boldsymbol{\theta},\boldsymbol{\pi}) = \sum_{i=1}^{N}\sum_{m=1}^M \gamma(\epsilon_m^i)\{\ln \pi_m^s + \ln p(Z_i|\theta_m)\}
- $$ {#eq-maximization-step}
-
-In @eq-maximization-step one can see, that we can evaluate the mixture parameters $\boldsymbol{\pi}$ and model parameter $\boldsymbol{\theta}$ idependently of each other.
-The mixture parameters can be calculated analytically:
+In this step, the mixture parameters $\pi_m^s$ and the model parameters $\boldsymbol{\theta}_m$ can be optimized separately. The mixture parameters $\pi_m^s$ can be updated analytically:
 
 $$
-\pi_m^s = \frac{1}{N^s}\sum_i^{N^s}\gamma(\epsilon_m^i)
-$$ {#eq-mixture-maximization}
+\pi_m^s = \frac{1}{N^s} \sum_i^{N^s} \gamma(\epsilon_m^i)
+$${#eq-mixture-maximization}
 
-The model parameters $\boldsymbol{\theta}$ can be optimized component-wise with an optimization algorithm, since an analytic solution is not feasible:
+On the other hand, the model parameters $\boldsymbol{\theta_m}$ are optimized component-wise using an optimization algorithm, as an analytic solution is not feasible.
+The updated log-likelihood for each model is:
 
 $$
-\ln p(\mathbf{Z},\boldsymbol{\epsilon}|\boldsymbol{\theta_m}) = \sum_{i=1}^{N} \gamma(\epsilon_m^i)\{\ln p(Z_i|\theta_m)\}
+\ln P(\mathbf{D}, \boldsymbol{\epsilon} \mid \boldsymbol{\theta}_m) = \sum_{i=1}^{N} \gamma(\epsilon_m^i) \ln P(\mathbf{X} = \boldsymbol{\xi}_i \mid \theta_m, T_i)
 $$
 
-By repeating these steps the algorithm is guaranteed to converge to a (local) maximum.
+By iterating these steps, the EM algorithm converges to a (local) maximum of the likelihood.
+
+$$
+\ln p(\mathbf{D}|\boldsymbol{\theta},\boldsymbol{\pi}) = \left\{\sum_{\boldsymbol{\epsilon}} p(\mathbf{D},\boldsymbol{\epsilon}|\boldsymbol{\theta}, \boldsymbol{\pi})\right\}
+$$ {#eq-mixture-llh}
 
@@
 Simlarly to the three component model, we can analyze the convergence over the iterations of the EM-algorithm.
-In @fig-fig-convergence4 we can see that in this more complex model, the likelihood space becomes more complex as at around 200 iterations, the negative log-likelihood starts to increase faster again.
+In @fig-convergence4 we can see that in this more complex model, the likelihood space becomes more complex as at around 200 iterations, the negative log-likelihood starts to increase faster again.
 
@@
 The y-axis on the right shows the sum of absulte difference between all model parameters.](figures/convergence_4_components_4_loc.png){#fig-convergence4}
+
+The component assignment is shown in @fig-4-matrix. Similarly to the 3-component model the different tumor locations are assigned to a one of the components....
+
+Here i probably should permute the components such that we have the same ordering as in the 3-component model.
+
+![Matrix representation of component assignment.
+Each row of the matrix corresponds to each ICD code.
+The collums represent the three different components](figures/mixture_components_4_t_staging_add_larynx.png){#fig-4-matrix}
+
+--> Add predictions and compare to the prevalence. Additionally we could compare to tumor location predictions.

### 2024-10-01 _tex/manuscript.tex
Built site for gh-pages

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  sn-mathphys-num,
]{sn-jnl}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

%%%% Standard Packages

\usepackage{graphicx}%
\usepackage{multirow}%
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{amsthm}%
\usepackage{mathrsfs}%
\usepackage[title]{appendix}%
\usepackage{xcolor}%
\usepackage{textcomp}%
\usepackage{manyfoot}%
\usepackage{booktabs}%
\usepackage{algorithm}%
\usepackage{algorithmicx}%
\usepackage{algpseudocode}%
\usepackage{listings}%

%%%%

\raggedbottom
\usepackage{multirow}
\usepackage{centernot}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={A mixture of hidden Markov models to predict the lymphatic spread in head and neck cancer depending on primary tumor location},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title[A mixture of hidden Markov models to predict the lymphatic spread
in head and neck cancer depending on primary tumor location]{A mixture
of hidden Markov models to predict the lymphatic spread in head and neck
cancer depending on primary tumor location}

% author setup
\author[1,2]{\fnm{Yoel Perez} \sur{Haas}}\email{yoel.perezhaas@usz.ch}\author*[1,2]{\fnm{Roman} \sur{Ludwig}}\email{roman.ludwig@usz.ch}\author[1,2]{\fnm{Julian} \sur{Brönnimann}}\author[1,2]{\fnm{Esmée Lauren} \sur{Looman}}\author[2]{\fnm{Panagiotis} \sur{Balermpas}}\author[11]{\fnm{Sergi} \sur{Benavente}}\author[3,4,7]{\fnm{Adrian} \sur{Schubert}}\author[8]{\fnm{Dorothea} \sur{Barbatei}}\author[8]{\fnm{Laurence} \sur{Bauwens}}\author[2]{\fnm{Jean-Marc} \sur{Hoffmann}}\author[3]{\fnm{Olgun} \sur{Elicin}}\author[6,10]{\fnm{Matthias} \sur{Dettmer}}\author[2]{\fnm{Bertrand} \sur{Pouymayou}}\author[4,5]{\fnm{Roland} \sur{Giger}}\author[8]{\fnm{Vincent} \sur{Grégoire}}\author[1,2]{\fnm{Jan} \sur{Unkelbach}}\email{jan.unkelbach@usz.ch}
% affil setup
\affil[1]{\orgdiv{Department of Physics}, \orgname{University of
Zurich}}
\affil[2]{\orgdiv{Radiation Oncology}, \orgname{University Hospital
Zurich}}
\affil[3]{\orgdiv{Department of Radiation Oncology}, \orgname{Bern
University Hospital}}
\affil[4]{\orgdiv{Department of ENT, Head \& Neck
Surgery}, \orgname{Bern University Hospital}}
\affil[5]{\orgdiv{Head and Neck Anticancer Center}, \orgname{Bern
University Hospital}}
\affil[6]{\orgdiv{Institute of Tissue Medicine and
Pathology}, \orgname{Bern University Hospital}}
\affil[7]{\orgdiv{Department of ENT, Head \& Neck
Surgery}, \orgname{Réseau Hospitalier Neuchâtelois}}
\affil[8]{\orgdiv{Department of Radiation Oncology}, \orgname{Centre
Léon Bérard}}
\affil[9]{\orgdiv{Department of Head and Neck Surgery}, \orgname{Centre
Léon Bérard}}
\affil[10]{\orgdiv{Institute of Pathology}, \orgname{Klinikum
Stuttgart}}
\affil[11]{\orgdiv{Departement of Radiation Oncology}, \orgname{Hospital
Vall d'Hebron}}

% abstract

\abstract{Purpose: to be done

Methods: to be done

Results: to be done

Conclusions: to be done}

% keywords

\begin{document}
\maketitle

\section{Introduction}\label{introduction}

Head and Neck Squamous Cell Carcinoma (HNSCC) are known to spread
through the lymphatic system often leading to metastases in the lymph
nodes \citep[@shah\_patterns\_1990]{mukherji_cervical_2001}. To minimize
nodal recurrences, lymph node levels (LNL) at risk of harboring occult
metastases are typically irradiated electively. Current guidelines for
different tumor locations are based on the overall prevalence of nodal
disease as reported in literature
\citep[@mukherji\_cervical\_2001, @shah\_patterns\_1990]{biau_selection_2019}.

To personalize the prediction of the risk of occult metastases, given a
patient's individual diagnosis, we previously published a large,
multi-centric dataset where the lymphatic involvement per LNL is
available for each
patient\citep[@ludwig\_multi-centric\_2023]{ludwig_dataset_2022}.Building
on this dataset, we introduced an interpretable hidden Markov model
(HMM), trained to predict the risk for occult nodal disease, given an
individual patient's diagnosis \citep{ludwig_hidden_2021}.

Personalized risk predictions could enable clinicians to safely reduce
the elective clinical target volume (CTV-N), potentially decreasing
treatment-related side effects that impair a patient's quality of life,
without compromising the efficacy of the treatment
\citep{batth_practical_2014}.

Initially, separate models were trained for distinct tumor locations,
such as the oropharynx and oral cavity. These tumor locations are also
used in guidelines to define the elective target volumes
\citep{biau_selection_2019}. However, this approach did not account for
variations in lymphatic spread between subsites within these tumor
regions. With data for more than 2700 patients available, we can now
further analyze subsite specific spread patterns. Closer analysis showed
that pooling subsites into a single model led to inaccurate predictions,
as it failed to capture distinct lymphatic spread patterns. To resolve
this, we propose using a mixture of HMMs, which allows us to model the
lymphatic spread more accurately for tumors located near anatomical
borders, such as those between the oropharynx and oral cavity (e.g.,
tumors in the palate).

Additionally, we extend the analysis to a broader mixture model that
encompasses tumors of the oral cavity, oropharynx, hypopharynx, and
larynx, offering a more comprehensive understanding of lymphatic spread
across these regions.

\section{Data on Lymphatic Progression Patterns}\label{sec-data}

For the analyses in this work, we used seven datasets from 5
institutions resulting in 2741 patients in total.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  0 oropharyngeal patients from the University of Zurich in Switzerland
\item
  0 oropharyngeal patients from the Centre Léon Bérard in France
\item
  0 oropharyngeal, larynx and oral cavity patients from the Inselspital
  Bern in Switzerland
\item
  0 oropharyngeal, larynx and oral cavity patients from the Centre Léon
  Bérard in France
\item
  0 oropharyngeal, larynx and oral cavity patients from the University
  of Zurich in Switzerland
\item
  164 oropharyngeal patients from the Hospital Vall d'Hebron in Spain
  (not yet public)
\item
  979 hypopharynx, larynx and oral cavity patients from University
  Medical Center Groningen (not yet public)
\end{enumerate}

The datasets 1-4 are publicly available as CSV tables
(\citet{ludwig_multi-centric_2023} \citet{ludwig_detailed_2022}) and can
be interactively explored on \href{https://lyprox.org}{LyProX}. For each
patient the primary tumor subsite is reportd and each indicidual LNL is
reported as either metastatic or healthy given the available diagnostic
modalities, which include pathology after neck dissection in some
patients. In this work we will stratify the tumor locations into
different ICD codes which are depicted in figure~\ref{fig-subsites}.

\begin{figure}

\centering{

\includegraphics{figures/Subsites.png}

}

\caption{\label{fig-subsites}Anatomical sketch of the tumor subsites and
their corresponding ICD-10 codes. Subsite C06 ``other parts of mouth''
has not been included. Further the The tumor locations are color coded
in the following pattern: blue-oral cavity, green-oropharynx,
red-hypopharynx, orange-larynx.}

\end{figure}%

The prevelance of involvement in LNLs I, II, III, IV and V is shown in
figure~\ref{fig-involvement}. The involvement is stratified per tumor
subsite and t-stage. The figure illustrates the variations in LNL
involvement between subsites within oral cavity (blue), oropharynx
(green), hypopharynx (red) and larynx (orange). The involvement pattern
presents a continuous change over the tumor subsites. Where tumors in
the oral cavity show the most prominent LNL I involvement. As the tumor
location moves towards the oropharynx LNL II involvement increases.
Moving the tumor location further in caudal direction towards the
hypopharynx increases LNL III involvement while LNL I and II involvement
decrease. Larnygeal tumors show the least LNL I involvement.

\begin{figure}

\centering{

\includegraphics{figures/involvement_I_to_V_all_sites_t_staging.png}

}

\caption{\label{fig-involvement}Prevalence of ipsilateral LNL
involvement stratified by subsite. The subsites are sorted in natural
order to represent the continuously changing LNL involvement. The
different tumo locations are color coded, where oral cavity subsistes
are depicted in blue, larynx in green, hypopharynx in red and larynx in
orange. The patient data is further stratified in early t-stage (0-2)
and late t-stage (3-4). The legend furter specifies the number of
patients in each subsite.}

\end{figure}%

\section{Unilateral Model for Lymphatic
Progression}\label{sec-unilateral}

In this chapter we will briefly summarize unilateral model for
ipsilateral lymph node involvement introduced in
\citet{ludwig_hidden_2021}, presenting the notation which is then needed
to extend the HMM to a mixture model encompassing multiple tumor
subsites.

The HMM describes each LNL \(v \in 1, 2, ..., V\) by a binary random
variable corresponding to the status of the LNL; healthy (0) or involved
(1). The entire state of a patient with \(V\) LNLs is defined by the
\(v\)-dimensional vector \(\mathbf{X} = [X_1, X_2, ...,X_V]\). In the
HMM, a patient's involvement is modeled over time \(t\). Thus, a
patient's state of lymph node involvement \(\mathbf{X}[t]\) evolves over
discrete time steps \(t\). Let us enumerate all \(2^V\) possible states,
representing all combinations of LNL involvement. In this paper, we
consider ipsilateral LNLs I, II, III, IV and V, which amounts to 32
possible states. The HMM is then specified by a transition matrix
\(\mathbf{A}\):

\begin{equation}\phantomsection\label{eq-transition-matrix}{
\mathbf{A} = \begin{pmatrix} A_{ij} \end{pmatrix} = P \left( \mathbf{X}[t+1] = \boldsymbol{\xi}_j \mid \mathbf{X}[t] = \boldsymbol{\xi}_i \right)
}\end{equation}

whose elements \(A_{ij}\) contain the conditional probabilities that a
state \(\mathbf{X}[t]=\boldsymbol{\xi}_i\) transitions to
\(\mathbf{X}[t+1]=\boldsymbol{\xi}_j\) over one time step. The
transition matrix is specified and parameterised via the graphical model
shown in figure~\ref{fig-graph}. The red arcs in the graph of
figure~\ref{fig-graph} are associated with the probability that the
primary tumor spreads directly to a LNL (parameters \(b_v\)). The blue
arcs describe the spread from an upstream LNL -- given it is already
metastatic -- to a downstream level (parameters
\(t_{v \rightarrow v+1}\)).

\begin{figure}

\centering{

\includegraphics[width=0.2\textwidth,height=\textheight]{figures/Model_5_LNL.png}

}

\caption{\label{fig-graph}Parametrized graphical model of the lymphatic
network considering four LNLs. Blue nodes represent the hidden states of
LNLs \(X_v\), while the red one is the tumor. Arcs represent possible
routes of metastatic spread, associated with a probability.}

\end{figure}%

Now, let \(\boldsymbol{\pi}\) be the \emph{starting distribution}

\begin{equation}\phantomsection\label{eq-starting-distribution}{
\boldsymbol{\pi} = \begin{pmatrix} \pi_i \end{pmatrix} = P \left( \mathbf{X}[0] = \boldsymbol{\xi}_i \right)
}\end{equation}

denoting the probability to start in state \(\boldsymbol{\xi}_i\) at
time step 0. Assuming that every patient started with all LNLs being
healthy, we set \(\pi_i\) to zero for all states except the completly
healthy state
\(\boldsymbol{\xi} = \begin{pmatrix} 0, 0, 0, 0, 0 \end{pmatrix}\),
which has probability one.

Using the quantities introduced so far, the probability
\(P \left( \mathbf{X}[t]=\boldsymbol{\xi}_i \right)\) to be in state
\(\boldsymbol{\xi}_i\) in time step \(t\) can now be conveniently
expressed as a matrix product:

\begin{equation}\phantomsection\label{eq-evolution}{
P \left( \mathbf{X}[t]=\boldsymbol{\xi}_i \right) = \left( \boldsymbol{\pi} \cdot \mathbf{A}^t \right)_i
}\end{equation}

This evolution implicitly marginalizes over all possible paths to arrive
at state \(\boldsymbol{\xi}_i\) after \(t\) time-steps. Additionally, we
must marginalize over the unknown time of diagnosis using a time-prior
\(P_T(t)\) which is defined by a binomial distribution. The t-stage of
the tumor can be included in the model by choosing different
parameterizations of the binomial distribution, considering that a tumor
in late t-stages was diagnosed later than a tumor in early t-stages,
therefore shifting the probability of diagnosis to later time steps.
This finally defines the probability distribution over all states of
lymph node involvement.

\begin{equation}\phantomsection\label{eq-marginalized-evolution}{
P \left( \mathbf{X}=\boldsymbol{\xi}_i \mid \boldsymbol{\theta}, \mathbf{T} \right) = \sum_{t=0}^{t_\text{max}} P_T(t) \left( \boldsymbol{\pi} \cdot \mathbf{A}^t \right)_i
}\end{equation}

where \(\boldsymbol{\theta}=\{ b_v, t_{v \rightarrow v+1} \}\) denotes
the set of all model parameters (7 in our case). Fortunately, the exact
length and shape of this distribution has little impact as previously
shown \citep{ludwig_hidden_2021}. We set \(t_\text{max}=\) 10 and
\(P_\text{early}(t)\) to a binomial distribution with parameter 0.3.
Further details on the HMM can be found in \citet{ludwig_hidden_2021}
and \citet{zora231470}.

With equation~\ref{eq-marginalized-evolution} we can compute the
probability of a patient being in any state \(\xi_i\). Therefore the
likelihood of observing a the For model training we assume that the
diagnoses in our data \(\mathbf{D}\) we observe correspond to the hidden
state \(\mathbf{X}\) of the patient. Thus, learning the model parameters
corresponds to maximizing the probability of observing the dataset
\(\mathbf{D}\):

\begin{equation}\phantomsection\label{eq-likelihood-HMM}{
P \left( \mathbf{D} \mid \boldsymbol{\theta}, \boldsymbol{\pi}\right) = \prod_i \left[P \left( \mathbf{X}=\boldsymbol{\xi}_i \mid \boldsymbol{\theta}, T_i \right) \right]^{N_{i}}
}\end{equation}

In equation~\ref{eq-likelihood-HMM} we take the likelihood of observing
a patient in t-stage \(T_i\) and with involvement \(\xi_i\) to the power
of \(N_i\), corresponding to the number of observations of the specific
diagnosis \(i\) in the dataset \(\mathbf{D}\). By multiplying over all
diagnoses \(i\) we get the likelihood of observing dataset
\(\mathbf{D}\) given the previously intoduced model with parameters
\(\boldsymbol{\theta}\)

\section{Mixture Model for Lymphatic Spread}\label{sec-mixture}

Primary tumors at different subsites exhibit distinct lymphatic spread
patterns. This presents a challenge when attempting to generalize
predictive models across subsites. One approach, as introduced in
\citep{ludwig_dynamic_2021}, uses a Hidden Markov Model (HMM) trained
specifically for oropharyngeal cancer. However, extending this model to
other subsites would either require generalizing over several subsites
or training a separate model for each. The former approach sacrifices
precision, particularly for subsites with fewer patients, while the
latter approach becomes computationally intensive and introduces large
uncertainties for subsites with limited patient data, such as C04 (Floor
of mouth) or C05 (Palate).

To address these challenges and exploit the anatomical similarities
between nearby subsites, we introduce a mixture model that combines data
from all subsites into a single model. This model accounts for
anatomical proximities, thereby improving predictive power while
maintaining computational efficiency.

\subsection{Mixture Model Formulation}\label{mixture-model-formulation}

Assume we have a dataset \(\mathbf{D}\) comprising the number of
patients \(N_{is}\) diagnosed with LNL involvement state \(i\) and
having a primary tutor in subsite \(s\). Let us further assume that we
want to describe this dataset using a mixture of \(M\) HMMs, each with a
different set of model parameters \(\boldsymbol{\theta_m}\).

As the generative model of the data, we assume that a patient with
subsite \(s\) is generated with probability \(\pi_m^s\) from model
\(m\),which satisfies the constraints \(\sum_{m = 1}^{M}\pi_m^s = 1\)
and \(\pi_m^s \geq 0\) for all \(m\) and all subsites \(s\).

The likelihood of the dataset is expressed as:

\begin{equation}\phantomsection\label{eq-mixture-distribution}{
P \left( \mathbf{D} \mid \boldsymbol{\theta}, \boldsymbol{\pi}\right) = \prod_s \prod_i \left[ \sum_{m=1}^M \pi_m^s P_m \left( \mathbf{X}=\boldsymbol{\xi}_i \mid \boldsymbol{\theta}_m, T_i \right) \right]^{N_{is}}
}\end{equation}

In equation~\ref{eq-mixture-distribution} we compute the probability of
each state \(\boldsymbol{\xi}_i\) under model \(m\) similarly to
equation~\ref{eq-likelihood-HMM}. However, we weight each model with the
mixing parameter \(\pi_m^s\). Since the mixing parameters are subsite
dependent, we differ between each subsite.

The model now contains two sets of parameters: the tumor spread
probabilities \(\mathbf{\theta_m}\) for each model, and the mixing
coefficients \(\pi_m^s\). The posterior distribution over these
parameters \(P(\boldsymbol{\theta},\boldsymbol{\pi}|\mathbf{D})\) can be
estimated assuming a uniform prior on the interval {[}0,1{]} for all
parameters. However,
\(P(\boldsymbol{\theta},\boldsymbol{\pi}|\mathbf{D})\) is a
multimodal-distribution as one can permute the different models and
achieve the same result. To address this problem, we apply an
expectation-maximization (EM) algorithm.

\subsection{Expectation-Maximization (EM)
Algorithm}\label{expectation-maximization-em-algorithm}

To solve for the model parameters, we apply the EM algorithm, which
iterates between two steps: the \textbf{Expectation (E) step} and
\textbf{the Maximization (M) step}.

\paragraph{Expectation Step}\label{expectation-step}

In the E-step, we compute the expected value of the latent variables
\(\boldsymbol{\epsilon}\), which represent the probability that each
observation \(\mathbf{X} = \boldsymbol{\xi}_i\) originates from model
\(m\). Given the current estimates of \(\boldsymbol{\theta}\) and
\(\boldsymbol{\pi}\), the expectation value \(\gamma(\epsilon_m^i)\) is
calculated as:

\begin{equation}\phantomsection\label{eq-expectation-step}{
\mathbb{E}[\epsilon_m] = \gamma(\epsilon_m^i) = \frac{\pi_m^s P(\mathbf{X} = \boldsymbol{\xi}_i \mid \boldsymbol{\theta_m}, T_i)}{\sum_{j=1}^M \pi_j^s P(\mathbf{X} = \boldsymbol{\xi}_i \mid \boldsymbol{\theta_j}, T_i)}
}\end{equation}

Here, the mixing parameter \(\pi_m^s\) depends on the subsite of
diagnosis, \(Z_i\), as denoted by the superscript \(s\).

\paragraph{Maximization Step}\label{maximization-step}

In the M-step, the model parameters are updated by maximizing the
following likelihood, based on the expectations computed in the E-step:

\begin{equation}\phantomsection\label{eq-maximization-step}{
\ln P(\mathbf{D}, \boldsymbol{\epsilon} \mid \boldsymbol{\theta}, \boldsymbol{\pi}) = \sum_{i=1}^{N} \sum_{m=1}^M \gamma(\epsilon_m^i) \left( \ln \pi_m^s + \ln P(\mathbf{X} = \boldsymbol{\xi}_i \mid \boldsymbol{\theta_m}, T_i) \right)
}\end{equation}

In this step, the mixture parameters \(\pi_m^s\) and the model
parameters \(\boldsymbol{\theta}_m\) can be optimized separately. The
mixture parameters \(\pi_m^s\) can be updated analytically:

\begin{equation}\phantomsection\label{eq-mixture-maximization}{
\pi_m^s = \frac{1}{N^s} \sum_i^{N^s} \gamma(\epsilon_m^i)
}\end{equation}

On the other hand, the model parameters \(\boldsymbol{\theta_m}\) are
optimized component-wise using an optimization algorithm, as an analytic
solution is not feasible. The updated log-likelihood for each model is:

\[
\ln P(\mathbf{D}, \boldsymbol{\epsilon} \mid \boldsymbol{\theta}_m) = \sum_{i=1}^{N} \gamma(\epsilon_m^i) \ln P(\mathbf{X} = \boldsymbol{\xi}_i \mid \theta_m, T_i)
\]

By iterating these steps, the EM algorithm converges to a (local)
maximum of the likelihood.

\begin{equation}\phantomsection\label{eq-mixture-llh}{
\ln p(\mathbf{D}|\boldsymbol{\theta},\boldsymbol{\pi}) = \left\{\sum_{\boldsymbol{\epsilon}} p(\mathbf{D},\boldsymbol{\epsilon}|\boldsymbol{\theta}, \boldsymbol{\pi})\right\}
}\end{equation}

\section{Three component Mixture Model}\label{sec-3comp}

We illustrate the methodology for a mixture model with M = 3 components,
considering the ipsilateral involvement of LNLs I, II, III, IV, and V.
We include the ICD codes as subsites for oral cavity, hypopharynx and
oropharynx. In figure~\ref{fig-convergence} the convergence of the
negative log-likelihood and change in model parameters is depicted.
After a random inizialization, the algorithm rapidly converges. The
algorithm was stopped when the difference of log-likelihood between two
iterations was below 0.01.

\begin{figure}

\centering{

\includegraphics{figures/Convergence_3_comp.png}

}

\caption{\label{fig-convergence}The y-axis on the left shows the
negative likelihood convergence depicted in the blue line. The y-axis on
the right shows the sum of absulte difference between all model
parameters showing that the parameter values stabilize rapidly as well.}

\end{figure}%

In figure~\ref{fig-3_simplex}, we visualize the resulting mixture
coefficients \(\boldsymbol{\pi}\) using a spatial representation, where
the vertices of the triangle correspond to the three components. In
figure~\ref{fig-3-matrix}, these mixture coefficients are presented in
matrix form, with the y-axis representing the ICD codes, showing how the
mixture components in each row add up to 1.

The spatial plot in figure~\ref{fig-3_simplex} illustrates how the model
assigns the three components to different tumor subsites. Component 0,
located at the bottom right of the triangle, primarily characterizes
oropharyngeal subsites. For instance, the base of tongue subsite (C01),
which exhibits the highest involvement of LNL II, is fully assigned to
this component. Similarly, subsite C10, which includes several
oropharyngeal regions, is assigned roughly 50\% to the oropharynx-like
component, with the remaining mixture distributed across the other two
components.

Hypopharyngeal subsites, on the other hand, are fully assigned to
Component 2, located at the top vertex of the triangle. Meanwhile, the
gum subsite (C03), with predominant LNL I involvement, is entirely
assigned to Component 1, situated at the bottom left. As subsites
anatomically approach the oropharynx, their mixture coefficients for the
oropharynx-like component increase. This is evident in the subsites C02
(tongue) and C05 (palate), which display a higher proportion of
oropharyngeal influence in their mixture. These results conform well
with the involvement patterns observed in the data.

\begin{figure}

\begin{minipage}{0.71\linewidth}

\begin{figure}[H]

\centering{

\includegraphics{figures/mixture_components_3_simplex_t_stages_new.png}

}

\caption{\label{fig-3_simplex}Assignment of each subsite to each of the
three components. The closer a subsite is to a vertex, the more it is
assigned to the corresponding component, with component 0 on the bottom
right, 1 on the bottom left and 2 on the top. The size of the marker
(area) corresponds to the number of patients in each subsite.}

\end{figure}%

\end{minipage}%
%
\begin{minipage}{0.29\linewidth}

\begin{figure}[H]

\centering{

\includegraphics{figures/mixture_components_3_t_stages_new.png}

}

\caption{\label{fig-3-matrix}Matrix representation of component
assignment. Each row of the matrix corresponds to each ICD code. The
collums represent the three different components}

\end{figure}%

\end{minipage}%

\end{figure}%

\section{Four component Mixture Model}\label{sec-4comp}

We can extend the mixture model to include the larynx. The larynx
patients are more finely divided into ICD codes C32.0, C32.1 and C32.2
as there is a notable difference between these ICD codes in
figure~\ref{fig-involvement}.

Simlarly to the three component model, we can analyze the convergence
over the iterations of the EM-algorithm. In
figure~\ref{fig-convergence4} we can see that in this more complex
model, the likelihood space becomes more complex as at around 200
iterations, the negative log-likelihood starts to increase faster again.

\begin{figure}

\centering{

\includegraphics{figures/convergence_4_components_4_loc.png}

}

\caption{\label{fig-convergence4}The y-axis on the left shows the
negative likelihood convergence depicted in the blue line. The y-axis on
the right shows the sum of absulte difference between all model
parameters.}

\end{figure}%

The component assignment is shown in figure~\ref{fig-4-matrix}.
Similarly to the 3-component model the different tumor locations are
assigned to a one of the components\ldots.

Here i probably should permute the components such that we have the same
ordering as in the 3-component model.

\begin{figure}

\centering{

\includegraphics{figures/mixture_components_4_t_staging_add_larynx.png}

}

\caption{\label{fig-4-matrix}Matrix representation of component
assignment. Each row of the matrix corresponds to each ICD code. The
collums represent the three different components}

\end{figure}%

--\textgreater{} Add predictions and compare to the prevalence.
Additionally we could compare to tumor location predictions.

  \bibliography{references.bib}

\end{document}

### 2024-10-03 manuscript.qmd
big change to the math

@@
 $$
-P \left( \mathbf{D} \mid \boldsymbol{\theta}, \boldsymbol{\pi}\right) = \prod_i \left[P \left( \mathbf{X}=\boldsymbol{\xi}_i \mid \boldsymbol{\theta}, T_i \right) \right]^{N_{i}}
+P \left( \mathbf{D} \mid \boldsymbol{\theta}, \boldsymbol{\pi}\right) = \prod_i \prod_j \left[P \left( \mathbf{X}=\boldsymbol{\xi}_i \mid \boldsymbol{\theta}, T_j \right) \right]^{N_{ij}}
 $$ {#eq-likelihood-HMM}
 
-In @eq-likelihood-HMM we take the likelihood of observing a patient in t-stage $T_i$ and with involvement $\xi_i$ to the power of $N_i$, corresponding to the number of observations of the specific diagnosis $i$ in the dataset $\mathbf{D}$.
-By multiplying over all diagnoses $i$ we get the likelihood of observing dataset $\mathbf{D}$ given the previously intoduced model with parameters $\boldsymbol{\theta}$
+In @eq-likelihood-HMM we take the likelihood of observing a patient in t-stage $T_j$ and with involvement $\xi_i$ to the power of $N_{ij}$, corresponding to the number of observations of the patients in state $i$ and t-stage $T_j$ (either early or late) in the dataset $\mathbf{D}$.
 
@@
 
-Assume we have a dataset $\mathbf{D}$ comprising the number of patients $N_{is}$ diagnosed with LNL involvement state $i$ and having a primary tutor in subsite $s$.
-Let us further assume that we want to describe this dataset using a mixture of $M$ HMMs, each with a different set of model parameters $\boldsymbol{\theta_m}$.
+The mixture model assumes that the data is generated by a set of $M$ different underlying models.
+In our case, we introduce $M$ distinct lymphatic spread models.
+Each patient $k$ is assumed to be generated by one specific model from this set of $M$ models.
 
-As the generative model of the data, we assume that a patient with subsite $s$ is generated with probability $\pi_m^s$ from model $m$,which satisfies the constraints $\sum_{m = 1}^{M}\pi_m^s = 1$ and $\pi_m^s \geq 0$ for all $m$ and all subsites $s$.
+To formalize this, each patient $k$ can be assigned to a specific model via a binary latent vector $\boldsymbol{\epsilon}_k$.
+The vector $\boldsymbol{\epsilon}_k$ has length $M$, with exactly one element set to 1, indicating which model generated the patient’s data, and all other elements set to 0.
+Thus, for patient $k$, the latent variable $\boldsymbol{\epsilon}_k$ can be interpreted as a categorical indicator variable that encodes the assignment to one of the $M$ lymphatic spread models.
 
-The likelihood of the dataset is expressed as:
+Next, we introduce subsite-specific mixing parameters $\boldsymbol{\pi}^s = \{\pi_1^s, \pi_2^s, \dots, \pi_M^s\}$, where each $\pi_m^s$ represents the prior probability that a patient with tumors in subsite $s$ is generated by model $m$.
+These mixing parameters must satisfy the condition:
 
 $$
-P \left( \mathbf{D} \mid \boldsymbol{\theta}, \boldsymbol{\pi}\right) = \prod_s \prod_i \left[ \sum_{m=1}^M \pi_m^s P_m \left( \mathbf{X}=\boldsymbol{\xi}_i \mid \boldsymbol{\theta}_m, T_i \right) \right]^{N_{is}}
-$$ {#eq-mixture-distribution}
+\sum_{m=1}^M \pi_m^s = 1, \quad \forall s
+$$
 
-In @eq-mixture-distribution we compute the probability of each state $\boldsymbol{\xi}_i$ under model $m$ similarly to @eq-likelihood-HMM.
+The mixing parameters $\boldsymbol{\pi}^s$ determine the proportion of patients with tumors in subsite $s$ that are expected to be generated by each model.
+Specifically, $\pi_m^s$ is the probability that a randomly chosen patient with a tumor in subsite $s$ is generated by model $m$.
+
+The latent variables $\boldsymbol{\epsilon}_{k}$, on the other hand, are patient-specific and represent the assignment of each patient to a particular model.
+We will later generalize patients with the same diagnosis to the latent variable $\boldsymbol{\epsilon}_{i,j}$.
+The subscripts $i$ and $j$ represent that the latent variables are only dependent on the diagnosis of the patient, which consists of lymphatic involvement $\boldsymbol{\xi}_i$ and t-stage $T_j$.
+The latent variables are central to the model, as they are hidden (unobserved) and must be inferred.
+
+The joint probability of the observed data $\mathbf{D}$ (i.e., patient data) and the latent variables $\boldsymbol{\epsilon}$ is given by:
+
+$$
+P(\mathbf{D}, \boldsymbol{\epsilon} | \boldsymbol{\theta}, \boldsymbol{\pi}) = \prod_{k=1}^{K} \prod_{m=1}^{M} \left[ \pi_m^s P(\mathbf{D}_k | \boldsymbol{\theta}_m, \boldsymbol{\pi}) \right]^{\epsilon_{k}^m}
+$${#eq-complete-likelihood}
+
+Here:
+
+* $\pi_m^s$ is the mixing parameter for subsite $s$ and model $m$,
+* $P(\mathbf{D}_k | \boldsymbol{\theta}_m)$ is the likelihood of patient $k$'s data given that it was generated by model $m$,
+* $\boldsymbol{\theta}_m$ represents the parameters of model $m$,
+* $\epsilon_{k}^m$ is the latent variable that indicates whether patient $k$ was generated by model $m$.
+
+**Incomplete data likelihood which can probably be skipped**
+
+Since the latent variables $\boldsymbol{\epsilon}_k$ are unobserved, the incomplete complete data likelihood marginalizes over all possible latent assignments to reflect the uncertainty about which model generated each patient’s data:
+
+$$
+P(\mathbf{D} | \boldsymbol{\theta}, \boldsymbol{\pi}) = \sum_{\boldsymbol{\epsilon}} P(\mathbf{D}, \boldsymbol{\epsilon} | \boldsymbol{\theta}, \boldsymbol{\pi})
+$$
+
+This allows us to integrate out the latent variables $\boldsymbol{\epsilon}$, as their true values are unknown.
+
+**Continues from here**
+
+The latent variables $\boldsymbol{\epsilon}_k$ are unknown and must be inferred from the observed data.
+The mixture model assumes that each patient is generated by exactly one of the $M$ models.
+However, in our application, this assumption does not reflect continuous or overlapping behaviors between subsites.
+Instead, we infer a probabilistic distribution over the latent variables, given the observed data and model parameters.
+The posterior distribution:
+
+$$
+P(\boldsymbol{\epsilon}_k | \mathbf{D}_k, \boldsymbol{\theta}, \boldsymbol{\pi})
+$$
+
+describes the probability that patient $k$ was generated by each of the $M$ models, allowing for soft assignments where each patient is associated with multiple models with certain probabilities, as opposed to being strictly assigned to just one.
+Given the probability distribution for the latent variables $\boldsymbol{\epsilon}_k$, we compute the expectation value $\mathbb{E}[\boldsymbol{\epsilon}_k] = \boldsymbol{\gamma}(\boldsymbol{\epsilon}_k)$ for each patient to then apply the complete data likelihood given in @eq-complete-likelihood.
+
+We can now apply the mixture model to our dataset $\mathbf{D}$ comprising the number of patients $N_{ijs}$ diagnosed with LNL involvement state $\mathbf{X} = \boldsymbol{\xi}_i$, t-stage $T_j$, with a primary tumor in subsite $s$.
+
+The likelihood of the complete dataset is then expressed as:
+
+$$
+P \left( \mathbf{D} \mid \boldsymbol{\theta}, \boldsymbol{\pi}\right) = \prod_i \prod_j \prod_s \left[ \prod_{m=1}^M \left[ \pi_m^s P_m \left( \mathbf{X}=\boldsymbol{\xi}_i \mid \boldsymbol{\theta}_m, T_j \right) \right]^{\gamma(\epsilon_{i,j}^m)} \right]^{N_{ijs}}
+$$ {#eq-complete-likelihood-explicite}
+
+In @eq-complete-likelihood we compute the likelihood of observing state $\boldsymbol{\xi}_i$ with t-stage $T_j$ under model $m$ similarly to @eq-likelihood-HMM.
 However, we weight each model with the mixing parameter $\pi_m^s$.
@@
 
-In the E-step, we compute the expected value of the latent variables $\boldsymbol{\epsilon}$, which represent the probability that each observation $\mathbf{X} = \boldsymbol{\xi}_i$ originates from model $m$.
-Given the current estimates of $\boldsymbol{\theta}$ and $\boldsymbol{\pi}$, the expectation value $\gamma(\epsilon_m^i)$ is calculated as:
+In the E-step, we compute the expected value of the latent variables $\boldsymbol{\epsilon}$, which represent the probabilities of a patient $k$ originating from one of the models $m$.
+Given the current estimates of $\boldsymbol{\theta}$ and $\boldsymbol{\pi}$, the expectation value $\gamma(\epsilon_m^k)$ is calculated as:
 
 $$
-\mathbb{E}[\epsilon_m] = \gamma(\epsilon_m^i) = \frac{\pi_m^s P(\mathbf{X} = \boldsymbol{\xi}_i \mid \boldsymbol{\theta_m}, T_i)}{\sum_{j=1}^M \pi_j^s P(\mathbf{X} = \boldsymbol{\xi}_i \mid \boldsymbol{\theta_j}, T_i)}
+\mathbb{E}[\epsilon^m] = \gamma(\epsilon^m_{i,j}) = \frac{\pi_m^s P(\mathbf{X} = \boldsymbol{\xi}_i \mid \boldsymbol{\theta_m}, T_j)}{\sum_{n=1}^M \pi_n^s P(\mathbf{X} = \boldsymbol{\xi}_i \mid \boldsymbol{\theta_n}, T_j)}
 $${#eq-expectation-step}
 
-Here, the mixing parameter $\pi_m^s$ depends on the subsite of diagnosis, $Z_i$, as denoted by the superscript $s$.
+THE OTHER OPTION
+
+$$
+\mathbb{E}[\epsilon^m] = \gamma(\epsilon^m_k) = \frac{\pi_m^s P(\mathbf{X} = \boldsymbol{\xi}_k \mid \boldsymbol{\theta_m}, T_k)}{\sum_{j=1}^M \pi_j^s P(\mathbf{X} = \boldsymbol{\xi}_k \mid \boldsymbol{\theta_j}, T_k)}
+$${#eq-expectation-step2}
+
+Here, the mixing parameter $\pi_m^s$ depends on the subsite location of the primary tumor of patient $k$.
 
@@
 
-In the M-step, the model parameters are updated by maximizing the following likelihood, based on the expectations computed in the E-step:
+In the M-step, the model parameters are updated by maximizing the complete data likelihood in equation @eq-complete-likelihood-explicite. For better handling, we are optimizing the log likelihood:
 
 $$
-\ln P(\mathbf{D}, \boldsymbol{\epsilon} \mid \boldsymbol{\theta}, \boldsymbol{\pi}) = \sum_{i=1}^{N} \sum_{m=1}^M \gamma(\epsilon_m^i) \left( \ln \pi_m^s + \ln P(\mathbf{X} = \boldsymbol{\xi}_i \mid \boldsymbol{\theta_m}, T_i) \right)
+\ln P(\mathbf{D}, \boldsymbol{\epsilon} \mid \boldsymbol{\theta}, \boldsymbol{\pi}) = \sum_{i} \sum_j \sum_{s} N_{i,j,s} \sum_{m=1}^M \gamma(\epsilon^m_{i,j}) \left( \ln \pi_m^s + \ln P(\mathbf{X} = \boldsymbol{\xi}_i \mid \boldsymbol{\theta_m}, T_j) \right)
 $${#eq-maximization-step}
@@
 $$
-\pi_m^s = \frac{1}{N^s} \sum_i^{N^s} \gamma(\epsilon_m^i)
+\pi_m^s = \frac{1}{N^s} \sum_k^{N^s} \gamma(\epsilon^m_k)
 $${#eq-mixture-maximization}
@@
 $$
-\ln P(\mathbf{D}, \boldsymbol{\epsilon} \mid \boldsymbol{\theta}_m) = \sum_{i=1}^{N} \gamma(\epsilon_m^i) \ln P(\mathbf{X} = \boldsymbol{\xi}_i \mid \theta_m, T_i)
+\ln P(\mathbf{D}, \boldsymbol{\epsilon} \mid \boldsymbol{\theta}_m) = \sum_{i} \sum_j \gamma(\epsilon^m_{i,j}) \ln P(\mathbf{X} = \boldsymbol{\xi}_i \mid \theta_m, T_j)
 $$
 
-By iterating these steps, the EM algorithm converges to a (local) maximum of the likelihood.
-
-$$
-\ln p(\mathbf{D}|\boldsymbol{\theta},\boldsymbol{\pi}) = \left\{\sum_{\boldsymbol{\epsilon}} p(\mathbf{D},\boldsymbol{\epsilon}|\boldsymbol{\theta}, \boldsymbol{\pi})\right\}
-$$ {#eq-mixture-llh}
+By iterating these steps, the EM algorithm converges to a (local) maximum of the complete data likelihood @eq-complete-likelihood.
 

### 2024-10-03 _tex/manuscript.tex
Built site for gh-pages

@@
 \begin{equation}\phantomsection\label{eq-likelihood-HMM}{
-P \left( \mathbf{D} \mid \boldsymbol{\theta}, \boldsymbol{\pi}\right) = \prod_i \left[P \left( \mathbf{X}=\boldsymbol{\xi}_i \mid \boldsymbol{\theta}, T_i \right) \right]^{N_{i}}
+P \left( \mathbf{D} \mid \boldsymbol{\theta}, \boldsymbol{\pi}\right) = \prod_i \prod_j \left[P \left( \mathbf{X}=\boldsymbol{\xi}_i \mid \boldsymbol{\theta}, T_j \right) \right]^{N_{ij}}
 }\end{equation}
@@
 In equation~\ref{eq-likelihood-HMM} we take the likelihood of observing
-a patient in t-stage \(T_i\) and with involvement \(\xi_i\) to the power
-of \(N_i\), corresponding to the number of observations of the specific
-diagnosis \(i\) in the dataset \(\mathbf{D}\). By multiplying over all
-diagnoses \(i\) we get the likelihood of observing dataset
-\(\mathbf{D}\) given the previously intoduced model with parameters
-\(\boldsymbol{\theta}\)
+a patient in t-stage \(T_j\) and with involvement \(\xi_i\) to the power
+of \(N_{ij}\), corresponding to the number of observations of the
+patients in state \(i\) and t-stage \(T_j\) (either early or late) in
+the dataset \(\mathbf{D}\).
 
@@
 
-Assume we have a dataset \(\mathbf{D}\) comprising the number of
-patients \(N_{is}\) diagnosed with LNL involvement state \(i\) and
-having a primary tutor in subsite \(s\). Let us further assume that we
-want to describe this dataset using a mixture of \(M\) HMMs, each with a
-different set of model parameters \(\boldsymbol{\theta_m}\).
+The mixture model assumes that the data is generated by a set of \(M\)
+different underlying models. In our case, we introduce \(M\) distinct
+lymphatic spread models. Each patient \(k\) is assumed to be generated
+by one specific model from this set of \(M\) models.
 
-As the generative model of the data, we assume that a patient with
-subsite \(s\) is generated with probability \(\pi_m^s\) from model
-\(m\),which satisfies the constraints \(\sum_{m = 1}^{M}\pi_m^s = 1\)
-and \(\pi_m^s \geq 0\) for all \(m\) and all subsites \(s\).
+To formalize this, each patient \(k\) can be assigned to a specific
+model via a binary latent vector \(\boldsymbol{\epsilon}_k\). The vector
+\(\boldsymbol{\epsilon}_k\) has length \(M\), with exactly one element
+set to 1, indicating which model generated the patient's data, and all
+other elements set to 0. Thus, for patient \(k\), the latent variable
+\(\boldsymbol{\epsilon}_k\) can be interpreted as a categorical
+indicator variable that encodes the assignment to one of the \(M\)
+lymphatic spread models.
 
-The likelihood of the dataset is expressed as:
+Next, we introduce subsite-specific mixing parameters
+\(\boldsymbol{\pi}^s = \{\pi_1^s, \pi_2^s, \dots, \pi_M^s\}\), where
+each \(\pi_m^s\) represents the prior probability that a patient with
+tumors in subsite \(s\) is generated by model \(m\). These mixing
+parameters must satisfy the condition:
 
-\begin{equation}\phantomsection\label{eq-mixture-distribution}{
-P \left( \mathbf{D} \mid \boldsymbol{\theta}, \boldsymbol{\pi}\right) = \prod_s \prod_i \left[ \sum_{m=1}^M \pi_m^s P_m \left( \mathbf{X}=\boldsymbol{\xi}_i \mid \boldsymbol{\theta}_m, T_i \right) \right]^{N_{is}}
+\[
+\sum_{m=1}^M \pi_m^s = 1, \quad \forall s
+\]
+
+The mixing parameters \(\boldsymbol{\pi}^s\) determine the proportion of
+patients with tumors in subsite \(s\) that are expected to be generated
+by each model. Specifically, \(\pi_m^s\) is the probability that a
+randomly chosen patient with a tumor in subsite \(s\) is generated by
+model \(m\).
+
+The latent variables \(\boldsymbol{\epsilon}_{k}\), on the other hand,
+are patient-specific and represent the assignment of each patient to a
+particular model. We will later generalize patients with the same
+diagnosis to the latent variable \(\boldsymbol{\epsilon}_{i,j}\). The
+subscripts \(i\) and \(j\) represent that the latent variables are only
+dependent on the diagnosis of the patient, which consists of lymphatic
+involvement \(\boldsymbol{\xi}_i\) and t-stage \(T_j\). The latent
+variables are central to the model, as they are hidden (unobserved) and
+must be inferred.
+
+The joint probability of the observed data \(\mathbf{D}\) (i.e., patient
+data) and the latent variables \(\boldsymbol{\epsilon}\) is given by:
+
+\begin{equation}\phantomsection\label{eq-complete-likelihood}{
+P(\mathbf{D}, \boldsymbol{\epsilon} | \boldsymbol{\theta}, \boldsymbol{\pi}) = \prod_{k=1}^{K} \prod_{m=1}^{M} \left[ \pi_m^s P(\mathbf{D}_k | \boldsymbol{\theta}_m, \boldsymbol{\pi}) \right]^{\epsilon_{k}^m}
 }\end{equation}
 
-In equation~\ref{eq-mixture-distribution} we compute the probability of
-each state \(\boldsymbol{\xi}_i\) under model \(m\) similarly to
-equation~\ref{eq-likelihood-HMM}. However, we weight each model with the
-mixing parameter \(\pi_m^s\). Since the mixing parameters are subsite
-dependent, we differ between each subsite.
+Here:
+
+\begin{itemize}
+\tightlist
+\item
+  \(\pi_m^s\) is the mixing parameter for subsite \(s\) and model \(m\),
+\item
+  \(P(\mathbf{D}_k | \boldsymbol{\theta}_m)\) is the likelihood of
+  patient \(k\)'s data given that it was generated by model \(m\),
+\item
+  \(\boldsymbol{\theta}_m\) represents the parameters of model \(m\),
+\item
+  \(\epsilon_{k}^m\) is the latent variable that indicates whether
+  patient \(k\) was generated by model \(m\).
+\end{itemize}
+
+\textbf{Incomplete data likelihood which can probably be skipped}
+
+Since the latent variables \(\boldsymbol{\epsilon}_k\) are unobserved,
+the incomplete complete data likelihood marginalizes over all possible
+latent assignments to reflect the uncertainty about which model
+generated each patient's data:
+
+\[
+P(\mathbf{D} | \boldsymbol{\theta}, \boldsymbol{\pi}) = \sum_{\boldsymbol{\epsilon}} P(\mathbf{D}, \boldsymbol{\epsilon} | \boldsymbol{\theta}, \boldsymbol{\pi})
+\]
+
+This allows us to integrate out the latent variables
+\(\boldsymbol{\epsilon}\), as their true values are unknown.
+
+\textbf{Continues from here}
+
+The latent variables \(\boldsymbol{\epsilon}_k\) are unknown and must be
+inferred from the observed data. The mixture model assumes that each
+patient is generated by exactly one of the \(M\) models. However, in our
+application, this assumption does not reflect continuous or overlapping
+behaviors between subsites. Instead, we infer a probabilistic
+distribution over the latent variables, given the observed data and
+model parameters. The posterior distribution:
+
+\[
+P(\boldsymbol{\epsilon}_k | \mathbf{D}_k, \boldsymbol{\theta}, \boldsymbol{\pi})
+\]
+
+describes the probability that patient \(k\) was generated by each of
+the \(M\) models, allowing for soft assignments where each patient is
+associated with multiple models with certain probabilities, as opposed
+to being strictly assigned to just one. Given the probability
+distribution for the latent variables \(\boldsymbol{\epsilon}_k\), we
+compute the expectation value
+\(\mathbb{E}[\boldsymbol{\epsilon}_k] = \boldsymbol{\gamma}(\boldsymbol{\epsilon}_k)\)
+for each patient to then apply the complete data likelihood given in
+equation~\ref{eq-complete-likelihood}.
+
+We can now apply the mixture model to our dataset \(\mathbf{D}\)
+comprising the number of patients \(N_{ijs}\) diagnosed with LNL
+involvement state \(\mathbf{X} = \boldsymbol{\xi}_i\), t-stage \(T_j\),
+with a primary tumor in subsite \(s\).
+
+The likelihood of the complete dataset is then expressed as:
+
+\begin{equation}\phantomsection\label{eq-complete-likelihood-explicite}{
+P \left( \mathbf{D} \mid \boldsymbol{\theta}, \boldsymbol{\pi}\right) = \prod_i \prod_j \prod_s \left[ \prod_{m=1}^M \left[ \pi_m^s P_m \left( \mathbf{X}=\boldsymbol{\xi}_i \mid \boldsymbol{\theta}_m, T_j \right) \right]^{\gamma(\epsilon_{i,j}^m)} \right]^{N_{ijs}}
+}\end{equation}
+
+In equation~\ref{eq-complete-likelihood} we compute the likelihood of
+observing state \(\boldsymbol{\xi}_i\) with t-stage \(T_j\) under model
+\(m\) similarly to equation~\ref{eq-likelihood-HMM}. However, we weight
+each model with the mixing parameter \(\pi_m^s\). Since the mixing
+parameters are subsite dependent, we differ between each subsite.
 
@@
 In the E-step, we compute the expected value of the latent variables
-\(\boldsymbol{\epsilon}\), which represent the probability that each
-observation \(\mathbf{X} = \boldsymbol{\xi}_i\) originates from model
-\(m\). Given the current estimates of \(\boldsymbol{\theta}\) and
-\(\boldsymbol{\pi}\), the expectation value \(\gamma(\epsilon_m^i)\) is
-calculated as:
+\(\boldsymbol{\epsilon}\), which represent the probabilities of a
+patient \(k\) originating from one of the models \(m\). Given the
+current estimates of \(\boldsymbol{\theta}\) and \(\boldsymbol{\pi}\),
+the expectation value \(\gamma(\epsilon_m^k)\) is calculated as:
 
 \begin{equation}\phantomsection\label{eq-expectation-step}{
-\mathbb{E}[\epsilon_m] = \gamma(\epsilon_m^i) = \frac{\pi_m^s P(\mathbf{X} = \boldsymbol{\xi}_i \mid \boldsymbol{\theta_m}, T_i)}{\sum_{j=1}^M \pi_j^s P(\mathbf{X} = \boldsymbol{\xi}_i \mid \boldsymbol{\theta_j}, T_i)}
+\mathbb{E}[\epsilon^m] = \gamma(\epsilon^m_{i,j}) = \frac{\pi_m^s P(\mathbf{X} = \boldsymbol{\xi}_i \mid \boldsymbol{\theta_m}, T_j)}{\sum_{n=1}^M \pi_n^s P(\mathbf{X} = \boldsymbol{\xi}_i \mid \boldsymbol{\theta_n}, T_j)}
 }\end{equation}
 
-Here, the mixing parameter \(\pi_m^s\) depends on the subsite of
-diagnosis, \(Z_i\), as denoted by the superscript \(s\).
+THE OTHER OPTION
+
+\begin{equation}\phantomsection\label{eq-expectation-step2}{
+\mathbb{E}[\epsilon^m] = \gamma(\epsilon^m_k) = \frac{\pi_m^s P(\mathbf{X} = \boldsymbol{\xi}_k \mid \boldsymbol{\theta_m}, T_k)}{\sum_{j=1}^M \pi_j^s P(\mathbf{X} = \boldsymbol{\xi}_k \mid \boldsymbol{\theta_j}, T_k)}
+}\end{equation}
+
+Here, the mixing parameter \(\pi_m^s\) depends on the subsite location
+of the primary tumor of patient \(k\).
 
@@
 In the M-step, the model parameters are updated by maximizing the
-following likelihood, based on the expectations computed in the E-step:
+complete data likelihood in equation
+equation~\ref{eq-complete-likelihood-explicite}. For better handling, we
+are optimizing the log likelihood:
 
 \begin{equation}\phantomsection\label{eq-maximization-step}{
-\ln P(\mathbf{D}, \boldsymbol{\epsilon} \mid \boldsymbol{\theta}, \boldsymbol{\pi}) = \sum_{i=1}^{N} \sum_{m=1}^M \gamma(\epsilon_m^i) \left( \ln \pi_m^s + \ln P(\mathbf{X} = \boldsymbol{\xi}_i \mid \boldsymbol{\theta_m}, T_i) \right)
+\ln P(\mathbf{D}, \boldsymbol{\epsilon} \mid \boldsymbol{\theta}, \boldsymbol{\pi}) = \sum_{i} \sum_j \sum_{s} N_{i,j,s} \sum_{m=1}^M \gamma(\epsilon^m_{i,j}) \left( \ln \pi_m^s + \ln P(\mathbf{X} = \boldsymbol{\xi}_i \mid \boldsymbol{\theta_m}, T_j) \right)
 }\end{equation}
@@
 \begin{equation}\phantomsection\label{eq-mixture-maximization}{
-\pi_m^s = \frac{1}{N^s} \sum_i^{N^s} \gamma(\epsilon_m^i)
+\pi_m^s = \frac{1}{N^s} \sum_k^{N^s} \gamma(\epsilon^m_k)
 }\end{equation}
@@
 \[
-\ln P(\mathbf{D}, \boldsymbol{\epsilon} \mid \boldsymbol{\theta}_m) = \sum_{i=1}^{N} \gamma(\epsilon_m^i) \ln P(\mathbf{X} = \boldsymbol{\xi}_i \mid \theta_m, T_i)
+\ln P(\mathbf{D}, \boldsymbol{\epsilon} \mid \boldsymbol{\theta}_m) = \sum_{i} \sum_j \gamma(\epsilon^m_{i,j}) \ln P(\mathbf{X} = \boldsymbol{\xi}_i \mid \theta_m, T_j)
 \]
@@
 By iterating these steps, the EM algorithm converges to a (local)
-maximum of the likelihood.
-
-\begin{equation}\phantomsection\label{eq-mixture-llh}{
-\ln p(\mathbf{D}|\boldsymbol{\theta},\boldsymbol{\pi}) = \left\{\sum_{\boldsymbol{\epsilon}} p(\mathbf{D},\boldsymbol{\epsilon}|\boldsymbol{\theta}, \boldsymbol{\pi})\right\}
-}\end{equation}
+maximum of the complete data likelihood
+equation~\ref{eq-complete-likelihood}.
 

### 2024-10-03 manuscript.qmd
Updates from Overleaf
reworked the math again to only us k

@@
 $$
-P \left( \mathbf{D} \mid \boldsymbol{\theta}, \boldsymbol{\pi}\right) = \prod_i \prod_j \left[P \left( \mathbf{X}=\boldsymbol{\xi}_i \mid \boldsymbol{\theta}, T_j \right) \right]^{N_{ij}}
+P \left( \mathbf{D} \mid \boldsymbol{\theta}, \boldsymbol{\pi}\right) = \prod_k^K \left[P \left( \mathbf{X}=\boldsymbol{\xi}_k \mid \boldsymbol{\theta}, T_k \right) \right]
 $$ {#eq-likelihood-HMM}
 
-In @eq-likelihood-HMM we take the likelihood of observing a patient in t-stage $T_j$ and with involvement $\xi_i$ to the power of $N_{ij}$, corresponding to the number of observations of the patients in state $i$ and t-stage $T_j$ (either early or late) in the dataset $\mathbf{D}$.
+In @eq-likelihood-HMM we compute the likelihood of observing patient the diagnosis of each patient $k$, i.e. t-stage $T_k$ and involvement $\xi_k$.
 
@@
 $$
-\sum_{m=1}^M \pi_m^s = 1, \quad \forall s
+\sum_{m}^M \pi_m^s = 1, \quad \forall s
 $$
@@
 The latent variables $\boldsymbol{\epsilon}_{k}$, on the other hand, are patient-specific and represent the assignment of each patient to a particular model.
-We will later generalize patients with the same diagnosis to the latent variable $\boldsymbol{\epsilon}_{i,j}$.
-The subscripts $i$ and $j$ represent that the latent variables are only dependent on the diagnosis of the patient, which consists of lymphatic involvement $\boldsymbol{\xi}_i$ and t-stage $T_j$.
 The latent variables are central to the model, as they are hidden (unobserved) and must be inferred.
@@
 $$
-P(\mathbf{D}, \boldsymbol{\epsilon} | \boldsymbol{\theta}, \boldsymbol{\pi}) = \prod_{k=1}^{K} \prod_{m=1}^{M} \left[ \pi_m^s P(\mathbf{D}_k | \boldsymbol{\theta}_m, \boldsymbol{\pi}) \right]^{\epsilon_{k}^m}
+P(\mathbf{D}, \boldsymbol{\epsilon} | \boldsymbol{\theta}, \boldsymbol{\pi}) = \prod_{k}^{K} \prod_{m}^{M} \left[ \pi_m^{s_k} P(\mathbf{D}_k | \boldsymbol{\theta}_m, \boldsymbol{\pi}) \right]^{\epsilon_{k}^m}
 $${#eq-complete-likelihood}
@@
 
-* $\pi_m^s$ is the mixing parameter for subsite $s$ and model $m$,
+* $\pi_m^{s_k}$ is the mixing parameter for patient $k$ with a primary tumor in subsite $s_k$ and model $m$,
 * $P(\mathbf{D}_k | \boldsymbol{\theta}_m)$ is the likelihood of patient $k$'s data given that it was generated by model $m$,
@@
 
-We can now apply the mixture model to our dataset $\mathbf{D}$ comprising the number of patients $N_{ijs}$ diagnosed with LNL involvement state $\mathbf{X} = \boldsymbol{\xi}_i$, t-stage $T_j$, with a primary tumor in subsite $s$.
-
 The likelihood of the complete dataset is then expressed as:
@@
 $$
-P \left( \mathbf{D} \mid \boldsymbol{\theta}, \boldsymbol{\pi}\right) = \prod_i \prod_j \prod_s \left[ \prod_{m=1}^M \left[ \pi_m^s P_m \left( \mathbf{X}=\boldsymbol{\xi}_i \mid \boldsymbol{\theta}_m, T_j \right) \right]^{\gamma(\epsilon_{i,j}^m)} \right]^{N_{ijs}}
+P \left( \mathbf{D} \mid \boldsymbol{\theta}, \boldsymbol{\pi}\right) = \prod_k^K \left[ \prod_{m}^M \pi_m^{s_k} P_m \left( \mathbf{X}=\boldsymbol{\xi}_k \mid \boldsymbol{\theta}_m, T_k \right) \right]^{\gamma(\epsilon_k^m)}
 $$ {#eq-complete-likelihood-explicite}
 
-In @eq-complete-likelihood we compute the likelihood of observing state $\boldsymbol{\xi}_i$ with t-stage $T_j$ under model $m$ similarly to @eq-likelihood-HMM.
-However, we weight each model with the mixing parameter $\pi_m^s$.
+In @eq-complete-likelihood we compute the likelihood of observing the patient's state $\boldsymbol{\xi}_k$ with t-stage $T_k$ under model $m$ similarly to @eq-likelihood-HMM.
+However, we weight each model with the mixing parameter $\pi_m^{s_k}$.
 Since the mixing parameters are subsite dependent, we differ between each subsite.
@@
 In the E-step, we compute the expected value of the latent variables $\boldsymbol{\epsilon}$, which represent the probabilities of a patient $k$ originating from one of the models $m$.
-Given the current estimates of $\boldsymbol{\theta}$ and $\boldsymbol{\pi}$, the expectation value $\gamma(\epsilon_m^k)$ is calculated as:
+Given the current estimates of $\boldsymbol{\theta}$ and $\boldsymbol{\pi}$, the expectation value $\gamma(\epsilon_k^m)$ is calculated as:
 
 $$
-\mathbb{E}[\epsilon^m] = \gamma(\epsilon^m_{i,j}) = \frac{\pi_m^s P(\mathbf{X} = \boldsymbol{\xi}_i \mid \boldsymbol{\theta_m}, T_j)}{\sum_{n=1}^M \pi_n^s P(\mathbf{X} = \boldsymbol{\xi}_i \mid \boldsymbol{\theta_n}, T_j)}
+\mathbb{E}[\epsilon^m_k] = \gamma(\epsilon^m_{k}) = \frac{\pi_m^{s_k} P(\mathbf{X} = \boldsymbol{\xi}_k \mid \boldsymbol{\theta_m}, T_k)}{\sum_{n}^M \pi_n^{s_k} P(\mathbf{X} = \boldsymbol{\xi}_k \mid \boldsymbol{\theta_n}, T_k)}
 $${#eq-expectation-step}
-
-THE OTHER OPTION
-
-$$
-\mathbb{E}[\epsilon^m] = \gamma(\epsilon^m_k) = \frac{\pi_m^s P(\mathbf{X} = \boldsymbol{\xi}_k \mid \boldsymbol{\theta_m}, T_k)}{\sum_{j=1}^M \pi_j^s P(\mathbf{X} = \boldsymbol{\xi}_k \mid \boldsymbol{\theta_j}, T_k)}
-$${#eq-expectation-step2}
-
-Here, the mixing parameter $\pi_m^s$ depends on the subsite location of the primary tumor of patient $k$.
 
@@
 $$
-\ln P(\mathbf{D}, \boldsymbol{\epsilon} \mid \boldsymbol{\theta}, \boldsymbol{\pi}) = \sum_{i} \sum_j \sum_{s} N_{i,j,s} \sum_{m=1}^M \gamma(\epsilon^m_{i,j}) \left( \ln \pi_m^s + \ln P(\mathbf{X} = \boldsymbol{\xi}_i \mid \boldsymbol{\theta_m}, T_j) \right)
+\ln P(\mathbf{D}, \boldsymbol{\epsilon} \mid \boldsymbol{\theta}, \boldsymbol{\pi}) = \sum_k^K \sum_{m}^M \gamma(\epsilon^m_{k}) \left( \ln \pi_m^{s_k} + \ln P(\mathbf{X} = \boldsymbol{\xi}_k \mid \boldsymbol{\theta_m}, T_k) \right)
 $${#eq-maximization-step}
@@
 $$
-\ln P(\mathbf{D}, \boldsymbol{\epsilon} \mid \boldsymbol{\theta}_m) = \sum_{i} \sum_j \gamma(\epsilon^m_{i,j}) \ln P(\mathbf{X} = \boldsymbol{\xi}_i \mid \theta_m, T_j)
+\ln P(\mathbf{D}, \boldsymbol{\epsilon} \mid \boldsymbol{\theta}_m) = \sum_{k}^K \gamma(\epsilon^m_{k}) \ln P(\mathbf{X} = \boldsymbol{\xi}_i \mid \theta_m, T_k)
 $$

### 2024-10-03 _tex/manuscript.tex
Built site for gh-pages

@@
 \begin{equation}\phantomsection\label{eq-likelihood-HMM}{
-P \left( \mathbf{D} \mid \boldsymbol{\theta}, \boldsymbol{\pi}\right) = \prod_i \prod_j \left[P \left( \mathbf{X}=\boldsymbol{\xi}_i \mid \boldsymbol{\theta}, T_j \right) \right]^{N_{ij}}
+P \left( \mathbf{D} \mid \boldsymbol{\theta}, \boldsymbol{\pi}\right) = \prod_k^K \left[P \left( \mathbf{X}=\boldsymbol{\xi}_k \mid \boldsymbol{\theta}, T_k \right) \right]
 }\end{equation}
 
-In equation~\ref{eq-likelihood-HMM} we take the likelihood of observing
-a patient in t-stage \(T_j\) and with involvement \(\xi_i\) to the power
-of \(N_{ij}\), corresponding to the number of observations of the
-patients in state \(i\) and t-stage \(T_j\) (either early or late) in
-the dataset \(\mathbf{D}\).
+In equation~\ref{eq-likelihood-HMM} we compute the likelihood of
+observing patient the diagnosis of each patient \(k\), i.e.~t-stage
+\(T_k\) and involvement \(\xi_k\).
 
@@
 \[
-\sum_{m=1}^M \pi_m^s = 1, \quad \forall s
+\sum_{m}^M \pi_m^s = 1, \quad \forall s
 \]
@@
 are patient-specific and represent the assignment of each patient to a
-particular model. We will later generalize patients with the same
-diagnosis to the latent variable \(\boldsymbol{\epsilon}_{i,j}\). The
-subscripts \(i\) and \(j\) represent that the latent variables are only
-dependent on the diagnosis of the patient, which consists of lymphatic
-involvement \(\boldsymbol{\xi}_i\) and t-stage \(T_j\). The latent
-variables are central to the model, as they are hidden (unobserved) and
-must be inferred.
+particular model. The latent variables are central to the model, as they
+are hidden (unobserved) and must be inferred.
 
@@
 \begin{equation}\phantomsection\label{eq-complete-likelihood}{
-P(\mathbf{D}, \boldsymbol{\epsilon} | \boldsymbol{\theta}, \boldsymbol{\pi}) = \prod_{k=1}^{K} \prod_{m=1}^{M} \left[ \pi_m^s P(\mathbf{D}_k | \boldsymbol{\theta}_m, \boldsymbol{\pi}) \right]^{\epsilon_{k}^m}
+P(\mathbf{D}, \boldsymbol{\epsilon} | \boldsymbol{\theta}, \boldsymbol{\pi}) = \prod_{k}^{K} \prod_{m}^{M} \left[ \pi_m^{s_k} P(\mathbf{D}_k | \boldsymbol{\theta}_m, \boldsymbol{\pi}) \right]^{\epsilon_{k}^m}
 }\end{equation}
@@
 \item
-  \(\pi_m^s\) is the mixing parameter for subsite \(s\) and model \(m\),
+  \(\pi_m^{s_k}\) is the mixing parameter for patient \(k\) with a
+  primary tumor in subsite \(s_k\) and model \(m\),
 \item
@@
 
-We can now apply the mixture model to our dataset \(\mathbf{D}\)
-comprising the number of patients \(N_{ijs}\) diagnosed with LNL
-involvement state \(\mathbf{X} = \boldsymbol{\xi}_i\), t-stage \(T_j\),
-with a primary tumor in subsite \(s\).
-
 The likelihood of the complete dataset is then expressed as:
@@
 \begin{equation}\phantomsection\label{eq-complete-likelihood-explicite}{
-P \left( \mathbf{D} \mid \boldsymbol{\theta}, \boldsymbol{\pi}\right) = \prod_i \prod_j \prod_s \left[ \prod_{m=1}^M \left[ \pi_m^s P_m \left( \mathbf{X}=\boldsymbol{\xi}_i \mid \boldsymbol{\theta}_m, T_j \right) \right]^{\gamma(\epsilon_{i,j}^m)} \right]^{N_{ijs}}
+P \left( \mathbf{D} \mid \boldsymbol{\theta}, \boldsymbol{\pi}\right) = \prod_k^K \left[ \prod_{m}^M \pi_m^{s_k} P_m \left( \mathbf{X}=\boldsymbol{\xi}_k \mid \boldsymbol{\theta}_m, T_k \right) \right]^{\gamma(\epsilon_k^m)}
 }\end{equation}
@@
 In equation~\ref{eq-complete-likelihood} we compute the likelihood of
-observing state \(\boldsymbol{\xi}_i\) with t-stage \(T_j\) under model
-\(m\) similarly to equation~\ref{eq-likelihood-HMM}. However, we weight
-each model with the mixing parameter \(\pi_m^s\). Since the mixing
-parameters are subsite dependent, we differ between each subsite.
+observing the patient's state \(\boldsymbol{\xi}_k\) with t-stage
+\(T_k\) under model \(m\) similarly to equation~\ref{eq-likelihood-HMM}.
+However, we weight each model with the mixing parameter \(\pi_m^{s_k}\).
+Since the mixing parameters are subsite dependent, we differ between
+each subsite.
 
@@
 current estimates of \(\boldsymbol{\theta}\) and \(\boldsymbol{\pi}\),
-the expectation value \(\gamma(\epsilon_m^k)\) is calculated as:
+the expectation value \(\gamma(\epsilon_k^m)\) is calculated as:
 
 \begin{equation}\phantomsection\label{eq-expectation-step}{
-\mathbb{E}[\epsilon^m] = \gamma(\epsilon^m_{i,j}) = \frac{\pi_m^s P(\mathbf{X} = \boldsymbol{\xi}_i \mid \boldsymbol{\theta_m}, T_j)}{\sum_{n=1}^M \pi_n^s P(\mathbf{X} = \boldsymbol{\xi}_i \mid \boldsymbol{\theta_n}, T_j)}
+\mathbb{E}[\epsilon^m_k] = \gamma(\epsilon^m_{k}) = \frac{\pi_m^{s_k} P(\mathbf{X} = \boldsymbol{\xi}_k \mid \boldsymbol{\theta_m}, T_k)}{\sum_{n}^M \pi_n^{s_k} P(\mathbf{X} = \boldsymbol{\xi}_k \mid \boldsymbol{\theta_n}, T_k)}
 }\end{equation}
-
-THE OTHER OPTION
-
-\begin{equation}\phantomsection\label{eq-expectation-step2}{
-\mathbb{E}[\epsilon^m] = \gamma(\epsilon^m_k) = \frac{\pi_m^s P(\mathbf{X} = \boldsymbol{\xi}_k \mid \boldsymbol{\theta_m}, T_k)}{\sum_{j=1}^M \pi_j^s P(\mathbf{X} = \boldsymbol{\xi}_k \mid \boldsymbol{\theta_j}, T_k)}
-}\end{equation}
-
-Here, the mixing parameter \(\pi_m^s\) depends on the subsite location
-of the primary tumor of patient \(k\).
 
@@
 \begin{equation}\phantomsection\label{eq-maximization-step}{
-\ln P(\mathbf{D}, \boldsymbol{\epsilon} \mid \boldsymbol{\theta}, \boldsymbol{\pi}) = \sum_{i} \sum_j \sum_{s} N_{i,j,s} \sum_{m=1}^M \gamma(\epsilon^m_{i,j}) \left( \ln \pi_m^s + \ln P(\mathbf{X} = \boldsymbol{\xi}_i \mid \boldsymbol{\theta_m}, T_j) \right)
+\ln P(\mathbf{D}, \boldsymbol{\epsilon} \mid \boldsymbol{\theta}, \boldsymbol{\pi}) = \sum_k^K \sum_{m}^M \gamma(\epsilon^m_{k}) \left( \ln \pi_m^{s_k} + \ln P(\mathbf{X} = \boldsymbol{\xi}_k \mid \boldsymbol{\theta_m}, T_k) \right)
 }\end{equation}
@@
 \[
-\ln P(\mathbf{D}, \boldsymbol{\epsilon} \mid \boldsymbol{\theta}_m) = \sum_{i} \sum_j \gamma(\epsilon^m_{i,j}) \ln P(\mathbf{X} = \boldsymbol{\xi}_i \mid \theta_m, T_j)
+\ln P(\mathbf{D}, \boldsymbol{\epsilon} \mid \boldsymbol{\theta}_m) = \sum_{k}^K \gamma(\epsilon^m_{k}) \ln P(\mathbf{X} = \boldsymbol{\xi}_i \mid \theta_m, T_k)
 \]
@@
 
---\textgreater{} Add predictions and compare to the prevalence.
-Additionally we could compare to tumor location predictions.
+add some analysis (level specific predictions and also comparison to
+single tumor location models.)
 

### 2024-10-10 manuscript.qmd
small comment added
extended methods

@@
 
-The mixture model assumes that the data is generated by a set of $M$ different underlying models.
-In our case, we introduce $M$ distinct lymphatic spread models.
-Each patient $k$ is assumed to be generated by one specific model from this set of $M$ models.
+The mixture model assumes that the data is generated by a set of $M$ different lymphatic spread models.
+Each patient $k$ is then assumed to be generated by one specific model from this set of $M$ models.
 
@@
 The mixture model assumes that each patient is generated by exactly one of the $M$ models.
-However, in our application, this assumption does not reflect continuous or overlapping behaviors between subsites.
-Instead, we infer a probabilistic distribution over the latent variables, given the observed data and model parameters.
-The posterior distribution:
+To do so a probabilistic distribution over the latent variables, given the observed data and model parameters is computed, the posterior distribution:
 
@@
 Since the mixing parameters are subsite dependent, we differ between each subsite.
+
+It is important to mention, that the latent variables $\boldsymbol{\epsilon}$ are only used for parameter estimation.
+The subsite specific mixture parameters $\boldsymbol{\pi}^s$ are the only parameters which are used for later risk predictions.
+In @fig-model-simple the mixture parameters are illustrated.
+Subsites with different spread patterns, such as Gum (C03) and Base of tongue (C01), are expected to get different model assignments.
+Nonetheless, the latent variables for two patients with the same diagnosis, i.e.
+same involvement $\boldsymbol{\xi}$ and and t-stage $T$, but different subsites are the same.
+
+![Illustration of mixture parameter assignment.
+Since Gum and Base of tongue express different spread patterns, the two models are expected to have different model assignments.
+The arrow visibility represents the value of the mixture parameter $\pi$, where the more visible the arrow, the larger the value for $\pi$](figures/mixture_model_simplified.png){#fig-model-simple}
 
@@
 
---> Add predictions and compare to the prevalence. Additionally we could compare to tumor location predictions.
+add some analysis (level specific predictions and also comparison to single tumor location models.)

### 2024-10-10 _tex/manuscript.tex
Built site for gh-pages

@@
 The mixture model assumes that the data is generated by a set of \(M\)
-different underlying models. In our case, we introduce \(M\) distinct
-lymphatic spread models. Each patient \(k\) is assumed to be generated
-by one specific model from this set of \(M\) models.
+different lymphatic spread models. Each patient \(k\) is then assumed to
+be generated by one specific model from this set of \(M\) models.
 
@@
 inferred from the observed data. The mixture model assumes that each
-patient is generated by exactly one of the \(M\) models. However, in our
-application, this assumption does not reflect continuous or overlapping
-behaviors between subsites. Instead, we infer a probabilistic
-distribution over the latent variables, given the observed data and
-model parameters. The posterior distribution:
+patient is generated by exactly one of the \(M\) models. To do so a
+probabilistic distribution over the latent variables, given the observed
+data and model parameters is computed, the posterior distribution:
 
@@
 each subsite.
+
+It is important to mention, that the latent variables
+\(\boldsymbol{\epsilon}\) are only used for parameter estimation. The
+subsite specific mixture parameters \(\boldsymbol{\pi}^s\) are the only
+parameters which are used for later risk predictions. In
+figure~\ref{fig-model-simple} the mixture parameters are illustrated.
+Subsites with different spread patterns, such as Gum (C03) and Base of
+tongue (C01), are expected to get different model assignments.
+Nonetheless, the latent variables for two patients with the same
+diagnosis, i.e.~same involvement \(\boldsymbol{\xi}\) and and t-stage
+\(T\), but different subsites are the same.
+
+\begin{figure}
+
+\centering{
+
+\includegraphics{figures/mixture_model_simplified.png}
+
+}
+
+\caption{\label{fig-model-simple}Illustration of mixture parameter
+assignment. Since Gum and Base of tongue express different spread
+patterns, the two models are expected to have different model
+assignments. The arrow visibility represents the value of the mixture
+parameter \(\pi\), where the more visible the arrow, the larger the
+value for \(\pi\)}
+
+\end{figure}%
 

### 2024-10-10 manuscript.qmd
use semicolons to separate multiple citations

@@
 # Introduction
-Head and Neck Squamous Cell Carcinoma (HNSCC) are known to spread through the lymphatic system  often leading to metastases in the lymph nodes [@mukherji_cervical_2001, @shah_patterns_1990].
+Head and Neck Squamous Cell Carcinoma (HNSCC) are known to spread through the lymphatic system  often leading to metastases in the lymph nodes [@mukherji_cervical_2001;@shah_patterns_1990].
 To minimize nodal recurrences, lymph node levels (LNL) at risk of harboring occult metastases are typically irradiated electively.
-Current guidelines for different tumor locations are based on the overall prevalence of nodal disease as reported in literature [@biau_selection_2019, @mukherji_cervical_2001, @shah_patterns_1990].
+Current guidelines for different tumor locations are based on the overall prevalence of nodal disease as reported in literature [@biau_selection_2019;@mukherji_cervical_2001;@shah_patterns_1990].
 
-To personalize the prediction of the risk of occult metastases, given a patient's individual diagnosis, we previously published a large, multi-centric dataset where the lymphatic involvement per LNL is available for each patient[@ludwig_dataset_2022, @ludwig_multi-centric_2023].Building on this dataset, we introduced an interpretable hidden Markov model (HMM), trained to predict the risk for occult nodal disease, given an individual patient's diagnosis [@ludwig_hidden_2021].
+To personalize the prediction of the risk of occult metastases, given a patient's individual diagnosis, we previously published a large, multi-centric dataset where the lymphatic involvement per LNL is available for each patient[@ludwig_dataset_2022;@ludwig_multi-centric_2023].Building on this dataset, we introduced an interpretable hidden Markov model (HMM), trained to predict the risk for occult nodal disease, given an individual patient's diagnosis [@ludwig_hidden_2021].
 

### 2024-10-10 manuscript.qmd
go over mixture model formalism (WIP)

@@
 # Introduction
-Head and Neck Squamous Cell Carcinoma (HNSCC) are known to spread through the lymphatic system  often leading to metastases in the lymph nodes [@mukherji_cervical_2001;@shah_patterns_1990].
-To minimize nodal recurrences, lymph node levels (LNL) at risk of harboring occult metastases are typically irradiated electively.
+
+Head and neck squamous cell carcinomas (HNSCCs) are known to spread through the lymphatic system often leading to metastases in the lymph nodes [@mukherji_cervical_2001;@shah_patterns_1990].
+To minimize nodal recurrences, lymph node levels (LNLs) at risk of harboring occult metastases are typically irradiated electively.
 Current guidelines for different tumor locations are based on the overall prevalence of nodal disease as reported in literature [@biau_selection_2019;@mukherji_cervical_2001;@shah_patterns_1990].
 
-To personalize the prediction of the risk of occult metastases, given a patient's individual diagnosis, we previously published a large, multi-centric dataset where the lymphatic involvement per LNL is available for each patient[@ludwig_dataset_2022;@ludwig_multi-centric_2023].Building on this dataset, we introduced an interpretable hidden Markov model (HMM), trained to predict the risk for occult nodal disease, given an individual patient's diagnosis [@ludwig_hidden_2021].
+To personalize the prediction of the risk of occult metastases, given a patient's individual diagnosis, we previously published a large, multi-centric dataset where the lymphatic involvement per LNL is available for each patient[@ludwig_dataset_2022;@ludwig_multi-centric_2023].
+Building on this dataset, we introduced an interpretable hidden Markov model (HMM), trained to predict the risk for occult nodal disease, given an individual patient's diagnosis [@ludwig_hidden_2021].
 
@@
 However, this approach did not account for variations in lymphatic spread between subsites within these tumor regions.
-With data for more than 2700 patients available, we can now further analyze subsite specific spread patterns.
+With data from more than 2700 patients available, we can now further analyze subsite specific spread patterns.
 Closer analysis showed that pooling subsites into a single model led to inaccurate predictions, as it failed to capture distinct lymphatic spread patterns.
@@
 
-Additionally, we extend the analysis to a broader mixture model that encompasses tumors of the oral cavity, oropharynx, hypopharynx, and larynx, offering a more comprehensive understanding of lymphatic spread across these regions.
+Additionally, we extend the analysis to a broader mixture model that encompasses tumors of the oral cavity, oropharynx, hypopharynx, and larynx, resulting in further personalized predictions of lymphatic spread across these regions.
 
@@
 
-The datasets 1-4 are publicly available as CSV tables (@ludwig_multi-centric_2023 @ludwig_detailed_2022) and can be interactively explored on [LyProX](https://lyprox.org).
+The datasets 1-4 are publicly available as CSV tables [@ludwig_multi-centric_2023;@ludwig_detailed_2022] and can be interactively explored on [LyProX](https://lyprox.org).
 For each patient the primary tumor subsite is reportd and each indicidual LNL is reported as either metastatic or healthy given the available diagnostic modalities, which include pathology after neck dissection in some patients.
@@
 The HMM describes each LNL $v \in 1, 2, ..., V$ by a binary random variable corresponding to the status of the LNL; healthy (0) or involved (1).
-The entire state of a patient with $V$ LNLs is defined by the $v$-dimensional vector $\mathbf{X} = [X_1, X_2, ...,X_V]$.
+The entire state of a patient with $V$ LNLs is defined by the $V$-dimensional vector $\mathbf{X} = [X_1, X_2, ...,X_V]$.
 In the HMM, a patient's involvement is modeled over time $t$.
@@
 The mixture model assumes that the data is generated by a set of $M$ different lymphatic spread models.
-Each patient $k$ is then assumed to be generated by one specific model from this set of $M$ models.
-
-To formalize this, each patient $k$ can be assigned to a specific model via a binary latent vector $\boldsymbol{\epsilon}_k$.
-The vector $\boldsymbol{\epsilon}_k$ has length $M$, with exactly one element set to 1, indicating which model generated the patient’s data, and all other elements set to 0.
-Thus, for patient $k$, the latent variable $\boldsymbol{\epsilon}_k$ can be interpreted as a categorical indicator variable that encodes the assignment to one of the $M$ lymphatic spread models.
-
-Next, we introduce subsite-specific mixing parameters $\boldsymbol{\pi}^s = \{\pi_1^s, \pi_2^s, \dots, \pi_M^s\}$, where each $\pi_m^s$ represents the prior probability that a patient with tumors in subsite $s$ is generated by model $m$.
-These mixing parameters must satisfy the condition:
+Each patient $k$, with their primary tumor in subsite $s \in (1, 2, \ldots, S)$, is assumed to be generated by one specific model $m \in (1, 2, \ldots, M)$ from this set of $M$ models with probability $\pi_m^s$.
+These so-called _mixing parameters_ $\boldsymbol{\pi}^s = \{\pi_1^s, \pi_2^s, \dots, \pi_M^s\}$ must satisfy the condition
 
@@
 
-The mixing parameters $\boldsymbol{\pi}^s$ determine the proportion of patients with tumors in subsite $s$ that are expected to be generated by each model.
-Specifically, $\pi_m^s$ is the probability that a randomly chosen patient with a tumor in subsite $s$ is generated by model $m$.
-
-The latent variables $\boldsymbol{\epsilon}_{k}$, on the other hand, are patient-specific and represent the assignment of each patient to a particular model.
-The latent variables are central to the model, as they are hidden (unobserved) and must be inferred.
+If we could record the componet $m$ from which a patient $k$ was drawn from, we could store this information in a binary latent vector $\boldsymbol{\epsilon}_k$.
+The vector $\boldsymbol{\epsilon}_k$ has length $M$, with exactly the $m$-th element set to 1, indicating which model generated the patient’s data, and all other elements set to 0.
+Thus, for patient $k$, the latent variable $\boldsymbol{\epsilon}_k$ can be interpreted as a categorical indicator variable that encodes the assignment to one of the $M$ lymphatic spread models.
+Unfortunately, the latent variables are hidden (unobserved) and must be inferred.
 

### 2024-10-11 manuscript.qmd
finish going through mixture & EM formalism

@@
 $$
-P \left( \mathbf{D} \mid \boldsymbol{\theta}, \boldsymbol{\pi}\right) = \prod_k^K \left[P \left( \mathbf{X}=\boldsymbol{\xi}_k \mid \boldsymbol{\theta}, T_k \right) \right]
+P \left( \mathbf{D} \mid \boldsymbol{\theta} \right) = \prod_k^K P \left( \mathbf{X}=\boldsymbol{\xi}_k \mid \boldsymbol{\theta}, T_k \right)
 $$ {#eq-likelihood-HMM}
@@
 Thus, for patient $k$, the latent variable $\boldsymbol{\epsilon}_k$ can be interpreted as a categorical indicator variable that encodes the assignment to one of the $M$ lymphatic spread models.
-Unfortunately, the latent variables are hidden (unobserved) and must be inferred.
+Typically in mixture models, this so-called _latent variable_ is unknown.
+However, it can be inferred and is useful for inferring the models' parameters.
 
-The joint probability of the observed data $\mathbf{D}$ (i.e., patient data) and the latent variables $\boldsymbol{\epsilon}$ is given by:
+The joint probability of the observed data $\mathbf{D}$ (i.e., patient data) and the latent variables $\boldsymbol{\epsilon}$ - sometimes called the _complete data likelihood_ - is given by:
 
 $$
-P(\mathbf{D}, \boldsymbol{\epsilon} | \boldsymbol{\theta}, \boldsymbol{\pi}) = \prod_{k}^{K} \prod_{m}^{M} \left[ \pi_m^{s_k} P(\mathbf{D}_k | \boldsymbol{\theta}_m, \boldsymbol{\pi}) \right]^{\epsilon_{k}^m}
-$${#eq-complete-likelihood}
+P(\mathbf{D}, \boldsymbol{\epsilon} \mid \boldsymbol{\theta}, \boldsymbol{\pi}) = \prod_{k}^{K} \prod_{m}^{M} \left[ \pi_m^{s_k} P(\mathbf{D}_k \mid \boldsymbol{\theta}_m) \right]^{\epsilon_{k}^m}
+$$ {#eq-complete-likelihood}
 
@@
 
-* $\pi_m^{s_k}$ is the mixing parameter for patient $k$ with a primary tumor in subsite $s_k$ and model $m$,
-* $P(\mathbf{D}_k | \boldsymbol{\theta}_m)$ is the likelihood of patient $k$'s data given that it was generated by model $m$,
-* $\boldsymbol{\theta}_m$ represents the parameters of model $m$,
-* $\epsilon_{k}^m$ is the latent variable that indicates whether patient $k$ was generated by model $m$.
+- $\pi_m^{s_k}$ is the mixing coefficient for subsite $s_k$ (where patient $k$ has their tumor) and model $m$,
+- $P(\mathbf{D}_k \mid \boldsymbol{\theta}_m)$ is the likelihood of patient $k$'s diagnosis, given that it was generated by model $m$,
+- $\boldsymbol{\theta}_m$ represents the parameters of model $m$,
+- $\epsilon_{k}^m$ is the latent variable that indicates patient $k$ was generated by model $m$.
 
-**Incomplete data likelihood which can probably be skipped**
+::: {.callout-note}
 
-Since the latent variables $\boldsymbol{\epsilon}_k$ are unobserved, the incomplete complete data likelihood marginalizes over all possible latent assignments to reflect the uncertainty about which model generated each patient’s data:
+## TODO: Discuss if we need this
+<!-- Not sure we can skip it. We want to maximize this w.r.t. the model parameters. -->
+
+In contrast, the _incomplete data likelihood_ marginalizes over all possible latent assignments to reflect the uncertainty about which model generated each patient’s data:
 
 $$
-P(\mathbf{D} | \boldsymbol{\theta}, \boldsymbol{\pi}) = \sum_{\boldsymbol{\epsilon}} P(\mathbf{D}, \boldsymbol{\epsilon} | \boldsymbol{\theta}, \boldsymbol{\pi})
-$$
+P(\mathbf{D}, \mid \boldsymbol{\theta}, \boldsymbol{\pi}) = \prod_{k}^{K} \sum_{m}^{M} \pi_m^{s_k} P(\mathbf{D}_k \mid \boldsymbol{\theta}_m)
+$$ {#eq-incomplete-likelihood}
 
-This allows us to integrate out the latent variables $\boldsymbol{\epsilon}$, as their true values are unknown.
+Ultimately, we want to find the parameters which maximize this likelihood function for the given data.
 
-**Continues from here**
+Note the summation inside the product.
+This structure of mixture models makes naive inference difficult, because the logarithm of this quantity is expensive to compute and not easy to differentiate.
+Thus, inferring the latent assignment is helpful, because the complete data (log-)likelihood (@eq-complete-likelihood) does not suffer from this shortcoming.
 
-The latent variables $\boldsymbol{\epsilon}_k$ are unknown and must be inferred from the observed data.
-The mixture model assumes that each patient is generated by exactly one of the $M$ models.
-To do so a probabilistic distribution over the latent variables, given the observed data and model parameters is computed, the posterior distribution:
+:::
+
+To infer the latent assignment $\boldsymbol{\epsilon}_k$, we start with its distribution, given the observed data and model parameters:
 
 $$
-P(\boldsymbol{\epsilon}_k | \mathbf{D}_k, \boldsymbol{\theta}, \boldsymbol{\pi})
-$$
+\gamma(\epsilon_k^m) := \mathbb{E}[\epsilon_k^m] = \frac{\pi_m^{s_k} P \left( \mathbf{D}_k \mid \boldsymbol{\theta}_m \right)}{\sum_{j \leq M} \pi_j^{s_k} P \left( \mathbf{D}_k \mid \boldsymbol{\theta}_j \right)}
+$$ {#eq-responsibilities}
 
-describes the probability that patient $k$ was generated by each of the $M$ models, allowing for soft assignments where each patient is associated with multiple models with certain probabilities, as opposed to being strictly assigned to just one.
-Given the probability distribution for the latent variables $\boldsymbol{\epsilon}_k$, we compute the expectation value $\mathbb{E}[\boldsymbol{\epsilon}_k] = \boldsymbol{\gamma}(\boldsymbol{\epsilon}_k)$ for each patient to then apply the complete data likelihood given in @eq-complete-likelihood.
-
-The likelihood of the complete dataset is then expressed as:
+This expectation value - often called the _responsibility_ - describes the probability that patient $k$ was generated by model $m$. It can be used to compute the expected complete data likelihood:
 
 $$
-P \left( \mathbf{D} \mid \boldsymbol{\theta}, \boldsymbol{\pi}\right) = \prod_k^K \left[ \prod_{m}^M \pi_m^{s_k} P_m \left( \mathbf{X}=\boldsymbol{\xi}_k \mid \boldsymbol{\theta}_m, T_k \right) \right]^{\gamma(\epsilon_k^m)}
-$$ {#eq-complete-likelihood-explicite}
+\mathbb{E}_{\boldsymbol{\epsilon}} \left[ P \left( \mathbf{D}, \boldsymbol{\epsilon} \mid \boldsymbol{\theta}, \boldsymbol{\pi} \right) \right] = \prod_{k=1}^K \prod_{m=1}^M \left[ \pi_m^{s_k} P \left( \mathbf{D}_k \mid \boldsymbol{\theta}_m \right) \right]^{\gamma(\epsilon_k^m)}
+$$ {#eq-expected-complete-likelihood}
 
-In @eq-complete-likelihood we compute the likelihood of observing the patient's state $\boldsymbol{\xi}_k$ with t-stage $T_k$ under model $m$ similarly to @eq-likelihood-HMM.
-However, we weight each model with the mixing parameter $\pi_m^{s_k}$.
-Since the mixing parameters are subsite dependent, we differ between each subsite.
+This expected complete data likelihood has the same tractable form as @eq-complete-likelihood and thus allows us to infer both the mixing coefficients $\boldsymbol{\pi}$ as well as each component model's parameters $\boldsymbol{\theta}_m$ using straightformward inference.
+Also, these two sets of parameters are the only ones used for the later risk prediction.
+The responsiblities $\gamma(\epsilon_k^m)$ are only used during inference.
 
-It is important to mention, that the latent variables $\boldsymbol{\epsilon}$ are only used for parameter estimation.
-The subsite specific mixture parameters $\boldsymbol{\pi}^s$ are the only parameters which are used for later risk predictions.
-In @fig-model-simple the mixture parameters are illustrated.
+In @fig-model-simple the mixture coefficients are illustrated.
 Subsites with different spread patterns, such as Gum (C03) and Base of tongue (C01), are expected to get different model assignments.
@@
 
-The model now contains two sets of parameters: the tumor spread probabilities $\mathbf{\theta_m}$ for each model, and the mixing coefficients $\pi_m^s$.
-The posterior distribution over these parameters $P(\boldsymbol{\theta},\boldsymbol{\pi}|\mathbf{D})$ can be estimated assuming a uniform prior on the interval [0,1] for all parameters.
-However, $P(\boldsymbol{\theta},\boldsymbol{\pi}|\mathbf{D})$ is a multimodal-distribution as one can permute the different models and achieve the same result.
-To address this problem, we apply an expectation-maximization (EM) algorithm.
+## Expectation-Maximization (EM) Algorithm
 
-## Expectation-Maximization (EM) Algorithm
-To solve for the model parameters, we apply the EM algorithm, which iterates between two steps: the **Expectation (E) step** and **the Maximization (M) step**.
+To actually find the mixing coefficients $\boldsymbol{\pi}$ and all models' parameters $\boldsymbol{\theta}_m$ that maximize @eq-incomplete-likelihood, we follow an iterative approach called _expectation-maximization_ or _EM-algorithm_.
+With arbitrarily initialized starting parameters, we alternate between the following two steps:
 
-#### Expectation Step
+1.
+In the **E**xpectation step, we compute the responsibilities $\gamma(\epsilon_k^m)$, which represent the probabilities of a patient $k$ originating from one of the models $m$, given the current estimates of $\boldsymbol{\theta}$ and $\boldsymbol{\pi}$, as given in @eq-responsibilities.
+2.
+During the **M**maximization step, we find a new set of parameters that maximize the new expected complete data likelihood (@eq-expected-complete-likelihood).
+For the mixture coefficients, we can even find an analytic solution to the new maximum:
+   $$
+   \pi_m^s = \frac{1}{|K_s|} \sum_{k \in K_s} \gamma(\epsilon_k^m)
+   $$
+   where we sum over the set of all patients $K_s$ with their tumor in subsite $s$.\
+   The models' new parameters are found by numerically maximizing the respective likelihood, weighted by the responsibilities:
+   $$
+   \ln P(\mathbf{D}, \boldsymbol{\epsilon} \mid \boldsymbol{\theta}_m) = \sum_k^K \gamma(\epsilon^m_{k}) \big[ \ln \pi_m^{s_k} + \ln P \left( \mathbf{D}_k \mid \boldsymbol{\theta}_m \right) \big]
+   $$ {#eq-maximization-step}
 
-In the E-step, we compute the expected value of the latent variables $\boldsymbol{\epsilon}$, which represent the probabilities of a patient $k$ originating from one of the models $m$.
-Given the current estimates of $\boldsymbol{\theta}$ and $\boldsymbol{\pi}$, the expectation value $\gamma(\epsilon_k^m)$ is calculated as:
-
-$$
-\mathbb{E}[\epsilon^m_k] = \gamma(\epsilon^m_{k}) = \frac{\pi_m^{s_k} P(\mathbf{X} = \boldsymbol{\xi}_k \mid \boldsymbol{\theta_m}, T_k)}{\sum_{n}^M \pi_n^{s_k} P(\mathbf{X} = \boldsymbol{\xi}_k \mid \boldsymbol{\theta_n}, T_k)}
-$${#eq-expectation-step}
-
-#### Maximization Step
-
-In the M-step, the model parameters are updated by maximizing the complete data likelihood in equation @eq-complete-likelihood-explicite. For better handling, we are optimizing the log likelihood:
-
-$$
-\ln P(\mathbf{D}, \boldsymbol{\epsilon} \mid \boldsymbol{\theta}, \boldsymbol{\pi}) = \sum_k^K \sum_{m}^M \gamma(\epsilon^m_{k}) \left( \ln \pi_m^{s_k} + \ln P(\mathbf{X} = \boldsymbol{\xi}_k \mid \boldsymbol{\theta_m}, T_k) \right)
-$${#eq-maximization-step}
-
-In this step, the mixture parameters $\pi_m^s$ and the model parameters $\boldsymbol{\theta}_m$ can be optimized separately. The mixture parameters $\pi_m^s$ can be updated analytically:
-
-$$
-\pi_m^s = \frac{1}{N^s} \sum_k^{N^s} \gamma(\epsilon^m_k)
-$${#eq-mixture-maximization}
-
-On the other hand, the model parameters $\boldsymbol{\theta_m}$ are optimized component-wise using an optimization algorithm, as an analytic solution is not feasible.
-The updated log-likelihood for each model is:
-
-$$
-\ln P(\mathbf{D}, \boldsymbol{\epsilon} \mid \boldsymbol{\theta}_m) = \sum_{k}^K \gamma(\epsilon^m_{k}) \ln P(\mathbf{X} = \boldsymbol{\xi}_i \mid \theta_m, T_k)
-$$
-
-By iterating these steps, the EM algorithm converges to a (local) maximum of the complete data likelihood @eq-complete-likelihood.
+By iterating these steps, the EM algorithm is guaranteed to converge to a (local) maximum of the complete data likelihood (@eq-complete-likelihood).
 

### 2024-10-14 _tex/manuscript.tex
Built site for gh-pages

@@
 \makeatletter
+\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
+\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
+\definecolor{quarto-callout-color}{HTML}{909090}
+\definecolor{quarto-callout-note-color}{HTML}{0758E5}
+\definecolor{quarto-callout-important-color}{HTML}{CC1914}
+\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
+\definecolor{quarto-callout-tip-color}{HTML}{00A047}
+\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
+\definecolor{quarto-callout-color-frame}{HTML}{acacac}
+\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
+\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
+\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
+\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
+\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
+\makeatother
+\makeatletter
 \@ifpackageloaded{caption}{}{\usepackage{caption}}
@@
 
-Head and Neck Squamous Cell Carcinoma (HNSCC) are known to spread
+Head and neck squamous cell carcinomas (HNSCCs) are known to spread
 through the lymphatic system often leading to metastases in the lymph
-nodes \citep[@shah\_patterns\_1990]{mukherji_cervical_2001}. To minimize
-nodal recurrences, lymph node levels (LNL) at risk of harboring occult
+nodes \citep{mukherji_cervical_2001, shah_patterns_1990}. To minimize
+nodal recurrences, lymph node levels (LNLs) at risk of harboring occult
 metastases are typically irradiated electively. Current guidelines for
@@
 disease as reported in literature
-\citep[@mukherji\_cervical\_2001, @shah\_patterns\_1990]{biau_selection_2019}.
+\citep{biau_selection_2019, mukherji_cervical_2001, shah_patterns_1990}.
 
@@
 available for each
-patient\citep[@ludwig\_multi-centric\_2023]{ludwig_dataset_2022}.Building
+patient\citep{ludwig_dataset_2022, ludwig_multi-centric_2023}. Building
 on this dataset, we introduced an interpretable hidden Markov model
@@
 variations in lymphatic spread between subsites within these tumor
-regions. With data for more than 2700 patients available, we can now
+regions. With data from more than 2700 patients available, we can now
 further analyze subsite specific spread patterns. Closer analysis showed
@@
 encompasses tumors of the oral cavity, oropharynx, hypopharynx, and
-larynx, offering a more comprehensive understanding of lymphatic spread
-across these regions.
+larynx, resulting in further personalized predictions of lymphatic
+spread across these regions.
 
@@
 The datasets 1-4 are publicly available as CSV tables
-(\citet{ludwig_multi-centric_2023} \citet{ludwig_detailed_2022}) and can
-be interactively explored on \href{https://lyprox.org}{LyProX}. For each
+\citep{ludwig_multi-centric_2023, ludwig_detailed_2022} and can be
+interactively explored on \href{https://lyprox.org}{LyProX}. For each
 patient the primary tumor subsite is reportd and each indicidual LNL is
@@
 (1). The entire state of a patient with \(V\) LNLs is defined by the
-\(v\)-dimensional vector \(\mathbf{X} = [X_1, X_2, ...,X_V]\). In the
+\(V\)-dimensional vector \(\mathbf{X} = [X_1, X_2, ...,X_V]\). In the
 HMM, a patient's involvement is modeled over time \(t\). Thus, a
@@
 \begin{equation}\phantomsection\label{eq-likelihood-HMM}{
-P \left( \mathbf{D} \mid \boldsymbol{\theta}, \boldsymbol{\pi}\right) = \prod_k^K \left[P \left( \mathbf{X}=\boldsymbol{\xi}_k \mid \boldsymbol{\theta}, T_k \right) \right]
+P \left( \mathbf{D} \mid \boldsymbol{\theta} \right) = \prod_k^K P \left( \mathbf{X}_k=\boldsymbol{\xi}_i \mid \boldsymbol{\theta}, T_k \right)
 }\end{equation}
@@
 observing patient the diagnosis of each patient \(k\), i.e.~t-stage
-\(T_k\) and involvement \(\xi_k\).
+\(T_k\) and involvement \(\mathbf{X}_k=\boldsymbol{\xi}_i\).
 
@@
 The mixture model assumes that the data is generated by a set of \(M\)
-different lymphatic spread models. Each patient \(k\) is then assumed to
-be generated by one specific model from this set of \(M\) models.
-
-To formalize this, each patient \(k\) can be assigned to a specific
-model via a binary latent vector \(\boldsymbol{\epsilon}_k\). The vector
-\(\boldsymbol{\epsilon}_k\) has length \(M\), with exactly one element
-set to 1, indicating which model generated the patient's data, and all
-other elements set to 0. Thus, for patient \(k\), the latent variable
-\(\boldsymbol{\epsilon}_k\) can be interpreted as a categorical
-indicator variable that encodes the assignment to one of the \(M\)
-lymphatic spread models.
-
-Next, we introduce subsite-specific mixing parameters
-\(\boldsymbol{\pi}^s = \{\pi_1^s, \pi_2^s, \dots, \pi_M^s\}\), where
-each \(\pi_m^s\) represents the prior probability that a patient with
-tumors in subsite \(s\) is generated by model \(m\). These mixing
-parameters must satisfy the condition:
+different lymphatic spread models. Each patient \(k\), with their
+primary tumor in subsite \(s \in (1, 2, \ldots, S)\), is assumed to be
+generated by one specific model \(m \in (1, 2, \ldots, M)\) from this
+set of \(M\) models with probability \(\pi_m^s\). These so-called
+\emph{mixing parameters}
+\(\boldsymbol{\pi}^s = \{\pi_1^s, \pi_2^s, \dots, \pi_M^s\}\) must
+satisfy the condition
 
@@
 
-The mixing parameters \(\boldsymbol{\pi}^s\) determine the proportion of
-patients with tumors in subsite \(s\) that are expected to be generated
-by each model. Specifically, \(\pi_m^s\) is the probability that a
-randomly chosen patient with a tumor in subsite \(s\) is generated by
-model \(m\).
-
-The latent variables \(\boldsymbol{\epsilon}_{k}\), on the other hand,
-are patient-specific and represent the assignment of each patient to a
-particular model. The latent variables are central to the model, as they
-are hidden (unobserved) and must be inferred.
+If we could record the componet \(m\) from which a patient \(k\) was
+drawn from, we could store this information in a binary latent vector
+\(\boldsymbol{\epsilon}_k\). The vector \(\boldsymbol{\epsilon}_k\) has
+length \(M\), with exactly the \(m\)-th element set to 1, indicating
+which model generated the patient's data, and all other elements set to
+0. Thus, for patient \(k\), the latent variable
+\(\boldsymbol{\epsilon}_k\) can be interpreted as a categorical
+indicator variable that encodes the assignment to one of the \(M\)
+lymphatic spread models. Typically in mixture models, this so-called
+\emph{latent variable} is unknown. However, it can be inferred and is
+useful for inferring the models' parameters.
 
 The joint probability of the observed data \(\mathbf{D}\) (i.e., patient
-data) and the latent variables \(\boldsymbol{\epsilon}\) is given by:
+data) and the latent variables \(\boldsymbol{\epsilon}\) - sometimes
+called the \emph{complete data likelihood} - is given by:
 
 \begin{equation}\phantomsection\label{eq-complete-likelihood}{
-P(\mathbf{D}, \boldsymbol{\epsilon} | \boldsymbol{\theta}, \boldsymbol{\pi}) = \prod_{k}^{K} \prod_{m}^{M} \left[ \pi_m^{s_k} P(\mathbf{D}_k | \boldsymbol{\theta}_m, \boldsymbol{\pi}) \right]^{\epsilon_{k}^m}
+P(\mathbf{D}, \boldsymbol{\epsilon} \mid \boldsymbol{\theta}, \boldsymbol{\pi}) = \prod_{k}^{K} \prod_{m}^{M} \left[ \pi_m^{s_k} P(\mathbf{D}_k \mid \boldsymbol{\theta}_m) \right]^{\epsilon_{k}^m}
 }\end{equation}
@@
 \item
-  \(\pi_m^{s_k}\) is the mixing parameter for patient \(k\) with a
-  primary tumor in subsite \(s_k\) and model \(m\),
+  \(\pi_m^{s_k}\) is the mixing coefficient for subsite \(s_k\) (where
+  patient \(k\) has their tumor) and model \(m\),
 \item
-  \(P(\mathbf{D}_k | \boldsymbol{\theta}_m)\) is the likelihood of
-  patient \(k\)'s data given that it was generated by model \(m\),
+  \(P(\mathbf{D}_k \mid \boldsymbol{\theta}_m)\) is the likelihood of
+  patient \(k\)'s diagnosis, given that it was generated by model \(m\),
 \item
@@
 \item
-  \(\epsilon_{k}^m\) is the latent variable that indicates whether
-  patient \(k\) was generated by model \(m\).
+  \(\epsilon_{k}^m\) is the latent variable that indicates patient \(k\)
+  was generated by model \(m\).
 \end{itemize}
 
-\textbf{Incomplete data likelihood which can probably be skipped}
+\begin{tcolorbox}[enhanced jigsaw, breakable, colbacktitle=quarto-callout-note-color!10!white, left=2mm, colback=white, coltitle=black, arc=.35mm, bottomrule=.15mm, rightrule=.15mm, bottomtitle=1mm, opacitybacktitle=0.6, title=\textcolor{quarto-callout-note-color}{\faInfo}\hspace{0.5em}{TODO: Discuss if we need this}, titlerule=0mm, toptitle=1mm, toprule=.15mm, colframe=quarto-callout-note-color-frame, leftrule=.75mm, opacityback=0]
 
-Since the latent variables \(\boldsymbol{\epsilon}_k\) are unobserved,
-the incomplete complete data likelihood marginalizes over all possible
-latent assignments to reflect the uncertainty about which model
+In contrast, the \emph{incomplete data likelihood} marginalizes over all
+possible latent assignments to reflect the uncertainty about which model
 generated each patient's data:
 
-\[
-P(\mathbf{D} | \boldsymbol{\theta}, \boldsymbol{\pi}) = \sum_{\boldsymbol{\epsilon}} P(\mathbf{D}, \boldsymbol{\epsilon} | \boldsymbol{\theta}, \boldsymbol{\pi})
-\]
-
-This allows us to integrate out the latent variables
-\(\boldsymbol{\epsilon}\), as their true values are unknown.
-
-\textbf{Continues from here}
-
-The latent variables \(\boldsymbol{\epsilon}_k\) are unknown and must be
-inferred from the observed data. The mixture model assumes that each
-patient is generated by exactly one of the \(M\) models. To do so a
-probabilistic distribution over the latent variables, given the observed
-data and model parameters is computed, the posterior distribution:
-
-\[
-P(\boldsymbol{\epsilon}_k | \mathbf{D}_k, \boldsymbol{\theta}, \boldsymbol{\pi})
-\]
-
-describes the probability that patient \(k\) was generated by each of
-the \(M\) models, allowing for soft assignments where each patient is
-associated with multiple models with certain probabilities, as opposed
-to being strictly assigned to just one. Given the probability
-distribution for the latent variables \(\boldsymbol{\epsilon}_k\), we
-compute the expectation value
-\(\mathbb{E}[\boldsymbol{\epsilon}_k] = \boldsymbol{\gamma}(\boldsymbol{\epsilon}_k)\)
-for each patient to then apply the complete data likelihood given in
-equation~\ref{eq-complete-likelihood}.
-
-The likelihood of the complete dataset is then expressed as:
-
-\begin{equation}\phantomsection\label{eq-complete-likelihood-explicite}{
-P \left( \mathbf{D} \mid \boldsymbol{\theta}, \boldsymbol{\pi}\right) = \prod_k^K \left[ \prod_{m}^M \pi_m^{s_k} P_m \left( \mathbf{X}=\boldsymbol{\xi}_k \mid \boldsymbol{\theta}_m, T_k \right) \right]^{\gamma(\epsilon_k^m)}
+\begin{equation}\phantomsection\label{eq-incomplete-likelihood}{
+P(\mathbf{D}, \mid \boldsymbol{\theta}, \boldsymbol{\pi}) = \prod_{k}^{K} \sum_{m}^{M} \pi_m^{s_k} P(\mathbf{D}_k \mid \boldsymbol{\theta}_m)
 }\end{equation}
 
-In equation~\ref{eq-complete-likelihood} we compute the likelihood of
-observing the patient's state \(\boldsymbol{\xi}_k\) with t-stage
-\(T_k\) under model \(m\) similarly to equation~\ref{eq-likelihood-HMM}.
-However, we weight each model with the mixing parameter \(\pi_m^{s_k}\).
-Since the mixing parameters are subsite dependent, we differ between
-each subsite.
+Ultimately, we want to find the parameters which maximize this
+likelihood function for the given data.
 
-It is important to mention, that the latent variables
-\(\boldsymbol{\epsilon}\) are only used for parameter estimation. The
-subsite specific mixture parameters \(\boldsymbol{\pi}^s\) are the only
-parameters which are used for later risk predictions. In
-figure~\ref{fig-model-simple} the mixture parameters are illustrated.
-Subsites with different spread patterns, such as Gum (C03) and Base of
-tongue (C01), are expected to get different model assignments.
-Nonetheless, the latent variables for two patients with the same
-diagnosis, i.e.~same involvement \(\boldsymbol{\xi}\) and and t-stage
-\(T\), but different subsites are the same.
+Note the summation inside the product. This structure of mixture models
+makes naive inference difficult, because the logarithm of this quantity
+is expensive to compute and not easy to differentiate. Thus, inferring
+the latent assignment is helpful, because the complete data
+(log-)likelihood (equation~\ref{eq-complete-likelihood}) does not suffer
+from this shortcoming.
+
+\end{tcolorbox}
+
+To infer the latent assignment \(\boldsymbol{\epsilon}_k\), we start
+with its distribution, given the observed data and model parameters:
+
+\begin{equation}\phantomsection\label{eq-responsibilities}{
+\gamma(\epsilon_k^m) := \mathbb{E}[\epsilon_k^m] = \frac{\pi_m^{s_k} P \left( \mathbf{D}_k \mid \boldsymbol{\theta}_m \right)}{\sum_{j \leq M} \pi_j^{s_k} P \left( \mathbf{D}_k \mid \boldsymbol{\theta}_j \right)}
+}\end{equation}
+
+This expectation value - often called the \emph{responsibility} -
+describes the probability that patient \(k\) was generated by model
+\(m\). It can be used to compute the expected complete data likelihood:
+
+\begin{equation}\phantomsection\label{eq-expected-complete-likelihood}{
+\mathbb{E}_{\boldsymbol{\epsilon}} \left[ P \left( \mathbf{D}, \boldsymbol{\epsilon} \mid \boldsymbol{\theta}, \boldsymbol{\pi} \right) \right] = \prod_{k=1}^K \prod_{m=1}^M \left[ \pi_m^{s_k} P \left( \mathbf{D}_k \mid \boldsymbol{\theta}_m \right) \right]^{\gamma(\epsilon_k^m)}
+}\end{equation}
+
+This expected complete data likelihood has the same tractable form as
+equation~\ref{eq-complete-likelihood} and thus allows us to infer both
+the mixing coefficients \(\boldsymbol{\pi}\) as well as each component
+model's parameters \(\boldsymbol{\theta}_m\) using straightformward
+inference. Also, these two sets of parameters are the only ones used for
+the later risk prediction. The responsiblities \(\gamma(\epsilon_k^m)\)
+are only used during inference.
+
+In figure~\ref{fig-model-simple} the mixture coefficients are
+illustrated. Subsites with different spread patterns, such as Gum (C03)
+and Base of tongue (C01), are expected to get different model
+assignments. Nonetheless, the latent variables for two patients with the
+same diagnosis, i.e.~same involvement \(\boldsymbol{\xi}\) and and
+t-stage \(T\), but different subsites are the same.
 
@@
 
-The model now contains two sets of parameters: the tumor spread
-probabilities \(\mathbf{\theta_m}\) for each model, and the mixing
-coefficients \(\pi_m^s\). The posterior distribution over these
-parameters \(P(\boldsymbol{\theta},\boldsymbol{\pi}|\mathbf{D})\) can be
-estimated assuming a uniform prior on the interval {[}0,1{]} for all
-parameters. However,
-\(P(\boldsymbol{\theta},\boldsymbol{\pi}|\mathbf{D})\) is a
-multimodal-distribution as one can permute the different models and
-achieve the same result. To address this problem, we apply an
-expectation-maximization (EM) algorithm.
-
 \subsection{Expectation-Maximization (EM)
@@
 
-To solve for the model parameters, we apply the EM algorithm, which
-iterates between two steps: the \textbf{Expectation (E) step} and
-\textbf{the Maximization (M) step}.
+To actually find the mixing coefficients \(\boldsymbol{\pi}\) and all
+models' parameters \(\boldsymbol{\theta}_m\) that maximize
+equation~\ref{eq-incomplete-likelihood}, we follow an iterative approach
+called \emph{expectation-maximization} or \emph{EM-algorithm}. With
+arbitrarily initialized starting parameters, we alternate between the
+following two steps:
 
-\paragraph{Expectation Step}\label{expectation-step}
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  In the \textbf{E}xpectation step, we compute the responsibilities
+  \(\gamma(\epsilon_k^m)\), which represent the probabilities of a
+  patient \(k\) originating from one of the models \(m\), given the
+  current estimates of \(\boldsymbol{\theta}\) and \(\boldsymbol{\pi}\),
+  as given in equation~\ref{eq-responsibilities}.
+\item
+  During the \textbf{M}maximization step, we find a new set of
+  parameters that maximize the new expected complete data likelihood
+  (equation~\ref{eq-expected-complete-likelihood}). For the mixture
+  coefficients, we can even find an analytic solution to the new
+  maximum: \[
+  \pi_m^s = \frac{1}{|K_s|} \sum_{k \in K_s} \gamma(\epsilon_k^m)
+  \] where we sum over the set of all patients \(K_s\) with their tumor
+  in subsite \(s\).\\
+  The models' new parameters are found by numerically maximizing the
+  respective likelihood, weighted by the responsibilities:
+  \begin{equation}\phantomsection\label{eq-maximization-step}{
+  \ln P(\mathbf{D}, \boldsymbol{\epsilon} \mid \boldsymbol{\theta}_m) = \sum_k^K \gamma(\epsilon^m_{k}) \big[ \ln \pi_m^{s_k} + \ln P \left( \mathbf{D}_k \mid \boldsymbol{\theta}_m \right) \big]
+  }\end{equation}
+\end{enumerate}
 
-In the E-step, we compute the expected value of the latent variables
-\(\boldsymbol{\epsilon}\), which represent the probabilities of a
-patient \(k\) originating from one of the models \(m\). Given the
-current estimates of \(\boldsymbol{\theta}\) and \(\boldsymbol{\pi}\),
-the expectation value \(\gamma(\epsilon_k^m)\) is calculated as:
-
-\begin{equation}\phantomsection\label{eq-expectation-step}{
-\mathbb{E}[\epsilon^m_k] = \gamma(\epsilon^m_{k}) = \frac{\pi_m^{s_k} P(\mathbf{X} = \boldsymbol{\xi}_k \mid \boldsymbol{\theta_m}, T_k)}{\sum_{n}^M \pi_n^{s_k} P(\mathbf{X} = \boldsymbol{\xi}_k \mid \boldsymbol{\theta_n}, T_k)}
-}\end{equation}
-
-\paragraph{Maximization Step}\label{maximization-step}
-
-In the M-step, the model parameters are updated by maximizing the
-complete data likelihood in equation
-equation~\ref{eq-complete-likelihood-explicite}. For better handling, we
-are optimizing the log likelihood:
-
-\begin{equation}\phantomsection\label{eq-maximization-step}{
-\ln P(\mathbf{D}, \boldsymbol{\epsilon} \mid \boldsymbol{\theta}, \boldsymbol{\pi}) = \sum_k^K \sum_{m}^M \gamma(\epsilon^m_{k}) \left( \ln \pi_m^{s_k} + \ln P(\mathbf{X} = \boldsymbol{\xi}_k \mid \boldsymbol{\theta_m}, T_k) \right)
-}\end{equation}
-
-In this step, the mixture parameters \(\pi_m^s\) and the model
-parameters \(\boldsymbol{\theta}_m\) can be optimized separately. The
-mixture parameters \(\pi_m^s\) can be updated analytically:
-
-\begin{equation}\phantomsection\label{eq-mixture-maximization}{
-\pi_m^s = \frac{1}{N^s} \sum_k^{N^s} \gamma(\epsilon^m_k)
-}\end{equation}
-
-On the other hand, the model parameters \(\boldsymbol{\theta_m}\) are
-optimized component-wise using an optimization algorithm, as an analytic
-solution is not feasible. The updated log-likelihood for each model is:
-
-\[
-\ln P(\mathbf{D}, \boldsymbol{\epsilon} \mid \boldsymbol{\theta}_m) = \sum_{k}^K \gamma(\epsilon^m_{k}) \ln P(\mathbf{X} = \boldsymbol{\xi}_i \mid \theta_m, T_k)
-\]
-
-By iterating these steps, the EM algorithm converges to a (local)
-maximum of the complete data likelihood
-equation~\ref{eq-complete-likelihood}.
+By iterating these steps, the EM algorithm is guaranteed to converge to
+a (local) maximum of the complete data likelihood
+(equation~\ref{eq-complete-likelihood}).
 

### 2024-10-17 manuscript.qmd
minor changes
extended

@@
 $$
-P \left( \mathbf{D} \mid \boldsymbol{\theta} \right) = \prod_k^K P \left( \mathbf{X}=\boldsymbol{\xi}_k \mid \boldsymbol{\theta}, T_k \right)
+P \left( \mathbf{D} \mid \boldsymbol{\theta} \right) = \prod_k^K P \left( \mathbf{X}_k=\boldsymbol{\xi}_i \mid \boldsymbol{\theta}, T_k \right)
 $$ {#eq-likelihood-HMM}
 
-In @eq-likelihood-HMM we compute the likelihood of observing patient the diagnosis of each patient $k$, i.e. t-stage $T_k$ and involvement $\xi_k$.
+In @eq-likelihood-HMM we compute the likelihood of observing patient the diagnosis of each patient $k$, i.e. t-stage $T_k$ and involvement  $\mathbf{X}_k=\boldsymbol{\xi}_i$.
 
@@
 - $\pi_m^{s_k}$ is the mixing coefficient for subsite $s_k$ (where patient $k$ has their tumor) and model $m$,
-- $P(\mathbf{D}_k \mid \boldsymbol{\theta}_m)$ is the likelihood of patient $k$'s diagnosis, given that it was generated by model $m$,
+- $P(\mathbf{D}_k \mid \boldsymbol{\theta}_m)$ is the likelihood of patient $k$'s diagnosis, i.e.
+The involvement state $\mathbf{X}_k = \boldsymbol{\xi}_i$ and t-stage $T_k$, given that it was generated by model $m$,
 - $\boldsymbol{\theta}_m$ represents the parameters of model $m$,

### 2024-10-21 _tex/manuscript.tex
Built site for gh-pages

@@
 \usepackage{centernot}
-\makeatletter
-\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
-\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
-\definecolor{quarto-callout-color}{HTML}{909090}
-\definecolor{quarto-callout-note-color}{HTML}{0758E5}
-\definecolor{quarto-callout-important-color}{HTML}{CC1914}
-\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
-\definecolor{quarto-callout-tip-color}{HTML}{00A047}
-\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
-\definecolor{quarto-callout-color-frame}{HTML}{acacac}
-\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
-\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
-\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
-\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
-\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
-\makeatother
 \makeatletter
@@
 
-\begin{tcolorbox}[enhanced jigsaw, breakable, colbacktitle=quarto-callout-note-color!10!white, left=2mm, colback=white, coltitle=black, arc=.35mm, bottomrule=.15mm, rightrule=.15mm, bottomtitle=1mm, opacitybacktitle=0.6, title=\textcolor{quarto-callout-note-color}{\faInfo}\hspace{0.5em}{TODO: Discuss if we need this}, titlerule=0mm, toptitle=1mm, toprule=.15mm, colframe=quarto-callout-note-color-frame, leftrule=.75mm, opacityback=0]
-
 In contrast, the \emph{incomplete data likelihood} marginalizes over all
@@
 from this shortcoming.
-
-\end{tcolorbox}
 
@@
 
+It is important to note that this mixture model algorithm diverges from
+the conventional formulation of mixture models as introduced in
+\citet{bishop_pattern_2006}. In Bishop's work, the mixture model is
+defined with a single set of mixture parameters \(\boldsymbol{\pi}\),
+implying that all data points are generated from the same mixture of
+models. In our approach, however, we categorize the data points
+(patients) into distinct cohorts (subsites), with each subsite
+characterized by its own set of mixture parameters. Consequently, we
+construct \(S\) separate mixture models, each possessing different
+mixture parameters while all models share the same underlying
+probability distributions, specifically the varying lymphatic spread
+models.
+
 \subsection{Expectation-Maximization (EM)
@@
 By iterating these steps, the EM algorithm is guaranteed to converge to
-a (local) maximum of the complete data likelihood
-(equation~\ref{eq-complete-likelihood}).
+a (local) maximum of the incomplete data likelihood
+(equation~\ref{eq-incomplete-likelihood}).
 

### 2024-10-22 _tex/manuscript.tex
Built site for gh-pages

@@
 For the analyses in this work, we used seven datasets from 5
-institutions resulting in 2741 patients in total.
+institutions resulting in 2437 patients in total.
 
@@
 \item
-  0 oropharyngeal patients from the University of Zurich in Switzerland
+  287 oropharyngeal patients from the University of Zurich in
+  Switzerland
 \item
-  0 oropharyngeal patients from the Centre Léon Bérard in France
+  263 oropharyngeal patients from the Centre Léon Bérard in France
 \item
-  0 oropharyngeal, larynx and oral cavity patients from the Inselspital
-  Bern in Switzerland
+  327 oropharyngeal, larynx and oral cavity patients from the
+  Inselspital Bern in Switzerland
 \item
-  0 oropharyngeal, larynx and oral cavity patients from the Centre Léon
-  Bérard in France
+  276 oropharyngeal, larynx and oral cavity patients from the Centre
+  Léon Bérard in France
 \item
-  0 oropharyngeal, larynx and oral cavity patients from the University
+  273 oropharyngeal, larynx and oral cavity patients from the University
   of Zurich in Switzerland
@@
 \item
-  979 hypopharynx, larynx and oral cavity patients from University
+  847 hypopharynx, larynx and oral cavity patients from University
   Medical Center Groningen (not yet public)
@@
 interactively explored on \href{https://lyprox.org}{LyProX}. For each
-patient the primary tumor subsite is reportd and each indicidual LNL is
+patient the primary tumor subsite is reported and each individual LNL is
 reported as either metastatic or healthy given the available diagnostic
@@
 patients. In this work we will stratify the tumor locations into
-different ICD codes which are depicted in figure~\ref{fig-subsites}.
+different ICD codes which are depicted in figure~\ref{fig-subsites}. For
+glottis tumors (C32.0), only t-stages T2,T3 and T4 have been taken into
+account as by definition glottis tumors are non-metastatic in the stages
+T0 and T1.
 

### 2024-12-06 manuscript.qmd
updated a figure

@@
 
+Patients with glottic laryngeal cancer (C32.0) at tumor stages T0 and T1 have been excluded from analysis because these stages, by definition, do not exhibit lymphatic involvement [need source from Panos].
 The datasets 1-4 are publicly available as CSV tables [@ludwig_multi-centric_2023;@ludwig_detailed_2022] and can be interactively explored on [LyProX](https://lyprox.org).
@@
 The patient data is further stratified in early t-stage (0-2) and late t-stage (3-4).
-The legend furter specifies the number of patients in each subsite.](figures/involvement_I_to_V_all_sites_t_staging.png){#fig-involvement}
+The legend furter specifies the number of patients in each subsite.
+For glottis (C32.0) early t-stage only includes T2.](figures/involvement_I_to_V_all_sites_t_staging.png){#fig-involvement}
 

### 2025-04-30 manuscript.qmd
temporal

--- 2024-10-11 manuscript.qmd
@@
 The datasets 1-4 are publicly available as CSV tables [@ludwig_multi-centric_2023;@ludwig_detailed_2022] and can be interactively explored on [LyProX](https://lyprox.org).
-For each patient the primary tumor subsite is reportd and each indicidual LNL is reported as either metastatic or healthy given the available diagnostic modalities, which include pathology after neck dissection in some patients.
+For each patient the primary tumor subsite is reported and each individual LNL is reported as either metastatic or healthy given the available diagnostic modalities, which include pathology after neck dissection in some patients.
 In this work we will stratify the tumor locations into different ICD codes which are depicted in @fig-subsites.
+For glottis tumors (C32.0), only t-stages T2,T3 and T4 have been taken into account as by definition glottis tumors are non-metastatic in the stages T0 and T1.
 
@@
 The patient data is further stratified in early t-stage (0-2) and late t-stage (3-4).
+For glottis patients t-stages 0 and 1 have been excluded from early t-stage.
 The legend furter specifies the number of patients in each subsite.](figures/involvement_I_to_V_all_sites_t_staging.png){#fig-involvement}
@@
 
-::: {.callout-note}
-
-## TODO: Discuss if we need this
-<!-- Not sure we can skip it. We want to maximize this w.r.t. the model parameters. -->
-
 In contrast, the _incomplete data likelihood_ marginalizes over all possible latent assignments to reflect the uncertainty about which model generated each patient’s data:
@@
 Thus, inferring the latent assignment is helpful, because the complete data (log-)likelihood (@eq-complete-likelihood) does not suffer from this shortcoming.
-
-:::
 
@@
 
+It is important to note that this mixture model algorithm diverges from the conventional formulation of mixture models as introduced in @bishop_pattern_2006.
+In Bishop's work, the mixture model is defined with a single set of mixture parameters $\boldsymbol{\pi}$, implying that all data points are generated from the same mixture of models.
+In our approach, however, we categorize the data points (patients) into distinct cohorts (subsites), with each subsite characterized by its own set of mixture parameters.
+Consequently, we construct $S$ separate mixture models, each possessing different mixture parameters while all models share the same underlying probability distributions, specifically the varying lymphatic spread models.
+
 ## Expectation-Maximization (EM) Algorithm
@@
 
-By iterating these steps, the EM algorithm is guaranteed to converge to a (local) maximum of the complete data likelihood (@eq-complete-likelihood).
+By iterating these steps, the EM algorithm is guaranteed to converge to a (local) maximum of the incomplete data likelihood (@eq-incomplete-likelihood).
 

### 2025-05-20 manuscript.qmd
added bootstrapping

--- 2024-12-06 manuscript.qmd
@@
 
-To actually find the mixing coefficients $\boldsymbol{\pi}$ and all models' parameters $\boldsymbol{\theta}_m$ that maximize @eq-incomplete-likelihood, we follow an iterative approach called _expectation-maximization_ or _EM-algorithm_.
+To find the mixing coefficients $\boldsymbol{\pi}$ and all models' parameters $\boldsymbol{\theta}_m$ that maximize @eq-incomplete-likelihood, we follow an iterative approach called _expectation-maximization_ or _EM-algorithm_.
 With arbitrarily initialized starting parameters, we alternate between the following two steps:
@@
 
+## Model Uncertainty estimation
+
+After fitting models using the Expectation-Maximization (EM) algorithm, we want to quantify the uncertainty associated with the estimated parameters and predictions.
+Understanding this uncertainty is critical, especially in applications where model outputs inform significant decisions.
+A robust estimation of uncertainty provides insights into the confidence we can place in the model and identifies potential limitations in its predictions.
+
+To estimate model uncertainty, we employ a Bayesian approach in which the model parameters are treated as random variables with prior distributions.
+Given the data \(\mathbf{D}\) and model parameters \(\boldsymbol{\theta}\), the posterior distribution is expressed as:
+
+$$
+P(\boldsymbol{\theta} \mid \mathbf{D}) \propto P(\mathbf{D} \mid \boldsymbol{\theta}) P(\boldsymbol{\theta})
+$$ {#eq-bayesian-posterior}
+
+where $P(\mathbf{D} \mid \boldsymbol{\theta})$ is the likelihood of the data, and $P(\boldsymbol{\theta})$ represents the prior distribution which we set to a uniform distribution.
+
+To address this, we adopt a simplified sampling strategy.
+In this approach, the mixture parameters $\boldsymbol{\pi}$ are fixed to the values obtained from the EM algorithm.
+We then sample only the model parameters $\boldsymbol{\theta}$ from the incomplete data likelihood:
+
+However, the AIP alorithm is computationally expensive for our application with a large latent space.
+Therefore, we employ a simpified parameter sampling where we fix the mixture parameters to the results obtained from the EM algorithm.
+We then only sample the model parameters $\mathbf{\theta}$ from the incomplete data likelihood @eq-incomplete-likelihood:
+
+$$
+P(\boldsymbol{\theta} \mid \mathbf{D}) \propto \prod_{k}^{K} \sum_{m}^{M} \pi_m^{s_k} P(\mathbf{D}_k \mid \boldsymbol{\theta}_m) P(\mathbf{\theta})
+$$
+
+While this method provides an efficient approximation, it underestimates model uncertainty.
+Fixing the latent variables $\mathbf{\epsilon}$ and mixture parameters $\boldsymbol{\pi}$ implies the assumption of no uncertainty in these components.
+Consequently, the resulting uncertainty estimates serve as a lower bound for the true model uncertainty.
+Despite its limitations, this approach balances computational efficiency with a practical estimation of uncertainty, making it a useful compromise.
+
 # Three component Mixture Model {#sec-3comp}
 
-We illustrate the methodology for a mixture model with M = 3 components, considering the ipsilateral involvement of LNLs I, II, III, IV, and V.
+We illustrate the methodology for a mixture model with M = 3 components, considering the ipsilateral involvement of LNLs I, II, III, IV, VI and VII.
 We include the ICD codes as subsites for oral cavity, hypopharynx and oropharynx.

### 2025-05-20 manuscript.qmd
new uncertainty estimation

--- 2025-04-30 manuscript.qmd
@@
 2.
-During the **M**maximization step, we find a new set of parameters that maximize the new expected complete data likelihood (@eq-expected-complete-likelihood).
+During the **M**aximization step, we find a new set of parameters that maximize the new expected complete data likelihood (@eq-expected-complete-likelihood).
 For the mixture coefficients, we can even find an analytic solution to the new maximum:
@@
 By iterating these steps, the EM algorithm is guaranteed to converge to a (local) maximum of the incomplete data likelihood (@eq-incomplete-likelihood).
+
+## Uncertainty estimation
+
+To estimate the uncertainty associated with both the model parameters and the resulting risk estimates, we employ a bootstrapping approach.
+Although more comprehensive sampling methods exist for the EM algorithm—such as the imputation-posterior (IP) algorithm, which samples from both model parameters and latent variables—we opt for a computationally more efficient alternative that still yields multiple estimates of all parameters.
+
+Our resampling procedure involves the following steps:
+
+1. Resample the observed data with replacement, maintaining the original number of samples per ICD code.
+
+2. Fit the model to each resampled dataset.
+
+3. Extract the model parameters and risk estimates from each fitted model.
+
+This method enables a straightforward uncertainty analysis while avoiding the computational cost of more intensive sampling techniques.
 

### 2025-05-20 manuscript.qmd

@@
 
-We illustrate the methodology for a mixture model with M = 3 components, considering the ipsilateral involvement of LNLs I, II, III, IV, and V.
+We illustrate the methodology for a mixture model with M = 3 components, considering the ipsilateral involvement of LNLs I, II, III, IV, VI and VII.
 We include the ICD codes as subsites for oral cavity, hypopharynx and oropharynx.

### 2025-05-21 manuscript.qmd
extended with first results

@@
 
-With @eq-marginalized-evolution we can compute the probability of a patient being in any state $\xi_i$. Therefore the likelihood of observing a the
-For model training we assume that the diagnoses in our data $\mathbf{D}$ we observe correspond to the hidden state $\mathbf{X}$ of the patient.
+With @eq-marginalized-evolution we can compute the probability of a patient being in any state $\xi_i$. 
+To train the model, we assume that the observed diagnoses in our dataset $\mathbf{D}$ directly reflect the underlying hidden states $\mathbf{X}$ of the patient.
 Thus, learning the model parameters corresponds to maximizing the probability of observing the dataset $\mathbf{D}$:
@@
 
-In @eq-likelihood-HMM we compute the likelihood of observing patient the diagnosis of each patient $k$, i.e. t-stage $T_k$ and involvement  $\mathbf{X}_k=\boldsymbol{\xi}_i$.
+In @eq-likelihood-HMM we compute the likelihood of observing the diagnosis of each patient $k$, i.e. t-stage $T_k$ and involvement  $\mathbf{X}_k=\boldsymbol{\xi}_i$.
+
+## Risk Estimation
+The goal of the HMM is to predict the risk of occult metastases in LNLs, given a patient's diagnosis.
+In this framework this corresponds to computing the probability of a hidden state $\mathbf{X}=\boldsymbol{\xi}_i$ given the observed diagnosis $\mathbf{Z}$ and the model parameters $\boldsymbol{\theta}$.
+With Baye's theorem we can compute this probability as follows:
+
+$$
+P(\mathbf{X} =\boldsymbol{\xi}_i \mid \mathbf{Z}, \boldsymbol{\theta}) = \frac{P(\mathbf{Z} \mid \mathbf{X}=\boldsymbol{\xi}_i, \boldsymbol{\theta}) P(\mathbf{X}_k=\boldsymbol{\xi}_i \mid \boldsymbol{\theta})}{P(\mathbf{Z} \mid \boldsymbol{\theta})}
+$$ {#eq-bayes}
+
+where $P(\mathbf{Z} \mid \boldsymbol{\theta})$ is the marginal likelihood of the data, which can be computed by summing over all possible hidden states $\boldsymbol{\xi}_i$:
+
+$$
+P(\mathbf{Z} \mid \boldsymbol{\theta}) = \sum_{\boldsymbol{\xi}_i} P(\mathbf{Z} \mid \boldsymbol{\xi}_i, \boldsymbol{\theta}) P(\boldsymbol{\xi}_i \mid \boldsymbol{\theta})
+$$ {#eq-marginal-likelihood}
+
+With this we can compute the probability for each hidden state and subsequently we can compute the probability for each LNL $v$ to be involved given the diagnosis $\mathbf{Z}$ and the model parameters $\boldsymbol{\theta}$ by summing over all possible hidden states $\boldsymbol{\xi}_{i,v}$ where the LNL $v$ is involved:
+
+$$
+P(X_v=1 \mid \mathbf{Z}, \boldsymbol{\theta}) = \sum_{\boldsymbol{\xi}_{i,v}} P(\mathbf{X} =\boldsymbol{\xi}_i \mid \mathbf{Z}, \boldsymbol{\theta})
+$$ {#eq-risk}
+
+As opposed to the model training, we do not assume anymore that the diagnoses correspond to the hidden states.
+Instead, we consider the sensitivity and specificity of the diagnostic modalities used to determine the involvement of LNLs.
+Enabling the uncertainty originating from diagnostic modalities to be considered in the model.
+Here we chose the sensitivity of 0.81 for clinical modalities corresponding to the literature value for imaging modalities [@de_bondt_detection_2007] and specificity of 1.0, indicating that all positively diagnosed LNLs are indeed involved.
 
@@
 
-## Uncertainty estimation
+## Uncertainty estimation {#sec-uncert}
 
@@
 
-# Three component Mixture Model {#sec-3comp}
+# Model Training Evaluation {#sec-model_train}
+In the following sections we will analyze the model training for three components first, considering Oral Cavity, Oropharynx and Hypopharynx data.
+Then we extend the model to four components, additionally including Larynx data.
+Both models will include LNLs I, II, III, IV and V.
+LNLs VI and VII are omitted for simplicity and since the prevalence for these LNLs never exceeds 11% (@fig-involvement), meaning that the risk of involvement is low in most cases.
 
-We illustrate the methodology for a mixture model with M = 3 components, considering the ipsilateral involvement of LNLs I, II, III, IV, VI and VII.
+## Three Component Mixture Model {#sec-3comp}
+
+We first evualuate the methodology for a mixture model with M = 3 components.
 We include the ICD codes as subsites for oral cavity, hypopharynx and oropharynx.
@@
 The spatial plot in @fig-3_simplex illustrates how the model assigns the three components to different tumor subsites.
-Component 0, located at the bottom right of the triangle, primarily characterizes oropharyngeal subsites.
-For instance, the base of tongue subsite (C01), which exhibits the highest involvement of LNL II, is fully assigned to this component.
-Similarly, subsite C10, which includes several oropharyngeal regions, is assigned roughly 50% to the oropharynx-like component, with the remaining mixture distributed across the other two components.
-
-Hypopharyngeal subsites, on the other hand, are fully assigned to Component 2, located at the top vertex of the triangle.
-Meanwhile, the gum subsite (C03), with predominant LNL I involvement, is entirely assigned to Component 1, situated at the bottom left.
-As subsites anatomically approach the oropharynx, their mixture coefficients for the oropharynx-like component increase.
+Component 0, located at the bottom right of the triangle, primarily characterizes hypopharyngeal subsites.
+Both hypopharyngeal subsites, piriform sinus (C12) and hypopharynx (C13), are strongly assigned to this component.
+Similarly, component 1, located at the bottom left, characterizes oral cavity subsites.
+For instance, the gum (C03), which has the highest LNL I involvement is fully assigned to this subsite.
+As subsites anatomically approach the oropharynx, their mixture coefficients for the oropharynx-like component, at the top of the triangle, increase.
 This is evident in the subsites C02 (tongue) and C05 (palate), which display a higher proportion of oropharyngeal influence in their mixture.
-These results conform well with the involvement patterns observed in the data.
+These results conform well with the involvement patterns observed in the data in figure @fig-involvement.
+The base of tongue subsite (C01), which exhibits the highest involvement of LNL II, is fully assigned to oropharynx-like component.
+Similarly, subsite C10, which includes several oropharyngeal regions, is assigned roughly 50% to the oropharynx-like component, with the remaining mixture distributed across the other two components, representing its large variation in spread patterns.
 
@@
 
-# Four component Mixture Model {#sec-4comp}
+To evaluate the uncertainty in the model parameters, the bootstrapping approach introduced in @sec-uncert has been applied.
+Specifically, we generated 200 bootstrap samples by resampling the observed data with replacement, maintaining the original number of samples per ICD code.
+For each bootstrap sample, the mixture model was re-fitted, and the resulting parameters were recorded.
+This process yielded a distribution of parameter estimates, allowing us to compute confidence intervals and assess the variability in the model's predictions.
+
+The results of the bootstrapping analysis are shown in @fig-simplex_uncert and @fig-3_comp_scatter_uncert.
+The figure illustrates the variability in the mixture coefficients $\boldsymbol{\pi}$ for each subsite in the same representation as @fig-3_simplex.
+The figure shows that the uncertainties match the continuity of the spread patterns.
+The base of tongue (C01) has most of its uncertainty along in between the oropharynx component and the hypopharynx component.
+Similarly, the hypopharynx subsite (C13) mostly presents uncertainty between these two components.
+The more general oropharnyx subsite (C10) shows more uncertainty in each components direction, due to its mixed spread pattern.
+The subsite for other pars of mouth (C03) however, only has uncertainty along the axis connecting the oropharyngeal component and the oral cavity componen, therefore never mixing with the hypopharynx-like component on the right.
+In figure @fig-3_comp_scatter_uncert the asymmetry of the uncertainty is better depicted.
+E.g.
+the base of tongue (C01) subsite and the hypopharynx (C13) subsite have strongly asymmetric uncertainties.
+Further, the plot better characterizes the uncertainty in the palate subsite (C05), where the mixture and uncertainty is mainly in the oropharynx-like and the oral cavity-like components if we center the uncertainty around the optimal values instead of the mean values as it is depicted in @fig-simplex_uncert.
+
+![Variability in mixture coefficients $\boldsymbol{\pi}$ across bootstrap samples for each subsite.
+Where the dots represent the optimal values and the ellipses represent the 68% percentile centered around the mean value.
+The 68% percentile are computed assuming a 2D-gaussian distribution.
+However, with the bounds of 0 and 1 the sampled mixture coefficients do not exactly follow a gaussian distribution.](figures/mixture_coefficients_uncert_simplex.png){#fig-simplex_uncert}
+
+![Variability in mixture coefficients $\boldsymbol{\pi}$ across bootstrap samples for each subsite.
+The dots represent the optimal values.
+The bars include 68% of all sample values centered around the optimal values.](figures/3_comp_mixture_coefficients_uncert_scatter.png){#fig-3_comp_scatter_uncert}
+
+Additionally, the bootstrapping approach enables us to quantify the uncertainty in the model predictions for each subsite and LNL.
+To benchmark the model we analyze the predicted and observed prevalences for LNLs I, II, III, IV, and V.
+These predictions are stratified by tumor subsite and t-stage, providing a comprehensive view of the model's performance and its associated uncertainty.
+
+In @fig-oral_cavity we present the predictions for the oral cavity subsites C03 and C05.
+Here we can clearly see in figure @fig-oral_cavity_C03_early and @fig-oral_cavity_C03_late that the difference in prevalence is substantial, even though both subsites are grouped in the same tumor location.
+The model predicts a higher prevalence for LNL II in subsite C05 (palate) compared to subsite C03 (gum).
+This is consistent with the observed data, where the prevalence of LNL II involvement is higher for the palate than for the gum.
+However, the model only predicts a slightly higher prevalence for LNL I in subsite C03 compared to subsite C05.
+This is not consistent with the observed data, where the prevalence of LNL I involvement is significantly higher for the gum than for the palate.
+Thus, showing a limitation in the model to properly split these different involvements.
+In late t-stages, depicted in figure @fig-oral_cavity_C03_late and @fig-oral_cavity_C05_late, the model predicts a higher prevalence for LNL II in subsite C05 compared to the early t-stage, but not fully capturing the increase in prevalence.
+For subsite C03 the LNL II involvement is predicted to increase  as well, capturing the increase in prevalence.
+However, for LNL I, while capturing the increase in prevalence, it does not optimally predict the prevalence, similarly to the early t-stage.
+For LNLs III, IV and V, the model consistently predicts the low prevalences observed in the data for both subsites.
+
+::: {#fig-oral_cavity layout-ncol=1}
+
+![](figures/3_predicted_prevalence_C03_early.png){#fig-oral_cavity_C03_early}
+
+![](figures/3_predicted_prevalence_C05_early.png){#fig-oral_cavity_C05_early}
+
+![](figures/3_predicted_prevalence_C03_late.png){#fig-oral_cavity_C03_late}
+
+![](figures/3_predicted_prevalence_C05_late.png){#fig-oral_cavity_C05_late}
+
+Observed prevalence (solid lines) and predicted prevalence (histograms) for subsites C03 and C05.
+Figures a and b show the prevalences for early t-stages.
+Figures c and d show the prevalences for late t-stages.
+:::
+
+In @fig-oropharynx we present the predictions for the oropharynx subsites C01 and C10. 
+In figures @fig-oropharynx_C01_early and @fig-oropharynx_C10_early we can see that the model predicts a higher prevalence for LNL II in subsite C01 (base of tongue) compared to subsite C10 (oropharynx).
+This is consistent with the observed data.
+In figures @fig-oropharynx_C01_late and @fig-oropharynx_C10_late we can see that the model precisely captures the increased prevalence in late t-stage for both subsites.
+LNLs I, IV and V are predicted to have low prevalences for both subsites, but the model does not perfectly fit the predicted prevalences to the observed prevalences.
+
+::: {#fig-oropharynx layout-ncol=1}
+
+![](figures/3_predicted_prevalence_C01_early.png){#fig-oropharynx_C01_early}
+
+![](figures/3_predicted_prevalence_C10_early.png){#fig-oropharynx_C10_early}
+
+![](figures/3_predicted_prevalence_C01_late.png){#fig-oropharynx_C01_late}
+
+![](figures/3_predicted_prevalence_C10_late.png){#fig-oropharynx_C10_late}
+
+Observed prevalence (solid lines) and predicted prevalence (histograms) for subsites C01 and C10.
+Figures a and b show the prevalences for early t-stages.
+Figures c and d show the prevalences for late t-stages.
+:::
+
+In summary, the mixture model successfully distinguishes between subsites, capturing the continuous variation in lymphatic spread patterns across tumor locations.
+However, the model's predictions are not always perfect.
+While it generally aligns with observed data, certain nuances, such as the higher prevalence of LNL I involvement in subsite C03 compared to C05, are not fully captured.
+
+## Four Component Mixture Model {#sec-4comp}
 
@@
 
-The component assignment is shown in @fig-4-matrix. Similarly to the 3-component model the different tumor locations are assigned to a one of the components....
-
-Here i probably should permute the components such that we have the same ordering as in the 3-component model.
+The component assignment is shown in @fig-4-matrix.
+Similarly to the 3-component model the different tumor locations are assigned to a one of the components.
+In this case, the new component is used to model laryngeal subsites.
 
@@
 add some analysis (level specific predictions and also comparison to single tumor location models.)
+
+# Model Predictions {#sec-model_pred}
+
+Given the model parameters, we can compute the risk of occult metastases given a patient's diagnosis.
+The model can be applied to predict the risk of involvement for each LNL, given the patient's t-stage and primary tumor subsite.
+
+The risk predictions are computed by applying the @eq-risk to each component and subsequently mixing the results using the mixture coefficients $\boldsymbol{\pi}$.
+The risk of involvement for each LNL $v$ in subsite $s$ is given by:
+$$
+P(X_v^s=1 \mid \mathbf{D}, \boldsymbol{\theta}) = \sum_{m=1}^M \pi_m^{s} P(X_v=1 \mid \mathbf{D}, \boldsymbol{\theta}_m)
+$$ {#eq-risk-mixing}
+
+where $s$ is the subsite and $\boldsymbol{\theta}_m$ are the parameters of model component $m$.
+
+For all estimates we will set a 5% threshold for the risk of involvement.
+This means that if the model predicts a risk of involvement of 5% or higher, we will consider the LNL to be at a "high" risk of involvement and should be irradiated.
+Consequently all LNLs with a risk of involvement below 5% will be considered to be at a "low" risk of involvement and should not be irradiated.
+This allows us to compare the model predictions with the current clinical guidelines for elective irradiation of LNLs.
+
+In the following sections we will analyze the model predictions for the four component model only and for selected clinical diagnoses.
+
+## Oral cavity
+
+Oral cavity tumors typically spread to LNLs I, II and III as shown in @fig-involvement.
+Consequently, the clinical guidelines recommend irradiating LNLs I, II, III for all tumors, irrespective of t-stage and even if there is no LNL involvement [@biau_selection_2019].
+
+We first analyze the risk predictions for the case where we diagnosed

### 2025-05-23 manuscript.qmd
Added further figures and added 4 comp analysis

@@
 The closer a subsite is to a vertex, the more it is assigned to the corresponding component, with component 0 on the bottom right, 1 on the bottom left and 2 on the top.
-The size of the marker (area) corresponds to the number of patients in each subsite.](figures/mixture_components_3_simplex_t_stages_new.png){#fig-3_simplex}
+The size of the marker (area) corresponds to the number of patients in each subsite.](figures/mixture_components_3_simplex.png){#fig-3_simplex}
 
@@
 Each row of the matrix corresponds to each ICD code.
-The collums represent the three different components](figures/mixture_components_3_t_stages_new.png){#fig-3-matrix}
+The collums represent the three different components](figures/mixture_components_3.png){#fig-3-matrix}
 :::
@@
 In this case, the new component is used to model laryngeal subsites.
+The uncertainty of the assignments is shown in @fig-4_comp_scatter_uncert.
+The most striking differences are in subsites C03 and C12, which were formerly assigned to a single component.
+Now, C03 mixes with the laryngeal component, while C12 is mixed with the laryngeal and the hypopharyngeal component.
+Another interesting observation is the mixture of the supraglottis subsite (C32.1) which is anatomically close to the hypopharynx.
+The model assigns a strong mixture of the larynx-like component and the hypopharynx-like component to this subsite.
+This is consistent with the observed data, where the supraglottis subsite has a high prevalence of LNL II and III involvement, which is also observed in the hypopharynx subsites.
+In contrast, the glottis subsite (C32.0) is strongly assigned to the laryngeal component, while the subsite C32.2 (subglottis) is assigned to the laryngeal component with little uncertainty.
+This is consistent with the observed data, where the glottis subsite has a much lower prevalence of involvement in any LNL.
+However, there is a strong t-stage dependency for this subsite.
+
+::: {layout="[30,75]" layout-valign="bottom"}
 
@@
 Each row of the matrix corresponds to each ICD code.
-The collums represent the three different components](figures/mixture_components_4_t_staging_add_larynx.png){#fig-4-matrix}
+The columns represent the four different components](figures/mixture_components_4.png){#fig-4-matrix}
 
-add some analysis (level specific predictions and also comparison to single tumor location models.)
+![Assignment of each subsite to each of the four components.
+The closer a subsite is to a vertex, the more it is assigned to the corresponding component, with component 0 on the bottom right, 1 on the bottom left, 2 on the top left and 3 on the top right.
+The size of the marker (area) corresponds to the number of patients in each subsite.](figures/4_comp_mixture_coefficients_uncert_scatter.png){#fig-4_comp_scatter_uncert}
+:::
+
+Similarly to the three component model, we can analyze the predicted and observed prevalences for LNLs I, II, III, IV and V.
+We will first revisit the oral cavity subsites C03 and C05.
+to check, whether the additional component improves the model predictions.
+In @fig-oral_cavity_4 the predictions for the oral cavity subsites C03 and C05 are shown.
+The model now predicts a higher prevalence for LNL I in subsite C03 (gum) compared to subsite C05 (palate) for both early and
+
+::: {#fig-oral_cavity_4 layout-ncol=1}
+
+![](figures/4_predicted_prevalence_C03_early.png){#fig-oral_cavity_C03_early}
+
+![](figures/4_predicted_prevalence_C05_early.png){#fig-oral_cavity_C05_early}
+
+![](figures/4_predicted_prevalence_C03_late.png){#fig-oral_cavity_C03_late}
+
+![](figures/4_predicted_prevalence_C05_late.png){#fig-oral_cavity_C05_late}
+
+Observed prevalence (solid lines) and predicted prevalence (histograms) for subsites C03 and C05.
+Figures a and b show the prevalences for early t-stages.
+Figures c and d show the prevalences for late t-stages.
+:::
+
+The predicted prevalences for for the laryngeal subsites C32.0 and C32.1 are shown in @fig-larynx_4.
+Here, the large difference between early and late t-stages is clearly visible.
+For subsite C32.0 (glottis) the model predicts a very low prevalence for LNL I, II and III in early t-stages, which is consistent with the observed data.
+For late t-stages, the model captures the increased involvement probabilities for LNLs II and III.
+For subsite C32.1 (supraglottis) the model overestimates the prevalence for LNLs II and III in early t-stages and slightly underestimates their prevalence in late t-stages.
+
+::: {#fig-larynx_4 layout-ncol=1}
+
+![](figures/4_predicted_prevalence_C32.0_early.png){#fig-oral_cavity_C32.0_early}
+
+![](figures/4_predicted_prevalence_C32.1_early.png){#fig-oral_cavity_C05_early}
+
+![](figures/4_predicted_prevalence_C32.0_late.png){#fig-oral_cavity_C03_late}
+
+![](figures/4_predicted_prevalence_C32.1_late.png){#fig-oral_cavity_C05_late}
+
+Observed prevalence (solid lines) and predicted prevalence (histograms) for subsites C32.0 and C32.1.
+Figures a and b show the prevalences for early t-stages.
+Figures c and d show the prevalences for late t-stages.
+:::
+
+The model parameters for the larynx-like component are shown in @fig-4_larynx_parameters.
+The parameter that governs the impact of t-stage $p_late$ is estimated around 1.
+Therefore, the model maximizes the increase of risk of involvement for late t-stages to capture the large difference in prevalence between early and late t-stages.
+
+![Model parameters for the larynx-like component.
+The optimal value and the 68% percentile around the optimal value are marked in red.
+The mean values and extreme values are indicated by the black bars.](figures/larynx_component_values.png){#fig-4_larynx_parameters}
 
@@
 We first analyze the risk predictions for the case where we diagnosed
+
+# Appendix
+Here I will probably add more figures such as all component parameter estimates and their uncertainties.

### 2025-05-23 _tex/manuscript.tex
Built site for gh-pages

@@
 
-The datasets 1-4 are publicly available as CSV tables
+Patients with glottic laryngeal cancer (C32.0) at tumor stages T0 and T1
+have been excluded from analysis because these stages, by definition, do
+not exhibit lymphatic involvement {[}need source from Panos{]}. The
+datasets 1-4 are publicly available as CSV tables
 \citep{ludwig_multi-centric_2023, ludwig_detailed_2022} and can be
 interactively explored on \href{https://lyprox.org}{LyProX}. For each
-patient the primary tumor subsite is reported and each individual LNL is
+patient the primary tumor subsite is reportd and each indicidual LNL is
 reported as either metastatic or healthy given the available diagnostic
@@
 patients. In this work we will stratify the tumor locations into
-different ICD codes which are depicted in figure~\ref{fig-subsites}. For
-glottis tumors (C32.0), only t-stages T2,T3 and T4 have been taken into
-account as by definition glottis tumors are non-metastatic in the stages
-T0 and T1.
+different ICD codes which are depicted in figure~\ref{fig-subsites}.
 
@@
 and late t-stage (3-4). The legend furter specifies the number of
-patients in each subsite.}
+patients in each subsite. For glottis (C32.0) early t-stage only
+includes T2.}
 
@@
 With equation~\ref{eq-marginalized-evolution} we can compute the
-probability of a patient being in any state \(\xi_i\). Therefore the
-likelihood of observing a the For model training we assume that the
-diagnoses in our data \(\mathbf{D}\) we observe correspond to the hidden
-state \(\mathbf{X}\) of the patient. Thus, learning the model parameters
+probability of a patient being in any state \(\xi_i\). To train the
+model, we assume that the observed diagnoses in our dataset
+\(\mathbf{D}\) directly reflect the underlying hidden states
+\(\mathbf{X}\) of the patient. Thus, learning the model parameters
 corresponds to maximizing the probability of observing the dataset
@@
 In equation~\ref{eq-likelihood-HMM} we compute the likelihood of
-observing patient the diagnosis of each patient \(k\), i.e.~t-stage
-\(T_k\) and involvement \(\mathbf{X}_k=\boldsymbol{\xi}_i\).
+observing the diagnosis of each patient \(k\), i.e.~t-stage \(T_k\) and
+involvement \(\mathbf{X}_k=\boldsymbol{\xi}_i\).
+
+\subsection{Risk Estimation}\label{risk-estimation}
+
+The goal of the HMM is to predict the risk of occult metastases in LNLs,
+given a patient's diagnosis. In this framework this corresponds to
+computing the probability of a hidden state
+\(\mathbf{X}=\boldsymbol{\xi}_i\) given the observed diagnosis
+\(\mathbf{Z}\) and the model parameters \(\boldsymbol{\theta}\). With
+Baye's theorem we can compute this probability as follows:
+
+\begin{equation}\phantomsection\label{eq-bayes}{
+P(\mathbf{X} =\boldsymbol{\xi}_i \mid \mathbf{Z}, \boldsymbol{\theta}) = \frac{P(\mathbf{Z} \mid \mathbf{X}=\boldsymbol{\xi}_i, \boldsymbol{\theta}) P(\mathbf{X}_k=\boldsymbol{\xi}_i \mid \boldsymbol{\theta})}{P(\mathbf{Z} \mid \boldsymbol{\theta})}
+}\end{equation}
+
+where \(P(\mathbf{Z} \mid \boldsymbol{\theta})\) is the marginal
+likelihood of the data, which can be computed by summing over all
+possible hidden states \(\boldsymbol{\xi}_i\):
+
+\begin{equation}\phantomsection\label{eq-marginal-likelihood}{
+P(\mathbf{Z} \mid \boldsymbol{\theta}) = \sum_{\boldsymbol{\xi}_i} P(\mathbf{Z} \mid \boldsymbol{\xi}_i, \boldsymbol{\theta}) P(\boldsymbol{\xi}_i \mid \boldsymbol{\theta})
+}\end{equation}
+
+With this we can compute the probability for each hidden state and
+subsequently we can compute the probability for each LNL \(v\) to be
+involved given the diagnosis \(\mathbf{Z}\) and the model parameters
+\(\boldsymbol{\theta}\) by summing over all possible hidden states
+\(\boldsymbol{\xi}_{i,v}\) where the LNL \(v\) is involved:
+
+\begin{equation}\phantomsection\label{eq-risk}{
+P(X_v=1 \mid \mathbf{Z}, \boldsymbol{\theta}) = \sum_{\boldsymbol{\xi}_{i,v}} P(\mathbf{X} =\boldsymbol{\xi}_i \mid \mathbf{Z}, \boldsymbol{\theta})
+}\end{equation}
+
+As opposed to the model training, we do not assume anymore that the
+diagnoses correspond to the hidden states. Instead, we consider the
+sensitivity and specificity of the diagnostic modalities used to
+determine the involvement of LNLs. Enabling the uncertainty originating
+from diagnostic modalities to be considered in the model. Here we chose
+the sensitivity of 0.81 for clinical modalities corresponding to the
+literature value for imaging modalities \citep{de_bondt_detection_2007}
+and specificity of 1.0, indicating that all positively diagnosed LNLs
+are indeed involved.
 
@@
   \(P(\mathbf{D}_k \mid \boldsymbol{\theta}_m)\) is the likelihood of
-  patient \(k\)'s diagnosis, given that it was generated by model \(m\),
+  patient \(k\)'s diagnosis, i.e.~The involvement state
+  \(\mathbf{X}_k = \boldsymbol{\xi}_i\) and t-stage \(T_k\), given that
+  it was generated by model \(m\),
 \item
@@
 
-To actually find the mixing coefficients \(\boldsymbol{\pi}\) and all
-models' parameters \(\boldsymbol{\theta}_m\) that maximize
+To find the mixing coefficients \(\boldsymbol{\pi}\) and all models'
+parameters \(\boldsymbol{\theta}_m\) that maximize
 equation~\ref{eq-incomplete-likelihood}, we follow an iterative approach
@@
 \item
-  During the \textbf{M}maximization step, we find a new set of
-  parameters that maximize the new expected complete data likelihood
+  During the \textbf{M}aximization step, we find a new set of parameters
+  that maximize the new expected complete data likelihood
   (equation~\ref{eq-expected-complete-likelihood}). For the mixture
@@
 
-\section{Three component Mixture Model}\label{sec-3comp}
-
-We illustrate the methodology for a mixture model with M = 3 components,
-considering the ipsilateral involvement of LNLs I, II, III, IV, and V.
-We include the ICD codes as subsites for oral cavity, hypopharynx and
-oropharynx. In figure~\ref{fig-convergence} the convergence of the
-negative log-likelihood and change in model parameters is depicted.
-After a random inizialization, the algorithm rapidly converges. The
-algorithm was stopped when the difference of log-likelihood between two
-iterations was below 0.01.
+\subsection{Uncertainty estimation}\label{sec-uncert}
+
+To estimate the uncertainty associated with both the model parameters
+and the resulting risk estimates, we employ a bootstrapping approach.
+Although more comprehensive sampling methods exist for the EM
+algorithm---such as the imputation-posterior (IP) algorithm, which
+samples from both model parameters and latent variables---we opt for a
+computationally more efficient alternative that still yields multiple
+estimates of all parameters.
+
+Our resampling procedure involves the following steps:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\item
+  Resample the observed data with replacement, maintaining the original
+  number of samples per ICD code.
+\item
+  Fit the model to each resampled dataset.
+\item
+  Extract the model parameters and risk estimates from each fitted
+  model.
+\end{enumerate}
+
+This method enables a straightforward uncertainty analysis while
+avoiding the computational cost of more intensive sampling techniques.
+
+\section{Model Training Evaluation}\label{sec-model_train}
+
+In the following sections we will analyze the model training for three
+components first, considering Oral Cavity, Oropharynx and Hypopharynx
+data. Then we extend the model to four components, additionally
+including Larynx data. Both models will include LNLs I, II, III, IV and
+V. LNLs VI and VII are omitted for simplicity and since the prevalence
+for these LNLs never exceeds 11\% (figure~\ref{fig-involvement}),
+meaning that the risk of involvement is low in most cases.
+
+\subsection{Three Component Mixture Model}\label{sec-3comp}
+
+We first evualuate the methodology for a mixture model with M = 3
+components. We include the ICD codes as subsites for oral cavity,
+hypopharynx and oropharynx. In figure~\ref{fig-convergence} the
+convergence of the negative log-likelihood and change in model
+parameters is depicted. After a random inizialization, the algorithm
+rapidly converges. The algorithm was stopped when the difference of
+log-likelihood between two iterations was below 0.01.
 
@@
 located at the bottom right of the triangle, primarily characterizes
-oropharyngeal subsites. For instance, the base of tongue subsite (C01),
-which exhibits the highest involvement of LNL II, is fully assigned to
-this component. Similarly, subsite C10, which includes several
-oropharyngeal regions, is assigned roughly 50\% to the oropharynx-like
-component, with the remaining mixture distributed across the other two
-components.
-
-Hypopharyngeal subsites, on the other hand, are fully assigned to
-Component 2, located at the top vertex of the triangle. Meanwhile, the
-gum subsite (C03), with predominant LNL I involvement, is entirely
-assigned to Component 1, situated at the bottom left. As subsites
+hypopharyngeal subsites. Both hypopharyngeal subsites, piriform sinus
+(C12) and hypopharynx (C13), are strongly assigned to this component.
+Similarly, component 1, located at the bottom left, characterizes oral
+cavity subsites. For instance, the gum (C03), which has the highest LNL
+I involvement is fully assigned to this subsite. As subsites
 anatomically approach the oropharynx, their mixture coefficients for the
-oropharynx-like component increase. This is evident in the subsites C02
-(tongue) and C05 (palate), which display a higher proportion of
-oropharyngeal influence in their mixture. These results conform well
-with the involvement patterns observed in the data.
+oropharynx-like component, at the top of the triangle, increase. This is
+evident in the subsites C02 (tongue) and C05 (palate), which display a
+higher proportion of oropharyngeal influence in their mixture. These
+results conform well with the involvement patterns observed in the data
+in figure figure~\ref{fig-involvement}. The base of tongue subsite
+(C01), which exhibits the highest involvement of LNL II, is fully
+assigned to oropharynx-like component. Similarly, subsite C10, which
+includes several oropharyngeal regions, is assigned roughly 50\% to the
+oropharynx-like component, with the remaining mixture distributed across
+the other two components, representing its large variation in spread
+patterns.
 
@@
 
-\includegraphics{figures/mixture_components_3_simplex_t_stages_new.png}
+\includegraphics{figures/mixture_components_3_simplex.png}
 
@@
 
-\includegraphics{figures/mixture_components_3_t_stages_new.png}
+\includegraphics{figures/mixture_components_3.png}
 
@@
 
-\section{Four component Mixture Model}\label{sec-4comp}
+To evaluate the uncertainty in the model parameters, the bootstrapping
+approach introduced in section~\ref{sec-uncert} has been applied.
+Specifically, we generated 200 bootstrap samples by resampling the
+observed data with replacement, maintaining the original number of
+samples per ICD code. For each bootstrap sample, the mixture model was
+re-fitted, and the resulting parameters were recorded. This process
+yielded a distribution of parameter estimates, allowing us to compute
+confidence intervals and assess the variability in the model's
+predictions.
+
+The results of the bootstrapping analysis are shown in
+figure~\ref{fig-simplex_uncert} and
+figure~\ref{fig-3_comp_scatter_uncert}. The figure illustrates the
+variability in the mixture coefficients \(\boldsymbol{\pi}\) for each
+subsite in the same representation as figure~\ref{fig-3_simplex}. The
+figure shows that the uncertainties match the continuity of the spread
+patterns. The base of tongue (C01) has most of its uncertainty along in
+between the oropharynx component and the hypopharynx component.
+Similarly, the hypopharynx subsite (C13) mostly presents uncertainty
+between these two components. The more general oropharnyx subsite (C10)
+shows more uncertainty in each components direction, due to its mixed
+spread pattern. The subsite for other pars of mouth (C03) however, only
+has uncertainty along the axis connecting the oropharyngeal component
+and the oral cavity componen, therefore never mixing with the
+hypopharynx-like component on the right. In figure
+figure~\ref{fig-3_comp_scatter_uncert} the asymmetry of the uncertainty
+is better depicted. E.g. the base of tongue (C01) subsite and the
+hypopharynx (C13) subsite have strongly asymmetric uncertainties.
+Further, the plot better characterizes the uncertainty in the palate
+subsite (C05), where the mixture and uncertainty is mainly in the
+oropharynx-like and the oral cavity-like components if we center the
+uncertainty around the optimal values instead of the mean values as it
+is depicted in figure~\ref{fig-simplex_uncert}.
+
+\begin{figure}
+
+\centering{
+
+\includegraphics{figures/mixture_coefficients_uncert_simplex.png}
+
+}
+
+\caption{\label{fig-simplex_uncert}Variability in mixture coefficients
+\(\boldsymbol{\pi}\) across bootstrap samples for each subsite. Where
+the dots represent the optimal values and the ellipses represent the
+68\% percentile centered around the mean value. The 68\% percentile are
+computed assuming a 2D-gaussian distribution. However, with the bounds
+of 0 and 1 the sampled mixture coefficients do not exactly follow a
+gaussian distribution.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\includegraphics{figures/3_comp_mixture_coefficients_uncert_scatter.png}
+
+}
+
+\caption{\label{fig-3_comp_scatter_uncert}Variability in mixture
+coefficients \(\boldsymbol{\pi}\) across bootstrap samples for each
+subsite. The dots represent the optimal values. The bars include 68\% of
+all sample values centered around the optimal values.}
+
+\end{figure}%
+
+Additionally, the bootstrapping approach enables us to quantify the
+uncertainty in the model predictions for each subsite and LNL. To
+benchmark the model we analyze the predicted and observed prevalences
+for LNLs I, II, III, IV, and V. These predictions are stratified by
+tumor subsite and t-stage, providing a comprehensive view of the model's
+performance and its associated uncertainty.
+
+In figure~\ref{fig-oral_cavity} we present the predictions for the oral
+cavity subsites C03 and C05. Here we can clearly see in figure
+figure~\ref{fig-oral_cavity_C03_early} and
+figure~\ref{fig-oral_cavity_C03_late} that the difference in prevalence
+is substantial, even though both subsites are grouped in the same tumor
+location. The model predicts a higher prevalence for LNL II in subsite
+C05 (palate) compared to subsite C03 (gum). This is consistent with the
+observed data, where the prevalence of LNL II involvement is higher for
+the palate than for the gum. However, the model only predicts a slightly
+higher prevalence for LNL I in subsite C03 compared to subsite C05. This
+is not consistent with the observed data, where the prevalence of LNL I
+involvement is significantly higher for the gum than for the palate.
+Thus, showing a limitation in the model to properly split these
+different involvements. In late t-stages, depicted in figure
+figure~\ref{fig-oral_cavity_C03_late} and
+figure~\ref{fig-oral_cavity_C05_late}, the model predicts a higher
+prevalence for LNL II in subsite C05 compared to the early t-stage, but
+not fully capturing the increase in prevalence. For subsite C03 the LNL
+II involvement is predicted to increase as well, capturing the increase
+in prevalence. However, for LNL I, while capturing the increase in
+prevalence, it does not optimally predict the prevalence, similarly to
+the early t-stage. For LNLs III, IV and V, the model consistently
+predicts the low prevalences observed in the data for both subsites.
+
+\begin{figure}
+
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\includegraphics{figures/3_predicted_prevalence_C03_early.png}
+
+}
+
+\subcaption{\label{fig-oral_cavity_C03_early}}
+
+\end{minipage}%
+\newline
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\includegraphics{figures/3_predicted_prevalence_C05_early.png}
+
+}
+
+\subcaption{\label{fig-oral_cavity_C05_early}}
+
+\end{minipage}%
+\newline
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\includegraphics{figures/3_predicted_prevalence_C03_late.png}
+
+}
+
+\subcaption{\label{fig-oral_cavity_C03_late}}
+
+\end{minipage}%
+\newline
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\includegraphics{figures/3_predicted_prevalence_C05_late.png}
+
+}
+
+\subcaption{\label{fig-oral_cavity_C05_late}}
+
+\end{minipage}%
+
+\caption{\label{fig-oral_cavity}Observed prevalence (solid lines) and
+predicted prevalence (histograms) for subsites C03 and C05. Figures a
+and b show the prevalences for early t-stages. Figures c and d show the
+prevalences for late t-stages.}
+
+\end{figure}%
+
+In figure~\ref{fig-oropharynx} we present the predictions for the
+oropharynx subsites C01 and C10. In figures
+figure~\ref{fig-oropharynx_C01_early} and
+figure~\ref{fig-oropharynx_C10_early} we can see that the model predicts
+a higher prevalence for LNL II in subsite C01 (base of tongue) compared
+to subsite C10 (oropharynx). This is consistent with the observed data.
+In figures figure~\ref{fig-oropharynx_C01_late} and
+figure~\ref{fig-oropharynx_C10_late} we can see that the model precisely
+captures the increased prevalence in late t-stage for both subsites.
+LNLs I, IV and V are predicted to have low prevalences for both
+subsites, but the model does not perfectly fit the predicted prevalences
+to the observed prevalences.
+
+\begin{figure}
+
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\includegraphics{figures/3_predicted_prevalence_C01_early.png}
+
+}
+
+\subcaption{\label{fig-oropharynx_C01_early}}
+
+\end{minipage}%
+\newline
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\includegraphics{figures/3_predicted_prevalence_C10_early.png}
+
+}
+
+\subcaption{\label{fig-oropharynx_C10_early}}
+
+\end{minipage}%
+\newline
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\includegraphics{figures/3_predicted_prevalence_C01_late.png}
+
+}
+
+\subcaption{\label{fig-oropharynx_C01_late}}
+
+\end{minipage}%
+\newline
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\includegraphics{figures/3_predicted_prevalence_C10_late.png}
+
+}
+
+\subcaption{\label{fig-oropharynx_C10_late}}
+
+\end{minipage}%
+
+\caption{\label{fig-oropharynx}Observed prevalence (solid lines) and
+predicted prevalence (histograms) for subsites C01 and C10. Figures a
+and b show the prevalences for early t-stages. Figures c and d show the
+prevalences for late t-stages.}
+
+\end{figure}%
+
+In summary, the mixture model successfully distinguishes between
+subsites, capturing the continuous variation in lymphatic spread
+patterns across tumor locations. However, the model's predictions are
+not always perfect. While it generally aligns with observed data,
+certain nuances, such as the higher prevalence of LNL I involvement in
+subsite C03 compared to C05, are not fully captured.
+
+\subsection{Four Component Mixture Model}\label{sec-4comp}
 
@@
 Similarly to the 3-component model the different tumor locations are
-assigned to a one of the components\ldots.
-
-Here i probably should permute the components such that we have the same
-ordering as in the 3-component model.
-
-\begin{figure}
-
-\centering{
-
-\includegraphics{figures/mixture_components_4_t_staging_add_larynx.png}
+assigned to a one of the components. In this case, the new component is
+used to model laryngeal subsites. The uncertainty of the assignments is
+shown in figure~\ref{fig-4_comp_scatter_uncert}. The most striking
+differences are in subsites C03 and C12, which were formerly assigned to
+a single component. Now, C03 mixes with the laryngeal component, while
+C12 is mixed with the laryngeal and the hypopharyngeal component.
+Another interesting observation is the mixture of the supraglottis
+subsite (C32.1) which is anatomically close to the hypopharynx. The
+model assigns a strong mixture of the larynx-like component and the
+hypopharynx-like component to this subsite. This is consistent with the
+observed data, where the supraglottis subsite has a high prevalence of
+LNL II and III involvement, which is also observed in the hypopharynx
+subsites. In contrast, the glottis subsite (C32.0) is strongly assigned
+to the laryngeal component, while the subsite C32.2 (subglottis) is
+assigned to the laryngeal component with little uncertainty. This is
+consistent with the observed data, where the glottis subsite has a much
+lower prevalence of involvement in any LNL. However, there is a strong
+t-stage dependency for this subsite.
+
+\begin{figure}
+
+\begin{minipage}{0.29\linewidth}
+
+\begin{figure}[H]
+
+\centering{
+
+\includegraphics{figures/mixture_components_4.png}
 
@@
 assignment. Each row of the matrix corresponds to each ICD code. The
-collums represent the three different components}
-
-\end{figure}%
-
-add some analysis (level specific predictions and also comparison to
-single tumor location models.)
+columns represent the four different components}
+
+\end{figure}%
+
+\end{minipage}%
+%
+\begin{minipage}{0.71\linewidth}
+
+\begin{figure}[H]
+
+\centering{
+
+\includegraphics{figures/4_comp_mixture_coefficients_uncert_scatter.png}
+
+}
+
+\caption{\label{fig-4_comp_scatter_uncert}Assignment of each subsite to
+each of the four components. The closer a subsite is to a vertex, the
+more it is assigned to the corresponding component, with component 0 on
+the bottom right, 1 on the bottom left, 2 on the top left and 3 on the
+top right. The size of the marker (area) corresponds to the number of
+patients in each subsite.}
+
+\end{figure}%
+
+\end{minipage}%
+
+\end{figure}%
+
+Similarly to the three component model, we can analyze the predicted and
+observed prevalences for LNLs I, II, III, IV and V. We will first
+revisit the oral cavity subsites C03 and C05. to check, whether the
+additional component improves the model predictions. In
+figure~\ref{fig-oral_cavity_4} the predictions for the oral cavity
+subsites C03 and C05 are shown. The model now predicts a higher
+prevalence for LNL I in subsite C03 (gum) compared to subsite C05
+(palate) for both early and
+
+\begin{figure}
+
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\includegraphics{figures/4_predicted_prevalence_C03_early.png}
+
+}
+
+\subcaption{\label{fig-oral_cavity_C03_early}}
+
+\end{minipage}%
+\newline
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\includegraphics{figures/4_predicted_prevalence_C05_early.png}
+
+}
+
+\subcaption{\label{fig-oral_cavity_C05_early}}
+
+\end{minipage}%
+\newline
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\includegraphics{figures/4_predicted_prevalence_C03_late.png}
+
+}
+
+\subcaption{\label{fig-oral_cavity_C03_late}}
+
+\end{minipage}%
+\newline
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\includegraphics{figures/4_predicted_prevalence_C05_late.png}
+
+}
+
+\subcaption{\label{fig-oral_cavity_C05_late}}
+
+\end{minipage}%
+
+\caption{\label{fig-oral_cavity_4}Observed prevalence (solid lines) and
+predicted prevalence (histograms) for subsites C03 and C05. Figures a
+and b show the prevalences for early t-stages. Figures c and d show the
+prevalences for late t-stages.}
+
+\end{figure}%
+
+The predicted prevalences for for the laryngeal subsites C32.0 and C32.1
+are shown in figure~\ref{fig-larynx_4}. Here, the large difference
+between early and late t-stages is clearly visible. For subsite C32.0
+(glottis) the model predicts a very low prevalence for LNL I, II and III
+in early t-stages, which is consistent with the observed data. For late
+t-stages, the model captures the increased involvement probabilities for
+LNLs II and III. For subsite C32.1 (supraglottis) the model
+overestimates the prevalence for LNLs II and III in early t-stages and
+slightly underestimates their prevalence in late t-stages.
+
+\begin{figure}
+
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\includegraphics{figures/4_predicted_prevalence_C32.0_early.png}
+
+}
+
+\subcaption{\label{fig-oral_cavity_C32.0_early}}
+
+\end{minipage}%
+\newline
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\includegraphics{figures/4_predicted_prevalence_C32.1_early.png}
+
+}
+
+\subcaption{\label{fig-oral_cavity_C05_early}}
+
+\end{minipage}%
+\newline
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\includegraphics{figures/4_predicted_prevalence_C32.0_late.png}
+
+}
+
+\subcaption{\label{fig-oral_cavity_C03_late}}
+
+\end{minipage}%
+\newline
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\includegraphics{figures/4_predicted_prevalence_C32.1_late.png}
+
+}
+
+\subcaption{\label{fig-oral_cavity_C05_late}}
+
+\end{minipage}%
+
+\caption{\label{fig-larynx_4}Observed prevalence (solid lines) and
+predicted prevalence (histograms) for subsites C32.0 and C32.1. Figures
+a and b show the prevalences for early t-stages. Figures c and d show
+the prevalences for late t-stages.}
+
+\end{figure}%
+
+The model parameters for the larynx-like component are shown in
+figure~\ref{fig-4_larynx_parameters}. The parameter that governs the
+impact of t-stage \(p_late\) is estimated around 1. Therefore, the model
+maximizes the increase of risk of involvement for late t-stages to
+capture the large difference in prevalence between early and late
+t-stages.
+
+\begin{figure}
+
+\centering{
+
+\includegraphics{figures/larynx_component_values.png}
+
+}
+
+\caption{\label{fig-4_larynx_parameters}Model parameters for the
+larynx-like component. The optimal value and the 68\% percentile around
+the optimal value are marked in red. The mean values and extreme values
+are indicated by the black bars.}
+
+\end{figure}%
+
+\section{Model Predictions}\label{sec-model_pred}
+
+Given the model parameters, we can compute the risk of occult metastases
+given a patient's diagnosis. The model can be applied to predict the
+risk of involvement for each LNL, given the patient's t-stage and
+primary tumor subsite.
+
+The risk predictions are computed by applying the equation~\ref{eq-risk}
+to each component and subsequently mixing the results using the mixture
+coefficients \(\boldsymbol{\pi}\). The risk of involvement for each LNL
+\(v\) in subsite \(s\) is given by:
+\begin{equation}\phantomsection\label{eq-risk-mixing}{
+P(X_v^s=1 \mid \mathbf{D}, \boldsymbol{\theta}) = \sum_{m=1}^M \pi_m^{s} P(X_v=1 \mid \mathbf{D}, \boldsymbol{\theta}_m)
+}\end{equation}
+
+where \(s\) is the subsite and \(\boldsymbol{\theta}_m\) are the
+parameters of model component \(m\).
+
+For all estimates we will set a 5\% threshold for the risk of
+involvement. This means that if the model predicts a risk of involvement
+of 5\% or higher, we will consider the LNL to be at a ``high'' risk of
+involvement and should be irradiated. Consequently all LNLs with a risk
+of involvement below 5\% will be considered to be at a ``low'' risk of
+involvement and should not be irradiated. This allows us to compare the
+model predictions with the current clinical guidelines for elective
+irradiation of LNLs.
+
+In the following sections we will analyze the model predictions for the
+four component model only and for selected clinical diagnoses.
+
+\subsection{Oral cavity}\label{oral-cavity}
+
+Oral cavity tumors typically spread to LNLs I, II and III as shown in
+figure~\ref{fig-involvement}. Consequently, the clinical guidelines
+recommend irradiating LNLs I, II, III for all tumors, irrespective of
+t-stage and even if there is no LNL involvement
+\citep{biau_selection_2019}.
+
+We first analyze the risk predictions for the case where we diagnosed
+
+\section{Appendix}\label{appendix}
+
+Here I will probably add more figures such as all component parameter
+estimates and their uncertainties.
 

### 2025-06-02 manuscript.qmd
added first risk discussion

@@
 
-We first analyze the risk predictions for the case where we diagnosed
+We first analyze the risk predictions for the case where we clinically diagnose no positive LNLs
+
+::: {#fig-oral_cavity_early_no_pos layout-ncol=2}
+
+![](figures/plot_C02_early_no_positive_diagnoses.png){#fig-oral_cavity_no_pos_C02_early}
+
+![](figures/plot_C03_early_no_positive_diagnoses.png){#fig-oral_cavity_no_pos_C03_early}
+
+![](figures/plot_C04_early_no_positive_diagnoses.png){#fig-oral_cavity_no_pos_C04_early}
+
+![](figures/plot_C05_early_no_positive_diagnoses.png){#fig-oral_cavity_no_pos_C05_early}
+
+![](figures/plot_C06_early_no_positive_diagnoses.png){#fig-oral_cavity_no_pos_C06_early}
+
+Probability of occult metastases in LNLs I, II, III, IV and V for early t-stages in oral cavity subsites C02 (tongue), C03 (gum), C04 (floor of mouth), C05 (palate) and C06 (other parts of mouth).
+The red line marks the 5% threshold for high risk of involvement.
+:::
+
+In @fig-oral_cavity_early_no_pos we can see that the model predicts different risks of involvement for each subsite.
+For subsite C03 (gum), predicts a higher risk of involvement for LNL I.
+In subsites C02 (tongue) and C05 (palate), the model predicts an above threshold risk of involvement for LNL II.
+While for subsites C02 (tongue) and C06 (other parts of mouth) all LNLs stay below the 5% threshold.
+Thus, the model would propose three different treatment plans for these subsites, with C03 (gum) LNL I, C02 (tongue) LNL II and C05 (palate) LNL II being irradiated, while C04 (floor of mouth) and C06 (other parts of mouth) would not need elective irradiation at all, which is a major deviation from the clinical guidelines which irradiate LNLs I, II and III for all oral cavity subsites.
 

### 2025-06-02 _tex/manuscript.tex
Built site for gh-pages

@@
 
-We first analyze the risk predictions for the case where we diagnosed
+We first analyze the risk predictions for the case where we clinically
+diagnose no positive LNLs
+
+\begin{figure}
+
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\includegraphics{figures/plot_C02_early_no_positive_diagnoses.png}
+
+}
+
+\subcaption{\label{fig-oral_cavity_no_pos_C02_early}}
+
+\end{minipage}%
+%
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\includegraphics{figures/plot_C03_early_no_positive_diagnoses.png}
+
+}
+
+\subcaption{\label{fig-oral_cavity_no_pos_C03_early}}
+
+\end{minipage}%
+\newline
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\includegraphics{figures/plot_C04_early_no_positive_diagnoses.png}
+
+}
+
+\subcaption{\label{fig-oral_cavity_no_pos_C04_early}}
+
+\end{minipage}%
+%
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\includegraphics{figures/plot_C05_early_no_positive_diagnoses.png}
+
+}
+
+\subcaption{\label{fig-oral_cavity_no_pos_C05_early}}
+
+\end{minipage}%
+\newline
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\includegraphics{figures/plot_C06_early_no_positive_diagnoses.png}
+
+}
+
+\subcaption{\label{fig-oral_cavity_no_pos_C06_early}}
+
+\end{minipage}%
+
+\caption{\label{fig-oral_cavity_early_no_pos}Probability of occult
+metastases in LNLs I, II, III, IV and V for early t-stages in oral
+cavity subsites C02 (tongue), C03 (gum), C04 (floor of mouth), C05
+(palate) and C06 (other parts of mouth). The red line marks the 5\%
+threshold for high risk of involvement.}
+
+\end{figure}%
+
+In figure~\ref{fig-oral_cavity_early_no_pos} we can see that the model
+predicts different risks of involvement for each subsite. For subsite
+C03 (gum), predicts a higher risk of involvement for LNL I. In subsites
+C02 (tongue) and C05 (palate), the model predicts an above threshold
+risk of involvement for LNL II. While for subsites C02 (tongue) and C06
+(other parts of mouth) all LNLs stay below the 5\% threshold. Thus, the
+model would propose three different treatment plans for these subsites,
+with C03 (gum) LNL I, C02 (tongue) LNL II and C05 (palate) LNL II being
+irradiated, while C04 (floor of mouth) and C06 (other parts of mouth)
+would not need elective irradiation at all, which is a major deviation
+from the clinical guidelines which irradiate LNLs I, II and III for all
+oral cavity subsites.
 
